---
title: "Parry_2023_all_joint_clusters"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

## Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Matrix)
library(igraph)
library(pagoda2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(conos)
library(stringr)
```

```{r}
# Check root path
here::here()
```

## Define paths

```{r}
path_to_sample_metadata <- here::here("metadata/metadata.csv")
path_to_obj <- here::here("reanalysis/R_objects/pagoda2_objects/")
path_to_cell_metadata <- here::here("reanalysis/metadata/pagoda2_cell_metadata/")
path_to_save_obj <- here::here("reanalysis/R_objects/1-all_cell_conos.rds")
path_to_save_cell_metadata <- here::here("reanalysis/metadata/1-all_cell_metadata.rds")
```

## Load data

```{r}
# List the file names
obj_files_to_load <- list.files(path_to_obj) # list object file names
cell_meta_files_to_load <- list.files(path_to_cell_metadata) # list cell metadta file names
```

```{r}
# Load the objects and collect in a list
gem_ids <- obj_files_to_load %>%
  str_remove("pagoda2_") %>%
  str_remove(".rds")
obj_files_to_load2 <- str_c(path_to_obj , obj_files_to_load, sep = "")
panel.preprocessed <- purrr::map(obj_files_to_load2, function(path) {
  pagoda2_obj <- readRDS(path)
  pagoda2_obj
})
names(panel.preprocessed) <- gem_ids
```

```{r}
# Load the cell metadata from all pagoda2 objects
gem_ids2 <- cell_meta_files_to_load %>%
  str_remove("cellmetadata_") %>%
  str_remove(".rds")
cell_meta_files_to_load2 <- str_c(path_to_cell_metadata , cell_meta_files_to_load, sep = "")
cell_metada.list <- purrr::map(cell_meta_files_to_load2, function(path) {
  pagoda2_cell_metadata <- readRDS(path)
  pagoda2_cell_metadata
})
names(cell_metada.list) <- gem_ids2
```

```{r}
# Combine all cell metadata together
cell_metadata_combined <- bind_rows(cell_metada.list) # rbind
```

## Integrate data with conos

```{r}
# Check barcode names for the first object in the list
head(rownames(panel.preprocessed[[1]]$counts))
```

```{r}
# Check duplicated barcode names
any(duplicated(unlist(lapply(panel.preprocessed, function(obj) rownames(obj$counts)))))
```

```{r}
# Create a Conos object
con <- Conos$new(panel.preprocessed, n.cores=1)
```

### Build the joint graph

```{r, results='hide'}
# Build the joint graph
con$buildGraph(k=30, k.self=5, space='PCA', ncomps=30, n.odgenes=2000, matching.method='mNN', metric='angular', score.component.variance=TRUE, verbose=TRUE)
```

### Graph-based clustering

```{r}
# Clustering
con$findCommunities(method=leiden.community, resolution=0.5)
```

```{r}
# Add clustering results to the cell metadata
cell_metadata_combined$leiden_cluster <- con$clusters$leiden$groups[rownames(cell_metadata_combined)] 

# Rename in some columns
colnames(cell_metadata_combined)[colnames(cell_metadata_combined) == "sra_id_col"] <- "sra_id"
colnames(cell_metadata_combined)[colnames(cell_metadata_combined) == "subject_id_col"] <- "subject_id"
colnames(cell_metadata_combined)[colnames(cell_metadata_combined) == "histologic_type_col"] <- "histologic_type"
```

### Generate embedding

```{r}
# Embedding
con$embedGraph(method='largeVis')
```

## Visualisation

### Clustering

```{r}
# Clustering
con$plotGraph(alpha=0.1, title="Leiden clusters")
```

### Subjects

```{r}
# Plot based on subjects
subject_id.vec <- setNames(cell_metadata_combined$subject_id, row.names(cell_metadata_combined))
con$plotGraph(alpha=0.1, 
              groups=subject_id.vec, 
              title="Subject IDs",
              show.legend = TRUE, 
              mark.groups=FALSE) 
```

### Histological types

```{r}
# Plot based on histologic origin
histo.vec <- setNames(cell_metadata_combined$histologic_type, row.names(cell_metadata_combined))
con$plotGraph(alpha=0.1, 
              groups=histo.vec, 
              title="Histological types",
              show.legend = TRUE) 
```

### Gene expression

#### B cell biomerkers

```{r, fig.height=4, fig.width=15}
# B cell biomerkers
CD19.plt <- con$plotGraph(gene="CD19", title="CD19")
CD5.plt <- con$plotGraph(gene="CD5", title="CD5")
CD79A.plt <- con$plotGraph(gene="CD79A", title="CD79A")

CD19.plt + CD5.plt + CD79A.plt
```

#### Erythroid biomarkers

```{r, fig.height=4, fig.width=15}
# Erythroid biomarkers
HBM.plt <- con$plotGraph(gene="HBM", title="HBM")
HBB.plt <- con$plotGraph(gene="HBB", title="HBB")
HBA1.plt <- con$plotGraph(gene="HBA1", title="HBA1")

HBM.plt + HBB.plt + HBA1.plt
```

#### T cell biomarkers

```{r, fig.height=4, fig.width=5}
# All T cells
con$plotGraph(gene="CD3D", title="CD3D")
```

```{r, fig.height=4, fig.width=5}
# CD8 T cells
con$plotGraph(gene="CD8A", title="CD8A")
```

```{r, fig.height=5, fig.width=12}
# CD4 T cells
CD4.plt <- con$plotGraph(gene="CD4", title="CD4")
IL7R.plt <- con$plotGraph(gene="IL7R", title="IL7R")

CD4.plt + IL7R.plt
```

#### NK cell biomarkers

```{r, fig.height=4, fig.width=15}
# NK cell biomarkers
NKG7.plt <- con$plotGraph(gene="NKG7", title="NKG7")
NCAM1.plt <- con$plotGraph(gene="NCAM1", title="NCAM1 (CD56)")
FCGR3A.plt <- con$plotGraph(gene="FCGR3A", title="FCGR3A (CD16)")

NKG7.plt + NCAM1.plt + FCGR3A.plt
```

#### Myeloid and monocyte biomarkers

Biomarkers for myeloid cells, such as CD11c (ITGAX), CD13 (ANPEP), CD14, CD15 (FUT4), CD64(FCGR1A), can be used to identify myeloid cells in the plot.

```{r, fig.height=4, fig.width=15}
# Myeloid and monocyte biomarkers
CD14.plt <- con$plotGraph(gene="CD14", title="CD14")
ITGAX.plt <- con$plotGraph(gene="ITGAX", title="ITGAX (CD11c)")
FCGR1A.plt <- con$plotGraph(gene="FCGR1A", title="FCGR1A (CD64)")

CD14.plt + ITGAX.plt + FCGR1A.plt
```

```{r}
# Use for find gene name of interest
grep("CD15", colnames(con$samples[[1]]$counts), value = TRUE)
```

## Cell type annotation

```{r}
# Add cell types to the cell metadata
cell_metadata_combined <- cell_metadata_combined %>%
  mutate(cell_type = case_when(
    leiden_cluster == "1" ~ "B cell",
    leiden_cluster == "2" ~ "CD8 T cell",
    leiden_cluster == "3" ~ "CD4 T cell",
    leiden_cluster == "4" ~ "B cell",
    leiden_cluster == "5" ~ "Myeloid cell",
    leiden_cluster == "6" ~ "Erythroid cell",
    leiden_cluster == "7" ~ "NK cell"
  ))
```

```{r}
# Visualise the annotation
cell_anno.vec <- setNames(cell_metadata_combined$cell_type, row.names(cell_metadata_combined))
con$plotGraph(alpha=0.1, 
              groups=cell_anno.vec, 
              title="Cell annotation",
              show.legend = TRUE) 
```

## Save the results

```{r, eval=FALSE}
saveRDS(con, path_to_save_obj)
saveRDS(cell_metadata_combined, path_to_save_cell_metadata)
```






