---
title: "Consistency_bulk_vs_sc_RNAseq"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This code we will evaluate the consistency of gene expression and functional enrichment analysis from our tumour integrated results and bulk transcriptomic data from both human and mouse studies. For human bulk transcriptomes, we used the data from [Nadeu et al. (2022)](https://www.nature.com/articles/s41591-022-01927-8) and [Parry et al. (2023a)](https://www.nature.com/articles/s41591-022-02113-6) while for mouse bulk transcriptomes, the data came from [Hing et al. (2023)](https://www.nature.com/articles/s41467-022-35778-1), [Ten Hacken et. al. (2023)](https://aacrjournals.org/bloodcancerdiscov/article/4/2/150/718388/In-Vivo-Modeling-of-CLL-Transformation-to-Richter), [Iyer et al. (2024)](https://www.science.org/doi/10.1126/scitranslmed.adg7915), and [Rompietti et al. (2025)](https://www.nature.com/articles/s41375-025-02557-y)

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Import the relavant packages
library(UpSetR)
library(ggvenn)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(lsa)
library(pheatmap)
library(viridis)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save files
path_to_bulk_analysis <- here::here("RNAseq/systematic_review")
path_to_single_cell_analysis <- here::here("scRNAseq/systematic_review")
path_to_sc_integrated_data <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs")
path_to_save_figure <- here::here("figures")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# List file names of all studies
study.lst <- c("1-Hacken_2023_GSE186137", "1-Nadeu_2022", "2-Iyer_2024_GSE263238", "2-Parry_2023a", 
               "3-Rompietti_2025", "5-Hing_2023_GSE183427")
study.lst
```

```{r}
# Create names for list later
names.lst <- sub(
  "^\\d+-([A-Za-z]+)_([0-9]{4}[A-Za-z]?).*$",  # capture name and year
  "\\1(\\2)",                         # output as Name(Year)
  study.lst
)

names.lst[length(names.lst) + 1] <- "Integrated scRNA-seq"
names.lst
```

```{r}
# Extract only name of studies
study_names <- sub("^\\d+-([^_0-9]+).*", "\\1", study.lst)
study_names
```

## Data from gene expression using DESeq2

```{r, message=FALSE, warning=FALSE}
# Load expression results in each study and collect in a list
path_to_dea_download <- str_c(path_to_bulk_analysis, "/", study.lst, "/expression_data/", study_names, "_dea.rds")
path_to_dea_download[3] <- "/mnt/hp-storage/users/pakorns/RNAseq/systematic_review/2-Iyer_2024_GSE263238/expression_data/Iyer_RT_vs_CLL_dea.rds" # Iyer was used in different file name
path_to_dea_download[6] <- "/mnt/hp-storage/users/pakorns/RNAseq/systematic_review/5-Hing_2023_GSE183427/expression_data/Hing_GSE183427_dea.rds" # Hing was used in different file name
path_to_dea_download[length(path_to_dea_download) + 1] <- str_c(path_to_sc_integrated_data, "/tumour_deseq2_dea.rds") # add path for integrated dea

dea.lst <- purrr::pmap(
  list(path_to_dea_download), 
  function(path){
    expr.mat <- readRDS(path)
    return(expr.mat)
  }
)

names(dea.lst) <- names.lst

# Convert some DESeq results to data frames
dea.lst$`Nadeu(2022)` <- as.data.frame(dea.lst$`Nadeu(2022)`)
dea.lst$`Parry(2023a)` <- as.data.frame(dea.lst$`Parry(2023a)`)
```

```{r}
# Clean the dea especially for column gene names to be consistent across all studies
names(dea.lst$`Hacken(2023)`)[names(dea.lst$`Hacken(2023)`) == "Gene.name"] <- "gene_name"
names(dea.lst$`Iyer(2024)`)[names(dea.lst$`Iyer(2024)`) == "Gene.name"] <- "gene_name"
names(dea.lst$`Rompietti(2025)`)[names(dea.lst$`Rompietti(2025)`) == "Gene.name"] <- "gene_name"
names(dea.lst$`Hing(2023)`)[names(dea.lst$`Hing(2023)`) == "Gene.name"] <- "gene_name"
dea.lst$`Integrated scRNA-seq`$gene_name <- rownames(dea.lst$`Integrated scRNA-seq`)
```

### Upregulated DEGs list

```{r}
# Extract adjusted significant upregulated genes and collect in a list
sig_upreg_gene.lst <- purrr::pmap(
  list(dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange > 0.25 & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)

# Drop empty ones
sig_upreg_gene.lst <- sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

### Downregulated DEGs list

```{r}
# Extract adjusted significant downregulated genes and collect in a list
sig_downreg_gene.lst <- purrr::pmap(
  list(dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange < (-0.25) & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)

# Drop empty ones
sig_downreg_gene.lst <- sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

## Data from GSEA

```{r}
# Hallmark
path_to_gsea_download <- str_c(path_to_bulk_analysis, "/", study.lst, "/enrichment_data/", study_names)
path_to_gsea_download[6] <- "/mnt/hp-storage/users/pakorns/RNAseq/systematic_review/5-Hing_2023_GSE183427/enrichment_data/Hing_GSE183427" # Hing was used in different file name
path_to_gsea_download[length(path_to_gsea_download) + 1] <- str_c(path_to_sc_integrated_data, "/tumour_deseq2") # add path for integrated gsea
path_to_hm <- str_c(path_to_gsea_download, "_hm_gsea.rds")
  
hm_gsea.lst <- purrr::pmap(
  list(path_to_hm), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)

names(hm_gsea.lst) <- names.lst

# GO:BP
path_to_go <- str_c(path_to_gsea_download, "_go_gsea.rds")
  
go_gsea.lst <- purrr::pmap(
  list(path_to_go), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)

names(go_gsea.lst) <- names.lst

# KEGG
path_to_kegg <- str_c(path_to_gsea_download, "_kegg_gsea.rds")
  
kegg_gsea.lst <- purrr::pmap(
  list(path_to_kegg), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)

names(kegg_gsea.lst) <- names.lst

# Reactome
path_to_reactome <- str_c(path_to_gsea_download, "_reactome_gsea.rds")
  
reactome_gsea.lst <- purrr::pmap(
  list(path_to_reactome), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)

names(reactome_gsea.lst) <- names.lst

# WikiPath
path_to_wikipath <- str_c(path_to_gsea_download, "_wikipath_gsea.rds")
  
wikipath_gsea.lst <- purrr::pmap(
  list(path_to_wikipath), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)

names(wikipath_gsea.lst) <- names.lst
```

# Mouse bulk RNA-seq consistency analysis

This section we evaluated the consistency of gene expression in mouse data sets.

## Direction consistency of the known DEGs

In this method, we showed the consistecy of expression direction of the known DEGs previously reported from the literature.

```{r, message=FALSE, warning=FALSE}
# Extract all expressed genes without considering significance 
mouse_studies <- c("Hacken(2023)", "Iyer(2024)", "Hing(2023)", "Rompietti(2025)")

common_mouse_genes <- purrr::map(
  dea.lst[mouse_studies],
  ~ .x %>%
    dplyr::filter(!is.na(`Mouse.gene.name`)) %>%
    pull(`Mouse.gene.name`)
) %>%
  Reduce(intersect, .)

common_mouse_gene_expr.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(`Mouse.gene.name` %in% common_mouse_genes) %>%
      dplyr::select(`Mouse.gene.name`, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
common_mouse_gene_expr.df <- common_mouse_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange
  ) 

DT::datatable(common_mouse_gene_expr.df, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# Extract all genes found in all study
all_mouse_gene_expr.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::select(`Mouse.gene.name`, gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
all_mouse_gene_expr.df <- all_mouse_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = list(log2FoldChange = ~mean(.x, na.rm = TRUE)) 
  ) %>%
  dplyr::mutate(
    Hacken_direction = dplyr::case_when(
      `Hacken(2023)` > 0 ~ "Up",
      `Hacken(2023)` < 0 ~ "Down",
      TRUE ~ "NA"
    ),
    Iyer_direction = dplyr::case_when(
      `Iyer(2024)` > 0 ~ "Up",
      `Iyer(2024)` < 0 ~ "Down",
      TRUE ~ "NA"
    ),
    Hing_direction = dplyr::case_when(
      `Hing(2023)` > 0 ~ "Up",
      `Hing(2023)` < 0 ~ "Down",
      TRUE ~ "NA"
    ),
    Rompietti_direction = dplyr::case_when(
      `Rompietti(2025)` > 0 ~ "Up",
      `Rompietti(2025)` < 0 ~ "Down",
      TRUE ~ "NA"
    )
  ) 

DT::datatable(all_mouse_gene_expr.df, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save the results
saveRDS(common_mouse_gene_expr.df, here::here("scRNAseq/meta_analysis/R_objects/common_mouse_gene_expr.rds"))
saveRDS(all_mouse_gene_expr.df, here::here("scRNAseq/meta_analysis/R_objects/all_mouse_gene_expr.rds"))
```

```{r}
# Define known mouse up- and downregulated DEGs
gene_mouse_categories <- list(
  # Upregulated DEGs 
  Cell_Cycle_Up = c("Mki67", "Cdk1", "Cdk6", "Ccna2", "Ccnb1", "Ccnd2", "Ccnd3"),
  Mitosis_DNA_Up = c("Aurkb", "Top2a"),
  Oncogenes_TF_Up = c("Myc", "Egr1"),
  Immune_Checkpoint_Up = c("Ctla4", "Cd274", "Il10"), # Immunosuppression and T-cell exhaustion markers (common in RT microenvironment)
  Microenv_Homing_Up = c("Cxcr5", "Ccr7", "Ccl17", "Ccl22", "Cx3cl1"), # Chemokines and Receptors for cell trafficking and interaction
  Signaling_Metabolism_Up = c("Ldha", "Tgfbr1", "Il16ra"),

  # Downregulated DEGs 
  Cycle_Inhibitors_Down = c("Cdkn1a", "Cdkn2b"), # Loss of cell cycle "brakes" (Tumor Suppressors)
  Oncogenes_TF_Down = c("Sox4", "Runx3", "Bcl6", "Bcl11b"),
  Immune_Checkpoint_Down = c("Cd74"),
  Microenv_Homing_Down = c("Itgb7"),
  Signaling_Metabolism_Down = c("Itk"),
  Immune_Reg_Down = c("Lair1") # Loss of inhibitory immune receptors
)

known_mouse_degs <- unlist(gene_mouse_categories)
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=2}
# Evaluate the consistency through the direction of the known DEGs by plotting the heatmap
desired_order <- c("Hing(2023)", "Hacken(2023)", "Iyer(2024)", "Rompietti(2025)")

# Extract expression matrix 
expr_mat <- as.matrix(common_mouse_gene_expr.df[ , -1])   # all columns except gene
rownames(expr_mat) <- common_mouse_gene_expr.df$`Mouse.gene.name`
valid_genes <- intersect(known_mouse_degs, rownames(expr_mat))
known_degs_expr_mat <- expr_mat[valid_genes, ]
known_degs_expr_mat <- apply(known_degs_expr_mat, 2, as.numeric)
rownames(known_degs_expr_mat) <- valid_genes
known_degs_expr_mat <- known_degs_expr_mat[ , desired_order]
colnames(known_degs_expr_mat) <- c("Hing(2023)", "Ten Hacken(2023)", "Iyer(2024)", "Rompietti(2025)")

# Set scale for heatmap
palette_length <- 100
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(palette_length)
value_limit <- max(abs(known_degs_expr_mat), na.rm = TRUE)
breaks <- seq(-value_limit, value_limit, length.out = palette_length + 1)

# Plot the heatmap
pheatmap(
  known_degs_expr_mat,
  scale = "none",
  color = colour_pal,
  breaks = breaks,
  cluster_rows = FALSE,
  cluster_cols = FALSE
)
```

## Common upregulated DEGs

```{r}
# Select mouse data sets and extract adjusted significant upregulated genes and collect in a list
mouse_sig_upreg_gene.lst <- purrr::pmap(
  list(dea.lst[mouse_studies]),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange > 0.25 & 
                         dea$padj < 0.05 & 
                         !is.na(dea$`Mouse.gene.name`) & 
                         dea$`Mouse.gene.name` != "", "Mouse.gene.name"]
    return(selected_gene)
  }
)

# Drop empty ones
mouse_sig_upreg_gene.lst <- mouse_sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

```{r}
# Plot the Venn diagram
mouse_up_degs_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_upreg_gene.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_upreg_gene.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_upreg_gene.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_upreg_gene.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_up_degs_venn.plt
```

```{r}
# Collect common upregulated DEGs
mouse_common_up_degs <- Reduce(intersect, mouse_sig_upreg_gene.lst)
mouse_common_up_degs
```

## Common downregulated DEGs

```{r}
# Select mouse data sets and extract adjusted significant downregulated genes and collect in a list
mouse_sig_downreg_gene.lst <- purrr::pmap(
  list(dea.lst[mouse_studies]),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange < (-0.25) & 
                         dea$padj < 0.05 & 
                         !is.na(dea$`Mouse.gene.name`) & 
                         dea$`Mouse.gene.name` != "", "Mouse.gene.name"]
    return(selected_gene)
  }
)

# Drop empty ones
mouse_sig_downreg_gene.lst <- mouse_sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

```{r}
# Plot the Venn diagram
mouse_down_degs_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_downreg_gene.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_downreg_gene.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_downreg_gene.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_downreg_gene.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_down_degs_venn.plt
```

```{r}
# Collect common downregulated DEGs
mouse_common_down_degs <- Reduce(intersect, mouse_sig_downreg_gene.lst)
mouse_common_down_degs
```

```{r, eval=FALSE}
# Save the results in the csv files
# 1. Pad the lists to the same length with NAs
max_len <- max(length(mouse_common_up_degs), length(mouse_common_down_degs))
length(mouse_common_up_degs) <- max_len 
length(mouse_common_down_degs) <- max_len

# 2. Combine into a Data Frame and set your Column Names here
mouse_common_degs <- data.frame(
  "Common_up_DEGs" = mouse_common_up_degs,   
  "Common_down_DEGs" = mouse_common_down_degs
)

# 3. Save as CSV
write.csv(
  mouse_common_degs, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/mouse_bulk_common_degs.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)
```

```{r, fig.height=2.5, fig.width=11}
# Visualise the heatmap for common up- and downregulated DEGs

# 1. Extract common DEGs expression data
mouse_common_degs <- c(mouse_common_up_degs, mouse_common_down_degs)
mouse_common_degs_expr.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(`Mouse.gene.name` %in% mouse_common_degs) %>%
      dplyr::select(`Mouse.gene.name`, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# 2. Change the data frame to wider form and convert to a matrix
mouse_common_degs_expr.df <- mouse_common_degs_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  ) %>%
  dplyr::mutate(`Mouse.gene.name` = factor(`Mouse.gene.name`, levels = mouse_common_degs)) %>%
  dplyr::arrange(`Mouse.gene.name`) %>%
  dplyr::rename("Ten Hacken(2023)" = `Hacken(2023)`) %>%
  dplyr::select(
    `Mouse.gene.name`, 
    `Hing(2023)`, 
    `Ten Hacken(2023)`, 
    `Iyer(2024)`, 
    `Rompietti(2025)`
  )

mouse_common_sig_expr_mat <- mouse_common_degs_expr.df %>% tibble::column_to_rownames(var = "Mouse.gene.name") %>% as.matrix()
mouse_common_sig_expr_mat <- t(mouse_common_sig_expr_mat)

# 4. Plot the heatmap
breaks <- seq(-10, 10, length.out = 101)
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)
pheatmap(
  mouse_common_sig_expr_mat,
  scale = "none",
  color = colour_pal,
  breaks = breaks,
  cluster_rows = FALSE,
  cluster_cols = FALSE
)
```

## Common GSEA terms

### Hallmark

```{r}
# Upregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(hm_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_up_hm_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_up_hm_venn.plt
```

```{r}
# Collect common upregulated terms
mouse_common_up_hm <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_up_hm
```

```{r}
# Downregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(hm_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_down_hm_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_down_hm_venn.plt
```

```{r}
# Collect common downregulated terms
mouse_common_down_hm <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_down_hm
```

### GO:BP

```{r}
# Upregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(go_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_up_go_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_up_go_venn.plt
```

```{r}
# Collect common upregulated terms
mouse_common_up_go <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_up_go
```

```{r}
# Downregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(go_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_down_go_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_down_go_venn.plt
```

```{r}
# Collect common downregulated terms
mouse_common_down_go <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_down_go
```

### KEGG

```{r}
# Upregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(kegg_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_up_kegg_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_up_kegg_venn.plt
```

```{r}
# Collect common upregulated terms
mouse_common_up_kegg <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_up_kegg
```

```{r}
# Downregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(kegg_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_down_kegg_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_down_kegg_venn.plt
```

```{r}
# Collect common downregulated terms
mouse_common_down_kegg <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_down_kegg
```

### Reactome

```{r}
# Upregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(reactome_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_up_reactome_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_up_reactome_venn.plt
```

```{r}
# Collect common upregulated terms
mouse_common_up_reactome <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_up_reactome
```

```{r}
# Downregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(reactome_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_down_reactome_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_down_reactome_venn.plt
```

```{r}
# Collect common downregulated terms
mouse_common_down_reactome <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_down_reactome
```

### WikiPath

```{r}
# Upregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(wikipath_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_up_wikipath_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_up_wikipath_venn.plt
```

```{r}
# Collect common upregulated terms
mouse_common_up_wikipath <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_up_wikipath
```

```{r}
# Downregulated terms
mouse_sig_term.lst <- purrr::pmap(
  list(wikipath_gsea.lst[mouse_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
mouse_sig_term.lst <- mouse_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
mouse_down_wikipath_venn.plt <- ggvenn(
  list("Hing (2023)" = mouse_sig_term.lst$`Hing(2023)`,
       "Ten Hacken (2023)" = mouse_sig_term.lst$`Hacken(2023)`,
       "Iyer (2024)" = mouse_sig_term.lst$`Iyer(2024)`,
       "Rompietti (2025)" = mouse_sig_term.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899", "#F1788D"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

mouse_down_wikipath_venn.plt
```

```{r}
# Collect common downregulated terms
mouse_common_down_wikipath <- Reduce(intersect, mouse_sig_term.lst)
mouse_common_down_wikipath
```

## Numerical consistency analysis across gene expression

We also measured the gene expression consistency across those studies by selecting genes with significant expression to evaluate the correlation of those genes, see the following steps. We measured numerically consistency between the expression level through correlations and similarities.

### All significant genes

```{r, message=FALSE, warning=FALSE}
# Extract significant genes 
common_sig_mouse_genes <- purrr::map(
  dea.lst[mouse_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05) %>%
    dplyr::filter(!is.na(`Mouse.gene.name`)) %>%
    pull(`Mouse.gene.name`)
) %>%
  Reduce(intersect, .)

sig_mouse_gene_expr.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(`Mouse.gene.name` %in% common_sig_mouse_genes) %>%
      dplyr::select(`Mouse.gene.name`, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
sig_mouse_gene_expr.df <- sig_mouse_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename and rearrange the column
sig_mouse_gene_expr.df <- sig_mouse_gene_expr.df %>%
  dplyr::rename("Ten Hacken(2023)" = `Hacken(2023)`)

sig_mouse_gene_expr.df <- sig_mouse_gene_expr.df %>%
  dplyr::select(
    `Mouse.gene.name`, 
    `Hing(2023)`, 
    `Ten Hacken(2023)`, 
    `Iyer(2024)`, 
    `Rompietti(2025)`
  )

head(sig_mouse_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- sig_mouse_gene_expr.df %>% dplyr::select(-`Mouse.gene.name`) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.4, fig.width=5.8}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

### Upregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract upregulated genes and their log2FC 
common_upreg_mouse_genes <- purrr::map(
  dea.lst[mouse_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
    dplyr::filter(!is.na(`Mouse.gene.name`)) %>%
    pull(`Mouse.gene.name`)
) %>%
  Reduce(intersect, .)

mouse_upreg_gene_expr.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(`Mouse.gene.name` %in% common_upreg_mouse_genes) %>%
      dplyr::select(`Mouse.gene.name`, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
mouse_upreg_gene_expr.df <- mouse_upreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename and rearrange the column
mouse_upreg_gene_expr.df <- mouse_upreg_gene_expr.df %>%
  dplyr::rename("Ten Hacken(2023)" = `Hacken(2023)`)

mouse_upreg_gene_expr.df <- mouse_upreg_gene_expr.df %>%
  dplyr::select(
    `Mouse.gene.name`, 
    `Hing(2023)`, 
    `Ten Hacken(2023)`, 
    `Iyer(2024)`, 
    `Rompietti(2025)`
  )

head(mouse_upreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- mouse_upreg_gene_expr.df %>% dplyr::select(-`Mouse.gene.name`) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.4, fig.width=5.8}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

### Downregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract downregulated genes and their log2FC 
common_downreg_mouse_genes <- purrr::map(
  dea.lst[mouse_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
    dplyr::filter(!is.na(`Mouse.gene.name`)) %>%
    pull(`Mouse.gene.name`)
) %>%
  Reduce(intersect, .)

mouse_downreg_gene_expr.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(`Mouse.gene.name` %in% common_downreg_mouse_genes) %>%
      dplyr::select(`Mouse.gene.name`, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
mouse_downreg_gene_expr.df <- mouse_downreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename and rearrange the column
mouse_downreg_gene_expr.df <- mouse_downreg_gene_expr.df %>%
  dplyr::rename("Ten Hacken(2023)" = `Hacken(2023)`)

mouse_downreg_gene_expr.df <- mouse_downreg_gene_expr.df %>%
  dplyr::select(
    `Mouse.gene.name`, 
    `Hing(2023)`, 
    `Ten Hacken(2023)`, 
    `Iyer(2024)`, 
    `Rompietti(2025)`
  )

head(mouse_downreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- mouse_downreg_gene_expr.df %>% dplyr::select(-`Mouse.gene.name`) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.4, fig.width=5.8}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

### Upregulated GSEA

```{r, message=FALSE, warning=FALSE}
# Hallmark 
mouse_common_up_hm_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- hm_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_up_hm) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# GO:BP 
mouse_common_up_go_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- go_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_up_go) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# KEGG 
mouse_common_up_kegg_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- kegg_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_up_kegg) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Reactome 
mouse_common_up_reactome_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- reactome_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_up_reactome) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# WikiPath 
mouse_common_up_wikipath_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- wikipath_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_up_wikipath) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Join the data frames using rbind()
mouse_common_up_terms <- bind_rows(mouse_common_up_hm_nes.df, 
                                   mouse_common_up_go_nes.df, 
                                   mouse_common_up_kegg_nes.df,
                                   mouse_common_up_reactome_nes.df,
                                   mouse_common_up_wikipath_nes.df)
mouse_common_up_terms <- mouse_common_up_terms %>%
  dplyr::rename("Ten Hacken(2023)" = `Hacken(2023)`) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(avrg_NES = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::select(
    ID,
    `Hing(2023)`, 
    `Ten Hacken(2023)`, 
    `Iyer(2024)`, 
    `Rompietti(2025)`,
    avrg_NES
  )

head(mouse_common_up_terms)
```

```{r}
# Remove the gene column for correlation
gsea_mat <- mouse_common_up_terms %>% dplyr::select(-c("ID", "avrg_NES")) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(gsea_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(gsea_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.4, fig.width=5.8}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson Correlation (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  spearman_cor,
  main = "Spearman Correlation (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)
```

### Downregulated GSEA

```{r, message=FALSE, warning=FALSE}
# Hallmark 
mouse_common_down_hm_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- hm_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_down_hm) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# GO:BP 
mouse_common_down_go_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- go_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_down_go) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# KEGG 
mouse_common_down_kegg_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- kegg_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_down_kegg) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Reactome 
mouse_common_down_reactome_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- reactome_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_down_reactome) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# WikiPath 
mouse_common_down_wikipath_nes.df <- purrr::map_dfr(
  mouse_studies,
  function(study_name){
    gsea <- wikipath_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_down_wikipath) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Join the data frames using rbind()
mouse_common_down_terms <- bind_rows(mouse_common_down_hm_nes.df, 
                                   mouse_common_down_go_nes.df, 
                                   mouse_common_down_kegg_nes.df,
                                   mouse_common_down_reactome_nes.df,
                                   mouse_common_down_wikipath_nes.df)
mouse_common_down_terms <- mouse_common_down_terms %>%
  dplyr::rename("Ten Hacken(2023)" = `Hacken(2023)`) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(avrg_NES = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::select(
    ID,
    `Hing(2023)`, 
    `Ten Hacken(2023)`, 
    `Iyer(2024)`, 
    `Rompietti(2025)`,
    avrg_NES
  )

head(mouse_common_down_terms)
```

```{r}
# Remove the gene column for correlation
gsea_mat <- mouse_common_down_terms %>% dplyr::select(-c("ID", "avrg_NES")) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(gsea_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(gsea_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.4, fig.width=5.8}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson Correlation (downregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  spearman_cor,
  main = "Spearman Correlation (downregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)
```

## GSEA summarisation

```{r}
# Rbind both common up- and downregulated term tables
mouse_common_terms <- bind_rows(mouse_common_up_terms, mouse_common_down_terms)
DT::datatable(mouse_common_terms, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# Plot the bar graph
selected_terms <- c("HALLMARK_E2F_TARGETS", "HALLMARK_MYC_TARGETS_V1", "HALLMARK_G2M_CHECKPOINT", "GOBP_MITOCHONDRIAL_TRANSLATION",
                    "KEGG_CELL_CYCLE", "WP_RETINOBLASTOMA_GENE_IN_CANCER", "HALLMARK_INFLAMMATORY_RESPONSE", 
                    "HALLMARK_KRAS_SIGNALING_UP", "GOBP_IMMUNE_RESPONSE_REGULATING_SIGNALING_PATHWAY", 
                    "GOBP_TYPE_II_INTERFERON_PRODUCTION", "GOBP_REGULATION_OF_T_CELL_ACTIVATION", 
                    "REACTOME_CLASS_A_1_RHODOPSIN_LIKE_RECEPTORS")

selected_enrichment.df <- mouse_common_terms %>%
  dplyr::filter(ID %in% selected_terms) %>%
  dplyr::select(ID, avrg_NES) 

selected_enrichment.df$ID <- gsub("HALLMARK_", "", selected_enrichment.df$ID)
selected_enrichment.df$ID <- gsub("GOBP_", "", selected_enrichment.df$ID)
selected_enrichment.df$ID <- gsub("KEGG_", "", selected_enrichment.df$ID)
selected_enrichment.df$ID <- gsub("REACTOME_", "", selected_enrichment.df$ID)
selected_enrichment.df$ID <- gsub("WP_", "", selected_enrichment.df$ID)
selected_enrichment.df$ID <- gsub("_V1", "", selected_enrichment.df$ID)
selected_enrichment.df$ID <- gsub("IMMUNE_RESPONSE_REGULATING_SIGNALING_PATHWAY", "IMMUNE_RESPONSE", selected_enrichment.df$ID)
selected_enrichment.df$ID <- gsub("_", " ", selected_enrichment.df$ID)

selected_enrichment.df$ID <- factor(selected_enrichment.df$ID, levels = selected_enrichment.df$ID)
selected_enrichment.df$Direction <- "Up"
selected_enrichment.df$Direction[selected_enrichment.df$avrg_NES<0] <- "Down"

gsea_bar.plt <- ggplot(selected_enrichment.df, aes(x=ID, y=avrg_NES, fill=Direction)) +
  geom_bar(stat = "identity", color="black", width = 0.75) +  
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, color = "black"),
    plot.margin = margin(t = 10, r = 10, b = 100, l = 65, unit = "pt"),
    legend.position = "none"
  ) +
  geom_hline(yintercept = 0) + 
  ylab("Average NES") + xlab(NULL) + 
  scale_fill_manual(values=c("#357EBD", "#D43F3A"))

gsea_bar.plt
```

# Human bulk and integrated single-cell RNA-seq consistency analysis

This section we evaluated the consistency of gene expression in human bulk and integrated data sets.

## Common upregulated DEGs

```{r}
# Select human data sets and extract adjusted significant upregulated genes and collect in a list
human_studies <- setdiff(names(dea.lst), mouse_studies)

human_sig_upreg_gene.lst <- purrr::pmap(
  list(dea.lst[human_studies]),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange > 0.25 & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)

# Drop empty ones
human_sig_upreg_gene.lst <- human_sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

```{r}
# Plot the Venn diagram
human_up_degs_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_upreg_gene.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_upreg_gene.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_upreg_gene.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_up_degs_venn.plt
```

```{r}
# Collect common upregulated DEGs
human_common_up_degs <- Reduce(intersect, human_sig_upreg_gene.lst)
human_common_up_degs
```

## Common downregulated DEGs

```{r}
# Select human data sets and extract adjusted significant upregulated genes and collect in a list
human_sig_downreg_gene.lst <- purrr::pmap(
  list(dea.lst[human_studies]),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange < (-0.25) & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)

# Drop empty ones
human_sig_downreg_gene.lst <- human_sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

```{r}
# Plot the Venn diagram
human_down_degs_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_downreg_gene.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_downreg_gene.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_downreg_gene.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_down_degs_venn.plt
```

```{r}
# Collect common downregulated DEGs
human_common_down_degs <- Reduce(intersect, human_sig_downreg_gene.lst)
human_common_down_degs
```

```{r, eval=FALSE}
# Save the results in the csv files
# 1. Pad the lists to the same length with NAs
max_len <- max(length(human_common_up_degs), length(human_common_down_degs))
length(human_common_up_degs) <- max_len 
length(human_common_down_degs) <- max_len

# 2. Combine into a Data Frame and set your Column Names here
human_bulk_integrated_common_degs <- data.frame(
  "Common_up_DEGs" = human_common_up_degs,   
  "Common_down_DEGs" = human_common_down_degs
)

# 3. Save as CSV
write.csv(
  human_bulk_integrated_common_degs, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/human_bulk_integrated_common_degs.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)
```

## Common GSEA terms

### Hallmark

```{r}
# Upregulated terms
human_sig_term.lst <- purrr::pmap(
  list(hm_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_up_hm_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_up_hm_venn.plt
```

```{r}
# Collect common upregulated terms
human_common_up_hm <- Reduce(intersect, human_sig_term.lst)
human_common_up_hm
```

```{r}
# Downregulated terms
human_sig_term.lst <- purrr::pmap(
  list(hm_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_down_hm_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_down_hm_venn.plt
```

### GO:BP

```{r}
# Upregulated terms
human_sig_term.lst <- purrr::pmap(
  list(go_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_up_go_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_up_go_venn.plt
```

```{r}
# Collect common upregulated terms
human_common_up_go <- Reduce(intersect, human_sig_term.lst)
human_common_up_go
```

```{r}
# Downregulated terms
human_sig_term.lst <- purrr::pmap(
  list(go_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_down_go_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_down_go_venn.plt
```

```{r}
# Collect common downregulated terms
human_common_down_go <- Reduce(intersect, human_sig_term.lst)
human_common_down_go
```

### KEGG

```{r}
# Upregulated terms
human_sig_term.lst <- purrr::pmap(
  list(kegg_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_up_kegg_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_up_kegg_venn.plt
```

```{r}
# Collect common upregulated terms
human_common_up_kegg <- Reduce(intersect, human_sig_term.lst)
human_common_up_kegg
```

```{r}
# Downregulated terms
human_sig_term.lst <- purrr::pmap(
  list(kegg_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_down_kegg_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_down_kegg_venn.plt
```

```{r}
# Collect common downregulated terms
human_common_down_kegg <- Reduce(intersect, human_sig_term.lst)
human_common_down_kegg
```

### Reactome

```{r}
# Upregulated terms
human_sig_term.lst <- purrr::pmap(
  list(reactome_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_up_reactome_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_up_reactome_venn.plt
```

```{r}
# Collect common upregulated terms
human_common_up_reactome <- Reduce(intersect, human_sig_term.lst)
human_common_up_reactome
```

```{r}
# Downregulated terms
human_sig_term.lst <- purrr::pmap(
  list(reactome_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_down_reactome_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_down_reactome_venn.plt
```

```{r}
# Collect common downregulated terms
human_common_down_reactome <- Reduce(intersect, human_sig_term.lst)
human_common_down_reactome
```

### WikiPath

```{r}
# Upregulated terms
human_sig_term.lst <- purrr::pmap(
  list(wikipath_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_up_wikipath_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_up_wikipath_venn.plt
```

```{r}
# Collect common upregulated terms
human_common_up_wikipath <- Reduce(intersect, human_sig_term.lst)
human_common_up_wikipath
```

```{r}
# Downregulated terms
human_sig_term.lst <- purrr::pmap(
  list(wikipath_gsea.lst[human_studies]),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Drop empty ones
human_sig_term.lst <- human_sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Plot the Venn diagram
human_down_wikipath_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_sig_term.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_sig_term.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_sig_term.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_down_wikipath_venn.plt
```

```{r}
# Collect common downregulated terms
human_common_down_wikipath <- Reduce(intersect, human_sig_term.lst)
human_common_down_wikipath
```

## Numerical consistency analysis across gene expression

We also measured the gene expression consistency across those studies by selecting genes with significant expression to evaluate the correlation of those genes, see the following steps. We measured numerically consistency between the expression level through correlations and similarities.

### All significant genes

```{r, message=FALSE, warning=FALSE}
# Extract significant genes 
common_sig_human_genes <- purrr::map(
  dea.lst[human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

sig_human_gene_expr.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_sig_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
sig_human_gene_expr.df <- sig_human_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename the column
sig_human_gene_expr.df <- sig_human_gene_expr.df %>%
  dplyr::rename(
    "Bulk Nadeu(2022)" = `Nadeu(2022)`,
    "Bulk Parry(2023a)" = `Parry(2023a)`
    ) 

head(sig_human_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- sig_human_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

### Upregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract upregulated genes and their log2FC 
common_upreg_human_genes <- purrr::map(
  dea.lst[human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

human_upreg_gene_expr.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_upreg_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
human_upreg_gene_expr.df <- human_upreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename the column
human_upreg_gene_expr.df <- human_upreg_gene_expr.df %>%
  dplyr::rename(
    "Bulk Nadeu(2022)" = `Nadeu(2022)`,
    "Bulk Parry(2023a)" = `Parry(2023a)`
    ) 

head(human_upreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- human_upreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

### Downregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract downregulated genes and their log2FC 
common_downreg_human_genes <- purrr::map(
  dea.lst[human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

human_downreg_gene_expr.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_downreg_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
human_downreg_gene_expr.df <- human_downreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename the column
human_downreg_gene_expr.df <- human_downreg_gene_expr.df %>%
  dplyr::rename(
    "Bulk Nadeu(2022)" = `Nadeu(2022)`,
    "Bulk Parry(2023a)" = `Parry(2023a)`
    ) 

head(human_downreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- human_downreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

### Upregulated GSEA

```{r, message=FALSE, warning=FALSE}
# Hallmark 
human_common_up_hm_nes.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    gsea <- hm_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% human_common_up_hm) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# GO:BP 
human_common_up_go_nes.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    gsea <- go_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% human_common_up_go) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# KEGG 
human_common_up_kegg_nes.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    gsea <- kegg_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% human_common_up_kegg) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Reactome 
human_common_up_reactome_nes.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    gsea <- reactome_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% human_common_up_reactome) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# WikiPath 
human_common_up_wikipath_nes.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    gsea <- wikipath_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% mouse_common_up_wikipath) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Join the data frames using rbind()
human_common_up_terms <- bind_rows(human_common_up_hm_nes.df, 
                                   human_common_up_go_nes.df, 
                                   human_common_up_kegg_nes.df,
                                   human_common_up_reactome_nes.df,
                                   human_common_up_wikipath_nes.df)

human_common_up_terms <- human_common_up_terms %>%
  dplyr::rowwise() %>%
  dplyr::mutate(avrg_NES = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(
    "Bulk Nadeu(2022)"    = `Nadeu(2022)`,
    "Bulk Parry(2023a)"   = `Parry(2023a)`
  )

head(human_common_up_terms)
```

```{r}
# Remove the gene column for correlation
gsea_mat <- human_common_up_terms %>% dplyr::select(-c("ID", "avrg_NES")) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(gsea_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(gsea_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson Correlation (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  spearman_cor,
  main = "Spearman Correlation (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)
```

### Downregulated GSEA

```{r, message=FALSE, warning=FALSE}
# Hallmark --> no common downregulated term found

# GO:BP 
human_common_down_go_nes.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    gsea <- go_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% human_common_down_go) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# KEGG 
human_common_down_kegg_nes.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    gsea <- kegg_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% human_common_down_kegg) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Reactome 
human_common_down_reactome_nes.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    gsea <- reactome_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% human_common_down_reactome) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# WikiPath 
human_common_down_wikipath_nes.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    gsea <- wikipath_gsea.lst[[study_name]]@result
    gsea_nes <- gsea %>%
      dplyr::filter(ID %in% human_common_down_wikipath) %>%
      dplyr::select(ID, NES) %>%
      dplyr::mutate(study = study_name) 
    return(gsea_nes)
  }
) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Join the data frames using rbind()
human_common_down_terms <- bind_rows(human_common_down_go_nes.df, 
                                     human_common_down_kegg_nes.df,
                                     human_common_down_reactome_nes.df,
                                     human_common_down_wikipath_nes.df)

human_common_down_terms <- human_common_down_terms %>%
  dplyr::rowwise() %>%
  dplyr::mutate(avrg_NES = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(
    "Bulk Nadeu(2022)"    = `Nadeu(2022)`,
    "Bulk Parry(2023a)"   = `Parry(2023a)`
  )

head(human_common_down_terms)
```

```{r}
# Remove the gene column for correlation
gsea_mat <- human_common_down_terms %>% dplyr::select(-c("ID", "avrg_NES")) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(gsea_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(gsea_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson Correlation (downregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  spearman_cor,
  main = "Spearman Correlation (downregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)
```

## Approach with the Seurat method for the integrated single-cell data

```{r, message=FALSE, warning=FALSE}
# Change the dea and gsea results from the integrated data to seurat approach
seurat_dea.lst <- dea.lst[human_studies]
seurat_dea.lst$`Integrated scRNA-seq` <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_dea.rds")) %>%
  tibble::rownames_to_column(var = "gene_name") %>%
      dplyr::rename(
       pvalue = "p_val",
       padj = "p_val_adj",
       log2FoldChange = "avg_log2FC"
      )

seurat_hm_gsea.lst <- hm_gsea.lst[human_studies]  
seurat_hm_gsea.lst$`Integrated scRNA-seq` <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_hm_gsea.rds"))

seurat_go_gsea.lst <- go_gsea.lst[human_studies]  
seurat_go_gsea.lst$`Integrated scRNA-seq` <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_go_gsea.rds"))

seurat_kegg_gsea.lst <- kegg_gsea.lst[human_studies]  
seurat_kegg_gsea.lst$`Integrated scRNA-seq` <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_kegg_gsea.rds"))

seurat_reactome_gsea.lst <- reactome_gsea.lst[human_studies]  
seurat_reactome_gsea.lst$`Integrated scRNA-seq` <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_reactome_gsea.rds"))

seurat_wikipath_gsea.lst <- wikipath_gsea.lst[human_studies]  
seurat_wikipath_gsea.lst$`Integrated scRNA-seq` <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_wikipath_gsea.rds"))
```

### Common upregulated DEGs

```{r}
# Select human data sets and extract adjusted significant upregulated genes and collect in a list
human_seurat_sig_upreg_gene.lst <- purrr::pmap(
  list(seurat_dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange > 0.25 & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)

# Drop empty ones
human_seurat_sig_upreg_gene.lst <- human_seurat_sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

```{r}
# Plot the Venn diagram
human_seurat_up_degs_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_seurat_sig_upreg_gene.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_seurat_sig_upreg_gene.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_seurat_sig_upreg_gene.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_seurat_up_degs_venn.plt
```

```{r}
# Collect common upregulated DEGs
human_seurat_common_up_degs <- Reduce(intersect, human_seurat_sig_upreg_gene.lst)
human_seurat_common_up_degs
```

### Common downregulated DEGs

```{r}
# Select human data sets and extract adjusted significant upregulated genes and collect in a list
human_seurat_sig_downreg_gene.lst <- purrr::pmap(
  list(seurat_dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange < (-0.25) & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)

# Drop empty ones
human_seurat_sig_downreg_gene.lst <- human_seurat_sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

```{r}
# Plot the Venn diagram
human_seurat_down_degs_venn.plt <- ggvenn(
  list("Bulk Nadeu (2022)" = human_seurat_sig_downreg_gene.lst$`Nadeu(2022)`,
       "Bulk Parry (2023a)" = human_seurat_sig_downreg_gene.lst$`Parry(2023a)`,
       "Integrated scRNA-seq" = human_seurat_sig_downreg_gene.lst$`Integrated scRNA-seq`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

human_seurat_down_degs_venn.plt
```

```{r}
# Collect common downregulated DEGs
human_seurat_common_down_degs <- Reduce(intersect, human_seurat_sig_downreg_gene.lst)
human_seurat_common_down_degs
```

```{r, eval=FALSE}
# Save the results in the csv files
# 1. Pad the lists to the same length with NAs
max_len <- max(length(human_seurat_common_up_degs), length(human_seurat_common_down_degs))
length(human_seurat_common_up_degs) <- max_len 
length(human_seurat_common_down_degs) <- max_len

# 2. Combine into a Data Frame and set your Column Names here
human_seurat_bulk_integrated_common_degs <- data.frame(
  "Common_up_DEGs" = human_seurat_common_up_degs,   
  "Common_down_DEGs" = human_seurat_common_down_degs
)

# 3. Save as CSV
write.csv(
  human_seurat_bulk_integrated_common_degs, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/human_seurat_bulk_integrated_common_degs.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)
```

### Numerical consistency analysis across gene expression

#### All significant genes

```{r, message=FALSE, warning=FALSE}
# Extract significant genes 
seurat_common_sig_human_genes <- purrr::map(
  seurat_dea.lst,
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

seurat_sig_human_gene_expr.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    dea <- seurat_dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% seurat_common_sig_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
seurat_sig_human_gene_expr.df <- seurat_sig_human_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename the column
seurat_sig_human_gene_expr.df <- seurat_sig_human_gene_expr.df %>%
  dplyr::rename(
    "Bulk Nadeu(2022)" = `Nadeu(2022)`,
    "Bulk Parry(2023a)" = `Parry(2023a)`
    ) 

head(seurat_sig_human_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- seurat_sig_human_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

#### Upregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract upregulated genes and their log2FC 
seurat_common_upreg_human_genes <- purrr::map(
  seurat_dea.lst,
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

seurat_human_upreg_gene_expr.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    dea <- seurat_dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% seurat_common_upreg_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
seurat_human_upreg_gene_expr.df <- seurat_human_upreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename the column
seurat_human_upreg_gene_expr.df <- seurat_human_upreg_gene_expr.df %>%
  dplyr::rename(
    "Bulk Nadeu(2022)" = `Nadeu(2022)`,
    "Bulk Parry(2023a)" = `Parry(2023a)`
    ) 

head(seurat_human_upreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- seurat_human_upreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

#### Downregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract downregulated genes and their log2FC 
seurat_common_downreg_human_genes <- purrr::map(
  seurat_dea.lst,
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

seurat_human_downreg_gene_expr.df <- purrr::map_dfr(
  human_studies,
  function(study_name){
    dea <- seurat_dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% seurat_common_downreg_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
seurat_human_downreg_gene_expr.df <- seurat_human_downreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename the column
seurat_human_downreg_gene_expr.df <- seurat_human_downreg_gene_expr.df %>%
  dplyr::rename(
    "Bulk Nadeu(2022)" = `Nadeu(2022)`,
    "Bulk Parry(2023a)" = `Parry(2023a)`
    ) 

head(seurat_human_downreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- seurat_human_downreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

#### Upregulated GSEA

```{r, message=FALSE, warning=FALSE}
# Hallmark 
seurat_human_common_up_hm_nes.df <- purrr::map_dfr(seurat_hm_gsea.lst, ~ .x@result, .id = "study") %>%
  dplyr::group_by(ID) %>%
  # Keep IDs present in ALL studies AND where NES > 0 in ALL studies
  dplyr::filter(n() == length(seurat_hm_gsea.lst) & all(NES > 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(ID, NES, study) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# GO:BP 
seurat_human_common_up_go_nes.df <- purrr::map_dfr(seurat_go_gsea.lst, ~ .x@result, .id = "study") %>%
  dplyr::group_by(ID) %>%
  # Keep IDs present in ALL studies AND where NES > 0 in ALL studies
  dplyr::filter(n() == length(seurat_go_gsea.lst) & all(NES > 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(ID, NES, study) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# KEGG 
seurat_human_common_up_kegg_nes.df <- purrr::map_dfr(seurat_kegg_gsea.lst, ~ .x@result, .id = "study") %>%
  dplyr::group_by(ID) %>%
  # Keep IDs present in ALL studies AND where NES > 0 in ALL studies
  dplyr::filter(n() == length(seurat_kegg_gsea.lst) & all(NES > 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(ID, NES, study) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Reactome 
seurat_human_common_up_reactome_nes.df <- purrr::map_dfr(seurat_reactome_gsea.lst, ~ .x@result, .id = "study") %>%
  dplyr::group_by(ID) %>%
  # Keep IDs present in ALL studies AND where NES > 0 in ALL studies
  dplyr::filter(n() == length(seurat_reactome_gsea.lst) & all(NES > 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(ID, NES, study) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# WikiPath 
seurat_human_common_up_wikipath_nes.df <- purrr::map_dfr(seurat_wikipath_gsea.lst, ~ .x@result, .id = "study") %>%
  dplyr::group_by(ID) %>%
  # Keep IDs present in ALL studies AND where NES > 0 in ALL studies
  dplyr::filter(n() == length(seurat_wikipath_gsea.lst) & all(NES > 0)) %>%
  dplyr::ungroup() %>%
  dplyr::select(ID, NES, study) %>%
  tidyr::pivot_wider(names_from = study, values_from = NES)

# Join the data frames using rbind()
seurat_human_common_up_terms <- bind_rows(seurat_human_common_up_hm_nes.df, 
                                          seurat_human_common_up_go_nes.df, 
                                          seurat_human_common_up_kegg_nes.df,
                                          seurat_human_common_up_reactome_nes.df,
                                          seurat_human_common_up_wikipath_nes.df)

seurat_human_common_up_terms <- seurat_human_common_up_terms %>%
  dplyr::rowwise() %>%
  dplyr::mutate(avrg_NES = mean(c_across(where(is.numeric)), na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(
    "Bulk Nadeu(2022)"    = `Nadeu(2022)`,
    "Bulk Parry(2023a)"   = `Parry(2023a)`
  )

head(seurat_human_common_up_terms)
```

```{r}
# Remove the gene column for correlation
gsea_mat <- seurat_human_common_up_terms %>% dplyr::select(-c("ID", "avrg_NES")) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(gsea_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(gsea_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(gsea_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson Correlation (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  spearman_cor,
  main = "Spearman Correlation (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated GSEA)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12
)
```

#### Downregulated GSEA

We did not run this step as no common downregulated GSEA terms are found.

# Candidate novel RT-related genes identified from the integrated single-cell data set

```{r}
# Load the DEGs results
# 1. Set the single-cell study name
sc_study.lst <- c("1-Penter_2021_GSE165087", "2-Nadeu_2022", "3-Hing_2023_GSE205704", "4-Parry_2023b", "5-Rigo_2024")
sc_names.lst <- sub(
  "^\\d+-([A-Za-z]+)_([0-9]{4}[A-Za-z]?).*$",  # capture name and year
  "\\1(\\2)",                         # output as Name(Year)
  sc_study.lst
)
sc_study_names <- sub("^\\d+-([^_0-9]+).*", "\\1", sc_study.lst)

# 2. Download the dea using Seurat in each study and collect in a list
path_to_download <- str_c(path_to_single_cell_analysis, "/", sc_study.lst, "/expression_data/", sc_study_names, "_dea.rds")
sc_seurat_dea.lst <- purrr::pmap(
  list(path_to_download), 
  function(path){
    expr.mat <- readRDS(path)
    expr.mat <- expr.mat %>% 
      tibble::rownames_to_column(var = "gene_name") %>%
      dplyr::rename(
       pvalue = "p_val",
       padj = "p_val_adj",
       log2FoldChange = "avg_log2FC"
      )
    return(expr.mat)
  }
)
names(sc_seurat_dea.lst) <- sc_names.lst

# Load the seurat dea for the integrated data
integrated_seurat_df <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_dea.rds")) %>%
  tibble::rownames_to_column(var = "gene_name") %>%
      dplyr::rename(
       pvalue = "p_val",
       padj = "p_val_adj",
       log2FoldChange = "avg_log2FC"
      )

# 3. Download the dea using DESeq2 in each study and collect in a list
path_to_download <- str_c(path_to_single_cell_analysis, "/", sc_study.lst, "/pseudo_bulk_deseq2/", sc_study_names, "_deseq2_dea.rds")
sc_deseq2_dea.lst <- purrr::pmap(
  list(path_to_download), 
  function(path){
    expr.mat <- readRDS(path)
    expr.mat <- expr.mat %>% rownames_to_column(var = "gene_name")
    return(expr.mat)
  }
)
names(sc_deseq2_dea.lst) <- sc_names.lst

# 4. Combine bulk, single-cell, and integrated dea
human_bulk_dea.lst <- dea.lst[human_studies]
names(human_bulk_dea.lst) <- c("Bulk Nadeu(2022)", "Bulk Parry(2023a)", "Integrated scRNA-seq")
combine_human_deseq2_dea.lst <- c(sc_deseq2_dea.lst, human_bulk_dea.lst)
combine_human_seurat_dea.lst <- c(sc_seurat_dea.lst, human_bulk_dea.lst)
combine_human_seurat_dea.lst$`Integrated scRNA-seq` <- integrated_seurat_df # Change the integrated dea to seurat form
```

## Seurat for single-cell data

We identified the DEGs from our integrated scRNA-seq data that were not found in bulk and single-cell individual analysis with Seurat method.

### Predicted novel upregulated DEGs

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=33}
# Extract adjusted significant upregulated genes
bulk_sc_sig_upreg_gene.lst <- purrr::pmap(
  list(combine_human_seurat_dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] 
    selected_gene <- dea %>% 
      dplyr::filter(!is.na(gene_name)) %>%
      dplyr::filter(!is.na(padj)) %>%
      dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
      pull(gene_name)
    return(selected_gene)
  }
)

# Reorder the list first
bulk_sc_sig_upreg_gene.lst <- bulk_sc_sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(bulk_sc_sig_upreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 1.2, 2.5, 1.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

### Predicted novel downregulated DEGs

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=30}
# Extract adjusted significant downregulated genes
bulk_sc_sig_downreg_gene.lst <- purrr::pmap(
  list(combine_human_seurat_dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] 
    selected_gene <- dea %>% 
      dplyr::filter(!is.na(gene_name)) %>%
      dplyr::filter(!is.na(padj)) %>%
      dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
      pull(gene_name)
    return(selected_gene)
  }
)

# Reorder the list first
bulk_sc_sig_downreg_gene.lst <- bulk_sc_sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(bulk_sc_sig_downreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 1.2, 2.5, 1.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

From the Seurat-based differential expression analysis of the integrated data, we observed that several thousand genes were identified as differentially expressed. Compared with other studies and datasets, this number appears unusually high. Consequently, we believe that the DEG results obtained from the pseudobulk approach implemented in Seurat after integration may not be reliable.

```{r, eval=FALSE}
# Save the results
# 1. Unique upregulated DEGs
target_degs <- bulk_sc_sig_upreg_gene.lst[["Integrated scRNA-seq"]]
other_lists <- bulk_sc_sig_upreg_gene.lst[names(bulk_sc_sig_upreg_gene.lst) != "Integrated scRNA-seq"]
degs_to_remove <- unique(unlist(other_lists)) # Union all data sets except the integrated
unique_integrated_up_degs <- setdiff(target_degs, degs_to_remove)

# 2. Unique downregulated DEGs
target_degs <- bulk_sc_sig_downreg_gene.lst[["Integrated scRNA-seq"]]
other_lists <- bulk_sc_sig_downreg_gene.lst[names(bulk_sc_sig_downreg_gene.lst) != "Integrated scRNA-seq"]
degs_to_remove <- unique(unlist(other_lists)) # Union all data sets except the integrated
unique_integrated_down_degs <- setdiff(target_degs, degs_to_remove)

# 3. Save the results in the csv files
max_len <- max(length(unique_integrated_up_degs), length(unique_integrated_down_degs))
length(unique_integrated_up_degs) <- max_len 
length(unique_integrated_down_degs) <- max_len

unique_seurat_integrated_degs <- data.frame(
  "UpDEGs" = unique_integrated_up_degs,   
  "DownDEGs" = unique_integrated_down_degs
)

write.csv(
  unique_seurat_integrated_degs, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/unique_seurat_integrated_degs.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)
```

## DESeq2 for single-cell data

We also identified the DEGs from our integrated scRNA-seq data that were not found in bulk and single-cell individual analysis with DESeq2 method.

### Predicted novel upregulated DEGs

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=12}
# Extract adjusted significant upregulated genes
bulk_sc_sig_upreg_gene.lst <- purrr::pmap(
  list(combine_human_deseq2_dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] 
    selected_gene <- dea %>% 
      dplyr::filter(!is.na(gene_name)) %>%
      dplyr::filter(!is.na(padj)) %>%
      dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
      pull(gene_name)
    return(selected_gene)
  }
)

# Reorder the list first
bulk_sc_sig_upreg_gene.lst <- bulk_sc_sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(bulk_sc_sig_upreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 1.2, 2.5, 1.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Predicted novel upregulated DEGs based on the integrated data
target_degs <- bulk_sc_sig_upreg_gene.lst[["Integrated scRNA-seq"]]
other_lists <- bulk_sc_sig_upreg_gene.lst[names(bulk_sc_sig_upreg_gene.lst) != "Integrated scRNA-seq"]
degs_to_remove <- unique(unlist(other_lists)) # Union all data sets except the integrated
unique_integrated_up_degs <- setdiff(target_degs, degs_to_remove)
unique_integrated_up_degs
```

### Predicted novel downregulated DEGs

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=12}
# Extract adjusted significant downregulated genes
bulk_sc_sig_downreg_gene.lst <- purrr::pmap(
  list(combine_human_deseq2_dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] 
    selected_gene <- dea %>% 
      dplyr::filter(!is.na(gene_name)) %>%
      dplyr::filter(!is.na(padj)) %>%
      dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
      pull(gene_name)
    return(selected_gene)
  }
)

# Reorder the list first
bulk_sc_sig_downreg_gene.lst <- bulk_sc_sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(bulk_sc_sig_downreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 1.2, 2.5, 1.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Predicted novel downregulated DEGs based on the integrated data
target_degs <- bulk_sc_sig_downreg_gene.lst[["Integrated scRNA-seq"]]
other_lists <- bulk_sc_sig_downreg_gene.lst[names(bulk_sc_sig_downreg_gene.lst) != "Integrated scRNA-seq"]
degs_to_remove <- unique(unlist(other_lists)) # Union all data sets except the integrated
unique_integrated_down_degs <- setdiff(target_degs, degs_to_remove)
unique_integrated_down_degs
```

```{r, eval=FALSE}
# Save the results in the csv files
max_len <- max(length(unique_integrated_up_degs), length(unique_integrated_down_degs))
length(unique_integrated_up_degs) <- max_len 
length(unique_integrated_down_degs) <- max_len

unique_seurat_integrated_degs <- data.frame(
  "UpDEGs" = unique_integrated_up_degs,   
  "DownDEGs" = unique_integrated_down_degs
)

write.csv(
  unique_seurat_integrated_degs, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/unique_deseq2_integrated_degs.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=5.5}
# Plot genes of interest
gene_plot = c(unique_integrated_up_degs, unique_integrated_down_degs)
integrated_expr.df <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_deseq2_dea.rds"))
selected_df <- integrated_expr.df[gene_plot, ]

novel_gene.plt <- ggplot() +
  # --- NEW LAYER: Background genes (All genes) ---
  # Color is set to pale grey ("grey85"). 
  geom_point(
    data = integrated_expr.df, 
    aes(x = log2FoldChange, y = -log10(pvalue)), 
    color = "grey90",    # Pale grey
    size = 2,            # Slightly smaller than the highlighted points
    alpha = 0.8          # Slight transparency
  ) +
  # --- EXISTING LAYER: Highlighted genes ---
  geom_point(
    data = selected_df, 
    aes(x = log2FoldChange, y = -log10(pvalue), fill = -log10(padj)), 
    shape = 21, 
    size = 4, 
    colour = "black", 
    stroke = 0.6
  ) +
  scale_fill_viridis_c(option = "viridis", direction = -1) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-0.25, 0.25), linetype = "dashed") +
  # Use selected_df for the labels
  geom_text_repel(
    data = selected_df, 
    aes(x = log2FoldChange, y = -log10(pvalue), label = rownames(selected_df)), 
    size = 3, 
    max.overlaps = 20, 
    color = "black"
  ) +
  scale_x_continuous(
    limits = c(-1.7, 1.7),
    breaks = seq(-6, 6, 2)
  ) +
  ylim(1, 5.1) +
  labs(
    x = "Fold-change RT vs CLL",
    y = "-log(P-Value)",
    fill = "-log(FDR)"
  ) +
  guides(
    fill = guide_colorbar(
      barheight = unit(10, "cm"),
      barwidth = unit(0.6, "cm"),
      direction = "vertical"
    )
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.line = element_line(color = "black", linewidth = 0.5),
    panel.grid = element_blank()
  )

novel_gene.plt
```

# Mouse and human RNA-seq consistency analysis

## Common upregulated DEGs

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Select all data sets and extract adjusted significant upregulated genes and collect in a list
desired_order <- c("Hing(2023)", "Ten Hacken(2023)", "Iyer(2024)", "Rompietti(2025)", 
                   "Nadeu(2022)", "Parry(2023a)", "Integrated scRNA-seq")

all_sig_upreg_gene.lst <- purrr::pmap(
  list(dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange > 0.25 & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)
names(all_sig_upreg_gene.lst)[names(all_sig_upreg_gene.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
all_sig_upreg_gene.lst <- all_sig_upreg_gene.lst[rev(desired_order)]

# Drop empty ones
all_sig_upreg_gene.lst <- all_sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Convert to UpSetR input
upset_data <- fromList(all_sig_upreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 1.2, 2.5, 1.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Collect common bulk upregulated DEGs but not integrated single-cell data set 
bulk_common_up_degs <- Reduce(intersect, all_sig_upreg_gene.lst[names(all_sig_upreg_gene.lst) != "Integrated scRNA-seq"])
bulk_common_up_degs
```

## Common downregulated DEGs

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Select all data sets and extract adjusted significant downregulated genes and collect in a list
all_sig_downreg_gene.lst <- purrr::pmap(
  list(dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange < (-0.25) & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)
names(all_sig_downreg_gene.lst)[names(all_sig_downreg_gene.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
all_sig_downreg_gene.lst <- all_sig_downreg_gene.lst[rev(desired_order)]

# Drop empty ones
all_sig_downreg_gene.lst <- all_sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Convert to UpSetR input
upset_data <- fromList(all_sig_downreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 1.2, 2.5, 1.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Collect common bulk downregulated DEGs but not integrated single-cell data set 
bulk_common_down_degs <- Reduce(intersect, all_sig_downreg_gene.lst[names(all_sig_downreg_gene.lst) != "Integrated scRNA-seq"])
bulk_common_down_degs
```

## Common GSEA terms

### Hallmark

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Upregulated terms
sig_term.lst <- purrr::pmap(
  list(hm_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Collect common upregulated terms
all_common_up_hm <- Reduce(intersect, sig_term.lst)
all_common_up_hm
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Downregulated terms
sig_term.lst <- purrr::pmap(
  list(hm_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

### GO:BP

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Upregulated terms
sig_term.lst <- purrr::pmap(
  list(go_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Collect common upregulated terms
all_common_up_go <- Reduce(intersect, sig_term.lst)
all_common_up_go
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Downregulated terms
sig_term.lst <- purrr::pmap(
  list(go_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

### KEGG

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Upregulated terms
sig_term.lst <- purrr::pmap(
  list(kegg_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Collect common upregulated terms
all_common_up_kegg <- Reduce(intersect, sig_term.lst)
all_common_up_kegg
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Downregulated terms
sig_term.lst <- purrr::pmap(
  list(kegg_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

### Reactome

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Upregulated terms
sig_term.lst <- purrr::pmap(
  list(reactome_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Collect common upregulated terms
all_common_up_reactome <- Reduce(intersect, sig_term.lst)
all_common_up_reactome
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Downregulated terms
sig_term.lst <- purrr::pmap(
  list(reactome_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

### WikiPath

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Upregulated terms
sig_term.lst <- purrr::pmap(
  list(wikipath_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Collect common upregulated terms
all_common_up_wikipath <- Reduce(intersect, sig_term.lst)
all_common_up_wikipath
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Downregulated terms
sig_term.lst <- purrr::pmap(
  list(wikipath_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)
names(sig_term.lst)[names(sig_term.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
sig_term.lst <- sig_term.lst[rev(desired_order)]

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0)    # drop empty ones

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 2.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

## Seurat-based approach for the integrated single-cell data

### Common upregulated DEGs

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Select all data sets and extract adjusted significant upregulated genes and collect in a list
desired_order <- c("Hing(2023)", "Ten Hacken(2023)", "Iyer(2024)", "Rompietti(2025)", 
                   "Nadeu(2022)", "Parry(2023a)", "Integrated scRNA-seq")

seurat_all_dea.lst <- dea.lst
seurat_all_dea.lst$`Integrated scRNA-seq` <- seurat_dea.lst$`Integrated scRNA-seq`

seurat_all_sig_upreg_gene.lst <- purrr::pmap(
  list(seurat_all_dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange > 0.25 & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)
names(seurat_all_sig_upreg_gene.lst)[names(seurat_all_sig_upreg_gene.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
seurat_all_sig_upreg_gene.lst <- seurat_all_sig_upreg_gene.lst[rev(desired_order)]

# Drop empty ones
seurat_all_sig_upreg_gene.lst <- seurat_all_sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Convert to UpSetR input
upset_data <- fromList(seurat_all_sig_upreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 1.2, 2.5, 1.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Collect common bulk upregulated DEGs but not integrated single-cell data set 
seurat_bulk_common_up_degs <- Reduce(intersect, seurat_all_sig_upreg_gene.lst)
seurat_bulk_common_up_degs
```

### Common downregulated DEGs

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=27}
# Select all data sets and extract adjusted significant upregulated genes and collect in a list
seurat_all_sig_downreg_gene.lst <- purrr::pmap(
  list(seurat_all_dea.lst),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange < (-0.25) & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)
names(seurat_all_sig_downreg_gene.lst)[names(seurat_all_sig_downreg_gene.lst) == "Hacken(2023)"] <- "Ten Hacken(2023)"
seurat_all_sig_downreg_gene.lst <- seurat_all_sig_downreg_gene.lst[rev(desired_order)]

# Drop empty ones
seurat_all_sig_downreg_gene.lst <- seurat_all_sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 

# Convert to UpSetR input
upset_data <- fromList(seurat_all_sig_downreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 1.2, 2.5, 1.5), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r}
# Collect common bulk downregulated DEGs but not integrated single-cell data set 
seurat_bulk_common_down_degs <- Reduce(intersect, seurat_all_sig_downreg_gene.lst[names(seurat_all_sig_downreg_gene.lst) != "Integrated scRNA-seq"])
seurat_bulk_common_down_degs
```

```{r}
# Collect common bulk downregulated DEGs but not Hing data set 
seurat_bulk_common_down_degs <- Reduce(intersect, seurat_all_sig_downreg_gene.lst[names(seurat_all_sig_downreg_gene.lst) != "Hing(2023)"])
seurat_bulk_common_down_degs
```

## Evaluation of representation of mouse to human RNA-seq

```{r}
# Select mouse and human data sets and extract adjusted significant upregulated genes and collect in a list
selected_studies <- c("Hacken(2023)", "Iyer(2024)", "Hing(2023)", "Rompietti(2025)", "Nadeu(2022)", "Parry(2023a)")

bulk_mouse_human_sig_upreg_gene.lst <- purrr::pmap(
  list(dea.lst[selected_studies]),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange > 0.25 & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)

# Drop empty ones
bulk_mouse_human_sig_upreg_gene.lst <- bulk_mouse_human_sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

```{r}
# Select mouse and human data sets and extract adjusted significant downregulated genes and collect in a list
bulk_mouse_human_sig_downreg_gene.lst <- purrr::pmap(
  list(dea.lst[selected_studies]),
  function(dea){
    dea <- dea[!is.na(dea$padj), ] # remove genes with padj = NA
    selected_gene <- dea[dea$log2FoldChange < (-0.25) & 
                         dea$padj < 0.05 & 
                         !is.na(dea$gene_name) & 
                         dea$gene_name != "", "gene_name"]
    return(selected_gene)
  }
)

# Drop empty ones
bulk_mouse_human_sig_downreg_gene.lst <- bulk_mouse_human_sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) 
```

```{r, eval=FALSE}
# Save only DEGs from mouse studies for further analysis
all_mouse_up_degs <- unlist(bulk_mouse_human_sig_upreg_gene.lst[mouse_studies])
all_mouse_down_degs <- unlist(bulk_mouse_human_sig_downreg_gene.lst[mouse_studies])
saveRDS(all_mouse_up_degs, here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/all_mouse_up_degs.rds"))
saveRDS(all_mouse_down_degs, here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/all_mouse_down_degs.rds"))
```

### Ten Hacken (2023)

#### Gene overlapping analysis

```{r}
# Plot the Venn diagram
hacken_human_up_degs_venn.plt <- ggvenn(
  list("Nadeu (2022)" = bulk_mouse_human_sig_upreg_gene.lst$`Nadeu(2022)`,
       "Parry (2023a)" = bulk_mouse_human_sig_upreg_gene.lst$`Parry(2023a)`,
       "Ten Hacken (2023)" = bulk_mouse_human_sig_upreg_gene.lst$`Hacken(2023)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

hacken_human_up_degs_venn.plt
```

```{r}
# Plot the Venn diagram
hacken_human_down_degs_venn.plt <- ggvenn(
  list("Nadeu (2022)" = bulk_mouse_human_sig_downreg_gene.lst$`Nadeu(2022)`,
       "Parry (2023a)" = bulk_mouse_human_sig_downreg_gene.lst$`Parry(2023a)`,
       "Ten Hacken (2023)" = bulk_mouse_human_sig_downreg_gene.lst$`Hacken(2023)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

hacken_human_down_degs_venn.plt
```

#### Numerical consistency analysis across gene expression

##### All significant genes

```{r}
# Set selected studies
hacken_human_studies <- c("Hacken(2023)", "Nadeu(2022)", "Parry(2023a)")
```

```{r, message=FALSE, warning=FALSE}
# Extract significant genes 
common_sig_hacken_human_genes <- purrr::map(
  dea.lst[hacken_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

sig_hacken_human_gene_expr.df <- purrr::map_dfr(
  hacken_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_sig_hacken_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
sig_hacken_human_gene_expr.df <- sig_hacken_human_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename the column
sig_hacken_human_gene_expr.df <- sig_hacken_human_gene_expr.df %>%
  dplyr::rename(
    "Ten Hacken(2023)" = `Hacken(2023)`
    ) 

head(sig_hacken_human_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- sig_hacken_human_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

##### Upregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract upregulated genes and their log2FC 
common_upreg_hacken_human_genes <- purrr::map(
  dea.lst[hacken_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

hacken_human_upreg_gene_expr.df <- purrr::map_dfr(
  hacken_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_upreg_hacken_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
hacken_human_upreg_gene_expr.df <- hacken_human_upreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename the column
hacken_human_upreg_gene_expr.df <- hacken_human_upreg_gene_expr.df %>%
  dplyr::rename(
    "Ten Hacken(2023)" = `Hacken(2023)`
    ) 

head(hacken_human_upreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- hacken_human_upreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

##### Downregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract downregulated genes and their log2FC 
common_downreg_hacken_human_genes <- purrr::map(
  dea.lst[hacken_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

hacken_human_downreg_gene_expr.df <- purrr::map_dfr(
  hacken_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_downreg_hacken_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
hacken_human_downreg_gene_expr.df <- hacken_human_downreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

# Rename the column
hacken_human_downreg_gene_expr.df <- hacken_human_downreg_gene_expr.df %>%
  dplyr::rename(
    "Ten Hacken(2023)" = `Hacken(2023)`
    ) 

head(hacken_human_downreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- hacken_human_downreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

### Hing (2023)

#### Gene overlapping analysis

```{r}
# Plot the Venn diagram
hing_human_up_degs_venn.plt <- ggvenn(
  list("Nadeu (2022)" = bulk_mouse_human_sig_upreg_gene.lst$`Nadeu(2022)`,
       "Parry (2023a)" = bulk_mouse_human_sig_upreg_gene.lst$`Parry(2023a)`,
       "Hing (2023)" = bulk_mouse_human_sig_upreg_gene.lst$`Hing(2023)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

hing_human_up_degs_venn.plt
```

```{r}
# Plot the Venn diagram
hing_human_down_degs_venn.plt <- ggvenn(
  list("Nadeu (2022)" = bulk_mouse_human_sig_downreg_gene.lst$`Nadeu(2022)`,
       "Parry (2023a)" = bulk_mouse_human_sig_downreg_gene.lst$`Parry(2023a)`,
       "Hing (2023)" = bulk_mouse_human_sig_downreg_gene.lst$`Hing(2023)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

hing_human_down_degs_venn.plt
```

#### Numerical consistency analysis across gene expression

##### All significant genes

```{r}
# Set selected studies
hing_human_studies <- c("Hing(2023)", "Nadeu(2022)", "Parry(2023a)")
```

```{r, message=FALSE, warning=FALSE}
# Extract significant genes 
common_sig_hing_human_genes <- purrr::map(
  dea.lst[hing_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

sig_hing_human_gene_expr.df <- purrr::map_dfr(
  hing_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_sig_hing_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
sig_hing_human_gene_expr.df <- sig_hing_human_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

head(sig_hing_human_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- sig_hing_human_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

##### Upregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract upregulated genes and their log2FC 
common_upreg_hing_human_genes <- purrr::map(
  dea.lst[hing_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

hing_human_upreg_gene_expr.df <- purrr::map_dfr(
  hing_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_upreg_hing_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
hing_human_upreg_gene_expr.df <- hing_human_upreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

head(hing_human_upreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- hing_human_upreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

##### Downregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract downregulated genes and their log2FC 
common_downreg_hing_human_genes <- purrr::map(
  dea.lst[hing_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

hing_human_downreg_gene_expr.df <- purrr::map_dfr(
  hing_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_downreg_hing_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
hing_human_downreg_gene_expr.df <- hing_human_downreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

head(hing_human_downreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- hing_human_downreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

### Iyer (2024)

#### Gene overlapping analysis

```{r}
# Plot the Venn diagram
iyer_human_up_degs_venn.plt <- ggvenn(
  list("Nadeu (2022)" = bulk_mouse_human_sig_upreg_gene.lst$`Nadeu(2022)`,
       "Parry (2023a)" = bulk_mouse_human_sig_upreg_gene.lst$`Parry(2023a)`,
       "Iyer (2024)" = bulk_mouse_human_sig_upreg_gene.lst$`Iyer(2024)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

iyer_human_up_degs_venn.plt
```

```{r}
# Plot the Venn diagram
iyer_human_down_degs_venn.plt <- ggvenn(
  list("Nadeu (2022)" = bulk_mouse_human_sig_downreg_gene.lst$`Nadeu(2022)`,
       "Parry (2023a)" = bulk_mouse_human_sig_downreg_gene.lst$`Parry(2023a)`,
       "Iyer (2024)" = bulk_mouse_human_sig_downreg_gene.lst$`Iyer(2024)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

iyer_human_down_degs_venn.plt
```

#### Numerical consistency analysis across gene expression

##### All significant genes

```{r}
# Set selected studies
iyer_human_studies <- c("Iyer(2024)", "Nadeu(2022)", "Parry(2023a)")
```

```{r, message=FALSE, warning=FALSE}
# Extract significant genes 
common_sig_iyer_human_genes <- purrr::map(
  dea.lst[iyer_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

sig_iyer_human_gene_expr.df <- purrr::map_dfr(
  iyer_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_sig_iyer_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
sig_iyer_human_gene_expr.df <- sig_iyer_human_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

head(sig_iyer_human_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- sig_iyer_human_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

##### Upregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract upregulated genes and their log2FC 
common_upreg_iyer_human_genes <- purrr::map(
  dea.lst[iyer_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

iyer_human_upreg_gene_expr.df <- purrr::map_dfr(
  iyer_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_upreg_iyer_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
iyer_human_upreg_gene_expr.df <- iyer_human_upreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

head(iyer_human_upreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- iyer_human_upreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

##### Downregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract downregulated genes and their log2FC 
common_downreg_iyer_human_genes <- purrr::map(
  dea.lst[iyer_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

iyer_human_downreg_gene_expr.df <- purrr::map_dfr(
  iyer_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_downreg_iyer_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
iyer_human_downreg_gene_expr.df <- iyer_human_downreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

head(iyer_human_downreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- iyer_human_downreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

### Rompietti (2025)

#### Gene overlapping analysis

```{r}
# Plot the Venn diagram
rompietti_human_up_degs_venn.plt <- ggvenn(
  list("Nadeu (2022)" = bulk_mouse_human_sig_upreg_gene.lst$`Nadeu(2022)`,
       "Parry (2023a)" = bulk_mouse_human_sig_upreg_gene.lst$`Parry(2023a)`,
       "Rompietti (2025)" = bulk_mouse_human_sig_upreg_gene.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

rompietti_human_up_degs_venn.plt
```

```{r}
# Plot the Venn diagram
rompietti_human_down_degs_venn.plt <- ggvenn(
  list("Nadeu (2022)" = bulk_mouse_human_sig_downreg_gene.lst$`Nadeu(2022)`,
       "Parry (2023a)" = bulk_mouse_human_sig_downreg_gene.lst$`Parry(2023a)`,
       "Rompietti (2025)" = bulk_mouse_human_sig_downreg_gene.lst$`Rompietti(2025)`
       ),
  fill_color = c("#0073C2FF", "#EFC000FF", "#6FB899"),
  stroke_size = 0.5,
  set_name_size = 4.5,   # Size of the outside labels (Default is 6)
  text_size = 4        # Size of the numbers inside the diagram (Default is 4)
)

rompietti_human_down_degs_venn.plt
```

#### Numerical consistency analysis across gene expression

##### All significant genes

```{r}
# Set selected studies
rompietti_human_studies <- c("Rompietti(2025)", "Nadeu(2022)", "Parry(2023a)")
```

```{r, message=FALSE, warning=FALSE}
# Extract significant genes 
common_sig_rompietti_human_genes <- purrr::map(
  dea.lst[rompietti_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

sig_rompietti_human_gene_expr.df <- purrr::map_dfr(
  rompietti_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_sig_rompietti_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
sig_rompietti_human_gene_expr.df <- sig_rompietti_human_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

head(sig_rompietti_human_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- sig_rompietti_human_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

##### Upregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract upregulated genes and their log2FC 
common_upreg_rompietti_human_genes <- purrr::map(
  dea.lst[rompietti_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

rompietti_human_upreg_gene_expr.df <- purrr::map_dfr(
  rompietti_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_upreg_rompietti_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
rompietti_human_upreg_gene_expr.df <- rompietti_human_upreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

head(rompietti_human_upreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- rompietti_human_upreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

##### Downregulated genes

```{r, message=FALSE, warning=FALSE}
# Extract downregulated genes and their log2FC 
common_downreg_rompietti_human_genes <- purrr::map(
  dea.lst[rompietti_human_studies],
  ~ .x %>%
    dplyr::filter(!is.na(padj)) %>%
    dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
    dplyr::filter(!is.na(gene_name)) %>%
    pull(gene_name)
) %>%
  Reduce(intersect, .)

rompietti_human_downreg_gene_expr.df <- purrr::map_dfr(
  rompietti_human_studies,
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene_name %in% common_downreg_rompietti_human_genes) %>%
      dplyr::select(gene_name, log2FoldChange) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
rompietti_human_downreg_gene_expr.df <- rompietti_human_downreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = log2FoldChange,
    values_fn = mean # Average duplicates to get a single number
  )

head(rompietti_human_downreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- rompietti_human_downreg_gene_expr.df %>% dplyr::select(-gene_name) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5.8, fig.width=6.2}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

# Session Information

```{r}
sessionInfo()
```
