---
title: "Rejeski_2022_GSE201704_reanalysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This source code we reanalysed single-cell data for a CLL patient with RT progression from [Rejeski et al. (2022)](https://ashpublications.org/blood/article/140/20/2175/485763/Oligoclonal-T-cell-expansion-in-a-patient-with). The patient received CAR T-cell therapy (CART) and experienced with the bone marrow failure with T-cell expansion.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Import the packages
library(Seurat)
library(SingleR)
library(celldex)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(DT)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save
path_to_merged_obj <- here::here("scRNAseq/GSE201704/GSM6069857/with_protein/GSM6069857_merged.rds")
path_to_save_downstream_obj <- here::here("scRNAseq/GSE201704/GSM6069857/reanalysis/R_objects/Rejeski_2022_GSE201704_downstream_seu.rds")
```

# Load data and function

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the merged object
seurat <- readRDS(path_to_merged_obj)
```

# Downstream analysis

```{r, message=FALSE, warning=FALSE}
# Downstream analysis pipeline
seurat <- JoinLayers(seurat)
seurat <- NormalizeData(seurat)
seurat <- FindVariableFeatures(seurat, nfeatures = 1000)
seurat <- ScaleData(seurat)
seurat <- RunPCA(seurat)
```

# Clustering and running t-SNE and UMAP

```{r, message=FALSE, warning=FALSE}
# Graph-based clustering and running UMAP
seurat <- FindNeighbors(seurat, dims = 1:30)
seurat <- FindClusters(seurat, resolution = 0.45)
seurat <- RunUMAP(seurat, dims = 1:30)
seurat <- RunTSNE(seurat, dims = 1:30)
```

# Data visualisation

## Clustering

```{r fig.height=4, fig.width=10}
# Plot UMAP based on clusters 
umap_cluster.plt <- DimPlot(seurat, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

# Plot tSNE based on clusters 
tsne_cluster.plt <- DimPlot(seurat, reduction = "tsne", group.by = "seurat_clusters", label = TRUE) + 
  labs(y= "tSNE 2", x = "tSNE 1")

cluster.plt <- umap_cluster.plt + tsne_cluster.plt
cluster.plt
```

## Gene expression

```{r, warning=FALSE, fig.height=7, fig.width=9}
# Plot UMAP based on SingleR
FeaturePlot(seurat, 
            features = c("CD8A", "IL7R", "CD4", "CD19", "CD79A", "CD79B", "CD14", "GNLY", "NKG7"),
            reduction = "umap",
            min.cutoff = 'q10',
            max.cutoff = 'q90') 
```

```{r, warning=FALSE, fig.height=7, fig.width=9}
# Plot UMAP based on SingleR
FeaturePlot(seurat, 
            features = c("CD8A", "IL7R", "CD4", "CD19", "CD79A", "CD79B", "CD14", "GNLY", "NKG7"),
            reduction = "tsne",
            min.cutoff = 'q10',
            max.cutoff = 'q90')
```

## Surface protein expression

```{r}
# Change default assay to antibody capture and normalisation
DefaultAssay(seurat) <- "ADT"
```

```{r fig.height=7, fig.width=6}
# Plot UMAP based on surface protein expression
umap_feature.plt <- FeaturePlot(
  seurat,
  features = c("CD3", "CD4.1", "CD8", "CD19.1", "CD11c", "CD56"),
  reduction = "umap",
  min.cutoff = 'q10',
  max.cutoff = 'q90'
)

umap_feature.plt
```

```{r fig.height=7, fig.width=6}
# Plot tSNE based on surface protein expression
tsne_feature.plt <- FeaturePlot(
  seurat,
  features = c("CD3", "CD4.1", "CD8", "CD19.1", "CD11c", "CD56"),
  reduction = "tsne",
  min.cutoff = 'q10',
  max.cutoff = 'q90'
)

tsne_feature.plt
```

## HTO sample origin

```{r fig.height=4, fig.width=10.5}
# Plot UMAP based on HTO 
umap_hto.plt <- DimPlot(seurat, reduction = "umap", group.by = 'HTO') + labs(y= "UMAP 2", x = "UMAP 1")

# Plot tSNE based on HTO 
tsne_hto.plt <- DimPlot(seurat, reduction = "tsne", group.by = 'HTO') + labs(y= "tSNE 2", x = "tSNE 1")

hto.plt <- umap_hto.plt + tsne_hto.plt
hto.plt
```

# Cell annotation

## Annotation using SingleR

```{r, message=FALSE, warning=FALSE}
# Cell annotation with SingleR
DefaultAssay(seurat) <- "RNA"
seurat <- singleR_cell_annotation(seurat) 
```

```{r fig.height=4, fig.width=11}
# Plot UMAP based on SingleR 
umap_singler.plt <- DimPlot(seurat, reduction = "umap", group.by = 'singleR.labels_main') +
  labs(y= "UMAP 2", x = "UMAP 1")

# Plot tSNE based on SingleR 
tsne_singler.plt <- DimPlot(seurat, reduction = "tsne", group.by = 'singleR.labels_main') +
  labs(y= "tSNE 2", x = "tSNE 1")

singler.plt <- umap_singler.plt + tsne_singler.plt
singler.plt
```

## Annotated based on manually clustering and biomarker identification

```{r}
# Define new cluster labels
new.cluster.ids <- c("T cell", "NK cell", "Myeloid cell", "B cell", "T cell", "T cell", "Myeloid cell", "Myeloid cell", 
                     "Myeloid cell", "Unclassified", "NK cell", "Unclassified", "B cell", "Myeloid cell", "Myeloid cell", 
                     "NK cell", "Unclassified")

# Map these to the cluster identities (i.e., Idents of the Seurat object)
names(new.cluster.ids) <- levels(seurat)

# Now map cluster IDs to the manual labels
seurat$manually_annotated <- plyr::mapvalues(
  x = as.character(Idents(seurat)),
  from = names(new.cluster.ids),
  to = new.cluster.ids
)
```

```{r fig.height=4, fig.width=11}
# Plot UMAP based on manually annotation 
umap_annotated.plt <- DimPlot(seurat, reduction = "umap", group.by = "manually_annotated", label = TRUE) +
  labs(y= "UMAP 2", x = "UMAP 1")

# Plot tSNE based on manually annotation 
tsne_annotated.plt <- DimPlot(seurat, reduction = "tsne", group.by = "manually_annotated", label = TRUE) +
  labs(y= "tSNE 2", x = "tSNE 1")

annotated.plt <- umap_annotated.plt + tsne_annotated.plt
annotated.plt
```

```{r, eval=FALSE}
# Save the downstream analysed object
saveRDS(seurat, path_to_save_downstream_obj)
```

# Gene expression and functional enrichment analysis between pre- and post CAR-T treatment

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(msigdbr)
library(org.Hs.eg.db)
```

## T cells

### GSEA with Hallmarks

```{r}
# Extract only T cells
t_cell_seurat <- subset(seurat, subset = manually_annotated == 'T cell')

# Find markers between pre- and post-treatment  
Idents(t_cell_seurat) <- "HTO" # change the default identity
t_cell_pre_post.ge <- FindMarkers(t_cell_seurat, ident.1 = 'CMO306', ident.2 = 'CMO305')
```

```{r}
# Manipulate the expression data
t_cell_dea <- cbind(gene = rownames(t_cell_pre_post.ge), t_cell_pre_post.ge)
lfc_vector <- t_cell_dea$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- t_cell_dea$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r}
# Show some gene expression results
head(t_cell_dea)
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_IL2_STAT5_SIGNALING, HALLMARK_MYC_TARGETS_V1, and HALLMARK_INFLAMMATORY_RESPONSE

IL2_STAT5_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_IL2_STAT5_SIGNALING", 
                          geneSetID = "HALLMARK_IL2_STAT5_SIGNALING",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

inflam_response_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_INFLAMMATORY_RESPONSE", 
                          geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

IL2_STAT5_gsea.plt + Myc_gsea.plt + inflam_response_gsea.plt
```

In our T cell GSEA hallmark results, the IL2_STAT5 signalling and MYC targets terms were consistent with the original results (Fig. S7A in the original supplemental materials) while inflammatory response in our results was not clear, compared to the original study that this term was downregulated.

## Myeloid cells

### GSEA with Hallmarks

```{r}
# Extract only myeloid cells
myeloid_cell_seurat <- subset(seurat, subset = manually_annotated == 'Myeloid cell')

# Find markers between pre- and post-treatment  
Idents(myeloid_cell_seurat) <- "HTO" # change the default identity
myeloid_cell_pre_post.ge <- FindMarkers(myeloid_cell_seurat, ident.1 = 'CMO306', ident.2 = 'CMO305')
```

```{r}
# Manipulate the expression data
myeloid_cell_dea <- cbind(gene = rownames(myeloid_cell_pre_post.ge), myeloid_cell_pre_post.ge)
lfc_vector <- myeloid_cell_dea$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- myeloid_cell_dea$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_ALLOGRAFT_REJECTION, HALLMARK_IL2_STAT5_SIGNALING, and HALLMARK_INFLAMMATORY_RESPONSE

allograft_reject_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_ALLOGRAFT_REJECTION", 
                          geneSetID = "HALLMARK_ALLOGRAFT_REJECTION",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

IL2_STAT5_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_IL2_STAT5_SIGNALING", 
                          geneSetID = "HALLMARK_IL2_STAT5_SIGNALING",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

inflam_response_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_INFLAMMATORY_RESPONSE", 
                          geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

allograft_reject_gsea.plt + IL2_STAT5_gsea.plt + inflam_response_gsea.plt
```

For our myeloid GSEA hallmark results, allograft rejection and IL2-STAT5 signalling terms were upregulated, consistent with the original study (Fig. S7B) while inflammatory in our results was downregulated while the original results was upregulated.

### GSEA with GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=15}
# GSEA plot for GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN and GOBP_MYELOID_LEUKOCYTE_MEDIATED_IMMUNITY

antigen_present_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN", 
                          geneSetID = "GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

myeloid_mediated_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MYELOID_LEUKOCYTE_MEDIATED_IMMUNITY", 
                          geneSetID = "GOBP_MYELOID_LEUKOCYTE_MEDIATED_IMMUNITY",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

antigen_present_gsea.plt + myeloid_mediated_gsea.plt 
```

For our myeloid GSEA GO:BP results, antigen processing and presentation was downregulated, similar to the original study (Fig. S7B) while myeloid leukocyte immunity in our results was not clear while the original results was downregulated.

# Session Information

```{r}
sessionInfo()
```



