---
title: "Hing_2023_mouse_GSE183426_reanalysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This source code we generated Seurat objects and reanalysed single-cell RNA-seq for RT- and CLL-like conditions in mouse experiments. We analysed for [Hing et. al. (2023)](https://www.nature.com/articles/s41467-022-35778-1) study which employed a mouse experimental model using Eµ-hTCL1 and Eµ-hPRMT5/hTCL1 mouse model for CLL and RT, respectively.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tidyr)
library(SingleR)
library(celldex)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save
path_to_metadata <- here::here("scRNAseq/Hing_2023_GSE183426/sample_metadata.csv")
path_to_count_matrices <- here::here("scRNAseq/Hing_2023_GSE183426")
path_to_save_obj_lst <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_reanalysed_seu_lst.rds")
path_to_cell_annotat_ref <- here::here("scRNAseq/Hing_2023_GSE183426/ImmGenData_ref.rds")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the sample metadata
sample_metadata <- read.csv(path_to_metadata)
DT::datatable(sample_metadata, options = list(scrollX = TRUE))
```

We will import count matrices and convert them to Seurat objects with quality control and pre-processing.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create the objects and collect in a list
path_to_download <- str_c(path_to_count_matrices, "/", sample_metadata$GSM_ID, "/count_matrices/")
seurat.lst <- purrr::pmap(
  list(path_to_download, sample_metadata$GSM_ID, sample_metadata$tissue, sample_metadata$Model, sample_metadata$cond), 
  function(path, gsm_id, tissue, model, cond){
    print(gsm_id)
    count_matrix <- Read10X(path)
    seurat.obj <- CreateSeuratObject(count_matrix, min.cells=10) %>% 
      RenameCells(add.cell.id = gsm_id)
    seurat.obj[["ref"]] <- "Hing_2022"
    seurat.obj[["case_id"]] <- gsm_id
    seurat.obj[["tissue"]] <- tissue
    seurat.obj[["model"]] <- model
    seurat.obj[["cond"]] <- cond
    seurat.obj <- subset(seurat.obj, subset = nCount_RNA > 600 & nFeature_RNA > 200) # filter out numerous zero-count cells
    seurat.obj <- seurat_quality_control(seurat.obj, species="mouse", mad_cutoff=TRUE)
    return(seurat.obj)
    }
  )
names(seurat.lst) <- sample_metadata$GSM_ID
```

```{r, eval=FALSE}
# Explore the list
seurat.lst
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat.lst, path_to_save_obj_lst)
```

# Comparison between Eµ-hPRMT5 and Eµ-hTCL1 in spleen

We will reanalyse the results according to Figure 3 and Supplemental Figure 2 in the original article.

```{r}
# Select samples of interest
selected_metadata <- sample_metadata %>%
  filter(tissue == "Spl" & Model %in% c("Eµ-hPRMT5", "Eµ-hTCL1"))
selected_metadata
```

```{r}
# Define path to save and load
path_to_save_merged_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_prmt5_tcl1_merged_seu.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_prmt5_tcl1_integrated_seu.rds")
path_to_save_markers <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_prmt5_tcl1_seu_marker.rds")
```

## Merge and integrate the data

```{r, eval=FALSE}
# Merge the objects
interested_lst <- seurat.lst[selected_metadata$GSM_ID]
seurat <- interested_lst[[1]]
for (i in 2:length(interested_lst)) {
  seurat <- merge(x = seurat, y = interested_lst[[i]])
  interested_lst[[i]] <- NA
}
rm(interested_lst)
```

```{r, eval=FALSE}
# Save the merged object
saveRDS(seurat, path_to_save_merged_obj)
```

We used Harmony package for data integration and batch correction, and annotated the cell types using SingleR package.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Reanalysis after merging and data integration 
DefaultAssay(seurat) <- "RNA"
seurat <- JoinLayers(seurat)
seurat <- NormalizeData(seurat)
seurat <- FindVariableFeatures(seurat)
seurat <- ScaleData(seurat)
seurat <- RunPCA(seurat)
seurat <- RunHarmony(seurat, group.by.vars = "case_id", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
seurat <- FindNeighbors(seurat, dims = 1:30, reduction = "harmony")              
seurat <- FindClusters(
    seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Cell annotation
ref <- readRDS(path_to_cell_annotat_ref)
seurat <- singleR_cell_annotation(seurat, ref=ref) 
```

```{r, eval=FALSE}
# Save the integrated object
saveRDS(seurat, path_to_save_integrated_obj)
```

## Visualisation after data integration

```{r}
# Define path
path_to_obj <- path_to_save_integrated_obj

# Load the data
seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=16}
# Plot based on case_id, conditions, cell types, and clusters
case_id.plt <- DimPlot(seurat, reduction = "umap", group.by = "case_id") + 
  labs(y= "UMAP 2", x = "UMAP 1")

model.plt <- DimPlot(seurat, reduction = "umap", group.by = "model") + 
  labs(y= "UMAP 2", x = "UMAP 1")

cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")

case_id.plt + model.plt + cell.plt
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("Hbb-bt", "Hba-a1", "Cd3d", "Cd8a", "Cd4", "Cd19", "Cd79a", "Cd14", "Ncr1"),
            reduction = "umap") 
```

```{r, fig.height=4, fig.width=8}
# Plot by spliting the model (Eµ-hPRMT5 and Eµ-hTCL1)
cell_split.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main", split.by = "model")
cell_split.plt 
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
varied_clusters.plt <- plot_varying_cluster_resolution(seurat, assay_to_use = "RNA")
varied_clusters.plt
```

We used seurat cluster at resolution = 0.3 for further analysis.

```{r, fig.height=4, fig.width=4.5}
# Plot clusters at resolution = 0.3
cluster.plt <- DimPlot(seurat, reduction = "umap", group.by = "RNA_snn_res.0.3", label = TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")
cluster.plt
```

## Cell type annotation

As automatic SingleR annotation labelled cells in more detail, we will manually annotated them.

```{r}
# Define new cluster labels
new.cluster.ids <- c("Cd8+ T", "B", "B", "Cd4+ T", "B", "B", "Cd8+ T", "Monocyte",
                     "B", "Neutrophil", "Cd8+ T", "NK", "Unassigned", "B", "Erythrocyte")

# Map these to the cluster identities (i.e., Idents of the Seurat object)
Idents(seurat) <- "RNA_snn_res.0.3"
names(new.cluster.ids) <- levels(seurat)

# Now map cluster IDs to the manual labels
seurat$manually_annotated <- plyr::mapvalues(
  x = as.character(Idents(seurat)),
  from = names(new.cluster.ids),
  to = new.cluster.ids
)
```

```{r, fig.height=4, fig.width=10}
# Plot clusters and cell types
cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "manually_annotated") +
  labs(y= "UMAP 2", x = "UMAP 1") 

cluster.plt + cell.plt
```

```{r, fig.height=3, fig.width=12}
# Dot plot for biomarker expression in each cell type

# 1. Create meta-data to concatenate clusters and cell types
seurat$cluster_celltype <- paste0("C", seurat$RNA_snn_res.0.3, " - ", seurat$manually_annotated)

# 2. Extract a data frame from Seurat dot plot
features = c("Cd3d", "Cd8a", "Cd4", "Ncr1", "Pdcd1", "Foxp3", "Cd19", "Itgam", "Cd177", "Cd14", "Hbb-bt")
dotdata <- DotPlot(
  seurat,
  features = features,
  group.by = "cluster_celltype"  
)$data

# 3. Preprocess the data frame compatible with ggplot
dotdata <- dotdata %>% 
  tidyr::separate(col = id, into = c("cluster", "celltype"), sep = " - ") %>%
  dplyr::mutate(proportion=pct.exp/100) %>%
  dplyr::rename(
    gene_name = features.plot,
    expression = avg.exp.scaled
  )

# 4. Factor levels
dotdata$cluster <- factor(dotdata$cluster, levels = unique(dotdata$cluster))
dotdata$gene_name <- factor(dotdata$gene_name, levels = rev(features))
celltype_order <- c("Cd8+ T", "B", "Cd4+ T", "NK", "Neutrophil", "Monocyte", "Erythrocyte", "Unassigned")
dotdata$celltype <- factor(dotdata$celltype, levels = celltype_order)

# 5. ggplot for the dot plot
biomarker_dot.plt <- ggplot(dotdata, aes(x = cluster, 
                                         y = gene_name, 
                                         color = expression,
                                         size = proportion)) +
  geom_point() +
  scale_size_area() +
  scale_color_viridis_c() +
  facet_wrap(~celltype, scales = "free_x", nrow=1) +
  theme_minimal(base_size = 8) +
  theme(strip.background = ggh4x::element_part_rect(side = "b", colour = "black", fill = "transparent")) +
  theme(axis.text.y = element_text(face = "italic")) +
  labs(x = NULL, y = NULL, size = "Proportion", color = "Expression")

biomarker_dot.plt
```

## Gene expression analysis between PRMT5 and TCL1 cells

```{r, fig.height=4, fig.width=8}
# Plot clusters by spliting the model (Eµ-hPRMT5 and Eµ-hTCL1)
cluster_split.plt <- DimPlot(seurat, reduction = "umap", group.by = "RNA_snn_res.0.3", split.by = "model", label = TRUE) +
  NoLegend()
cluster_split.plt 
```

From the clustering results, we can see that cluster 4 and 13 of tumour cells predominated in TCL1 model (similar to cluster 3.3 in the original article in Figure 3d) while cluster 5 predominated in PRMT5 model (similar to cluster 3.4 and 3.6 in the original study).

```{r, eval=FALSE}
# Biomarker identification
Idents(seurat) <- "RNA_snn_res.0.3"
seurat.markers <- FindAllMarkers(seurat, only.pos = TRUE)
```

```{r, eval=FALSE}
# Save the marker result
saveRDS(seurat.markers, path_to_save_markers)
```

## Heatmap for top 500 markers

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(ComplexHeatmap)
library(circlize)
library(grid)
```

```{r}
# Take top 500 per cluster
seurat.markers <- readRDS(path_to_save_markers)
top_n_genes <- seurat.markers %>%
  group_by(cluster) %>%
  top_n(n = 500, wt = avg_log2FC)

# Filter to specific clusters of interest (e.g. "4", "5")
top_markers <- top_n_genes %>%
  filter(cluster %in% c("4", "5", "13")) %>%  
  pull(gene) %>%
  unique()
```

```{r, message=FALSE, warning=FALSE}
# Extract scaled expression matrix
Bcell_seurat <- subset(seurat, subset = RNA_snn_res.0.3 %in% c(1, 2, 4, 5, 8, 13))
avg_expr <- AverageExpression(
  Bcell_seurat,
  features = top_markers,
  assays = "RNA",     
  slot = "data"       
)$RNA

# Scale genes 
expr_mat <- t(scale(t(avg_expr)))
colnames(expr_mat) <- c("C1", "C2", "C4", "C5", "C8", "C13")
```

```{r}
# Set up colours and annotation
colfun <- circlize::colorRamp2(
  breaks = c(min(expr_mat), 0, max(expr_mat)),
  colors = c("#2166AC", "white", "#B2182B")
)

# highlight genes (as in your example)
highlights <- c("Hmgb1","Cd69","Hmgb2","Junb","Top2a","Ccr7","Birc5","Cd83",
                "Atf4","Tuba1b","Tubb5","Hspa5","Ccdc34","Ube2c","Il2rg",
                "Cks1b","Smc2","Stmn1","Tyms","Ubc","Gapdh","Apoe","Xrcc1",
                "Cd79a","Zfp36","Ppp1r15a","Ldha","Mki67","Cdk1","Gpx4","Btg1",
                "Itm2b","Xbp1","Rel","Aurkb","Eif5a","C1qbp","Cd79b","Tk1",
                "Nfkbia","Hsp90aa1","Cenpm","Nr4a1","Pfdn5","Eif4a2","Pim1",
                "Cdca3","Pclaf","Ccnb2","Rrm2")

fig_anno <- rowAnnotation(
  link = anno_mark(
    at = which(rownames(expr_mat) %in% highlights),
    labels = rownames(expr_mat)[rownames(expr_mat) %in% highlights],
    labels_gp = gpar(fontsize = 5),
    padding = 1.2
  )
)
```

```{r, fig.height=6, fig.width=5}
# Draw heatmap
heatmap.plt <- ComplexHeatmap::Heatmap(
  expr_mat,
  col = colfun,
  name = "Expression",
  show_row_names = FALSE,
  right_annotation = fig_anno,
  row_dend_width = unit(3, "mm"),
  column_dend_height = unit(3, "mm"),
  heatmap_legend_param = list(
    legend_direction = "vertical",
    title_position = "lefttop-rot",
    title_gp = gpar(fontsize = 6)
  )
)

draw(heatmap.plt)
```

In original study they reported these genes in the top 50 genes while we can show these in the top 500 genes although the genes were in the same clusters (PRMT5 and TCL1 cells). We think that using different analysis tools such as monocle and seurat can result in different results.

## Enrichment analysis

We also performed gene list-based enrichment analysis based on the top 500 markers identified from cluster 5, 4, and 13.

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(clusterProfiler)
library(org.Mm.eg.db)  # Mouse annotation database
library(enrichplot)    
```

### TCL1 cells

#### Cluster 4

```{r, message=FALSE, warning=FALSE}
# Extract top 500 gene name from cluster 4
top_markers_4 <- top_n_genes %>%
  filter(cluster == "4") %>%  
  pull(gene)

# Convert symbols to Entrez IDs
gene_df <- bitr(top_markers_4, fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Mm.eg.db)

entrez_genes <- gene_df$ENTREZID
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
gobp_enrich <- enrichGO(
  gene = entrez_genes,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE       
)

# View results
gobb_df <- as.data.frame(gobp_enrich)
DT::datatable(gobb_df, options = list(scrollX = TRUE))
```

```{r, fig.height=9, fig.width=7}
# Dotplot of top GO terms
dotplot(gobp_enrich, showCategory = 20) + ggtitle("GO Biological Process")
```

#### Cluster 13

```{r, message=FALSE, warning=FALSE}
# Extract top 500 gene name from cluster 13
top_markers_13 <- top_n_genes %>%
  filter(cluster == "13") %>%  
  pull(gene)

# Convert symbols to Entrez IDs
gene_df <- bitr(top_markers_13, fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Mm.eg.db)

entrez_genes <- gene_df$ENTREZID
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
gobp_enrich <- enrichGO(
  gene = entrez_genes,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE       
)

# View results
gobb_df <- as.data.frame(gobp_enrich)
DT::datatable(gobb_df, options = list(scrollX = TRUE))
```

```{r, fig.height=4, fig.width=7}
# Dotplot of top GO terms
dotplot(gobp_enrich, showCategory = 20) + ggtitle("GO Biological Process")
```

### PRMT5 cells

```{r, message=FALSE, warning=FALSE}
# Extract top 500 gene name from cluster 5
top_markers_5 <- top_n_genes %>%
  filter(cluster == "5") %>%  
  pull(gene)

# Convert symbols to Entrez IDs
gene_df <- bitr(top_markers_5, fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Mm.eg.db)

entrez_genes <- gene_df$ENTREZID
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
gobp_enrich <- enrichGO(
  gene = entrez_genes,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE       
)

# View results
gobb_df <- as.data.frame(gobp_enrich)
DT::datatable(gobb_df, options = list(scrollX = TRUE))
```

```{r, fig.height=9, fig.width=7}
# Dotplot of top GO terms
dotplot(gobp_enrich, showCategory = 20) + ggtitle("GO Biological Process")
```

## Visulaise other genes of interest

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=20}
# Visualise feature expression according to the original study

# 1. Define genes of interest
features <- c("Ighm", "Pax5", "Ighd", "Ebf1", "Cd93", "Cd69", "Spn", "Myc", "Mki67")

# 2. Compute global min and max expression across all selected genes
expr_matrix <- GetAssayData(seurat, assay = "RNA", layer = "data")
expr_min <- min(expr_matrix)
expr_max <- max(expr_matrix)

# 3. Generate plots
feature_plots <- lapply(features, function(gene) {
  
  # Return individual ggplots for each split
  plt_list <- FeaturePlot(
    seurat,
    features = gene,
    reduction = "umap",
    pt.size = 1,
    split.by = "model",
    combine = FALSE,         # IMPORTANT: return list of ggplots
    keep.scale = "feature"
  )
  
  # Apply the same global colour scale and labels to each split
  plt_list <- lapply(plt_list, function(x) {
    x + 
      scale_color_viridis_c(
        option = "viridis",
        limits = c(expr_min, expr_max)  # same scale for all
      ) +
      labs(y = "UMAP 2", x = "UMAP 1", title = gene)
  })
  
  # Combine splits for the same gene
  wrap_plots(plt_list)
})

# 4. Combine all genes and keep only one legend
combined_plot <- wrap_plots(feature_plots, ncol = 3, guides = "collect")

# 5. Make legend appear once, at the bottom
combined_plot & theme(legend.position = "bottom")
```

# Comparison between Eµ-hPRMT5/hTCL1 and Eµ-hTCL1 in spleen

We will reanalyse the results according to Figure 5 and Supplemental Figure 4 in the original article.

```{r}
# Select samples of interest
selected_metadata <- sample_metadata %>%
  filter(tissue == "Spl" & Model %in% c("Eµ-hPRMT5/hTCL1", "Eµ-hTCL1"))
selected_metadata
```

```{r}
# Define path to save and load
path_to_save_merged_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_rt_cll_spl_merged_seu.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_rt_cll_spl_integrated_seu.rds")
path_to_save_markers <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_rt_cll_spl_seu_marker.rds")
```

## Merge and integrate the data

```{r, eval=FALSE}
# Merge the objects
interested_lst <- seurat.lst[selected_metadata$GSM_ID]
seurat <- interested_lst[[1]]
for (i in 2:length(interested_lst)) {
  seurat <- merge(x = seurat, y = interested_lst[[i]])
  interested_lst[[i]] <- NA
}
rm(interested_lst)
```

```{r, eval=FALSE}
# Save the merged object
saveRDS(seurat, path_to_save_merged_obj)
```

We used Harmony package for data integration and batch correction, and annotated the cell types using SingleR package.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Reanalysis after merging and data integration 
DefaultAssay(seurat) <- "RNA"
seurat <- JoinLayers(seurat)
seurat <- NormalizeData(seurat)
seurat <- FindVariableFeatures(seurat)
seurat <- ScaleData(seurat)
seurat <- RunPCA(seurat)
seurat <- RunHarmony(seurat, group.by.vars = "case_id", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
seurat <- FindNeighbors(seurat, dims = 1:30, reduction = "harmony")              
seurat <- FindClusters(
    seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )
seurat <- singleR_cell_annotation(seurat, ref=ref) 
```

```{r, eval=FALSE}
# Save the integrated object
saveRDS(seurat, path_to_save_integrated_obj)
```

## Visualisation after data integration

```{r}
# Define path
path_to_obj <- path_to_save_integrated_obj

# Load the data
seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=16}
# Plot based on case_id, conditions, cell types, and clusters
case_id.plt <- DimPlot(seurat, reduction = "umap", group.by = "case_id") + 
  labs(y= "UMAP 2", x = "UMAP 1")

model.plt <- DimPlot(seurat, reduction = "umap", group.by = "model") + 
  labs(y= "UMAP 2", x = "UMAP 1")

cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")

case_id.plt + model.plt + cell.plt
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("Hbb-bt", "Hba-a1", "Cd3d", "Cd8a", "Cd4", "Cd19", "Cd79a", "Cd14", "Nkg7"),
            reduction = "umap") 
```

```{r, fig.height=4, fig.width=8}
# Plot by spliting the model (Eµ-hPRMT5/hTCL1 and Eµ-hTCL1)
cell_split.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main", split.by = "model")
cell_split.plt 
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
varied_clusters.plt <- plot_varying_cluster_resolution(seurat, assay_to_use = "RNA")
varied_clusters.plt
```

We used seurat cluster at resolution = 0.3 for further analysis.

```{r, fig.height=4, fig.width=4.5}
# Plot clusters at resolution = 0.3
cluster.plt <- DimPlot(seurat, reduction = "umap", group.by = "RNA_snn_res.0.3", label = TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")
cluster.plt
```

## Cell type annotation

As automatic SingleR annotation labelled cells in more detail, we will manually annotated them.

```{r}
# Define new cluster labels
new.cluster.ids <- c("B", "Cd8+ T", "Cd4+ T", "B", "B", "B", "B", "Monocyte",
                     "B", "Cd8+ T", "Unassigned", "B", "NK", "Neutrophil", "Neutrophil", "Erythrocyte")

# Map these to the cluster identities (i.e., Idents of the Seurat object)
Idents(seurat) <- "RNA_snn_res.0.3"
names(new.cluster.ids) <- levels(seurat)

# Now map cluster IDs to the manual labels
seurat$manually_annotated <- plyr::mapvalues(
  x = as.character(Idents(seurat)),
  from = names(new.cluster.ids),
  to = new.cluster.ids
)
```

```{r, fig.height=4, fig.width=10}
# Plot clusters and cell types
cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "manually_annotated") +
  labs(y= "UMAP 2", x = "UMAP 1") 

cluster.plt + cell.plt
```

```{r, fig.height=3, fig.width=12}
# Dot plot for biomarker expression in each cell type

# 1. Create meta-data to concatenate clusters and cell types
seurat$cluster_celltype <- paste0("C", seurat$RNA_snn_res.0.3, " - ", seurat$manually_annotated)

# 2. Extract a data frame from Seurat dot plot
features = c("Cd3d", "Cd8a", "Cd4", "Ncr1", "Pdcd1", "Foxp3", "Cd19", "Itgam", "Cd177", "Cd14", "Hbb-bt")
dotdata <- DotPlot(
  seurat,
  features = features,
  group.by = "cluster_celltype"  
)$data

# 3. Preprocess the data frame compatible with ggplot
dotdata <- dotdata %>% 
  tidyr::separate(col = id, into = c("cluster", "celltype"), sep = " - ") %>%
  dplyr::mutate(proportion=pct.exp/100) %>%
  dplyr::rename(
    gene_name = features.plot,
    expression = avg.exp.scaled
  )

# 4. Factor levels
dotdata$cluster <- factor(dotdata$cluster, levels = unique(dotdata$cluster))
dotdata$gene_name <- factor(dotdata$gene_name, levels = rev(features))
celltype_order <- c("Cd8+ T", "B", "Cd4+ T", "NK", "Neutrophil", "Monocyte", "Erythrocyte", "Unassigned")
dotdata$celltype <- factor(dotdata$celltype, levels = celltype_order)

# 5. ggplot for the dot plot
biomarker_dot.plt <- ggplot(dotdata, aes(x = cluster, 
                                         y = gene_name, 
                                         color = expression,
                                         size = proportion)) +
  geom_point() +
  scale_size_area() +
  scale_color_viridis_c() +
  facet_wrap(~celltype, scales = "free_x", nrow=1) +
  theme_minimal(base_size = 8) +
  theme(strip.background = ggh4x::element_part_rect(side = "b", colour = "black", fill = "transparent")) +
  theme(axis.text.y = element_text(face = "italic")) +
  labs(x = NULL, y = NULL, size = "Proportion", color = "Expression")

biomarker_dot.plt
```

## Gene expression analysis between PRMT5/TCL1 and TCL1 cells

```{r, fig.height=4, fig.width=8}
# Plotc lusters by spliting the model (Eµ-hPRMT5/hTCL1 and Eµ-hTCL1)
cluster_split.plt <- DimPlot(seurat, reduction = "umap", group.by = "RNA_snn_res.0.3", split.by = "model", label = TRUE) +
  NoLegend()
cluster_split.plt 
```

From the clustering results, cluster 6 of tumour cells predominates in the PRMT5/TCL1 model, similar to cluster 5.6 in the original study.

```{r, eval=FALSE}
# Biomarker identification
Idents(seurat) <- "RNA_snn_res.0.3"
seurat.markers <- FindAllMarkers(seurat, only.pos = TRUE)
```

```{r, eval=FALSE}
# Save the marker result
saveRDS(seurat.markers, path_to_save_markers)
```

## Heatmap for top 5000 markers

```{r}
# Take top 5000 per cluster
seurat.markers <- readRDS(path_to_save_markers)
top_n_genes <- seurat.markers %>%
  group_by(cluster) %>%
  top_n(n = 5000, wt = avg_log2FC)

# Filter to specific clusters of interest 
top_markers <- top_n_genes %>%
  filter(cluster %in% c("6", "5", "11")) %>%  
  pull(gene) %>%
  unique()
```

```{r, message=FALSE, warning=FALSE}
# Extract scaled expression matrix
Bcell_seurat <- subset(seurat, subset = RNA_snn_res.0.3 %in% c(0, 3, 4, 5, 6, 8, 11))
avg_expr <- AverageExpression(
  Bcell_seurat,
  features = top_markers,
  assays = "RNA",     
  slot = "data"       
)$RNA

# Scale genes 
expr_mat <- t(scale(t(avg_expr)))
colnames(expr_mat) <- c("C0", "C3", "C4", "C5", "C6", "C8", "C11")
```

```{r}
# Set up colours and annotation
colfun <- circlize::colorRamp2(
  breaks = c(min(expr_mat), 0, max(expr_mat)),
  colors = c("#2166AC", "white", "#B2182B")
)

# highlight genes (as in your example)
highlights <-
  c(
    "Tubb5",
    "Tuba1b",
    "Npm1",
    "Marcksl1",
    "Actb",
    "Gapdh",
    "Napsa",
    "Cd37",
    "Ldha",
    "Cd19",
    "Coro1a",
    "Klf2",
    "Eif5a",
    "Hnrnpk",
    "Rac2",
    "Ms4a1",
    "Rpl3"
  )

fig_anno <- rowAnnotation(
  link = anno_mark(
    at = which(rownames(expr_mat) %in% highlights),
    labels = rownames(expr_mat)[rownames(expr_mat) %in% highlights],
    labels_gp = gpar(fontsize = 5),
    padding = 1.2
  )
)
```

```{r, fig.height=6, fig.width=5}
# Draw heatmap
heatmap.plt <- ComplexHeatmap::Heatmap(
  expr_mat,
  col = colfun,
  name = "Expression",
  show_row_names = FALSE,
  right_annotation = fig_anno,
  row_dend_width = unit(3, "mm"),
  column_dend_height = unit(3, "mm"),
  heatmap_legend_param = list(
    legend_direction = "vertical",
    title_position = "lefttop-rot",
    title_gp = gpar(fontsize = 6)
  )
)

draw(heatmap.plt)
```

## Enrichment analysis

We also performed gene list-based enrichment analysis based on the top 5000 markers identified from cluster 6.

```{r, message=FALSE, warning=FALSE}
# Extract top 5000 gene name from cluster 6
top_markers_6 <- top_n_genes %>%
  filter(cluster == "6") %>%  
  pull(gene)

# Convert symbols to Entrez IDs
gene_df <- bitr(top_markers_5, fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Mm.eg.db)

entrez_genes <- gene_df$ENTREZID
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
gobp_enrich <- enrichGO(
  gene = entrez_genes,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE       
)

# View results
gobb_df <- as.data.frame(gobp_enrich)
DT::datatable(gobb_df, options = list(scrollX = TRUE))
```

```{r, fig.height=9, fig.width=7}
# Dotplot of top GO terms
dotplot(gobp_enrich, showCategory = 20) + ggtitle("GO Biological Process")
```

## Visulaise other genes of interest

```{r, message=FALSE, warning=FALSE, fig.height=12, fig.width=15}
# Visualise feature expression according to the original study

# 1. Define genes of interest
features <- c("Myc", "Mki67", "Egr1", "Cxcr5", "Ccr7", "Il10", "Ctla4", "Cd274")

# 2. Compute global min and max expression across all selected genes
expr_matrix <- GetAssayData(seurat, assay = "RNA", layer = "data")
expr_min <- min(expr_matrix)
expr_max <- max(expr_matrix)

# 3. Generate plots
feature_plots <- lapply(features, function(gene) {
  
  # Return individual ggplots for each split
  plt_list <- FeaturePlot(
    seurat,
    features = gene,
    reduction = "umap",
    pt.size = 1,
    split.by = "model",
    combine = FALSE,         # IMPORTANT: return list of ggplots
    keep.scale = "feature"
  )
  
  # Apply the same global colour scale and labels to each split
  plt_list <- lapply(plt_list, function(x) {
    x + 
      scale_color_viridis_c(
        option = "viridis",
        limits = c(expr_min, expr_max)  # same scale for all
      ) +
      labs(y = "UMAP 2", x = "UMAP 1", title = gene)
  })
  
  # Combine splits for the same gene
  wrap_plots(plt_list)
})

# 4. Combine all genes and keep only one legend
combined_plot <- wrap_plots(feature_plots, ncol = 2, guides = "collect")

# 5. Make legend appear once, at the bottom
combined_plot & theme(legend.position = "bottom")
```

# Comparison between Eµ-hPRMT5/hTCL1 and Eµ-hTCL1 in lymph nodes

We will reanalyse the results according to Supplemental Figure 5 in the original article.

```{r}
# Select samples of interest
selected_metadata <- sample_metadata %>%
  filter(tissue == "LN" & Model %in% c("Eµ-hPRMT5/hTCL1", "Eµ-hTCL1"))
selected_metadata
```

```{r}
# Define path to save and load
path_to_save_merged_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_rt_cll_ln_merged_seu.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_rt_cll_ln_integrated_seu.rds")
path_to_save_markers <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_rt_cll_ln_seu_marker.rds")
```

## Merge and integrate the data

```{r, eval=FALSE}
# Merge the objects
interested_lst <- seurat.lst[selected_metadata$GSM_ID]
seurat <- interested_lst[[1]]
for (i in 2:length(interested_lst)) {
  seurat <- merge(x = seurat, y = interested_lst[[i]])
  interested_lst[[i]] <- NA
}
rm(interested_lst)
```

```{r, eval=FALSE}
# Save the merged object
saveRDS(seurat, path_to_save_merged_obj)
```

We used Harmony package for data integration and batch correction, and annotated the cell types using SingleR package.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Reanalysis after merging and data integration 
DefaultAssay(seurat) <- "RNA"
seurat <- JoinLayers(seurat)
seurat <- NormalizeData(seurat)
seurat <- FindVariableFeatures(seurat)
seurat <- ScaleData(seurat)
seurat <- RunPCA(seurat)
seurat <- RunHarmony(seurat, group.by.vars = "case_id", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
seurat <- FindNeighbors(seurat, dims = 1:30, reduction = "harmony")              
seurat <- FindClusters(
    seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )
seurat <- singleR_cell_annotation(seurat, ref=ref) 
```

```{r, eval=FALSE}
# Save the integrated object
saveRDS(seurat, path_to_save_integrated_obj)
```

## Visualisation after data integration

```{r}
# Define path
path_to_obj <- path_to_save_integrated_obj

# Load the data
seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=17}
# Plot based on case_id, conditions, cell types, and clusters
case_id.plt <- DimPlot(seurat, reduction = "umap", group.by = "case_id") + 
  labs(y= "UMAP 2", x = "UMAP 1")

model.plt <- DimPlot(seurat, reduction = "umap", group.by = "model") + 
  labs(y= "UMAP 2", x = "UMAP 1")

cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")

case_id.plt + model.plt + cell.plt
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("Hbb-bt", "Hba-a1", "Cd3d", "Cd8a", "Cd4", "Cd19", "Cd79a", "Cd14", "Nkg7"),
            reduction = "umap") 
```

```{r, fig.height=4, fig.width=8.5}
# Plot by spliting the model (Eµ-hPRMT5/hTCL1 and Eµ-hTCL1)
cell_split.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main", split.by = "model")
cell_split.plt 
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
varied_clusters.plt <- plot_varying_cluster_resolution(seurat, assay_to_use = "RNA")
varied_clusters.plt
```

We used seurat cluster at resolution = 0.3 for further analysis.

```{r, fig.height=4, fig.width=4.5}
# Plot clusters at resolution = 0.3

# Reorder and relabel clusters numerically
seurat$RNA_snn_res.0.3 <- factor(
  seurat$RNA_snn_res.0.3,
  levels = as.character(sort(as.numeric(levels(seurat$RNA_snn_res.0.3))))
)

cluster.plt <- DimPlot(seurat, reduction = "umap", group.by = "RNA_snn_res.0.3", label = TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

## Cell type annotation

As automatic SingleR annotation labelled cells in more detail, we will manually annotated them.

```{r}
# Define new cluster labels
new.cluster.ids <- c("Cd8+ T", "B", "B", "Cd4+ T", "Cd8+ T", "B", "B", 
                     "B", "B", "Cd8+ T", "B", "Monocyte", "B", "NK")

# Map these to the cluster identities (i.e., Idents of the Seurat object)
Idents(seurat) <- "RNA_snn_res.0.3"
names(new.cluster.ids) <- levels(seurat)

# Now map cluster IDs to the manual labels
seurat$manually_annotated <- plyr::mapvalues(
  x = as.character(Idents(seurat)),
  from = names(new.cluster.ids),
  to = new.cluster.ids
)
```

```{r, fig.height=4, fig.width=10}
# Plot clusters and cell types
cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "manually_annotated") +
  labs(y= "UMAP 2", x = "UMAP 1") 

cluster.plt + cell.plt
```

```{r, fig.height=3, fig.width=12}
# Dot plot for biomarker expression in each cell type

# 1. Create meta-data to concatenate clusters and cell types
seurat$cluster_celltype <- paste0("C", seurat$RNA_snn_res.0.3, " - ", seurat$manually_annotated)

# 2. Extract a data frame from Seurat dot plot
features = c("Cd3d", "Cd8a", "Cd4", "Ncr1", "Pdcd1", "Foxp3", "Cd19", "Itgam", "Cd177", "Cd14", "Hbb-bt")
dotdata <- DotPlot(
  seurat,
  features = features,
  group.by = "cluster_celltype"  
)$data

# 3. Preprocess the data frame compatible with ggplot
dotdata <- dotdata %>% 
  tidyr::separate(col = id, into = c("cluster", "celltype"), sep = " - ") %>%
  dplyr::mutate(proportion=pct.exp/100) %>%
  dplyr::rename(
    gene_name = features.plot,
    expression = avg.exp.scaled
  )

# 4. Factor levels
dotdata$cluster <- factor(dotdata$cluster, levels = unique(dotdata$cluster))
dotdata$gene_name <- factor(dotdata$gene_name, levels = rev(features))
celltype_order <- c("Cd8+ T", "B", "Cd4+ T", "NK", "Neutrophil", "Monocyte", "Erythrocyte", "Unassigned")
dotdata$celltype <- factor(dotdata$celltype, levels = celltype_order)

# 5. ggplot for the dot plot
biomarker_dot.plt <- ggplot(dotdata, aes(x = cluster, 
                                         y = gene_name, 
                                         color = expression,
                                         size = proportion)) +
  geom_point() +
  scale_size_area() +
  scale_color_viridis_c() +
  facet_wrap(~celltype, scales = "free_x", nrow=1) +
  theme_minimal(base_size = 8) +
  theme(strip.background = ggh4x::element_part_rect(side = "b", colour = "black", fill = "transparent")) +
  theme(axis.text.y = element_text(face = "italic")) +
  labs(x = NULL, y = NULL, size = "Proportion", color = "Expression")

biomarker_dot.plt
```

## Gene expression analysis between PRMT5/TCL1 and TCL1 cells

```{r, fig.height=4, fig.width=7.5}
# Plotc lusters by spliting the model (Eµ-hPRMT5/hTCL1 and Eµ-hTCL1)
cluster_split.plt <- DimPlot(seurat, reduction = "umap", group.by = "RNA_snn_res.0.3", split.by = "model", label = TRUE) +
  NoLegend()
cluster_split.plt 
```

From the clustering results, we can see that cluster 2, 6 and 7 of tumour cells predominated in PRMT5/TCL1 model (similar to cluster LN2 and LN7 in the original article in Supplemental Figure 5e) while cluster 5 predominated in TCL1 model (similar to cluster LN4 in the original study).

```{r, eval=FALSE}
# Biomarker identification
Idents(seurat) <- "RNA_snn_res.0.3"
seurat.markers <- FindAllMarkers(seurat, only.pos = TRUE)
```

```{r, eval=FALSE}
# Save the marker result
saveRDS(seurat.markers, path_to_save_markers)
```

## Heatmap for top 5000 markers

```{r}
# Take top 5000 per cluster
seurat.markers <- readRDS(path_to_save_markers)
top_n_genes <- seurat.markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5000)

# Filter to specific clusters of interest 
top_markers <- top_n_genes %>%
  filter(cluster %in% c("2", "5", "6", "7")) %>%  
  pull(gene) %>%
  unique()
```

```{r, message=FALSE, warning=FALSE}
# Extract scaled expression matrix
Bcell_seurat <- subset(seurat, subset = RNA_snn_res.0.3 %in% c(1, 2, 5, 6, 7, 8, 10, 12))
avg_expr <- AverageExpression(
  Bcell_seurat,
  features = top_markers,
  assays = "RNA",     
  slot = "data"       
)$RNA

# Scale genes 
expr_mat <- t(scale(t(avg_expr)))
colnames(expr_mat) <- c("C1", "C2", "C5", "C6", "C7", "C8", "C10", "C12")
```

```{r}
# Set up colours and annotation
colfun <- circlize::colorRamp2(
  breaks = c(min(expr_mat), 0, max(expr_mat)),
  colors = c("#2166AC", "white", "#B2182B")
)

# highlight genes (as in your example)
highlights <-
  c(
    "Rac2",
    "Coro1a",
    "Ms4a1",
    "Cd37",
    "Actb",
    "Cd83",
    "Cd79b",
    "Napsa",
    "Cks1b",
    "Cenpm",
    "Mki67",
    "Ube2c",
    "Cdk1",
    "Birc5",
    "Hmgb1",
    "Tubb5",
    "Smc2",
    "Aurkb",
    "Stmn1",
    "Top2a",
    "Hmgb2",
    "Tk1",
    "Tyms",
    "Tuba1b"
  )

fig_anno <- rowAnnotation(
  link = anno_mark(
    at = which(rownames(expr_mat) %in% highlights),
    labels = rownames(expr_mat)[rownames(expr_mat) %in% highlights],
    labels_gp = gpar(fontsize = 5),
    padding = 1.2
  )
)
```

```{r, fig.height=6, fig.width=5}
# Draw heatmap
heatmap.plt <- ComplexHeatmap::Heatmap(
  expr_mat,
  col = colfun,
  name = "Expression",
  show_row_names = FALSE,
  right_annotation = fig_anno,
  row_dend_width = unit(3, "mm"),
  column_dend_height = unit(3, "mm"),
  heatmap_legend_param = list(
    legend_direction = "vertical",
    title_position = "lefttop-rot",
    title_gp = gpar(fontsize = 6)
  )
)

draw(heatmap.plt)
```

## Enrichment analysis

We also performed gene list-based enrichment analysis based on the top 5000 markers identified from cluster 2, 5, 6, and 7.

### TCL1 cells

```{r, message=FALSE, warning=FALSE}
# Extract top 5000 gene name from cluster 5
top_markers_5 <- top_n_genes %>%
  filter(cluster == "5") %>%  
  pull(gene)

# Convert symbols to Entrez IDs
gene_df <- bitr(top_markers_5, fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Mm.eg.db)

entrez_genes <- gene_df$ENTREZID
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
gobp_enrich <- enrichGO(
  gene = entrez_genes,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE       
)

# View results
gobb_df <- as.data.frame(gobp_enrich)
DT::datatable(gobb_df, options = list(scrollX = TRUE))
```

```{r, fig.height=9, fig.width=7}
# Dotplot of top GO terms
dotplot(gobp_enrich, showCategory = 20) + ggtitle("GO Biological Process")
```

### PRMT5/TCL1 cells

#### Cluster 2

```{r, message=FALSE, warning=FALSE}
# Extract top 5000 gene name from cluster 2
top_markers_2 <- top_n_genes %>%
  filter(cluster == "2") %>%  
  pull(gene)

# Convert symbols to Entrez IDs
gene_df <- bitr(top_markers_2, fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Mm.eg.db)

entrez_genes <- gene_df$ENTREZID
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
gobp_enrich <- enrichGO(
  gene = entrez_genes,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE       
)

# View results
gobb_df <- as.data.frame(gobp_enrich)
DT::datatable(gobb_df, options = list(scrollX = TRUE))
```

```{r, fig.height=13, fig.width=7}
# Dotplot of top GO terms
dotplot(gobp_enrich, showCategory = 20) + ggtitle("GO Biological Process")
```

#### Cluster 6

```{r, message=FALSE, warning=FALSE}
# Extract top 5000 gene name from cluster 6
top_markers_6 <- top_n_genes %>%
  filter(cluster == "6") %>%  
  pull(gene)

# Convert symbols to Entrez IDs
gene_df <- bitr(top_markers_6, fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Mm.eg.db)

entrez_genes <- gene_df$ENTREZID
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
gobp_enrich <- enrichGO(
  gene = entrez_genes,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE       
)

# View results
gobb_df <- as.data.frame(gobp_enrich)
DT::datatable(gobb_df, options = list(scrollX = TRUE))
```

```{r, fig.height=13, fig.width=7}
# Dotplot of top GO terms
dotplot(gobp_enrich, showCategory = 20) + ggtitle("GO Biological Process")
```

#### Cluster 7

```{r, message=FALSE, warning=FALSE}
# Extract top 5000 gene name from cluster 7
top_markers_7 <- top_n_genes %>%
  filter(cluster == "7") %>%  
  pull(gene)

# Convert symbols to Entrez IDs
gene_df <- bitr(top_markers_7, fromType = "SYMBOL",
                toType = "ENTREZID",
                OrgDb = org.Mm.eg.db)

entrez_genes <- gene_df$ENTREZID
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
gobp_enrich <- enrichGO(
  gene = entrez_genes,
  OrgDb = org.Mm.eg.db,
  keyType = "ENTREZID",
  ont = "BP",          
  pAdjustMethod = "BH",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  readable = TRUE       
)

# View results
gobb_df <- as.data.frame(gobp_enrich)
DT::datatable(gobb_df, options = list(scrollX = TRUE))
```

```{r, fig.height=13, fig.width=7}
# Dotplot of top GO terms
dotplot(gobp_enrich, showCategory = 20) + ggtitle("GO Biological Process")
```

## Visulaise other genes of interest

```{r, message=FALSE, warning=FALSE, fig.height=20, fig.width=20}
# Visualise feature expression according to the original study

# 1. Define genes of interest
features <- c("Myc", "Ly6a", "Mki67", "Ccr7", "Cxcr5", "Il10", "Ctla4", "Egr1", "Cd274", "Cd93")

# 2. Compute global min and max expression across all selected genes
expr_matrix <- GetAssayData(seurat, assay = "RNA", layer = "data")
expr_min <- min(expr_matrix)
expr_max <- max(expr_matrix)

# 3. Generate plots
feature_plots <- lapply(features, function(gene) {
  
  # Return individual ggplots for each split
  plt_list <- FeaturePlot(
    seurat,
    features = gene,
    reduction = "umap",
    pt.size = 1,
    split.by = "model",
    combine = FALSE,         # IMPORTANT: return list of ggplots
    keep.scale = "feature"
  )
  
  # Apply the same global colour scale and labels to each split
  plt_list <- lapply(plt_list, function(x) {
    x + 
      scale_color_viridis_c(
        option = "viridis",
        limits = c(0, 2)  # same scale for all
      ) +
      labs(y = "UMAP 2", x = "UMAP 1", title = gene)
  })
  
  # Combine splits for the same gene
  wrap_plots(plt_list)
})

# 4. Combine all genes and keep only one legend
combined_plot <- wrap_plots(feature_plots, ncol = 2, guides = "collect")

# 5. Make legend appear once, at the bottom
combined_plot & theme(legend.position = "bottom")
```

# Session Information

```{r}
sessionInfo()
```


