---
title: "Parry_2023a_reanalysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This source code we reanalysed single-cell RNA-seq datasets for RT patients from [Parry et. al. (2023)](https://www.nature.com/articles/s41591-022-02113-6). The original study sorted cell types, such as CLL and RT tumour cells, using flow cytometry before single-cell sequencing. Then, the samples were based on the sorted cells.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(dplyr)
library(DoubletFinder)
library(harmony)
library(SingleR)
library(celldex)
library(clustree)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_metadata <- here::here("scRNAseq/Parry_2023a/sample_metadata.csv")
path_to_count_matrices <- here::here("scRNAseq/Parry_2023a/expression_data/")
path_to_save_obj_lst <- here::here("scRNAseq/Parry_2023a/reanalysis/R_objects/Parry_2023a_reanalysis_seu_lst.rds")
path_to_save_merged_obj_lst <- here::here("scRNAseq/Parry_2023a/reanalysis/R_objects/Parry_2023a_reanalysis_merged_seu_lst.rds")
path_to_save_harmony_integrated_obj_lst <- here::here("scRNAseq/Parry_2023a/reanalysis/R_objects/Parry_2023a_reanalysis_harmony_integrated_seu_lst.rds")
path_to_save_cca_integrated_obj_lst <- here::here("scRNAseq/Parry_2023a/reanalysis/R_objects/Parry_2023a_reanalysis_cca_integrated_seu_lst.rds")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the sample metadata
sample_metadata <- read.csv(path_to_metadata)
DT::datatable(sample_metadata, options = list(scrollX = TRUE))
```

We will import count matrices and convert them to Seurat objects with quality control and pre-processing. We used the parameters same as the original study.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create the objects and collect in a list
path_to_download <- str_c(path_to_count_matrices, "/", sample_metadata$Run, "/count_matrices/outs/filtered_feature_bc_matrix/")
seurat.lst <- purrr::pmap(
  list(path_to_download, 
       sample_metadata$Run, 
       sample_metadata$Case, 
       sample_metadata$histological_type,
       sample_metadata$Condition, 
       sample_metadata$Tissue), 
  function(path, sra_run, case, histological_type, cond, tissue){
    count_matrix <- Read10X(path)
    seurat.obj <- CreateSeuratObject(count_matrix) %>% 
      RenameCells(add.cell.id = sra_run)
    seurat.obj[["ref"]] <- "Parry_2023a"
    seurat.obj[["case_id"]] <- sra_run
    seurat.obj[["histological_type"]] <- histological_type
    seurat.obj[["case"]] <- case
    seurat.obj[["cond"]] <- cond
    seurat.obj[["tissue"]] <- tissue
    seurat.obj <- seurat_quality_control(seurat.obj, 
                                         min_nCount = 500, max_nCount = 50000,
                                         min_nFeature = 0, max_nFeature = 20000,
                                         max_percent_mt = 10, 
                                         min_percent_ribo = 0, max_percent_ribo = 100,
                                         mad_cutoff=FALSE)
    seurat.obj <- seurat_preprocess(seurat.obj)
    if (ncol(seurat.obj) > 500) {  # pass doublet detection if objects have too low cells
      seurat.obj <- doublet_filter(seurat.obj, sct=FALSE)
    } else {
      message("Skipping doublet filtering for object with <500 cells")
    }
    return(seurat.obj)
    }
  )
names(seurat.lst) <- sample_metadata$Run
```

```{r, eval=FALSE}
# See the preprocessing results
seurat.lst
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat.lst, path_to_save_obj_lst)
```

# Merge the objects

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Merge the objects based on individual patients
seurat_merged_case.lst <- vector("list", 3)  # create empty list of length 3
names(seurat_merged_case.lst) <- c("patient_43", "patient_4", "patient_10")

# Loop to merge the Seurat objects per patient
for (i in c(43, 4, 10)) {
  sra_related_case <- sample_metadata[sample_metadata$Case == i, "Run"]
  seurat_sublst <- seurat.lst[sra_related_case]
  
  # Start with the first object
  seurat.obj <- seurat_sublst[[1]]
  
  # Merge remaining ones (if any)
  if (length(seurat_sublst) > 1) {
    for (j in 2:length(seurat_sublst)) {
      seurat.obj <- merge(x = seurat.obj, y = seurat_sublst[[j]])
    }
  }
  
  # Save into list
  seurat_merged_case.lst[[paste0("patient_", i)]] <- seurat.obj
}
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat_merged_case.lst, path_to_save_merged_obj_lst)
```

# Integrate the objects

## Harmony approach

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Data integration and re-downstream analysis
seurat_integrated_case.lst <- purrr::map(seurat_merged_case.lst, function(seurat.obj){
  DefaultAssay(seurat.obj) <- "RNA"
  seurat.obj <- JoinLayers(seurat.obj)
  seurat.obj <- NormalizeData(seurat.obj)
  seurat.obj <- FindVariableFeatures(seurat.obj)
  seurat.obj <- ScaleData(seurat.obj)
  seurat.obj <- RunPCA(seurat.obj)
  seurat.obj <- RunHarmony(seurat.obj, group.by.vars = "case", dims = 1:30) 
  seurat.obj <- RunUMAP(seurat.obj, dims = 1:30, reduction = "harmony")
  seurat.obj <- FindNeighbors(seurat.obj, reduction = "harmony", dims = 1:30) 
  seurat.obj <- FindClusters(
      seurat.obj,
      random.seed = 12345,
      verbose = FALSE,
      resolution = seq(0.1, 0.9, by = 0.1)
    )
  seurat.obj <- singleR_cell_annotation(seurat.obj) 
})
names(seurat_integrated_case.lst) <- names(seurat_merged_case.lst)
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat_integrated_case.lst, path_to_save_harmony_integrated_obj_lst)
```

## Canonical correlation analysis (CCA) approach

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Integrate the objects based on individual patients
seurat_integrated_case.lst <- vector("list", 3)  # create empty list of length 3
names(seurat_integrated_case.lst) <- c("patient_43", "patient_4", "patient_10")

# Loop to integrate the Seurat objects per patient
for (i in c(43, 4, 10)) {
  sra_related_case <- sample_metadata[sample_metadata$Case == i, "Run"]
  seurat_sublst <- seurat.lst[sra_related_case]
  
  seurat_sublst <- lapply(seurat_sublst, function(obj) {
    obj <- NormalizeData(obj)
    obj <- FindVariableFeatures(obj, selection.method = "vst")
    return(obj)
    })
  
  # Select feature, find anchor, and integrate the objects
  features <- SelectIntegrationFeatures(object.list = seurat_sublst)
  anchors <- FindIntegrationAnchors(object.list = seurat_sublst, anchor.features = features)
  integrated_seurat <- IntegrateData(anchorset = anchors)
  
  # Downstream analysis after integration
  DefaultAssay(integrated_seurat) <- "integrated"
  integrated_seurat <- ScaleData(integrated_seurat)
  integrated_seurat <- RunPCA(integrated_seurat)
  integrated_seurat <- RunUMAP(integrated_seurat, dims = 1:30)
  integrated_seurat <- FindNeighbors(integrated_seurat, dims = 1:30)
  integrated_seurat <- FindClusters(
      integrated_seurat,
      random.seed = 12345,
      verbose = FALSE,
      resolution = seq(0.1, 0.9, by = 0.1)
    )
  
  # Save into list
  seurat_integrated_case.lst[[paste0("patient_", i)]] <- integrated_seurat
}
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat_integrated_case.lst, path_to_save_cca_integrated_obj_lst)
```

After we preprocessed, merged, and integrated the objects, we will analyse individual patients. From the original article, they used CCA for data integration. Then we will use the results from CCA for further analysis.

```{r}
# Load the integrated list as the previous steps take a long time
seurat_integrated_case.lst <- readRDS(path_to_save_cca_integrated_obj_lst)
```

# Patient 43 analysis

```{r}
# Extract seurat object case 43 from the integrated list
case43_seurat <- seurat_integrated_case.lst$patient_43
```

## Clustering evaluation

```{r, height=20, widght=10}
# Carry out clustree to evaluate clustering stability at various resolution
clustree.plt <- clustree(case43_seurat, prefix = "integrated_snn_res.")
clustree.plt
```

We can see that clustering was stable at the resolution between 0.1 to 0.3. Then we will use resolution 0.1 for further analysis.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Clustering plot
cluster.plt <- DimPlot(case43_seurat, reduction = "umap", group.by = "integrated_snn_res.0.1") +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

## Noise removal

Due to many noises from the integration, we will remove them using DBSCAN algorithm.

```{r, message=FALSE, warning=FALSE}
# Import the package
library(dbscan)
```

```{r, message=FALSE, warning=FALSE}
# Remove noises using DBSCAN algorithm
umap <- Embeddings(case43_seurat, "umap")
db <- dbscan(umap, eps = 1.5, minPts = 800)  # tune eps and minPts

# db$cluster gives 0 for noise/outliers
keep.cells <- rownames(umap)[db$cluster != 0]
filtered_case43_seurat <- subset(case43_seurat, cells = keep.cells)
```

```{r, message=FALSE, warning=FALSE}
# Reclustering after noise removal
filtered_case43_seurat <- FindClusters(
      filtered_case43_seurat,
      random.seed = 12345,
      verbose = FALSE,
      resolution = seq(0.1, 0.9, by = 0.1)
    )
```

```{r, height=20, widght=10}
# Carry out clustree to evaluate clustering stability at various resolution
clustree.plt <- clustree(filtered_case43_seurat, prefix = "integrated_snn_res.")
clustree.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Clustering plot after noise removal
cluster.plt <- DimPlot(filtered_case43_seurat, reduction = "umap", group.by = "integrated_snn_res.0.1", label = TRUE) +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

## Cell biomarker expression

```{r, warning=FALSE, fig.height=7, fig.width=9}
# Cell biomarker feature plot
DefaultAssay(filtered_case43_seurat) <- "RNA"
FeaturePlot(filtered_case43_seurat, 
            features = c("HBM", "HBA2", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap",
            min.cutoff = 'q10',
            max.cutoff = 'q90')
```

From the feature plot, we can see that no non-tumour cells were detected.

## CLL and RT cell identification

```{r}
# Define features from the original study
features <- rev(c("EZH2", "CD83", "TRAF4", "IRF4", "IRF5", "CCNB1", "PRC1", "HMGN2", "LEF1"))
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Dot plot to identify CLL and RT cells
DotPlot(filtered_case43_seurat, features = features, group.by = "integrated_snn_res.0.1") + 
  coord_flip()
```

We can see from the dot plot that cluster 0, 1, and 4 expressed EZH2, CD83, TRAF4, IRF4, and IRF5 similar to cluster 3 and 4 for the original study (Fig. 5d). Therefore, these clusters were RT cells while cluster 2 and 3 were CLL cells, similar to cluster 1 and 2 in the original results.

# Patient 10 analysis

```{r}
# Extract seurat object case 10 from the integrated list
case10_seurat <- seurat_integrated_case.lst$patient_10
```

## Clustering evaluation

```{r, height=20, widght=10}
# Carry out clustree to evaluate clustering stability at various resolution
clustree.plt <- clustree(case10_seurat, prefix = "integrated_snn_res.")
clustree.plt
```

We can see that clustering was stable at the resolution between 0.2 to 0.3. Then we will use resolution 0.1 for further analysis.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Clustering plot
cluster.plt <- DimPlot(case10_seurat, reduction = "umap", group.by = "integrated_snn_res.0.2") +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

## Noise removal

Due to many noises from the integration, we will remove them using DBSCAN algorithm.

```{r, message=FALSE, warning=FALSE}
# Remove noises using DBSCAN algorithm
umap <- Embeddings(case10_seurat, "umap")
db <- dbscan(umap, eps = 1.5, minPts = 400)  # tune eps and minPts

# db$cluster gives 0 for noise/outliers
keep.cells <- rownames(umap)[db$cluster != 0]
filtered_case10_seurat <- subset(case10_seurat, cells = keep.cells)
```

```{r, message=FALSE, warning=FALSE}
# Reclustering after noise removal
filtered_case10_seurat <- FindClusters(
      filtered_case10_seurat,
      random.seed = 12345,
      verbose = FALSE,
      resolution = seq(0.1, 0.9, by = 0.1)
    )
```

```{r, height=20, widght=10}
# Carry out clustree to evaluate clustering stability at various resolution
clustree.plt <- clustree(filtered_case10_seurat, prefix = "integrated_snn_res.")
clustree.plt
```

For clustree, the resolution was quite stable at 0.3.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Clustering plot after noise removal
cluster.plt <- DimPlot(filtered_case10_seurat, reduction = "umap", group.by = "integrated_snn_res.0.3", label = TRUE) +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

## Cell biomarker expression

```{r, warning=FALSE, fig.height=7, fig.width=9}
# Cell biomarker feature plot
DefaultAssay(filtered_case10_seurat) <- "RNA"
FeaturePlot(filtered_case10_seurat, 
            features = c("HBM", "HBA2", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap",
            min.cutoff = 'q10',
            max.cutoff = 'q90')
```

## Filtering tumour cells

We will select tumour cells (cluster 4 and 7).

```{r, message=FALSE, warning=FALSE}
# Subset tumoure cells and reclustering after subset
DefaultAssay(filtered_case10_seurat) <- "integrated"
case10_tumour_seurat <- subset(filtered_case10_seurat, subset = integrated_snn_res.0.3 %in% c(4, 7))
case10_tumour_seurat <- FindClusters(
      case10_tumour_seurat,
      random.seed = 12345,
      verbose = FALSE,
      resolution = seq(0.1, 0.9, by = 0.1)
    )
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Clustering plot after noise removal
cluster.plt <- DimPlot(case10_tumour_seurat, reduction = "umap", group.by = "integrated_snn_res.0.1", label = TRUE) +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

## CLL and RT cell identification

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Dot plot to identify CLL and RT cells
DefaultAssay(case10_tumour_seurat) <- "RNA"
DotPlot(case10_tumour_seurat, features = features, group.by = "integrated_snn_res.0.1") + 
  coord_flip()
```

We observed from the dot plot that cluster 1 expressed EZH2, CD83, TRAF4, IRF4, and IRF5, consistent with the transcriptional features of RT cells, and comparable to cluster 3 in the original study (Fig. 5e). Cluster 0 displayed transcriptional profiles characteristic of CLL cells, resembling cluster 1 in the published results. Meanwhile, cluster 2 showed an intermediate expression pattern between RT and CLL, similar to cluster 2 in the reference study.

# Patient 4 analysis

```{r}
# Extract seurat object case 4 from the integrated list
case4_seurat <- seurat_integrated_case.lst$patient_4
```

## Clustering evaluation

```{r, height=20, widght=10}
# Carry out clustree to evaluate clustering stability at various resolution
clustree.plt <- clustree(case4_seurat, prefix = "integrated_snn_res.")
clustree.plt
```

We can see that clustering was stable at the resolution between 0.2 to 0.3. Then we will use resolution 0.1 for further analysis.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Clustering plot
cluster.plt <- DimPlot(case4_seurat, reduction = "umap", group.by = "integrated_snn_res.0.2") +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

## Noise removal

Due to many noises from the integration, we will remove them using DBSCAN algorithm.

```{r, message=FALSE, warning=FALSE}
# Remove noises using DBSCAN algorithm
umap <- Embeddings(case4_seurat, "umap")
db <- dbscan(umap, eps = 1.5, minPts = 400)  # tune eps and minPts

# db$cluster gives 0 for noise/outliers
keep.cells <- rownames(umap)[db$cluster != 0]
filtered_case4_seurat <- subset(case4_seurat, cells = keep.cells)
```

```{r, message=FALSE, warning=FALSE}
# Reclustering after noise removal
filtered_case4_seurat <- FindClusters(
      filtered_case4_seurat,
      random.seed = 12345,
      verbose = FALSE,
      resolution = seq(0.1, 0.9, by = 0.1)
    )
```

```{r, height=20, widght=10}
# Carry out clustree to evaluate clustering stability at various resolution
clustree.plt <- clustree(filtered_case4_seurat, prefix = "integrated_snn_res.")
clustree.plt
```

For clustree, the resolution was quite stable at 0.1.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Clustering plot after noise removal
cluster.plt <- DimPlot(filtered_case4_seurat, reduction = "umap", group.by = "integrated_snn_res.0.1", label = TRUE) +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

## Cell biomarker expression

```{r, warning=FALSE, fig.height=7, fig.width=9}
# Cell biomarker feature plot
DefaultAssay(filtered_case4_seurat) <- "RNA"
FeaturePlot(filtered_case4_seurat, 
            features = c("HBM", "HBA2", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap",
            min.cutoff = 'q10',
            max.cutoff = 'q90')
```

## Filtering tumour cells

We will select tumour cells (cluster 4).

```{r, message=FALSE, warning=FALSE}
# Subset tumoure cells and reclustering after subset
DefaultAssay(filtered_case4_seurat) <- "integrated"
case4_tumour_seurat <- subset(filtered_case4_seurat, subset = integrated_snn_res.0.3 == 4)
case4_tumour_seurat <- FindClusters(
      case4_tumour_seurat,
      random.seed = 12345,
      verbose = FALSE,
      resolution = seq(0.1, 0.9, by = 0.1)
    )
```

```{r, height=20, widght=10}
# Carry out clustree to evaluate clustering stability at various resolution
clustree.plt <- clustree(case4_tumour_seurat, prefix = "integrated_snn_res.")
clustree.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Clustering plot after noise removal
cluster.plt <- DimPlot(case4_tumour_seurat, reduction = "umap", group.by = "integrated_snn_res.0.3", label = TRUE) +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

## CLL and RT cell identification

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Dot plot to identify CLL and RT cells
DefaultAssay(case4_tumour_seurat) <- "RNA"
DotPlot(case4_tumour_seurat, features = features, group.by = "integrated_snn_res.0.3") + 
  coord_flip()
```

From the dot plot, clusters 0 and 1 expressed transcriptional profiles similar to RT cells while cluster 2 and 3 exhibited the intermediate profiles between RT and CLL cells. As the original study did not show the results for patient 4, we cannot conclude clearly about this result.

# Session Information

```{r}
sessionInfo()
```




