---
title: "Parry_2023b_Seurat_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(SeuratWrappers)
library(harmony)
library(SingleR)
library(celldex)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(stringr)
```

```{r}
# Check root path
here::here()
```

# Define path

```{r}
path_to_sample_metadata <- here::here("scRNAseq/phs002458.v2.p1/metadata/metadata.csv") 
path_to_counts <- here::here("scRNAseq/phs002458.v2.p1/expression_data/")
path_to_save_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/seurat_objects/1-seurat_integrated.rds")
```

# Import metadata

```{r}
sample_metadata <- read.csv(path_to_sample_metadata)
rownames(sample_metadata) <- sample_metadata$SRA_RUN
head(sample_metadata)
```

# Create Seurat objects

```{r, eval=FALSE}
sra.vec <- rownames(sample_metadata)
seurat.list <- purrr::map(sra.vec, function(vector) {
  path <- str_c(path_to_counts, vector, "/count_matrices/outs/filtered_feature_bc_matrix/", sep = "")
  count.data <- Read10X(path)
  seurat.obj <- CreateSeuratObject(counts=count.data, min.features=200, min.cells=10)
  seurat.obj <- RenameCells(seurat.obj, add.cell.id = vector) # Rename barcodes to prevent duplication
  seurat.obj[["sra_id"]] <- sample_metadata[vector, "SRA_RUN"] 
  seurat.obj[["subject_id"]] <- sample_metadata[vector, "subject_id"] 
  seurat.obj[["histological_type"]] <- sample_metadata[vector, "histological_type"]
  seurat.obj[["percent.mt"]] <- PercentageFeatureSet(seurat.obj, pattern="^MT-.")
  max_total_counts <- ifelse(sample_metadata[vector, "histological_type"] == "RS", 12000, 8000)
  seurat.obj <- subset(seurat.obj, subset = nCount_RNA > 500 & nCount_RNA < max_total_counts & percent.mt < 10)
  seurat.obj
})
names(seurat.list) <- sra.vec
```

# Merge the objects

```{r, eval=FALSE}
seurat <- seurat.list[[1]]
for (i in 2:length(seurat.list)) {
  seurat <- merge(x = seurat, y = seurat.list[[i]])
  seurat.list[[i]] <- NA
}
rm(seurat.list)
```

# Normalization, Dimensionality reduction and batch effect correction

```{r, message=FALSE, warning=FALSE, eval=FALSE}
seurat <- NormalizeData(seurat, normalization.method = "LogNormalize", scale.factor = 10000) 
seurat <- FindVariableFeatures(seurat) 
seurat <- ScaleData(seurat) 
seurat <- RunPCA(seurat) 
seurat <- RunHarmony(seurat, group.by.vars = "sra_id", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
```

# Cell annotation using SingleR

```{r, eval=FALSE}
seurat <- JoinLayers(seurat)
DefaultAssay(seurat) <- "RNA"
ref <- celldex::DatabaseImmuneCellExpressionData()
seurat_counts <- GetAssayData(seurat, layer = 'data')

pred <- SingleR(test = seurat_counts,
                ref = ref,
                labels = ref$label.main)

seurat$singleR.labels_main <- pred$labels[match(rownames(seurat@meta.data), rownames(pred))]
```

# Save the integrated object

```{r, eval=FALSE}
saveRDS(seurat, path_to_save_obj)
```

# Visualisation

```{r}
# Define path
path_to_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/seurat_objects/1-seurat_integrated.rds")

# Load the data
seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE}
# Plot based on subjects
DimPlot(seurat, reduction = "umap", group.by = "subject_id") +
  labs(y= "UMAP 2", x = "UMAP 1")
```

```{r, message=FALSE, warning=FALSE}
# Plot based on histological type
DimPlot(seurat, reduction = "umap", group.by = "histological_type") +
  labs(y= "UMAP 2", x = "UMAP 1")
```

```{r, message=FALSE, warning=FALSE}
# Plot based on cell annotation
DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")
```

# Pseudobulk analysis of tumour cells between RT and CLL conditions

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(DESeq2)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(msigdbr)
```

## Differential expression analysis

```{r}
# Create condition column and select only tumour cells
RT_subjects <- unique(sample_metadata[sample_metadata$histological_type == "RS", "subject_id"])
seurat$cond <- ifelse(as.character(seurat$subject_id) %in% RT_subjects, "RT", "CLL")
tumour_seurat <- subset(seurat, subset = singleR.labels_main == "B cells")
```

```{r, message=FALSE, warning=FALSE}
# Data aggregation
cts <- AggregateExpression(tumour_seurat,
                           group.by = c("subject_id"),
                           assays = "RNA",
                           slot = "counts",
                           resturn.seurat = FALSE
                           )
cts <- cts$RNA

# Create pseudocount matrix for tumour cells
pseudo_count.mat <- as.data.frame(cts)
```

```{r}
# Create metadata for DESeq2 object
colData <- data.frame(samples = colnames(pseudo_count.mat))
colData$condition <- ifelse(colData$samples %in% RT_subjects, "RT", "CLL")
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# Create a DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = pseudo_count.mat,
                              colData = colData,
                              design = ~ condition)
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# Run DESeq2
dds$condition <- relevel(dds$condition, ref = "CLL")
dds <- DESeq(dds)
```

```{r, message=FALSE, warning=FALSE}
# Show the results
res <- results(dds, contrast = c("condition", "RT", "CLL"))
res <- lfcShrink(dds, contrast = c("condition", "RT", "CLL"), res=res, type = 'normal')
dea.df <- as.data.frame(res)
head(dea.df)
```

```{r, message=FALSE, warning=FALSE}
# Volcano plot
FC <- 1
p <- 0.05
 
keyvals <- rep('grey75', nrow(dea.df))
names(keyvals) <- rep('NS', nrow(dea.df))
 
keyvals[which(abs(dea.df$log2FoldChange) > FC & dea.df$padj > p)] <- 'grey50'
names(keyvals)[which(abs(dea.df$log2FoldChange) > FC & dea.df$padj > p)] <- 'log2FoldChange'
 
keyvals[which(abs(dea.df$log2FoldChange) < FC & dea.df$padj < p)] <- 'grey25'
names(keyvals)[which(abs(dea.df$log2FoldChange)  < FC & dea.df$padj < p)] <- '-Log10Q'
 
keyvals[which(dea.df$log2FoldChange < -FC & dea.df$padj < p)] <- '#0000CD'
names(keyvals)[which(dea.df$log2FoldChange  < -FC & dea.df$padj < p)] <- 'Signif. down-regulated'
 
keyvals[which(dea.df$log2FoldChange > FC & dea.df$padj < p)] <- '#DC143C'
names(keyvals)[which(dea.df$log2FoldChange > FC & dea.df$padj < p)] <- 'Signif. up-regulated'
 
unique(keyvals)
unique(names(keyvals))
 
volcano.plt <- EnhancedVolcano(dea.df,
  lab = rownames(dea.df),
  x = 'log2FoldChange',
  y = 'padj',
  #selectLab = rownames(res2)[which(names(keyvals) %in% c('NS','log2FoldChange','-Log10Q','low','high'))],
  xlim = c(-5,5),
  ylim = c(-0.1,12),
  xlab = bquote(~Log[2]~ 'fold change'),
  ylab = bquote(~-Log[10] ~ italic(P)),
  title = 'Custom colour over-ride',
  pCutoff = 0.05,
  FCcutoff = 1,
  pointSize = 1,
  labSize = 3,
  #shape = c(6, 4, 2, 11, 15),
  colCustom = keyvals,
  colAlpha = 0.5,
  legendPosition = 'right',
  legendLabSize = 12,
  legendIconSize = 5.0,
  drawConnectors = FALSE,
  widthConnectors = 0.5,
  colConnectors = 'grey50',
  gridlines.major = TRUE,
  gridlines.minor = FALSE,
  border = 'partial',
  borderWidth = 1.5,
  borderColour = 'black')
 
volcano.plt
```

## Gene Set Enrichment Analysis (GSEA)

```{r}
# Hallmark set
hs_hm_sets <- read.csv("/mnt/hp-storage/users/pakorns/scRNAseq/GSE201704/GSM6069857/hs_hm_sets.csv")

# GO:BP set
hs_go_sets <- read.csv("/mnt/hp-storage/users/pakorns/scRNAseq/GSE201704/GSM6069857/hs_go_sets.csv")
```

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df) 
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

### Hallmark

```{r, message=FALSE, warning=FALSE}
# Run GSEA (Hallmarks)
set.seed(2020)
HM_gse <- GSEA(
  geneList = lfc_vector, # Ordered ranked gene list
  minGSSize = 25, # Minimum gene set size
  maxGSSize = 500, # Maximum gene set set
  pvalueCutoff = 0.05, # p-value cutoff
  eps = 0, # Boundary for calculating the p value
  seed = TRUE, # Set seed to make results reproducible
  pAdjustMethod = "BH", # Benjamini-Hochberg correction
  TERM2GENE = dplyr::select(
    hs_hm_sets,
    gs_name,
    gene_symbol
  )
)
```

```{r}
# Collect the GSEA result
HM_GSEA.result <- HM_gse@result
HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))
```

```{r, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

### GO:BP

```{r, message=FALSE, warning=FALSE}
# GSEA in GO-BP term
set.seed(2020)
GOBP_gsea <- GSEA(
  geneList = lfc_vector, # Ordered ranked gene list
  minGSSize = 25, # Minimum gene set size
  maxGSSize = 500, # Maximum gene set set
  pvalueCutoff = 0.05, # p-value cutoff
  eps = 0, # Boundary for calculating the p value
  seed = TRUE, # Set seed to make results reproducible
  pAdjustMethod = "BH", # Benjamini-Hochberg correction
  TERM2GENE = dplyr::select(
    hs_go_sets,
    gs_name,
    gene_symbol
  )
)
```

```{r}
# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))
```

```{r, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

# Session Information

```{r}
sessionInfo()
```




