---
title: "Parry_2023b_all_cell_annotation"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Matrix)
library(igraph)
library(pagoda2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(conos)
library(stringr)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
path_to_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/1-all_joint_clusters_conos.rds")
path_to_cell_metadata <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/metadata/1-all_joint_clusters_cell_metadata.rds")
path_to_save_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/2-all_annotate_conos.rds")
path_to_save_cell_metadata <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/metadata/2-all_annotate_cell_metadata.rds")
```

# Load data

```{r}
con <- readRDS(path_to_obj)
cell_metadata <- readRDS(path_to_cell_metadata)
```

# General visualisation

```{r, fig.height=5, fig.width=15}
# Plots for clusters, subject_id, and histological types
cluster.plt <- con$plotGraph(alpha=0.1, title="Leiden clusters", show.legend = TRUE) 

subject_id.vec <- setNames(cell_metadata$subject_id, row.names(cell_metadata))
subject_id.plt <- con$plotGraph(alpha=0.1, 
                                groups=subject_id.vec, 
                                title="Subject IDs",
                                show.legend = TRUE, 
                                mark.groups=FALSE) 

histo.vec <- setNames(cell_metadata$histologic_type, row.names(cell_metadata))
histo.plt <- con$plotGraph(alpha=0.1, 
                           groups=histo.vec, 
                           title="Histological types",
                           show.legend = TRUE) 

umap.plt <- cluster.plt + subject_id.plt + histo.plt
umap.plt
```

# Gene expression

## B cell biomerkers

```{r, fig.height=5, fig.width=12}
# B cell biomerkers
CD19.plt <- con$plotGraph(gene="CD19", title="CD19")
CD5.plt <- con$plotGraph(gene="CD5", title="CD5")
CD79A.plt <- con$plotGraph(gene="CD79A", title="CD79A")

CD19.plt + CD5.plt + CD79A.plt
```

## Erythroid biomarkers

```{r, fig.height=5, fig.width=12}
# Erythroid biomarkers
HBM.plt <- con$plotGraph(gene="HBM", title="HBM")
HBB.plt <- con$plotGraph(gene="HBB", title="HBB")
HBA1.plt <- con$plotGraph(gene="HBA1", title="HBA1")

HBM.plt + HBB.plt + HBA1.plt
```

## T cell biomarkers

```{r, fig.height=5, fig.width=15}
# All T cells
CD3D.plt <- con$plotGraph(gene="CD3D", title="CD3D")

# CD8 T cells
CD8A.plt <- con$plotGraph(gene="CD8A", title="CD8A")

# CD4 T cells
CD4.plt <- con$plotGraph(gene="CD4", title="CD4")
IL7R.plt <- con$plotGraph(gene="IL7R", title="IL7R")

(CD3D.plt + CD8A.plt) | (CD4.plt + IL7R.plt)
```

## NK cell biomarkers

```{r, fig.height=5, fig.width=12}
# NK cell biomarkers
NKG7.plt <- con$plotGraph(gene="NKG7", title="NKG7")
NCAM1.plt <- con$plotGraph(gene="NCAM1", title="NCAM1 (CD56)")
FCGR3A.plt <- con$plotGraph(gene="FCGR3A", title="FCGR3A (CD16)")

NKG7.plt + NCAM1.plt + FCGR3A.plt
```

## Myeloid and monocyte biomarkers

Biomarkers for myeloid cells, such as CD11c (ITGAX), CD13 (ANPEP), CD14, CD15 (FUT4), CD64(FCGR1A), can be used to identify myeloid cells in the plot.

```{r, fig.height=5, fig.width=12}
# Myeloid and monocyte biomarkers
CD14.plt <- con$plotGraph(gene="CD14", title="CD14")
ITGAX.plt <- con$plotGraph(gene="ITGAX", title="ITGAX (CD11c)")
FCGR1A.plt <- con$plotGraph(gene="FCGR1A", title="FCGR1A (CD64)")

CD14.plt + ITGAX.plt + FCGR1A.plt
```

```{r}
# Use for find gene name of interest
grep("CD15", colnames(con$samples[[1]]$counts), value = TRUE)
```

```{r}
# Create plots
gene.list <- c("CD19", "CD5", "CD79A", "HBM", "HBB", "HBA1", "CD3D", "CD8A", "CD4", "IL7R", "NKG7", "NCAM1", "FCGR3A", "CD14",
               "ITGAX", "FCGR1A")

plot.list <- purrr::map(gene.list, function(gene){
  plt <- con$plotGraph(gene=gene, title=gene)
  plt
})

names(plot.list) <- str_c(gene.list, ".plt", sep="")
```

```{r, fig.height=15, fig.width=12}
# Feature plots
feature.plt <- wrap_plots(plot.list, ncol = 4, nrow = 4)
feature.plt
```

# Cell type annotation

```{r}
# Add cell types to the cell metadata
cell_metadata <- cell_metadata %>%
  mutate(cell_type = case_when(
    leiden_cluster == "1" ~ "B cell",
    leiden_cluster == "2" ~ "CD8 T cell",
    leiden_cluster == "3" ~ "CD4 T cell",
    leiden_cluster == "4" ~ "Mixed B and T cell",
    leiden_cluster == "5" ~ "Erythoid cell",
    leiden_cluster == "6" ~ "Myeloid cell",
    leiden_cluster == "7" ~ "NK cell"
  ))
```

```{r, fig.height=10, fig.width=10}
# Visualise the annotation
cell_anno.vec <- setNames(cell_metadata$cell_type, row.names(cell_metadata))
cell.plt <- con$plotGraph(alpha=0.1, 
              groups=cell_anno.vec, 
              title="Cell annotation",
              show.legend = TRUE) 

umap.plt <- cluster.plt + subject_id.plt + histo.plt + cell.plt
umap.plt
```

# Save the results

```{r, eval=FALSE}
saveRDS(con, path_to_save_obj)
saveRDS(cell_metadata, path_to_save_cell_metadata)
```

# Session Information

```{r}
sessionInfo()
```



