---
title: "Parry_2023b_tumour_psedobulk_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Matrix)
library(igraph)
library(pagoda2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(conos)
library(stringr)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
path_to_sample_metadata <- here::here("scRNAseq/phs002458.v2.p1/metadata/metadata.csv")
path_to_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/pagoda2_objects/")
path_to_cell_metadata <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/metadata/pagoda2_cell_metadata/")
path_to_save_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/1-tumour_joint_clusters_conos.rds")
path_to_save_cell_metadata <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/metadata/1-tumour_joint_clusters_cell_metadata.rds")
```

# Load data

```{r}
# Load sample metadata
sample_metadata <- read.csv(path_to_sample_metadata)
head(sample_metadata)
```

```{r}
# Collect sra with CLL
sra_tumour <- sample_metadata[sample_metadata$is_tumor == "Yes", "SRA_RUN"]
sra_tumour
```

```{r}
# Collect RT subjects
RT_subject_ids <- sample_metadata[sample_metadata$histological_type == "RS", "subject_id"]
RT_subject_ids <- unique(RT_subject_ids)
RT_subject_ids
```

```{r, eval=FALSE}
# Load the objects and collect in a list
obj_files_to_load <- str_c(path_to_obj , "pagoda2_", sra_tumour, ".rds", sep = "")
panel.preprocessed <- purrr::map(obj_files_to_load, function(path) {
  pagoda2_obj <- readRDS(path)
  pagoda2_obj
})
names(panel.preprocessed) <- sra_tumour
```

```{r, eval=FALSE}
# Load the cell metadata from all pagoda2 objects
cell_meta_files_to_load <- str_c(path_to_cell_metadata, "cellmetadata_", sra_tumour, ".rds", sep = "")
cell_metada.list <- purrr::map(cell_meta_files_to_load, function(path) {
  pagoda2_cell_metadata <- readRDS(path)
  pagoda2_cell_metadata
})
names(cell_metada.list) <- sra_tumour
```

```{r, eval=FALSE}
# Combine all cell metadata together
cell_metadata_combined <- bind_rows(cell_metada.list) # rbind

# Rename in some columns
colnames(cell_metadata_combined)[colnames(cell_metadata_combined) == "sra_id_col"] <- "sra_id"
colnames(cell_metadata_combined)[colnames(cell_metadata_combined) == "subject_id_col"] <- "subject_id"
colnames(cell_metadata_combined)[colnames(cell_metadata_combined) == "histologic_type_col"] <- "histologic_type"
```

```{r, eval=FALSE}
# Create conditions column to the cell metadata
cell_metadata_combined$cond <- ifelse(cell_metadata_combined$subject_id %in% RT_subject_ids, "RT", "CLL")
```

# Create a conos object with pre-processing

```{r, eval=FALSE}
# Create an object
con <- Conos$new(panel.preprocessed, n.cores=1)
```

```{r, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
# Build a joint graph
con$buildGraph(k=30, k.self=5, space='PCA', ncomps=30, n.odgenes=2000, matching.method='mNN', metric='angular', score.component.variance=TRUE, verbose=TRUE)
```

```{r, eval=FALSE}
# Leiden clustering
set.seed(1234)
con$findCommunities(method=leiden.community, resolution=1)
```

```{r, eval=FALSE}
# Add clustering results to the cell metadata
cell_metadata_combined$leiden_cluster <- con$clusters$leiden$groups[rownames(cell_metadata_combined)]
```

```{r, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
# Graph embedding
con$embedGraph(method='UMAP')
```

# Save the results

```{r, eval=FALSE}
saveRDS(con, path_to_save_obj)
saveRDS(cell_metadata_combined, path_to_save_cell_metadata)
```

# Visualisation

```{r}
# Define path
path_to_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/1-tumour_joint_clusters_conos.rds")
path_to_cell_metadata <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/metadata/1-tumour_joint_clusters_cell_metadata.rds")

# Load the data
con <- readRDS(path_to_obj)
cell_metadata_combined <- readRDS(path_to_cell_metadata)
```

```{r, fig.height=10, fig.width=10}
# Plots for clusters, subject_id, histological types, and conditions
cluster.plt <- con$plotGraph(alpha=0.1, title="Leiden clusters", show.legend = TRUE) 

subject_id.vec <- setNames(cell_metadata_combined$subject_id, row.names(cell_metadata_combined))
subject_id.plt <- con$plotGraph(alpha=0.1, 
                                groups=subject_id.vec, 
                                title="Subject IDs",
                                show.legend = TRUE, 
                                mark.groups=FALSE) 

histo.vec <- setNames(cell_metadata_combined$histologic_type, row.names(cell_metadata_combined))
histo.plt <- con$plotGraph(alpha=0.1, 
                           groups=histo.vec, 
                           title="Histological types",
                           show.legend = TRUE) 

cond.vec <- setNames(cell_metadata_combined$cond, row.names(cell_metadata_combined))
cond.plt <- con$plotGraph(alpha=0.1, 
                          groups=cond.vec, 
                          title="Conditions",
                          show.legend = TRUE, 
                          mark.groups=FALSE) 

umap.plt <- cluster.plt + subject_id.plt + histo.plt + cond.plt
umap.plt
```

# Biomarker identification

```{r, fig.height=5, fig.width=12}
# B cell biomerkers
CD19.plt <- con$plotGraph(gene="CD19", title="CD19")
CD5.plt <- con$plotGraph(gene="CD5", title="CD5")
CD79A.plt <- con$plotGraph(gene="CD79A", title="CD79A")

CD19.plt + CD5.plt + CD79A.plt
```

```{r, fig.height=5, fig.width=12}
# Erythroid biomarkers
HBM.plt <- con$plotGraph(gene="HBM", title="HBM")
HBB.plt <- con$plotGraph(gene="HBB", title="HBB")
HBA1.plt <- con$plotGraph(gene="HBA1", title="HBA1")

HBM.plt + HBB.plt + HBA1.plt
```

```{r, fig.height=5, fig.width=15}
# All T cells
CD3D.plt <- con$plotGraph(gene="CD3D", title="CD3D")

# CD8 T cells
CD8A.plt <- con$plotGraph(gene="CD8A", title="CD8A")

# CD4 T cells
CD4.plt <- con$plotGraph(gene="CD4", title="CD4")
IL7R.plt <- con$plotGraph(gene="IL7R", title="IL7R")

(CD3D.plt + CD8A.plt) | (CD4.plt + IL7R.plt)
```

```{r, fig.height=5, fig.width=12}
# NK cell biomarkers
NKG7.plt <- con$plotGraph(gene="NKG7", title="NKG7")
NCAM1.plt <- con$plotGraph(gene="NCAM1", title="NCAM1 (CD56)")
FCGR3A.plt <- con$plotGraph(gene="FCGR3A", title="FCGR3A (CD16)")

NKG7.plt + NCAM1.plt + FCGR3A.plt
```

```{r, fig.height=5, fig.width=12}
# Myeloid and monocyte biomarkers
CD14.plt <- con$plotGraph(gene="CD14", title="CD14")
ITGAX.plt <- con$plotGraph(gene="ITGAX", title="ITGAX (CD11c)")
FCGR1A.plt <- con$plotGraph(gene="FCGR1A", title="FCGR1A (CD64)")

CD14.plt + ITGAX.plt + FCGR1A.plt
```

```{r}
# Create plots
gene.list <- c("CD19", "CD5", "CD79A", "HBM", "HBB", "HBA1", "CD3D", "CD8A", "CD4", "IL7R", "NKG7", "NCAM1", "FCGR3A", "CD14",
               "ITGAX", "FCGR1A")

plot.list <- purrr::map(gene.list, function(gene){
  plt <- con$plotGraph(gene=gene, title=gene)
  plt
})

names(plot.list) <- str_c(gene.list, ".plt", sep="")
```

```{r, fig.height=15, fig.width=12}
# Feature plots
feature.plt <- wrap_plots(plot.list, ncol = 4, nrow = 4)
feature.plt
```

From the feature plots, we can see that cluster 1, 2, 3, 4, and 8 are tumour cells. We then extract them for further analysis.

```{r}
# Extract cell barcodes which are tumour cells 
tumour_barcodes <- rownames(cell_metadata_combined)[cell_metadata_combined$leiden_cluster %in% c(1, 2, 3, 4, 8)]
head(tumour_barcodes)
```

# Pseudobulk analysis between RT and CLL conditions

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(DESeq2)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(msigdbr)
```

## Aggregate the raw matrices and gene expression analysis using DESeq2

```{r}
# Define function for quality control for raw counts and keep only tumour cells
my_QC <- function(count) {
  total_counts <- Matrix::colSums(count) # total UMIs
  cell_counts <- Matrix::rowSums(count) # total cells per gene
  
  # Calculate mitochondrial percentage
  genes <- rownames(count) # get gene names
  mito_genes <- grep("^MT-", genes, value = TRUE) # identify mitochondrial genes
  mito_counts <- Matrix::colSums(count[mito_genes, ]) # mitochondrial UMIs per cell
  percent_mito <- mito_counts / total_counts * 100
  
  # Keep only cells with <10% mitochondrial expression and 500 < UMI_count < 8000
  keep_cells <- which(percent_mito < 10 & (total_counts > 500 & total_counts < 8000)) 
  
  # Keep only genes with cells > 10
  keep_genes <- which(cell_counts > 10) 
  
  # Filter counts matrix
  filtered_count <- count[keep_genes, keep_cells]
  
  # Keep tumour cells
  keep_barcodes <- colnames(filtered_count)[colnames(filtered_count) %in% tumour_barcodes]
  filtered_count <- filtered_count[, keep_barcodes, drop=FALSE] 
  
  # Return the filtered matrix
  return(filtered_count)
}
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# Load raw count matrix data
path_to_rawcount <- here::here("scRNAseq/phs002458.v2.p1/expression_data/")
rawcounts_files_to_load <- str_c(path_to_rawcount , sra_tumour, "/count_matrices/outs/filtered_feature_bc_matrix/", sep = "")
cell_count.list <- purrr::map2(sra_tumour, rawcounts_files_to_load, function(id, path) {
  raw_counts <- read10xMatrix(path)
  colnames(raw_counts) <- paste0("sample_", id, "-", colnames(raw_counts)) # Change barcode names based on the samples
  filtered_count <- my_QC(raw_counts)
  filtered_count
})
names(cell_count.list) <- sra_tumour
```

```{r}
# Aggregate count in each barcode to same genes
aggregate.list <- lapply(cell_count.list, rowSums)
```

```{r}
# Create raw count matrix merging all samples
all_genes <- Reduce(union, lapply(aggregate.list, names)) # Get all gene names

aligned.list <- lapply(aggregate.list, function(x) {
  # Create a vector for all genes, fill with 0
  vec <- numeric(length(all_genes))
  names(vec) <- all_genes
  
  # Fill with existing counts
  vec[names(x)] <- x
  return(vec)
})

all_counts <- do.call(cbind, aligned.list)
colnames(all_counts) <- names(aggregate.list)
rownames(all_counts) <- all_genes
```

```{r}
# Generate metadata for DESeq2
RT_sra <- sample_metadata[sample_metadata$subject_id %in% RT_subject_ids, "SRA_RUN"]
colData <- data.frame(samples = colnames(all_counts))
colData$condition <- ifelse(colData$samples %in% RT_sra, "RT", "CLL")
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# Create a DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = all_counts,
                              colData = colData,
                              design = ~ condition)
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# Run DESeq2
dds$condition <- relevel(dds$condition, ref = "CLL")
dds <- DESeq(dds)
```

```{r, message=FALSE, warning=FALSE}
# Show the results
res <- results(dds, contrast = c("condition", "RT", "CLL"))
res <- lfcShrink(dds, contrast = c("condition", "RT", "CLL"), res=res, type = 'normal')
dea.df <- as.data.frame(res)
head(dea.df)
```

```{r, message=FALSE, warning=FALSE}
# Volcano plot
FC <- 1
p <- 0.05
 
keyvals <- rep('grey75', nrow(dea.df))
names(keyvals) <- rep('NS', nrow(dea.df))
 
keyvals[which(abs(dea.df$log2FoldChange) > FC & dea.df$padj > p)] <- 'grey50'
names(keyvals)[which(abs(dea.df$log2FoldChange) > FC & dea.df$padj > p)] <- 'log2FoldChange'
 
keyvals[which(abs(dea.df$log2FoldChange) < FC & dea.df$padj < p)] <- 'grey25'
names(keyvals)[which(abs(dea.df$log2FoldChange)  < FC & dea.df$padj < p)] <- '-Log10Q'
 
keyvals[which(dea.df$log2FoldChange < -FC & dea.df$padj < p)] <- '#0000CD'
names(keyvals)[which(dea.df$log2FoldChange  < -FC & dea.df$padj < p)] <- 'Signif. down-regulated'
 
keyvals[which(dea.df$log2FoldChange > FC & dea.df$padj < p)] <- '#DC143C'
names(keyvals)[which(dea.df$log2FoldChange > FC & dea.df$padj < p)] <- 'Signif. up-regulated'
 
unique(keyvals)
unique(names(keyvals))
 
volcano.plt <- EnhancedVolcano(dea.df,
  lab = rownames(dea.df),
  x = 'log2FoldChange',
  y = 'padj',
  #selectLab = rownames(res2)[which(names(keyvals) %in% c('NS','log2FoldChange','-Log10Q','low','high'))],
  xlim = c(-5,5),
  ylim = c(-0.1,12),
  xlab = bquote(~Log[2]~ 'fold change'),
  ylab = bquote(~-Log[10] ~ italic(P)),
  title = 'Custom colour over-ride',
  pCutoff = 0.05,
  FCcutoff = 1,
  pointSize = 1,
  labSize = 3,
  #shape = c(6, 4, 2, 11, 15),
  colCustom = keyvals,
  colAlpha = 0.5,
  legendPosition = 'right',
  legendLabSize = 12,
  legendIconSize = 5.0,
  drawConnectors = FALSE,
  widthConnectors = 0.5,
  colConnectors = 'grey50',
  gridlines.major = TRUE,
  gridlines.minor = FALSE,
  border = 'partial',
  borderWidth = 1.5,
  borderColour = 'black')
 
volcano.plt
```

## Gene Set Enrichment Analysis (GSEA)

```{r}
# Hallmark set
hs_hm_sets <- read.csv("/mnt/hp-storage/users/pakorns/scRNAseq/GSE201704/GSM6069857/hs_hm_sets.csv")

# GO:BP set
hs_go_sets <- read.csv("/mnt/hp-storage/users/pakorns/scRNAseq/GSE201704/GSM6069857/hs_go_sets.csv")
```

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df) 
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

### Hallmark

```{r, message=FALSE, warning=FALSE}
# Run GSEA (Hallmarks)
set.seed(2020)
HM_gse <- GSEA(
  geneList = lfc_vector, # Ordered ranked gene list
  minGSSize = 25, # Minimum gene set size
  maxGSSize = 500, # Maximum gene set set
  pvalueCutoff = 0.05, # p-value cutoff
  eps = 0, # Boundary for calculating the p value
  seed = TRUE, # Set seed to make results reproducible
  pAdjustMethod = "BH", # Benjamini-Hochberg correction
  TERM2GENE = dplyr::select(
    hs_hm_sets,
    gs_name,
    gene_symbol
  )
)
```

```{r}
# Collect the GSEA result
HM_GSEA.result <- HM_gse@result
HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))
```

```{r, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

### GO:BP

```{r, message=FALSE, warning=FALSE}
# GSEA in GO-BP term
set.seed(2020)
GOBP_gsea <- GSEA(
  geneList = lfc_vector, # Ordered ranked gene list
  minGSSize = 25, # Minimum gene set size
  maxGSSize = 500, # Maximum gene set set
  pvalueCutoff = 0.05, # p-value cutoff
  eps = 0, # Boundary for calculating the p value
  seed = TRUE, # Set seed to make results reproducible
  pAdjustMethod = "BH", # Benjamini-Hochberg correction
  TERM2GENE = dplyr::select(
    hs_go_sets,
    gs_name,
    gene_symbol
  )
)
```

```{r}
# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))
```

```{r, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

# Session Information

```{r}
sessionInfo()
```





