---
title: "Parry_2023b_microenvironment_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Matrix)
library(igraph)
library(pagoda2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(conos)
library(stringr)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
path_to_sample_metadata <- here::here("scRNAseq/phs002458.v2.p1/metadata/metadata.csv")
path_to_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/pagoda2_objects/")
path_to_cell_metadata <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/metadata/pagoda2_cell_metadata/")
path_to_save_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/1-microenviron_joint_clusters_conos.rds")
path_to_save_cell_metadata <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/metadata/1-microenviron_joint_clusters_cell_metadata.rds")
```

# Load data

```{r}
# Load sample metadata
sample_metadata <- read.csv(path_to_sample_metadata)
head(sample_metadata)
```

```{r}
# Collect sra with microenvironment
sra_microenviron <- sample_metadata[sample_metadata$is_tumor == "No" & sample_metadata$histological_type != "Erythroid", "SRA_RUN"]
sra_microenviron
```

```{r, eval=FALSE}
# Load the objects and collect in a list
obj_files_to_load <- str_c(path_to_obj , "pagoda2_", sra_microenviron, ".rds", sep = "")
panel.preprocessed <- purrr::map(obj_files_to_load, function(path) {
  pagoda2_obj <- readRDS(path)
  pagoda2_obj
})
names(panel.preprocessed) <- sra_microenviron
```

```{r, eval=FALSE}
# Load the cell metadata from all pagoda2 objects
cell_meta_files_to_load <- str_c(path_to_cell_metadata, "cellmetadata_", sra_microenviron, ".rds", sep = "")
cell_metada.list <- purrr::map(cell_meta_files_to_load, function(path) {
  pagoda2_cell_metadata <- readRDS(path)
  pagoda2_cell_metadata
})
names(cell_metada.list) <- sra_microenviron
```

```{r, eval=FALSE}
# Combine all cell metadata together
cell_metadata_combined <- bind_rows(cell_metada.list) # rbind

# Rename in some columns
colnames(cell_metadata_combined)[colnames(cell_metadata_combined) == "sra_id_col"] <- "sra_id"
colnames(cell_metadata_combined)[colnames(cell_metadata_combined) == "subject_id_col"] <- "subject_id"
colnames(cell_metadata_combined)[colnames(cell_metadata_combined) == "histologic_type_col"] <- "histologic_type"
```

```{r, eval=FALSE}
# Create conditions column to the cell metadata
RT_subject_ids <- unique(sample_metadata[sample_metadata$histological_type == "RS", "subject_id"])
cell_metadata_combined$cond <- ifelse(cell_metadata_combined$subject_id %in% RT_subject_ids, "RT", "CLL")
```

# Create a conos object with pre-processing

```{r, eval=FALSE}
# Create an object
con <- Conos$new(panel.preprocessed, n.cores=1)
```

```{r, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
# Build a joint graph
con$buildGraph(k=30, k.self=5, space='PCA', ncomps=30, n.odgenes=2000, matching.method='mNN', metric='angular', score.component.variance=TRUE, verbose=TRUE)
```

```{r, eval=FALSE}
# Leiden clustering
set.seed(1234)
con$findCommunities(method=leiden.community, resolution=2.5)
```

```{r, eval=FALSE}
# Add clustering results to the cell metadata
cell_metadata_combined$leiden_cluster <- con$clusters$leiden$groups[rownames(cell_metadata_combined)]
```

```{r, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
# Graph embedding
con$embedGraph(method='UMAP')
```

# Save the results

```{r, eval=FALSE}
saveRDS(con, path_to_save_obj)
saveRDS(cell_metadata_combined, path_to_save_cell_metadata)
```

# Visualisation

```{r}
# Define path
path_to_obj <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/R_objects/1-microenviron_joint_clusters_conos.rds")
path_to_cell_metadata <- here::here("scRNAseq/phs002458.v2.p1/reanalysis/metadata/1-microenviron_joint_clusters_cell_metadata.rds")

# Load the data
con <- readRDS(path_to_obj)
cell_metadata_combined <- readRDS(path_to_cell_metadata)
```

```{r, fig.height=5, fig.width=15}
# Plots for clusters, subject_id, and histological types
cluster.plt <- con$plotGraph(alpha=0.1, title="Leiden clusters", show.legend = TRUE) 

subject_id.vec <- setNames(cell_metadata_combined$subject_id, row.names(cell_metadata_combined))
subject_id.plt <- con$plotGraph(alpha=0.1, 
                                groups=subject_id.vec, 
                                title="Subject IDs",
                                show.legend = TRUE, 
                                mark.groups=FALSE) 

cond.vec <- setNames(cell_metadata_combined$cond, row.names(cell_metadata_combined))
cond.plt <- con$plotGraph(alpha=0.1, 
                          groups=cond.vec, 
                          title="Conditions",
                          show.legend = TRUE, 
                          mark.groups=FALSE)  

umap.plt <- cluster.plt + subject_id.plt + cond.plt
umap.plt
```

# Biomarker identification

## B cell biomerkers

```{r, fig.height=5, fig.width=12}
# B cell biomerkers
CD19.plt <- con$plotGraph(gene="CD19", title="CD19")
CD5.plt <- con$plotGraph(gene="CD5", title="CD5")
CD79A.plt <- con$plotGraph(gene="CD79A", title="CD79A")

CD19.plt + CD5.plt + CD79A.plt
```

## Erythroid cell biomarkers

```{r, fig.height=5, fig.width=12}
# Erythroid biomarkers
HBM.plt <- con$plotGraph(gene="HBM", title="HBM")
HBB.plt <- con$plotGraph(gene="HBB", title="HBB")
HBA1.plt <- con$plotGraph(gene="HBA1", title="HBA1")

HBM.plt + HBB.plt + HBA1.plt
```

## T cell biomarkers

```{r, fig.height=5, fig.width=15}
# All T cells
CD3D.plt <- con$plotGraph(gene="CD3D", title="CD3D")

# CD8 T cells
CD8A.plt <- con$plotGraph(gene="CD8A", title="CD8A")

# CD4 T cells
CD4.plt <- con$plotGraph(gene="CD4", title="CD4")
IL7R.plt <- con$plotGraph(gene="IL7R", title="IL7R")

(CD3D.plt + CD8A.plt) | (CD4.plt + IL7R.plt)
```

## NK cell biomarkers

```{r, fig.height=5, fig.width=12}
# NK cell biomarkers
NKG7.plt <- con$plotGraph(gene="NKG7", title="NKG7")
NCAM1.plt <- con$plotGraph(gene="NCAM1", title="NCAM1 (CD56)")
FCGR3A.plt <- con$plotGraph(gene="FCGR3A", title="FCGR3A (CD16)")

NKG7.plt + NCAM1.plt + FCGR3A.plt
```

## Myeloid and monocyte biomarkers

```{r, fig.height=5, fig.width=12}
# Myeloid and monocyte biomarkers
CD14.plt <- con$plotGraph(gene="CD14", title="CD14")
ITGAX.plt <- con$plotGraph(gene="ITGAX", title="ITGAX (CD11c)")
FCGR1A.plt <- con$plotGraph(gene="FCGR1A", title="FCGR1A (CD64)")

CD14.plt + ITGAX.plt + FCGR1A.plt
```

```{r}
# Create plots
gene.list1 <- c("CD19", "CD5", "CD79A", "HBM", "HBB", "HBA1", "CD3D", "CD8A", "CD4", "IL7R", "NKG7", "NCAM1", "FCGR3A", "CD14",
               "ITGAX", "FCGR1A")

plot.list1 <- purrr::map(gene.list1, function(gene){
  plt <- con$plotGraph(gene=gene, title=gene)
  plt
})

names(plot.list1) <- str_c(gene.list1, ".plt", sep="")
```

```{r, fig.height=15, fig.width=12}
# Feature plots
feature.plt <- wrap_plots(plot.list1, ncol = 4, nrow = 4)
feature.plt
```

## Other biomarkers for lymphocyte subclusters

```{r}
# Create plots
gene.list2 <- c("CD3D", "CD8A", "CD4", "FCGR3A", "PRF1", "GZMB", "XCL1", "GZMK", "IL7R", "SELL", "CX3CR1", 
               "CCR7", "FOXP3", "TCF7", "TOX")

plot.list2 <- purrr::map(gene.list2, function(gene){
  plt <- con$plotGraph(gene=gene, title=gene)
  plt
})

names(plot.list2) <- str_c(gene.list2, ".plt", sep="")
```

```{r, fig.height=9, fig.width=12}
# Feature plots
feature2.plt <- wrap_plots(plot.list2, ncol = 5, nrow = 3)
feature2.plt
```

```{r, fig.height=4, fig.width=4}
# Feature plot for ZNF683 expression
gene_interest <- "ZNF683"
con$plotGraph(gene=gene_interest, title=gene_interest)
```

# Cell type annotation

```{r}
# Add cell types to the cell metadata
cell_metadata_combined <- cell_metadata_combined %>%
  mutate(cell_type = case_when(
    leiden_cluster == "1" ~ "T E/EM, CD8+",
    leiden_cluster == "2" ~ "T naive, CD4+",
    leiden_cluster == "3" ~ "Erythroid progenitor",
    leiden_cluster == "4" ~ "T cytotoxic, CD4+",
    leiden_cluster == "5" ~ "T, CD4+",
    leiden_cluster == "6" ~ "T E/EM, CD8+",
    leiden_cluster == "7" ~ "T reg, CD4+",
    leiden_cluster == "8" ~ "NK, CD16+",
    leiden_cluster == "9" ~ "T naive, CD4+",
    leiden_cluster == "10" ~ "T exhausted, CD8+",
    leiden_cluster == "11" ~ "T mem., GZMK+CD8+",
    leiden_cluster == "12" ~ "Malignant B",
    leiden_cluster == "13" ~ "NK, CD16+",
    leiden_cluster == "14" ~ "Erythroid progenitor",
    leiden_cluster == "15" ~ "NK, CD16-",
    leiden_cluster == "16" ~ "monocytes, CD14+",
    leiden_cluster == "17" ~ "monocytes, CD16+",
    leiden_cluster == "18" ~ "Malignant B",
    leiden_cluster == "19" ~ "Suspected doublets"
  ))
```

```{r, fig.height=4, fig.width=5}
# Visualise the annotation
cell_anno.vec <- setNames(cell_metadata_combined$cell_type, row.names(cell_metadata_combined))
cell.plt <- con$plotGraph(alpha=0.1, 
              groups=cell_anno.vec, 
              mark.groups=FALSE,
              title="Cell annotation",
              show.legend = TRUE) +
  guides(
    colour = guide_legend(override.aes = list(size=5, alpha = 1))  # increase dot size in legend
  ) 
cell.plt
```

```{r, fig.height=10, fig.width=10}
# Visualise all
umap.plt2 <- umap.plt + cell.plt
umap.plt2
```

# Pseudobulk analysis of lymphocyte subtypes

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(DESeq2)
library(tibble)
library(pheatmap)
library(RColorBrewer)
```

## Create a count matrix for lymphocyte subclusters

```{r}
# Create a dictionary-like data structure recording cluster numbers and lymphocyte subclusters
lymphocyte_subclusters.dict <- c("1" = "T E/EM, CD8+",
                                 "2" = "T naive, CD4+",
                                 "4" = "T cytotoxic, CD4+",
                                 "5" = "T, CD4+",
                                 "6" = "T E/EM, CD8+",
                                 "7" = "T reg, CD4+",
                                 "8" = "NK, CD16+",
                                 "9" = "T naive, CD4+",
                                 "10" = "T exhausted, CD8+",
                                 "11" = "T mem., GZMK+CD8+",
                                 "13" = "NK, CD16+",
                                 "15" = "NK, CD16-"
                                 )
```

```{r}
# Collect cell barcodes in each subclusters and collect in a list
lymphocyte_barcodes.list <- purrr::map(lymphocyte_subclusters.dict, function(subtype){
  barcodes <- rownames(cell_metadata_combined)[cell_metadata_combined$cell_type == subtype]
  barcodes
})
names(lymphocyte_barcodes.list) <- lymphocyte_subclusters.dict

# Barcodes for all lymphocytes
lymphocyte_barcodes <- unique(unlist(lymphocyte_barcodes.list))
```

```{r}
# Define function for quality control for raw counts and keep only tumour cells
my_QC <- function(count) {
  total_counts <- Matrix::colSums(count) # total UMIs
  cell_counts <- Matrix::rowSums(count) # total cells per gene
  
  # Calculate mitochondrial percentage
  genes <- rownames(count) # get gene names
  mito_genes <- grep("^MT-", genes, value = TRUE) # identify mitochondrial genes
  mito_counts <- Matrix::colSums(count[mito_genes, ]) # mitochondrial UMIs per cell
  percent_mito <- mito_counts / total_counts * 100
  
  # Keep only cells with <10% mitochondrial expression and 500 < UMI_count < 8000
  keep_cells <- which(percent_mito < 10 & (total_counts > 500 & total_counts < 8000)) 
  
  # Keep only genes with cells > 10
  keep_genes <- which(cell_counts > 10) 
  
  # Filter counts matrix
  filtered_count <- count[keep_genes, keep_cells]
  
  # Keep lymphocytes
  keep_barcodes <- colnames(filtered_count)[colnames(filtered_count) %in% lymphocyte_barcodes]
  filtered_count <- filtered_count[, keep_barcodes, drop=FALSE] 
  
  # Return the filtered matrix
  return(filtered_count)
}
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# Load raw count matrix data
path_to_rawcount <- here::here("scRNAseq/phs002458.v2.p1/expression_data/")
rawcounts_files_to_load <- str_c(path_to_rawcount , 
                                 sra_microenviron, 
                                 "/count_matrices/outs/filtered_feature_bc_matrix/", 
                                 sep = ""
                                 )
cell_count.list <- purrr::map2(sra_microenviron, rawcounts_files_to_load, function(id, path) {
  raw_counts <- read10xMatrix(path)
  colnames(raw_counts) <- paste0("sample_", id, "-", colnames(raw_counts)) # Change barcode names based on the samples
  filtered_count <- my_QC(raw_counts)
  filtered_count
})
names(cell_count.list) <- sra_microenviron
```

```{r}
# Create raw count matrix merging all samples
all_genes <- Reduce(union, lapply(cell_count.list, rownames)) # Get all gene names

aligned.list <- lapply(cell_count.list, function(x) {
  # Create sparse matrix for all genes
  mat <- Matrix(0, nrow = length(all_genes), ncol = ncol(x), sparse = TRUE)
  rownames(mat) <- all_genes
  colnames(mat) <- colnames(x)
  
  # Fill existing counts (sparse assignment)
  mat[rownames(x),] <- x
  return(mat)
})

lymphocyte_counts <- do.call(cbind, aligned.list) # Unlist aligned.list to merge only one count matrix
```

```{r}
# Create list collecting counts in each lymphocyte subcluster
lymphocyte_subcluster_count.list <- purrr::map(lymphocyte_barcodes.list, function(barcode){
  subcluster_mat <- lymphocyte_counts[, colnames(lymphocyte_counts) %in% barcode, drop=FALSE]
  subcluster_mat
})
names(lymphocyte_subcluster_count.list) <- names(lymphocyte_barcodes.list)

# Aggregate count in each barcode to same genes
aggregate.list <- lapply(lymphocyte_subcluster_count.list, rowSums)

# Final count matrix
all_counts <- do.call(cbind, aggregate.list)
```

## Gene expression analysis

```{r}
# Generate metadata for DESeq2
colData <- data.frame(samples = colnames(all_counts))
colData$condition <- ifelse(grepl("NK", colData$samples), "NK",
                            ifelse(grepl("CD8", colData$samples), "CD8", "CD4"))
```

```{r, message=FALSE, warning=FALSE}
# Generate a DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = all_counts,
                              colData = colData,
                              design = ~ condition)
```

```{r}
# Data transformation
vsd <- vst(dds, blind=FALSE)       # Variance-stabilizing transformation
#rld <- rlog(dds_filtered, blind=FALSE)      # Regularized log transformation

# Extract transformed expression matrix
expr_mat <- assay(vsd)

# Calculate z-scores
zscore_mat <- t(scale(t(expr_mat)))
```

## Heatmap visualisation

```{r}
# List of genes
genes_to_plot <- c(
  "CD3D", "CD8A", "CD4", "FCGR3A", "NCR1", "TCF7", "SELL", "LEF1", "CCR7", "FOXP3", "IL2RA", "CTLA4", "LAG3", "TIGIT",
  "PDCD1", "HAVCR2", "CD160", "ENTPD1", "CD27", "CD28", "TNFRSF4", "ICOS", "TNFRSF9", "CD40LG", "CD226", "GZMA", "GZMB",
  "PRF1", "GZMK", "NKG7", "TBX21", "EOMES", "TOX", "ZNF683", "IL7R", "CX3CR1", "CD69", "CXCR4", "CXCR3", "KLRG1", "IFNG",
  "TGFB1", "XCL1", "XCL2"
)
```

```{r}
# Change column names of the matrix
new_col_names <- c("E/EM_1", "Naive_1", "Cytotoxicity", "CD4", "E/EM_2", "T reg", "CD16+_1", "Naive_2", "Exhausted", 
                   "GZMK+ mem.", "CD16+_2", "CD16-")

# Copy zscore_mat
zscore_mat_selected_genes <- zscore_mat[rownames(zscore_mat) %in% genes_to_plot, ]
colnames(zscore_mat_selected_genes) <- new_col_names
```

```{r}
# Extract main and sub- clusters info from column names
primary_clusters <-  ifelse(grepl("NK", colnames(zscore_mat)), "NK",
                            ifelse(grepl("CD8", colnames(zscore_mat)), "CD8", "CD4"))
secondary_clusters <- colnames(zscore_mat_selected_genes)

# Create annotation dataframe
annotation_col <- data.frame(
  MainType = factor(primary_clusters, levels = c("NK", "CD8", "CD4")),
  SubType = factor(secondary_clusters, levels = c("CD16+_1", "CD16+_2", "CD16-", "Exhausted", "E/EM_1", "E/EM_2",
                                                  "GZMK+ mem.", "Cytotoxicity", "T reg", "Naive_1", "Naive_2", "CD4"))
)
annotation_col <- annotation_col[, c("SubType", "MainType")] # Put MainType above SubType
rownames(annotation_col) <- colnames(zscore_mat_selected_genes)

# Create ordered genes
genes_present <- genes_to_plot[genes_to_plot %in% rownames(zscore_mat_selected_genes)]
order_rows <- match(genes_present, rownames(zscore_mat_selected_genes))

# Order by primary and then secondary annotation and gene list
order_cols <- order(annotation_col$SubType, annotation_col$MainType)
zscore_mat_ordered <- zscore_mat_selected_genes[order_rows, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]
```

```{r, fig.height=8, fig.width=4}
# Plot
pheatmap(
  zscore_mat_ordered,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = TRUE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  gaps_col = which(diff(as.numeric(factor(annotation_col_ordered$MainType))) != 0)
)
```

## Principal component plot of the samples

```{r, message=FALSE, warning=FALSE}
# PCA plot
pcaData <- plotPCA(vsd, intgroup=c("condition", "samples"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=samples, shape=condition)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  xlim(c(-30, 30)) +
  ylim(c(-30, 30))
```

# Session Information

```{r}
sessionInfo()
```





