---
title: "Rigo_2024_reanalysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(stringr)
library(DoubletFinder)
library(SingleR)
library(celldex)
library(DT)
```

```{r}
# Check root path
here::here()
```

# Define paths and load the functions

```{r}
path_to_cnt <- here::here("scRNAseq/Rigo_2024_scRNAseq/count_matrices")
path_to_save_obj_lst <- here::here("scRNAseq/Rigo_2024_scRNAseq/reanalysis/R_objects/0-individual_seurat_list.rds")
path_to_save_merged_obj <- here::here("scRNAseq/Rigo_2024_scRNAseq/reanalysis/R_objects/1-merged_seurat.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/Rigo_2024_scRNAseq/reanalysis/R_objects/2-integrated_seurat.rds")
path_to_save_tumour_obj <- here::here("scRNAseq/Rigo_2024_scRNAseq/reanalysis/R_objects/3-tumour_seurat.rds")
path_to_save_dea <- here::here("scRNAseq/Rigo_2024_scRNAseq/reanalysis/R_objects/4-rt_cll_dea.rds")
```

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

# Define function for doublet detecttion

We will use multiple rate table from [10X](https://kb.10xgenomics.com/hc/en-us/articles/360001378811-What-is-the-maximum-number-of-cells-that-can-be-profiled) and use the rate based on our number of cell barcodes to run DoubletFinder.

```{r}
# Create 10X multiple rate table 
multiple_rate.tbl <- data.frame(
  'Multiplet_rate'= c(0.004, 0.008, 0.016, 0.024, 0.032, 0.04, 0.048, 0.056, 0.064, 0.072, 0.08),
  'Loaded_cells' = c(825, 1650, 3300, 4950, 6600, 8250, 9900, 11550, 13200, 14850, 16500),
  'Recovered_cells' = c(500, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 8000, 9000, 10000)
  )

DT::datatable(multiple_rate.tbl, options = list(scrollX = TRUE))
```

```{r}
# Define a function to filter out doublets based DoubletFinder package
my_doublet_filter <- function(seurat.obj){
  multiplet_rate <- multiple_rate.tbl %>% 
    dplyr::filter(Recovered_cells < nrow(seurat.obj@meta.data)) %>% 
    dplyr::slice(which.max(Recovered_cells)) %>% # select the min threshold depending on your number of samples
    dplyr::select(Multiplet_rate) %>% as.numeric(as.character()) 
  
  # pK Identification (no ground-truth)
  sweep_list <- paramSweep(seurat.obj, PCs = 1:30, sct = FALSE) 
  sweep_stats <- summarizeSweep(sweep_list)
  bcmvn <- find.pK(sweep_stats)
  optimal.pk <- bcmvn %>% 
    dplyr::filter(BCmetric == max(BCmetric)) %>%
    dplyr::select(pK)
  optimal.pk <- as.numeric(as.character(optimal.pk[[1]]))
  
  # Homotypic Doublet Proportion Estimate
  annotations <- seurat.obj@meta.data$seurat_clusters
  homotypic.prop <- modelHomotypic(annotations)
  nExp.poi <- round(multiplet_rate * nrow(seurat.obj@meta.data))
  nExp.poi.adj <- round(nExp.poi*(1-homotypic.prop))
  
  # Run DoubletFinder
  seurat.obj <- doubletFinder(seu = seurat.obj, PCs = 1:30, pK = optimal.pk, nExp = nExp.poi.adj) 
  
  # Change name of metadata column with Singlet/Doublet information
  colnames(seurat.obj@meta.data)[grepl('DF.classifications.*', colnames(seurat.obj@meta.data))] <- "doublet_finder"
  
  # Filter out doublets
  singlets <- subset(seurat.obj, doublet_finder == "Singlet")
  
  return(singlets)
} 
```

# Load count data and create Seurat objects with quality control, preprocessing, cell annotation, and filtering out doublets

```{r}
# Define a function for cell type annotation using SingleR
my_cell_annotation <- function(seurat.obj){
  ref <- celldex::DatabaseImmuneCellExpressionData()
  seurat_counts <- GetAssayData(seurat.obj, layer = 'data')
  pred <- SingleR(test = seurat_counts,
                ref = ref,
                labels = ref$label.main)
  seurat.obj$singleR.labels_main <- pred$labels[match(rownames(seurat.obj@meta.data), rownames(pred))]
  return(seurat.obj)
}
```

```{r, message=FALSE, warning=FALSE, eval=FALSE, results='hide'}
# Create the objects and collect in a list
cnt_files_to_load <- list.files(path_to_cnt) # list count file names
path_to_load <- str_c(path_to_cnt, "/", cnt_files_to_load, "/outs/filtered_feature_bc_matrix/")
name_ids <- cnt_files_to_load %>% str_remove("case14_")
seurat.lst <- purrr::map2(name_ids, path_to_load, function(name, path){
  count_matrix <- Read10X(path)
  obj <- CreateSeuratObject(count_matrix) %>% 
    RenameCells(add.cell.id = name)
  obj[["cond"]] <- name
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern="^MT-.")
  obj[["percent.ribo"]] <- PercentageFeatureSet(obj, pattern="^RP[SL]")
  obj <- subset(obj, 
                subset = 
                  nCount_RNA > 600 & nCount_RNA < 15000 & 
                  nFeature_RNA > 200 & nFeature_RNA < 4000 &
                  percent.ribo > 10 & percent.ribo < 60 &
                  percent.mt < 10
                )
  obj <- NormalizeData(obj, normalization.method = "LogNormalize", scale.factor = 10000) 
  obj <- FindVariableFeatures(obj) 
  obj <- ScaleData(obj) 
  obj <- RunPCA(obj) 
  obj <- RunUMAP(obj, dims = 1:30)
  obj <- FindNeighbors(obj, dims = 1:30)              
  obj <- FindClusters(obj, resolution = 0.1)
  obj <- my_doublet_filter(obj)
  obj
})
names(seurat.lst) <- name_ids
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat.lst, path_to_save_obj_lst)
```

# Merge and integrate the objects

```{r, eval=FALSE}
# Merge the objects
seurat <- seurat.lst[[1]]
for (i in 2:length(seurat.lst)) {
  seurat <- merge(x = seurat, y = seurat.lst[[i]])
  seurat.lst[[i]] <- NA
}
rm(seurat.lst)
```

```{r, eval=FALSE}
## Save the merged object
saveRDS(seurat, path_to_save_merged_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Reanalysis after merging and data integration 
seurat <- JoinLayers(seurat)
seurat <- NormalizeData(seurat, normalization.method = "LogNormalize", scale.factor = 10000) 
seurat <- FindVariableFeatures(seurat) 
seurat <- ScaleData(seurat) 
seurat <- RunPCA(seurat) 
seurat <- RunHarmony(seurat, group.by.vars = "cond", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
seurat <- RunTSNE(seurat, dims = 1:20, reduction = "harmony")
seurat <- FindNeighbors(seurat, dims = 1:30)              
seurat <- FindClusters(seurat, resolution = 0.1)
seurat <- my_cell_annotation(seurat) 
```

# Save the integrated object

```{r, eval=FALSE}
# Save the object
saveRDS(seurat, path_to_save_integrated_obj)
```

# Visualisation

```{r}
# Load the integrated data
seurat <- readRDS(path_to_save_integrated_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=15}
# Plot based on conditions, cell types, and clusters
cond.plt <- DimPlot(seurat, reduction = "umap", group.by = "cond") + 
  labs(y= "UMAP 2", x = "UMAP 1")

cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt <- DimPlot(seurat, reduction = "umap", group.by = "seurat_clusters") +
  labs(y= "UMAP 2", x = "UMAP 1")

cond.plt + cell.plt + cluster.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=6, fig.width=7}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("LDHA", "HRK", "PMAIP1", "PRMT5"),
            reduction = "umap")
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("CD8A", "IL7R", "CD4", "CD19", "CD79A", "CD79B", "CD14", "GNLY", "NKG7"),
            reduction = "umap") 
```

# B cell cluster analysis

```{r, message=FALSE, warning=FALSE, results='hide'}
# Extract B cells and re-downstream analysis
Bcell_seurat <- subset(seurat, subset = singleR.labels_main == "B cells")
Bcell_seurat <- NormalizeData(Bcell_seurat, normalization.method = "LogNormalize", scale.factor = 10000) 
Bcell_seurat <- FindVariableFeatures(Bcell_seurat) 
Bcell_seurat <- ScaleData(Bcell_seurat) 
Bcell_seurat <- RunPCA(Bcell_seurat) 
Bcell_seurat <- RunHarmony(Bcell_seurat, group.by.vars = "cond", dims = 1:30) 
Bcell_seurat <- RunUMAP(Bcell_seurat, dims = 1:30, reduction = "harmony")
Bcell_seurat <- FindNeighbors(Bcell_seurat, dims = 1:30)              
Bcell_seurat <- FindClusters(Bcell_seurat, resolution = 0.1)
```

```{r, fig.height=5, fig.width=6}
# Plot based on clusters
DimPlot(Bcell_seurat, reduction = "umap", group.by = "seurat_clusters") +
  labs(y= "UMAP 2", x = "UMAP 1")
```

```{r, fig.height=5, fig.width=6}
# Remove outlier cells outside the main cluster
Bcell_seurat <- subset(Bcell_seurat, subset = seurat_clusters %in% c("0", "1"))

# Plot based on clusters
DimPlot(Bcell_seurat, reduction = "umap", group.by = "seurat_clusters", pt.size = 0.5) +
  labs(y= "UMAP 2", x = "UMAP 1")
```

```{r, fig.height=5, fig.width=5.5}
# Manually remove outlier cells outside the main cluster
#noise_cells <- CellSelector(DimPlot(Bcell_seurat))
noise_cells <- c("CLL_CCCTGATGTCACAGAG-1", 
                 "CLL_GGACGTCGTTGTTGAC-1", 
                 "RT-like_TGCAGGCGTGCTGCAC-1", 
                 "RT-like_TGGAGAGGTTTGTTCT-1")
Bcell_seurat <- subset(Bcell_seurat, cell = setdiff(colnames(Bcell_seurat), noise_cells))

# Plot based on clusters
DimPlot(Bcell_seurat, reduction = "umap", group.by = "seurat_clusters", pt.size = 0.5) +
  labs(y= "UMAP 2", x = "UMAP 1")
```

## Apoptotic-related gene feature plots

```{r, fig.height=5, fig.width=5.5}
# Plot based on conditions
DimPlot(Bcell_seurat, reduction = "umap", pt.size = 0.5, group.by = "cond") +
  scale_color_manual(values = c("CLL" = "steelblue", "RT-like" = "firebrick")) +
  labs(y= "UMAP 2", x = "UMAP 1")
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=15}
# Feature plots
HRK.plt <- FeaturePlot(Bcell_seurat, features = "HRK", reduction = "umap", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(y= "UMAP 2", x = "UMAP 1")

BCL2A1.plt <- FeaturePlot(Bcell_seurat, features = "BCL2A1", reduction = "umap", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(y= "UMAP 2", x = "UMAP 1")

BID.plt <- FeaturePlot(Bcell_seurat, features = "BID", reduction = "umap", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(y= "UMAP 2", x = "UMAP 1")

HRK.plt | BCL2A1.plt | BID.plt
```

## Cell Cycle Score

```{r, warning=FALSE, message=FALSE, fig.height=4, fig.width=15}
# Consider cell cycle genes based on cell cycle upregulation is a hallmark of Richter transformation
Bcell_seurat <- CellCycleScoring(
  Bcell_seurat,
  s.features = cc.genes$s.genes,
  g2m.features = cc.genes$g2m.genes,
  set.ident = FALSE
)

cell_cycle.plt <- DimPlot(Bcell_seurat, group.by = "Phase", pt.size = 1) + labs(y= "UMAP 2", x = "UMAP 1")

umap_s_score <- FeaturePlot(Bcell_seurat, features = "S.Score", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "S Score", y= "UMAP 2", x = "UMAP 1") 
  
umap_g2m_score <- FeaturePlot(Bcell_seurat, features = "G2M.Score", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "G2M Score", y= "UMAP 2", x = "UMAP 1") 

cell_cycle.plt + umap_s_score + umap_g2m_score
```

# Pseudobulk gene expression analysis

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

## Find biomarkers of B cells between RT and CLL conditions

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Find markers between CLL and RT-like conditions  
Idents(Bcell_seurat) <- "cond" # change the default identity
Bcell_RT_CLL.ge <- FindMarkers(
  Bcell_seurat,
  ident.1 = "RT-like",
  ident.2 = "CLL",
  test.use = "t",
  only.pos = FALSE,
  logfc.threshold = 0
)
```

```{r, eval=FALSE}
# Save the dea results
saveRDS(Bcell_RT_CLL.ge, path_to_save_dea)
```

```{r}
# Manipulate the expression data
Bcell_RT_CLL.ge <- readRDS(path_to_save_dea)
Bcell.ge <- cbind(gene = rownames(Bcell_RT_CLL.ge), Bcell_RT_CLL.ge)
lfc_vector <- Bcell.ge$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- Bcell.ge$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, fig.height=5, fig.width=5.5}
# Plot genes of interest
gene_plot = c("HRK", "PMAIP1", "BCL2L11", "BAX", "BAK1", "BBC3", "BMF", "BAD", "BID")
selected_df <- Bcell.ge[Bcell.ge$gene %in% gene_plot, ]

ggplot(selected_df, aes(x = avg_log2FC, y = -log10(p_val), fill = -log10(p_val_adj))) +
  geom_point(shape = 21, size = 4, colour = "black", stroke = 0.6) +
  scale_fill_viridis_c(option = "viridis", direction = -1) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_text_repel(aes(label = gene), size = 4, max.overlaps = 20, color = "black") +
  scale_x_continuous(
    limits = c(-5, 5),
    breaks = seq(-6, 6, 2)
  ) +
  ylim(-10, 180) +
  labs(
    x = "Fold-change RS-like phase vs CLL phase",
    y = "-log(P-Value)",
    fill = "-log(FDR)"
  ) +
  guides(
    fill = guide_colorbar(
      barheight = unit(10, "cm"),   # adjust height
      barwidth = unit(0.6, "cm"),  # adjust width
      direction = "vertical"
    )
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.line = element_line(color = "black", linewidth = 0.5),
    panel.grid = element_blank()  # optional, remove grid lines
  )
```

## Gene Set Enrichment Analysis (GSEA)

### Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector, minGSSize = 25, maxGSSize = 500)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

### GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector, minGSSize = 25, maxGSSize = 500)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

We also performed GSEA with additional enrichment systems such as KEGG pathways, Reactome, and WikiPath.

### KEGG pathway

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector, minGSSize = 25, maxGSSize = 500)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

### Reactome

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector, minGSSize = 25, maxGSSize = 500)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

### WikiPath

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector, minGSSize = 25, maxGSSize = 500)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r}
# Bar plot of GSEA 
gsea_bp <- rbind(HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_OXIDATIVE_PHOSPHORYLATION",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_FATTY_ACID_METABOLISM",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == 
                                                 "REACTOME_AEROBIC_RESPIRATION_AND_RESPIRATORY_ELECTRON_TRANSPORT",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_MITOCHONDRIAL_TRANSLATION",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_CELL_CYCLE_PHASE_TRANSITION",],
                 WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_GLYCOLYSIS_AND_GLUCONEOGENESIS",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_MTORC1_SIGNALING",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_E2F_TARGETS",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_G2M_CHECKPOINT",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_APOPTOSIS",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_REGULATION_OF_APOPTOTIC_SIGNALING_PATHWAY",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_PROGRAMMED_NECROTIC_CELL_DEATH",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_POSITIVE_REGULATION_OF_CARBOHYDRATE_METABOLIC_PROCESS",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == 
                                                 "REACTOME_REGULATION_OF_ENDOGENOUS_RETROELEMENTS",],
                 WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_CILIOPATHIES",]
)

gsea_bp$Description <- gsub("HALLMARK_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("GOBP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("KEGG_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("REACTOME_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("WP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("_V1", "", gsea_bp$Description)
gsea_bp$Description <- gsub("OXIDATIVE_PHOSPHORYLATION", "OXPHOS", gsea_bp$Description)
gsea_bp$Description <- gsub("AEROBIC_RESPIRATION_AND_RESPIRATORY_ELECTRON_TRANSPORT", "RESPIRATORY_ELECTRON_TRANSPORT", gsea_bp$Description)
gsea_bp$Description <- gsub("REGULATION_OF_APOPTOTIC_SIGNALING_PATHWAY", "APOPTOTIC_SIGNALING_PATHWAY", gsea_bp$Description)
gsea_bp$Description <- gsub("POSITIVE_REGULATION_OF_CARBOHYDRATE_METABOLIC_PROCESS", "CARBOHYDRATE_METABOLIC_PROCESS", gsea_bp$Description)
gsea_bp$Description <- gsub("REGULATION_OF_ENDOGENOUS_RETROELEMENTS", "ENDOGENOUS_RETROELEMENTS", gsea_bp$Description)
gsea_bp$Description <- gsub("_", " ", gsea_bp$Description)

gsea_bp$Description <- factor(gsea_bp$Description, levels = gsea_bp$Description)
gsea_bp$Direction <- "Up"
gsea_bp$Direction[gsea_bp$NES<0] <- "Down"

gsea_bar.plt <- ggplot(gsea_bp, aes(x=Description, y=NES, fill=Direction)) +
  geom_bar(stat = "identity", color="black", width = 0.75) +  
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_hline(yintercept = 0) + 
  ylab("NES") + xlab(NULL) + 
  scale_fill_manual(values=c("#8882BA", "#EF7F6A")) + theme(legend.position = "none")

gsea_bar.plt
```

# Trajectory analysis

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in trajectory analysis
library(monocle3)
library(SeuratWrappers)
```

## Transfer seurat data to monocle

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create monocle object and generate graph (reanalysis in monocle)
#cds <- as.cell_data_set(Bcell_seurat) # Convert the seurat object to monocle object using SeuratWrappers
#cds <- preprocess_cds(cds, num_dim = 30)       
#cds <- reduce_dimension(cds, preprocess_method = "PCA", reduction_method = "UMAP")  
#cds <- cluster_cells(cds, reduction_method = "UMAP")
#cds <- learn_graph(cds)
```

```{r, message=FALSE, warning=FALSE}
# Create monocle object and generate graph (use data from seurat)
cds <- as.cell_data_set(Bcell_seurat) # Convert the seurat object to monocle object using SeuratWrappers

# Assign seurat clusters to monocle partitions
recreate.partitions <- c(rep(1, length(cds@colData@rownames)))
names(recreate.partitions) <- cds@colData@rownames
recreate.partitions <- as.factor(recreate.partitions)
cds@clusters@listData[["UMAP"]][["partitions"]] <- recreate.partitions
cds@clusters@listData[["UMAP"]][["clusters"]] <- Bcell_seurat@active.ident

# Assign UMAP coordination
cds@int_colData@listData[["reducedDims"]]@listData[["UMAP"]] <- Bcell_seurat@reductions$umap@cell.embeddings
```

```{r, message=FALSE, warning=FALSE, fig.height=3, fig.width=4}
# Monocle plot before trajectory analysis
plot_cells(cds, 
           color_cells_by = "cluster", 
           label_groups_by_cluster = FALSE, 
           group_label_size = 5
           ) + 
  theme(legend.position = "right")
```

## Trajectory analysis

```{r}
# Learn trajectory
cds <- learn_graph(cds, use_partition = FALSE)
```

```{r, fig.height=3.5, fig.width=9}
# Plot after learning trajectory
cond.plt <- plot_cells(cds, 
                       color_cells_by = "cond",
                       graph_label_size=5,
                       show_trajectory_graph = TRUE
                       ) + 
  ggtitle("cond") +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )

cluster.plt <- plot_cells(cds, 
                       color_cells_by = "seurat_clusters",
                       graph_label_size=5,
                       show_trajectory_graph = TRUE
                       ) + 
  ggtitle("seurat clusters") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

cond.plt + cluster.plt
```

```{r, fig.height=4, fig.width=5}
root.plt <- plot_cells(cds, label_principal_points = TRUE) + 
  ggtitle("root principal node") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

root.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Order cells in pseudotime
cds <- order_cells(cds, reduction_method = "UMAP", root_pr_nodes = "Y_2")

plot_cells(cds, 
           color_cells_by = "pseudotime",
           graph_label_size=5,
           show_trajectory_graph = FALSE
           ) +
  ggtitle("Pseudotime progression") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )
```

```{r}
# Transfer pseudotime to seurat object
Bcell_seurat$pseudotime <- pseudotime(cds)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=17}
# Feature plots between pseudotime and gene expression
pseudotime.plt <- FeaturePlot(Bcell_seurat, features = "pseudotime", reduction = "umap", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(y= "UMAP 2", x = "UMAP 1")

LDHA.plt <- FeaturePlot(Bcell_seurat, features = "LDHA", reduction = "umap", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(y= "UMAP 2", x = "UMAP 1")

HRK.plt <- FeaturePlot(Bcell_seurat, features = "HRK", reduction = "umap", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(y= "UMAP 2", x = "UMAP 1")

PMAIP1.plt <- FeaturePlot(Bcell_seurat, features = "PMAIP1", reduction = "umap", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(y= "UMAP 2", x = "UMAP 1")

pseudotime.plt | LDHA.plt | HRK.plt | PMAIP1.plt
```

# Save the object

```{r, eval=FALSE}
# Save the tumour object
saveRDS(Bcell_seurat, path_to_save_tumour_obj) 
```

# Session Information

```{r}
sessionInfo()
```


