---
title: "Nadeu_2022_tumour_cluster_case19"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

## Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(SeuratWrappers)
library(harmony)
library(tidyverse)
library(DT)
```

```{r}
# Check working path
getwd()
```

## Define parameters

```{r}
# Paths
path_to_obj <- here::here("reanalysis/results/R_objects/4.seurat_leukemic.rds")
path_to_save <- here::here("reanalysis/results/R_objects/5.seurat_clustered_19.rds")


# Colors
color_palette <- c("black", "gray", "red", "yellow", "violet", "green4",
                   "blue", "mediumorchid2", "coral2", "blueviolet",
                   "indianred4", "deepskyblue1", "dimgray", "deeppink1",
                   "green", "lightgray", "hotpink1")


# Source functions
source(here::here("bin/utils.R"))


# Thresholds
max_pct_mt <- 18
```

## Load data

```{r}
seurat <- readRDS(path_to_obj)
```

## Subset and reprocess

```{r, warning=FALSE, message=FALSE}
# Re-downstream analysis after subsetting
seurat <- subset(seurat, donor_id == "19")
seurat <- process_seurat(seurat, dims = 1:20)
DimPlot(seurat, group.by = "subproject")
```

```{r, warning=FALSE, message=FALSE}
# Analyse in BCLLATLAS_10
seurat <- subset(seurat, subproject == "BCLLATLAS_10")
seurat <- process_seurat(seurat, dims = 1:20)
DimPlot(seurat, group.by = "subproject")
FeaturePlot(seurat, "pct_mt") +
  scale_color_viridis_c(option = "magma")
FeaturePlot(seurat, "nFeature_RNA") +
  scale_color_viridis_c(option = "magma")
```

```{r, warning=FALSE, message=FALSE}
# Re-clustering after selecting BCLLATLAS_10
seurat <- FindNeighbors(seurat, reduction = "pca", dims = 1:20)
seurat <- FindClusters(seurat, resolution = 0.3)
DimPlot(seurat)
```

```{r, warning=FALSE, message=FALSE}
# Filter out low quality cells
seurat <- subset(seurat, seurat_clusters != "3")
seurat <- subset(seurat, pct_mt < max_pct_mt)
seurat <- process_seurat(seurat, dims = 1:20)

# Visualisation
DimPlot(seurat)
DimPlot(seurat, group.by = "time_point")
```

```{r}
# Filter out outliers
#outliers <- CellSelector(DimPlot(seurat))
outliers <- c(
  "j7z61du3_vyxon6q4_CATCCGTAGGATCATA-1",
  "j7z61du3_vyxon6q4_GTCCTCATCAGACCTA-1",
  "j7z61du3_vyxon6q4_GTCTCACTCAAGGTGG-1",
  "j7z61du3_vyxon6q4_TGCACGGTCGGTCTAA-1",
  "te04uhdq_2uatjn5y_AAACCCATCGTTCAGA-1",
  "te04uhdq_2uatjn5y_GTCCTCAAGTTTGGCT-1",
  "te04uhdq_2uatjn5y_TCCCAGTAGGTACAAT-1"
)
seurat$is_outlier <- ifelse(colnames(seurat) %in% outliers, TRUE, FALSE)
seurat <- subset(seurat, is_outlier == FALSE)
```

## Cluster

```{r, warning=FALSE, message=FALSE}
seurat <- FindNeighbors(seurat, reduction = "pca", dims = 1:20)
seurat <- FindClusters(seurat, resolution = 0.6)
DimPlot(seurat)
```

From the original results they found that one cluster (cluster 7) was low quality as it had low log2FC of biomarkers and the cells of the cluster were scattered. Then they removed it. However, in our reanalysed results we did not find any clusters having such behaviours. Therefore, we did not remove any clusters.

```{r, warning=FALSE, message=FALSE}
# Is cluster 7 composed of poor-quality cells?
#markers_7 <- FindMarkers(
  #seurat, ident.1 = "7",
  #only.pos = TRUE,
  #logfc.threshold = 0.3
#)
#DT::datatable(markers_7)
```

```{r, warning=FALSE, message=FALSE}
#seurat <- subset(seurat, seurat_clusters != "7")
#seurat <- FindNeighbors(seurat, reduction = "pca", dims = 1:20)
#seurat <- FindClusters(seurat, resolution = 0.6)
#DimPlot(seurat)
#DimPlot(seurat, group.by = "time_point")

# Show CXCR4 expression
FeaturePlot(seurat, "CXCR4") +
  scale_color_viridis_c(option = "magma")
```

As cluster 0 and 3 have the same CXCR4 expression pattern (upregulation) while cluster 1 and 4 have the same trend (downregulation), we will group them into two main clusters.

Moreover, clusters 2 and 5 have no distinctive markers as fold changes are very small:

```{r}
markers_rt <- FindMarkers(
  seurat,
  ident.1 = "2",
  ident.2 = "5",
  only.pos = FALSE
)
DT::datatable(arrange(markers_rt, desc(avg_log2FC)))
```

Rename clusters:

```{r}
seurat$final_clusters <- as.character(seurat$seurat_clusters)
seurat$final_clusters <- factor(case_when(
  seurat$final_clusters == "0" ~ "0",
  seurat$final_clusters == "1" ~ "1",
  seurat$final_clusters == "2" ~ "2",
  seurat$final_clusters == "3" ~ "0",
  seurat$final_clusters == "4" ~ "1",
  seurat$final_clusters == "5" ~ "2",
  seurat$final_clusters == "6" ~ "3",
  seurat$final_clusters == "7" ~ "4",
  seurat$final_clusters == "8" ~ "5"
))
Idents(seurat) <- "final_clusters"
DimPlot(seurat, cols = color_palette)
```

## Save

```{r}
saveRDS(seurat, path_to_save)
```



