---
title: "Nadeu_2022_tumour_annotate_case365"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

## Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(SeuratWrappers)
library(harmony)
library(tidyverse)
library(DT)
```

```{r}
# Check working path
getwd()
```

## Define parameters

```{r}
# Paths
path_to_obj <- here::here("reanalysis/results/R_objects/5.seurat_clustered_365.rds")
#path_to_obj_sampl_artifacts <- here::here("results/R_objects/cll_seurat_annotated.rds")
path_to_save <- here::here("reanalysis/results/R_objects/6.seurat_annotated_365.rds")
path_to_save_markers <- here::here("reanalysis/results/2_clustering_annotation/markers_clusters_365.rds")

# Colors
color_palette <- c("black", "gray", "red", "yellow", "violet", "green4",
                   "blue", "mediumorchid2", "coral2", "blueviolet",
                   "indianred4", "deepskyblue1", "dimgray", "deeppink1",
                   "green", "lightgray", "hotpink1")

# Source functions
source(here::here("bin/utils.R"))

# Thresholds
min_log2FC <- 0.3
alpha <- 0.001
```

## Load data

```{r, fig.height=4, fig.width=5.2}
seurat <- readRDS(path_to_obj)
seurat
DimPlot(seurat, cols = color_palette)
#seurat_bias <- readRDS(path_to_obj_sampl_artifacts)
```

## Cell Cycle Score

```{r, fig.height=4, fig.width=5}
# Consider cell cycle genes based on cell cycle upregulation is a hallmark of Richter transformation
DimPlot(seurat, group.by = "time_point")
FeaturePlot(seurat, "CD3D")
```

```{r, warning=FALSE, message=FALSE, fig.height=4, fig.width=5}
# Consider cell cycle genes based on cell cycle upregulation is a hallmark of Richter transformation
seurat <- CellCycleScoring(
  seurat,
  s.features = cc.genes.updated.2019$s.genes,
  g2m.features = cc.genes.updated.2019$g2m.genes,
  set.ident = FALSE
)

DimPlot(seurat, group.by = "Phase")

umap_s_score <- FeaturePlot(seurat, features = "S.Score") +
  scale_color_viridis_c(option = "magma") +
  labs(title = "S Score") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "plain"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )

umap_g2m_score <- FeaturePlot(seurat, features = "G2M.Score") +
  scale_color_viridis_c(option = "magma") +
  labs(title = "G2M Score") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "plain"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )

umap_cc_combined <- ggpubr::ggarrange(
  plotlist = list(umap_s_score, umap_g2m_score),
  nrow = 2,
  ncol = 1,
  common.legend = FALSE
)

umap_cc_combined
```

## Find Markers

```{r, warning=FALSE, message=FALSE}
markers <- FindAllMarkers(seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)
DT::datatable(markers, options = list(scrollX = TRUE))
```

## Annotation

Important literature to annotate the cells (From Nadeu et al. (2022)):

* [Intraclonal Complexity in Chronic Lymphocytic Leukemia:Fractions Enriched in Recently Born/Divided andOlder/Quiescent Cells](https://molmed.biomedcentral.com/articles/10.2119/molmed.2011.00360).
* [MicroRNA-155influences B-cell receptor signaling and associateswith aggressive disease in chronic lymphocytic leukemia](https://ashpublications.org/blood/article/124/4/546/33343/MicroRNA-155-influences-B-cell-receptor-signaling).
* [Of Lymph Nodes and CLL Cells: Deciphering the Role of CCR7 in the Pathogenesis of CLL and Understanding Its Potential as Therapeutic Target](https://www.frontiersin.org/articles/10.3389/fimmu.2021.662866/full)

| Cluster | Markers                             | Annotation                 |
|---------|-------------------------------------|----------------------------|
| 0       | CXCR4                               | CLL                        |
| 1_0     | CD27                                | CXCR4loCD27hi RT           |
| 1_1     | MIR155HG                            | MIR155HGhi RT              |
| 1_2     | ENO1, TCL1A                         | RT quiescent               |
| 1_3     | PCNA, TOP2A                         | RT proliferative           |

```{r}
# Cell annotation
seurat$annotation_final <- factor(
  seurat$final_clusters,
  levels = c("0", "1_0", "1_1", "1_2", "1_3")
)
new_levels_365 <- c("CLL", "CXCR4loCD27hi RT", "MIR155HGhi RT",
                    "RT quiescent", "RT proliferative")
levels(seurat$annotation_final) <- new_levels_365
seurat$annotation_final <- factor(
  seurat$annotation_final,
  color_annotations$`365`$annotation_final
)
Idents(seurat) <- "annotation_final"


# Plot UMAP
(umap_annotation <- plot_annotation(
  seurat_obj = seurat,
  pt_size = 0.5,
  colors_reference = color_annotations,
  patient_id = "365",
  nothing = TRUE
))
```

## Visualise markers

```{r, fig.height=4, fig.width=5}
# UMAPs
genes_interest <- c("CXCR4", "CD27", "MIR155HG", "ENO1", "TCL1A",
                    "PCNA", "TOP2A")
(feature_plots <- purrr::map(genes_interest, function(x) {
  p <- FeaturePlot(seurat, x, pt.size = 0.5) +
    scale_color_viridis_c(option = "magma")
  p
}))
```

```{r, warning=FALSE, message=FALSE}
# Dot plots
(dot_plot <- plot_dot_plot(
  seurat,
  goi = rev(genes_interest),
  colors_reference = color_annotations,
  patient_id = "365"
))

# Violin plots
(vln_plot_s <- plot_violin_plot(
  seurat,
  continuous_var = "S.Score",
  ylab = "S Phase Score",
  colors_reference = color_annotations,
  patient_id = "365"
))
(vln_plot_g2m <- plot_violin_plot(
  seurat,
  continuous_var = "G2M.Score",
  ylab = "G2M Phase Score",
  colors_reference = color_annotations,
  patient_id = "365"
))
```

## Save

```{r, eval=FALSE}
# Save Seurat object
saveRDS(seurat, path_to_save)
```

```{r, eval=FALSE}
# Save markers
markers$annotation <- factor(markers$cluster)
levels(markers$annotation) <- new_levels_365
markers_list <- purrr::map(levels(markers$annotation), function(x) {
  df <- markers[markers$annotation == x, ]
  df <- df[, c(7, 1, 5, 2:4, 6, 8)]
  df
})
names(markers_list) <- levels(markers$annotation)
markers_list <- markers_list[color_annotations$`365`$annotation_final]
markers_final <- bind_rows(markers_list)
saveRDS(markers_list, path_to_save_markers)
saveRDS(
  markers_final,
  here::here("reanalysis/results/tables/markers/markers_annotated_clusters_patient_365.rds")
)
saveRDS(markers_list, path_to_save_markers)
openxlsx::write.xlsx(
  x = markers_list,
  file = "reanalysis/results/tables/markers/markers_annotated_clusters_patient_365.xlsx"
)
```






