---
title: "GSE165087_repeat"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

## Import relevant packages

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
```

## Set a working path directory

```{r}
setwd('/mnt/hp-storage/users/pakorns/scRNAseq/GSE165087')
```

## Import the merged object

```{r}
merged.seu <- readRDS('merged_annotated.rds')
```

## Visualise cell annotation

### Primary cell annotation

```{r fig.height=4.5, fig.width=6}
primary_annotation_umap <- DimPlot(merged.seu, reduction = 'umap', group.by = 'singleR.labels_main', raster=FALSE) +
  labs(y= "UMAP 2", x = "UMAP 1")
primary_annotation_umap
```

### Secondary cell annotation

```{r}
secondary_annotation_umap <- DimPlot(merged.seu, reduction = 'umap', group.by = 'singleR.labels_fine', raster=FALSE) +
  labs(y= "UMAP 2", x = "UMAP 1")
secondary_annotation_umap
```

## Other visualisation

### UMAP based on case origins

```{r fig.height=4.5, fig.width=5}
case_origin_umap <- DimPlot(merged.seu, reduction = 'umap', group.by = 'case', raster=FALSE, pt.size = 0.5) +
  labs(y= "UMAP 2", x = "UMAP 1")
case_origin_umap
```

### UMAP based on conditions

```{r fig.height=4.5, fig.width=5}
cond_origin_umap <- DimPlot(merged.seu, reduction = 'umap', group.by = 'cond', raster=FALSE) +
  labs(y= "UMAP 2", x = "UMAP 1")
cond_origin_umap
```

### UMAP based on tissue origins

```{r fig.height=4.5, fig.width=5}
tissue_origin_umap <- DimPlot(merged.seu, reduction = 'umap', group.by = 'tissue', raster=FALSE) +
  labs(y= "UMAP 2", x = "UMAP 1")
tissue_origin_umap
```

## Proof for reproducibility

### Select Patient 1 to 6 with B cells, and pre- and post-treatment stage

```{r}
# Explore categorical data in time column
unique(merged.seu@meta.data$time)
```

```{r}
# Select patient 1-6 with B cells, and pre- and post-treatment stage
selected.seu <- subset(
  merged.seu, 
  subset = singleR.labels_main == "B cells" & 
    case %in% c(1, 2, 3, 4, 5, 6) &
    time %in% c("Pretreatment_FCR", "Relapse_post_FCR", "Relapse_post_RIC")
)
```

```{r}
# Create a new column which is case-time
selected.seu$case_time <- paste0("CLL", selected.seu$case, "_", selected.seu$time)
```

```{r}
# Count B cells in each case-time condition
table(selected.seu$case_time)  # replace 'condition' with your actual column name
```

```{r}
# Change colour manually
my_colors <- c(
  "CLL1_Pretreatment_FCR" = "#a6cee3",   # Pale Blue
  "CLL1_Relapse_post_FCR" = "#1f78b4",  # Dark Blue
  "CLL2_Pretreatment_FCR" = "#b2df8a",   # Pale Green
  "CLL2_Relapse_post_FCR" = "#33a02c",  # Dark Green
  "CLL3_Pretreatment_FCR" = "#fb9a99",   # Pale Red
  "CLL3_Relapse_post_FCR" = "#e31a1c",  # Dark Red
  "CLL4_Pretreatment_FCR" = "#fdbf6f",   # Pale Orange
  "CLL4_Relapse_post_RIC" = "#ff7f00",  # Dark Orange
  "CLL5_Pretreatment_FCR" = "#cab2d6",   # Pale Purple
  "CLL5_Relapse_post_RIC" = "#6a3d9a",  # Dark Purple
  "CLL6_Pretreatment_FCR" = "#ffff99",   # Pale Yellow
  "CLL6_Relapse_post_RIC" = "#b15928"   # Dark Brown
)
```

```{r fig.height=4.5, fig.width=7}
# UMAP plot based on case_time
DimPlot(selected.seu, 
        reduction = "umap", 
        group.by = "case_time",
        raster = FALSE) +
  scale_color_manual(values = my_colors) +
  labs(y= "UMAP 2", x = "UMAP 1")
```

```{r}
# Find markers
Idents(selected.seu) <- "case_time" # change the default identity
seu.markers <- FindAllMarkers(selected.seu, only.pos = TRUE)
```

```{r}
# Plot some gene expressions (CD24, MEF2C, CXCR4, RGS1, RGS2)
CD24_vinplot <- VlnPlot(selected.seu, features = c("CD24"), pt.size = 0, raster = FALSE) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_fill_manual(values = my_colors)
MEF2C_vinplot <- VlnPlot(selected.seu, features = c("MEF2C"), pt.size = 0, raster = FALSE) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_fill_manual(values = my_colors)
CXCR4_vinplot <- VlnPlot(selected.seu, features = c("CXCR4"), pt.size = 0, raster = FALSE) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_fill_manual(values = my_colors)
RGS1_vinplot <- VlnPlot(selected.seu, features = c("RGS1"), pt.size = 0, raster = FALSE) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_fill_manual(values = my_colors)
RGS2_vinplot <- VlnPlot(selected.seu, features = c("RGS2"), pt.size = 0, raster = FALSE) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_fill_manual(values = my_colors)
```

```{r}
# Show some gene expressions (CD24, MEF2C) downregulated at relapse phase
CD24_vinplot + MEF2C_vinplot + patchwork::plot_layout(ncol = 2)
```

```{r fig.height=4, fig.width=9}
# Show some gene expressions (CXCR4, RGS1, RGS2) upregulated at relapse phase
CXCR4_vinplot + RGS1_vinplot + RGS2_vinplot + patchwork::plot_layout(ncol = 3)
```

```{r}
# Measure gene expression between pre- and post-treatment
selected.seu$pre_post <- with(
  selected.seu@meta.data,
  ifelse(grepl("pre", time, ignore.case = TRUE), "pre",
         ifelse(grepl("post", time, ignore.case = TRUE), "post", NA))
)
```

```{r}
# Find markers between pre- and post-treatment  
Idents(selected.seu) <- "pre_post" # change the default identity
pre_post.ge <- FindMarkers(selected.seu, ident.1 = 'post', ident.2 = 'pre')
```

```{r}
# Show some expression results
pre_post.ge$gene <- rownames(pre_post.ge)
pre_post.ge$neg_log_p <- -log10(pre_post.ge$p_val_adj)
pre_post.ge <- pre_post.ge %>%
  mutate(adapt_log = ifelse(is.infinite(neg_log_p), 302, neg_log_p))
head(pre_post.ge)
```

```{r}
# Find second maximum negative log p-value (apart from infinity)
second_max_unique <- sort(unique(pre_post.ge$neg_log_p), decreasing = TRUE)[2]
second_max_unique
```

```{r}
# Plot a volcano plot
genes_of_interest <- c("CD24", "MEF2C", "RGS1", "RGS2", "ID3", "CXCR4", "ZFP36L2", "ZFP36")
ggplot(pre_post.ge, aes(x = avg_log2FC, y = adapt_log)) +
  geom_point(aes(color = p_val_adj < 0.05 & abs(avg_log2FC) > 0.25), size = 0.5) +
  scale_color_manual(values = c("grey", "orange")) +
  geom_text_repel(data = subset(pre_post.ge, gene %in% genes_of_interest),
                  aes(label = gene),
                  max.overlaps = Inf, size = 4) +
  labs(x = "Log2 Fold Change", y = "-log10 Adjusted P-value") +
  scale_y_continuous(limits = c(0, 310)) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Select Patient 6 to 8 with B cells, and pre- and ibrutinib stage

```{r}
# Subset B cells from case 6 to 8
ibrutinib.seu <- subset(
  merged.seu,
  subset = singleR.labels_main == 'B cells' &
    case %in% c(6, 7, 8))
```

```{r}
# Explore categorical data in time column
unique(ibrutinib.seu@meta.data$time)
```

```{r}
# Create a new column which is ibrutinib_course
ibrutinib.seu$ibrutinib_course <- ifelse(grepl('Pre', ibrutinib.seu$time), 'Baseline', 
                                         ifelse(grepl('Relapse', ibrutinib.seu$time), 'Baseline', 'Ibrutinib'))
```

```{r}
# Find markers
Idents(ibrutinib.seu) <- "ibrutinib_course" # change the default identity
ibrutinib_course.ge <- FindAllMarkers(ibrutinib.seu, only.pos = TRUE)
```

```{r}
# Create a metadata column to identify Baseline vs Ibrutinib
ibrutinib.seu$patient_phase <- paste0(
  "CLL", ibrutinib.seu$case, "_", ibrutinib.seu$ibrutinib_course
)
Idents(ibrutinib.seu) <- "patient_phase"

# Genes to plot
genes_to_plot <- c("RAC2", "TPM3", "DBI", "CD99", "LINC01857", 
                   "DUSP1", "FOS", "TMSB4X", "CD52", "TXNIP", 
                   "BTG2", "MS4A1", "CCDC88A", "RGS10", "PLPP5", 
                   "IGFLR1", "FMOD", "LINC01480", "MIR155HG", "PIGR")

# Calculate average expression
avg_exp <- AverageExpression(ibrutinib.seu, features = genes_to_plot, group.by = "patient_phase")
mat <- avg_exp$RNA
mat_scaled <- t(scale(t(mat)))
mat_raw <- as.matrix(mat)
```

```{r}
# Identify NA and inf before visualisation
summary(as.vector(mat_scaled))
any(is.na(mat_scaled))
any(is.infinite(mat_scaled))
```

```{r}
# Remove rows with NA/Inf
mat_scaled <- mat_scaled[complete.cases(mat_scaled) & apply(mat_scaled, 1, function(x) all(is.finite(x))), ]
mat_raw <- mat_raw[complete.cases(mat_raw) & apply(mat_raw, 1, function(x) all(is.finite(x))), ]
```

```{r}
# Extract treatment phase from colnames (e.g., CLL1_Baseline → Baseline)
treatment_phase <- sub(".*-(Baseline|Ibrutinib)$", "\\1", colnames(mat_scaled))

# Annotation dataframe
annotation_col <- data.frame(Treatment = treatment_phase)
rownames(annotation_col) <- colnames(mat_scaled)

# Ensure correct labels and factors
annotation_col$Treatment <- factor(annotation_col$Treatment, levels = c("Baseline", "Ibrutinib"))

# Order so Baseline samples come first, then Ibrutinib
order_cols <- order(annotation_col$Treatment)
mat_scaled_ordered <- mat_scaled[, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]

# Colour settings
ann_colors <- list(Treatment = c("Baseline" = "#1f78b4", "Ibrutinib" = "#33a02c"))

# Plot
pheatmap(
  mat_scaled_ordered,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  annotation_colors = ann_colors,
  fontsize_row = 8,
  fontsize_col = 8
)
```

```{r fig.height=4, fig.width=4}
# Heatmap based on the scaled expression

# Extract patient and treatment info from column names
patient_id <- sub("^(CLL\\d+)-.*", "\\1", colnames(mat_scaled))
treatment_phase <- sub(".*-(Baseline|Ibrutinib)$", "\\1", colnames(mat_scaled))

# Create annotation dataframe
annotation_col <- data.frame(
  Patient = factor(patient_id, levels = c("CLL6", "CLL7", "CLL8")),
  Treatment = factor(treatment_phase, levels = c("Baseline", "Ibrutinib"))
)
rownames(annotation_col) <- colnames(mat_scaled)

# Order by patient ID and then by treatment phase
order_cols <- order(annotation_col$Treatment, annotation_col$Patient)
mat_scaled_ordered <- mat_scaled[, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]

# Colour settings
ann_colors <- list(
  Treatment = c("Baseline" = "#1f78b4", "Ibrutinib" = "#de2d26"),
  Patient = c("CLL6" = "#d95f0e", "CLL7" = "#2ca25f", "CLL8" = "#fec44f") 
)

# Plot
pheatmap(
  mat_scaled_ordered,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = TRUE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  annotation_colors = ann_colors,
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  gaps_col = which(diff(as.numeric(factor(annotation_col_ordered$Treatment))) != 0)
)
```

```{r fig.height=4, fig.width=4}
# Heatmap based on the raw expression

# Extract patient and treatment info from column names
patient_id <- sub("^(CLL\\d+)-.*", "\\1", colnames(mat_scaled))
treatment_phase <- sub(".*-(Baseline|Ibrutinib)$", "\\1", colnames(mat_scaled))

# Create annotation dataframe
annotation_col <- data.frame(
  Patient = factor(patient_id, levels = c("CLL6", "CLL7", "CLL8")),
  Treatment = factor(treatment_phase, levels = c("Baseline", "Ibrutinib"))
)
rownames(annotation_col) <- colnames(mat_raw)

# Order by patient ID and then by treatment phase
order_cols <- order(annotation_col$Treatment, annotation_col$Patient)
mat_raw_ordered <- mat_raw[, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]

# Colour settings
ann_colors <- list(
  Treatment = c("Baseline" = "#1f78b4", "Ibrutinib" = "#de2d26"),
  Patient = c("CLL6" = "#d95f0e", "CLL7" = "#2ca25f", "CLL8" = "#fec44f") 
)

# Plot
pheatmap(
  mat_raw_ordered,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = TRUE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  annotation_colors = ann_colors,
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  gaps_col = which(diff(as.numeric(factor(annotation_col_ordered$Treatment))) != 0)
)
```

```{r fig.height=4, fig.width=4}
# Extract patient and treatment info from column names
patient_id <- sub("^(CLL\\d+)-.*", "\\1", colnames(mat_scaled))
treatment_phase <- sub(".*-(Baseline|Ibrutinib)$", "\\1", colnames(mat_scaled))

# Create annotation dataframe
annotation_col <- data.frame(
  Patient = factor(patient_id, levels = c("CLL6", "CLL7", "CLL8")),
  Treatment = factor(treatment_phase, levels = c("Baseline", "Ibrutinib"))
)
rownames(annotation_col) <- colnames(mat_scaled)

# Order by patient ID and then by treatment phase
order_cols <- order(annotation_col$Treatment, annotation_col$Patient)
mat_scaled_ordered <- mat_scaled[, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]

# Colour settings
ann_colors <- list(
  Treatment = c("Baseline" = "#1f78b4", "Ibrutinib" = "#de2d26"),
  Patient = c("CLL6" = "#d95f0e", "CLL7" = "#2ca25f", "CLL8" = "#fec44f") 
)

# Get min and max of matrix for breaks and labels
min_val <- round(min(mat_scaled_ordered), 2)
max_val <- round(max(mat_scaled_ordered), 2)

# Plot
pheatmap(
  mat_scaled_ordered,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = TRUE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  annotation_colors = ann_colors,
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  gaps_col = which(diff(as.numeric(factor(annotation_col_ordered$Treatment))) != 0),
  legend_breaks = c(min_val, max_val),
  legend_labels = c("min", "max"),
  legend = TRUE
)
```

### Select Patient 6

```{r}
# Subset B cells from case 6 to 8
Case6.seu <- subset(
  merged.seu,
  subset = singleR.labels_main == 'B cells' &
    case == 6)
```

```{r}
# Explore categorical data in time column
unique(Case6.seu$time)
```

```{r}
# Create a new column which is course for ibrutinib treatment
Case6.seu$course <- ifelse(Case6.seu$time %in% c("Pretreatment_FCR", "Relapse_post_RIC"), "Baseline",
                    ifelse(Case6.seu$time == "Ibrutinib_1_month", "Ibrutinib",
                    ifelse(Case6.seu$time == "Ibrutinib_21_month", "Relapse", NA)))
```

```{r}
# Find markers
Idents(Case6.seu) <- "course" # change the default identity
Case6_course.ge <- FindAllMarkers(Case6.seu, only.pos = TRUE)
```

```{r}
# Plot some gene expressions (MS4A1, MIR155HG)
Case6_colors <- c(
  "Baseline" = "#e31a1c",   # Dark Red
  "Ibrutinib" = "#33a02c",  # Dark Green
  "Relapse" = "#1f78b4"     # Dark Blue
)

MS4A1_vinplot <- VlnPlot(Case6.seu, features = c("MS4A1"), pt.size = 0, raster = FALSE) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  scale_fill_manual(values = Case6_colors)
MIR155HG_vinplot <- VlnPlot(Case6.seu, features = c("MIR155HG"), pt.size = 0, raster = FALSE) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
  scale_fill_manual(values = Case6_colors)

MS4A1_vinplot + MIR155HG_vinplot + patchwork::plot_layout(ncol = 2)
```

```{r}
# Genes to plot
genes_to_plot <- c("RAC2", "TPM3", "DBI", "CD99", "LINC01857", 
                   "DUSP1", "FOS", "TMSB4X", "CD52", "TXNIP", 
                   "BTG2", "MS4A1", "CCDC88A", "RGS10", "PLPP5", 
                   "IGFLR1", "FMOD", "LINC01480", "MIR155HG", "PIGR")

Case6.seu <- ScaleData(Case6.seu, features = genes_to_plot)

DoHeatmap(Case6.seu, features = genes_to_plot) + 
  scale_fill_distiller(palette = "RdYlBu", direction = 1) + 
  NoLegend()
```

```{r fig.height=5, fig.width=5}
# Gene expression plot (make columns in each categories equally)

set.seed(123)  # for reproducibility

# Extract metadata and add cell names as a column
meta <- Case6.seu@meta.data %>%
  tibble::rownames_to_column(var = "cell")  # make cell names a column

# Sample equal number of cells per group
min_cells <- min(table(Case6.seu$course))

subset_cells <- meta %>%
  dplyr::group_by(course) %>%
  dplyr::sample_n(min_cells) %>%
  dplyr::pull(cell)

Case6.seu.down <- subset(Case6.seu, cells = subset_cells)

DoHeatmap(Case6.seu.down, 
          features = genes_to_plot, 
          group.by = "course", 
          angle = 30,
          ) +
  NoLegend() +
  scale_fill_distiller(palette = "RdYlBu", direction = 1)
```

### Select Patient 9 with B cells, and CLL and RT stage

```{r}
# Subset B cells from case 9
Case9.seu <- subset(
  merged.seu,
  subset = singleR.labels_main == 'B cells' &
    case == 9)
```

```{r}
# Create a new column which is cond_tissue
Case9.seu$cond_tissue <- paste0(Case9.seu$cond, "_", Case9.seu$tissue)
```

```{r}
# Explore categorical data in cond_tissue column
unique(Case9.seu$cond_tissue)
```

```{r}
# Change colour manually
case9_colors <- c(
  "CLL_PB" = "#fff700",   # Lemon Yellow
  "RT_PB" = "#e31a1c",  # Dark Red
  "RT_BM" = "#b15928"   # Dark Brown
)
```

```{r fig.height=4.5, fig.width=6}
# UMAP plot based on cond_tissue

DimPlot(Case9.seu, 
        reduction = "umap", 
        group.by = "cond_tissue",
        raster = FALSE) +
  scale_color_manual(values = case9_colors) +
  labs(y= "UMAP 2", x = "UMAP 1")
```

```{r}
# Find gene expression between RT and CLL
Idents(Case9.seu) <- "cond" # change the default identity
RT.ge <- FindMarkers(Case9.seu, ident.1 = 'RT', ident.2 = 'CLL')
```

```{r}
# Show some results
RT.ge$gene <- rownames(RT.ge)
RT.ge$neg_log_p <- -log10(RT.ge$p_val_adj)
RT.ge <- RT.ge %>%
  mutate(adapt_log = ifelse(is.infinite(neg_log_p), 302, neg_log_p))
head(RT.ge)
```

```{r}
# Find second maximum negative log p-value (apart from infinity)
second_max_unique <- sort(unique(RT.ge$neg_log_p), decreasing = TRUE)[2]
second_max_unique
```

```{r}
# Plot a volcano plot
genes_of_interest <- c("IGHV3-7", "NFKBIA", "FOS", "FOSB", "JUN", "JUNB", "IGLC7", "CD99", "CD83", "CD24", "LTB", "CD70")
ggplot(RT.ge, aes(x = avg_log2FC, y = adapt_log)) +
  geom_point(aes(color = p_val_adj < 0.05 & abs(avg_log2FC) > 0.25), size = 0.5) +
  scale_color_manual(values = c("grey", "orange")) +
  geom_text_repel(data = subset(RT.ge, gene %in% genes_of_interest),
                  aes(label = gene),
                  max.overlaps = Inf, size = 4) +
  labs(x = "Log2 Fold Change", y = "-log10 Adjusted P-value") +
  scale_y_continuous(limits = c(0, 400)) +
  theme_minimal() +
  theme(legend.position = "none")
```

