---
title: "Rigo_2024_systematic_review"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This source code we analysed single-cell RNA-seq for RT and CLL conditions both gene expression and functional enrichment analysis before we will compare the results to other studies for our systematic review. We analysed for [Rigo et. al. (2024)](https://www.nature.com/articles/s41419-024-06707-5) study which performed single-cell analysis between CLL and RT-like conditions in one patient. RT-like condition is defined by large B cells detection in peripheral blood with B symptoms and elevated serum LDH after ibrutinib interruption.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(DoubletFinder)
library(SingleR)
library(celldex)
library(sctransform)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_count_matrices <- here::here("scRNAseq/Rigo_2024_scRNAseq/count_matrices")
path_to_save_obj_lst <- here::here("scRNAseq/systematic_review/5-Rigo_2024/R_objects/Rigo_seu_lst.rds")
path_to_save_merged_obj <- here::here("scRNAseq/systematic_review/5-Rigo_2024/R_objects/Rigo_merged_seu.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/systematic_review/5-Rigo_2024/R_objects/Rigo_integrated_seu.rds")
path_to_save_tumour_obj <- here::here("scRNAseq/systematic_review/5-Rigo_2024/R_objects/Rigo_tumour_seu.rds")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Explore directories in count matrix path
lst_files <- list.files(path_to_count_matrices)
lst_files
```

For Rigo et al. (2024), they performed single-cell analysis for case 14. We will import count matrices and convert them to Seurat objects with quality control and pre-processing.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create the objects and collect in a list
path_to_download <- str_c(path_to_count_matrices, "/", lst_files, "/outs/filtered_feature_bc_matrix/")
conditions <- sub(".*_", "", lst_files) # Replace everything before underscore to "" (empty string)
seurat.lst <- purrr::pmap(
  list(path_to_download, lst_files, conditions), 
  function(path, case_id, cond){
    count_matrix <- Read10X(path)
    seurat.obj <- CreateSeuratObject(count_matrix, min.cells=10) %>% 
      RenameCells(add.cell.id = case_id)
    seurat.obj[["ref"]] <- "Rigo_2024"
    seurat.obj[["case_id"]] <- case_id
    seurat.obj[["case"]] <- "14"
    seurat.obj[["cond"]] <- cond
    seurat.obj[["tissue"]] <- "PB"
    seurat.obj <- seurat_quality_control(seurat.obj, mad_cutoff=TRUE)
    seurat.obj <- seurat_preprocess(seurat.obj)
    seurat.obj <- doublet_filter(seurat.obj, sct=FALSE)
    return(seurat.obj)
    }
  )
names(seurat.lst) <- lst_files
```

```{r, eval=FALSE}
# Explore data structures of Seurat objects in the list
seurat.lst
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat.lst, path_to_save_obj_lst)
```

# Merge and integrate the objects

```{r, eval=FALSE}
# Merge the objects
seurat <- seurat.lst[[1]]
for (i in 2:length(seurat.lst)) {
  seurat <- merge(x = seurat, y = seurat.lst[[i]])
  seurat.lst[[i]] <- NA
}
rm(seurat.lst)
```

```{r, eval=FALSE}
# Save the merged object
saveRDS(seurat, path_to_save_merged_obj)
```

We used Harmony package for data integration and batch correction, and annotated the cell types using SingleR package.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Reanalysis after merging and data integration 
DefaultAssay(seurat) <- "RNA"
seurat <- JoinLayers(seurat)
seurat <- SCTransform(seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
seurat <- RunPCA(seurat)
seurat <- RunHarmony(seurat, group.by.vars = "case_id", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
seurat <- FindNeighbors(seurat, dims = 1:30)              
seurat <- FindClusters(
    seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )
seurat <- singleR_cell_annotation(seurat) 
```

# Save the integrated object

```{r, eval=FALSE}
saveRDS(seurat, path_to_save_integrated_obj)
```

# Visualisation after data integration

```{r}
# Define path
path_to_obj <- path_to_save_integrated_obj

# Load the data
seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=10}
# Plot based on case_id, conditions, cell types, and clusters
case_id.plt <- DimPlot(seurat, reduction = "umap", group.by = "case_id") + 
  labs(y= "UMAP 2", x = "UMAP 1")

case.plt <- DimPlot(seurat, reduction = "umap", group.by = "case") + 
  labs(y= "UMAP 2", x = "UMAP 1")

ref.plt <- DimPlot(seurat, reduction = "umap", group.by = "ref") + 
  labs(y= "UMAP 2", x = "UMAP 1")

cond.plt <- DimPlot(seurat, reduction = "umap", group.by = "cond") + 
  labs(y= "UMAP 2", x = "UMAP 1")

tissue.plt <- DimPlot(seurat, reduction = "umap", group.by = "tissue") +
  labs(y= "UMAP 2", x = "UMAP 1")

cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")

(case_id.plt + case.plt) / (ref.plt + cond.plt) / (tissue.plt + cell.plt)
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
varied_clusters.plt <- plot_varying_cluster_resolution(seurat)
varied_clusters.plt
```

```{r, fig.height=4, fig.width=5}
# Plot seurat clusters at resolution=0.2
clusters.plt <- DimPlot(seurat, reduction = "umap", group.by = "SCT_snn_res.0.2", raster=FALSE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

clusters.plt
```

# Tumour cell extraction

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Extract tumour cells 
tumour_seurat <- subset(seurat, subset = SCT_snn_res.0.2 %in% c(0, 1))
tumour_seurat <- SCTransform(tumour_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
tumour_seurat <- RunPCA(tumour_seurat)
tumour_seurat <- RunHarmony(tumour_seurat, group.by.vars = "case_id", dims = 1:30) 
tumour_seurat <- RunUMAP(tumour_seurat, dims = 1:30, reduction = "harmony")
tumour_seurat <- FindNeighbors(tumour_seurat, dims = 1:30)              
tumour_seurat <- FindClusters(tumour_seurat, resolution=0.25)
```

```{r, eval=FALSE}
# Save the tumour cells
saveRDS(tumour_seurat, path_to_save_tumour_obj)
```

```{r}
# Define path
path_to_obj <- path_to_save_tumour_obj

# Load the data
tumour_seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=9.5}
# Plot based on sample_id, conditions, tissue, and cell types
case_id.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "case_id") +
  labs(y= "UMAP 2", x = "UMAP 1")

cond.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "cond") + 
  labs(y= "UMAP 2", x = "UMAP 1")

tissue.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "tissue") +
  labs(y= "UMAP 2", x = "UMAP 1")

clusters.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "seurat_clusters") +
  labs(y= "UMAP 2", x = "UMAP 1")

(case_id.plt + cond.plt) / (tissue.plt + clusters.plt)
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(tumour_seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

# Pseudobulk gene expression analysis and GSEA

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

```{r}
# Define the paths to save and load
path_to_save_dea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/expression_data/Rigo_dea.rds")
path_to_save_hm_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/enrichment_data/Rigo_hm_gsea.rds")
path_to_save_go_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/enrichment_data/Rigo_go_gsea.rds")
path_to_save_kegg_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/enrichment_data/Rigo_kegg_gsea.rds")
path_to_save_reactome_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/enrichment_data/Rigo_reactome_gsea.rds")
path_to_save_wikipath_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/enrichment_data/Rigo_wikipath_gsea.rds")
```

## Find biomarkers of B cells between RT and CLL conditions

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Find markers between CLL and RT-like conditions  
Idents(tumour_seurat) <- "cond" # change the default identity
tumour_seurat <- PrepSCTFindMarkers(tumour_seurat) # Prepare SCT assay for DE testing
tumour.ge <- FindMarkers(
  tumour_seurat,
  ident.1 = "RT-like",
  ident.2 = "CLL",
  test.use = "wilcox",
  only.pos = FALSE,
  logfc.threshold = 0
)
dea.df <- cbind(gene = rownames(tumour.ge), tumour.ge)
```

```{r, eval=FALSE}
# Save gene expression data
saveRDS(dea.df, path_to_save_dea)
```

## Gene Set Enrichment Analysis (GSEA)

```{r}
# Load the saved expression data
path_to_dea <- path_to_save_dea
dea.df <- readRDS(path_to_dea)
```

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea.df$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

### Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will explore the trends of other hallmark terms reported from other studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as E2F targets, Myc targets, and oxidative phosphorylation (OXPHOS) were upregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

### GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will explore the trends of other GO:BP terms reported from other studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as OXPHOS and mitochondrial translation were upregulated while BCR signalling was downregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

We also performed GSEA with additional enrichment systems such as KEGG pathways, Reactome, and WikiPath.

### KEGG pathway

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

### Reactome

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

### WikiPath

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save GSEA results
saveRDS(HM_gsea, path_to_save_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_go_gsea)
saveRDS(KEGG_gsea, path_to_save_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_wikipath_gsea)
```

# Pseudobulk analysis using DESeq2

As this study perform single-cell analysis on only one patient, we cannot directly perform pseudobulk analysis with DESeq2 that aggregate read counts based on cases. Then we will split the cells from both conditions (RT-like and CLL) into two main part using `createDataPartition` from the caret package and create pseudo cases (case 1 and 2 from first and second part, respectively). After that, we can perform DESeq2 which requires the variant within the groups.

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(DESeq2)
library(caret)
```

```{r}
# Define paths to save the results
path_to_save_deseq2_dea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/pseudo_bulk_deseq2/Rigo_deseq2_dea.rds")
path_to_save_deseq2_hm_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/pseudo_bulk_deseq2/Rigo_deseq2_hm_gsea.rds")
path_to_save_deseq2_go_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/pseudo_bulk_deseq2/Rigo_deseq2_go_gsea.rds")
path_to_save_deseq2_kegg_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/pseudo_bulk_deseq2/Rigo_deseq2_kegg_gsea.rds")
path_to_save_deseq2_reactome_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/pseudo_bulk_deseq2/Rigo_deseq2_reactome_gsea.rds")
path_to_save_deseq2_wikipath_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/pseudo_bulk_deseq2/Rigo_deseq2_wikipath_gsea.rds")
```

```{r}
# Split the cells in the object to two main parts using caret package
set.seed(123)
first_idx <- createDataPartition(tumour_seurat@meta.data$cond, p = 0.5, list = FALSE)
first_cells <- colnames(tumour_seurat)[first_idx]

# Add new columns naming pseudo_case to the object
tumour_seurat$pseudo_case <- ifelse(colnames(tumour_seurat) %in% first_cells, 1, 2)
tumour_seurat$pseudo_case_cond <- str_c(tumour_seurat$cond, "-", tumour_seurat$pseudo_case) # create CLL-1, RT-like-1 etc.
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Pseudobulk analysis by aggregation based on cases
RT_subjects <- c("RT-like-1", "RT-like-2")
dea.df <- pseudobulk_deseq2(tumour_seurat, aggregate_group.by = "pseudo_case_cond", RT_subjects)

# Save the expression results
saveRDS(dea.df, path_to_save_deseq2_dea)
```

```{r}
# Show some expression results
dea.df <- readRDS(path_to_save_deseq2_dea)
head(dea.df)
```

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df) 
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

## GSEA

### Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

### GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

### KEGG pathway

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

### Reactome

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

### WikiPath

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

The results from DESeq2 were consistent with other studies. However, to be consistent with our standard pipeline for analysis, we will also collect the result from `FindMarkers()` for further steps.

```{r, eval=FALSE}
# Save the enrichment results
saveRDS(HM_gsea, path_to_save_deseq2_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_deseq2_go_gsea)
saveRDS(KEGG_gsea, path_to_save_deseq2_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_deseq2_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_deseq2_wikipath_gsea)
```

# Session Information

```{r}
sessionInfo()
```
