---
title: "Rejeski_2022_GSE201704_preprocessing_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This source code we generated Seurat objects for single-cell RNA-seq datasets for a RT patient from [Rejeski et. al. (2022)](https://ashpublications.org/blood/article/140/20/2175/485763/Oligoclonal-T-cell-expansion-in-a-patient-with). This study contains two multiplexed data sets such as before and after CD19 CAR-T therapy.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(DoubletFinder)
library(SingleR)
library(celldex)
library(sctransform)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_count_matrices <- here::here("scRNAseq/GSE201704/GSM6069857")
path_to_save_obj_lst <- here::here("scRNAseq/GSE201704/GSM6069857/R_objects/Rejeski_2022_GSE201704_seu_lst.rds")
path_to_save_merged_obj <- here::here("scRNAseq/GSE201704/GSM6069857/R_objects/Rejeski_2022_GSE201704_merged_seu.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/GSE201704/GSM6069857/R_objects/Rejeski_2022_GSE201704_integrated_seu.rds")
path_to_save_tumour_obj <- here::here("scRNAseq/GSE201704/GSM6069857/R_objects/Rejeski_2022_GSE201704_tumour_seu.rds")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

We will import count matrices, demultiplex, and convert them to Seurat objects with quality control and pre-processing.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create objects with demultiplexing and then collect as a list
sample_lst <- c("CMO305", "CMO306")
path_to_download <- str_c(path_to_count_matrices, "/", sample_lst, "/count_matrices/count/filtered_feature_bc_matrix")
seurat.lst <- purrr::pmap(
  list(path_to_download, sample_lst),
  function(path, sample){
    count_matrix <- Read10X(path)
    seurat.obj <- CreateSeuratObject(count_matrix[["Gene Expression"]], min.cells=10) %>% 
      RenameCells(add.cell.id = sample)
    seurat.obj[["ref"]] <- "Rejeski_2022"
    seurat.obj[["case_id"]] <- "GSM6069857"
    seurat.obj[["case"]] <- 1
    seurat.obj[["sample_id"]] <- sample
    seurat.obj[["cond"]] <- "RT"
    seurat.obj[["tissue"]] <- "PB"
    seurat.obj <- seurat_quality_control(seurat.obj)
    seurat.obj <- seurat_preprocess(seurat.obj)
    seurat.obj <- doublet_filter(seurat.obj, sct=FALSE)
    return(seurat.obj)
  }
)
names(seurat.lst) <- sample_lst
```

```{r, eval=FALSE}
# Explore objects in the list
seurat.lst
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat.lst, path_to_save_obj_lst)
```

# Merge and integrate the objects

```{r, eval=FALSE}
# Merge the objects
seurat <- seurat.lst[[1]]
for (i in 2:length(seurat.lst)) {
  seurat <- merge(x = seurat, y = seurat.lst[[i]])
  seurat.lst[[i]] <- NA
}
rm(seurat.lst)
```

```{r, eval=FALSE}
# Save the merged object
saveRDS(seurat, path_to_save_merged_obj)
```

We used Harmony package for data integration and batch correction, and annotated the cell types using SingleR package.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Reanalysis after merging and data integration 
DefaultAssay(seurat) <- "RNA"
seurat <- JoinLayers(seurat)
seurat <- SCTransform(seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
seurat <- RunPCA(seurat)
seurat <- RunHarmony(seurat, group.by.vars = "sample_id", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
seurat <- FindNeighbors(seurat, dims = 1:30)              
seurat <- FindClusters(
    seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )
seurat <- singleR_cell_annotation(seurat) 
```

```{r, eval=FALSE}
# Save the integrated object
saveRDS(seurat, path_to_save_integrated_obj)
```

# Visualisation after data integration

```{r}
# Define path
path_to_obj <- path_to_save_integrated_obj

# Load the data
seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=11}
# Plot based on sample_id and cell types
sample_id.plt <- DimPlot(seurat, reduction = "umap", group.by = "sample_id") + 
  labs(y= "UMAP 2", x = "UMAP 1")

cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")

(sample_id.plt + cell.plt) 
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
varied_clusters.plt <- plot_varying_cluster_resolution(seurat)
varied_clusters.plt
```

# Tumour cell extraction

```{r, message=FALSE, warning=FALSE}
# Extract tumour cells
tumour_seurat <- subset(seurat, subset = SCT_snn_res.0.1 == 2)
tumour_seurat <- SCTransform(tumour_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
tumour_seurat <- RunPCA(tumour_seurat)
tumour_seurat <- FindNeighbors(tumour_seurat, dims = 1:30)              
tumour_seurat <- FindClusters(tumour_seurat, resolution=0.25)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
# Plot based on clusters
cluster.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "seurat_clusters") +
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

# Annotate the RT cells

```{r fig.height=5, fig.width=15}
# Visualise co-expression of CD5 and CD23(FCER2)
features <- c("CD5", "FCER2")
feature_plots <- lapply(features, function(gene) {
  FeaturePlot(tumour_seurat, features = gene, reduction = "umap", pt.size = 1) +
    scale_color_viridis_c(option = "magma") +
    labs(y = "UMAP 2", x = "UMAP 1") +
    ggtitle(gene) 
})

# Combine all plots (patchwork automatically combines lists)
wrap_plots(feature_plots, ncol = 3)  
```

From the plots, all B cell clusters co-expressed in both CD5 and CD23. These B cells are all CLL cells. Next step we will differentiate tumour cells into CLL and RT cells. We have known that RT cells have higher cell proliferation activity than CLL cells. Therefore, we will calculate cell cycle score to each of cluster.

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=18}
# Consider cell cycle genes based on cell cycle upregulation is a hallmark of Richter transformation
tumour_seurat <- CellCycleScoring(
  tumour_seurat,
  s.features = cc.genes$s.genes,
  g2m.features = cc.genes$g2m.genes,
  set.ident = FALSE
)

cell_cycle.plt <- DimPlot(tumour_seurat, group.by = "Phase", pt.size = 1) + labs(y= "UMAP 2", x = "UMAP 1")

umap_s_score <- FeaturePlot(tumour_seurat, features = "S.Score", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "S Score", y= "UMAP 2", x = "UMAP 1") 
  
umap_g2m_score <- FeaturePlot(tumour_seurat, features = "G2M.Score", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "G2M Score", y= "UMAP 2", x = "UMAP 1") 

cell_cycle.plt + umap_s_score + umap_g2m_score
```

From the cell cycle scoring, we suspect that all clusters could be RT cells due to moderate-to-high cell cycle scores in G2M and S phases. Moreover, we will use our identified biomarkers resulted from our systematic review to confirm if the cluster contains RT cells. We have sets of up- and downregulated DEGs with rank-consistence between RT and CLL conditions. We will show their feature plots.

```{r, message=FALSE, warning=FALSE, fig.height=14, fig.width=17}
# Visualise downregulated genes that we found from the systematic review
features <- c("HMGA1", "VDAC1", "PPA1", "LDHA", "GGCT", "HSPD1", "IMPDH2", "PAICS", "NME1")
feature_plots <- lapply(features, function(gene) {
  FeaturePlot(tumour_seurat, features = gene, reduction = "umap", pt.size = 1) +
    scale_color_viridis_c(option = "magma") +
    labs(y = "UMAP 2", x = "UMAP 1") +
    ggtitle(gene) 
})

# Combine all plots (patchwork automatically combines lists)
wrap_plots(feature_plots, ncol = 3)  
```

We also visualised upregulated genes obtained from the intersection between common DEGs and RS-rank consistent genes.

```{r, message=FALSE, warning=FALSE, fig.height=14, fig.width=17}
# Visualise downregulated genes that we found from the systematic review
features <- c("COX5A", "SELENOH", "PRDX5", "ATP5PD", "NPM1", "NME2", "HSPA8", "MIF", "STMN1")
feature_plots <- lapply(features, function(gene) {
  FeaturePlot(tumour_seurat, features = gene, reduction = "umap", pt.size = 1) +
    scale_color_viridis_c(option = "magma") +
    labs(y = "UMAP 2", x = "UMAP 1") +
    ggtitle(gene) 
})

# Combine all plots (patchwork automatically combines lists)
wrap_plots(feature_plots, ncol = 3)  
```

In downregulated genes, we visualised from both RP- and RP-rank consistent genes overlapping with common downregulated DEGs.

```{r, message=FALSE, warning=FALSE, fig.height=14, fig.width=17}
# Visualise downregulated genes that we found from the systematic review
features <- c("PNRC1", "OTUD1", "IGKC", "RPL21", "SF1", "VAMP2", "MDM4", "SRRM2", "SRSF7")
feature_plots <- lapply(features, function(gene) {
  FeaturePlot(tumour_seurat, features = gene, reduction = "umap", pt.size = 1) +
    scale_color_viridis_c(option = "magma") +
    labs(y = "UMAP 2", x = "UMAP 1") +
    ggtitle(gene) 
})

# Combine all plots (patchwork automatically combines lists)
wrap_plots(feature_plots, ncol = 3)  
```

```{r, eval=FALSE}
# Save the tumour object
saveRDS(tumour_seurat, path_to_save_tumour_obj)
```

# Session Information

```{r}
sessionInfo()
```

