---
title: "Overlapping_genes_seurat"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

After we carried out psedobulk gene expression analysis and GSEA for several RT scRNA-seq datasets, we will evaluate the consistency of gene expression results across those studies with calculated from Pseudobulk analysis using `FindMarkers()` in Seurat. We will explore overlapping genes among those studies.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Import the relavant packages
library(UpSetR)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(lsa)
library(pheatmap)
library(viridis)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save files
path_to_single_cell_analysis <- here::here("scRNAseq/systematic_review")
path_to_save_up_common_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/1.upregulated_common_degs.txt")
path_to_save_down_common_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/1.downregulated_common_degs.txt")
path_to_save_up_consist_rp <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/2.upregulated_rp.rds")
path_to_save_down_consist_rp <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/2.downregulated_rp.rds")
path_to_save_up_consist_rs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/2.upregulated_rs.rds")
path_to_save_down_consist_rs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/2.downregulated_rs.rds")
path_to_save_up_union_rp_rs_consist_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/3.upregulated_union_rp_rs_consist_degs.txt")
path_to_save_down_union_rp_rs_consist_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/3.downregulated_union_rp_rs_consist_degs.txt")
path_to_save_up_intersct_rp_rs_consist_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/3.upregulated_intersct_rp_rs_consist_degs.txt")
path_to_save_down_intersct_rp_rs_consist_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/3.downregulated_intersct_rp_rs_consist_degs.txt")
path_to_save_figure <- here::here("figures")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# List file names of all studies
study.lst <- c("1-Penter_2021_GSE165087", "2-Nadeu_2022", "3-Hing_2023_GSE205704", "4-Parry_2023b", "5-Rigo_2024")
study.lst
```

```{r}
# Create names for list later
names.lst <- sub(
  "^\\d+-([A-Za-z]+)_([0-9]{4}[A-Za-z]?).*$",  # capture name and year
  "\\1(\\2)",                         # output as Name(Year)
  study.lst
)
names.lst
```

```{r}
# Extract only name of studies
study_names <- sub("^\\d+-([^_0-9]+).*", "\\1", study.lst)
study_names
```

# Pseudobulk gene expression using `FindMarkers()` in Seurat

```{r}
# Load expression results in each study and collect in a list
path_to_download <- str_c(path_to_single_cell_analysis, "/", study.lst, "/expression_data/", study_names, "_dea.rds")
dea.lst <- purrr::pmap(
  list(path_to_download), 
  function(path){
    expr.mat <- readRDS(path)
    return(expr.mat)
  }
)
names(dea.lst) <- names.lst
```

## Upregulated genes

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=12}
# Extract adjusted significant upregulated genes
sig_upreg_gene.lst <- purrr::pmap(
  list(dea.lst),
  function(dea){
    selected_gene <- dea[dea$avg_log2FC > 0.25 & dea$p_val_adj < 0.05, "gene"]
    return(selected_gene)
  }
)

# Reorder the list first
sig_upreg_gene.lst <- sig_upreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_upreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 1.7), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r, eval=FALSE}
# Save the figure
png(filename = paste0(path_to_save_figure, "/upset_plt_seurat_up_degs.png"), width=12, height=8, units="in", res=500)
print(upset.plt)
dev.off()
```

```{r}
# Collect upregulated DEGs
five_common_up_degs <- Reduce(intersect, sig_upreg_gene.lst)
five_common_up_degs
```

## Downregulated genes

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=12}
# Extract adjusted significant downregulated genes
sig_downreg_gene.lst <- purrr::pmap(
  list(dea.lst),
  function(dea){
    selected_gene <- dea[dea$avg_log2FC < (-0.25) & dea$p_val_adj < 0.05, "gene"]
    return(selected_gene)
  }
)

# Reorder the list first
sig_downreg_gene.lst <- sig_downreg_gene.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_downreg_gene.lst)

# Plot UpSetR
upset.plt <- upset(
  upset_data,
  sets = colnames(upset_data),
  keep.order = TRUE,
  nintersects = NA,
  point.size = 3.5,
  line.size = 1,
  text.scale = c(2, 2, 2, 2, 2.5, 1.7), # defined as c(intersection size title, intersection size tick labels, set size title, set size tick labels, set names, numbers above bars)
  mb.ratio = c(0.55, 0.45) # Adjust margin: give more space to the bottom/left for those large labels
)

upset.plt
```

```{r, eval=FALSE}
# Save the figure
png(filename = paste0(path_to_save_figure, "/upset_plt_seurat_down_degs.png"), width=12, height=8, units="in", res=500)
print(upset.plt)
dev.off()
```

```{r}
# Collect downregulated DEGs
five_common_down_degs <- Reduce(intersect, sig_downreg_gene.lst)
five_common_down_degs
```

```{r, eval=FALSE}
# Save the common DEG results
write.table(
  five_common_up_degs,
  file = path_to_save_up_common_degs,   
  sep = "\t",        
  quote = FALSE,     
  row.names = FALSE,
  col.names = FALSE
)

write.table(
  five_common_down_degs,
  file = path_to_save_down_common_degs,   
  sep = "\t",        
  quote = FALSE,     
  row.names = FALSE,
  col.names = FALSE
)
```

# Consistency analysis of expression genes

## Identify the consistent direction of expression

We compared the known DEGs between RT versus CLL reported from the previous publications and compared the consistency through the direction between upregulation and downregulation.

```{r}
# Extract all expressed genes without considering significance 
common_genes <- purrr::map(
  dea.lst,
  ~ .x %>%
    pull(gene)
) %>%
  Reduce(intersect, .)

common_gene_expr.df <- purrr::map_dfr(
  names(dea.lst),
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene %in% common_genes) %>%
      dplyr::select(gene, avg_log2FC) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
common_gene_expr.df <- common_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = avg_log2FC
  )

DT::datatable(common_gene_expr.df, options = list(scrollX = TRUE))
```

```{r}
# Define known up- and downregulated DEGs
gene_categories <- list(
  # Upregulated DEGs
  Cell_Cycle_Up = c("MKI67", "PCNA", "CDK1", "CDK2", "CDK4", "CDK5", "CDK6", "CCND2", "CCND3", "CDKN2A", "CDKN3", "CKS1B", "E2F7", 
                    "ANP32B"),
  Mitosis_Up = c("FOXM1", "PLK1", "PLK4", "AURKA", "AURKB", "TUBB", "TUBA1C", "STMN1", "TPX2", "HMMR", "KIF14", "KIF18B", "BUB1", 
                 "UBE2C", "UBE2S", "BIRC5"),
  DNA_Repair_Up = c("POLH", "POLQ", "POLE2", "TERT", "BRCA1", "XRCC2", "AICDA", "TOP2A", "NPM1"),
  Oncogenes_Wnt_Up = c("MYC", "MYCBP", "EZH2", "PRMT5", "WNT3", "WNT5A", "WNT6", "WNT10A", "GAS5"),
  Immune_Surface_Up = c("TLR8", "TLR9", "TLR10", "CD24", "CD70", "LTB", "MS4A1", "BID"),
  
  # Downregulated DEGs
  Apoptosis_Down = c("BAX", "BAK1", "BAD", "BMF", "HRK", "BBC3", "PMAIP1", "BCL2L11"),
  Chromatin_TumorSup_Down = c("CDKN1B", "CREBBP", "EP300", "CHD2", "CHD7", "ARID1A", "ARID4B", "ARID5B", "SPEN", "BAZ2A"),
  B_Identity_MHC_Down = c("PAX5", "CD79A", "CD74", "CD37", "CD83", "CD99", "IGHV3-7", "IGLC7", "HLA-A", "HLA-B"),
  Signaling_Homing_Down = c("CXCR4", "ITGB1", "LYN", "TRAF3", "TRAF5", "NFKBIA", "WNT16", "CDK17"),
  Transcription_Stress_Down = c("FOS", "FOSB", "JUN", "JUNB", "ZFP36", "BTG1", "UBC")
)

known_degs <- unlist(gene_categories)
```

```{r, fig.height=14, fig.width=2.5}
# Evaluate the consistency through the direction of the known DEGs by plotting the heatmap
library(pheatmap)

# Extract expression matrix 
expr_mat <- as.matrix(common_gene_expr.df[ , -1])   # all columns except gene
rownames(expr_mat) <- common_gene_expr.df$gene
valid_genes <- intersect(known_degs, rownames(expr_mat))
known_degs_expr_mat <- expr_mat[valid_genes, ]
colnames(known_degs_expr_mat) <- names.lst

# Set scale for heatmap
palette_length <- 100
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(palette_length)
value_limit <- max(abs(known_degs_expr_mat), na.rm = TRUE)
breaks <- seq(-value_limit, value_limit, length.out = palette_length + 1)

# Plot the heatmap
pheatmap(
  known_degs_expr_mat,
  scale = "none",
  color = colour_pal,
  breaks = breaks,
  cluster_rows = FALSE,
  cluster_cols = FALSE
)
```

For our results, several well-established upregulated DEGs, including PCNA, CDK4, CDK5, CKS1B, TUBA1C, STMN1, NPM1, MYC, PRMT5, WNT3, GAS5, TLR9, and BID, exhibited consistent directionality across all studies. Similarly, known downregulated DEGs, such as CHD7, CD74, NFKBIA, JUNB, and BTG1, showed consistent downregulation across all studies.

We also measured the gene expression consistency across those studies by selecting genes with significant expression to evaluate the correlation of those genes, see the following steps. We measured numerically consistency between the expression level through correlations and similarities.

## All significant genes

```{r}
# Extract significant genes 
common_sig_genes <- purrr::map(
  dea.lst,
  ~ .x %>%
    dplyr::filter(p_val_adj < 0.05) %>%
    pull(gene)
) %>%
  Reduce(intersect, .)

sig_gene_expr.df <- purrr::map_dfr(
  names(dea.lst),
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene %in% common_sig_genes) %>%
      dplyr::select(gene, avg_log2FC) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
sig_gene_expr.df <- sig_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = avg_log2FC
  )

DT::datatable(sig_gene_expr.df, options = list(scrollX = TRUE))
```

```{r}
# Remove the gene column for correlation
expr_mat <- sig_gene_expr.df %>% dplyr::select(-gene) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5, fig.width=5.5}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson Correlation (Significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12         
)

pheatmap(
  spearman_cor,
  main = "Spearman Correlation (Significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (Significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (Significant genes)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

## Upregulated genes

```{r}
# Extract upregulated genes and their log2FC
common_upreg_genes <- purrr::map(
  dea.lst,
  ~ .x %>%
    dplyr::filter(avg_log2FC > 0.25, p_val_adj < 0.05) %>%
    pull(gene)
) %>%
  Reduce(intersect, .)

upreg_gene_expr.df <- purrr::map_dfr(
  names(dea.lst),
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene %in% common_upreg_genes) %>%
      dplyr::select(gene, avg_log2FC) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
upreg_gene_expr.df <- upreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = avg_log2FC
  )

head(upreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- upreg_gene_expr.df %>% dplyr::select(-gene) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5, fig.width=5.5}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (upregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

## Downregulated genes

```{r}
# Extract downregulated genes and their log2FC
common_downreg_genes <- purrr::map(
  dea.lst,
  ~ .x %>%
    dplyr::filter(avg_log2FC < (-0.25), p_val_adj < 0.05) %>%
    pull(gene)
) %>%
  Reduce(intersect, .)

downreg_gene_expr.df <- purrr::map_dfr(
  names(dea.lst),
  function(study_name){
    dea <- dea.lst[[study_name]]
    gene_expr <- dea %>%
      dplyr::filter(gene %in% common_downreg_genes) %>%
      dplyr::select(gene, avg_log2FC) %>%
      dplyr::mutate(study = study_name) 
    return(gene_expr)
  }
)

# Change the data frame to wider form
downreg_gene_expr.df <- downreg_gene_expr.df %>%
  tidyr::pivot_wider(
    names_from = study,
    values_from = avg_log2FC
  )

head(downreg_gene_expr.df)
```

```{r}
# Remove the gene column for correlation
expr_mat <- downreg_gene_expr.df %>% dplyr::select(-gene) %>% as.matrix()

# Compute correlations and cosine similarity
pearson_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "pearson")
spearman_cor <- cor(expr_mat, use = "pairwise.complete.obs", method = "spearman")
cosine_similar <- lsa::cosine(expr_mat)

# Compute Euclidean similarity
euclidean_dist <- as.matrix(dist(t(expr_mat), method = "euclidean"))
euclidean_similar <- 1 / (1 + euclidean_dist)
```

```{r, fig.height=5, fig.width=5.5}
# Plot with viridis colour palette
pheatmap(
  pearson_cor,
  main = "Pearson correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  spearman_cor,
  main = "Spearman correlation (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  cosine_similar,
  main = "Cosine similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)

pheatmap(
  euclidean_similar,
  main = "Euclidean similarity (downregulated DEGs)",
  color = viridis(100),
  breaks = seq(0, 1, length.out = length(viridis(100)) + 1),
  display_numbers = TRUE,
  number_color = "black",
  border_color = NA,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_number = 12,
  fontsize_row = 12,        
  fontsize_col = 12 
)
```

# Evaluate gene-ranked consistency

From the previous step, we got 78 and 10 for common up- and downregulated DEGs from all studies. However, we are not sure if these DEGs are common from ranking by chance. We will evaluate thier rank based on log2FC by using RankProd package. The package evaluates the statistics of gene rank across several gene expression studies by modelling ranked genes as probability distribution for the sum of outcomes by rolling *K* dices with *N* faces, similar to measuring *N* genes in expression analysis across *K* replicates or studies.

```{r, message=FALSE, warning=FALSE}
# Import the relavant packages
library(RankProd)
```

```{r}
# Join dea tables from all studies
common_dea.df <- dea.lst[[1]] %>%
  dplyr::mutate(log2FC_1 = avg_log2FC) %>%
  dplyr::select(gene, log2FC_1)

# Loop over the rest
for (i in 2:length(dea.lst)) {
  
  # Rename dynamically with :=
  df_i <- dea.lst[[i]] %>%
    dplyr::mutate(!!paste0("log2FC_", i) := avg_log2FC) %>% 
    dplyr::select(gene, !!paste0("log2FC_", i))
  
  # Join onto the growing table
  common_dea.df <- full_join(common_dea.df, df_i, by = "gene")
}

# Result: one data frame with gene and log2FC_x columns
head(common_dea.df)
```

## Rank product

```{r, message=FALSE, warning=FALSE}
# Perform Rank Product
expr_mat <- as.matrix(common_dea.df[ , -1])   # all columns except gene
rownames(expr_mat) <- common_dea.df$gene
cl=rep(1, ncol(expr_mat))
RP.out <- RankProducts(expr_mat, cl, na.rm=FALSE, logged=TRUE, rand=123)
```

```{r}
# Summarise results of significant consistent genes
RP_sig_results <- topGene(RP.out, cutoff=0.05, method="pfp", logged=TRUE, logbase=2, gene.names=rownames(expr_mat))
```

We will explore the RP results for both upregulated and downregulated genes.

```{r}
# Upregulated genes table
RP_up_genes.df <- RP_sig_results[["Table2"]]
DT::datatable(RP_up_genes.df, options = list(scrollX = TRUE))
```

```{r}
# Downregulated genes table
RP_down_genes.df <- RP_sig_results[["Table1"]]
DT::datatable(RP_down_genes.df, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save the RP results
saveRDS(RP_up_genes.df, path_to_save_up_consist_rp)
saveRDS(RP_down_genes.df, path_to_save_down_consist_rp)
```

## Rank sum

```{r, message=FALSE, warning=FALSE}
# Perform Rank Sum
RS.out <- RankProducts(expr_mat, cl, na.rm=FALSE, calculateProduct=FALSE, logged=TRUE, rand=123)
```

```{r}
# Summarise results of significant consistent genes
RS_sig_results <- topGene(RS.out, cutoff=0.05, method="pfp", logged=TRUE, logbase=2, gene.names=rownames(expr_mat))
```

We will explore the RP results for both upregulated and downregulated genes.

```{r}
# Upregulated genes table
RS_up_genes.df <- RS_sig_results[["Table2"]]
DT::datatable(RS_up_genes.df, options = list(scrollX = TRUE))
```

```{r}
# Downregulated genes table
RS_down_genes.df <- RS_sig_results[["Table1"]]
DT::datatable(RS_down_genes.df, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save the RP results
saveRDS(RS_up_genes.df, path_to_save_up_consist_rs)
saveRDS(RS_down_genes.df, path_to_save_down_consist_rs)
```

# Find overlapping between common DEGs and significant consistent ranked genes

We have identified common DEGs and significantly consistent ranked genes. We will next identify the common DEGs that are rank-consistent in order to evaluate their expression-rank robustness across all studies.

```{r}
# Collect consistent gene names from both up- and downregulated genes
consist_rp_up_genes <- rownames(RP_sig_results[["Table2"]])
consist_rp_down_genes <- rownames(RP_sig_results[["Table1"]])
consist_rs_up_genes <- rownames(RS_sig_results[["Table2"]])
consist_rs_down_genes <- rownames(RS_sig_results[["Table1"]])
```

After that, we will identify overlappping genes in both DEGs and consistent ranked genes.

## Overlapping in upregulated genes

```{r}
# Overlapping between upregulated common DEGs and upregulated RP-rank consistent genes
up_deg_rp_intersct <- intersect(five_common_up_degs, consist_rp_up_genes)
up_deg_rp_intersct
```

```{r}
# Overlapping between upregulated common DEGs and upregulated RS-rank consistent genes
up_deg_rs_intersct <- intersect(five_common_up_degs, consist_rs_up_genes)
up_deg_rs_intersct
```

We can see that `up_deg_rp_intersct` is a subset in `up_deg_rs_intersct`. We also proved from the union below.

```{r}
# Finding union between up_deg_rp_intersct and up_deg_rs_intersct
union_consist_rank_up_degs <- union(up_deg_rp_intersct, up_deg_rs_intersct)
union_consist_rank_up_degs
```

```{r}
# Finding overlapping upregulated genes between DEGs and all consistent ranking
consist_rank_up_degs <- Reduce(intersect, list(five_common_up_degs, consist_rp_up_genes, consist_rs_up_genes))
consist_rank_up_degs
```

## Overlapping in downregulated genes

```{r}
# Overlapping between upregulated common DEGs and upregulated RP-rank consistent genes
down_deg_rp_intersct <- intersect(five_common_down_degs, consist_rp_down_genes)
down_deg_rp_intersct
```

```{r}
# Overlapping between upregulated common DEGs and upregulated RS-rank consistent genes
down_deg_rs_intersct <- intersect(five_common_down_degs, consist_rs_down_genes)
down_deg_rs_intersct
```

It is clear that `down_deg_rp_intersct` is a subset in `down_deg_rs_intersct`. We also proved from the union below.

```{r}
# Finding union between up_deg_rp_intersct and up_deg_rs_intersct
union_consist_rank_down_degs <- union(down_deg_rp_intersct, down_deg_rs_intersct)
union_consist_rank_down_degs
```

```{r}
# Finding overlapping downregulated genes between DEGs and all consistent ranking
consist_rank_down_degs <- Reduce(intersect, list(five_common_down_degs, consist_rp_down_genes, consist_rs_down_genes))
consist_rank_down_degs
```

We also visualised the normalised expression in common DEGs consistent with RP and RS

```{r, fig.height=10, fig.width=2.5}
# Visualise the heatmap (only intersection between DEGs and all ranking methods: RP and RS)
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

# Extract expression matrix with rank-consistent DEGs
consist_sig_genes <- c(consist_rank_up_degs, consist_rank_down_degs)
consist_sig_expr_mat <- expr_mat[consist_sig_genes, ]
colnames(consist_sig_expr_mat) <- names.lst

# Plot the heatmap
pheatmap(
  consist_sig_expr_mat,
  scale = "none",
  color = colour_pal,
  cluster_rows = FALSE,
  cluster_cols = FALSE
)
```

```{r, fig.height=14, fig.width=2.5}
# Visualise the heatmap (union between all ranking methods: RP and RS)
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

# Extract expression matrix with rank-consistent DEGs
consist_sig_genes <- c(union_consist_rank_up_degs, union_consist_rank_down_degs)
consist_sig_expr_mat <- expr_mat[consist_sig_genes, ]
colnames(consist_sig_expr_mat) <- names.lst

# Plot the heatmap
pheatmap(
  consist_sig_expr_mat,
  scale = "none",
  color = colour_pal,
  cluster_rows = FALSE,
  cluster_cols = FALSE
)
```

We will save these genes for further study and discussion.

```{r, eval=FALSE}
# Save the results
write.table(
  union_consist_rank_up_degs,
  file = path_to_save_up_union_rp_rs_consist_degs,   
  sep = "\t",        
  quote = FALSE,     
  row.names = FALSE,
  col.names = FALSE
)

write.table(
  union_consist_rank_down_degs,
  file = path_to_save_down_union_rp_rs_consist_degs,   
  sep = "\t",        
  quote = FALSE,     
  row.names = FALSE,
  col.names = FALSE
)

write.table(
  consist_rank_up_degs,
  file = path_to_save_up_intersct_rp_rs_consist_degs,   
  sep = "\t",        
  quote = FALSE,     
  row.names = FALSE,
  col.names = FALSE
)

write.table(
  consist_rank_down_degs,
  file = path_to_save_down_intersct_rp_rs_consist_degs,   
  sep = "\t",        
  quote = FALSE,     
  row.names = FALSE,
  col.names = FALSE
)
```

# Session Information

```{r}
sessionInfo()
```
