---
title: "Hing_2023_GSE183426_common_DEGs_validation"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This source code we validated the rank-consistent common DEGs from our systematic review with single-cell RNA-seq for RT- and CLL-like conditions in mouse experiments. We analysed for [Hing et. al. (2023)](https://www.nature.com/articles/s41467-022-35778-1) study which employed a mouse experimental model using Eµ-hTCL1 and Eµ-hPRMT5/hTCL1 mouse model for CLL and RT, respectively.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tidyr)
library(SingleR)
library(celldex)
library(sctransform)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save
path_to_metadata <- here::here("scRNAseq/Hing_2023_GSE183426/sample_metadata.csv")
path_to_obj_lst <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_reanalysed_seu_lst.rds")
path_to_save_merged_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_merged_seu.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_integrated_seu.rds")
path_to_save_tumour_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_tumour_seu.rds")
path_to_save_filtered_tumour_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_filtered_tumour_seu.rds")
path_to_save_markers <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_markers.rds")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the sample metadata
sample_metadata <- read.csv(path_to_metadata)
DT::datatable(sample_metadata, options = list(scrollX = TRUE))
```

```{r}
# Load the seurat list
seurat.lst <- readRDS(path_to_obj_lst)
```

# Data merging and integration

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Merge the objects 
seurat <- seurat.lst[[1]]
for (i in 2:length(seurat.lst)) {
  seurat <- merge(x = seurat, y = seurat.lst[[i]])
}
rm(seurat.lst); gc()

# Save the object
saveRDS(seurat, path_to_save_merged_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Data integration and downstream analysis
DefaultAssay(seurat) <- "RNA"
seurat <- JoinLayers(seurat)
seurat <- SCTransform(seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
seurat <- RunPCA(seurat)
seurat <- RunHarmony(seurat, group.by.vars = "case_id", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
seurat <- FindNeighbors(seurat, dims = 1:30, reduction = "harmony")              
seurat <- FindClusters(
    seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )

# Save the object
saveRDS(seurat, path_to_save_integrated_obj)
```

# Visualisation after integration

```{r}
# Load the data
seurat <- readRDS(path_to_save_integrated_obj)
```

```{r, message=FALSE, warning=FALSE}
# Plot based on case_id
case_id.plt <- DimPlot(seurat, reduction = "umap", group.by = "case_id") + 
  labs(y= "UMAP 2", x = "UMAP 1")
case_id.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=10}
# Plot based on models and tissues
model.plt <- DimPlot(seurat, reduction = "umap", group.by = "model") + 
  labs(y= "UMAP 2", x = "UMAP 1")

tissue.plt <- DimPlot(seurat, reduction = "umap", group.by = "tissue") + 
  labs(y= "UMAP 2", x = "UMAP 1")

model.plt + tissue.plt
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("Hbb-bt", "Hba-a1", "Cd3d", "Cd8a", "Cd4", "Cd19", "Cd79a", "Cd14", "Ncr1"),
            reduction = "umap") 
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
varied_clusters.plt <- plot_varying_cluster_resolution(seurat)
varied_clusters.plt
```

```{r, fig.height=4, fig.width=5}
# Plot seurat clusters at resolution=0.1
clusters.plt <- DimPlot(seurat, 
                        reduction = "umap", 
                        group.by = "SCT_snn_res.0.1", 
                        raster=FALSE, 
                        label = TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

clusters.plt
```

# Tumour cell extraction

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Extract tumour cells and reprocessing after subset
tumour_seurat <- subset(seurat, subset = !(SCT_snn_res.0.1 %in% c(0, 2, 4, 10)))
tumour_seurat <- SCTransform(tumour_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
tumour_seurat <- RunPCA(tumour_seurat)
tumour_seurat <- RunUMAP(tumour_seurat, dims = 1:30, reduction = "harmony")
tumour_seurat <- FindNeighbors(tumour_seurat, dims = 1:30, reduction = "harmony")              
tumour_seurat <- FindClusters(
    tumour_seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )

# Save the object
saveRDS(tumour_seurat, path_to_save_tumour_obj)
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
tumour_seurat <- readRDS(path_to_save_tumour_obj)
varied_clusters.plt <- plot_varying_cluster_resolution(tumour_seurat)
varied_clusters.plt
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(tumour_seurat, 
            features = c("Hbb-bt", "Hba-a1", "Cd3d", "Cd8a", "Cd4", "Cd19", "Cd79a", "Cd14", "Ncr1"),
            reduction = "umap") 
```

```{r, fig.height=8, fig.width=8, warning=FALSE, message=FALSE}
# Calculate subcluster for cluster 5 at res=0.5 as we will remove contaminated cells
Idents(tumour_seurat) <- "SCT_snn_res.0.5"

tumour_seurat <- FindSubCluster(
  object    = tumour_seurat,
  cluster   = 5,                  
  graph.name = "SCT_snn",     
  subcluster.name = "seurat_subclusters",
  resolution = 0.3                 
)

DimPlot(tumour_seurat, group.by = "seurat_subclusters", label = TRUE, repel = TRUE)
```

# Tumour cells QC

We removed other contaminated cells such as erythrocytes and T cells.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Filter out contaminated cells and reprocessing
filtered_tumour_seurat <- subset(tumour_seurat, subset = !(seurat_subclusters %in% c("5_2", "18", "24")))
filtered_tumour_seurat <- SCTransform(filtered_tumour_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered_tumour_seurat <- RunPCA(filtered_tumour_seurat)
filtered_tumour_seurat <- RunUMAP(filtered_tumour_seurat, dims = 1:30, reduction = "harmony")
filtered_tumour_seurat <- FindNeighbors(filtered_tumour_seurat, dims = 1:30, reduction = "harmony")              
filtered_tumour_seurat <- FindClusters(
    filtered_tumour_seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )

# Save the object
saveRDS(filtered_tumour_seurat, path_to_save_filtered_tumour_obj)
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
filtered_tumour_seurat <- readRDS(path_to_save_filtered_tumour_obj)
varied_clusters.plt <- plot_varying_cluster_resolution(filtered_tumour_seurat)
varied_clusters.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
# Plot based on clustering
Idents(filtered_tumour_seurat) <- "SCT_snn_res.0.3"

# Create subclusters for cluster 5
filtered_tumour_seurat <- FindSubCluster(
    object          = filtered_tumour_seurat,
    cluster         = 5,
    graph.name      = "SCT_snn",
    subcluster.name = "seurat_subclusters",
    resolution      = 0.05)

clusters.plt <- DimPlot(filtered_tumour_seurat, 
                        reduction = "umap", 
                        group.by = "seurat_subclusters", 
                        label = TRUE,
                        repel = TRUE,
                        raster=FALSE) + 
  labs(y= "UMAP 2", x = "UMAP 1") 

clusters.plt
```

# RT and CLL cell annotation

## Tumour cell cluster distribution split by the models

```{r, fig.height=5, fig.width=10}
# Plot clusters by spliting the model (Eµ-hPRMT5, Eµ-hTCL1, and Eµ-hPRMT5/hTCL1)
cluster_split.plt <- DimPlot(filtered_tumour_seurat, 
                             reduction = "umap", 
                             group.by = "seurat_subclusters", 
                             split.by = "model", 
                             label = TRUE) +
  NoLegend()
cluster_split.plt 
```

## Biomarker identification

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Thresholds
min_log2FC <- 0.3
alpha <- 0.001

# Calculate markers
Idents(filtered_tumour_seurat) <- "seurat_subclusters"
markers <- FindAllMarkers(filtered_tumour_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_markers)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

## Cell cycle scoring

```{r}
# Convert human cell cycle genes to mouse
human_to_mouse_map <- c(
  # S phase
  MCM5="Mcm5", PCNA="Pcna", TYMS="Tyms", FEN1="Fen1", MCM2="Mcm2", MCM4="Mcm4", RRM1="Rrm1", UNG="Ung", GINS2="Gins2", MCM6="Mcm6", CDCA7="Cdca7", DTL="Dtl", PRIM1="Prim1",
  UHRF1="Uhrf1", MLF1IP="Mlf1ip", HELLS="Hells", RFC2="Rfc2", RPA2="Rpa2", NASP="Nasp", RAD51AP1="Rad51ap1", GMNN="Gmnn", WDR76="Wdr76", SLBP="Slbp", CCNE2="Ccne2", UBR7="Ubr7", 
  POLD3="Pold3", MSH2="Msh2", ATAD2="Atad2", RAD51="Rad51", RRM2="Rrm2", CDC45="Cdc45", CDC6="Cdc6", EXO1="Exo1", TIPIN="Tipin", DSCC1="Dscc1", BLM="Blm", CASP8AP2="Casp8ap2",
  USP1="Usp1", CLSPN="Clspn", POLA1="Pola1", CHAF1B="Chaf1b", BRIP1="Brip1", E2F8="E2f8",

  # G2M phase
  HMGB2="Hmgb2", CDK1="Cdk1", NUSAP1="Nusap1", UBE2C="Ube2c", BIRC5="Birc5", TPX2="Tpx2", TOP2A="Top2a", NDC80="Ndc80", CKS2="Cks2", NUF2="Nuf2", CKS1B="Cks1b", MKI67="Mki67",
  TMPO="Tmpo", CENPF="Cenpf", TACC3="Tacc3", FAM64A="Fam64a", SMC4="Smc4", CCNB2="Ccnb2", CKAP2L="Ckap2l", CKAP2="Ckap2", AURKB="Aurkb", BUB1="Bub1", KIF11="Kif11", ANP32E="Anp32e",
  TUBB4B="Tubb4b", GTSE1="Gtse1", KIF20B="Kif20b", HJURP="Hjurp", CDCA3="Cdca3", HN1="Hn1", CDC20="Cdc20", TTK="Ttk", CDC25C="Cdc25c", KIF2C="Kif2c", RANGAP1="Rangap1", 
  NCAPD2="Ncapd2", DLGAP5="Dlgap5", CDCA2="Cdca2", CDCA8="Cdca8", ECT2="Ect2", KIF23="Kif23", HMMR="Hmmr", AURKA="Aurka", PSRC1="Psrc1", ANLN="Anln", LBR="Lbr", CKAP5="Ckap5", 
  CENPE="Cenpe", CTCF="Ctcf", NEK2="Nek2", G2E3="G2e3", GAS2L3="Gas2l3", CBX5="Cbx5", CENPA="Cenpa"
)

convert_to_mouse <- function(human_genes, map) {
  mouse_genes <- unname(map[human_genes])
  mouse_genes <- mouse_genes[!is.na(mouse_genes)]  # remove unmapped
  return(mouse_genes)
}

mouse_s_genes  <- convert_to_mouse(cc.genes$s.genes,  human_to_mouse_map)
mouse_g2m_genes <- convert_to_mouse(cc.genes$g2m.genes, human_to_mouse_map)
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=18}
# Consider cell cycle genes based on cell cycle upregulation is a hallmark of Richter transformation
filtered_tumour_seurat <- CellCycleScoring(
  filtered_tumour_seurat,
  s.features = mouse_s_genes,
  g2m.features = mouse_g2m_genes,
  set.ident = FALSE
)

cell_cycle.plt <- DimPlot(filtered_tumour_seurat, group.by = "Phase", pt.size = 1) + labs(y= "UMAP 2", x = "UMAP 1")

umap_s_score <- FeaturePlot(filtered_tumour_seurat, features = "S.Score", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "S Score", y= "UMAP 2", x = "UMAP 1") 
  
umap_g2m_score <- FeaturePlot(filtered_tumour_seurat, features = "G2M.Score", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "G2M Score", y= "UMAP 2", x = "UMAP 1") 

cell_cycle.plt + umap_s_score + umap_g2m_score
```

From the cell cycle scoring, we can see that cells in Eµ-hPRMT5/hTCL1 models had high cell cycle scores in G2M and S phases.

## Subtype annotation

We annotated tumour subtypes as literatures below:

-   [Detection of early seeding of Richter transformation in chronic lymphocytic leukemia](https://www.nature.com/articles/s41591-022-01927-8#code-availability)
-   [Transcriptomic clustering of chronic lymphocytic leukemia: molecular subtypes based on Bruton’s tyrosine kinase expression levels](https://www.nature.com/articles/s41408-024-01196-3)
-   [Molecular map of chronic lymphocytic leukemia and its impact on outcome](https://www.nature.com/articles/s41588-022-01140-w)
-   [Subgroup-specific gene expression profiles and mixed epistasis in chronic lymphocytic leukemia](https://haematologica.org/article/view/haematol.2022.281869)
-   [CD69 is independently prognostic in chronic lymphocytic leukemia: a comprehensive clinical and biological profiling study](https://haematologica.org/article/view/6219)
-   [MYC target gene activation in chronic lymphocytic leukemia and richter transformation: links to aggressiveness and tumor microenvironment interactions](https://www.frontiersin.org/journals/pharmacology/articles/10.3389/fphar.2025.1642458/full)
-   [P594: THE CXCR4-LOW/CD5-HIGH PROLIFERATIVE FRACTION IS ENRICHED IN BTK MUTATIONS AND ANTICIPATES RELAPSE IN IBRUTINIB-TREATED CLL](https://journals.lww.com/hemasphere/fulltext/2023/08003/p594__the_cxcr4_low_cd5_high_proliferative.494.aspx)
-   [The transcription factor Foxo1 controls germinal center B cell proliferation in response to T cell help](https://rupress.org/jem/article/214/4/1181/42216/The-transcription-factor-Foxo1-controls-germinal)
-   [KLF2– A Negative Regulator of Pre-B Cell Clonal Expansion and B Cell Activation](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0097953)

| Cluster | Markers                         | Annotation                   |
|---------|---------------------------------|------------------------------|
| 0       | Osbpl5, Hck                     | M-CLL                        |
| 1       | Myc, Fcer2a, Cd69, Slamf1, Cd38 | U-CLL                        |
| 2       | Ighm, Mzb1, Xbp1                | MZB1hiIGHMhiXBP1hi CLL       |
| 3       | Btk                             | Btkhi CLL                    |
| 4       | Cxcr4, Mir155hg                 | Cxcr4hiCd27lowMir155hghi CLL |
| 5_0     | Cd3d, Cd3e, Cd8a                | contaminated CD8+ T          |
| 5_1     | Bcat1, Pcna, Mki67              | Bcat1hi RT                   |
| 6       | Top2a, Pcna, Mki67              | proliferative RT             |
| 7       | Ccnd2                           | Ccnd2hi RT                   |
| 8       | Slc5a5, Gnat3, Trpm5            | contaminated epithelial      |
| 9       | Btk, Pcna, Pik3ip1              | Btkhi RT                     |
| 10      | Btla, Camk2d                    | exhausted CLL                |
| 11      | Cd5                             | Cxcr4loCd5hi CLL             |
| 12      | Cd3e, Cd3g, Cd8b1               | contaminated CD8+ T          |
| 13      | Cd5, Cd38, Pdcd1                | aggressive CLL               |
| 14      | Cd4, Il7r                       | contaminated CD4+ T          |
| 15      | Ifit1bl1, Isg15, Hif1a          | stimulated RT                |
| 16      | Cd79a, Pou2af1, Ifit1           | IFN-I–responsive B           |
| 17      | Cd27                            | Cxcr4loCd27hi CLL            |
| 18      | Eno1, Klf2                      | quiescent RT                 |
| 19      | Bach2, Ccr6, Ccr7               | memory B                     |
| 20      | Cxcr4, Foxo1, Cdkn2c            | quiescent CLL                |
| 21      | Mx1, Ly6d, Crtam                | APC-like B                   |
| 22      | Gzma, Gzmk, Nkg7                | contaminated NK              |

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=12}
# Plot UMAP for tumour subtypes

# -------------------------------------------------------------------
# 1. Map Seurat clusters and create the annotation column
# -------------------------------------------------------------------
cluster_to_annotation <- c(
  "0" = "M-CLL", "1" = "U-CLL", "2" = "MZB1hiIGHMhiXBP1hi CLL", "3" = "Btkhi CLL", "4" = "Cxcr4hiCd27lowMir155hghi CLL",
  "5_0" = "contaminated CD8+ T (1)", "5_1" = "Bcat1hi RT", "6" = "proliferative RT", "7" = "Ccnd2hi RT", 
  "8" = "contaminated epithelial", "9" = "Btkhi RT", "10" = "exhausted CLL", "11" = "Cxcr4loCd5hi CLL", 
  "12" = "contaminated CD8+ T (2)", "13" = "aggressive CLL", "14" = "contaminated CD4+ T", "15" = "stimulated RT", 
  "16" = "IFN-I–responsive B", "17" = "Cxcr4loCd27hi CLL", "18" = "quiescent RT", "19" = "memory B", "20" = "quiescent CLL", 
  "21" = "APC-like B", "22" = "contaminated NK"
)

filtered_tumour_seurat$tumour_annotate <- 
  as.vector(cluster_to_annotation[as.character(filtered_tumour_seurat$seurat_subclusters)])

# -------------------------------------------------------------------
# 2. Set level order
# -------------------------------------------------------------------
new_levels <- c(
  "M-CLL", "U-CLL", "MZB1hiIGHMhiXBP1hi CLL", "Btkhi CLL", "Cxcr4hiCd27lowMir155hghi CLL", "contaminated CD8+ T (1)", "Bcat1hi RT",   "proliferative RT", "Ccnd2hi RT", "contaminated epithelial", "Btkhi RT", "exhausted CLL", "Cxcr4loCd5hi CLL", 
  "contaminated CD8+ T (2)", "aggressive CLL", "contaminated CD4+ T", "stimulated RT", "IFN-I–responsive B", "Cxcr4loCd27hi CLL", 
  "quiescent RT", "memory B", "quiescent CLL", "APC-like B", "contaminated NK"
)

filtered_tumour_seurat$tumour_annotate <- factor(
  filtered_tumour_seurat$tumour_annotate,
  levels = new_levels
)

# -------------------------------------------------------------------
# 3. Define colours and labels
# -------------------------------------------------------------------
new_colours <- c(
  "#349043", "#80BF61", "#AADA79", "#7BB379", "#66A853", "#C3C3C3", "#FF6800", "#ED564E", "#E25644", "#C2C2C2", "#d65c7a", 
  "#4A8C1C", "#C6E297", "#C3C3C3", "#53C100", "#B1B1B1", "#BB777A", "#6FAA66", "#3A7BAA", "#93CD6A", "#EC6857", "#7AC7E2", 
  "#80BF61", "#3F8BBA", "#A1A1A1"
)

custom_labels <- list(
  bquote("M-CLL"), bquote("U-CLL"), bquote(MZB1^hi~IGHM^hi~XBP1^hi~"CLL"), bquote(Btk^hi~"CLL"), 
  bquote(Cxcr4^hi~Cd27^lo~Mir155hg^hi~"CLL"), bquote("contaminated CD8+ T"), bquote(Bcat1^hi~"RT"), bquote("proliferative RT"),
  bquote(Ccnd2^hi~"RT"), bquote("contaminated epithelial"), bquote(Btk^hi~"RT"), bquote("exhausted CLL"), 
  bquote(Cxcr4^lo~Cd5^hi~"CLL"), bquote("contaminated CD8+ T"), bquote("aggressive CLL"), bquote("contaminated CD4+ T"), 
  bquote("stimulated RT"), bquote("IFN-I–responsive B"), bquote(Cxcr4^lo~Cd27^hi~"CLL"), bquote("quiescent RT"), 
  bquote("memory B"), bquote("quiescent CLL"), bquote("APC-like B"), bquote("contaminated NK")
)

# -------------------------------------------------------------------
# 4. Plot UMAP
# -------------------------------------------------------------------
tumour_subtype.plt <- DimPlot(filtered_tumour_seurat, 
                              group.by = "tumour_annotate", 
                              label = TRUE,
                              repel = TRUE,
                              pt.size = 0.5) +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels,
    labels = custom_labels
  ) +
  labs(y = "UMAP 2", x = "UMAP 1") +
  ggtitle("Tumour subtypes")

tumour_subtype.plt
```

# Validation for identified common DEGs

Moreover, we will use our identified biomarkers resulted from our systematic review to confirm if the cluster contains RT cells. We have sets of up- and downregulated DEGs with RP- and RS-rank consistence between RT and CLL conditions. We will show their feature plots.

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=25}
# Dot plot for common DEGs based on annotated cells

path_to_common_up_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/3.upregulated_union_rp_rs_consist_degs.txt")
path_to_common_down_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/3.downregulated_union_rp_rs_consist_degs.txt")

up_genes <- read.table(path_to_common_up_degs, stringsAsFactors = FALSE, header = FALSE)[, 1]
down_genes <- read.table(path_to_common_down_degs, stringsAsFactors = FALSE, header = FALSE)[, 1]
features <- c(up_genes, down_genes)
mouse_genes <- paste0(toupper(substr(features, 1, 1)),
                      tolower(substr(features, 2, nchar(features))))

colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

common_degs.plt <- DotPlot(filtered_tumour_seurat, features=mouse_genes, group.by = "tumour_annotate") +
  RotatedAxis() +
  scale_colour_gradientn(colours = colour_pal) +
  scale_y_discrete(
      limits = rev(new_levels),
      labels = rev(custom_labels)
    ) +
  labs(
    colour = "Average\nExpression",
    size = "Percent\nExpressed"
  )

common_degs.plt
```

We can see from the plot that our identified upregulated RT DEGs were consistent mostly with proliferative, Bcat1hi, and Btkhi RT subtypes and partially with Ccnd2hi, stimulated, and quiescent RT subtypes. While downregulated RT DEGs were consistent mostly in both RT subtypes. Interestingly, Igkc was downregulated in RT and aggressive CLL cells.

```{r, message=FALSE, warning=FALSE, fig.height=15, fig.width=20}
# Visualise upregulated genes that we found from the systematic review
mouse_features <- c("Hmga1", "Vdac1", "Ppa1", "Ldha", "Ggct", "Hspd1", "Impdh2", "Paics", "Nme1", "Nme2", "Cox5a", "Selenoh", 
  "Clta", "Prdx5", "Npm1", "Hspa8", "Mif", "Stmn1", "Cacybp", "Mrpl4")
feature_plots <- lapply(mouse_features, function(gene) {
  FeaturePlot(filtered_tumour_seurat, features = gene, reduction = "umap", pt.size = 1) +
    scale_color_viridis_c(option = "magma", limits=c(0, 3)) +
    labs(y = "UMAP 2", x = "UMAP 1") +
    ggtitle(gene) 
})

# Combine all plots (patchwork automatically combines lists)
degs_umap.plt <- wrap_plots(feature_plots, ncol = 5)  
degs_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=12, fig.width=15}
# Visualise downregulated genes that we found from the systematic review
mouse_features <- c("Pnrc1", "Otud1", "Igkc", "Rpl21", "Sf1", "Vamp2", "Mdm4", "Srrm2", "Srsf7")
feature_plots <- lapply(mouse_features, function(gene) {
  FeaturePlot(filtered_tumour_seurat, features = gene, reduction = "umap", pt.size = 1) +
    scale_color_viridis_c(option = "magma", limits=c(0, 3)) +
    labs(y = "UMAP 2", x = "UMAP 1") +
    ggtitle(gene) 
})

# Combine all plots (patchwork automatically combines lists)
degs_umap.plt <- wrap_plots(feature_plots, ncol = 3)  
degs_umap.plt
```

# Session Information

```{r}
sessionInfo()
```
