---
title: "Rigo_2024_systematic_review"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

This source code we analysed single-cell RNA-seq for RT and CLL conditions both gene expression and functional enrichment analysis before we will compare the results to other studies for our systematic review. We analysed for [Rigo et. al. (2024)](https://www.nature.com/articles/s41419-024-06707-5) study which performed single-cell analysis between CLL and RT-like conditions in one patient. RT-like condition is defined by large B cells detection in peripheral blood with B symptoms and elevated serum LDH after ibrutinib interruption.

## Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(DoubletFinder)
library(SingleR)
library(celldex)
library(sctransform)
```

```{r}
# Check root path
here::here()
```

## Define paths

```{r}
# Define paths to load and save 
path_to_count_matrices <- here::here("scRNAseq/Rigo_2024_scRNAseq/count_matrices")
path_to_save_obj_lst <- here::here("scRNAseq/systematic_review/5-Rigo_2024/R_objects/Rigo_seu_lst.rds")
path_to_save_merged_obj <- here::here("scRNAseq/systematic_review/5-Rigo_2024/R_objects/Rigo_merged_seu.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/systematic_review/5-Rigo_2024/R_objects/Rigo_integrated_seu.rds")
path_to_save_tumour_obj <- here::here("scRNAseq/systematic_review/5-Rigo_2024/R_objects/Rigo_tumour_seu.rds")
path_to_save_filtered_tumour_obj <- here::here("scRNAseq/systematic_review/5-Rigo_2024/R_objects/Rigo_filtered_tumour_seu.rds")
path_to_save_dea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/expression_data/Rigo_dea.rds")
path_to_save_hm_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/enrichment_data/Rigo_hm_gsea.rds")
path_to_save_go_gsea <- here::here("scRNAseq/systematic_review/5-Rigo_2024/enrichment_data/Rigo_go_gsea.rds")
```

## Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Explore directories in count matrix path
lst_files <- list.files(path_to_count_matrices)
lst_files
```

For Rigo et al. (2024), they performed single-cell analysis for case 14. We will import count matrices and convert them to Seurat objects with quality control and pre-processing.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create the objects and collect in a list
path_to_download <- str_c(path_to_count_matrices, "/", lst_files, "/outs/filtered_feature_bc_matrix/")
conditions <- sub(".*_", "", lst_files) # Replace everything before underscore to "" (empty string)
seurat.lst <- purrr::pmap(
  list(path_to_download, lst_files, conditions), 
  function(path, case_id, cond){
    count_matrix <- Read10X(path)
    seurat.obj <- CreateSeuratObject(count_matrix, min.cells=10) %>% 
      RenameCells(add.cell.id = case_id)
    seurat.obj[["ref"]] <- "Rigo_2024"
    seurat.obj[["case_id"]] <- case_id
    seurat.obj[["case"]] <- "14"
    seurat.obj[["cond"]] <- cond
    seurat.obj[["tissue"]] <- "PB"
    seurat.obj <- seurat_quality_control(seurat.obj)
    seurat.obj <- seurat_preprocess(seurat.obj)
    seurat.obj <- doublet_filter(seurat.obj, sct=FALSE)
    return(seurat.obj)
    }
  )
names(seurat.lst) <- lst_files
```

```{r, eval=FALSE}
# Explore data structures of Seurat objects in the list
seurat.lst
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat.lst, path_to_save_obj_lst)
```

## Merge and integrate the objects

```{r, eval=FALSE}
# Merge the objects
seurat <- seurat.lst[[1]]
for (i in 2:length(seurat.lst)) {
  seurat <- merge(x = seurat, y = seurat.lst[[i]])
  seurat.lst[[i]] <- NA
}
rm(seurat.lst)
```

```{r, eval=FALSE}
# Save the merged object
saveRDS(seurat, path_to_save_merged_obj)
```

We used Harmony package for data integration and batch correction, and annotated the cell types using SingleR package.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Reanalysis after merging and data integration 
DefaultAssay(seurat) <- "RNA"
seurat <- JoinLayers(seurat)
seurat <- SCTransform(seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
seurat <- RunPCA(seurat)
seurat <- RunHarmony(seurat, group.by.vars = "case_id", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
seurat <- FindNeighbors(seurat, dims = 1:30)              
seurat <- FindClusters(
    seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )
seurat <- singleR_cell_annotation(seurat) 
```

## Save the integrated object

```{r, eval=FALSE}
saveRDS(seurat, path_to_save_integrated_obj)
```

## Visualisation after data integration

```{r}
# Define path
path_to_obj <- path_to_save_integrated_obj

# Load the data
seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=10}
# Plot based on case_id, conditions, cell types, and clusters
case_id.plt <- DimPlot(seurat, reduction = "umap", group.by = "case_id") + 
  labs(y= "UMAP 2", x = "UMAP 1")

case.plt <- DimPlot(seurat, reduction = "umap", group.by = "case") + 
  labs(y= "UMAP 2", x = "UMAP 1")

ref.plt <- DimPlot(seurat, reduction = "umap", group.by = "ref") + 
  labs(y= "UMAP 2", x = "UMAP 1")

cond.plt <- DimPlot(seurat, reduction = "umap", group.by = "cond") + 
  labs(y= "UMAP 2", x = "UMAP 1")

tissue.plt <- DimPlot(seurat, reduction = "umap", group.by = "tissue") +
  labs(y= "UMAP 2", x = "UMAP 1")

cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")

(case_id.plt + case.plt) / (ref.plt + cond.plt) / (tissue.plt + cell.plt)
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
varied_clusters.plt <- plot_varying_cluster_resolution(seurat)
varied_clusters.plt
```

```{r, fig.height=4, fig.width=5}
# Plot seurat clusters at resolution=0.1
clusters.plt <- DimPlot(seurat, reduction = "umap", group.by = "SCT_snn_res.0.1", raster=FALSE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

clusters.plt
```

## Tumour cell extraction

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Extract tumour cells 
tumour_seurat <- subset(seurat, subset = singleR.labels_main == "B cells")
tumour_seurat <- SCTransform(tumour_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
tumour_seurat <- RunPCA(tumour_seurat)
tumour_seurat <- RunHarmony(tumour_seurat, group.by.vars = "case_id", dims = 1:30) 
tumour_seurat <- RunUMAP(tumour_seurat, dims = 1:30, reduction = "harmony")
tumour_seurat <- FindNeighbors(tumour_seurat, dims = 1:30)              
tumour_seurat <- FindClusters(tumour_seurat, resolution=0.25)
```

```{r, eval=FALSE}
# Save the tumour cells
saveRDS(tumour_seurat, path_to_save_tumour_obj)
```

```{r}
# Define path
path_to_obj <- path_to_save_tumour_obj

# Load the data
tumour_seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=9.5}
# Plot based on sample_id, conditions, tissue, and cell types
case_id.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "case_id") +
  labs(y= "UMAP 2", x = "UMAP 1")

cond.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "cond") + 
  labs(y= "UMAP 2", x = "UMAP 1")

tissue.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "tissue") +
  labs(y= "UMAP 2", x = "UMAP 1")

clusters.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "seurat_clusters") +
  labs(y= "UMAP 2", x = "UMAP 1")

(case_id.plt + cond.plt) / (tissue.plt + clusters.plt)
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(tumour_seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

As there are some T and NK cells (cluster 3) and erythrocytes, we will remove this cluster before psedobulk analysis.

```{r, fig.height=4, fig.width=5}
# Filter out T and NK cells
filtered_tumour_seurat <- subset(tumour_seurat, subset = seurat_clusters != 3) 

# Visualise clusters
DimPlot(filtered_tumour_seurat, reduction = "umap", group.by = "seurat_clusters") +
  labs(y= "UMAP 2", x = "UMAP 1")
```

```{r, fig.height=4, fig.width=5}
# Filter out erythrocytes by manually removal
#noise_cells <- CellSelector(DimPlot(filtered_tumour_seurat))
noise_cells <- c(
  "case14_CLL_AAATGGAAGTCGGGAT-1", "case14_CLL_ACTACGATCAACACCA-1", "case14_CLL_AGTACCAGTGCTTCAA-1",
  "case14_CLL_ATTCCTAAGGAACGCT-1", "case14_CLL_CATTCTATCTCTGGTC-1", "case14_CLL_CGACAGCGTGTACAGG-1",
  "case14_CLL_CGTGCTTGTCACTCTC-1", "case14_CLL_GAGAAATCAGTCTGGC-1", "case14_CLL_GCAGTTATCATAGGCT-1",
  "case14_CLL_GGCGTCAGTGGAGGTT-1", "case14_CLL_TAAGCGTGTAGACTGG-1", "case14_CLL_TCGACCTAGAGTTGAT-1",
  "case14_CLL_TCGCACTAGTTGAATG-1", "case14_CLL_TGATCAGTCATTGTGG-1", "case14_CLL_TGCTGAATCTTTGCGC-1",
  "case14_CLL_TTGCATTAGAGCAAGA-1", "case14_RT-like_AACAAAGCACAACGAG-1", "case14_RT-like_CTGGTCTGTACATTGC-1",
  "case14_RT-like_GATGAGGAGTCACTGT-1", "case14_RT-like_GCCTGTTGTGGAGGTT-1", "case14_RT-like_GGGATGATCGCATTAG-1",
  "case14_RT-like_GTGAGTTCAGATGCGA-1", "case14_RT-like_TAATTCCCATCCGTGG-1", "case14_RT-like_TCAAGACAGGTCCAGA-1",
  "case14_RT-like_TCACATTGTAAGGAGA-1", "case14_RT-like_TCTGGCTGTCGAGCTC-1", "case14_RT-like_TGCAGGCGTGCTGCAC-1",
  "case14_RT-like_TGGAGAGGTTTGTTCT-1", "case14_RT-like_TTTGTTGTCATAGAGA-1"
)
filtered_tumour_seurat <- subset(filtered_tumour_seurat, cell = setdiff(colnames(filtered_tumour_seurat), noise_cells))

# Visualise clusters
DimPlot(filtered_tumour_seurat, reduction = "umap", group.by = "seurat_clusters") +
  labs(y= "UMAP 2", x = "UMAP 1")
```

```{r, eval=FALSE}
# Save the filtered tumour object
saveRDS(filtered_tumour_seurat, path_to_save_filtered_tumour_obj)
```

## Pseudobulk gene expression analysis and GSEA

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

### Find biomarkers of B cells between RT and CLL conditions

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Find markers between CLL and RT-like conditions  
Idents(filtered_tumour_seurat) <- "cond" # change the default identity
tumour_seurat <- PrepSCTFindMarkers(filtered_tumour_seurat) # Prepare SCT assay for DE testing
tumour.ge <- FindMarkers(
  filtered_tumour_seurat,
  ident.1 = "RT-like",
  ident.2 = "CLL",
  test.use = "wilcox",
  only.pos = FALSE,
  logfc.threshold = 0
)
dea.df <- cbind(gene = rownames(tumour.ge), tumour.ge)
```

```{r, eval=FALSE}
# Save gene expression data
saveRDS(dea.df, path_to_save_dea)
```

### Gene Set Enrichment Analysis (GSEA)

```{r}
# Load the saved expression data
path_to_dea <- path_to_save_dea
dea.df <- readRDS(path_to_dea)
```

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea.df$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

#### Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will explore the trends of other hallmark terms reported from other studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as E2F targets, Myc targets, and oxidative phosphorylation (OXPHOS) were upregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

#### GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will explore the trends of other GO:BP terms reported from other studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as OXPHOS and mitochondrial translation were upregulated while BCR signalling was downregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

```{r, eval=FALSE}
# Save gene expression and GSEA data
saveRDS(HM_gsea, path_to_save_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_go_gsea)
```

As this study perform single-cell analysis on only one pateint, we cannot perform pseudobulk analysis with DESeq2 that aggregate read counts based on cases.  




