---
title: "Parry_2023a_bulkRNA-seq_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

In this code we will analyse bulk RNA-seq between Richter's transformation (RT) and chronic lymphocytic leukaemia (CLL) patients. The data comes from [Parry et al. (2023a)](https://www.nature.com/articles/s41591-022-02113-6). After we performed Salmon to convert FASTQ files to pseudocount data in each sample, in this code we will convert them to a count matrix before we use the matrix for DESeq2 for further gene expression analysis.

## Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Import the relavant packages
library(tximport)
library(ensembldb)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(DESeq2)
```

```{r}
# Check root path
here::here()
```

## Define paths

```{r}
# Define paths to load and save files
path_to_samples <- here::here("RNAseq/Parry_2023a")
path_to_sample_metadata <- here::here("RNAseq/systematic_review/2-Parry_2023a/sample_metadata.csv")
path_to_ref_transcriptome <- "/home/pakorns/hc-storage/fastq_files/RNAseq/transcriptome_ref"
path_to_save_txi <- here::here("RNAseq/systematic_review/2-Parry_2023a/count_matrices/Parry_txi.rds")
path_to_save_dea <- here::here("RNAseq/systematic_review/2-Parry_2023a/expression_data/Parry_dea.rds")
path_to_save_hm_gsea <- here::here("RNAseq/systematic_review/2-Parry_2023a/enrichment_data/Parry_hm_gsea.rds")
path_to_save_go_gsea <- here::here("RNAseq/systematic_review/2-Parry_2023a/enrichment_data/Parry_go_gsea.rds")
```

## Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r, message=FALSE, warning=FALSE}
# Load sample metadata
sample_metadata <- read.csv(path_to_sample_metadata) 
DT::datatable(sample_metadata, options = list(scrollX = TRUE))
```

```{r}
# Create a vector collecting Salmon quant.sf path in each sample
path_to_download <- str_c(path_to_samples, "/", sample_metadata$Run, "/quant.sf")
names(path_to_download) <- sample_metadata$Run
path_to_download
```

## Convert salmon pseudocounts to an expression matrix

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Load the ensembl hg38 gtf file
path_to_hg38_transcriptome <- str_c(path_to_ref_transcriptome, "/Homo_sapiens.GRCh38.115.gtf.gz")
edb <- ensDbFromGtf(path_to_hg38_transcriptome) # Return edb = "./Homo_sapiens.GRCh38.115.sqlite"
```

```{r, message=FALSE, warning=FALSE}
# Make the transcript-to-gene map (tx2gene)
edb <- "./Homo_sapiens.GRCh38.115.sqlite"
edb_obj <- EnsDb(edb)
tx2gene <- transcripts(edb_obj,
                       columns = c("tx_id","gene_id","gene_name"),
                       return.type="DataFrame")

# convert to a plain data.frame for tximport
tx2gene <- as.data.frame(tx2gene)
head(tx2gene)
```

```{r, message=FALSE, warning=FALSE}
# Load pseudocount files
txi <- tximport(path_to_download, 
                type = "salmon", 
                tx2gene = tx2gene, 
                ignoreTxVersion = TRUE 
               )

# This gives you txi$counts, txi$abundance (TPM), txi$length
head(txi$counts)
```

```{r, eval=FALSE}
# Save the expression file
saveRDS(txi, path_to_save_txi)
```

## Transcriptomics analysis using DESeq2

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(DESeq2)
library(limma)
```

### Create a DESeq2 object

```{r}
# Create metadata for DESeq2 object
colData <- sample_metadata[, c("Run", "histological_type", "case")] %>%
  rename(
    samples = "Run",
    condition = "histological_type"
  ) %>%
  mutate(
    batch = ifelse(case == "7", 1, 
                   ifelse(case == "19", 2, 
                          ifelse(case == "24", 3, 
                                 ifelse(case == "27", 4, 5))))
  )

rownames(colData) <- colData$samples
colData$condition <- factor(colData$condition)
colData$batch <- factor(colData$batch)
colData
```

```{r, message=FALSE, warning=FALSE}
# Create the object
dds <- DESeqDataSetFromTximport(txi, colData, ~ batch + condition)
```

```{r, message=FALSE, warning=FALSE}
# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
  
# Run DESeq2
dds$condition <- relevel(dds$condition, ref = "CLL")
dds <- DESeq(dds)
```

### Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6}
# Variance stabilised transformation
vsdata <- vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
vs_mat_corrected <- removeBatchEffect(vs_mat, batch = colData(dds)$batch)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     sample = rownames(colData(dds)))

# Plot
ggplot(pca_df, aes(PC1, PC2, color = condition)) +
  geom_point(size = 3) +
  theme_bw()
```

After batch correction using `limma()`, we can see that the samples were grouped based on the conditions rather than the case origin.

### Differential gene expression analysis

```{r, message=FALSE, warning=FALSE}
# Generate a table for gene expression
res <- lfcShrink(dds, 
                 contrast = c("condition", "RS", "CLL"), 
                 type = 'normal')
head(res)
```

```{r}
# Add gene names to a column in expression result
dea.df <- res
gene_map <- unique(tx2gene[, c("gene_id", "gene_name")])
dea.df$gene_id <- rownames(dea.df)
dea.df$gene_name <- gene_map$gene_name[match(rownames(dea.df), gene_map$gene_id)]
head(dea.df)
```

```{r, eval=FALSE}
# Save the expression result
saveRDS(dea.df, path_to_save_dea)
```

### Plot heatmap of gene expression

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(pheatmap)
library(RColorBrewer)
```

```{r}
# Data transformation
vsd <- vst(dds, blind=FALSE)       # Variance-stabilizing transformation
#rld <- rlog(dds_filtered, blind=FALSE)      # Regularized log transformation

# Extract transformed expression matrix
expr_mat <- assay(vsd)

# Calculate z-scores
zscore_mat <- t(scale(t(expr_mat)))
```

```{r}
# Change column names of the matrix
new_col_names <- str_c(
  colData$case[match(colnames(zscore_mat), colData$samples)], 
  "_", 
  colData$condition[match(colnames(zscore_mat), colData$samples)]
  ) 
colnames(zscore_mat) <- new_col_names
```

```{r}
# Extract main and sub- clusters info from column names
conditions <-  ifelse(grepl("RS", colnames(zscore_mat)), "RT", "CLL")
patient_id <- sub("_.*", "", colnames(zscore_mat))

# Create annotation dataframe
annotation_col <- data.frame(
  Condition = factor(conditions, levels = c("CLL", "RT")),
  Case = factor(patient_id, levels = c("42","27","19","7","24"))
)
annotation_col <- annotation_col[, c("Case", "Condition")] 
rownames(annotation_col) <- colnames(zscore_mat)

# Order by Condition and then Case 
order_cols <- order(annotation_col$Condition, annotation_col$Case)
zscore_mat_ordered <- zscore_mat[, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]
```

```{r, fig.height=8, fig.width=4}
# Filter for significant DE genes
sig_genes <- dea.df[!is.na(dea.df$padj) & dea.df$padj < 0.05, ]

# Order genes by log2FoldChange (descending: up -> down)
gene_order <- sig_genes$gene_id[order(sig_genes$log2FoldChange, decreasing = TRUE)]

# Subset zscore matrix
zscore_mat_sig <- zscore_mat_ordered[gene_order, ]

# Plot heatmap
pheatmap(
  zscore_mat_sig,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  gaps_col = which(diff(as.numeric(factor(annotation_col_ordered$Condition))) != 0)
)
```

### Volcanol plot

```{r, warning=FALSE, fig.height=6, fig.width=5}
# Make a new column for plotting
dea.df$neglog10padj <- -log10(dea.df$padj)

# Volcano plot
genes_of_interest <- c("PAX5", "CD79A", "CD74", "ITGB1", "CD37", "LYN", "POLQ", "AURKA", "XRCC2", "KIF18B", "PLK1",
                       "E2F7", "KIF14", "AICDA", "PLK4", "CDK1", "EZH2", "AURKB", "E2F8", "BRCA1", "POLE2", "TERT", "CDK2")

ggplot(dea.df, aes(x = log2FoldChange, y = neglog10padj)) +
  geom_point(aes(color = padj < 0.25 & abs(log2FoldChange) > 1), size = 0.5) +
  scale_color_manual(values = c("grey", "orange")) +
  geom_text_repel(
    data = subset(dea.df, gene_name %in% genes_of_interest),
    aes(label = gene_name),
    max.overlaps = Inf,
    size = 4
  ) +
  labs(x = "Log2 Fold Change", y = "-log10 Adjusted P-value") +
  scale_y_continuous(limits = c(0, 20)) +
  theme_minimal() +
  theme(legend.position = "none")
```

Comparison of our volcano plot with the original results (Fig. 5b) shows consistency, supporting the validity of the pipeline used in our study.

## Gene Set Enrichment Analysis (GSEA)

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df)
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector, gene_ident = "ensembl_gene")
```

```{r}
# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will explore the trends of hallmark terms reported from other single-cell pseudobulk studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as E2F targets, Myc targets, and oxidative phosphorylation (OXPHOS) were upregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

From the plots, terms such as E2F and Myc targets were consistent with the two single-cell studies while the OXPHOS trend was not clear.

#### GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector, gene_ident = "ensembl_gene")
```

```{r}
# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

 We will also explore the trends of GO:BP terms reported from other single-cell studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as OXPHOS and mitochondrial translation were upregulated while BCR signalling was downregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

Although the terms such as OXPHOS and mitochondrial translation were not siginificantly enriched, the trends were quite upregulated, consistent with the two single-cell studies. Furthermore, BCR signalling was significantly downregulated, similar to that reported in the Nadeu et al. (2022). 

```{r, eval=FALSE}
# Save gene expression and GSEA data
saveRDS(HM_gsea, path_to_save_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_go_gsea)
```





