---
title: "Batch_method_evaluation"
author: "Pakorn Sagulkoo"
format:
  html: 
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This source code we evaluated performance of batch removal methods in both running time, batch mixing, and biological preservation. We adapted codes and concepts from [Integrative single-cell meta-analysis reveals disease-relevant vascular cell states and markers in human atherosclerosis](https://www.cell.com/cell-reports/fulltext/S2211-1247(23)01392-X) that integrated single-cell data sets for human atherosclerosis study.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(SingleR)
library(celldex)
library(sctransform)
library(ggsci)
library(tidyverse)
library(dplyr)
library(data.table)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save
path_to_object <- here::here("scRNAseq/meta_analysis/R_objects")
path_to_merged_obj <- paste0(path_to_object, "/merged_seu.rds")
path_to_save_sample_merged_obj <- paste0(path_to_object, "/subsampled_merged_seu.rds")
path_to_save_sample_split_lst <- paste0(path_to_object, "/subsampled_split_seu.rds")
path_to_save_integrated_lst <- paste0(path_to_object, "/integrated_method_lst.rds")
path_to_save_running_time_lst <- paste0(path_to_object, "/running_time_lst.rds")
path_to_save_metric_evaluate_lst <- paste0(path_to_object, "/metric_eval_lst.rds")
path_to_save_figure <- here::here("figures")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r, eval=FALSE}
# Load the integrated seurat object
seurat <- readRDS(path_to_merged_obj)
```

# Subset some stuides to evaluate batch removal

As our merged dataset contains a large number of cells, we subsampled data from selected studies, such as Nadeu (2022), Hing (2023), and Rigo (2024), to evaluate performance more efficiently in terms of both time and memory.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Subsample the merged object and reprocessing
DefaultAssay(seurat) <- "RNA"
seurat <- subset(seurat, subset = ref %in% c("Nadeu_2022", "Hing_2022", "Rigo_2024"))
seurat <- JoinLayers(seurat)
seurat[["RNA"]] <- split(seurat[["RNA"]], f = seurat$case_id) 
seurat <- SCTransform(seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
seurat <- RunPCA(seurat)
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "pca")
seurat <- singleR_cell_annotation(seurat)

# Save the object
saveRDS(seurat, path_to_save_sample_merged_obj)
```

```{r, fig.height=7, fig.width=11}
# Visualisation for merged results
seurat <- readRDS(path_to_save_sample_merged_obj)

# Plot based on patient, ref, cond, and cell types
group_by <- c("patient", "ref", "cond", "singleR.labels_main")
umap_plots <- lapply(group_by, function (group){
  DimPlot(seurat, reduction = "umap", group.by = group, raster=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})
merged_umap.plt <- wrap_plots(umap_plots, ncol=2)
merged_umap.plt
```

# Batch removals

We evaluated five different batch correction methods for single-cell data integration: canonical correlation analysis (CCA) with mutual nearest neighbors (MNN), reciprocal principal component analysis (rPCA), joint PCA (jPCA), harmony, and Scanorama from Python.

```{r, eval=FALSE}
# Create lists collecting running time and integrated objects
running_time.lst <- list()
integrated_lst <- list()
```

```{r, message=FALSE, warning=FALSE}
# Read the objects used for visualisation as the all processes took a long time 
running_time.lst <- readRDS(path_to_save_running_time_lst)
integrated_lst <- readRDS(path_to_save_integrated_lst)
```

## CCA + MNN

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Perform the batch removal method with recording running time
start_time <- Sys.time()
integrated_seurat <- IntegrateLayers(object = seurat, 
                                     method = CCAIntegration, 
                                     normalization.method = "SCT", 
                                     verbose = FALSE)
end_time <- Sys.time()
running_time <- as.numeric(difftime(end_time, start_time, units = "mins"))
running_time.lst$cca <- running_time
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Processing and downstream analysis
integrated_seurat <- RunPCA(integrated_seurat)
integrated_seurat <- RunUMAP(integrated_seurat, dims = 1:30, reduction = "integrated.dr")
integrated_seurat <- FindNeighbors(integrated_seurat, dims = 1:30, reduction = "integrated.dr") 
integrated_seurat <- FindClusters(
  integrated_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Collect the object in the list
integrated_lst$cca <- integrated_seurat
```

```{r, fig.height=7, fig.width=11}
# Visualisation for integrated results
integrated_seurat <- integrated_lst$cca

# Plot based on patient, ref, cond, and cell types
group_by <- c("patient", "ref", "cond", "singleR.labels_main")
umap_plots <- lapply(group_by, function (group){
  DimPlot(integrated_seurat, reduction = "umap", group.by = group, raster=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})
cca_umap.plt <- wrap_plots(umap_plots, ncol=2)
cca_umap.plt
```

## rPCA

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Perform the batch removal method with recording running time
start_time <- Sys.time()
integrated_seurat <- IntegrateLayers(object = seurat, 
                                     method = RPCAIntegration, 
                                     normalization.method = "SCT", 
                                     verbose = FALSE)
end_time <- Sys.time()
running_time <- as.numeric(difftime(end_time, start_time, units = "mins"))
running_time.lst$rpca <- running_time
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Processing and downstream analysis
integrated_seurat <- RunPCA(integrated_seurat)
integrated_seurat <- RunUMAP(integrated_seurat, dims = 1:30, reduction = "integrated.dr")
integrated_seurat <- FindNeighbors(integrated_seurat, dims = 1:30, reduction = "integrated.dr") 
integrated_seurat <- FindClusters(
  integrated_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Collect the object in the list
integrated_lst$rpca <- integrated_seurat
```

```{r, fig.height=7, fig.width=11}
# Visualisation for integrated results
integrated_seurat <- integrated_lst$rpca

# Plot based on patient, ref, cond, and cell types
group_by <- c("patient", "ref", "cond", "singleR.labels_main")
umap_plots <- lapply(group_by, function (group){
  DimPlot(integrated_seurat, reduction = "umap", group.by = group, raster=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})
rpca_umap.plt <- wrap_plots(umap_plots, ncol=2)
rpca_umap.plt
```

## jPCA

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Perform the batch removal method with recording running time
start_time <- Sys.time()
integrated_seurat <- IntegrateLayers(object = seurat, 
                                     method = JointPCAIntegration, 
                                     normalization.method = "SCT", 
                                     verbose = FALSE)
end_time <- Sys.time()
running_time <- as.numeric(difftime(end_time, start_time, units = "mins"))
running_time.lst$jpca <- running_time
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Processing and downstream analysis
integrated_seurat <- RunPCA(integrated_seurat)
integrated_seurat <- RunUMAP(integrated_seurat, dims = 1:30, reduction = "integrated.dr")
integrated_seurat <- FindNeighbors(integrated_seurat, dims = 1:30, reduction = "integrated.dr") 
integrated_seurat <- FindClusters(
  integrated_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Collect the object in the list
integrated_lst$jpca <- integrated_seurat
```

```{r, fig.height=7, fig.width=11}
# Visualisation for integrated results
integrated_seurat <- integrated_lst$jpca

# Plot based on patient, ref, cond, and cell types
group_by <- c("patient", "ref", "cond", "singleR.labels_main")
umap_plots <- lapply(group_by, function (group){
  DimPlot(integrated_seurat, reduction = "umap", group.by = group, raster=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})
jpca_umap.plt <- wrap_plots(umap_plots, ncol=2)
jpca_umap.plt
```

## Harmony

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Perform the batch removal method with recording running time
start_time <- Sys.time()
integrated_seurat <- RunHarmony(seurat, group.by.vars = "case_id", dims = 1:30)
end_time <- Sys.time()
running_time <- as.numeric(difftime(end_time, start_time, units = "mins"))
running_time.lst$harmony <- running_time
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Processing and downstream analysis
integrated_seurat <- RunUMAP(integrated_seurat, dims = 1:30, reduction = "harmony")
integrated_seurat <- FindNeighbors(integrated_seurat, dims = 1:30, reduction = "harmony") 
integrated_seurat <- FindClusters(
  integrated_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Collect the object in the list
integrated_lst$harmony <- integrated_seurat
```

```{r, fig.height=7, fig.width=11}
# Visualisation for integrated results
integrated_seurat <- integrated_lst$harmony

# Plot based on patient, ref, cond, and cell types
group_by <- c("patient", "ref", "cond", "singleR.labels_main")
umap_plots <- lapply(group_by, function (group){
  DimPlot(integrated_seurat, reduction = "umap", group.by = group, raster=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})
harmony_umap.plt <- wrap_plots(umap_plots, ncol=2)
harmony_umap.plt
```

## Scanorama

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set the environment to run python and import scanorama
library(reticulate)
use_condaenv("scanorama_env", required = TRUE)
scanorama <- import("scanorama")
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Prepare objects and features for integration
seurat.lst <- SplitObject(seurat, split.by = "case_id")
seurat.lst <- lapply(seurat.lst, function(obj){
  obj <- SCTransform(obj, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
})
saveRDS(seurat.lst, path_to_save_sample_split_lst) # Save the list

# Prepare matrices and gene lists for Scanorama
sct_mat.lst <- list()
gene.lst <- list()

for (i in seq_along(seurat.lst)) {
  mat <- GetAssayData(seurat.lst[[i]], assay = "SCT", slot = "data") # Extract SCT data (genes × cells)
  sct_mat.lst[[i]] <- t(as.matrix(mat)) # Transpose to cells × genes (required by Scanorama)
  gene.lst[[i]] <- rownames(mat) # Correct gene list
}
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Perform the batch removal method with recording running time
start_time <- Sys.time()
integrated.corrected.data <- scanorama$correct(sct_mat.lst, gene.lst, return_dimred=TRUE, return_dense=TRUE)
end_time <- Sys.time()
running_time <- as.numeric(difftime(end_time, start_time, units = "mins"))
running_time.lst$scanor <- running_time
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Add the batch corrected data to the object
intdata <- lapply(integrated.corrected.data[[2]], t)
panorama <- do.call(cbind, intdata)
rownames(panorama) <- as.character(integrated.corrected.data[[3]]) # genes
cell_names <- unlist(lapply(sct_mat.lst, rownames))
colnames(panorama) <- cell_names

# Get PCA embeddings
intdimred <- do.call(rbind, integrated.corrected.data[[1]])
rownames(intdimred) <- cell_names
colnames(intdimred) <- paste0("PC_", 1:ncol(intdimred))
stdevs <- apply(intdimred, MARGIN = 2, FUN = sd) # dd standard deviations in order to draw Elbow Plots in Seurat

# Combined meta.data
meta.data <- do.call(rbind, lapply(seurat.lst, function(obj) obj@meta.data))
rownames(meta.data) <- sub("^[^.]+\\.", "", rownames(meta.data)) # strip rownames from the original to merge with scanor results
meta.data <- meta.data[cell_names, ]  # reorder to match columns of panorama
rownames(meta.data) <- cell_names

# Create a seurat object
integrated_seurat <- CreateSeuratObject(
  counts = panorama,
  assay = "pano",
  project = "rt_cll_metaanalysis",
  meta.data = meta.data
)
integrated_seurat[["pca"]] <- CreateDimReducObject(embeddings = intdimred, stdev = stdevs, key = "PC_", assay = "pano")
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Downstream analysis
integrated_seurat <- RunUMAP(integrated_seurat, dims = 1:30)
integrated_seurat <- FindNeighbors(integrated_seurat, dims = 1:30) 
integrated_seurat <- FindClusters(
  integrated_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Collect the object in the list
integrated_lst$scanor <- integrated_seurat
```

```{r, fig.height=7, fig.width=11}
# Visualisation for integrated results
integrated_seurat <- integrated_lst$scanor

# Plot based on patient, ref, cond, and cell types
group_by <- c("patient", "ref", "cond", "singleR.labels_main")
umap_plots <- lapply(group_by, function (group){
  DimPlot(integrated_seurat, reduction = "umap", group.by = group, raster=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})
scanor_umap.plt <- wrap_plots(umap_plots, ncol=2)
scanor_umap.plt
```

## Save the results

```{r, eval=FALSE}
# Save the running time and seurat objects
saveRDS(running_time.lst, path_to_save_running_time_lst)
saveRDS(integrated_lst, path_to_save_integrated_lst)
```

# Evaulation in running time

```{r}
# Set label and colour pallete
method_map <- c(
  "cca"     = "CCA_MNN",
  "rpca"    = "rPCA",
  "jpca"    = "jPCA",
  "harmony" = "Harmony",
  "scanor"  = "Scanorama"
)

method_cols <- c(
  "Scanorama" = "#F05C3B",
  "CCA_MNN"   = "#197EC0",
  "jPCA"      = "#C80813",
  "rPCA"      = "#FD8CC1",
  "Harmony"   = "#46732E"
)
```

```{r}
# Create a dataframe collecting running time in each method
running_time.df <- data.frame(
  method = c("CCA_MNN", "rPCA", "jPCA", "Harmony", "Scanorama"),
  running_time_min = c(
    running_time.lst$cca,
    running_time.lst$rpca,
    running_time.lst$jpca,
    running_time.lst$harmony,
    running_time.lst$scanor
  )
)

# Plot a bar graph for running time
running_time.plt <- running_time.df %>%
  dplyr::mutate(method = fct_reorder(method, running_time_min)) %>%
  ggplot(aes(x=method, y=running_time_min, fill=method)) +
  geom_col(width = 0.6) +
  xlab("Method") +
  ylab("Running time (mins)") +
  scale_fill_manual(values = method_cols) +  
  theme(aspect.ratio = 1.3,
        axis.text.x = element_text(angle=45, hjust=1),
        legend.position = "none")
  
running_time.plt  
```

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/all_running_time.png"), plot=running_time.plt, dpi=500, height=3, width=4)
```

We can see from the plot that harmony provided the significantly lowest running time compared to other methods.

# Batch removal and biological conservation performance

We also evaluated the batch removal performance as well as biological conservation among the five techniques. We computed four metrices, such as local inverse Simpson's index (LISI), k-nearest-neighbours batch effect test (kBET), cluster average silhouette widths (aASW), and principal component regression (PCRegression), to assess the performance.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Import the relevant packages
library(kBET)
library(FNN)
library(lisi)
library(cluster)
library(stats)
library(DescTools)

# Create a list collecting metric results
eval_matric.lst <-list()
```

```{r}
# Load the existing results
eval_matric.lst <-readRDS(path_to_save_metric_evaluate_lst)
```

## LISI

LISI is employed to quantify batch diversity among neighbouring cells using a k-nearest neighbour (KNN) graph. A higher LISI score indicates superior batch mixing among adjacent cells, signifying enhanced batch-effect removal performance.

We commenced by calculating batch LISI (bLISI), using references or studies as the batch variable, in order to assess the batch mixing achieved by these batch removal techniques.

Furthermore, we also accounted for the deleterious effects of batch overcorrection, which could potentially disrupt biological structures. Subsequently, we evaluated biological conservation through cell LISI (cLISI) and cluster LISI (clLISI), which utilise cell annotation and clusters as batch variables. Elevated cLISI and clLISI scores reflect a high degree of mixing for these variables across their subgroups,for instance, high mixing between two different cell types or clusters, consequently leading to poorer biological conservation.

```{r, eval=FALSE}
# Define function to compute batch LISI (bLISI), cell LISIS (cLISI), and cluster LISI (clLISI)
compute_lisi_scores <- function(seu, batch_var = "ref") {
  pca_list <- seu@reductions$pca
  emb <- pca_list@cell.embeddings[, 1:30]
  cluster_candidates <- c("SCT_snn_res.0.3", "pano_snn_res.0.3")
  cluster_var <- cluster_candidates[cluster_candidates %in% colnames(seu@meta.data)][1]
  label_vars <- c("singleR.labels_main", cluster_var)
  meta <- seu@meta.data[, c(batch_var, label_vars), drop = FALSE]
  lisi_res <- compute_lisi(emb, meta, c(batch_var, label_vars))
  df <- as.data.frame(lisi_res) # convert to data frame with clear names
  colnames(df) <- c("bLISI", "cLISI", "clLISI")
  return(df)
}
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Compute LISI
lisi_results <- lapply(names(integrated_lst), function(name) {
    obj <- integrated_lst[[name]]
    df <- compute_lisi_scores(obj) # compute
    df$method <- name
    return(df)
  })

lisi_df <- rbindlist(lisi_results, use.names = TRUE, fill = TRUE)
lisi_df <- lisi_df %>% dplyr::mutate(method_name = method_map[method])
eval_matric.lst$lisi <- lisi_df
```

```{r, message=FALSE, warning=FALSE}
# Plot mean LISI scores
lisi_df <- eval_matric.lst$lisi

## bLISI score
mean_blisi.plt <- lisi_df %>%
  dplyr::select(c("bLISI", "method_name")) %>%
  dplyr::group_by(method_name) %>%
  dplyr::summarize(mean_bLISI = mean(bLISI)) %>%
  dplyr::mutate(method_name = fct_reorder(as.factor(method_name), mean_bLISI, .desc=TRUE)) %>%
  ggplot(aes(x=method_name, y=mean_bLISI, fill=method_name)) + 
  geom_col(width = 0.6) + 
  xlab("Method") + 
  ylab("Mean bLISI") + 
  scale_fill_manual(values = method_cols) + 
  theme_bw() +
  theme(aspect.ratio = 1.3,
        legend.position = "none",
        axis.text.x = element_text(angle=45, hjust=1))

## cLISI score
mean_clisi.plt <- lisi_df %>%
  dplyr::select(c("cLISI", "method_name")) %>%
  dplyr::group_by(method_name) %>%
  dplyr::summarize(mean_cLISI = mean(cLISI)) %>%
  dplyr::mutate(method_name = fct_reorder(as.factor(method_name), mean_cLISI, .desc=TRUE)) %>%
  ggplot(aes(x=method_name, y=mean_cLISI, fill=method_name)) + 
  geom_col(width = 0.6) + 
  xlab("Method") + 
  ylab("Mean cLISI") + 
  scale_fill_manual(values = method_cols) + 
  theme_bw() +
  theme(aspect.ratio = 1.3,
        legend.position = "none",
        axis.text.x = element_text(angle=45, hjust=1))

## clLISI score
mean_cllisi.plt <- lisi_df %>%
  dplyr::select(c("clLISI", "method_name")) %>%
  dplyr::group_by(method_name) %>%
  dplyr::summarize(mean_clLISI = mean(clLISI)) %>%
  dplyr::mutate(method_name = fct_reorder(as.factor(method_name), mean_clLISI, .desc=TRUE)) %>%
  ggplot(aes(x=method_name, y=mean_clLISI, fill=method_name)) + 
  geom_col(width = 0.6) + 
  xlab("Method") + 
  ylab("Mean clLISI") + 
  scale_fill_manual(values = method_cols) + 
  theme_bw() +
  theme(aspect.ratio = 1.3,
        legend.position = "none",
        axis.text.x = element_text(angle=45, hjust=1))

mean_lisi.plt <- mean_blisi.plt + mean_clisi.plt + mean_cllisi.plt
mean_lisi.plt
```

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/all_blisi.png"), plot=mean_blisi.plt, dpi=500, height=3, width=4)
ggsave(filename=paste0(path_to_save_figure, "/all_clisi.png"), plot=mean_clisi.plt, dpi=500, height=3, width=4)
ggsave(filename=paste0(path_to_save_figure, "/all_cllisi.png"), plot=mean_cllisi.plt, dpi=500, height=3, width=4)
```

From the results we found that CCA, Harmony, rPCA, and jPCA provided the highest bLISI score (1.23) while Scanorama had the lowest score as 1.04. For cLISI, all methods provided the same score as 1.07. CCA also provided the highest clLISI (1.48), following with jPCA (1.44), rPCA (1.37), Harmony (1.35), and Scanorma (1.05).

Overall when cosidering both batch removal and biological conservation with LISI metric, Harmony had the best performance.

## kBET

The kBET assesses how closely the label composition of a cell's k-nearest neighbourhood (KNN) aligns with the expected label composition, with the result reported as rejection rates. Lower rejection rates indicate superior batch mixing.

We computed the rejection rates by varying the cell's neighbourhood size (k) and subsequently calculated the Area Under the Curve (AUC) between the rejection rates and k. A lower AUC value signifies greater batch removal efficacy.

```{r, eval=FALSE}
# Define function to compute kBET
compute_kbet <- function(seu, batch_var = "ref", k_range) {
  pca_list <- seu@reductions$pca
  emb <- pca_list@cell.embeddings[, 1:30]
  batch <- seu@meta.data[[batch_var]]
  kbet_list <- vector("list", length(k_range))
  
  for (i in seq_along(k_range)) {
    
    batch_est <- kBET(
      df = emb, 
      batch = batch, 
      k0 = k_range[i], 
      do.pca = FALSE, 
      plot = FALSE, 
      heuristic = FALSE,
      adapt = FALSE,
      verbose = TRUE
    )
    
    df <- data.frame(
      class = rep(c("observed", "expected"),
                  each = length(batch_est$stats$kBET.observed)),
      data  = c(batch_est$stats$kBET.observed,
                batch_est$stats$kBET.expected),
      k     = paste0("k_", k_range[i])
    )
    
    kbet_list[[i]] <- df
  }
  return(rbindlist(kbet_list, use.names = TRUE, fill = TRUE))
}

# Define a base for k range
base_k <- c(10, 25, 50, 100, 500, 1000, 2500)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Compute kBET
kbet_results <- lapply(names(integrated_lst), function(name) {
    obj <- integrated_lst[[name]]
    
    # Set k_range
    batch_vec <- obj@meta.data$ref
    batch_labels <- as.factor(batch_vec)
    k0 = floor(mean(table(batch_labels))) / 4 # set k0 as the original article suggested
    k_range <- unique(sort(c(base_k, k0)))
    n_cells <- ncol(obj)
    k_range <- k_range[k_range < n_cells] # remove impossible k
    
    # Run kBET
    df <- compute_kbet(obj, batch_var = "ref", k_range)
    df$method <- name
    return(df)
})
    
kbet_df <- rbindlist(kbet_results, use.names = TRUE, fill = TRUE)
kbet_df <- kbet_df %>% dplyr::mutate(method_name = method_map[method])
eval_matric.lst$kbet <- kbet_df
```

```{r, message=FALSE, warning=FALSE, fig.width=10}
# Plot kBET observed rejection rates curve
kbet_df <- eval_matric.lst$kbet

kbet_rates_curve.plt = kbet_df %>%
  dplyr::filter(class == "observed") %>%
  dplyr::group_by(method_name, k) %>%
  dplyr::summarize(mean_obs_rejection_rate = mean(data)) %>%
  ggplot(aes(x=fct_reorder(k, mean_obs_rejection_rate), 
             mean_obs_rejection_rate, 
             color=method_name)) + 
  geom_line(aes(x=fct_reorder(k, mean_obs_rejection_rate),
                y=mean_obs_rejection_rate, group=method_name), 
            linetype="dashed") + 
  geom_point(size=1.5) +
  #geom_col(position = "dodge") + 
  labs(x="Neighborhood size (k)",
       y="Mean kBET observed rejection rate",
       color="Method") + 
  theme_bw() +
  theme(aspect.ratio = 1.3,
        legend.position = "right") + 
  scale_color_manual(values= method_cols)

kbet_rates_curve.plt
```

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/all_kbet.png"), plot=kbet_rates_curve.plt, dpi=500, height=5, width=7)
```

```{r, message=FALSE, warning=FALSE}
# Calculate AUC in each method
auc.lst <- list()
methods <- c("CCA_MNN", "rPCA", "jPCA", "Harmony", "Scanorama")

for (i in seq_along(methods)) { 
  
  # Keep only observed rejection rates and compute means
  sorted_rej_rates <- kbet_df %>% 
    dplyr::filter(class == "observed" & method_name == methods[i]) %>%
    dplyr::group_by(method_name, k) %>%
    dplyr::summarize(mean_obs_rejection_rate = mean(data), .groups = "drop") %>%
    dplyr::arrange(mean_obs_rejection_rate) %>%
    dplyr::mutate(k_index = seq(1, length(unique(k)), by=1))
  
  # Calculate AUC 
  auc <- DescTools::AUC(x=sorted_rej_rates$k_index,
                       y=sorted_rej_rates$mean_obs_rejection_rate,
                       method="spline")
  auc.lst[[i]] <- auc
  names(auc.lst)[i] <- methods[i]

}

auc.lst
```

From the AUC analysis, Harmony provided the lowest score (6.857699). This was closely followed by rPCA (6.85806), then CCA (6.858154), and jPCA (6.858713). Scanorama had the highest AUC score (6.999909). Given that a lower AUC indicates greater batch removal efficacy, Harmony provided the best results for batch removal according to this metric.

## cASW

The Silhouette Score is typically employed to evaluate the appropriate parameter settings for clustering. We adapted this concept to assess the clustering results derived from the batch removal methods by varying the resolution parameter for the single-cell data from 0.1 to 0.9. A higher cASW indicates superior clustering, which in this context signifies improved biological conservation.

```{r, eval=FALSE}
# Define a function to compute cluster average silhouette width (clASW)
compute_sil <- function(seu, res){
  pca_list <- seu@reductions$pca
  emb <- pca_list@cell.embeddings[, 1:30]
  dist_mat <- dist(emb)
  sil_list <- vector("list", length(res))
  cluster_candidates <- c("SCT_snn_res.", "pano_snn_res.")
  matches <- cluster_candidates[sapply(cluster_candidates, function(x) any(grepl(x, colnames(seu@meta.data))))]
  cluster_var <- matches[1]
  
  for (i in seq_along(res)){
    clusters <- seu[[paste0(cluster_var, res[i])]][,1]
    sil <- silhouette(x = as.numeric(x = as.factor(x = clusters)), dist = dist_mat)
    df <- data.frame(cluster=sil[, 1], neighbor=sil[, 2], sil_width=sil[, 3], res=res[i])
    sil_list[[i]] <- df
  }
  return(rbindlist(sil_list, use.names = TRUE, fill = TRUE))
}

# Define a cluster resolution range
res <- seq(0.1, 0.9, by = 0.1)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Compute silhouette score
sil_results <- lapply(names(integrated_lst), function(name) {
  obj <- integrated_lst[[name]]
  df <- compute_sil(obj, res)
  df$method <- name
  return(df)
})

sil_df <- rbindlist(sil_results, use.names = TRUE, fill = TRUE)
sil_df <- sil_df %>% dplyr::mutate(method_name = method_map[method])
eval_matric.lst$sil <- sil_df
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=5.5}
# Plot silhouette scores
sil_df <- eval_matric.lst$sil

sil_scores.plt <- sil_df %>%
  dplyr::group_by(res, method_name) %>%
  dplyr::summarise(mean_sil = mean(sil_width), .groups = "drop") %>%
  ggplot(aes(x = factor(res), y = mean_sil, fill = method_name)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  xlab("Clustering resolution") +
  ylab("Mean Silhouette score") +
  scale_fill_manual(values = method_cols) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(fill = "Method") 

sil_scores.plt
```

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/all_sil.png"), plot=sil_scores.plt, dpi=500, height=5, width=5.5)
```

When varying the resolution parameter, Scanorama demonstrated robustness in preserving biological conservation, as it yielded positive cASW scores across all resolutions tested. Harmony, conversely, exhibited the lowest positive cASW score at resolution 0.1 compared to the other methods; however, all positive scores were observed at every resolution except 0.8 and 0.9.

## PCRegression

This metric is an adaptation designed to calculate the Principal Component-based variance using a linear model, with the effect size of the batch attributed to the *R^2^* value. A higher *R^2^* value indicates a greater batch effect, thereby signifying poorer batch removal performance. As the observed *R^2^* values were very low, we transformed the metric using the negative logarithmic form for reporting. Consequently, higher reported values reflect superior batch removal.

```{r, eval=FALSE}
# Define a function to compute PCRegress
compute_pcr <- function(seu, batch_var){
  pca_list <- seu@reductions$pca
  pca_data <- list(x = pca_list@cell.embeddings, sdev = pca_list@stdev)
  class(pca_data) <- "prcomp"  # pcRegression expects a prcomp object
  batch <- as.factor(seu[[batch_var, drop = TRUE]])  # use column dynamically
  pcr <- kBET::pcRegression(pca.data=pca_data,
                            batch=batch,
                            n_top=30)
  return(pcr$R2Var)
}
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Compute PCR 
pcr_results <- lapply(names(integrated_lst), function(name) { 
  obj <- integrated_lst[[name]]
  R2Var <- compute_pcr(obj, batch_var = "ref")
  df <- data.frame(R2Var = R2Var, method = name)
  return(df)
})
  
pcr_df <- rbindlist(pcr_results, use.names = TRUE, fill = TRUE)
pcr_df <- pcr_df %>% dplyr::mutate(method_name = method_map[method])
eval_matric.lst$pcr <- pcr_df
```

```{r}
# Plot the -log10 values of variance explained by batch effects
pcr_df <- eval_matric.lst$pcr

r2var.plt <- pcr_df %>%
  dplyr::mutate(method_name=fct_reorder(method_name, -log10(R2Var), .desc=TRUE)) %>%
  ggplot(aes(x=method_name, y=-log10(R2Var), fill=method_name)) + 
  geom_col(width = 0.6) + 
  xlab("Method") + 
  ylab("-Log10(R2Var)") +
  scale_fill_manual(values = method_cols) + 
  theme_bw() +
  theme(aspect.ratio = 1.3,
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust=1))

r2var.plt
```

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/all_r2var.png"), plot=r2var.plt, dpi=500, height=3, width=4)
```

From the plots, it was observed that all methods except Scanorama exhibited the highest negative *R^2^* variance.

## Save the metric results

```{r, eval=FALSE}
# Save the metric results
saveRDS(eval_matric.lst, path_to_save_metric_evaluate_lst)
```

Considering running time, batch removal efficacy, and biological conservation performance collectively, Harmony proved to be the optimal option across these metrics. We subsequently employed Harmony for all further downstream analyses.

# Session Information

```{r}
sessionInfo()
```
