---
title: "Data_cleaning_and_cell_extraction"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This source code we cleaned the integrated data sets from several CLL and RT studies by removing some noises or ambient cells. Moreover. we extracted tumour and microenvironmental cells for further analysis,

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tidyr)
library(SingleR)
library(celldex)
library(sctransform)
library(cowplot)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_integrated_obj <- here::here("scRNAseq/meta_analysis/R_objects/integrated_seu.rds")
path_to_save_filtered_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered_seu.rds")
path_to_save_tumour_obj <- here::here("scRNAseq/meta_analysis/R_objects/tumour_seu.rds")
path_to_save_microenvir_obj <- here::here("scRNAseq/meta_analysis/R_objects/microenvir_seu.rds")
path_to_save_figure <- here::here("figures")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the integrated seurat object
seurat <- readRDS(path_to_integrated_obj)
seurat$patient <- paste0(seurat$case, "_", seurat$ref)
seurat$adapted_ref <- sub("^(.*)_([0-9].*)$", "\\1 (\\2)", seurat$ref) # Substitiute string like "Parry-2023b" to "Parry (2023b)"
seurat$adapted_ref <- factor(seurat$adapted_ref, level = c("Penter (2021)", "Nadeu (2022)", "Rejeski (2022)", "Hing (2023)", 
                                                 "Parry (2023a)", "Parry (2023b)", "Rigo (2024)"))
#seurat$group <- ifelse(grepl("CLL", seurat$cond), "CLL", "RT")
```

# Data visualisation

```{r, message=FALSE, warning=FALSE}
# Plot summarisation for cell population

study_data <- data.frame(
  Study = c("Penter (2021)", "Nadeu (2022)", "Rejeski (2022)", 
            "Hing (2023) Human", "Hing (2023) Mouse", 
            "Parry (2023a)", "Parry (2023b)", "Rigo (2024)"),
  Total_Cells = c(309282, 18676, 24596, 27257, 81954, 44541, 186939, 7072),
  Avg_Cells = c(11454, 1098, 12298, 4542, 4097, 1310, 5341, 3536)
)
study_data$Study <- factor(study_data$Study, levels = study_data$Study) # Reorder the Factor 

study_long <- study_data %>% # Reshape to "Long" format for easy plotting with ggplot
  pivot_longer(cols = c("Total_Cells", "Avg_Cells"), 
               names_to = "Metric", 
               values_to = "Count") %>%
  mutate(Metric_Label = ifelse(Metric == "Total_Cells", "Total Cells", "Avg Cells / Sample"))

cell_bar.plt <- ggplot(study_long, aes(x = Study, y = Count, fill = Study)) +
  geom_bar(stat = "identity", width = 0.7, color = "black") +
  facet_wrap(~Metric_Label, scales = "free_y") +
  geom_text(aes(label = round(Count)), vjust = -0.5, size = 3) +
  theme_minimal() +
  labs(title = "Summary of Single-Cell Datasets",
       y = "Number of Cells",
       x = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
        strip.text = element_text(size = 12, face = "bold"),
        legend.position = "none") + # Hide legend since x-axis labels are sufficient
  scale_fill_brewer(palette = "Set3")

cell_bar.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/study_cell_pop.png"), plot=cell_bar.plt, dpi=500, height=5, width=10)
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=10}
# Plot based on case
case.plt <- DimPlot(seurat, reduction = "umap", group.by = "patient", raster=TRUE) +                        
  labs(y= "UMAP 2", x = "UMAP 1")
case.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=9}
# Plot based on reference, conditions, tissues, and cell types
group_by <- c("adapted_ref", "cond", "tissue", "singleR.labels_main")
umap_plots <- lapply(group_by, function (group){
  DimPlot(seurat, reduction = "umap", group.by = group, raster=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})
integrated_umap.plt <- wrap_plots(umap_plots, ncol=2)
integrated_umap.plt
```

We can see from the UMAP plots that harmony removed batches and integrated all data sets clearly as no case or study were grouped separately. The clusters were separated based on the cell types.

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
varied_clusters.plt <- plot_varying_cluster_resolution(seurat)
varied_clusters.plt
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

```{r, fig.height=8, fig.width=9}
cluster.plt <- DimPlot(seurat, 
                       reduction = "umap", 
                       group.by = "SCT_snn_res.0.3", 
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt
```

```{r, fig.height=8, fig.width=9}
cell.plt <- DimPlot(seurat, 
                    reduction = "umap", 
                    group.by = "singleR.labels_main", 
                    repel=TRUE, 
                    raster=TRUE, 
                    label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

cell.plt
```

# Cell noise removal

We detected erythrocytes (cluster 28 at SCT_snn_res.0.3) and other ambient cells (cluster 18, 31, 34, and 36). Then we removed them.

```{r, eval=FALSE}
# Remove ambient cells and processing after noise removal
filtered_seurat <- subset(seurat, subset = !(SCT_snn_res.0.3 %in% c(18, 28, 31, 34, 36)))
filtered_seurat <- SCTransform(filtered_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered_seurat <- RunPCA(filtered_seurat)
filtered_seurat <- RunUMAP(filtered_seurat, dims = 1:30, reduction = "harmony")
filtered_seurat <- FindNeighbors(filtered_seurat, dims = 1:30, reduction = "harmony") 
filtered_seurat <- FindClusters(
  filtered_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Save the object
saveRDS(filtered_seurat, path_to_save_filtered_obj)
```

# Visualisation for integrated data

```{r}
# Read the dbs filtered object for visualisation
filtered_seurat <- readRDS(path_to_save_filtered_obj)
```

```{r, message=FALSE, warning=FALSE}
# Plot UMAP for cell annotation

cell_to_annotation <- c(
  "B cells" = "B cell",
  "Monocytes" = "Monocyte/Macrophage",
  "NK cells" = "NK cell",
  "T cells, CD4+" = "CD4+ T cell",
  "T cells, CD8+" = "CD8+ T cell"
)

filtered_seurat$cell_annotate <- 
  as.vector(cell_to_annotation[as.character(filtered_seurat$singleR.labels_main)])

new_levels <- c(
  "B cell",
  "Monocyte/Macrophage",
  "NK cell",
  "CD4+ T cell",
  "CD8+ T cell"
)

filtered_seurat$cell_annotate <- factor(
  filtered_seurat$cell_annotate,
  levels = new_levels
)

new_colours <- c(
  "#D43F3A", 
  "#EEA236",
  "#5CB85C",
  "#357EBD",
  "#46B8DA"
)

cell.plt <- DimPlot(filtered_seurat, 
                    group.by = "cell_annotate", 
                    pt.size = 0.05,
                    raster = FALSE) +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels
  ) +
  labs(y = "UMAP 2", 
       x = "UMAP 1",
       title = "Harmony integration",
       subtitle = "(n=593,424)") +
  theme(
    plot.title = element_text(hjust = 0.5), # Center title
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  coord_fixed(ratio = 1)

cell.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/harmony_cell_umap.png"), plot=cell.plt, dpi=500, height=7, width=7)
```

```{r, message=FALSE, warning=FALSE}
# Plot cells by spliting the condition 
filtered_seurat$group <- ifelse(grepl("CLL", filtered_seurat$cond), "CLL", "RT")
filtered_seurat$group_label <- ifelse(grepl("CLL", filtered_seurat$cond), "CLL (n=361,266)", "RT (n=232,158)")
cell_split.plt <- DimPlot(filtered_seurat, 
                             reduction = "umap", 
                             group.by = "cell_annotate", 
                             split.by = "group_label") +
  NoLegend() +
  labs(y = "UMAP 2", 
       x = "UMAP 1",
       title = "") +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels
  ) +
  coord_fixed(ratio = 1)

cell_split.plt 
```

```{r, message=FALSE, warning=FALSE}
# Plot cells by spliting the tissues 
filtered_seurat$tissue_label <- ifelse(grepl("LN", filtered_seurat$tissue), "LN (n=42,359)", 
                                       ifelse(grepl("PB", filtered_seurat$tissue), "PB (n=340,431)", "BM (n=210,634)"))

tissue_split.plt <- DimPlot(filtered_seurat, 
                             reduction = "umap", 
                             group.by = "cell_annotate", 
                             split.by = "tissue_label") +
  NoLegend() +
  labs(y = "UMAP 2", 
       x = "UMAP 1",
       title = "") +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels
  ) +
  coord_fixed(ratio = 1)

tissue_split.plt 
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/harmony_cell_split.png"), plot=cell_split.plt , dpi=500, height=7, width=7)
ggsave(filename=paste0(path_to_save_figure, "/harmony_tissue_split.png"), plot=tissue_split.plt , dpi=500, height=7, width=7)
```

```{r}
# Plot cell population

proportion_study_df <- filtered_seurat@meta.data %>%
  dplyr::mutate(cell_annotate = factor(cell_annotate, levels = new_levels)) %>%
  dplyr::count(adapted_ref, cell_annotate) %>%
  dplyr::group_by(adapted_ref) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n) # Remove n column

proportion_group_df <- filtered_seurat@meta.data %>%
  dplyr::mutate(cell_annotate = factor(cell_annotate, levels = new_levels)) %>%
  dplyr::count(group, cell_annotate) %>%
  dplyr::group_by(group) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n)

proportion_tissue_df <- filtered_seurat@meta.data %>%
  dplyr::mutate(cell_annotate = factor(cell_annotate, levels = new_levels)) %>%
  dplyr::count(tissue, cell_annotate) %>%
  dplyr::group_by(tissue) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n)

study_cell_pop.plt <- ggplot(proportion_study_df, aes(x = adapted_ref, 
                                                    y = proportion, 
                                                    fill = cell_annotate)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "Main cell type" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Study", 
    title = "Study"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

study_cell_pop.plt

cond_cell_pop.plt <- ggplot(proportion_group_df, aes(x = group, 
                                                    y = proportion, 
                                                    fill = cell_annotate)) +
  geom_col(position = "stack", width = 0.4) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "Main cell type" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Condition", 
    title = "Condition"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

cond_cell_pop.plt

tissue_cell_pop.plt <- ggplot(proportion_tissue_df, aes(x = tissue, 
                                                    y = proportion, 
                                                    fill = cell_annotate)) +
  geom_col(position = "stack", width = 0.4) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "Main cell type" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Tissue", 
    title = "Tissue"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

tissue_cell_pop.plt
```

```{r}
# Plot cell population based tissues split conditions
proportion_tissue_cond_df <- filtered_seurat@meta.data %>%
  dplyr::mutate(cell_annotate = factor(cell_annotate, levels = new_levels)) %>%
  dplyr::count(tissue, group, cell_annotate) %>%
  dplyr::group_by(tissue, group) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n)

tissue_cond_cell_pop.plt <- ggplot(proportion_tissue_cond_df, 
                                   aes(x = group, 
                                       y = proportion, 
                                       fill = cell_annotate)) +
  
  geom_col(position = "stack", width = 0.8) + # Increased width slightly for readability
  facet_wrap(~tissue, scales = "free_x") + # Use free_x if tissues have different numbers of bars
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "Main cell type"
  ) +
  labs(
    y = "Cell proportion",
    x = "Condition", # The x-axis is now the condition
    title = "Condition within Tissue"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_rect(fill = "grey90", color = NA),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

tissue_cond_cell_pop.plt
```

```{r}
# Plot cell population based cases

proportion_patient_df <- filtered_seurat@meta.data %>%
  dplyr::mutate(cell_annotate = factor(cell_annotate, levels = new_levels)) %>%
  dplyr::count(patient, cell_annotate) %>%
  dplyr::group_by(patient) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n)

patient_cell_pop.plt <- ggplot(proportion_patient_df, aes(x = patient, 
                                                    y = proportion, 
                                                    fill = cell_annotate)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "Main cell type" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Case", 
    title = "Case"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

patient_cell_pop.plt
```

```{r}
# Combined bar plots
adapted_study_cell_pop_no_legend.plt <- study_cell_pop.plt + theme(legend.position = "none", axis.title.x = element_blank())

adapted_cond_cell_pop.plt <- cond_cell_pop.plt + 
  theme(legend.position = "bottom", 
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) 

adapted_tissue_cell_pop.plt <- tissue_cell_pop.plt +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) 

combined_cell_pop.plt <- adapted_study_cell_pop_no_legend.plt + adapted_cond_cell_pop.plt + adapted_tissue_cell_pop.plt
combined_cell_pop.plt
```

```{r, eval=FALSE}
# Save the figures and object
ggsave(filename=paste0(path_to_save_figure, "/cell_pop_study.png"), plot=study_cell_pop.plt , dpi=500, height=4, width=5)
ggsave(filename=paste0(path_to_save_figure, "/cell_pop_cond.png"), plot=cond_cell_pop.plt , dpi=500, height=4, width=5)
ggsave(filename=paste0(path_to_save_figure, "/cell_pop_tissue.png"), plot=tissue_cell_pop.plt , dpi=500, height=4, width=5)
ggsave(filename=paste0(path_to_save_figure, "/cell_pop_case.png"), plot=patient_cell_pop.plt , dpi=500, height=4, width=10)
ggsave(filename=paste0(path_to_save_figure, "/cell_pop_combined.png"), plot=combined_cell_pop.plt , dpi=500, height=4, width=7)
ggsave(filename=paste0(path_to_save_figure, "/cell_pop_tissue_cond.png"), plot=tissue_cond_cell_pop.plt , dpi=500, height=4, width=7)
saveRDS(filtered_seurat, path_to_save_filtered_obj)
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(filtered_seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

# Cell biomarkers

```{r, message=FALSE, warning=FALSE}
# Dot plot for cell biomarkers
Idents(filtered_seurat) <- filtered_seurat$cell_annotate
features <- c("CD3D", "CD4", "IL7R", "CD8A", "CD8B", "CD19", "CD79A", "MS4A1", "NCAM1", "NKG7", "CD14", "CD16", "FCGR3A", "CD68")
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

biomark.plt <- DotPlot(filtered_seurat, features=features) +
  RotatedAxis() +
  scale_colour_gradientn(colours = colour_pal)

biomark.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/harmony_marker_dot.png"), plot=biomark.plt, dpi=500, height=4, width=8)
```

```{r, message=FALSE, warning=FALSE, fig.height=14, fig.width=12}
# Visualise cell biomarker expression
DefaultAssay(filtered_seurat) <- "SCT"
features <- c("CD3D", "CD4", "IL7R", "CD8A", "CD8B", "CD19", "CD79A", "MS4A1", "NKG7", "CD14", "FCGR3A", "CD68")
feature_plots <- lapply(features, function(gene) {
  FeaturePlot(filtered_seurat, 
              features = gene, 
              reduction = "umap", 
              pt.size = 2,
              slot = "data",
              order = TRUE) +
    scale_colour_gradientn(colours=colour_pal, limits=c(0, 2)) +
    coord_fixed(ratio = 1) +
    labs(y = "UMAP 2", x = "UMAP 1") +
    ggtitle(gene) 
})

# Combine all plots (patchwork automatically combines lists)
exprs_umap_pre.plt <- wrap_plots(feature_plots, ncol = 3)  
exprs_umap_pre.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=14, fig.width=8}
# Visualise cell biomarker expression for publication
DefaultAssay(filtered_seurat) <- "SCT"
feature_plots <- lapply(features, function(gene) {
  FeaturePlot(filtered_seurat, 
              features = gene, 
              reduction = "umap", 
              pt.size = 2,
              slot = "data",
              order = TRUE) +
    scale_colour_gradientn(colours=colour_pal, limits=c(0, 2)) +
    coord_fixed(ratio = 1) +
    theme_void() +
    theme(legend.position="none",
          plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
          ) +
    labs(y = "UMAP 2", x = "UMAP 1") +
    ggtitle(gene) 
})

# Combine all plots (patchwork automatically combines lists)
exprs_umap.plt <- wrap_plots(feature_plots, ncol = 3)  
exprs_umap.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/harmony_exprs_umap.png"), plot=exprs_umap.plt, dpi=500, height=12, width=8)
ggsave(filename=paste0(path_to_save_figure, "/harmony_pre_exprs_umap.png"), plot=exprs_umap_pre.plt, dpi=500, height=12, width=10)
```

# Cell extraction

## Tumour cell extraction

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Extract tumour cell with processing
tumour_seurat <- subset(filtered_seurat, subset = singleR.labels_main == "B cells")
tumour_seurat <- SCTransform(tumour_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
tumour_seurat <- RunPCA(tumour_seurat)
tumour_seurat <- RunUMAP(tumour_seurat, dims = 1:30, reduction = "harmony")
tumour_seurat <- FindNeighbors(tumour_seurat, dims = 1:30, reduction = "harmony") 
tumour_seurat <- FindClusters(
  tumour_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Save the object
saveRDS(tumour_seurat, path_to_save_tumour_obj)
```

## Microenvironmental cell extraction

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Extract non-tumour cell with processing
microenvir_seurat <- subset(filtered_seurat, subset = singleR.labels_main != "B cells")
microenvir_seurat <- SCTransform(microenvir_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
microenvir_seurat <- RunPCA(microenvir_seurat)
microenvir_seurat <- RunUMAP(microenvir_seurat, dims = 1:30, reduction = "harmony")
microenvir_seurat <- FindNeighbors(microenvir_seurat, dims = 1:30, reduction = "harmony")        
microenvir_seurat <- FindClusters(
  microenvir_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Save the object
saveRDS(microenvir_seurat, path_to_save_microenvir_obj)
```

# Session Information

```{r}
sessionInfo()
```
