---
title: "Tumour_cell_pseudobulk_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This source code we analysed the tumour cells after extracting from the harmony-integrated seurat objects. We analysed for pseudobulk gene expression and GSEA analysis between RT versus CLL conditions. 

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tibble)
library(SingleR)
library(celldex)
library(sctransform)
library(cowplot)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_tumour_obj <- here::here("scRNAseq/meta_analysis/R_objects/tumour_seu.rds")
path_to_save_figure <- here::here("figures")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the tumour object
tumour_seurat <- readRDS(path_to_tumour_obj)
```

# Filter out ambient cells

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(tumour_seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

As no contaminated or ambient cells were found, then we used the same object for further analyses

# Pseudobulk gene expression analysis between RT versus CLL

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(ggpubr)
```

## Based on seurat command

```{r}
# Define paths to save and load
path_to_save_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_dea.rds")
path_to_save_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_hm_gsea.rds")
path_to_save_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_go_gsea.rds")
path_to_save_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_kegg_gsea.rds")
path_to_save_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_reactome_gsea.rds")
path_to_save_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_wikipath_gsea.rds")
```

### Find biomarkers of tumour cells between RT and CLL conditions

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Find markers between CLL and RT conditions  
Idents(tumour_seurat) <- "group" # we used this variable as it contains RT and CLL
tumour_seurat <- PrepSCTFindMarkers(tumour_seurat) # Prepare SCT assay for DE testing
tumour.ge <- FindMarkers(
  tumour_seurat,
  ident.1 = "RT",
  ident.2 = "CLL",
  test.use = "wilcox",
  only.pos = FALSE,
  logfc.threshold = 0
)
dea.df <- cbind(gene = rownames(tumour.ge), tumour.ge)
avg_exp <- AverageExpression(tumour_seurat, features = dea.df$gene, assay = "SCT") # Calculate the average expression for the genes
dea.df$baseMean <- rowMeans(avg_exp$SCT) # Calculate the mean across all relevant groups (e.g., global mean for the gene)

# Save the expression results
saveRDS(dea.df, path_to_save_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/tumour_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Load the saved expression data
dea.df <- readRDS(path_to_save_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=10}
# Volcano plot for Seurat pseudobulk analysis
genes_of_interest <- c("MKI67", "CDK1", "AURKB", "PRMT5", "CDKN3", "CDKN1B", "CREBBP", "CHD7", "TYMS", "SF1", "NPM1", "TOP2A",
                       "TPX2", "UBE2C", "BIRC5", "KIF14", "LDHA", "MTHFD2", "HMGA1", "CKS2", "IGKC")

non_zero_min_pval <- sort(unique(dea.df$p_val_adj), decreasing = FALSE)[2] # Find minimum p-value which is not 0
dea.df$adapted_p_val <- ifelse(dea.df$p_val_adj == 0, non_zero_min_pval, dea.df$p_val_adj) # Change zero adj_p_val to non-zero minimum adj_p_val to avoid plotting error

volcano.plt <- volcano_plot(dea.df, 
                            FC=0.25, 
                            x_col="avg_log2FC",
                            y_col="adapted_p_val",
                            xlim=c(-10,10), 
                            ylim=c(-0.1,320), 
                            genes_of_interest=genes_of_interest,
                            boxedLabels = TRUE) 

volcano.plt
```

```{r, message=FALSE, warning=FALSE}
# MA plot for Seurat pseudobulk analysis
ma_data <- dea.df
colnames(ma_data)[colnames(ma_data) == "avg_log2FC"] <- "log2FoldChange"
colnames(ma_data)[colnames(ma_data) == "p_val_adj"]  <- "padj"

ma.plt <- ggmaplot(ma_data, 
                   fdr = 0.05,            # Threshold for significance
                   fc = 2,                # Threshold for FC
                   size = 0.4,            # Dot size
                   palette = c("#B31B21", "#1465AC", "darkgray"),
                   genenames = as.vector(ma_data$gene),
                   top = 0, label.select = genes_of_interest, # set top = 0, then specify genes using label.select argument
                   legend = "right", 
                   font.label = c("bold", 11), 
                   label.rectangle = TRUE,
                   font.legend = "bold",
                   font.main = "bold",
                   ggtheme = ggplot2::theme_minimal(),
                   xlab = "Mean Expression",
                   ylab = "Log2 Fold Change"
                   )

ma.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/seurat_volcano.png"), plot=volcano.plt, dpi=500, height=7, width=9)
ggsave(filename=paste0(path_to_save_figure, "/seurat_ma.png"), plot=ma.plt, dpi=500, height=3.5, width=5)
```

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea.df$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save GSEA results
saveRDS(HM_gsea, path_to_save_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_go_gsea)
saveRDS(KEGG_gsea, path_to_save_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_wikipath_gsea.csv"))
```

## Based on DESeq2

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(DESeq2)
library(limma)
```

```{r}
# Define paths to save the results
path_to_save_deseq2_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_deseq2_dea.rds")
path_to_save_deseq2_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_deseq2_hm_gsea.rds")
path_to_save_deseq2_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_deseq2_go_gsea.rds")
path_to_save_deseq2_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_deseq2_kegg_gsea.rds")
path_to_save_deseq2_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_deseq2_reactome_gsea.rds")
path_to_save_deseq2_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_deseq2_wikipath_gsea.rds")
```

### Create an DESeq2 object

```{r, message=FALSE, warning=FALSE}
# Prepare single-cell meta-data for DESeq2
tumour_seurat$patient_group <- paste0("case_", tumour_seurat$patient, "_", tumour_seurat$group) 
patient_group_cls <- unique(tumour_seurat$patient_group)
RT_subjects <- patient_group_cls[grep("RT", patient_group_cls)]
RT_subjects <- gsub(pattern = "_", replacement = "-", x=RT_subjects) # change "_" to "-" as DESeq2 converted sample names from "_" to "-"

# Create pseudocount matrix for tumour cell object
cts <- AggregateExpression(tumour_seurat, 
                           group.by = "patient_group",
                           assays = "RNA",
                           slot = "counts",
                           resturn.seurat = FALSE) 
cts <- cts$RNA
pseudo_count.mat <- as.data.frame(cts)

# Create metadata for DESeq2 object
batch_map <- c("Penter (2021)" = "1", "Nadeu (2022)" = "2", "Rejeski (2022)" = "3", "Hing (2023)" = "4", "Parry (2023a)" = "5", 
               "Parry (2023b)" = "6", "Rigo (2024)" = "7")
colData <- data.frame(samples = colnames(pseudo_count.mat))
colData$condition <- ifelse(colData$samples %in% RT_subjects, "RT", "CLL")
colData$study <- stringr::word(colData$samples, start = -3, end = -2, sep = fixed("-"))
colData$study <- sub("^(.*)-([0-9].*)$", "\\1 (\\2)", colData$study) # Substitiute string like "Parry-2023b" to "Parry (2023b)"
colData$study <- factor(colData$study, level = c("Penter (2021)", "Nadeu (2022)", "Rejeski (2022)", "Hing (2023)", 
                                                 "Parry (2023a)", "Parry (2023b)", "Rigo (2024)"))
colData$batch <- batch_map[colData$study]
colData$batch <- factor(colData$batch)
rownames(colData) <- colData$samples

# Create a DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = pseudo_count.mat, 
                              colData = colData,
                              design = ~ batch + condition)

# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

### Prepare the input for Genome-scale metabolic models (GEMs)

We also converted the raw count matrix to TPM matrix for GEMsx with ftINIT for the next step.

```{r, eval=FALSE}
# Import the biomaRt package to get the human gene length
library(biomaRt)

# 1. Connect to Human Ensembl (hg38)
human_mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# 2. Get Gene Symbols and Transcript Lengths
gene_attributes <- getBM(
  attributes = c("hgnc_symbol", "transcript_length"),
  filters = "hgnc_symbol",
  values = rownames(pseudo_count.mat), 
  mart = human_mart
)

# 3. Take the MAX length for each gene (Longest Isoform)
# Since one gene has many transcripts, we pick the longest one as a proxy
gene_lengths_max <- gene_attributes %>%
  filter(hgnc_symbol != "") %>%        # Remove rows with no gene name
  group_by(hgnc_symbol) %>%            # Group by gene name
  summarise(Length = max(transcript_length))

saveRDS(gene_lengths_max, here::here("scRNAseq/meta_analysis/R_objects/gene_lengths_max.rds"))
```

```{r}
# Read the gene length
gene_lengths_max <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_lengths_max.rds"))
head(gene_lengths_max)
```

```{r, message=FALSE, warning=FALSE}
# Calculate TPM
matrix_genes <- rownames(pseudo_count.mat)
match_indices <- match(matrix_genes, gene_lengths_max$hgnc_symbol)
valid_lengths <- gene_lengths_max$Length[match_indices]
na_count <- sum(is.na(valid_lengths)) # evaluate missing data

if (na_count > 0) {
  median_len <- median(gene_lengths_max$Length, na.rm = TRUE)
  valid_lengths[is.na(valid_lengths)] <- median_len
}

calculate_tpm <- function(counts, lengths) {
  rpk <- counts / (lengths / 1000) # Reads Per Kilobase
  scale_factor <- colSums(rpk) # Sum the RPKs for each sample (scaling factor)
  tpm <- t(t(rpk) / scale_factor) * 1e6 # Divide RPK by the scaling factor (per million) 
  # transpose is needed because R divides matrices by rows, not columns
  return(tpm)
}

tpm_matrix <- calculate_tpm(pseudo_count.mat, valid_lengths)
colSums(tpm_matrix) # Check the output (Column sums should all be exactly 1,000,000)
```

```{r, eval=FALSE}
# Convert to data frames
raw_count.df <- as_tibble(pseudo_count.mat, rownames = "Gene")
tpm_count.df <- as_tibble(tpm_matrix, rownames = "Gene")

# Save the data frames
write.table(raw_count.df, here::here("scRNAseq/meta_analysis/df_files/raw_count.txt"), row.names=FALSE, quote=FALSE, sep="\t")
write.table(tpm_count.df, here::here("scRNAseq/meta_analysis/df_files/tpm_count.txt"), row.names=FALSE, quote=FALSE, sep="\t")
```

### Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6.5}
# Variance stabilised transformation
vsdata <- DESeq2::vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
design_matrix <- model.matrix(~ condition, data = colData(dds))
vs_mat_corrected <- limma::removeBatchEffect(
  vs_mat, 
  batch = colData(dds)$batch, 
  design = design_matrix 
)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2               # variance of each PC
pca_var_percent <- pca_var / sum(pca_var) * 100

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     study = colData(dds)$study)

# Plot PCA
pca.plt <- ggplot(pca_df, aes(PC1, PC2, color = study, shape = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  scale_color_manual(values = c(
    "Penter (2021)" = "#1f77b4", 
    "Nadeu (2022)" = "#ff7f0e",
    "Rejeski (2022)" = "#9632B8" ,
    "Hing (2023)" = "#D43F3A" ,
    "Parry (2023a)" = "#5CB85C",
    "Parry (2023b)" = "#FFB900",
    "Rigo (2024)" = "#46B8DA")) +
  scale_shape_manual(values = c("CLL" = 16, "RT" = 17)) +
  guides(
    color = guide_legend(title = "Study"),
    shape = guide_legend(title = "Condition")
  ) +
  theme_bw() +
  theme(
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 11)
  )

pca.plt
```

From the PCA plot we detected an outlier from the right lower corner with the highest PC1 value we then removed it before further analysis

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6.5}
# Plot the PCA plot after outlier removal

# Identify and Remove the Outlier 
outlier_id <- rownames(pca_df)[which.max(pca_df$PC1)] # the outlier is the sample with the highest PC1 value
dds_clean <- dds[, colnames(dds) != outlier_id] # Filter the outlier out of the original dds object

# Re-run VST on the cleaned dataset
vsdata_clean <- vst(dds_clean, blind = FALSE)
vs_mat_clean <- assay(vsdata_clean)

# Adjust for batch (assume batch info is in colData(dds)$batch)
design_matrix <- model.matrix(~ condition, data = colData(dds_clean))
vs_mat_corrected <- limma::removeBatchEffect(
  vs_mat_clean, 
  batch = colData(dds_clean)$batch, 
  design = design_matrix # <--- This argument is key for separation!
)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2
pca_var_percent <- pca_var / sum(pca_var) * 100

# Make a dataframe for ggplot
pca_df_clean <- data.frame(pca$x,
                     condition = colData(dds_clean)$condition,
                     study = colData(dds_clean)$study)

# Plot PCA
corrected_pca.plt <- ggplot(pca_df_clean, aes(PC1, PC2, color = study, shape = condition)) + 
  geom_point(size = 3) +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  scale_color_manual(values = c(
    "Penter (2021)" = "#1f77b4", 
    "Nadeu (2022)" = "#ff7f0e",
    "Rejeski (2022)" = "#9632B8" ,
    "Hing (2023)" = "#D43F3A" ,
    "Parry (2023a)" = "#5CB85C",
    "Parry (2023b)" = "#FFB900",
    "Rigo (2024)" = "#46B8DA")) +
  scale_shape_manual(values = c("CLL" = 16, "RT" = 17)) +
  guides(
    color = guide_legend(title = "Study"),
    shape = guide_legend(title = "Condition")
  ) +
  theme_bw() +
  theme(
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 11)
  ) 

corrected_pca.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/deseq2_pca.png"), plot=pca.plt, dpi=500, height=4.5, width=6.5)
ggsave(filename=paste0(path_to_save_figure, "/deseq2_corrected_pca.png"), plot=corrected_pca.plt, dpi=500, height=4.5, width=6.5)
```

We can see that after outlier removal and batch correction, the PCA successfully separated the conditions along the PC1 axis. Most RT samples (triangles) are located on the negative side of PC1, while most CLL samples (circles) are found on the positive side. 

### Run DESeq2

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Run DESeq2 for the data after outlier removal
counts_clean <- counts(dds_clean)
colData_clean <- colData(dds_clean)
colData_clean$condition <- factor(colData_clean$condition, levels = c("CLL", "RT"))
colData_clean$batch <- factor(colData_clean$batch) 

dds_final <- DESeqDataSetFromMatrix(
  countData = counts_clean,
  colData = colData_clean,
  design = ~ batch + condition  
)

keep <- rowSums(counts(dds_final) >= 10) >= 3 # Keep genes with at least 10 counts in at least 3 samples
dds_final <- dds_final[keep, ]
dds_final$condition <- relevel(dds_final$condition, ref = "CLL")
dds_final <- DESeq(dds_final)

# Generate a table for gene expression
res <- results(dds_final, contrast = c("condition", "RT", "CLL"), independentFiltering = FALSE)
res <- lfcShrink(dds_final, contrast = c("condition", "RT", "CLL"), res = res, type = "normal")
res$direction <- "NS"
res$direction[res$padj < 0.05 & res$log2FoldChange > 0.25] <- "Up"
res$direction[res$padj < 0.05 & res$log2FoldChange < (-0.25)] <- "Down"
dea.df <- as.data.frame(res)

# Save the expression results
saveRDS(dea.df, path_to_save_deseq2_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/tumour_deseq2_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Show the expression results
dea.df <- readRDS(path_to_save_deseq2_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

We listed the DEGs for both up- and downregulation below.

```{r}
# Upregulated DEGs arranged by log2FC from high to low
up_degs <- dea.df %>%
  dplyr::filter(padj < 0.05 & log2FoldChange > 0.25) %>%
  dplyr::arrange(desc(log2FoldChange)) %>%
  tibble::rownames_to_column(var = "gene") %>%
  dplyr::pull(gene)

up_degs
```

```{r}
# Downregulated DEGs arranged by log2FC from high to low
down_degs <- dea.df %>%
  dplyr::filter(padj < 0.05 & log2FoldChange < (-0.25)) %>%
  dplyr::arrange(log2FoldChange) %>%
  tibble::rownames_to_column(var = "gene") %>%
  dplyr::pull(gene)

down_degs
```

```{r, eval=FALSE}
# Save the results
write.table(
  up_degs,
  file = here::here("scRNAseq/meta_analysis/supplm_file/tumour_deseq2_upregulated_degs.txt"),   
  sep = "\t",        
  quote = FALSE,     
  row.names = FALSE,
  col.names = FALSE
)

write.table(
  down_degs,
  file = here::here("scRNAseq/meta_analysis/supplm_file/tumour_deseq2_downregulated_degs.txt"),   
  sep = "\t",        
  quote = FALSE,     
  row.names = FALSE,
  col.names = FALSE
)
```

We compared our identified DEGs with those reported in previous human RT versus CLL transcriptomic studies and found that MKI67, NPM1, TOP2A, TPX2, CDK1, UBE2C, BIRC5, AURKB, and KIF14, were significantly upregulated and consistent with prior findings, whereas PRMT5 was upregulated but did not reach statistical significance (padj = 0.0556). Downregulated DEGs, such as CREBBP, CHD7, and CDKN1B, were also consistent.

In addition, we identified other significant genes from the same functional families that have not been reported in previous studies. For example, CDKN3 was identified as an upregulated DEG. We detected that TYMS had the highest overexpression level (log2FC = 2.219). 

We also identified the DEGs that were found from both DESeq2-based pseudobulk analysis and the common DEGs from individual study pseudobulk analysis (see the following sections). For example, LDHA, MTHFD2, HMGA1, and CKS2, upregulated DEGs, have been reported the roles in tumour metabolic reprogramming, proliferation, and aggressiveness for other cancers. Moreover, SF1 was a downregulated DEG from the both methods.

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=10}
# Volcano plot
genes_of_interest <- c("MKI67", "CDK1", "AURKB", "PRMT5", "CDKN3", "CDKN1B", "CREBBP", "CHD7", "TYMS", "SF1", "NPM1", "TOP2A",
                       "TPX2", "UBE2C", "BIRC5", "KIF14", "LDHA", "MTHFD2", "HMGA1", "CKS2")

volcano.plt <- volcano_plot(dea.df, 
                            FC=0.25, 
                            x_col="log2FoldChange",
                            y_col="padj",
                            xlim=c(-2.5,2.5), 
                            ylim=c(-0.1,8), 
                            genes_of_interest=genes_of_interest,
                            boxedLabels = TRUE) 

volcano.plt
```

```{r, message=FALSE, warning=FALSE}
# MA plot 
ma_data <- dea.df

ma.plt <- ggmaplot(ma_data, 
                   fdr = 0.05,            # Threshold for significance
                   fc = 1.19,                # Threshold for FC
                   size = 0.4,            # Dot size
                   palette = c("#B31B21", "#1465AC", "darkgray"),
                   genenames = as.vector(ma_data$gene),
                   top = 0, label.select = genes_of_interest, # set top = 0, then specify genes using label.select argument
                   legend = "right", 
                   font.label = c("bold", 11), 
                   label.rectangle = TRUE,
                   font.legend = "bold",
                   font.main = "bold",
                   ggtheme = ggplot2::theme_minimal(),
                   xlab = "Mean Expression",
                   ylab = "Log2 Fold Change"
                   )

ma.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/deseq2_volcano.png"), plot=volcano.plt, dpi=500, height=7, width=9)
ggsave(filename=paste0(path_to_save_figure, "/deseq2_ma.png"), plot=ma.plt, dpi=500, height=3.5, width=5)
```

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df) 
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save the enrichment results
saveRDS(HM_gsea, path_to_save_deseq2_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_deseq2_go_gsea)
saveRDS(KEGG_gsea, path_to_save_deseq2_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_deseq2_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_deseq2_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_deseq2_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_deseq2_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_deseq2_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_deseq2_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/tumour_deseq2_wikipath_gsea.csv"))
```

## GSEA summary 

We then summarised the common terms found from Hallmark, GO:BP, and canonical pathways.

```{r, message=FALSE, warning=FALSE}
# Bar plot of GSEA 
seurat_reactome_result <- readRDS(path_to_save_reactome_gsea)
seurat_reactome_selected.result <- seurat_reactome_result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

gsea_bp <- rbind(HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_OXIDATIVE_PHOSPHORYLATION",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_FATTY_ACID_METABOLISM",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_ADIPOGENESIS",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_CHOLESTEROL_HOMEOSTASIS",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_GLYCOLYSIS",],
                 KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_PYRUVATE_METABOLISM",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_DNA_REPAIR",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_MITOCHONDRIAL_TRANSLATION",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_CELL_CYCLE_PHASE_TRANSITION",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_KEAP1_NFE2L2_PATHWAY",],
                 WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_NRF2_PATHWAY",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_MTORC1_SIGNALING",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_IL2_STAT5_SIGNALING",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_MYC_TARGETS_V1",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_E2F_TARGETS",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_G2M_CHECKPOINT",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_MAPK6_MAPK4_SIGNALING",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_SIGNALING_BY_NOTCH4",],
                 WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_INSULIN_SIGNALING",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_TNFA_SIGNALING_VIA_NFKB",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_TGF_BETA_SIGNALING",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_REGULATION_OF_JNK_CASCADE",],
                 seurat_reactome_selected.result[seurat_reactome_selected.result$Description == "REACTOME_SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_RAF_ACTIVATION",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_SIGNALING_BY_FGFR3",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_SIGNALING_BY_FGFR4",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_INTERLEUKIN_17_SIGNALING",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_FOXO_MEDIATED_TRANSCRIPTION",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_SIGNALING_BY_EGFR",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_SIGNALING_BY_MET",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_MYD88_INDEPENDENT_TLR4_CASCADE",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_TOLL_LIKE_RECEPTOR_9_TLR9_CASCADE",],
                 WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL1_SIGNALING",]
)

gsea_bp$Description <- gsub("HALLMARK_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("GOBP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("KEGG_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("REACTOME_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("WP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("_V1", "", gsea_bp$Description)
gsea_bp$Description <- gsub("SIGNALING_BY_NOTCH4", "NOTCH4_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("SIGNALING_BY_FGFR3", "FGFR3_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("SIGNALING_BY_FGFR4", "FGFR4_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("SIGNALING_BY_EGFR", "EGFR_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("SIGNALING_BY_MET", "MET_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("TOLL_LIKE_RECEPTOR_9_TLR9_CASCADE", "TLR9_CASCADE", gsea_bp$Description)
gsea_bp$Description <- gsub("REGULATION_OF_JNK_CASCADE", "JNK_CASCADE_REGULATION", gsea_bp$Description)
gsea_bp$Description <- gsub("SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR", "BCR_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("_", " ", gsea_bp$Description)

gsea_bp$Description <- factor(gsea_bp$Description, levels = gsea_bp$Description)
gsea_bp$Direction <- "Up"
gsea_bp$Direction[gsea_bp$NES<0] <- "Down"

gsea_bar.plt <- ggplot(gsea_bp, aes(x=Description, y=NES, fill=Direction)) +
  geom_bar(stat = "identity", color="black", width = 0.75) +  
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, color = "black"),
    plot.margin = margin(t = 10, r = 10, b = 100, l = 65, unit = "pt"),
    legend.position = "none"
  ) +
  geom_hline(yintercept = 0) + 
  ylab("NES") + xlab(NULL) + 
  scale_fill_manual(values=c("#357EBD", "#D43F3A"))

gsea_bar.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=30, fig.width=13}
# GSEA plots
terms_of_interest <- c("HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_FATTY_ACID_METABOLISM", "HALLMARK_ADIPOGENESIS",
                       "HALLMARK_CHOLESTEROL_HOMEOSTASIS", "HALLMARK_GLYCOLYSIS", "KEGG_PYRUVATE_METABOLISM",
                       "REACTOME_DNA_REPAIR", "GOBP_MITOCHONDRIAL_TRANSLATION", "GOBP_CELL_CYCLE_PHASE_TRANSITION",
                       "WP_NRF2_PATHWAY", "HALLMARK_MTORC1_SIGNALING", "HALLMARK_IL2_STAT5_SIGNALING", "HALLMARK_MYC_TARGETS_V1",
                       "HALLMARK_MYC_TARGETS_V2", "HALLMARK_E2F_TARGETS", "HALLMARK_G2M_CHECKPOINT", "REACTOME_MAPK6_MAPK4_SIGNALING",
                       "REACTOME_SIGNALING_BY_NOTCH4", "WP_INSULIN_SIGNALING", "HALLMARK_TNFA_SIGNALING_VIA_NFKB", 
                       "HALLMARK_TGF_BETA_SIGNALING", "REACTOME_SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR",
                       "GOBP_REGULATION_OF_JNK_CASCADE", "REACTOME_RAF_ACTIVATION", "REACTOME_SIGNALING_BY_FGFR3",
                       "REACTOME_SIGNALING_BY_FGFR4", "REACTOME_INTERLEUKIN_17_SIGNALING", "REACTOME_FOXO_MEDIATED_TRANSCRIPTION",
                       "REACTOME_SIGNALING_BY_EGFR", "REACTOME_SIGNALING_BY_MET", "REACTOME_MYD88_INDEPENDENT_TLR4_CASCADE",
                       "REACTOME_TOLL_LIKE_RECEPTOR_9_TLR9_CASCADE", "WP_IL1_SIGNALING"
                       )

gsea_plot.lst <- lapply(terms_of_interest, function(term){
  term_system <- strsplit(term, split = "_")[[1]][1]
  
  adapted_term <- term %>% 
    str_remove("^(HALLMARK|GOBP|KEGG|REACTOME|WP)_") %>%
    str_replace("SIGNALING_BY_NOTCH4", "NOTCH4_SIGNALING") %>%
    str_replace("SIGNALING_BY_FGFR3", "FGFR3_SIGNALING") %>%
    str_replace("SIGNALING_BY_FGFR4", "FGFR4_SIGNALING") %>%
    str_replace("SIGNALING_BY_EGFR", "EGFR_SIGNALING") %>%
    str_replace("SIGNALING_BY_MET", "MET_SIGNALING") %>%
    str_replace("TOLL_LIKE_RECEPTOR_9_TLR9_CASCADE", "TLR9_CASCADE") %>%
    str_replace("REGULATION_OF_JNK_CASCADE", "JNK_CASCADE_REGULATION") %>%
    str_replace("SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR", "BCR_SIGNALING") %>%
    str_replace_all("_", " ")
  
  if (grepl("BCR", term)){
    gseaplot(seurat_reactome_result, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "HALLMARK"){
    gseaplot(HM_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "GOBP"){
    gseaplot(GOBP_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "KEGG") {
    gseaplot(KEGG_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "REACTOME") {
    gseaplot(REACTOME_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else {
    gseaplot(WIKIPATH_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  }
})

gsea.plt <- wrap_plots(gsea_plot.lst, ncol=3)
gsea.plt
```

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/gsea.png"), plot=gsea_bar.plt, dpi=500, height=5, width=10)
ggsave(filename=paste0(path_to_save_figure, "/gsea_plt.png"), plot=gsea.plt, dpi=500, height=30, width=13)
```

# Test overlapping with the common DEGs from the consistency evaluation

We evaluated the integration and batch correction efficacy for our method by comparing the common DEGs from individual analysis. In theory, we should find the all common DEGs in the expression analysis results from the data merging and integration.

```{r}
# Load the common DEGs list from the previous analysis
upregulated_common_degs <- read.table(here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/1.upregulated_common_degs.txt"))[,1]
downregulated_common_degs <- read.table(here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/1.downregulated_common_degs.txt"))[,1]
```

## DEGs from Seurat

```{r}
# Load the seurat expression results
seurat_dea.df <- readRDS(path_to_save_dea)
```

```{r}
# Overlapping in upregulated DEGs
seurat_up_degs <- seurat_dea.df %>%
  dplyr::filter(p_val_adj < 0.05 & avg_log2FC > 0.25) %>%
  dplyr::pull(gene)

setdiff(upregulated_common_degs, seurat_up_degs)
```

```{r}
# Overlapping in downregulated DEGs
seurat_down_degs <- seurat_dea.df %>%
  dplyr::filter(p_val_adj < 0.05 & avg_log2FC < (-0.25)) %>%
  dplyr::pull(gene)

setdiff(downregulated_common_degs, seurat_down_degs)
```

Although we detected some upregulated and downregulated common DEGs that were non-significant from the integrated data, the integrated data consisted of seven studies while the common DEGs came from only five studies. Then we extracted the five studies and performed pseudobulk analysis.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Extract only five studies and perform pseudobulk analysis
selected_studies_seurat <- subset(tumour_seurat, subset = !(ref %in% c("Rejeski_2022", "Parry_2023a"))) 
selected_studies_seurat <- SCTransform(selected_studies_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
selected_studies_seurat <- RunPCA(selected_studies_seurat)
selected_studies_seurat <- RunUMAP(selected_studies_seurat, dims = 1:30, reduction = "harmony")
Idents(selected_studies_seurat) <- "group" 
selected_studies_seurat <- PrepSCTFindMarkers(selected_studies_seurat) # Prepare SCT assay for DE testing
tumour.ge <- FindMarkers(
  selected_studies_seurat,
  ident.1 = "RT",
  ident.2 = "CLL",
  test.use = "wilcox",
  only.pos = FALSE,
  logfc.threshold = 0
)
dea.df <- cbind(gene = rownames(tumour.ge), tumour.ge)

# Save the expression results
path_to_save_selected_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/selected_study_tumour_dea.rds")
saveRDS(dea.df, path_to_save_selected_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/selected_study_tumour_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Load the five study seurat expression results 
path_to_save_selected_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/selected_study_tumour_dea.rds")
five_seurat_dea.df <- readRDS(path_to_save_selected_dea)
DT::datatable(five_seurat_dea.df, options = list(scrollX = TRUE))
```

```{r}
# Overlapping in upregulated DEGs
five_seurat_up_degs <- five_seurat_dea.df %>%
  dplyr::filter(p_val_adj < 0.05 & avg_log2FC > 0.25) %>%
  dplyr::pull(gene)

setdiff(upregulated_common_degs, five_seurat_up_degs)
```

```{r}
# Plot a Venn diagram
library(ggvenn)

up_degs_venn.plt <- ggvenn(
  list("Common\nUpregulated DEGs" = upregulated_common_degs,
       "Integrated\nupregulated DEGs" = five_seurat_up_degs),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5,
  set_name_size = 6,   # Size of the outside labels (Default is 6)
  text_size = 6        # Size of the numbers inside the diagram (Default is 4)
)

up_degs_venn.plt
```

```{r}
# Overlapping in downregulated DEGs
five_seurat_down_degs <- five_seurat_dea.df %>%
  dplyr::filter(p_val_adj < 0.05 & avg_log2FC < (-0.25)) %>%
  dplyr::pull(gene)

setdiff(downregulated_common_degs, five_seurat_down_degs)
```

```{r}
# Plot a Venn diagram
down_degs_venn.plt <- ggvenn(list("Common\ndownregulated DEGs" = downregulated_common_degs,
                                "Integrated\ndownregulated DEGs" = five_seurat_down_degs),
                           fill_color = c("#0073C2FF", "#EFC000FF"),
                           stroke_size = 0.5, 
  set_name_size = 6,   # Size of the outside labels (Default is 6)
  text_size = 6        # Size of the numbers inside the diagram (Default is 4)
)

down_degs_venn.plt
```

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/venn_up_degs.png"), plot=up_degs_venn.plt, dpi=500, height=4.5, width=7)
ggsave(filename=paste0(path_to_save_figure, "/venn_down_degs.png"), plot=down_degs_venn.plt, dpi=500, height=4.5, width=7)
```

Although our analysis included five studies and largely aligned with common DEG findings, we observed notable discrepancies. Among the candidate upregulated genes, ALKBH7, UFC1, and ATP5PO reached statistical significance but displayed marginal effect sizes (log2FC < 0.25). Furthermore, EIF3K failed to show significant upregulation, exhibiting a non-significant trend toward downregulation (negative log2FC). 

Additionally, for common downregulated DEGs, RPL21 and CCNI were significant but did not pass the expression cut-point while MDM4 and SRRM2 were slightly upregulated in the integrated results.

In summary, pseudobulk results derived from SCT-normalized matrices based on the integrated data were unreliable due to the discordance with the common DEGs from individual studies. Consequently, we adopted a DESeq2-based pseudobulk approach using raw counts, explicitly modeling batch effects within the experimental design. The DEGs identified through this robust pipeline were selected for all subsequent analyses.

## DEGs from DESeq2

```{r}
# Load the DESeq2 expression result
deseq2_dea.df <- readRDS(path_to_save_deseq2_dea)
```

```{r}
# Overlapping in upregulated DEGs
deseq2_up_degs <- deseq2_dea.df %>%
  dplyr::filter(padj < 0.05 & log2FoldChange > 0.25) %>%
  tibble::rownames_to_column(var = "gene") %>%
  dplyr::pull(gene)

setdiff(upregulated_common_degs, deseq2_up_degs)
```

```{r}
# Overlapping in downregulated DEGs
deseq2_down_degs <- deseq2_dea.df %>%
  dplyr::filter(padj < 0.05 & log2FoldChange < (-0.25)) %>%
  tibble::rownames_to_column(var = "gene") %>%
  dplyr::pull(gene)

setdiff(downregulated_common_degs, deseq2_down_degs)
```

We also identified the consensus DEGs between the common DEGs from individual data analysis and DESeq2-based pseudobulk analysis from the integrated data.

```{r}
# Upregulated consensus DEGs
intersect(upregulated_common_degs, deseq2_up_degs)
```

```{r}
# Downregulated consensus DEGs
intersect(downregulated_common_degs, deseq2_down_degs)
```

# Heatmap of gene expression for DESeq2

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(ComplexHeatmap)
library(circlize)
```

```{r}
# Define known up- and downregulated DEGs from previous publications
gene_categories <- list(
  # Upregulated DEGs
  Cell_Cycle_Up = c("MKI67", "PCNA", "CDK1", "CDK2", "CDK4", "CDK5", "CDK6", "CCND2", "CCND3", "CDKN2A", "CDKN3", "CKS1B", "E2F7", 
                    "ANP32B"),
  Mitosis_Up = c("FOXM1", "PLK1", "PLK4", "AURKA", "AURKB", "TUBB", "TUBA1C", "STMN1", "TPX2", "HMMR", "KIF14", "KIF18B", "BUB1", 
                 "UBE2C", "UBE2S", "BIRC5"),
  DNA_Repair_Up = c("POLH", "POLQ", "POLE2", "TERT", "BRCA1", "XRCC2", "AICDA", "TOP2A", "NPM1"),
  Oncogenes_Wnt_Up = c("MYC", "MYCBP", "EZH2", "PRMT5", "WNT3", "WNT5A", "WNT6", "WNT10A", "GAS5"),
  Immune_Surface_Up = c("TLR8", "TLR9", "TLR10", "CD24", "CD70", "LTB", "MS4A1", "BID"),
  
  # Downregulated DEGs
  Apoptosis_Down = c("BAX", "BAK1", "BAD", "BMF", "HRK", "BBC3", "PMAIP1", "BCL2L11"),
  Chromatin_TumorSup_Down = c("CDKN1B", "CREBBP", "EP300", "CHD2", "CHD7", "ARID1A", "ARID4B", "ARID5B", "SPEN", "BAZ2A"),
  B_Identity_MHC_Down = c("PAX5", "CD79A", "CD74", "CD37", "CD83", "CD99", "IGHV3-7", "IGLC7", "HLA-A", "HLA-B"),
  Signaling_Homing_Down = c("CXCR4", "ITGB1", "LYN", "TRAF3", "TRAF5", "NFKBIA", "WNT16", "CDK17"),
  Transcription_Stress_Down = c("FOS", "FOSB", "JUN", "JUNB", "ZFP36", "BTG1", "UBC")
)

known_degs <- unlist(gene_categories)

# Define common DEGs from our analysis
common_degs <- c(upregulated_common_degs, downregulated_common_degs)

# Define genes of interest
genes_of_interest <- c("MKI67", "CDK1", "AURKB", "PRMT5", "CDKN3", "CDKN1B", "CREBBP", "CHD7", "TYMS", "SF1", "NPM1", "TOP2A",
                       "TPX2", "UBE2C", "BIRC5", "KIF14", "LDHA", "MTHFD2", "HMGA1", "CKS2", "RAD51", "FASN", "CCNB1", 
                       "CCNB2", "CDCA3", "CDCA5", "CDCA7", "CDCA8", "CYLD", "PTPRO", "EPC1", "EPC2", "KAT7", "JMJD1C", 
                       "MAP3K1", "USP3", "USP15")

# Import mouse DEGs
all_mouse_up_degs <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/all_mouse_up_degs.rds"))
all_mouse_down_degs <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/all_mouse_down_degs.rds"))
mouse_expr.df <- readRDS(here::here("scRNAseq/meta_analysis/R_objects/all_mouse_gene_expr.rds")) %>%
  dplyr::select(gene_name, Hacken_direction, Hing_direction, Iyer_direction, Rompietti_direction) %>%
  dplyr::mutate(gene_name = case_when(
    gene_name == "GATD3"  ~ "GATD3B",
    gene_name == "BLTP3B" ~ "UHRF1BP1L",
    TRUE ~ gene_name  # This is the 'default' in case_when
  ))
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=12}
# Plot the heatmap

sample_order <- c("1_Penter_CLL", "2_Penter_CLL", "3_Penter_CLL", "4_Penter_CLL", "5_Penter_CLL", "6_Penter_CLL", "7_Penter_CLL", 
                  "8_Penter_CLL", "9_Penter_CLL", "12_Nadeu_CLL", "19_Nadeu_CLL", "3299_Nadeu_CLL", "365_Nadeu_CLL", "1_Hing_CLL", 
                  "2_Hing_CLL", "MDAC-ACC-17_Parry_CLL", "MDAC-ACC-18_Parry_CLL", "MDAC-ACC-29_Parry_CLL", "14_Rigo_CLL", 
                  "9_Penter_RT", "12_Nadeu_RT", "19_Nadeu_RT", "3299_Nadeu_RT", "365_Nadeu_RT", "1_Rejeski_RT", "1_Hing_RT", 
                  "2_Hing_RT", "4_Parry_RT", "10_Parry_RT", "43_Parry_RT", "DFCI-5448_Parry_RT", "MDAC-ACC-09_Parry_RT", 
                  "MDAC-ACC-14_Parry_RT", "14_Rigo_RT"
                 )

# 1. Generate the scaled expression matrix
scaled_expr_mat <- t(scale(t(vs_mat))) # Generate the scaled expression matrix
colnames(scaled_expr_mat) <- sub("^case-(.+)-([^-]+)-([^-]+)-([^-]+)$", "\\1_\\2_\\4", colnames(scaled_expr_mat))
scaled_expr_deg_mat <- scaled_expr_mat[rownames(scaled_expr_mat) %in% c(up_degs, down_degs),] # Select only DEGs
scaled_expr_deg_mat <- scaled_expr_deg_mat[, match(sample_order, colnames(scaled_expr_deg_mat))] # Reorder samples

# 2. Generate sample_data for heatmap
sample_data_heatmap <- colData
sample_data_heatmap$condition <- factor(sample_data_heatmap$condition, level = c("CLL", "RT"))
sample_data_heatmap$samples <- sub("^case-(.+)-([^-]+)-([^-]+)-([^-]+)$", "\\1_\\2_\\4", sample_data_heatmap$samples)
sample_data_heatmap <- sample_data_heatmap[match(sample_order, sample_data_heatmap$samples),]

sample_data_heatmap$included_dea <- "Yes"
sample_data_heatmap$included_dea[sample_data_heatmap$samples == "2_Hing_CLL"] <- "No"
sample_data_heatmap$included_dea <- factor(sample_data_heatmap$included_dea, level = c("Yes", "No"))

# 3. Heatmap annotations (columns)
heatmap_anno_col <- HeatmapAnnotation(Condition = sample_data_heatmap$condition, DEA = sample_data_heatmap$included_dea, 
                        Study = sample_data_heatmap$study, 
                        col = list(Condition = c("CLL" = "#E6E6E6", "RT" = "#E15759"), 
                                   DEA = c("Yes" = "#FFE099", "No" = "#E6E6E6"), 
                                   Study = c("Penter (2021)" = "#1f77b4", "Nadeu (2022)" = "#ff7f0e", "Rejeski (2022)" = "#9632B8",
                                             "Hing (2023)" = "#D43F3A", "Parry (2023a)" = "#5CB85C", "Parry (2023b)" = "#FFB900",
                                             "Rigo (2024)" = "#46B8DA")
                                   ),
                        show_legend = TRUE, border = TRUE, show_annotation_name = TRUE
                        )

# 4. Heatmap annotations (rows)
deg_data_heatmap <- deseq2_dea.df %>%
  dplyr::filter(direction != "NS") %>%
  dplyr::select(direction) %>%
  tibble::rownames_to_column(var = "gene") %>%
  dplyr::mutate(
    known_deg  = ifelse(gene %in% known_degs, "Yes", "No"),
    common_deg = ifelse(gene %in% common_degs, "Yes", "No"),
    mouse_deg = dplyr::case_when(
      direction == "Up"   & gene %in% all_mouse_up_degs   ~ "Yes",
      direction == "Down" & gene %in% all_mouse_down_degs ~ "Yes",
      TRUE ~ "No" 
    )
  ) %>%
  dplyr::left_join(mouse_expr.df, by = c("gene" = "gene_name")) %>%
  dplyr::mutate(across(ends_with("_direction"), ~tidyr::replace_na(.x, "NA")))

deg_data_heatmap <- deg_data_heatmap[match(rownames(scaled_expr_deg_mat), deg_data_heatmap$gene, ),] 
deg_data_heatmap$direction <- factor(deg_data_heatmap$direction, level = c("Up", "Down"))

heatmap_anno_row <- rowAnnotation(foo = anno_mark(at = which(deg_data_heatmap$gene %in% genes_of_interest), 
                                                  labels = deg_data_heatmap$gene[which(deg_data_heatmap$gene %in% genes_of_interest)],
                                                  labels_gp = gpar(fontsize = 6)))

# 5. Plot the heatmap
heatmap.plt  <- Heatmap(scaled_expr_deg_mat, cluster_rows = TRUE, show_row_dend = FALSE, split = 2, show_row_names = FALSE,
                        cluster_columns = FALSE, column_order = sample_order, top_annotation = heatmap_anno_col, 
                        show_column_names = TRUE, column_split = c(rep(1, 19), rep(2, 15)), column_names_rot = 45,
                        row_title = c(paste0("n=", length(up_degs)), paste0("n=", length(down_degs))), row_title_rot = 0, 
                        border="black", col = colorRamp2(c(-3, -1, 0, 1, 3), c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15")),
                        row_names_gp = gpar(fontsize = 4), heatmap_legend_param = list(title = "Gene z-scores")
                        ) 

heatmap.plt <- heatmap.plt +
  Heatmap(deg_data_heatmap$direction, name = "Direction", width = unit(0.1, "cm"), 
          border = TRUE, col = c("Down"="#343A8F", "Up"="#D63027")) +
  Heatmap(deg_data_heatmap$known_deg, name = "Previously reported DEG", width = unit(0.5, "cm"), border = TRUE, 
          col = c("Yes"="#0B652E", "No"="#FFFFBF")) +
  Heatmap(deg_data_heatmap$common_deg, name = "Common DEG", width = unit(0.5, "cm"), border = TRUE, 
          col = c("Yes"="#0B652E", "No"="#FFFFBF"), right_annotation = heatmap_anno_row)

heatmap.plt
```

```{r, eval=FALSE}
# Save the figure
png(filename = paste0(path_to_save_figure, "/deseq2_heatmap.png"), width=14, height=8, units="in", res=500)
print(heatmap.plt)
dev.off()
```

We also plotted the heatmap for the unique DEGs found only in the integrated data (See in Consistency_bulk_vs_sc_RNAseq section). 

```{r}
# List the unique degs
unique_integrated_up_degs <- c("PDPN", "CYTOR", "LINC00884", "SLC25A4", "AC002467.1", "LRTOMT", "IPO4", "GINS2", "CST3", "GATD3B", "AC133644.2", 
                               "CPVL", "FAH", "C1orf74", "HIST1H3B", "HIST1H4I", "HYKK", "AL355076.2")
unique_integrated_down_degs <- c("LINC01781", "EDRF1", "SENP1", "KANSL2", "UHRF1BP1L", "COG3", "KAT7", "RBM39", "P2RY14", "ZSWIM5")
unique_integrated_degs <- c(unique_integrated_up_degs, unique_integrated_down_degs)

# List the common DEGs between DESeq2 and Seurat (See the next section)
seurat_deseq2_common_degs.df <- read.csv(here::here("scRNAseq/meta_analysis/supplm_file/seurat_deseq2_integrated_common_degs.csv"))
seurat_deseq2_common_degs <- c(seurat_deseq2_common_degs.df$UpDEGs, seurat_deseq2_common_degs.df$DownDEGs)
seurat_deseq2_common_degs <- seurat_deseq2_common_degs[seurat_deseq2_common_degs != ""]
```

```{r, message=FALSE, warning=FALSE, fig.height=6, fig.width=11}
# Plot the heatmap for unique degs
# 1. Extract the unique genes from the main scaled matrix 
sample_data_heatmap_sorted <- sample_data_heatmap[match(sample_order, sample_data_heatmap$samples), ]
unique_deg_mat <- scaled_expr_mat[rownames(scaled_expr_mat) %in% unique_integrated_degs, , drop=FALSE]
unique_deg_mat <- unique_deg_mat[ , sample_order] 

# 2. Annotation
heatmap_anno_col <- HeatmapAnnotation(
  Condition = sample_data_heatmap_sorted$condition, 
  DEA = sample_data_heatmap_sorted$included_dea,
  Study = sample_data_heatmap_sorted$study,
  col = list(
    Condition = c("CLL" = "#E6E6E6", "RT" = "#E15759"),
    DEA = c("Yes" = "#FFE099", "No" = "#E6E6E6"),
    Study = c("Penter (2021)" = "#1f77b4", "Nadeu (2022)" = "#ff7f0e", 
              "Rejeski (2022)" = "#9632B8", "Hing (2023)" = "#D43F3A", 
              "Parry (2023a)" = "#5CB85C", "Parry (2023b)" = "#FFB900",
              "Rigo (2024)" = "#46B8DA")
  ),
  show_legend = TRUE, 
  border = TRUE, 
  show_annotation_name = TRUE
)

unique_deg_data_heatmap <- deg_data_heatmap[match(rownames(unique_deg_mat), deg_data_heatmap$gene), ] %>%
  dplyr::mutate(seurat_deg = ifelse(gene %in% seurat_deseq2_common_degs, "Yes", "No"))
unique_deg_data_heatmap$direction <- factor(unique_deg_data_heatmap$direction, levels = c("Up", "Down"))
unique_deg_data_heatmap$Hacken_direction <- factor(unique_deg_data_heatmap$Hacken_direction, levels = c("Up", "Down", "NA"))

unique_heatmap_anno_row <- rowAnnotation(
  foo = anno_mark(at = 1:nrow(unique_deg_data_heatmap), 
                  labels = unique_deg_data_heatmap$gene,
                  labels_gp = gpar(fontsize = 8))
)

# 3. Plot
n_up_shown <- sum(unique_deg_data_heatmap$direction == "Up", na.rm = TRUE)
n_down_shown <- sum(unique_deg_data_heatmap$direction == "Down", na.rm = TRUE)

unique_degs_heatmap.plt <- Heatmap(
  unique_deg_mat, 
  cluster_rows = TRUE, 
  show_row_dend = FALSE, 
  split = unique_deg_data_heatmap$direction, 
  cluster_row_slices = FALSE,
  show_row_names = FALSE, 
  cluster_columns = FALSE, 
  show_column_names = TRUE, 
  column_split = c(rep(1, 19), rep(2, 15)), 
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 8),
  top_annotation = heatmap_anno_col, 
  row_title = c(paste0("n=", n_up_shown), paste0("n=", n_down_shown)), 
  row_title_rot = 0, 
  border = "black", 
  col = colorRamp2(c(-3, -1, 0, 1, 3), c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15")),
  heatmap_legend_param = list(title = "Gene z-scores")
) 

# 4. Add the annotation bars
dir_cols <- c("Down" = "#343A8F", "Up" = "#D63027", "NA" = "#D0D0D0")
direction_annos <- rowAnnotation(
  "Integrated dataset" = unique_deg_data_heatmap$direction,
  "Hing (2023)"       = unique_deg_data_heatmap$Hing_direction,
  "Ten Hacken (2023)" = unique_deg_data_heatmap$Hacken_direction,
  "Iyer (2024)"       = unique_deg_data_heatmap$Iyer_direction,
  "Rompietti (2025)"  = unique_deg_data_heatmap$Rompietti_direction,
  col = list(
    "Integrated dataset" = dir_cols, "Hing (2023)" = dir_cols, 
    "Ten Hacken (2023)" = dir_cols, "Iyer (2024)" = dir_cols, "Rompietti (2025)" = dir_cols
  ),
  gap = unit(1.5, "mm"),
  show_legend = c(FALSE, FALSE, TRUE, FALSE, FALSE), # ONLY show the first one
  annotation_legend_param = list(
    "Ten Hacken (2023)" = list(title = "Expression Direction")
  ),
  simple_anno_size = unit(0.3, "cm"),
  border = TRUE
)

unique_degs_heatmap.plt <- unique_degs_heatmap.plt + 
                           direction_annos + 
                           unique_heatmap_anno_row

draw(unique_degs_heatmap.plt, merge_legend = TRUE)
```

```{r, eval=FALSE}
# Save the figure
png(filename = paste0(path_to_save_figure, "/novel_and_mouse_degs_heatmap.png"), width=11, height=6, units="in", res=500)
draw(unique_degs_heatmap.plt, merge_legend = TRUE)
dev.off()
```

# Consistenct assessment between DESeq2- and Seurat-based expression analyses

In this section we evaluated the consistency of the gene expression between the two different approaches.

```{r}
# Rename the columns from the seurat results comparable to the DESeq2 form
seurat_dea.df <- seurat_dea.df %>%
      dplyr::rename(
       pvalue = "p_val",
       padj = "p_val_adj",
       log2FoldChange = "avg_log2FC"
      )

# Convert the rowname in DESeq2 result to be a column
deseq2_dea.df <- deseq2_dea.df %>%
  dplyr::mutate(gene = rownames(deseq2_dea.df))
```

## Overlapping between DEGs

```{r}
# Upregulated DEGs
seurat_up_degs <- seurat_dea.df %>% 
  dplyr::filter(!is.na(gene)) %>%
  dplyr::filter(!is.na(padj)) %>%
  dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
  pull(gene)

deseq2_up_degs <- deseq2_dea.df %>% 
  dplyr::filter(!is.na(gene)) %>%
  dplyr::filter(!is.na(padj)) %>%
  dplyr::filter(padj < 0.05, log2FoldChange > 0.25) %>%
  pull(gene)

# Plot a Venn diagram
seurat_deseq2_up_degs_venn.plt <- ggvenn(
  list("DESeq2-based\nupregulated DEGs" = deseq2_up_degs,
       "Seurat-based\nupregulated DEGs" = seurat_up_degs),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5,
  set_name_size = 6,   # Size of the outside labels (Default is 6)
  text_size = 6        # Size of the numbers inside the diagram (Default is 4)
)

seurat_deseq2_up_degs_venn.plt
```

```{r}
# Show the overlapped upregulated DEGs
seurat_deseq2_common_up_degs <- intersect(deseq2_up_degs, seurat_up_degs)
seurat_deseq2_common_up_degs
```

```{r}
# Downregulated DEGs
seurat_down_degs <- seurat_dea.df %>% 
  dplyr::filter(!is.na(gene)) %>%
  dplyr::filter(!is.na(padj)) %>%
  dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
  pull(gene)

deseq2_down_degs <- deseq2_dea.df %>% 
  dplyr::filter(!is.na(gene)) %>%
  dplyr::filter(!is.na(padj)) %>%
  dplyr::filter(padj < 0.05, log2FoldChange < (-0.25)) %>%
  pull(gene)

# Plot a Venn diagram
seurat_deseq2_down_degs_venn.plt <- ggvenn(
  list("DESeq2-based\ndownregulated DEGs" = deseq2_down_degs,
       "Seurat-based\ndownregulated DEGs" = seurat_down_degs),
  fill_color = c("#0073C2FF", "#EFC000FF"),
  stroke_size = 0.5,
  set_name_size = 6,   # Size of the outside labels (Default is 6)
  text_size = 6        # Size of the numbers inside the diagram (Default is 4)
)

seurat_deseq2_down_degs_venn.plt
```

```{r}
# Show the overlapped downregulated DEGs
seurat_deseq2_common_down_degs <- intersect(deseq2_down_degs, seurat_down_degs)
seurat_deseq2_common_down_degs
```

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/seurat_deseq2_up_degs_venn.png"), 
       plot=seurat_deseq2_up_degs_venn.plt, dpi=500, height=4.5, width=7)
ggsave(filename=paste0(path_to_save_figure, "/seurat_deseq2_down_degs_venn.png"), 
       plot=seurat_deseq2_down_degs_venn.plt, dpi=500, height=4.5, width=7)
```

```{r, eval=FALSE}
# Save the results as a csv file
# 1. Pad the lists to the same length with NAs
max_len <- max(length(seurat_deseq2_common_up_degs), length(seurat_deseq2_common_down_degs))
length(seurat_deseq2_common_up_degs) <- max_len 
length(seurat_deseq2_common_down_degs) <- max_len

# 2. Combine into a Data Frame and set your Column Names here
seurat_deseq2_integrated_common_degs <- data.frame(
  "UpDEGs" = seurat_deseq2_common_up_degs,   
  "DownDEGs" = seurat_deseq2_common_down_degs
)

# 3. Save as CSV
write.csv(
  seurat_deseq2_integrated_common_degs, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/seurat_deseq2_integrated_common_degs.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)
```

## Numerical metric assessment

### All significant genes

```{r}
# Generate the log2FC dataframe for both methods
seurat_deseq2_common_sig_genes.df <- list(
  "Seurat approach" = seurat_dea.df, 
  "DESeq2 approach" = deseq2_dea.df
) %>%
  purrr::map_dfr(~ .x, .id = "approach") %>%
  dplyr::filter(padj < 0.05) %>%                       # 1. Filter significant genes first (Efficiency)
  dplyr::group_by(gene) %>%
  dplyr::filter(n_distinct(approach) == 2) %>%         # 2. Keep genes present in BOTH (Safety: use n_distinct)
  dplyr::ungroup() %>%
  dplyr::select(gene, log2FoldChange, approach) %>%
  tidyr::pivot_wider(names_from = approach, values_from = log2FoldChange)

head(seurat_deseq2_common_sig_genes.df)
```

```{r}
# Show the metric results
expr_mat <- seurat_deseq2_common_sig_genes.df %>% 
  tibble::column_to_rownames("gene") %>% 
  as.matrix()

similarity_summary.df <- tibble::tibble(
  Metric = c("Pearson Correlation", "Spearman Correlation", "Cosine Similarity", "Euclidean Similarity"),
  Value = c(
    cor(expr_mat, method = "pearson")[1, 2],
    cor(expr_mat, method = "spearman")[1, 2],
    lsa::cosine(expr_mat)[1, 2],
    1 / (1 + dist(t(expr_mat), method = "euclidean")[1]) # dist returns a single value for 2 cols
  )
)

similarity_summary.df
```

### Upregulated genes

```{r}
# Generate the log2FC dataframe for both methods
seurat_deseq2_common_up_degs.df <- list(
  "Seurat approach" = seurat_dea.df, 
  "DESeq2 approach" = deseq2_dea.df
) %>%
  purrr::map_dfr(~ .x, .id = "approach") %>%
  dplyr::filter(padj < 0.05 & log2FoldChange > 0.25) %>%      # 1. Filter significant genes first (Efficiency)
  dplyr::group_by(gene) %>%
  dplyr::filter(n_distinct(approach) == 2) %>%                # 2. Keep genes present in BOTH (Safety: use n_distinct)
  dplyr::ungroup() %>%
  dplyr::select(gene, log2FoldChange, approach) %>%
  tidyr::pivot_wider(names_from = approach, values_from = log2FoldChange)

head(seurat_deseq2_common_up_degs.df)
```

```{r}
# Show the metric results
expr_mat <- seurat_deseq2_common_up_degs.df %>% 
  tibble::column_to_rownames("gene") %>% 
  as.matrix()

similarity_summary.df <- tibble::tibble(
  Metric = c("Pearson Correlation", "Spearman Correlation", "Cosine Similarity", "Euclidean Similarity"),
  Value = c(
    cor(expr_mat, method = "pearson")[1, 2],
    cor(expr_mat, method = "spearman")[1, 2],
    lsa::cosine(expr_mat)[1, 2],
    1 / (1 + dist(t(expr_mat), method = "euclidean")[1]) # dist returns a single value for 2 cols
  )
)

similarity_summary.df
```

### Downregulated genes

```{r}
# Generate the log2FC dataframe for both methods
seurat_deseq2_common_down_degs.df <- list(
  "Seurat approach" = seurat_dea.df, 
  "DESeq2 approach" = deseq2_dea.df
) %>%
  purrr::map_dfr(~ .x, .id = "approach") %>%
  dplyr::filter(padj < 0.05 & log2FoldChange < (-0.25)) %>%      # 1. Filter significant genes first (Efficiency)
  dplyr::group_by(gene) %>%
  dplyr::filter(n_distinct(approach) == 2) %>%                # 2. Keep genes present in BOTH (Safety: use n_distinct)
  dplyr::ungroup() %>%
  dplyr::select(gene, log2FoldChange, approach) %>%
  tidyr::pivot_wider(names_from = approach, values_from = log2FoldChange)

head(seurat_deseq2_common_down_degs.df)
```

```{r}
# Show the metric results
expr_mat <- seurat_deseq2_common_down_degs.df %>% 
  tibble::column_to_rownames("gene") %>% 
  as.matrix()

similarity_summary.df <- tibble::tibble(
  Metric = c("Pearson Correlation", "Spearman Correlation", "Cosine Similarity", "Euclidean Similarity"),
  Value = c(
    cor(expr_mat, method = "pearson")[1, 2],
    cor(expr_mat, method = "spearman")[1, 2],
    lsa::cosine(expr_mat)[1, 2],
    1 / (1 + dist(t(expr_mat), method = "euclidean")[1]) # dist returns a single value for 2 cols
  )
)

similarity_summary.df
```

# Session Information

```{r}
sessionInfo()
```


















