---
title: "Cell_cell_communication_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

This section we analysed the cell-cell communication between tumour and non-tumour cells for the integrated dataset. We applied CellChat package mainly for the analysis.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tibble)
library(sctransform)
library(cowplot)
library(CellChat)
library(ComplexHeatmap)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_integrated_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered_seu.rds")
path_to_save_myeloid_obj <- here::here("scRNAseq/meta_analysis/R_objects/myeloid_seu.rds")
path_to_save_rt_obj <- here::here("scRNAseq/meta_analysis/R_objects/rt_seu.rds")
path_to_save_cll_obj <- here::here("scRNAseq/meta_analysis/R_objects/cll_seu.rds")
path_to_save_myeloid_markers <- here::here("scRNAseq/meta_analysis/R_objects/myeloid_marker.rds")
path_to_save_cll_cellchat <- here::here("scRNAseq/meta_analysis/R_objects/cll_cellchat.rds")
path_to_save_rt_cellchat <- here::here("scRNAseq/meta_analysis/R_objects/rt_cellchat.rds")
path_to_save_merged_cellchat_list <- here::here("scRNAseq/meta_analysis/R_objects/merged_cellchat_lst.rds")
path_to_save_merged_cellchat <- here::here("scRNAseq/meta_analysis/R_objects/merged_cellchat.rds")
path_to_save_figure <- here::here("figures")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the integrated object
seurat <- readRDS(path_to_integrated_obj)
```

```{r, message=FALSE, warning=FALSE}
# Plot UMAP for cell annotation

cell_to_annotation <- c(
  "B cells" = "B cell",
  "Monocytes" = "Monocyte/Macrophage",
  "NK cells" = "NK cell",
  "T cells, CD4+" = "CD4+ T cell",
  "T cells, CD8+" = "CD8+ T cell"
)

seurat$cell_annotate <- 
  as.vector(cell_to_annotation[as.character(seurat$singleR.labels_main)])

new_levels <- c(
  "B cell",
  "Monocyte/Macrophage",
  "NK cell",
  "CD4+ T cell",
  "CD8+ T cell"
)

seurat$cell_annotate <- factor(
  seurat$cell_annotate,
  levels = new_levels
)

new_colours <- c(
  "#D43F3A", 
  "#EEA236",
  "#5CB85C",
  "#357EBD",
  "#46B8DA"
)

cell.plt <- DimPlot(seurat, 
                    group.by = "cell_annotate", 
                    pt.size = 0.05,
                    raster = FALSE) +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels
  ) +
  labs(y = "UMAP 2", 
       x = "UMAP 1",
       title = "Harmony integration",
       subtitle = "(n=593,424)") +
  theme(
    plot.title = element_text(hjust = 0.5), # Center title
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  coord_fixed(ratio = 1)

cell.plt
```

# Myeloid subtype annotation

From the SingleR package, it annotated only for the main myeloid cells as monocytes. We manually annotated them to monocytes and macrophages with known cell biomarkers.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Subset only myeloid cells with reanalysis
myeloid_seurat <- subset(seurat, subset = singleR.labels_main == "Monocytes")
myeloid_seurat <- SCTransform(myeloid_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
myeloid_seurat <- RunPCA(myeloid_seurat)
myeloid_seurat <- RunUMAP(myeloid_seurat, dims = 1:30, reduction = "harmony")
myeloid_seurat <- FindNeighbors(myeloid_seurat, dims = 1:30, reduction = "harmony")        
myeloid_seurat <- FindClusters(
  myeloid_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Save the object
saveRDS(myeloid_seurat, path_to_save_myeloid_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the myeloid cell clusters
myeloid_seurat <- readRDS(path_to_save_myeloid_obj)

myeloid_cluster.plt <- DimPlot(myeloid_seurat, 
                       reduction = "umap", 
                       group.by = "SCT_snn_res.0.1", 
                       pt.size = 2,
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

myeloid_tissue.plt <- DimPlot(myeloid_seurat, 
                       reduction = "umap", 
                       group.by = "tissue", 
                       pt.size = 2,
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

myeloid_cond.plt <- DimPlot(myeloid_seurat, 
                       reduction = "umap", 
                       group.by = "group", 
                       pt.size = 2,
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

myeloid_cluster.plt + myeloid_tissue.plt + myeloid_cond.plt
```

```{r, fig.height=7, fig.width=10, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(myeloid_seurat, 
            features = c("LYZ", "S100A8", "FCN1", "VCAN", "C1QA", "APOE", "CTSD", "LILRB1", "CD163", "CD68", "TNFSF13", "TNFSF13B"),
            reduction = "umap") 
```

After the cell clustering, we will identified positive biomarkers for each cluster.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Thresholds
min_log2FC <- 0.3
alpha <- 0.001

# Calculate markers
Idents(myeloid_seurat) <- "SCT_snn_res.0.1"
markers <- FindAllMarkers(myeloid_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_myeloid_markers)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_myeloid_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We annotated the myeloid subtypes as literatures below:

- [The monocytic population in chronic lymphocytic leukemia shows altered composition and deregulation of genes involved in phagocytosis and inflammation](https://haematologica.org/article/view/6722)
- [Single-cell transcriptomic analysis of human colonic macrophages reveals niche-specific subsets](https://rupress.org/jem/article/219/3/e20211846/212998/Single-cell-transcriptomic-analysis-of-human)
- [Generation of a blockage monoclonal antibody of LILRB1 against HLA-G](https://www.sciencedirect.com/science/article/abs/pii/S1046592823001341)
- [Human Monocyte Subsets and Phenotypes in Major Chronic Inflammatory Diseases](https://www.frontiersin.org/journals/immunology/articles/10.3389/fimmu.2019.02035/full)
- [Single cell analysis reveals a subset of cytotoxic-like plasmacytoid dendritic cells in people with HIV-1](https://www.sciencedirect.com/science/article/pii/S2589004223017054?via%3Dihub)

| Cluster | Markers                         | Subtype annotation           | CellChat annotation          |
|---------|---------------------------------|------------------------------|------------------------------|
| 0       | C1QA, C1QB, C1QC, LILRB1        | Macrophage                   | Macrophage                   |
| 1       | FCGR3A(CD16)                    | CD16+ Monocyte               | Monocyte                     |
| 2       | CD14, S100A8                    | CD14+ Monocyte               | Monocyte                     |
| 3       | CD14, S100A8                    | CD14+ Monocyte               | Monocyte                     |
| 4       | LILRA4, CLEC4C, IL3RA           | pDC                          | pDC                          |                         
| 5       | CD79A, CD79B, MS4A1             | contaminated B cell          | Ambient cell                 |

```{r, eval=FALSE}
# Label the CellChat annotation
myeloid_seurat$cellchat_annotate <- "Monocyte"
myeloid_seurat$cellchat_annotate[myeloid_seurat$SCT_snn_res.0.1 == 0] <- "Macrophage"
myeloid_seurat$cellchat_annotate[myeloid_seurat$SCT_snn_res.0.1 == 4] <- "pDC"
myeloid_seurat$cellchat_annotate[myeloid_seurat$SCT_snn_res.0.1 == 5] <- "Ambient cell"
```

# Prepare the data for CellChat

After identifying the myeloid subtypes, we mapped the new annotations back to the main Seurat object.

```{r, eval=FALSE}
# Map the myeloid subtypes back to the main object
seurat$cellchat_annotate <- as.character(seurat$cell_annotate)
seurat$cellchat_annotate[seurat$cellchat_annotate == "B cell"] <- "Tumour cell"
seurat$cellchat_annotate[colnames(myeloid_seurat)] <- myeloid_seurat$cellchat_annotate
seurat <- subset(seurat, subset = cellchat_annotate != "Ambient cell")
```

As CellChat prefers the normalised RNA slot, we changed the default assay from SCT to RNA instead. 

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set assay default to RNA and normalise the values
DefaultAssay(seurat) <- "RNA"
Idents(seurat) <- "cellchat_annotate"
seurat <- NormalizeData(seurat, normalization.method="LogNormalize", scale.factor=10000)
```

After that, we split the object into RT and CLL according the CellChat recommendation.

```{r, eval=FALSE}
# Split the object based on the conditions
rt_seurat <- subset(seurat, subset = group == "RT")
cll_seurat <- subset(seurat, subset = group == "CLL")

# Save the objects
saveRDS(rt_seurat, path_to_save_rt_obj)
saveRDS(cll_seurat, path_to_save_cll_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set the ligand-receptor interaction database
future::plan(sequential) # do sequential
CellChatDB <- CellChatDB.human
CellChatDB.use <- subsetDB(CellChatDB, search = "Secreted Signaling", key = "annotation") # use Secreted Signaling
```

# Cell-cell communication analysis with CellChat

For cell–cell interaction analysis, CLL and RT samples were analysed separately as individual objects. The results were subsequently integrated to enable joint visualisation and comparative analysis.

## CLL cell-cell communication

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create cellchat object
data.input <- cll_seurat[["RNA"]]$data
labels <- Idents(cll_seurat)
meta <- data.frame(labels = labels, row.names = names(labels)) # create a dataframe of the cell labels
cellchat <- createCellChat(object = cll_seurat, group.by = "ident", assay = "RNA")
cellchat@DB <- CellChatDB.use

# Preprocessing the expression data for cell-cell communication analysis
cellchat <- subsetData(cellchat) 
#future::plan("multisession", workers = 4) # do parallel
cellchat <- identifyOverExpressedGenes(cellchat, do.fast = FALSE)
cellchat <- identifyOverExpressedInteractions(cellchat)

# Compute the communication probability and infer cellular communication network
cellchat <- computeCommunProb(cellchat, type = "triMean")
cellchat <- filterCommunication(cellchat, min.cells = 10) # filter out few cells in certain cell groups
cellchat <- computeCommunProbPathway(cellchat) # Infer a signaling pathway level
cellchat <- aggregateNet(cellchat) # Calculate the aggregated cell-cell communication network
cellchat <- netAnalysis_computeCentrality(cellchat, slot.name = "netP") # Compute the network centrality scores

# Save the cellchat object
saveRDS(cellchat, path_to_save_cll_cellchat)
```

```{r, message=FALSE, warning=FALSE}
# Import the cellchat object
cellchat <- readRDS(path_to_save_cll_cellchat)
```

```{r, message=FALSE, warning=FALSE}
# Plot the number/weight of interactions
groupSize <- as.numeric(table(cellchat@idents))
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize, weight.scale = TRUE, label.edge= FALSE, title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize, weight.scale = TRUE, label.edge= FALSE, title.name = "Interaction weights/strength")
```

```{r}
# Plot the signaling sending to each cell group
mat <- cellchat@net$weight
par(mfrow = c(2, 3), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```

```{r}
# Plot the signaling received by each cell group
mat <- t(cellchat@net$weight)
par(mfrow = c(2, 3), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```

We showed all ligand-receptor interactions found in all cell types for the CLL conditions.

```{r, fig.height=4, fig.width=11, message=FALSE, warning=FALSE}
# Show all the significant interactions (L-R pairs) 
cll_bubble.plt <- netVisual_bubble(cellchat, remove.isolate = FALSE) +
  labs(x = "Source-target cell type", y = "Ligand-receptor interaction") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(size = 10, margin = margin(t = 12)), 
    axis.title.y = element_text(size = 10, angle = 90, margin = margin(r = 20)), 
    plot.margin = unit(c(1, 1, 1, 1), "cm") # Add space around plot so titles aren't cut off
  ) 

cll_bubble.plt
```

```{r, eval=FALSE}
# Save the bubble plot
ggsave(filename=paste0(path_to_save_figure, "/cellchat_cll_bubble.png"), plot=cll_bubble.plt, dpi=500, height=4, width=11)
```

From the bubble plot, we can see that the tumour cells in the CLL conditions had the autocrine signalling via MIF->(CD74+CXCR4) interaction. The tumour cells also received with the same interaction from NK and T cells. Moreover, the tumour cells were received via LGALS9->CD44 and LGALS9->CD45 interactions from monocytes and macrophages.

For the role as the sender, the tumour cells interacted with macrophages and monocytes via MIF->(CD74+CD44).

```{r, fig.height=8, fig.width=8}
# Plot the chord diagram
cll_chord.plt <- netVisual_chord_gene(cellchat, lab.cex = 0.5,legend.pos.y = 30) 
```

```{r, eval=FALSE}
# Save the chord plot
png(filename = paste0(path_to_save_figure, "/cellchat_cll_chord.png"), 
    width = 8, height = 8, units = "in", res = 500)

par(mar = c(1, 1, 1, 1)) 
netVisual_chord_gene(cellchat, lab.cex = 0.5, legend.pos.y = 30)

dev.off()
```

```{r, fig.height=4, fig.width=5, message=FALSE, warning=FALSE}
# Visualize dominant senders (sources) and receivers (targets) in a 2D space
dominant_2d.plt <- netAnalysis_signalingRole_scatter(cellchat)
dominant_2d.plt
```

From the plot in the CLL condition, we can see that NK and CD4+ T cells contributed the most as the senders while monocytes and macrophages provided the highest role as the receivers.

```{r, fig.height=6, fig.width=10, message=FALSE, warning=FALSE}
# Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
out_heatmap.plt <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing")
in_heatmap.plt <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming")
commun_heatmap.plt <- out_heatmap.plt + in_heatmap.plt
commun_heatmap.plt
```

```{r, eval=FALSE}
# Save the heatmap
png(filename = paste0(path_to_save_figure, "/cellchat_cll_heatmap.png"), width = 10, height = 6, units = "in", res = 500)
draw(commun_heatmap.plt)
dev.off()
```

## RT cell-cell communication

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create cellchat object
data.input <- rt_seurat[["RNA"]]$data
labels <- Idents(rt_seurat)
meta <- data.frame(labels = labels, row.names = names(labels)) # create a dataframe of the cell labels
cellchat <- createCellChat(object = rt_seurat, group.by = "ident", assay = "RNA")
cellchat@DB <- CellChatDB.use

# Preprocessing the expression data for cell-cell communication analysis
cellchat <- subsetData(cellchat) 
#future::plan("multisession", workers = 4) # do parallel
cellchat <- identifyOverExpressedGenes(cellchat, do.fast = FALSE)
cellchat <- identifyOverExpressedInteractions(cellchat)

# Compute the communication probability and infer cellular communication network
cellchat <- computeCommunProb(cellchat, type = "triMean")
cellchat <- filterCommunication(cellchat, min.cells = 10) # filter out few cells in certain cell groups
cellchat <- computeCommunProbPathway(cellchat) # Infer a signaling pathway level
cellchat <- aggregateNet(cellchat) # Calculate the aggregated cell-cell communication network
cellchat <- netAnalysis_computeCentrality(cellchat, slot.name = "netP") # Compute the network centrality scores

# Save the cellchat object
saveRDS(cellchat, path_to_save_rt_cellchat)
```

```{r, message=FALSE, warning=FALSE}
# Import the cellchat object
cellchat <- readRDS(path_to_save_rt_cellchat)
```

```{r, message=FALSE, warning=FALSE}
# Plot the number/weight of interactions
groupSize <- as.numeric(table(cellchat@idents))
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize, weight.scale = TRUE, label.edge= FALSE, title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize, weight.scale = TRUE, label.edge= FALSE, title.name = "Interaction weights/strength")
```

```{r}
# Plot the signaling sending to each cell group
mat <- cellchat@net$weight
par(mfrow = c(2, 3), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```

```{r}
# Plot the signaling received by each cell group
mat <- t(cellchat@net$weight)
par(mfrow = c(2, 3), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```

We showed all ligand-receptor interactions found in all cell types for the CLL conditions.

```{r, fig.height=4, fig.width=11, message=FALSE, warning=FALSE}
# Show all the significant interactions (L-R pairs) 
rt_bubble.plt <- netVisual_bubble(cellchat, remove.isolate = FALSE) +
  labs(x = "Source-target cell type", y = "Ligand-receptor interaction") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(size = 10, margin = margin(t = 12)), 
    axis.title.y = element_text(size = 10, angle = 90, margin = margin(r = 20)), 
    plot.margin = unit(c(1, 1, 1, 1), "cm") # Add space around plot so titles aren't cut off
  ) 

rt_bubble.plt
```

```{r, eval=FALSE}
# Save the bubble plot
ggsave(filename=paste0(path_to_save_figure, "/cellchat_rt_bubble.png"), plot=rt_bubble.plt, dpi=500, height=4, width=11)
```

Compared with CLL, tumour cells in the RT condition exhibited additional autocrine ligand–receptor interactions, including MIF–>(CD74+CD44), LGALS9–>P4HB, LGALS9–>CD44, and LGALS9–>CD45. Moreover, RT tumour cells received additional paracrine interactions from myeloid populations, such as TNFSF13B–>TNFRSF13C from monocytes and macrophages, and LGALS9–>P4HB from monocytes, relative to the CLL group.

As the sender role, the tumour cells interacted with more cell types apart from the myeloid cell types, such as NK and T cells, in comparison with the CLL condition. They interacted those cells via PPIA->BSG, MIF->(CD74+CD44), LGALS9->P4HB, LGALS9->CD44, and LGALS9->CD45.

```{r, fig.height=8, fig.width=8}
# Plot the chord diagram
rt_chord.plt <- netVisual_chord_gene(cellchat, lab.cex = 0.5,legend.pos.y = 30) 
```

```{r, eval=FALSE}
# Save the chord plot
png(filename = paste0(path_to_save_figure, "/cellchat_rt_chord.png"), 
    width = 8, height = 8, units = "in", res = 500)

par(mar = c(1, 1, 1, 1)) 
netVisual_chord_gene(cellchat, lab.cex = 0.5, legend.pos.y = 30)

dev.off()
```

```{r, fig.height=4, fig.width=5, message=FALSE, warning=FALSE}
# Visualize dominant senders (sources) and receivers (targets) in a 2D space
dominant_2d.plt <- netAnalysis_signalingRole_scatter(cellchat)
dominant_2d.plt
```

Compared with the CLL condition, tumour cells in RT assumed the dominant roles as both signal senders and receivers, replacing monocytes/macrophages and CD4⁺ T and NK cells, respectively. This pattern indicates a shift towards enhanced intrinsic signalling mediated by autocrine interactions, rather than signalling through interactions with other cell types.

```{r, fig.height=6, fig.width=10, message=FALSE, warning=FALSE}
# Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
out_heatmap.plt <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing")
in_heatmap.plt <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming")
commun_heatmap.plt <- out_heatmap.plt + in_heatmap.plt
commun_heatmap.plt
```

```{r, eval=FALSE}
# Save the heatmap
png(filename = paste0(path_to_save_figure, "/cellchat_rt_heatmap.png"), width = 10, height = 6, units = "in", res = 500)
draw(commun_heatmap.plt)
dev.off()
```

## Comparative analysis

After we analysed the cell-cell communication in both CLL and RT conditions with the different cellchat objects, we performed the comparative analysis for the interaction in this section.

```{r, eval=FALSE}
# Load the objects and merge into one object
cll_cellchat <- readRDS(path_to_save_cll_cellchat)
rt_cellchat <- readRDS(path_to_save_rt_cellchat)
cellchat.lst <- list(CLL = cll_cellchat, RT = rt_cellchat)
cellchat <- mergeCellChat(cellchat.lst, add.names = names(cellchat.lst))

# Save the files
saveRDS(cellchat.lst, path_to_save_merged_cellchat_list)
saveRDS(cellchat, path_to_save_merged_cellchat)
```

### Number of interactions and interaction strength

```{r}
# Load the merged object and list
cellchat.lst <- readRDS(path_to_save_merged_cellchat_list)
cellchat <- readRDS(path_to_save_merged_cellchat)
cll_cellchat <- cellchat.lst$CLL
rt_cellchat <- cellchat.lst$RT

# Compare the total number of interactions and interaction strength
assigned_colours <- c("#1465AC", "#B31B21")
compare_number.plt <- compareInteractions(cellchat, show.legend = FALSE, group = c(1,2), color.use = assigned_colours)
compare_weight.plt <- compareInteractions(cellchat, show.legend = FALSE, group = c(1,2), measure = "weight", 
                                          color.use = assigned_colours)
compare_number.plt + compare_weight.plt
```

```{r}
# Circle plots for differential number of interactions or interaction strength
par(mfrow = c(1,2), xpd=TRUE)
netVisual_diffInteraction(cellchat, weight.scale = TRUE)
netVisual_diffInteraction(cellchat, weight.scale = TRUE, measure = "weight")
```

For the circle plots representing the number of interactions and interaction strength, the red and blue line indicate the edge for RT and CLL, respectively. In comparison between RT and CLL, the tumour cells had the higher number of interactions and the strength. 

```{r, message=FALSE, warning=FALSE}
# Heatmap for differential number of interactions or interaction strength
compare_number_heatmap.plt <- netVisual_heatmap(cellchat)
compare_weigth_heatmap.plt <- netVisual_heatmap(cellchat, measure = "weight")
compare_heatmap.plt <- compare_number_heatmap.plt + compare_weigth_heatmap.plt
compare_heatmap.plt
```

```{r, eval=FALSE}
# Save the figure
png(filename = paste0(path_to_save_figure, "/cellchat_compare_interaction_heatmap.png"), width = 8, height = 4, units = "in", res = 500)
draw(compare_heatmap.plt)
dev.off()
```

In the heatmaps, colour intensity represents relative changes in both the number of interactions and interaction strength between the RT and CLL conditions. In addition, the top coloured bar plot represents the column-wise sum of the absolute relative values, indicating the overall interaction magnitude for receiver cell types. The coloured bar on the right shows the row-wise summation, representing the overall interaction magnitude for sender cell types.

```{r, message=FALSE, warning=FALSE}
# Circle plots compare the number of interactions 
weight.max <- getMaxWeight(cellchat.lst, attribute = c("idents","count"))
par(mfrow = c(1,2), xpd=TRUE)
for (i in 1:length(cellchat.lst)) {
  netVisual_circle(cellchat.lst[[i]]@net$count, 
                   weight.scale = TRUE, label.edge= FALSE, 
                   edge.weight.max = weight.max[2], edge.width.max = 12, 
                   title.name = paste0("Number of interactions - ", names(cellchat.lst)[i]))
}
```

```{r, message=FALSE, warning=FALSE}
# Circle plots compare the interaction strength 
par(mfrow = c(1,2), xpd=TRUE)
for (i in 1:length(cellchat.lst)) {
  netVisual_circle(cellchat.lst[[i]]@net$weight, 
                   weight.scale = TRUE, label.edge= FALSE, edge.width.max = 12, 
                   title.name = paste0("Interaction strength - ", names(cellchat.lst)[i]))
}
```

For the heatmaps and circle pltos, tumour cells in the RT conditions exhibited higher interaction activity as signal senders towards all non-tumour cell types, as well as towards themselves. As signal receivers, RT tumour cells showed increased interactions with monocytes, CD4⁺ T cells, CD8⁺ T cells, and with themselves.

```{r, message=FALSE, warning=FALSE, fig.height=3, fig.width=8}
# Identify cell populations with significant changes in sending or receiving signals
num.link <- sapply(cellchat.lst, function(x) {rowSums(x@net$count) + colSums(x@net$count)-diag(x@net$count)})
weight.MinMax <- c(min(num.link), max(num.link)) # control the dot size in the different datasets
gg <- list()
for (i in 1:length(cellchat.lst)) {
  gg[[i]] <- netAnalysis_signalingRole_scatter(cellchat.lst[[i]], title = names(cellchat.lst)[i], weight.MinMax = weight.MinMax) +
    xlim(0, 1.1) +  
    ylim(0, 1.1)    
}
patchwork::wrap_plots(plots = gg)
```

Consistent with the previous plots, tumour cells in RT exhibited the most prominent roles as both signal senders and receivers. In contrast, in the CLL condition, monocytes and macrophages predominated as signal senders, whereas NK cells and CD4⁺ T cells were dominant in the receiver role.  

### Signalling comparison

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=14}
# Identify the signaling changes of specific cell populations
assigned_colours <- c("Shared" = "black",           
                      "CLL specific" = "#1465AC",   
                      "RT specific" = "#B31B21")    

tumour_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Tumour cell")
monocyte_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Monocyte")
macrophage_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Macrophage")
pdc_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "pDC")
nk_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "NK cell")
cd4_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "CD4+ T cell")
cd8_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "CD8+ T cell")

all_plots <- list(tumour_signal_diff.plt, 
                  monocyte_signal_diff.plt, 
                  macrophage_signal_diff.plt, 
                  pdc_signal_diff.plt,
                  nk_signal_diff.plt, 
                  cd4_signal_diff.plt, 
                  cd8_signal_diff.plt)

all_plots_coloured <- lapply(all_plots, function(p) {
  p + 
    scale_color_manual(values = assigned_colours) +
    scale_fill_manual(values = assigned_colours)
})

patchwork::wrap_plots(plots = all_plots_coloured)
```

```{r, message=FALSE, warning=FALSE}
# Compare the overall information flow of each signaling pathway or ligand-receptor pair
assigned_colours <- c("#1465AC", "#B31B21")

compare_inform_stacked.plt <- rankNet(cellchat, 
                                      mode = "comparison", 
                                      measure = "weight", 
                                      sources.use = NULL, 
                                      targets.use = NULL, 
                                      stacked = TRUE, 
                                      do.stat = TRUE,
                                      color.use = assigned_colours) +
  theme(legend.position = "bottom")

compare_inform_bar.plt <- rankNet(cellchat, 
                                  mode = "comparison", 
                                  measure = "weight", 
                                  sources.use = NULL, 
                                  targets.use = NULL, 
                                  stacked = FALSE, 
                                  do.stat = TRUE,
                                  color.use = assigned_colours) +
  theme(legend.position = "bottom")

compare_inform_stacked.plt + compare_inform_bar.plt
```

```{r, fig.height=6, fig.width=12, message=FALSE, warning=FALSE}
# Compare outgoing and incoming signalling patterns 
pathway.union <- union(cll_cellchat@netP$pathways, rt_cellchat@netP$pathways)
cll_outgoing_heatmap <- netAnalysis_signalingRole_heatmap(cll_cellchat, pattern = "outgoing", signaling = pathway.union, title = "CLL", width = 5, height = 6)
rt_outgoing_heatmap <- netAnalysis_signalingRole_heatmap(rt_cellchat, pattern = "outgoing", signaling = pathway.union, title = "RT", width = 5, height = 6)
cll_incoming_heatmap <- netAnalysis_signalingRole_heatmap(cll_cellchat, pattern = "incoming", signaling = pathway.union, title = "CLL", width = 5, height = 6)
rt_incoming_heatmap <- netAnalysis_signalingRole_heatmap(rt_cellchat, pattern = "incoming", signaling = pathway.union, title = "RT", width = 5, height = 6)

draw(cll_outgoing_heatmap + cll_incoming_heatmap + rt_outgoing_heatmap + rt_incoming_heatmap, ht_gap = unit(0.5, "cm"))
```

```{r, eval=FALSE}
# Save the heatmap
png(filename = paste0(path_to_save_figure, "/cellchat_compare_signal_heatmap.png"), width = 12, height = 6, units = "in", res = 500)
draw(cll_outgoing_heatmap + cll_incoming_heatmap + rt_outgoing_heatmap + rt_incoming_heatmap, ht_gap = unit(0.5, "cm"))
dev.off()
```

```{r}
# Circle plot for the BAFF signalling in the RT conditions
pathways.show <- c("BAFF") 
par(mfrow=c(1,1))
netVisual_aggregate(rt_cellchat, 
                    signaling = pathways.show, 
                    layout = "circle", 
                    signaling.name = pathways.show)
```

The results showed increased BAFF signalling in the RT condition, with tumour cells acting as receivers and monocytes and macrophages as the primary senders.

### Ligand-receptor analysis

```{r, fig.height=4, fig.width=17, message=FALSE, warning=FALSE}
# Identify dysfunctional signaling by comparing the communication probabities 
compare_prob_bubble.plt <- netVisual_bubble(cellchat, 
                                            comparison = c(1, 2), 
                                            remove.isolate = TRUE)

x_axis_labels <- levels(compare_prob_bubble.plt$data$source.target)
axis_colours <- rep("black", length(x_axis_labels))
axis_colours[str_detect(x_axis_labels, "\\(CLL\\)")] <- "#1465AC"
axis_colours[str_detect(x_axis_labels, "\\(RT\\)")] <- "#B31B21"

compare_prob_bubble.plt <- compare_prob_bubble.plt +
  labs(x = "Source-target cell type", y = "Ligand-receptor interaction") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, colour = axis_colours),
    axis.title.x = element_text(size = 10, margin = margin(t = 12)), 
    axis.title.y = element_text(size = 10, angle = 90, margin = margin(r = 20)), 
    plot.margin = unit(c(1, 1, 1, 1), "cm") # Add space around plot so titles aren't cut off
  ) 

compare_prob_bubble.plt
```

```{r, fig.height=4, fig.width=11, message=FALSE, warning=FALSE}
# Identify dysfunctional signaling by comparing the communication probabities of tumour->non-tumour and non-tumour->tumour
compare_tumour_prob_bubble.plt <- netVisual_bubble_custom(cellchat, 
                                                          comparison = c(1, 2), 
                                                          focus.cell = "Tumour cell", 
                                                          remove.isolate = TRUE)

compare_tumour_prob_bubble.plt$data <- compare_tumour_prob_bubble.plt$data[!is.na(compare_tumour_prob_bubble.plt$data$prob), ]
compare_tumour_prob_bubble.plt$data$source.target <- droplevels(compare_tumour_prob_bubble.plt$data$source.target)
x_axis_labels <- levels(compare_tumour_prob_bubble.plt$data$source.target)
axis_colours <- rep("black", length(x_axis_labels))
axis_colours[str_detect(x_axis_labels, "\\(CLL\\)")] <- "#1465AC"
axis_colours[str_detect(x_axis_labels, "\\(RT\\)")] <- "#B31B21"

compare_tumour_prob_bubble.plt <- compare_tumour_prob_bubble.plt + 
  labs(x = "Source-target cell type", y = "Ligand-receptor interaction") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, colour = axis_colours),
        axis.title.x = element_text(size = 10, margin = margin(t = 12)), 
        axis.title.y = element_text(size = 10, angle = 90, margin = margin(r = 20)), 
        plot.margin = unit(c(1, 1, 1, 1), "cm")) # Add space around plot so titles aren't cut off

compare_tumour_prob_bubble.plt
```

```{r, eval=FALSE}
# Save the bubble plots
ggsave(filename=paste0(path_to_save_figure, "/cellchat_compare_all_bubble.png"), 
       plot=compare_prob_bubble.plt, dpi=500, height=4, width=17)
ggsave(filename=paste0(path_to_save_figure, "/cellchat_compare_tumour_bubble.png"), 
       plot=compare_tumour_prob_bubble.plt, dpi=500, height=4, width=11)
```

The bubble plot showed that, in the RT condition, tumour cells exhibited additional ligand–receptor interactions both with themselves (autocrine signalling) and with non-tumour cells. For example, the TNFSF13B->TNFRSF13C interaction was found in monocytes/macrophages -> tumour cells. LGALS9->(P4HB/CD44/CD45) was also identified in tumour->tumour and tumour->non-tumour communication.

```{r, fig.height=6, fig.width=7, message=FALSE, warning=FALSE}
# Identify dysfunctional signaling from and to tumour
group_tumour <- "Tumour cell"  
all_groups <- levels(cellchat@idents)
group_others <- setdiff(all_groups, group_tumour)

plot_outgoing <- netVisual_bubble(cellchat, 
                                  sources.use = group_tumour, 
                                  targets.use = all_groups, 
                                  comparison = c(1, 2), 
                                  remove.isolate = TRUE) +
                  labs(title = "Signaling from Tumour") + 
                  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

plot_incoming <- netVisual_bubble(cellchat, 
                                  sources.use = group_others, 
                                  targets.use = group_tumour, 
                                  comparison = c(1, 2), 
                                  remove.isolate = TRUE) +
                  labs(title = "Signaling to Tumour") + 
                  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

tumour_bubble.plt <- plot_outgoing / plot_incoming
tumour_bubble.plt
```

```{r, fig.height=8, fig.width=20, message=FALSE, warning=FALSE}
# Identify the up-regulated (increased) and down-regulated (decreased) signaling ligand-receptor pairs
LR_up_prob_bubble.plt <- netVisual_bubble(cellchat, 
                                          max.dataset = 2,
                                          comparison = c(1, 2),
                                          title.name = "Increased signaling in RT", 
                                          angle.x = 45, 
                                          remove.isolate = TRUE)

LR_down_prob_bubble.plt <- netVisual_bubble(cellchat, 
                                            max.dataset = 1,
                                            comparison = c(1, 2),
                                            title.name = "Decreased signaling in RT", 
                                            angle.x = 45, 
                                            remove.isolate = TRUE)

LR_up_prob_bubble.plt / LR_down_prob_bubble.plt
```

```{r, message=FALSE, warning=FALSE}
# Calculate differential expression analysis between RT and CLL for each cell type
pos.dataset = "RT"
features.name = paste0(pos.dataset, ".merged")
cellchat <- identifyOverExpressedGenes(cellchat, 
                                       group.dataset = "datasets", 
                                       pos.dataset = pos.dataset, 
                                       features.name = features.name, 
                                       only.pos = FALSE, 
                                       thresh.pc = 0.1, thresh.fc = 0.05,thresh.p = 0.05, 
                                       do.fast = FALSE,
                                       group.DE.combined = FALSE) 
net <- netMappingDEG(cellchat, features.name = features.name, variable.all = TRUE)
net.up <- subsetCommunication(cellchat, net = net, datasets = "RT",ligand.logFC = 0.05, receptor.logFC = NULL)
net.down <- subsetCommunication(cellchat, net = net, datasets = "CLL",ligand.logFC = -0.05, receptor.logFC = NULL)
```

```{r, eval=FALSE}
# Save the ligand-receptor interaction results as the csv files
write.csv(
  net.up, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/cellchat_rt_up_lr.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)

write.csv(
  net.down, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/cellchat_rt_down_lr.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)
```

```{r, fig.height=8, fig.width=20, message=FALSE, warning=FALSE}
# Plot the bubble plot 
pairLR.use.up = net.up[, "interaction_name", drop = FALSE]
LR_up_dea_bubble.plt <- netVisual_bubble(cellchat, 
                        pairLR.use = pairLR.use.up, 
                        comparison = c(1,2),
                        angle.x = 45, 
                        remove.isolate = TRUE, 
                        title.name = "Up-regulated signaling in RT")
                        
pairLR.use.down = net.down[, "interaction_name", drop = FALSE]
LR_down_dea_bubble.plt <- netVisual_bubble(cellchat, 
                        pairLR.use = pairLR.use.down, 
                        comparison = c(1,2),
                        angle.x = 45, 
                        remove.isolate = TRUE, 
                        title.name = "Down-regulated signaling in RT")

LR_up_dea_bubble.plt / LR_down_dea_bubble.plt
```

```{r, message=FALSE, warning=FALSE}
# Violin plot ligand-receptor expression in BAFF signalling
cellchat@meta$datasets = factor(cellchat@meta$datasets, levels = c("CLL", "RT")) # set factor level
plotGeneExpression(cellchat, signaling = "BAFF", split.by = "datasets", colors.ggplot = TRUE, type = "violin")
```

# Session Information

```{r}
sessionInfo()
```









