---
title: "Tumour_subtype_annotation"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

After performing level-1 cell-type annotation using the SingleR package on the integrated dataset, level-2 annotation of tumour subtypes was carried out based on previously reported cell types and established biomarker genes from the literature.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tibble)
library(SingleR)
library(celldex)
library(sctransform)
library(cowplot)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_tumour_obj <- here::here("scRNAseq/meta_analysis/R_objects/tumour_seu.rds")
path_to_save_markers <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/tumour_markers.rds")
path_to_save_annotated_tumour_obj <- here::here("scRNAseq/meta_analysis/R_objects/annotated_tumour_seu.rds")
path_to_save_figure <- here::here("figures")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the tumour object
tumour_seurat <- readRDS(path_to_tumour_obj)
```

# Cluster visualisation

As Louvain clustering was performed on the tumour cell subset, the resulting clustering structure was visualised prior to the identification of cluster-specific marker genes.

```{r, fig.height=8, fig.width=28}
# Plot the cluster
cluster.plt <- DimPlot(tumour_seurat, 
                       reduction = "umap", 
                       group.by = "SCT_snn_res.0.1", 
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

study.plt <- DimPlot(tumour_seurat, 
                       reduction = "umap", 
                       group.by = "ref", 
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

case.plt <- DimPlot(tumour_seurat, 
                       reduction = "umap", 
                       group.by = "patient", 
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

cluster.plt + study.plt + case.plt
```

```{r, fig.height=8, fig.width=16, message=FALSE, warning=FALSE}
# Plot clusters by splitting the condition
cluster_split.plt <- DimPlot(tumour_seurat, 
                             reduction = "umap", 
                             group.by = "SCT_snn_res.0.1", 
                             split.by = "group", 
                             label = TRUE) +
  NoLegend()
cluster_split.plt 
```

From the plot, we used the clustering results at the resolution = 0.1 to calculate cluster-specific marker genes.

# Clustering-specific marker gene identification

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Thresholds
min_log2FC <- 0.25
alpha <- 0.001

# Calculate markers
Idents(tumour_seurat) <- "SCT_snn_res.0.1"
markers <- FindAllMarkers(tumour_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_markers)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

# Cell cycle scoring

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=22}
# Consider cell cycle genes based on cell cycle upregulation is a hallmark of Richter transformation
tumour_seurat <- CellCycleScoring(
  tumour_seurat,
  s.features = cc.genes$s.genes,
  g2m.features = cc.genes$g2m.genes,
  set.ident = FALSE
)

cell_cycle.plt <- DimPlot(tumour_seurat, group.by = "Phase", pt.size = 1) + labs(y= "UMAP 2", x = "UMAP 1")

umap_s_score <- FeaturePlot(tumour_seurat, features = "S.Score", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "S Score", y= "UMAP 2", x = "UMAP 1") 
  
umap_g2m_score <- FeaturePlot(tumour_seurat, features = "G2M.Score", pt.size = 1) +
  scale_color_viridis_c(option = "magma") +
  labs(title = "G2M Score", y= "UMAP 2", x = "UMAP 1") 

cell_cycle.plt + umap_s_score + umap_g2m_score
```

# Subtype annotation

We annotated tumour subtypes as literatures below:

- [Detection of early seeding of Richter transformation in chronic lymphocytic leukemia](https://www.nature.com/articles/s41591-022-01927-8#code-availability)
- [Low BACH2 Expression Predicts Adverse Outcome in Chronic Lymphocytic Leukaemia](https://www.mdpi.com/2072-6694/14/1/23)
- [Transcriptomic clustering of chronic lymphocytic leukemia: molecular subtypes based on Bruton’s tyrosine kinase expression levels](https://www.nature.com/articles/s41408-024-01196-3)
- [MYC target gene activation in chronic lymphocytic leukemia and richter transformation: links to aggressiveness and tumor microenvironment interactions](https://www.frontiersin.org/journals/pharmacology/articles/10.3389/fphar.2025.1642458/full)
- [Molecular map of chronic lymphocytic leukemia and its impact on outcome](https://www.nature.com/articles/s41588-022-01140-w)
- [Subgroup-specific gene expression profiles and mixed epistasis in chronic lymphocytic leukemia](https://haematologica.org/article/view/haematol.2022.281869)
- [MicroRNA-155 influences B-cell receptor signaling and associates with aggressive disease in chronic lymphocytic leukemia](https://www.sciencedirect.com/science/article/pii/S0006497120399365#:~:text=CLL%20with%20high%2Dlevel%20miR,outcome%20in%20patients%20with%20CLL.)
- [SLC22A5 (OCTN2) Carnitine Transporter—Indispensable for Cell Metabolism, a Jekyll and Hyde of Human Cancer](https://www.mdpi.com/1420-3049/25/1/14)
- [Regulatory network of BLIMP1, IRF4, and XBP1 triad in plasmacytic differentiation and multiple myeloma pathogenesis](https://www.sciencedirect.com/science/article/abs/pii/S0008874922001198?via%3Dihub)
- [Temporal Tracking of Plasma Cells in vivo Using J-chain CreERT2 Reporter System](https://www.biorxiv.org/content/10.1101/2023.12.02.569736v1)
- [ELL2 Influences Transcription Elongation, Splicing, Ig Secretion and Growth](https://www.omicsonline.org/open-access/ell2-influences-transcription-elongation-splicing-ig-secretion-and-growth-109185.html)

Moreover, RT cells were also annotated according to their proliferative status based on the expression levels of cell-proliferation–related genes. Quiescent RT was defined by the expression of RT-associated genes, such as members of the MYC family and PDCD1, in the absence of detectable expression of cell-proliferation–related genes.

| Cluster | Markers                                                       | Annotation                             |
|---------|---------------------------------------------------------------|----------------------------------------|
| 0       | CD27                                                          | CXCR4loCD27hi CLL                      |  
| 1       | TCL1A, BACH2                                                  | quiescent CLL                          |  
| 2       | ENO1, TCL1A, PDCD1, CCND2                                     | CCND2hi RT                             |  
| 3       | PCNA, ENO1, TCL1A                                             | quiescent_to_low-proliferative RT      |  
| 4       | MYC, MYCBP, ENO1, BACH2, LDHA                                 | low-proliferative RT                   |  
| 5       | MYC, MYCBP2, MIR155HG, BCAT1, CCND2                           | BCAT1hiCCND2hi RT                      | 
| 6       | ENO1, MIR155HG, LDHA, BACH2                                   | quiescent RT                           | 
| 7       | MYCBP2                                                        | quiescent RT                           |  
| 8       | SLC22A5                                                       | SLC22A5hi CLL                          |
| 9       | ELL2, IRF4                                                    | plasma cell                            | 
| 10      | PCNA, MYCBP, ENO1, PDCD1, LDHA, CCND2                         | intermediate-proliferative RT          | 
| 11      | TOP2A, PCNA, MYCBP, LDHA, CCND2                               | intermediate_to_high-proliferative RT  | 
| 12      | TOP2A, PCNA, MYC, MYCBP, LDHA                                 | intermediate_to_high-proliferative RT  | 
| 13      | ELL2, XBP1, JCHAIN                                            | plasma cell                            | 
| 14      | TOP2A, PCNA, MKI67, MYC, MYCBP, MIR155HG, LDHA, PDCD1, CCND2  | high-proliferative RT                  | 
| 15      | TOP2A, MKI67, MYCBP, MIR155HG, LDHA, PDCD1, CCND2             | high-proliferative RT                  | 
| 16      | CXCR4                                                         | CXCR4hiCD27lo CLL                      | 
| 17      | CXCR4                                                         | CXCR4hiCD27lo CLL                      | 
| 18      | CXCR4                                                         | CXCR4hiCD27lo CLL                      | 
| 19      | CXCR4                                                         | CXCR4hiCD27lo CLL                      | 
| 20      | RSPH9, RSPH10B, CCDC73, DNAH11                                | ciliated cell                          | 
| 21      | CD3D, CD3E, Nkg7                                              | non-B cell lymphocyte                  | 
| 22      | CD3D, CD3E, Nkg7                                              | non-B cell lymphocyte                  | 

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=11}
# Plot UMAP for tumour subtypes

# 1. Map Seurat clusters and create the annotation column
cluster_to_annotation <- c(
  "0" = "CXCR4loCD27hi CLL", "1" = "quiescent CLL", "2" = "CCND2hi RT", "3" = "quiescent_to_low-proliferative RT", "4" = "low-proliferative RT",
  "5" = "BCAT1hiCCND2hi RT", "6" = "quiescent RT", "7" = "quiescent RT", "8" = "SLC22A5hi CLL", "9" = "plasma cell", 
  "10" = "intermediate-proliferative RT", "11" = "intermediate_to_high-proliferative RT", "12" = "intermediate_to_high-proliferative RT", 
  "13" = "plasma cell", "14" = "high-proliferative RT", "15" = "high-proliferative RT", "16" = "CXCR4hiCD27lo CLL", 
  "17" = "CXCR4hiCD27lo CLL", "18" = "CXCR4hiCD27lo CLL", "19" = "CXCR4hiCD27lo CLL", "20" = "ciliated cell", 
  "21" = "non-B cell lymphocyte", "22" = "non-B cell lymphocyte"
)



tumour_seurat$tumour_annotate <- 
  as.vector(cluster_to_annotation[as.character(tumour_seurat$SCT_snn_res.0.1)])

# 2. Set level order
new_levels <- c(
  # Group1: indolent/resting CLL  
  "CXCR4hiCD27lo CLL",      # Homing to lymph node (oldest/resting)
  "quiescent CLL",          # Stable disease
  "CXCR4loCD27hi CLL",      # Recently born/exiting lymph node
  "SLC22A5hi CLL",          # Distinct metabolic subclone 

  # Group2: The RT trajectory (metabolic priming -> proliferation) 
  "quiescent RT",                          # The "seed" (non-cycling, transformed)
  "CCND2hi RT",                            # G1 priming (cyclin D2 upregulation)
  "BCAT1hiCCND2hi RT",                     # Metabolic switch (Warburg effect active)
  "quiescent_to_low-proliferative RT",     # S-phase entry transition
  "low-proliferative RT",                  # Early S-phase
  "intermediate-proliferative RT",         # Mid S-phase
  "intermediate_to_high-proliferative RT", # Late S/G2 phase
  "high-proliferative RT",                 # M-phase (mitosis/explosion)

  # Group3: differentiation endpoint 
  "plasma cell",            # Terminal differentiation 

  # Group4: microenvironment/contaminated cells 
  "non-B cell lymphocyte",  # T-cells/NK cells
  "ciliated cell"           # Normal epithelial contaminant
)

tumour_seurat$tumour_annotate <- factor(
  tumour_seurat$tumour_annotate,
  levels = new_levels
)

# 3. Define colours and labels
new_colours <- c(
  "#7AC7E2", "#0084A7", "#5489C3", "#7C8FE1", # Group 1 (4 colors)
  "#FCFFC9", "#E8C36A", "#FFCAAE", "#FF6800", "#E25644", "#ED564E", "#d65c7a", "#BB777A", # Group 2 (8 colors)
  "#D5ADDD", # Group 3 (1 color)
  "#B1B1B1", "#A1A1A1" # Group 4 (2 colors)
)

custom_labels <- list(
  bquote(CXCR4^hi~CD27^lo~"CLL"), bquote("quiescent CLL"), bquote(CXCR4^lo~CD27^hi~"CLL"), bquote(SLC22A5^hi~"CLL"), 
  bquote("quiescent RT"), bquote(CCND2^hi~"RT"), bquote(BCAT1^hi~CCND2^hi~"RT"), bquote("quiescent_to_low-proliferative RT"), 
  bquote("low-proliferative RT"), bquote("intermediate-proliferative RT"), bquote("intermediate_to_high-proliferative RT"), 
  bquote("high-proliferative RT"), bquote("plasma cell"), bquote("non-B cell lymphocyte"), bquote("ciliated cell")
)

# 4. Plot UMAP
tumour_subtype.plt <- DimPlot(tumour_seurat, 
                              group.by = "tumour_annotate", 
                              label = TRUE,
                              repel = TRUE,
                              pt.size = 0.1,
                              alpha = 0.5,
                              raster = FALSE
                              ) +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels,
    labels = custom_labels
  ) +
  labs(y = "UMAP 2", x = "UMAP 1") +
  ggtitle("Tumour subtypes")

tumour_subtype.plt
```

It is worth noting that we were unable to merge all tumour cells into a small number of clusters, even after applying batch-correction methods due to the high inter-patient heterogeneity of tumour cells.

# Identify the rank-based consistent common DEG expressions

Moreover, biomarkers identified through our systematic review, derived from rank-based consistently common DEGs, were used to evaluate gene expression patterns in the tumour subtypes.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=25}
# Dot plot for common DEGs based on annotated cells

path_to_common_up_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/3.upregulated_union_rp_rs_consist_degs.txt")
path_to_common_down_degs <- here::here("scRNAseq/systematic_review/6-consistency_assessment/overlapping_genes/seurat/3.downregulated_union_rp_rs_consist_degs.txt")

up_genes <- read.table(path_to_common_up_degs, stringsAsFactors = FALSE, header = FALSE)[, 1]
down_genes <- read.table(path_to_common_down_degs, stringsAsFactors = FALSE, header = FALSE)[, 1]
features <- c(up_genes, down_genes)
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

common_degs.plt <- DotPlot(tumour_seurat, features=features, group.by = "tumour_annotate") +
  RotatedAxis() +
  scale_colour_gradientn(colours = colour_pal) +
  scale_y_discrete(
      limits = rev(new_levels),
      labels = rev(custom_labels)
    ) +
  labs(
    colour = "Average\nExpression",
    size = "Percent\nExpressed"
  )

common_degs.plt
```

From the expression results, the tumour subtype annotation was consistent with the rank-based common DEGs from the indiviudal analyses. We saved the tumour object and subtype annotation results for further analyses.

```{r, eval=FALSE}
# Save the annotation results
saveRDS(tumour_seurat, path_to_save_annotated_tumour_obj)

# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/tumour_subtype_umap.png"), plot=tumour_subtype.plt, dpi=500, height=8, width=11)
ggsave(filename=paste0(path_to_save_figure, "/tumour_subtype_dot.png"), plot=common_degs.plt, dpi=500, height=8, width=25)
```

# Session Information

```{r}
sessionInfo()
```
