---
title: "Microenvironment_monocyte_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

In this analysis, we focused on monocytes and other microenvironmental cells from the integrated dataset by performing pseudobulk analysis between RT and CLL in this cell type. Moreover, we also annotated level-2 subtypes of these cells.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tibble)
library(SingleR)
library(celldex)
library(sctransform)
library(cowplot)
library(DT)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_microenvir_obj <- here::here("scRNAseq/meta_analysis/R_objects/microenvir_seu.rds")
path_to_save_mn_obj <- here::here("scRNAseq/meta_analysis/R_objects/mn_seu.rds")
path_to_save_filtered_mn_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered_mn_seu.rds")
path_to_save_filtered_mn_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered_mn_marker.rds")
path_to_save_filtered2_mn_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered2_mn_seu.rds")
path_to_save_filtered2_mn_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered2_mn_marker.rds")
path_to_save_filtered3_mn_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered3_mn_seu.rds")
path_to_save_filtered3_mn_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered3_mn_marker.rds")
path_to_save_filtered4_mn_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered4_mn_seu.rds")
path_to_save_filtered4_mn_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered4_mn_marker.rds")
path_to_save_filtered5_mn_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered5_mn_seu.rds")
path_to_save_filtered5_mn_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered5_mn_marker.rds")
path_to_save_filtered6_mn_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered6_mn_seu.rds")
path_to_save_filtered6_mn_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered6_mn_marker.rds")
path_to_save_pure_mn_obj <- here::here("scRNAseq/meta_analysis/R_objects/pure_mn_seu.rds")
path_to_save_pure_mn_markers <- here::here("scRNAseq/meta_analysis/R_objects/pure_mn_marker.rds")
path_to_save_figure <- here::here("figures")
```

```{r}
# Thresholds
min_log2FC <- 0.25
alpha <- 0.001
```

```{r}
# Set seed for reproducibility
set.seed(1)
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the microenvironment object
microenvir_seurat <- readRDS(path_to_microenvir_obj)
```

```{r, message=FALSE, warning=FALSE}
# Plot UMAP for cell annotation

cell_to_annotation <- c(
  "Monocytes" = "Monocyte/Macrophage",
  "NK cells" = "NK cell",
  "T cells, CD4+" = "CD4+ T cell",
  "T cells, CD8+" = "CD8+ T cell"
)

microenvir_seurat$cell_annotate <- 
  as.vector(cell_to_annotation[as.character(microenvir_seurat$singleR.labels_main)])

new_levels <- c(
  "Monocyte/Macrophage",
  "NK cell",
  "CD4+ T cell",
  "CD8+ T cell"
)

microenvir_seurat$cell_annotate <- factor(
  microenvir_seurat$cell_annotate,
  levels = new_levels
)

new_colours <- c(
  "#EEA236",
  "#5CB85C",
  "#357EBD",
  "#46B8DA"
)

cell.plt <- DimPlot(microenvir_seurat, 
                    group.by = "cell_annotate", 
                    pt.size = 0.05,
                    raster = FALSE) +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels
  ) +
  labs(y = "UMAP 2", 
       x = "UMAP 1",
       title = "Microenvironmental cells",
       subtitle = "(n=192,550)") +
  theme(
    plot.title = element_text(hjust = 0.5), # Center title
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  coord_fixed(ratio = 1)

cell.plt
```

# Extract monocytes

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Extract monocytes with reanalysis
mn_seurat <- subset(microenvir_seurat, subset = singleR.labels_main == "Monocytes")
mn_seurat <- SCTransform(mn_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
mn_seurat <- RunPCA(mn_seurat)
mn_seurat <- RunUMAP(mn_seurat, dims = 1:30, reduction = "harmony")
mn_seurat <- FindNeighbors(mn_seurat, dims = 1:30, reduction = "harmony")        
mn_seurat <- FindClusters(
  mn_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Save the object
saveRDS(mn_seurat, path_to_save_mn_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the myeloid cell clusters
mn_seurat <- readRDS(path_to_save_mn_obj)

group_by <- c("SCT_snn_res.0.1", "tissue", "group")
umap_plots <- lapply(group_by, function (group){
  DimPlot(mn_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 2,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

mn_umap.plt <- wrap_plots(umap_plots, ncol=3)
mn_umap.plt
```

# Filtering out contaminated cells

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on cell biomarker expression
FeaturePlot(mn_seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

As we detected contaminated B cells (cluster 5 at SCT_snn_res.0.1), we removed them and reprocessed after filtering.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Filter out B cells with reprocessing
filtered_mn_seurat <- subset(mn_seurat, subset = SCT_snn_res.0.1 != 5)
filtered_mn_seurat <- SCTransform(filtered_mn_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered_mn_seurat <- RunPCA(filtered_mn_seurat)
filtered_mn_seurat <- RunUMAP(filtered_mn_seurat, dims = 1:30, reduction = "harmony")
filtered_mn_seurat <- FindNeighbors(filtered_mn_seurat, dims = 1:30, reduction = "harmony")        
filtered_mn_seurat <- FindClusters(
  filtered_mn_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Save the object
saveRDS(filtered_mn_seurat, path_to_save_filtered_mn_obj)
```

We also identified cell biomarkers for further contaminated cell removal.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the clusters
filtered_mn_seurat <- readRDS(path_to_save_filtered_mn_obj)
group_by <- c("SCT_snn_res.0.7", "SCT_snn_res.0.8", "SCT_snn_res.0.9")
umap_plots <- lapply(group_by, function (group){
  DimPlot(filtered_mn_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 2,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

filtered_mn_umap.plt <- wrap_plots(umap_plots, ncol=3)
filtered_mn_umap.plt
```

After the cluster visualisation, we identified positive biomarkers for each cluster based on `SCT_snn_res.0.9`.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(filtered_mn_seurat) <- "SCT_snn_res.0.9"
markers <- FindAllMarkers(filtered_mn_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered_mn_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered_mn_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered_mn_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

Based on the marker gene analysis, we identified 20 clusters containing contaminating cell populations. Clusters 5, 14, 17, 21, and 23 expressed B-cell and plasma-cell markers, including IGKV, IGHV, TNFRSF17 (BCMA), IGHG1, and IGKV + TRBV.

Clusters 3 and 10 showed upregulation of CD247 (CD3ζ), KLRB1 (CD161), FGFBP2, and CAMK4, indicating T-cell and NK-cell identities. Clusters 0 and 1 expressed FCER1A, CD1C, CLEC10A, and NRIP3, characteristic of dendritic cells. Additional clusters (20, 22, and 26) exhibited atypical expression patterns, including epithelial markers such as KRT5 or cell-cycle–associated genes.

These contaminating clusters were excluded from downstream analyses, retaining only clusters 2, 4, 6, 7, 8, 9, 11, 12, 13, 15, 16, 19, 24, and 25 for further investigation.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the monocyte clusters with reprocessing
filtered2_mn_seurat <- subset(filtered_mn_seurat, idents = c(2, 4, 6, 7, 8, 9, 11, 12, 13, 15, 16, 19, 24, 25))
filtered2_mn_seurat <- SCTransform(filtered2_mn_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered2_mn_seurat <- RunPCA(filtered2_mn_seurat)
filtered2_mn_seurat <- RunUMAP(filtered2_mn_seurat, dims = 1:30, reduction = "harmony")
filtered2_mn_seurat <- FindNeighbors(filtered2_mn_seurat, dims = 1:30, reduction = "harmony")        
filtered2_mn_seurat <- FindClusters(
  filtered2_mn_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 1, by = 0.1)
)

# Save the object
saveRDS(filtered2_mn_seurat, path_to_save_filtered2_mn_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
filtered2_mn_seurat <- readRDS(path_to_save_filtered2_mn_obj)
Idents(filtered2_mn_seurat) <- "SCT_snn_res.1"
markers <- FindAllMarkers(filtered2_mn_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered2_mn_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered2_mn_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered2_mn_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We still found non-monocytic cells in 4 clusters: Cluster 10 and 19 (dendritic cells, markers: CH25H, FLT3, FCER1A, NDRG2, and MRC1; Cluster 3, 5, 7, 8, 13, and 22 (T cells/NK cells/B-Cells), markers: TRBV, IGKV, IL32, IGKV, IGLV, IGHG1, and TNFRSF13B; and Cluster 6 (platelets), markers: PPBP and PF4.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the monocyte clusters with reprocessing
filtered3_mn_seurat <- subset(filtered2_mn_seurat, idents = c(0, 1, 2, 4, 9, 10, 11, 12, 14, 15, 16, 17, 18, 20, 21, 23))
filtered3_mn_seurat <- SCTransform(filtered3_mn_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered3_mn_seurat <- RunPCA(filtered3_mn_seurat)
filtered3_mn_seurat <- RunUMAP(filtered3_mn_seurat, dims = 1:30, reduction = "harmony")
filtered3_mn_seurat <- FindNeighbors(filtered3_mn_seurat, dims = 1:30, reduction = "harmony")        
filtered3_mn_seurat <- FindClusters(
  filtered3_mn_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 1, by = 0.1)
)

# Save the object
saveRDS(filtered3_mn_seurat, path_to_save_filtered3_mn_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
filtered3_mn_seurat <- readRDS(path_to_save_filtered3_mn_obj)
Idents(filtered3_mn_seurat) <- "SCT_snn_res.0.3"
markers <- FindAllMarkers(filtered3_mn_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered3_mn_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered3_mn_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered3_mn_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

Clusters for contaminated cells were still detected: Cluster 5 (dendritic cells/granulocytes), markers: CH25H, FLT3, PADI4, and BPI; Cluster 7 (platelets/T cells), markers: IGKV, TRBC1, PPBP, and PF4; and Cluster 3 (T cells/monocytes), markers: CD2, CD163, and TRBC.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the monocyte clusters with reprocessing
filtered4_mn_seurat <- subset(filtered3_mn_seurat, idents = c(0, 1, 2, 4, 6, 8))
filtered4_mn_seurat <- SCTransform(filtered4_mn_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered4_mn_seurat <- RunPCA(filtered4_mn_seurat)
filtered4_mn_seurat <- RunUMAP(filtered4_mn_seurat, dims = 1:30, reduction = "harmony")
filtered4_mn_seurat <- FindNeighbors(filtered4_mn_seurat, dims = 1:30, reduction = "harmony")        
filtered4_mn_seurat <- FindClusters(
  filtered4_mn_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 1, by = 0.1)
)

# Save the object
saveRDS(filtered4_mn_seurat, path_to_save_filtered4_mn_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
filtered4_mn_seurat <- readRDS(path_to_save_filtered4_mn_obj)
Idents(filtered4_mn_seurat) <- "SCT_snn_res.1"
markers <- FindAllMarkers(filtered4_mn_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered4_mn_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered4_mn_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered4_mn_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We removed cluster 0, 2, 7, 10, and 16 as they were not monocytes.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the monocyte clusters with reprocessing
filtered5_mn_seurat <- subset(filtered4_mn_seurat, idents = c(1, 3, 4, 5, 6, 8, 9, 11, 12, 13, 14, 15))
filtered5_mn_seurat <- SCTransform(filtered5_mn_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered5_mn_seurat <- RunPCA(filtered5_mn_seurat)
filtered5_mn_seurat <- RunUMAP(filtered5_mn_seurat, dims = 1:30, reduction = "harmony")
filtered5_mn_seurat <- FindNeighbors(filtered5_mn_seurat, dims = 1:30, reduction = "harmony")        
filtered5_mn_seurat <- FindClusters(
  filtered5_mn_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 1, by = 0.1)
)

# Save the object
saveRDS(filtered5_mn_seurat, path_to_save_filtered5_mn_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
filtered5_mn_seurat <- readRDS(path_to_save_filtered5_mn_obj)
Idents(filtered5_mn_seurat) <- "SCT_snn_res.1"
markers <- FindAllMarkers(filtered5_mn_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered5_mn_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered5_mn_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered5_mn_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We removed cluster 8 as it contained B cell doublets.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the monocyte clusters with reprocessing
filtered6_mn_seurat <- subset(filtered5_mn_seurat, idents = c(8), invert = TRUE)
filtered6_mn_seurat <- SCTransform(filtered6_mn_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered6_mn_seurat <- RunPCA(filtered6_mn_seurat)
filtered6_mn_seurat <- RunUMAP(filtered6_mn_seurat, dims = 1:30, reduction = "harmony")
filtered6_mn_seurat <- FindNeighbors(filtered6_mn_seurat, dims = 1:30, reduction = "harmony")        
filtered6_mn_seurat <- FindClusters(
  filtered6_mn_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 1, by = 0.1)
)

# Save the object
saveRDS(filtered6_mn_seurat, path_to_save_filtered6_mn_obj)
```

In conclusion, we removed and processed the monocyte object for 6 times till we obtained the pure monocyte object for further analysis.

# Subtype annotation

```{r, eval=FALSE}
# Change the obeject name
filtered6_mn_seurat <- readRDS(path_to_save_filtered6_mn_obj)
pure_mn_seurat <- filtered6_mn_seurat
saveRDS(pure_mn_seurat, path_to_save_pure_mn_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the clusters
pure_mn_seurat <- readRDS(path_to_save_pure_mn_obj)
group_by <- c("SCT_snn_res.0.1", "SCT_snn_res.0.2", "SCT_snn_res.0.3")
umap_plots <- lapply(group_by, function (group){
  DimPlot(pure_mn_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 3,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

pure_mn_umap.plt <- wrap_plots(umap_plots, ncol=3)
pure_mn_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the umap
group_by <- c("ref", "tissue", "group")
umap_plots <- lapply(group_by, function (group){
  DimPlot(pure_mn_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 3,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

pure_mn_umap.plt <- wrap_plots(umap_plots, ncol=3)
pure_mn_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=12}
# Visualise the cases
pure_mn_seurat$patient <- paste0(pure_mn_seurat$case, "_", pure_mn_seurat$ref)
case_mn.plt <- DimPlot(pure_mn_seurat, 
                       reduction = "umap", 
                       group.by = "patient", 
                       pt.size = 3,
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

case_mn.plt
```

Based on Parry (2023b), although the original clinical condition for cases MDAC-ACC-17_Parry_2023b, MDAC-ACC-18_Parry_2023b, and MDAC-ACC-29_Parry_2023b was CLL, their tumour cells were annotated as CCND2hi RT based on RT-associated marker expression and predominant S-phase cell-cycle scores (see Tumour Subtype Annotation section).

Similarly, in Penter (2021), tumour cells from case 1 were annotated as BCAT1hiCCND2hi RT, with high G2/M-phase cell-cycle scores. Accordingly, we reclassified the clinical condition of these samples from CLL to RT for downstream analyses.

```{r, eval=FALSE}
# Change the new condition for case MDAC-ACC-17_Parry_2023b, MDAC-ACC-18_Parry_2023b, and MDAC-ACC-29_Parry_2023b
selected_cases <- c("MDAC-ACC-17_Parry_2023b", "MDAC-ACC-18_Parry_2023b", "MDAC-ACC-29_Parry_2023b", "1_Penter_2021")
pure_mn_seurat$new_cond <- pure_mn_seurat$group
pure_mn_seurat@meta.data[pure_mn_seurat$patient %in% selected_cases, "new_cond"] <- "RT"
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Visualise the adapted condition
new_cond_mn.plt <- DimPlot(pure_mn_seurat, 
                       reduction = "umap", 
                       group.by = "new_cond", 
                       pt.size = 3,
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

new_cond_mn.plt
```

```{r, fig.height=7, fig.width=12, warning=FALSE, message=FALSE}
# Plot based on cell biomarker expression
FeaturePlot(pure_mn_seurat, 
            features = c("CD14", "FCGR3A", "CD68", "CD163", "HLA-DRA", "HLA-DRB1", "HLA-DRB5", 
                         "CD163", "TNFSF13B", "LGALS9", "C1QA", "C1QB", "ITGAM"),
            reduction = "umap") 
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(pure_mn_seurat) <- "SCT_snn_res.0.3"
markers <- FindAllMarkers(pure_mn_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_pure_mn_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/pure_mn_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_pure_mn_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We annotated the myeloid subtypes as literatures below:

- [The monocytic population in chronic lymphocytic leukemia shows altered composition and deregulation of genes involved in phagocytosis and inflammation](https://haematologica.org/article/view/6722)
- [Generation of a blockage monoclonal antibody of LILRB1 against HLA-G](https://www.sciencedirect.com/science/article/abs/pii/S1046592823001341)
- [Human Monocyte Subsets and Phenotypes in Major Chronic Inflammatory Diseases](https://www.frontiersin.org/journals/immunology/articles/10.3389/fimmu.2019.02035/full)
- [Distinct Populations of Immune-Suppressive Macrophages Differentiate from Monocytic Myeloid-Derived Suppressor Cells in Cancer](https://www.cell.com/cell-reports/fulltext/S2211-1247(20)31560-6?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS2211124720315606%3Fshowall%3Dtrue)
- [S100A8/A9-alarmin promotes local myeloid-derived suppressor cell activation restricting severe autoimmune arthritis](https://www.sciencedirect.com/science/article/pii/S2211124723010173)
- [Non-Classical monocytes display inflammatory features: Validation in Sepsis and Systemic Lupus Erythematous](https://www.nature.com/articles/srep13886)
- [687 MDSC gene expression analysis in patients with cancer and the response to inhibition of Bruton's tyrosine kinase](https://jitc.bmj.com/content/9/Suppl_2/A715)
- [Myeloid-derived suppressor cells as immunosuppressive regulators and therapeutic targets in cancer](https://www.nature.com/articles/s41392-021-00670-9)
- [Multi-omics analysis identifies an M-MDSC-like immunosuppressive phenotype in lineage-switched AML with KMT2A rearrangement](https://www.nature.com/articles/s41467-025-63271-y)
- [Distinctive features of “nurselike” cells that differentiate in the context of chronic lymphocytic leukemia](https://ashpublications.org/blood/article/99/3/1030/53487/Distinctive-features-of-nurselike-cells-that)
- [Nurse like cells: chronic lymphocytic leukemia associated macrophages](https://www.tandfonline.com/doi/abs/10.3109/10428194.2014.991731)
- [Factors involved in CLL pathogenesis and cell survival are disrupted by differentiation of CLL B-cells into antibody-secreting cells](https://www.oncotarget.com/article/3941/text/)
- [Tumor-associated macrophages-derived exosomes promote the migration of gastric cancer cells by transfer of functional Apolipoprotein E](https://www.nature.com/articles/s41419-018-0465-5)
- [IL2RA+VSIG4+ tumor-associated macrophage is a key subpopulation of the immunosuppressive microenvironment in anaplastic thyroid cancer](https://www.sciencedirect.com/science/article/pii/S0925443922002629)
- [Hypoxia-driven remodeling of SELENOP+ macrophages shapes T cell dynamics and promotes ovarian cancer metastasis](https://www.nature.com/articles/s41467-025-67859-2)

| Cluster | Markers                                                             | General annotation                  | Disease-based annotation                         |
|---------|---------------------------------------------------------------------|-------------------------------------|--------------------------------------------------|
| 0       | FCGR3A(CD16a), C1QA, C1QB                                           | Non-classical monocyte              | Non-classical monocyte                           |
| 1       | CD14, CD36, CD163, RNASE1, ITGAM(CD11b), ITGB2(CD18), TNFSF13B      | Classical monocyte (1)              | Classical monocyte (NLC-precursor)               |
| 2       | IL6, TNF, IL1B, CCL20, CCL4L2, LGALS9                               | Inflammatory classical monocyte (2) | Inflammatory classical monocyte (cytokine-high)  |
| 3       | CD14, CD36, ITGB2(CD18), S100A8, S100A9, CLU, VCAN                  | Early-classical monocyte            | M-MDSC-like monocyte                             |
| 4       | FCGR3A(CD16a), HLA-DRA, HLA-DRB1, HLA-DRB5, C1QA, C1QB, C1QC, C1QBP | Intermediate monocyte               | Intermediate monocyte                            |
| 5       | CD14, CD36, CD163, RNASE1, ITGAM(CD11b), TNFSF13B, GAS6             | Classical monocyte (2)              | Scavenger NLC-like monocyte                      |
| 6       | CD14, IL1B, CXCL1, IL1R2, CXCL3                                     | Inflammatory classical monocyte (1) | Inflammatory classical monocyte (chemokine-high) |
| 7       | FCGR3A(CD16a), APOE, SELENOP, VSIG4, C1QA                           | Macrophage-like monocyte            | TAM-like monocyte                                |

Cluster 1 is a NLC-precursor as it expresses TNFSF13B (BAFF) but lacks the full GAS6 signature, likely the stage just before becoming an NLC. While cluster 4 is scavenger NLC-like monocyte because It expresses GAS6 + BAFF + CD163. Then, It is functionally a Nurse-Like Cell circulating in the blood. Cluster 3 is early classical monocyte because of being CD14+ monocyte without strong inflammatory or scavenger features. However, as this cluster was exclusively observed in RT samples and exhibited expression of S100A8, S100A9, CLU, and VCAN, we annotated it as monocytic myeloid-derived suppressor cells (M-MDSCs)-like monocyte. Cluster 2 and 6 are inflammatory classical monocyte with the former has a cytokine-high pattern defined by IL6/IL1B and the latter has a chemokine-high pattern defined by CXCL1/CXCL3. Cluster 7 is TAM-like monocyte becasue it expresses APOE/SELENOP which are macrophage markers. In a cancer context, these are TAM precursors.

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=10}
# Plot UMAP for monocyte subtypes

# 1. Map Seurat clusters and create the annotation column
cluster_to_annotation <- c(
  "0" = "Non-classical monocyte", "1" = "Classical monocyte (NLC-precursor)", "2" = "Inflammatory classical monocyte (cytokine-predominant)", "3" = "M-MDSC-like monocyte", 
  "4" = "Intermediate monocyte", "5" = "Scavenger NLC-like monocyte", "6" = "Inflammatory classical monocyte (chemokine-predominant)", "7" = "TAM-like monocyte"
  )

pure_mn_seurat$level2_annotate <- 
  as.vector(cluster_to_annotation[as.character(pure_mn_seurat$SCT_snn_res.0.3)])

# 2. Set level order
new_levels <- c(
  # Group1: inflammatory monocyte 
  "Inflammatory classical monocyte (chemokine-predominant)", "Inflammatory classical monocyte (cytokine-predominant)",
  
  # Group2: non-inflammatory classical monocyte  
  "Classical monocyte (NLC-precursor)", "Scavenger NLC-like monocyte", "M-MDSC-like monocyte",     

  # Group3: non-classical monocyte 
  "Intermediate monocyte", "Non-classical monocyte", "TAM-like monocyte" 
)

pure_mn_seurat$level2_annotate <- factor(
  pure_mn_seurat$level2_annotate,
  levels = new_levels
)

# 3. Define colours and labels
new_colours <- c(
  "#FDBF6F", "#E25644",            # Group 1 (2 colours)
  "#1F78B4", "#C28BCD", "#A6CEE3", # Group 2 (3 colours)
  "#1FB4A5", "#33A02C", "#FFD92F"  # Group 3 (3 colours)
  
)

custom_labels <- list(
  bquote("Inflammatory classical monocyte (chemokine-predominant)"), bquote("Inflammatory classical monocyte (cytokine-predominant)"),
  bquote("Classical monocyte (NLC-precursor)"), bquote("Scavenger NLC-like monocyte"), bquote("M-MDSC-like monocyte"), 
  bquote("Intermediate monocyte"), bquote("Non-classical monocyte"), bquote("TAM-like monocyte")
  )

# 4. Plot UMAP
mn_subtype.plt <- DimPlot(pure_mn_seurat, 
                          group.by = "level2_annotate", 
                          label = FALSE,
                          repel = TRUE,
                          pt.size = 1.5,
                          alpha = 1,
                          raster = FALSE
                          ) +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels,
    labels = custom_labels
  ) +
  labs(y = "UMAP 2", x = "UMAP 1") +
  ggtitle("Monocyte subtypes")

mn_subtype.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# Dot plot for biomarker expressions based on annotated cells
features <- c("CD14", "FCGR3A", "IL1B", "TNF", "CXCL1", "IL1R2", "CXCL3", "IL6", "CCL20", "CCL4L2", "LGALS9", 
              "CD36", "CD163", "RNASE1", "ITGAM", "ITGB2", "TNFSF13B", "GAS6", "S100A8", "S100A9", "CLU", "VCAN", 
              "HLA-DRA", "HLA-DRB1", "HLA-DRB5", "APOE", "SELENOP", "VSIG4", "C1QA", "C1QB", "C1QC", "C1QBP")
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

mn_markers.plt <- DotPlot(pure_mn_seurat, features=features, group.by = "level2_annotate") +
  RotatedAxis() +
  scale_colour_gradientn(colours = colour_pal) +
  scale_y_discrete(
      limits = rev(new_levels),
      labels = rev(custom_labels)
    ) +
  labs(
    colour = "Average\nExpression",
    size = "Percent\nExpressed"
  )

mn_markers.plt
```

```{r, message=FALSE, warning=FALSE}
# Plot subtype population based on tissue and condition
proportion_tissue_df <- pure_mn_seurat@meta.data %>%
  dplyr::mutate(level2_annotate = factor(level2_annotate, levels = new_levels)) %>%
  dplyr::count(tissue, level2_annotate) %>%
  dplyr::group_by(tissue) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n) # Remove n column

proportion_cond_df <- pure_mn_seurat@meta.data %>%
  dplyr::mutate(level2_annotate = factor(level2_annotate, levels = new_levels)) %>%
  dplyr::count(new_cond, level2_annotate) %>%
  dplyr::group_by(new_cond) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n) # Remove n column

tissue_cell_pop.plt <- ggplot(proportion_tissue_df, aes(x = tissue, 
                                                    y = proportion, 
                                                    fill = level2_annotate)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "Monocyte subtype" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Tissue", 
    title = "Tissue"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

cond_cell_pop.plt <- ggplot(proportion_cond_df, aes(x = new_cond, 
                                                    y = proportion, 
                                                    fill = level2_annotate)) +
  geom_col(position = "stack", width = 0.5) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "Monocyte subtype" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Condition", 
    title = "Condition"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

# Combine the plots
adapted_tissue_cell_pop.plt <- tissue_cell_pop.plt +
  theme(legend.position = "none",
        axis.title.x = element_blank()) 

adapted_cond_cell_pop.plt <- cond_cell_pop.plt + 
  theme(legend.position = "right", 
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) 

combined_cell_pop.plt <- adapted_tissue_cell_pop.plt + adapted_cond_cell_pop.plt
combined_cell_pop.plt
```

From the cell-type composition, we observed that monocytes in the CLL condition were predominantly detected in peripheral blood and were mainly present in inflammatory states. In contrast, monocytes in the RT condition were more frequently found in bone marrow and lymph node samples and were composed primarily of two major subsets: classical and non-classical monocytes.

```{r, eval=FALSE}
# Save the monocyte object
saveRDS(pure_mn_seurat, path_to_save_pure_mn_obj)

# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/mn_umap.png"), plot=mn_subtype.plt, dpi=500, height=5, width=10)
ggsave(filename=paste0(path_to_save_figure, "/mn_dot.png"), plot=mn_markers.plt, dpi=500, height=4, width=18)
ggsave(filename=paste0(path_to_save_figure, "/mn_cell_pop.png"), plot=combined_cell_pop.plt , dpi=500, height=4, width=7)
```

# Trajectory analysis

In this section, we used the Monocle 3 package to identify developmental trajectories of monocytes from the inflammatory state toward M-MDSC-like monocyte, nurse-like cell (NLC), non-classical, and tumour-associated macrophage (TAM) states, and to characterise gene expression changes along these trajectories.

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(monocle3)
library(igraph)
library(SeuratWrappers)
library(paletteer)
```

```{r}
# Define path to save
path_to_save_mn_monocle_obj <- here::here("scRNAseq/meta_analysis/R_objects/mn_monocle_cds.rds")
path_to_save_mn_monocle_dea <- here::here("scRNAseq/meta_analysis/R_objects/mn_monocle_dea.rds")
path_to_save_mn_monocle_module <- here::here("scRNAseq/meta_analysis/R_objects/mn_monocle_module.rds")
path_to_save_mn_monocle_agg_mat <- here::here("scRNAseq/meta_analysis/R_objects/mn_monocle_agg_mat.rds")
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create monocle object and generate graph (use data from seurat)
cds <- as.cell_data_set(pure_mn_seurat) # Convert the seurat object to monocle object using SeuratWrappers
cds <- preprocess_cds(cds)

# Create gene_short_name
rowData(cds)$gene_short_name <- row.names(rowData(cds))

# Assign seurat clusters to monocle and create a partition
cds <- monocle3::cluster_cells(cds, reduction_method = "UMAP", 
                               cluster_method = "louvain", 
                               resolution = 0.5)
cds@clusters$UMAP$clusters <- pure_mn_seurat$SCT_snn_res.0.3

# Assign UMAP coordination
cds@int_colData$reducedDims$UMAP <- pure_mn_seurat@reductions$umap@cell.embeddings
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=17}
# Monocle plot before trajectory analysis
cds <- readRDS(path_to_save_mn_monocle_obj)
group_by <- c("cluster", "partition", "level2_annotate")
umap_plots <- lapply(group_by, function (group){
  plot_cells(cds, 
             color_cells_by = group, 
             label_groups_by_cluster = FALSE, 
             group_label_size = 5,
             cell_size = 0.7) +  
    theme(legend.position = "right")
})

monocle_umap.plt <- wrap_plots(umap_plots, ncol=3)
monocle_umap.plt
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Learn trajectory
cds <- learn_graph(cds, use_partition = TRUE, close_loop = FALSE)
```

```{r, fig.height=5, fig.width=7}
# Plot after learning trajectory
monocle_cluster.plt <- plot_cells(cds, 
                                  color_cells_by = "cluster",
                                  graph_label_size=5,
                                  cell_size = 0.7,
                                  show_trajectory_graph = TRUE
                                  ) + 
  ggtitle("Seurat clusters") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

monocle_cluster.plt
```

```{r, fig.height=5, fig.width=7}
# Plot root
root.plt <- plot_cells(cds, label_principal_points = TRUE) + 
  ggtitle("Root principal node") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

root.plt
```

## Calculate pseudotime

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set a root to order cells in pseudotime --> we did not use this command, we used the command in the box below instead.
get_earliest_principal_node <- function(cds, root_cluster="6"){
  # Identify the cells in the starting cluster
  cell_ids <- which(colData(cds)[, "SCT_snn_res.0.3"] == root_cluster)
  
  # Find the principal graph node closest to these cells
  closest_vertex <- cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  closest_vertex <- as.matrix(closest_vertex[colnames(cds), ])
  root_pr_nodes <- igraph::V(principal_graph(cds)[["UMAP"]])$name[as.numeric(names(which.max(table(closest_vertex[cell_ids,]))))]
  
  return(root_pr_nodes)
}

cds <- order_cells(cds, root_pr_nodes=get_earliest_principal_node(cds))
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set a root to order cells in pseudotime
cds <- order_cells(cds, reduction_method = "UMAP", root_pr_nodes = "Y_80")
```

```{r, fig.height=5, fig.width=7}
# Plot the pseudotime
monocle_pseudotime.plt <- plot_cells(cds, 
                                     color_cells_by = "pseudotime",
                                     label_cell_groups=FALSE,
                                     label_leaves=FALSE,
                                     label_branch_points=FALSE,
                                     graph_label_size=1.5,
                                     cell_size = 0.7) +
  ggtitle("Pseudotime") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

monocle_pseudotime.plt
```

```{r, fig.height=5, fig.width=5}
# Plot the pseudotime with trajectory path with Seurat stype

# 1. Extract principal graph
pg_aux <- principal_graph_aux(cds)$UMAP

# 2. Convert to edge list
edges <- as_data_frame(principal_graph(cds)$UMAP, what = "edges")
nodes <- as.data.frame(t(pg_aux$dp_mst))
nodes$name <- rownames(nodes)
root_node <- cds@principal_graph_aux$UMAP$root_pr_nodes

# 3. Merge edges with coordinates
trajectory_df <- edges %>%
  dplyr::left_join(nodes, by = c("from" = "name")) %>%
  dplyr::rename(
    UMAP_1_from = umap_1,
    UMAP_2_from = umap_2
  ) %>%
  dplyr::left_join(nodes, by = c("to" = "name")) %>%
  dplyr::rename(
    UMAP_1_to = umap_1,
    UMAP_2_to = umap_2
  )

seurat_df <- data.frame(
  Embeddings(pure_mn_seurat, "umap"),
  pseudotime = pseudotime(cds)
)

root_df <- nodes %>%
  dplyr::filter(name %in% root_node)

# 4. Plot
trajectory.plt <- ggplot(seurat_df, aes(umap_1, umap_2)) +
  geom_point(aes(color = pseudotime), size = 2, alpha=1) +
  geom_segment(
    data = trajectory_df,
    aes(
      x = UMAP_1_from,
      y = UMAP_2_from,
      xend = UMAP_1_to,
      yend = UMAP_2_to
    ),
    inherit.aes = FALSE,
    colour = "black",
    linewidth = 0.7
  ) +
  geom_point(
    data = root_df,
    aes(umap_1, umap_2),
    inherit.aes = FALSE,
    shape = 21,          # circle with fill
    size = 6,
    fill = "white",
    colour = "black",
    stroke = 1
  ) +
  geom_text(
    data = root_df,
    aes(umap_1, umap_2),
    label = "1",
    inherit.aes = FALSE,
    size = 5,
    fontface = "bold"
  ) +
  scale_color_gradientn(
    name = "Pseudotime",
    colors = paletteer_c("grDevices::RdYlBu", 30, direction = -1)
  ) +
  theme_void() +
  theme(
    legend.position = "top" 
  ) + 
  annotate("text", x = -2, y = 7.6, label = "Inflam. mono.", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -7, y = 6.6, label = "CD14+ NLC", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 8, y = -7.7, label = "CD16+ TAM-like mono.", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -5.5, y = -8.3, label = "M-MDSC-like mono.", 
             color = "black", fontface = "plain", size = 4)

trajectory.plt
```

```{r, eval=FALSE}
# Save the monocle object
saveRDS(cds, path_to_save_mn_monocle_obj)

# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/mn_trajectory.png"), plot=trajectory.plt, dpi=500, height=5, width=5)
```

## Gene expression across pseudotime

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Group genes into modules
mn_cds_pr_test_res <- monocle3::graph_test(cds, neighbor_graph = "principal_graph", cores = 4)
monocle_degs <- row.names(subset(mn_cds_pr_test_res, q_value < 0.05))
gene_modules_df <- find_gene_modules(cds[monocle_degs,], resolution=5e-3)
cell_group_df <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=colData(cds)$level2_annotate)
agg_mat <- aggregate_gene_expression(cds, gene_modules_df, cell_group_df)
row.names(agg_mat) <- paste0("Module ", row.names(agg_mat))

# Save the module results
write.csv(mn_cds_pr_test_res, here::here("scRNAseq/meta_analysis/supplm_file/mn_trajectory_gene_expr.csv"))
saveRDS(mn_cds_pr_test_res, path_to_save_mn_monocle_dea)
saveRDS(gene_modules_df, path_to_save_mn_monocle_module)
saveRDS(agg_mat, path_to_save_mn_monocle_agg_mat)
```

```{r, message=FALSE, warning=FALSE}
# Show the monocle dae
monocle_dea <- readRDS(path_to_save_mn_monocle_dea)
DT::datatable(monocle_dea, options = list(scrollX = TRUE))
```

```{r, fig.height=10, fig.width=8}
# Plot heatmap 
library(pheatmap)
gene_modules_df <- readRDS(path_to_save_mn_monocle_module)
agg_mat <- readRDS(path_to_save_mn_monocle_agg_mat)
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)
pseudotime_heatmap.plt <- pheatmap::pheatmap(agg_mat, scale="column", fontsize = 11,
                                             color = colour_pal,
                                             angle_col = 45,
                                             clustering_method="ward.D2")

pseudotime_heatmap.plt
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Test a range of resolutions to appropriate modules
resolutions <- c(1e-3, 2e-3, 3e-3, 4e-3, 5e-3)

for(res in resolutions){
  modules <- find_gene_modules(cds[monocle_degs,], resolution = res)
  num_mods <- length(unique(modules$module))
  print(paste("Resolution:", res, "->", num_mods, "Modules"))
}
```

```{r}
# Collect gene signatures for fate subtypes based on gene modules
chemokine_gene <- gene_modules_df %>% filter(module %in% c(1, 9)) %>% pull(id)
cytokine_gene <- gene_modules_df %>% filter(module %in% c(3, 7)) %>% pull(id)
nlc_gene <- gene_modules_df %>% filter(module %in% c(8, 15)) %>% pull(id)
classical_gene <- gene_modules_df %>% filter(module %in% c(17)) %>% pull(id)
mdsc_gene <- gene_modules_df %>% filter(module %in% c(10)) %>% pull(id)
intermediate_gene <- gene_modules_df %>% filter(module %in% c(23, 25)) %>% pull(id)
non_classical_gene <- gene_modules_df %>% filter(module %in% c(6, 11)) %>% pull(id)
tam_gene <- gene_modules_df %>% filter(module %in% c(22)) %>% pull(id)
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 1 Inflammatory classical monocyte > TAM
cells_types_of_interest <- c("Inflammatory classical monocyte (cytokine-predominant)", 
                             "Inflammatory classical monocyte (chemokine-predominant)",
                             "Intermediate monocyte", "Non-classical monocyte", "TAM-like monocyte")
pseudotime_genes <- c("CD14", "FCGR3A", "C1QA", "CEBPB", "TIAM2", "PDK4")

tam_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                          genes = pseudotime_genes, 
                                                          cell_annotations = cells_types_of_interest, 
                                                          pseudotime_boundary = 20,
                                                          facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Inflam. mono. to TAM-like mono.") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

tam_genes_pseudotime
```

During the transition from an active inflammatory state towards a TAM-like phenotype, monocytes exhibited a shift from classical to non-classical states, characterised by reduced expression of CD14 and increased expression of FCGR3A (CD16a). In addition, expression of the complement-related gene C1QA increased at later pseudotime stages. Several other genes showed elevated expression during TAM late-stage differentiation, including CEBPB, a key transcriptional regulator implicated in the suppression of T cell activity; TIAM2, a pro-tumourigenic factor associated with cancer progression through activation of oncogenic signalling pathways, immune cell modulation, and interactions with stromal cells. Collectively, the upregulation of these genes is consistent with a tumour-supportive myeloid programme that promotes tumour proliferation and survival; and PDK4, an enzyme that promotes metabolic reprogramming towards a Warburg-like phenotype. Collectively, the upregulation of these genes is consistent with a tumour-supportive myeloid programme that promotes tumour proliferation and survival.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 2 Inflammatory classical monocyte > M-MDSC
cells_types_of_interest <- c("Inflammatory classical monocyte (cytokine-predominant)",
                            "Inflammatory classical monocyte (chemokine-predominant)", 
                            "Classical monocyte (NLC-precursor)", "M-MDSC-like monocyte")
pseudotime_genes <- c("VCAN", "S100A9", "HLA-DRB5", "LAG3", "EGR1", "GLUL")

mdsc_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                          genes = pseudotime_genes, 
                                                          cell_annotations = cells_types_of_interest, 
                                                          pseudotime_boundary = 15,
                                                          facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Inflam. mono. to M-MDSC-like mono.") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

mdsc_genes_pseudotime
```

The transition from an inflammatory to an M-MDSC-like monocyte state was also observed during RT progression, characterised by increased expression of VCAN and S100A9, together with reduced expression of HLA-DRB5 at later pseudotime stages. In addition, several genes associated with a TAM-like programme showed increased expression in the late state, including LAG3, an inhibitory immune checkpoint molecule; EGR1, a transcription factor implicated in the establishment of an immunosuppressive microenvironment; and GLUL (glutamine synthetase), which contributes to metabolic reprogramming and immune suppression within the tumour microenvironment. Collectively, these transcriptional changes support a progressive shift towards immunosuppressive, tumour-supportive myeloid states during RT progression.    

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 3 Inflammatory classical monocyte > NLC
cells_types_of_interest <- c("Inflammatory classical monocyte (cytokine-predominant)", 
                             "Inflammatory classical monocyte (chemokine-predominant)", 
                             "Classical monocyte (NLC-precursor)", 
                             "Scavenger NLC-like monocyte")
pseudotime_genes <- c("CD36", "CD163", "CXCL12", "TNFSF13B", "IL10RB", "LGALS1", "HPSE")

nlc_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                          genes = pseudotime_genes, 
                                                          cell_annotations = cells_types_of_interest, 
                                                          pseudotime_boundary = 15,
                                                          facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Inflam. mono. to NLC") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

nlc_genes_pseudotime
```

Another monocyte fate identified during RT progression is the differentiation from active inflammatory monocytes towards NLCs, as indicated by the increased expression of CD36 and CD163, established markers of NLCs, at later pseudotime stages. In contrast to other monocyte fates that promote RT progression indirectly through immune suppression and metabolic reprogramming, this NLC-like subtype appears to support tumour survival and proliferation more directly through the upregulation of TNFSF13B (BAFF), a key factor enhancing tumour cell activation and persistence.


In addition, several genes associated with immune suppression and tumour invasion were upregulated in this trajectory, including IL10RB, the receptor for the immunosuppressive cytokine IL-10; LGALS1 (Galectin-1), a potent immunomodulatory protein known to induce apoptosis in infiltrating T cells; and HPSE (Heparanase), an extracellular matrix–degrading enzyme that facilitates tumour invasion by clearing physical barriers. Collectively, these transcriptional changes support a distinct NLC-like monocyte fate that directly promotes tumour growth and immune evasion during RT progression.

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/mn_tam_gene_flow.png"), plot=tam_genes_pseudotime, dpi=500, height=8, width=6)
ggsave(filename=paste0(path_to_save_figure, "/mn_mdsc_gene_flow.png"), plot=mdsc_genes_pseudotime, dpi=500, height=8, width=6)
ggsave(filename=paste0(path_to_save_figure, "/mn_nlc_gene_flow.png"), plot=nlc_genes_pseudotime, dpi=500, height=8, width=6)
```

# Pseudobulk gene expression analysis between RT versus CLL

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(DESeq2)
library(limma)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(ggpubr)
```

## Based on Seurat command

```{r}
# Define paths to save and load
path_to_save_seurat_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_seurat_dea.rds")
path_to_save_seurat_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_seurat_hm_gsea.rds")
path_to_save_seurat_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_seurat_go_gsea.rds")
path_to_save_seurat_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_seurat_kegg_gsea.rds")
path_to_save_seurat_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_seurat_reactome_gsea.rds")
path_to_save_seurat_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_seurat_wikipath_gsea.rds")
```

### Find biomarkers of monocytes between RT and CLL conditions

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Find markers between CLL and RT conditions  
Idents(pure_mn_seurat) <- "new_cond" # we used this variable as it contains RT and CLL
pure_mn_seurat <- PrepSCTFindMarkers(pure_mn_seurat) # Prepare SCT assay for DE testing
mn.ge <- FindMarkers(
  pure_mn_seurat,
  ident.1 = "RT",
  ident.2 = "CLL",
  test.use = "wilcox",
  only.pos = FALSE,
  logfc.threshold = 0
)
dea.df <- cbind(gene = rownames(mn.ge), mn.ge)
avg_exp <- AverageExpression(pure_mn_seurat, features = dea.df$gene, assay = "SCT") # Calculate the average expression for the genes
dea.df$baseMean <- rowMeans(avg_exp$SCT) # Calculate the mean across all relevant groups (e.g., global mean for the gene)

# Save the expression results
saveRDS(dea.df, path_to_save_seurat_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/mn_seurat_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Load the saved expression data
dea.df <- readRDS(path_to_save_seurat_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=10}
# Volcano plot for Seurat pseudobulk analysis
genes_of_interest <- c(
  "RNASE1", "CLU", "LAG3", "C1QB", "C1QC", "EGR1", "FCGR3A", "NR4A3",                      # Upregulated
  "IL1B", "CXCL1", "CCL3L1", "FOS", "FOSB", "JUN", "NFKBIA", "CCL2", "TNF"                 # Downregulated
)

non_zero_min_pval <- sort(unique(dea.df$p_val_adj), decreasing = FALSE)[2] # Find minimum p-value which is not 0
dea.df$adapted_p_val <- ifelse(dea.df$p_val_adj == 0, non_zero_min_pval, dea.df$p_val_adj) # Change zero adj_p_val to non-zero minimum adj_p_val to avoid plotting error

volcano.plt <- volcano_plot(dea.df, 
                            FC=0.25, 
                            x_col="avg_log2FC",
                            y_col="adapted_p_val",
                            xlim=c(-12,12), 
                            ylim=c(-0.1,320), 
                            genes_of_interest=genes_of_interest,
                            boxedLabels = TRUE) 

volcano.plt
```

```{r, message=FALSE, warning=FALSE}
# MA plot for Seurat pseudobulk analysis
ma_data <- dea.df
colnames(ma_data)[colnames(ma_data) == "avg_log2FC"] <- "log2FoldChange"
colnames(ma_data)[colnames(ma_data) == "p_val_adj"]  <- "padj"

ma.plt <- ggmaplot(ma_data, 
                   fdr = 0.05,            # Threshold for significance
                   fc = 2,                # Threshold for FC
                   size = 0.4,            # Dot size
                   palette = c("#B31B21", "#1465AC", "darkgray"),
                   genenames = as.vector(ma_data$gene),
                   top = 0, label.select = genes_of_interest, # set top = 0, then specify genes using label.select argument
                   legend = "right", 
                   font.label = c("bold", 11), 
                   label.rectangle = TRUE,
                   font.legend = "bold",
                   font.main = "bold",
                   ggtheme = ggplot2::theme_minimal(),
                   xlab = "Mean Expression",
                   ylab = "Log2 Fold Change"
                   )

ma.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/seurat_mn_volcano.png"), plot=volcano.plt, dpi=500, height=7, width=9)
ggsave(filename=paste0(path_to_save_figure, "/seurat_mn_ma.png"), plot=ma.plt, dpi=500, height=3.5, width=5)
```

From the Seurat-based gene expression analysis we highlighted some interested DEGs. The results strongly confirm the biological story: RT monocytes switch OFF the inflammation and switch ON a Scavenger/Immunosuppressive program.

1.  Upregulated DEGs in RT. These genes represent the Scavenger / Nurse-Like Cell (NLC) phenotype that supports the tumor.

`RNASE1`: The top scavenger marker.

`CLU` (Clusterin): A scavenger chaperone protein involved in clearing debris.

`C1QB` and `C1QC`: Part of the complement pathway, often upregulated in TAMs.

`EGR1`: A transcription factor often associated with monocyte differentiation and the NLC phenotype.

`FCGR3A (CD16)`: A marker for non-classical monocytes 


2.  Downregulated DEGs in RT (Enriched in CLL). These genes represent the Inflammatory / Cytokine Storm phenotype that is lost during transformation.

`IL1B` and `TNF`: The master regulator of inflammation.

`CXCL1` and `CCL3L1`: Chemokines responsible for recruiting other immune cells.

`FOS`, `FOSB` and `JUN`: Immediate-early genes involved in acute activation and stress response.

`NFKBIA`: A regulator of the NF-kB pathway, further confirming the loss of inflammatory signaling.

`CCL2`: A major monocyte chemoattractant.

`NR4A3`: role in monocyte differentiation to dendritic cells

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea.df$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

### GSEA summary 

We then summarised the common terms found from Hallmark, GO:BP, and canonical pathways.

```{r, message=FALSE, warning=FALSE}
# Bar plot of GSEA 
gsea_bp <- rbind(
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_COMPLEMENT",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_B_CELL_MEDIATED_IMMUNITY",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_CD4_POSITIVE_ALPHA_BETA_T_CELL_ACTIVATION",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_GRANULOCYTE_CHEMOTAXIS",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_HYPOXIA",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_INFLAMMATORY_RESPONSE",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_APOPTOSIS",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_TNFA_SIGNALING_VIA_NFKB",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_IL2_STAT5_SIGNALING",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_KRAS_SIGNALING_UP",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_IL6_JAK_STAT3_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL1_SIGNALING",],
  REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_INTERLEUKIN_10_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL18_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL26_SIGNALING",],
  REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_SIGNALING_BY_GPCR",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_NOD_LIKE_RECEPTOR_SIGNALING_PATHWAY",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_P53_PATHWAY",]
)

gsea_bp$Description <- gsub("HALLMARK_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("GOBP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("KEGG_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("REACTOME_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("WP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("_V1", "", gsea_bp$Description)
gsea_bp$Description <- gsub("CD4_POSITIVE_ALPHA_BETA_T_CELL_ACTIVATION", "CD4_T_CELL_ACTIVATION", gsea_bp$Description)
gsea_bp$Description <- gsub("KRAS_SIGNALING_UP", "KRAS_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("INTERLEUKIN_10_SIGNALING", "IL10_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("SIGNALING_BY_GPCR", "GPCR_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("NOD_LIKE_RECEPTOR_SIGNALING_PATHWAY", "NOD_LIKE_RECEPTOR_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION", "CYTOKINE_RECEPTOR_INTERACTION", gsea_bp$Description)
gsea_bp$Description <- gsub("_", " ", gsea_bp$Description)

gsea_bp$Description <- factor(gsea_bp$Description, levels = gsea_bp$Description)
gsea_bp$Direction <- "Up"
gsea_bp$Direction[gsea_bp$NES<0] <- "Down"

gsea_bar.plt <- ggplot(gsea_bp, aes(x=Description, y=NES, fill=Direction)) +
  geom_bar(stat = "identity", color="black", width = 0.75) +  
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, color = "black"),
    plot.margin = margin(t = 10, r = 10, b = 100, l = 65, unit = "pt"),
    legend.position = "none"
  ) +
  geom_hline(yintercept = 0) + 
  ylab("NES") + xlab(NULL) + 
  scale_fill_manual(values=c("#357EBD", "#D43F3A"))

gsea_bar.plt
```

Based on the GSEA results comparing RT and CLL monocytes, we identified several major biological alterations, as described below.

- Increased complement expression. complement-high monocytes act as precursors to immunosuppressive TAMs.

- Suppressed inflammatory and chemotactic processes in RT. RT monocytes exhibited a marked downregulation of immune and inflammatory responses, including reduced immune cell adhesion, impaired T-cell activation, and decreased activity of multiple cytokine signalling pathways.

- Enhanced interactions with B cells. RT monocytes showed increased tumour cell–associated interactions, characterised by upregulation of B-cell–related immune signalling pathways.

- Maintenance of RT monocyte survival through reduced apoptosis. Pathways associated with apoptosis and p53 signalling were downregulated, suggesting enhanced survival and maintenance of the RT monocyte population.

- Reduced response to hypoxic stress. RT monocytes demonstrated decreased hypoxia-responsive signalling, which may contribute to a metabolic shift towards oxidative phosphorylation.

```{r, message=FALSE, warning=FALSE, fig.height=18, fig.width=12}
# GSEA plots
terms_of_interest <- c(
  "HALLMARK_COMPLEMENT", "GOBP_B_CELL_MEDIATED_IMMUNITY", "GOBP_CD4_POSITIVE_ALPHA_BETA_T_CELL_ACTIVATION", 
  "GOBP_GRANULOCYTE_CHEMOTAXIS", "HALLMARK_INFLAMMATORY_RESPONSE", "HALLMARK_APOPTOSIS", "HALLMARK_TNFA_SIGNALING_VIA_NFKB", 
  "HALLMARK_IL2_STAT5_SIGNALING", "HALLMARK_KRAS_SIGNALING_UP", "HALLMARK_IL6_JAK_STAT3_SIGNALING", 
  "WP_IL1_SIGNALING", "REACTOME_INTERLEUKIN_10_SIGNALING", "WP_IL18_SIGNALING", "WP_IL26_SIGNALING", 
  "REACTOME_SIGNALING_BY_GPCR", "KEGG_NOD_LIKE_RECEPTOR_SIGNALING_PATHWAY", 
  "KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION", "HALLMARK_P53_PATHWAY", "HALLMARK_HYPOXIA"
  )

gsea_plot.lst <- lapply(terms_of_interest, function(term){
  term_system <- strsplit(term, split = "_")[[1]][1]

  adapted_term <- term %>% 
    str_remove("^(HALLMARK|GOBP|KEGG|REACTOME|WP)_") %>%
    str_replace("CD4_POSITIVE_ALPHA_BETA_T_CELL_ACTIVATION", "CD4_T_CELL_ACTIVATION") %>%
    str_replace("KRAS_SIGNALING_UP", "KRAS_SIGNALING") %>%
    str_replace("INTERLEUKIN_10_SIGNALING", "IL10_SIGNALING") %>%
    str_replace("POSITIVE_REGULATION_OF_DEFENSE_RESPONSE", "DEFENSE_RESPONSE_REGULATION") %>%
    str_replace("SIGNALING_BY_GPCR", "GPCR_SIGNALING") %>%
    str_replace("NOD_LIKE_RECEPTOR_SIGNALING_PATHWAY", "NOD_LIKE_RECEPTOR_SIGNALING") %>%
    str_replace("CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION", "CYTOKINE_RECEPTOR_INTERACTION") %>%
    str_replace_all("_", " ")
  
  if (term_system == "HALLMARK"){
    gseaplot(HM_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "GOBP"){
    gseaplot(GOBP_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "KEGG") {
    gseaplot(KEGG_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "REACTOME") {
    gseaplot(REACTOME_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else {
    gseaplot(WIKIPATH_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  }
})

gsea.plt <- wrap_plots(gsea_plot.lst, ncol=3)
gsea.plt
```

```{r, eval=FALSE}
# Save GSEA results
saveRDS(HM_gsea, path_to_save_seurat_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_seurat_go_gsea)
saveRDS(KEGG_gsea, path_to_save_seurat_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_seurat_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_seurat_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_seurat_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_seurat_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_seurat_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_seurat_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_seurat_wikipath_gsea.csv"))

# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/mn_gsea.png"), plot=gsea_bar.plt, dpi=500, height=5, width=10)
ggsave(filename=paste0(path_to_save_figure, "/mn_gsea_plt.png"), plot=gsea.plt, dpi=500, height=18, width=12)
```

## Based on DESeq2

```{r}
# Define paths to save the results
path_to_save_deseq2_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_deseq2_dea.rds")
path_to_save_deseq2_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_deseq2_hm_gsea.rds")
path_to_save_deseq2_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_deseq2_go_gsea.rds")
path_to_save_deseq2_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_deseq2_kegg_gsea.rds")
path_to_save_deseq2_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_deseq2_reactome_gsea.rds")
path_to_save_deseq2_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/mn_deseq2_wikipath_gsea.rds")
```

### Create an DESeq2 object

```{r, message=FALSE, warning=FALSE}
# Prepare single-cell meta-data for DESeq2
pure_mn_seurat$ref[pure_mn_seurat$ref == "Hing_2022"] <- "Hing_2023" # Update reference name
pure_mn_seurat$patient_cond <- paste0("case_", pure_mn_seurat$case, "_", pure_mn_seurat$ref, "_", pure_mn_seurat$new_cond) 
patient_cond_cls <- unique(pure_mn_seurat$patient_cond)
RT_subjects <- patient_cond_cls[grep("RT", patient_cond_cls)]
RT_subjects <- gsub(pattern = "_", replacement = "-", x=RT_subjects) # change "_" to "-" as DESeq2 converted sample names from "_" to "-"

# Create pseudocount matrix for tumour cell object
cts <- AggregateExpression(pure_mn_seurat, 
                           group.by = "patient_cond",
                           assays = "RNA",
                           slot = "counts",
                           resturn.seurat = FALSE) 
cts <- cts$RNA
pseudo_count.mat <- as.data.frame(cts)

# Create metadata for DESeq2 object
batch_map <- c("Penter (2021)" = "1", "Nadeu (2022)" = "2", "Rejeski (2022)" = "3", "Hing (2023)" = "4", "Parry (2023a)" = "5", 
               "Parry (2023b)" = "6", "Rigo (2024)" = "7")
colData <- data.frame(samples = colnames(pseudo_count.mat))
colData$condition <- ifelse(colData$samples %in% RT_subjects, "RT", "CLL")
colData$study <- stringr::word(colData$samples, start = -3, end = -2, sep = fixed("-"))
colData$study <- sub("^(.*)-([0-9].*)$", "\\1 (\\2)", colData$study) # Substitiute string like "Parry-2023b" to "Parry (2023b)"
colData$study <- factor(colData$study, level = c("Penter (2021)", "Nadeu (2022)", "Rejeski (2022)", "Hing (2023)", 
                                                 "Parry (2023a)", "Parry (2023b)", "Rigo (2024)"))
colData$batch <- batch_map[colData$study]
colData$batch <- factor(colData$batch)
rownames(colData) <- colData$samples

# Create a DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = pseudo_count.mat, 
                              colData = colData,
                              design = ~ batch + condition)

# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

### Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6.5}
# Variance stabilised transformation
vsdata <- vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
design_matrix <- model.matrix(~ condition, data = colData(dds))
vs_mat_corrected <- limma::removeBatchEffect(
  vs_mat, 
  batch = colData(dds)$batch, 
  design = design_matrix 
)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2               # variance of each PC
pca_var_percent <- pca_var / sum(pca_var) * 100

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     study = colData(dds)$study)

# Plot PCA
pca.plt <- ggplot(pca_df, aes(PC1, PC2, color = study, shape = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  scale_color_manual(values = c(
    "Penter (2021)" = "#1f77b4", 
    "Nadeu (2022)" = "#ff7f0e",
    "Rejeski (2022)" = "#9632B8" ,
    "Hing (2023)" = "#D43F3A" ,
    "Parry (2023a)" = "#5CB85C",
    "Parry (2023b)" = "#FFB900",
    "Rigo (2024)" = "#46B8DA")) +
  scale_shape_manual(values = c("CLL" = 16, "RT" = 17)) +
  guides(
    color = guide_legend(title = "Study"),
    shape = guide_legend(title = "Condition")
  ) +
  theme_bw() +
  theme(
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 11)
  )

pca.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/deseq2_mn_pca.png"), plot=pca.plt, dpi=500, height=4.5, width=6.5)
```

### Run DESeq2

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Run DESeq2 for the data after outlier removal
keep <- rowSums(counts(dds) >= 10) >= 3 # Keep genes with at least 10 counts in at least 3 samples
noise_genes <- c(
  grep("^IG[HKL][VDJ]", rownames(dds), value = TRUE),  # Ig genes
  grep("^HLA-", rownames(dds), value = TRUE),          # Patient ID (HLA)
  grep("^MT-", rownames(dds), value = TRUE),           # Mitochondrial stress
  "XIST", "RPS4Y1", "KDM5D", "UTY", "DDX3Y",           # Sex genes
  "S100A8", "S100A9", "S100A12", "CXCL8"               # Acute stress response
)

dds <- dds[keep & (!rownames(dds) %in% noise_genes), ]
dds$condition <- relevel(dds$condition, ref = "CLL")
dds <- DESeq(dds)

# Generate a table for gene expression
res <- results(dds, contrast = c("condition", "RT", "CLL"), independentFiltering = FALSE)
res <- lfcShrink(dds, contrast = c("condition", "RT", "CLL"), res = res, type = "normal")
res$direction <- "NS"
res$direction[res$padj < 0.05 & res$log2FoldChange > 0.25] <- "Up"
res$direction[res$padj < 0.05 & res$log2FoldChange < (-0.25)] <- "Down"
dea.df <- as.data.frame(res)

# Save the expression results
saveRDS(dea.df, path_to_save_deseq2_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/mn_deseq2_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Show the expression results
dea.df <- readRDS(path_to_save_deseq2_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

We listed the DEGs for both up- and downregulation below.

```{r}
# Upregulated DEGs arranged by log2FC from high to low
up_degs <- dea.df %>%
  dplyr::filter(padj < 0.05 & log2FoldChange > 0.25) %>%
  dplyr::arrange(desc(log2FoldChange)) %>%
  tibble::rownames_to_column(var = "gene") %>%
  dplyr::pull(gene)

up_degs
```

```{r}
# Downregulated DEGs arranged by log2FC from high to low
down_degs <- dea.df %>%
  dplyr::filter(padj < 0.05 & log2FoldChange < (-0.25)) %>%
  dplyr::arrange(log2FoldChange) %>%
  tibble::rownames_to_column(var = "gene") %>%
  dplyr::pull(gene)

down_degs
```

From the results, we observed that pseudobulk analysis using DESeq2 failed to detect DEGs, likely due to low read counts resulting from monocyte aggregation.

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=10}
# Volcano plot
volcano.plt <- volcano_plot(dea.df, 
                            FC=0.25, 
                            x_col="log2FoldChange",
                            y_col="padj",
                            xlim=c(-1.2,1.2), 
                            ylim=c(-0.1,2), 
                            genes_of_interest=NULL,
                            boxedLabels = TRUE) 

volcano.plt
```

```{r, message=FALSE, warning=FALSE}
# MA plot 
ma_data <- dea.df

ma.plt <- ggmaplot(ma_data, 
                   fdr = 0.05,            # Threshold for significance
                   fc = 1.19,                # Threshold for FC
                   size = 0.4,            # Dot size
                   palette = c("#B31B21", "#1465AC", "darkgray"),
                   genenames = as.vector(ma_data$gene),
                   top = 0, label.select = NULL, # set top = 0, then specify genes using label.select argument
                   legend = "right", 
                   font.label = c("bold", 11), 
                   label.rectangle = TRUE,
                   font.legend = "bold",
                   font.main = "bold",
                   ggtheme = ggplot2::theme_minimal(),
                   xlab = "Mean Expression",
                   ylab = "Log2 Fold Change"
                   )

ma.plt
```

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df) 
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save the enrichment results
saveRDS(HM_gsea, path_to_save_deseq2_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_deseq2_go_gsea)
saveRDS(KEGG_gsea, path_to_save_deseq2_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_deseq2_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_deseq2_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_deseq2_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_deseq2_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_deseq2_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_deseq2_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/mn_deseq2_wikipath_gsea.csv"))
```

# Session Information

```{r}
sessionInfo()
```
