---
title: "Microenvironment_CD4+_T_cell_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

In this analysis, we focused on CD4+ T cells, from the integrated dataset by performing pseudobulk analysis between RT and CLL in this cell type. Moreover, we also annotated level-2 subtypes of these cells.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tibble)
library(SingleR)
library(celldex)
library(sctransform)
library(cowplot)
library(DT)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_tcell_obj <- here::here("scRNAseq/meta_analysis/R_objects/t_seu.rds")
path_to_save_filtered_tcell_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered_tcell_seu.rds")
path_to_save_filtered_tcell_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered_tcell_markers.rds")
path_to_save_cd4_obj <- here::here("scRNAseq/meta_analysis/R_objects/cd4_seu.rds")
path_to_save_cd8_obj <- here::here("scRNAseq/meta_analysis/R_objects/cd8_seu.rds")
path_to_save_figure <- here::here("figures")
```

```{r}
# Thresholds
min_log2FC <- 0.25
alpha <- 0.001
```

```{r}
# Set seed for reproducibility
set.seed(1)
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r, eval=FALSE}
# Load the t cell object
tcell_seurat <- readRDS(path_to_tcell_obj)
```

# Filtering out non-T celll contamination

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# T cell purification

# 1.Calculate Percentages of Contaminants
# RBC (Hemoglobin)
tcell_seurat[["percent.rbc"]] <- PercentageFeatureSet(tcell_seurat, pattern = "^HB[AB]")
# B cells (Immunoglobulins) 
tcell_seurat[["percent.ig"]] <- PercentageFeatureSet(tcell_seurat, pattern = "^IG[HKL]")
# Platelets
platelet_genes <- c("PF4", "PPBP", "TUBB1")
tcell_seurat[["percent.platelet"]] <- PercentageFeatureSet(tcell_seurat, features = platelet_genes)

# 2.Filter out RBCs, platelets, and high immunoglobulin (plasma/B cell) noise
filtered_tcell_seurat <- subset(tcell_seurat, 
                            subset = percent.rbc < 1 & 
                                     percent.ig < 1 & 
                                     percent.platelet < 1) 

# 3.Filter out contaminated B cells and NK cells
filtered_tcell_seurat <- subset(filtered_tcell_seurat, subset = MS4A1 < 1 & KLRF1 < 1)

# 4.Filter out contaminated monocytes
filtered_tcell_seurat <- subset(filtered_tcell_seurat, subset = CD14 < 1 & LYZ < 2 & S100A9 < 1) 

# 5.Reanalysis 
filtered_tcell_seurat <- SCTransform(filtered_tcell_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE) 
filtered_tcell_seurat <- RunPCA(filtered_tcell_seurat)
filtered_tcell_seurat <- RunUMAP(filtered_tcell_seurat, dims = 1:30, reduction = "harmony") 
filtered_tcell_seurat <- FindNeighbors(filtered_tcell_seurat, dims = 1:30, reduction = "harmony")        
filtered_tcell_seurat <- FindClusters(
  filtered_tcell_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# 6. Save the object
saveRDS(filtered_tcell_seurat, path_to_save_filtered_tcell_obj)
```

```{r, fig.height=7, fig.width=12, warning=FALSE, message=FALSE}
# Plot based on cell biomarker expression
filtered_tcell_seurat <- readRDS(path_to_save_filtered_tcell_obj)
FeaturePlot(filtered_tcell_seurat, 
            features = c("HBM", "HBA1", "CD79A", "CD14", "CD3D", "CD3E", 
                         "CD4", "CD8A", "CD8B", "KLRF1", "FCGR3A", "TYROBP"),
            reduction = "umap") 
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Cluster visualisation
group_by <- c("SCT_snn_res.0.1", "SCT_snn_res.0.2", "SCT_snn_res.0.3")
umap_plots <- lapply(group_by, function (group){
  DimPlot(filtered_tcell_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 2,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

tcell_umap.plt <- wrap_plots(umap_plots, ncol=3)
tcell_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=9}
# Dot plot for biomarker expressions 
features <- c("HBM", "HBA1", "CD79A", "CD14", "CD3D", "CD3E", 
              "CD4", "CD8A", "CD8B", "KLRF1", "FCGR3A", "TYROBP")
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

tcell_markers.plt <- DotPlot(filtered_tcell_seurat, features=features, group.by = "SCT_snn_res.0.1") +
  RotatedAxis() +
  scale_colour_gradientn(colours = colour_pal) +
  labs(
    colour = "Average\nExpression",
    size = "Percent\nExpressed"
  )

tcell_markers.plt
```

# Differentiate CD4+ and CD8+ T cells

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(filtered_tcell_seurat) <- "SCT_snn_res.0.1"
markers <- FindAllMarkers(filtered_tcell_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered_tcell_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered_tcell_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered_tcell_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

From the marker identification we found that cluster 1 and 2 were CD4+ T cells while cluster 0, 3, and 6 were CD8+ T cells. We then extracted them for further analysis.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Extract CD4+ T cells with reanalysis
cd4_seurat <- subset(filtered_tcell_seurat, subset = SCT_snn_res.0.1 %in% c(1, 2))
cd4_seurat <- subset(cd4_seurat, subset = CD4 > 0 & CD8A < 1 & CD8B < 1 & TRDC < 1) # Filter out CD8+ and Gamma-Delta T cells
cd4_seurat <- SCTransform(cd4_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
cd4_seurat <- RunPCA(cd4_seurat)
cd4_seurat <- RunUMAP(cd4_seurat, dims = 1:30, reduction = "harmony")
cd4_seurat <- FindNeighbors(cd4_seurat, dims = 1:30, reduction = "harmony")        
cd4_seurat <- FindClusters(
  cd4_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# Save the object
saveRDS(cd4_seurat, path_to_save_cd4_obj)
```

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Extract CD8+ T cells with reanalysis
cd8_seurat <- subset(filtered_tcell_seurat, subset = SCT_snn_res.0.1 %in% c(0, 3, 6))
cd8_seurat <- subset(cd8_seurat, subset = (CD8A > 0 | CD8B > 0) & CD4 < 1) # Filter out CD4+ T cells
cd8_seurat <- SCTransform(cd8_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
cd8_seurat <- RunPCA(cd8_seurat)
cd8_seurat <- RunUMAP(cd8_seurat, dims = 1:30, reduction = "harmony")
cd8_seurat <- FindNeighbors(cd8_seurat, dims = 1:30, reduction = "harmony")        
cd8_seurat <- FindClusters(
  cd8_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# Save the object
saveRDS(cd8_seurat, path_to_save_cd8_obj)
```

# CD4+ T cell purification

```{r}
# Define paths to load and save 
path_to_save_cd4_markers <- here::here("scRNAseq/meta_analysis/R_objects/cd4_marker.rds")
path_to_save_filtered_cd4_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered_cd4_seu.rds")
path_to_save_filtered_cd4_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered_cd4_marker.rds")
path_to_save_filtered2_cd4_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered2_cd4_seu.rds")
path_to_save_filtered2_cd4_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered2_cd4_marker.rds")
path_to_save_pure_cd4_obj <- here::here("scRNAseq/meta_analysis/R_objects/pure_cd4_seu.rds")
path_to_save_pure_cd4_markers <- here::here("scRNAseq/meta_analysis/R_objects/pure_cd4_marker.rds")
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(cd4_seurat) <- "SCT_snn_res.1.2"
markers <- FindAllMarkers(cd4_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_cd4_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/cd4_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_cd4_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We found that cluster 2, 10, and 14 were contaminated cells (B cell, RBC, and platelets).

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the CD4+ T cell clusters with reprocessing
filtered_cd4_seurat <- subset(cd4_seurat, idents = c(2, 10, 14), invert = TRUE)
filtered_cd4_seurat <- SCTransform(filtered_cd4_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered_cd4_seurat <- RunPCA(filtered_cd4_seurat)
filtered_cd4_seurat <- RunUMAP(filtered_cd4_seurat, dims = 1:30, reduction = "harmony")
filtered_cd4_seurat <- FindNeighbors(filtered_cd4_seurat, dims = 1:30, reduction = "harmony")        
filtered_cd4_seurat <- FindClusters(
  filtered_cd4_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# Save the object
saveRDS(filtered_cd4_seurat, path_to_save_filtered_cd4_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(filtered_cd4_seurat) <- "SCT_snn_res.2"
markers <- FindAllMarkers(filtered_cd4_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered_cd4_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered_cd4_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered_cd4_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We found contaminated cells from cluster 3, 4, 7, 8, 9, 10, 11, 14, 15, 16, 19, and 26. We then removed them.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the CD4+ T cell clusters with reprocessing
filtered2_cd4_seurat <- subset(filtered_cd4_seurat, idents = c(3, 4, 7, 8, 9, 10, 11, 14, 15, 16, 19, 26), invert = TRUE)
filtered2_cd4_seurat <- SCTransform(filtered2_cd4_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered2_cd4_seurat <- RunPCA(filtered2_cd4_seurat)
filtered2_cd4_seurat <- RunUMAP(filtered2_cd4_seurat, dims = 1:30, reduction = "harmony")
filtered2_cd4_seurat <- FindNeighbors(filtered2_cd4_seurat, dims = 1:30, reduction = "harmony")        
filtered2_cd4_seurat <- FindClusters(
  filtered2_cd4_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# Save the object
saveRDS(filtered2_cd4_seurat, path_to_save_filtered2_cd4_obj)
```

In conclusion, we removed and processed the CD4+ cell object for 2 times till we obtained the pure CD4+ T cell object for further analysis.

# Subtype annotation

```{r, eval=FALSE}
# Change the obeject name
pure_cd4_seurat <- filtered2_cd4_seurat
saveRDS(pure_cd4_seurat, path_to_save_pure_cd4_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the clusters
pure_cd4_seurat <- readRDS(path_to_save_pure_cd4_obj)
group_by <- c("SCT_snn_res.0.4", "SCT_snn_res.0.5", "SCT_snn_res.0.6")
umap_plots <- lapply(group_by, function (group){
  DimPlot(pure_cd4_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 4,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

pure_cd4_umap.plt <- wrap_plots(umap_plots, ncol=3)
pure_cd4_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the umap
group_by <- c("ref", "tissue", "group")
umap_plots <- lapply(group_by, function (group){
  DimPlot(pure_cd4_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 4,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

pure_cd4_umap.plt <- wrap_plots(umap_plots, ncol=3)
pure_cd4_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=10}
# Visualise the cases
pure_cd4_seurat$patient <- paste0(pure_cd4_seurat$case, "_", pure_cd4_seurat$ref)
case_cd4.plt <- DimPlot(pure_cd4_seurat, 
                       reduction = "umap", 
                       group.by = "patient", 
                       pt.size = 3,
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

case_cd4.plt
```

Based on Parry (2023b), although the original clinical condition for cases MDAC-ACC-17_Parry_2023b, MDAC-ACC-18_Parry_2023b, and MDAC-ACC-29_Parry_2023b was CLL, their tumour cells were annotated as CCND2hi RT based on RT-associated marker expression and predominant S-phase cell-cycle scores (see Tumour Subtype Annotation section).

Similarly, in Penter (2021), tumour cells from case 1 were annotated as BCAT1hiCCND2hi RT, with high G2/M-phase cell-cycle scores. Accordingly, we reclassified the clinical condition of these samples from CLL to RT for downstream analyses.

```{r, eval=FALSE}
# Change the new condition for case MDAC-ACC-17_Parry_2023b, MDAC-ACC-18_Parry_2023b, and MDAC-ACC-29_Parry_2023b
selected_cases <- c("MDAC-ACC-17_Parry_2023b", "MDAC-ACC-18_Parry_2023b", "MDAC-ACC-29_Parry_2023b", "1_Penter_2021")
pure_cd4_seurat$new_cond <- pure_cd4_seurat$group
pure_cd4_seurat@meta.data[pure_cd4_seurat$patient %in% selected_cases, "new_cond"] <- "RT"
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=6}
# Visualise the adapted condition
new_cond_cd4.plt <- DimPlot(pure_cd4_seurat, 
                       reduction = "umap", 
                       group.by = "new_cond", 
                       pt.size = 3,
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

new_cond_cd4.plt
```

```{r, fig.height=7, fig.width=12, warning=FALSE, message=FALSE}
# Plot based on cell biomarker expression
FeaturePlot(pure_cd4_seurat, 
            features = c("CD3D", "CD4", "TCF7", "CXCR3", "NKG7", "CCR8", "CDKN1A",
                         "BACH2", "CXCR5", "CCR4", "HAVCR2", "MKI67"),
            reduction = "umap") 
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(pure_cd4_seurat) <- "SCT_snn_res.0.6"
markers <- FindAllMarkers(pure_cd4_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_pure_cd4_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/pure_cd4_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_pure_cd4_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We annotated the myeloid subtypes as literature below:

- [Role of CD4+ T cells in cancer immunity: a single-cell sequencing exploration of tumor microenvironment](https://link.springer.com/article/10.1186/s12967-025-06167-1)
- [Selective addition of CXCR3+CCR4-CD4+ Th1 cells enhances generation of cytotoxic T cells by dendritic cells in vitro](https://www.nature.com/articles/emm200920)
- [GATA-3 promotes Th2 responses through three different mechanisms: induction of Th2 cytokine production, selective growth of Th2 cells and inhibition of Th1 cell-specific factors](https://www.nature.com/articles/7310002)
- [Bcl6-Mediated Transcriptional Regulation of Follicular Helper T cells (TFH)](https://www.cell.com/trends/immunology/fulltext/S1471-4906(21)00026-0?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1471490621000260%3Fshowall%3Dtrue)
- [BACH2 drives quiescence and maintenance of resting Treg cells to promote homeostasis and cancer immunosuppression](https://rupress.org/jem/article/217/9/e20190711/151846/BACH2-drives-quiescence-and-maintenance-of-resting)
- [CDKN1A and cellular senescence are associated with immune resistance in advanced lung adenocarcinoma](https://link.springer.com/article/10.1007/s10238-025-01926-2)
- [A Multi-Omics Analysis of an Exhausted T Cells’ Molecular Signature in Pan-Cancer](https://www.mdpi.com/2075-4426/14/7/765)
- [Intratumoral CD4+ T Cells Mediate Anti-tumor Cytotoxicity in Human Bladder Cancer](https://www.sciencedirect.com/science/article/pii/S0092867420306127)
- [Inhibition of topoisomerase IIA (Top2α) induces telomeric DNA damage and T cell dysfunction during chronic viral infection](https://www.nature.com/articles/s41419-020-2395-2?utm_source=researchgate.net&utm_medium=article)
- [A Population of Tumor-Infiltrating CD4+ T Cells Co-Expressing CD38 and CD39 Is Associated with Checkpoint Inhibitor Resistance](https://aacrjournals.org/clincancerres/article-abstract/29/20/4242/729426/A-Population-of-Tumor-Infiltrating-CD4-T-Cells-Co?redirectedFrom=fulltext)
- [Stem-like progenitor and terminally differentiated TFH-like CD4+ T cell exhaustion in the tumor microenvironment](https://www.cell.com/cell-reports/fulltext/S2211-1247(24)00125-6)
- [Discovery and validation of proliferative exhausted T cells as a favorable prognostic biomarker in esophageal squamous cell carcinoma](https://www.nature.com/articles/s41598-025-25535-x#Sec2)

| Cluster | Markers                                              | Annotation                |
|---------|------------------------------------------------------|---------------------------|
| 0       | CXCR3                                                | Th1                       |
| 1       | NKG7, GZMA, GZMK                                     | CD4+ CTL                  |
| 2       | TCF7, LEF1, SELL                                     | Tn                        |
| 3       | FOXP3                                                | Treg                      |
| 4       | BACH2, CDKN1A                                        | BACH2hi quiescent CD4+ T  |
| 5       | CCR4, GATA3                                          | Th2                       |
| 6       | CXCR5, BCL6                                          | Tfh                       |
| 7       | HAVCR2, LAG3, PDCD1, TIGIT, CD38, GZMH               | Tex                       |
| 8       | MKI67, PCNA, TOP2A, CD38, CXCL13, GZMH, CDK4, ESCO2  | Prolif-like Tex           |

Abbreviations: Tn, naive T cell; Th, helper T cell; Treg, regulatory T cell; Tfh, T follicular helper cell; Tex, exhausted T cell; Prolif-like Tex, proliferative-like Tex; 

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=7}
# Plot UMAP for CD4+ T cell subtypes

# 1. Map Seurat clusters and create the annotation column
cluster_to_annotation <- c(
  "0" = "Th1", "1" = "CD4+ CTL", "2" = "Tn", "3" = "Treg", "4" = "BACH2hi quiescent CD4+ T", 
  "5" = "Th2", "6" = "Tfh", "7" = "Tex", "8" = "Prolif-like Tex"
  )

pure_cd4_seurat$level2_annotate <- 
  as.vector(cluster_to_annotation[as.character(pure_cd4_seurat$SCT_snn_res.0.6)])

# 2. Set level order
new_levels <- c(
  # Group1: naive, helper, and cytotoxic CD4+  
  "Tn", "Th1", "Th2", "CD4+ CTL", "Tfh",
  
  # Group2: regulatory and quiescent CD4+  
  "Treg", "BACH2hi quiescent CD4+ T",     

  # Group3: exhausted CD4+  
  "Tex", "Prolif-like Tex" 
)

pure_cd4_seurat$level2_annotate <- factor(
  pure_cd4_seurat$level2_annotate,
  levels = new_levels
) 

# 3. Define colours and labels
new_colours <- c(
  "#B3B3B3", "#1F78B4", "#33A02C", "#008080", "#C28BCD",  # Group 1 (5 colours)
  "#FDBF6F", "#FFD92F",                                   # Group 2 (2 colours)
  "#E25644", "#B04968"                                    # Group 3 (2 colours)
) 

custom_labels <- list(
  bquote("Tn"), bquote("Th1"), bquote("Th2"), bquote("CD4+ CTL"), bquote("Tfh"),
  bquote("Treg"), bquote(BACH2^hi~"quiescent CD4+ T"), 
  bquote("Tex"), bquote("Prolif-like Tex")
  )

# 4. Plot UMAP
cd4_subtype.plt <- DimPlot(pure_cd4_seurat, 
                          group.by = "level2_annotate", 
                          label = FALSE,
                          repel = TRUE,
                          pt.size = 1,
                          alpha = 1,
                          raster = FALSE
                          ) +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels,
    labels = custom_labels
  ) +
  labs(y = "UMAP 2", x = "UMAP 1") +
  ggtitle("CD4+ T cell subtypes")

cd4_subtype.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Dot plot for biomarker expressions based on annotated cells
features <- c("TCF7", "LEF1", "SELL", "CXCR3", "CCR4", "GATA3", "NKG7", "GZMA", "GZMK",
              "CXCR5", "BCL6", "FOXP3", "BACH2", "CDKN1A", "HAVCR2", "LAG3", "PDCD1", "TIGIT", 
              "CD38", "GZMH", "CXCL13", "CDK4", "ESCO2", "MKI67", "PCNA", "TOP2A"
              )
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

cd4_markers.plt <- DotPlot(pure_cd4_seurat, features=features, group.by = "level2_annotate") +
  RotatedAxis() +
  scale_colour_gradientn(colours = colour_pal) +
  scale_y_discrete(
      limits = rev(new_levels),
      labels = rev(custom_labels)
    ) +
  labs(
    colour = "Average\nExpression",
    size = "Percent\nExpressed"
  )

cd4_markers.plt
```

```{r, message=FALSE, warning=FALSE}
# Plot subtype population based on tissue and condition
proportion_tissue_df <- pure_cd4_seurat@meta.data %>%
  dplyr::mutate(level2_annotate = factor(level2_annotate, levels = new_levels)) %>%
  dplyr::count(tissue, level2_annotate) %>%
  dplyr::group_by(tissue) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n) # Remove n column

proportion_cond_df <- pure_cd4_seurat@meta.data %>%
  dplyr::mutate(level2_annotate = factor(level2_annotate, levels = new_levels)) %>%
  dplyr::count(new_cond, level2_annotate) %>%
  dplyr::group_by(new_cond) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n) # Remove n column

tissue_cell_pop.plt <- ggplot(proportion_tissue_df, aes(x = tissue, 
                                                    y = proportion, 
                                                    fill = level2_annotate)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "CD4+ T cell subtype" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Tissue", 
    title = "Tissue"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

cond_cell_pop.plt <- ggplot(proportion_cond_df, aes(x = new_cond, 
                                                    y = proportion, 
                                                    fill = level2_annotate)) +
  geom_col(position = "stack", width = 0.5) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "CD4+ T cell subtype" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Condition", 
    title = "Condition"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

# Combine the plots
adapted_tissue_cell_pop.plt <- tissue_cell_pop.plt +
  theme(legend.position = "none",
        axis.title.x = element_blank()) 

adapted_cond_cell_pop.plt <- cond_cell_pop.plt + 
  theme(legend.position = "right", 
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) 

combined_cell_pop.plt <- adapted_tissue_cell_pop.plt + adapted_cond_cell_pop.plt
combined_cell_pop.plt
```

```{r, eval=FALSE}
# Save the CD4 T cell object
saveRDS(pure_cd4_seurat, path_to_save_pure_cd4_obj)

# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/cd4_umap.png"), plot=cd4_subtype.plt, dpi=500, height=4.5, width=7)
ggsave(filename=paste0(path_to_save_figure, "/cd4_dot.png"), plot=cd4_markers.plt, dpi=500, height=4, width=13)
ggsave(filename=paste0(path_to_save_figure, "/cd4_cell_pop.png"), plot=combined_cell_pop.plt , dpi=500, height=4, width=7)
```

# Trajectory analysis

In this section, we used the Monocle 3 package to identify developmental trajectories of CD4+ T cells using naive T cell subtype as a root node.

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(monocle3)
library(igraph)
library(SeuratWrappers)
library(paletteer)
```

```{r}
# Define path to save
path_to_save_cd4_monocle_obj <- here::here("scRNAseq/meta_analysis/R_objects/cd4_monocle_cds.rds")
path_to_save_cd4_monocle_dea <- here::here("scRNAseq/meta_analysis/R_objects/cd4_monocle_dea.rds")
path_to_save_cd4_monocle_module <- here::here("scRNAseq/meta_analysis/R_objects/cd4_monocle_module.rds")
path_to_save_cd4_monocle_agg_mat <- here::here("scRNAseq/meta_analysis/R_objects/cd4_monocle_agg_mat.rds")
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create monocle object and generate graph (use data from seurat)
cds <- as.cell_data_set(pure_cd4_seurat) # Convert the seurat object to monocle object using SeuratWrappers
cds <- preprocess_cds(cds)

# Create gene_short_name
rowData(cds)$gene_short_name <- row.names(rowData(cds))

# Assign seurat clusters to monocle and create a partition
cds <- monocle3::cluster_cells(cds, reduction_method = "UMAP", 
                               cluster_method = "louvain", 
                               resolution = 0.5)
cds@clusters$UMAP$clusters <- pure_cd4_seurat$SCT_snn_res.0.6

# Assign UMAP coordination
cds@int_colData$reducedDims$UMAP <- pure_cd4_seurat@reductions$umap@cell.embeddings
```

```{r, eval=FALSE}
# Set cluster 5 as partition 2 as Th2 cannot differenriate from CTL
all_cells <- colnames(cds)
assigned_partitions <- rep("1", length(all_cells))
names(assigned_partitions) <- all_cells
cluster5_cells <- colnames(pure_cd4_seurat)[pure_cd4_seurat$SCT_snn_res.0.6 == "5"]
assigned_partitions[cluster5_cells] <- "2"
cds@clusters$UMAP$partitions <- as.factor(assigned_partitions)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=17}
# Monocle plot before trajectory analysis
cds <- readRDS(path_to_save_cd4_monocle_obj)
group_by <- c("cluster", "partition", "level2_annotate")
umap_plots <- lapply(group_by, function (group){
  plot_cells(cds, 
             color_cells_by = group, 
             label_groups_by_cluster = FALSE, 
             group_label_size = 5,
             cell_size = 0.7) +  
    theme(legend.position = "right")
})

monocle_umap.plt <- wrap_plots(umap_plots, ncol=3)
monocle_umap.plt
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Learn trajectory
cds <- learn_graph(cds, use_partition = TRUE, close_loop = FALSE,
                   learn_graph_control = list(minimal_branch_len = 5, rann.k = 50))
```

```{r, fig.height=5, fig.width=7}
# Plot after learning trajectory
monocle_cluster.plt <- plot_cells(cds, 
                                  color_cells_by = "cluster",
                                  graph_label_size=5,
                                  cell_size = 0.7,
                                  show_trajectory_graph = TRUE
                                  ) + 
  ggtitle("Seurat clusters") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

monocle_cluster.plt
```

```{r, fig.height=5, fig.width=7}
# Plot root
root.plt <- plot_cells(cds, label_principal_points = TRUE) + 
  ggtitle("Root principal node") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

root.plt
```

```{r, fig.height=9, fig.width=8, warning=FALSE, message=FALSE}
# Plot based on cell biomarker expression
FeaturePlot(pure_cd4_seurat, 
            features = c("CCR7", "LEF1", "TCF7", "SELL", "BACH2", "GATA3"),
            reduction = "umap") 
```

From the feature plots, we saw that Y_267 area showed the highest expression of CCR7, LEF1, TCf7, and SELL in the cluster 2. We then set this point as the root node. For cluster 5, we set Y_306 as the root node as it expressed the high BACH2 (from cluster 4) while lowest expressed in GATA3 (signature marker for Th2).

## Calculate pseudotime

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set a root to order cells in pseudotime
cds <- order_cells(cds, reduction_method = "UMAP", root_pr_nodes = c("Y_267", "Y_306"))
```

```{r, fig.height=5, fig.width=7}
# Plot the pseudotime
monocle_pseudotime.plt <- plot_cells(cds, 
                                     color_cells_by = "pseudotime",
                                     label_cell_groups=FALSE,
                                     label_leaves=FALSE,
                                     label_branch_points=FALSE,
                                     graph_label_size=1.5,
                                     cell_size = 0.7) +
  ggtitle("Pseudotime") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

monocle_pseudotime.plt
```

```{r, fig.height=5, fig.width=5}
# Plot the pseudotime with trajectory path with Seurat style

# 1. Extract principal graph
pg_aux <- principal_graph_aux(cds)$UMAP

# 2. Convert to edge list
edges <- as_data_frame(principal_graph(cds)$UMAP, what = "edges")
nodes <- as.data.frame(t(pg_aux$dp_mst))
nodes$name <- rownames(nodes)
root_node <- cds@principal_graph_aux$UMAP$root_pr_nodes

# 3. Merge edges with coordinates
trajectory_df <- edges %>%
  dplyr::left_join(nodes, by = c("from" = "name")) %>%
  dplyr::rename(
    UMAP_1_from = umap_1,
    UMAP_2_from = umap_2
  ) %>%
  dplyr::left_join(nodes, by = c("to" = "name")) %>%
  dplyr::rename(
    UMAP_1_to = umap_1,
    UMAP_2_to = umap_2
  )

seurat_df <- data.frame(
  Embeddings(pure_cd4_seurat, "umap"),
  pseudotime = pseudotime(cds)
)

root_df <- nodes %>%
  dplyr::filter(name %in% root_node)

# 4. Plot
trajectory_no_annotate.plt <- ggplot(seurat_df, aes(umap_1, umap_2)) +
  geom_point(aes(color = pseudotime), size = 1, alpha=1) +
  geom_segment(
    data = trajectory_df,
    aes(
      x = UMAP_1_from,
      y = UMAP_2_from,
      xend = UMAP_1_to,
      yend = UMAP_2_to
    ),
    inherit.aes = FALSE,
    colour = "black",
    linewidth = 0.7
  ) +
  geom_point(
    data = root_df,
    aes(umap_1, umap_2),
    inherit.aes = FALSE,
    shape = 21,          # circle with fill
    size = 6,
    fill = "white",
    colour = "black",
    stroke = 1
  ) +
  geom_text(
    data = root_df,
    aes(umap_1, umap_2),
    label = c("1", "2"),
    inherit.aes = FALSE,
    size = 5,
    fontface = "bold"
  ) +
  scale_color_gradientn(
    name = "Pseudotime",
    colors = paletteer_c("grDevices::RdYlBu", 30, direction = -1)
  ) +
  theme_void() +
  theme(
    legend.position = "bottom" 
  )  
  
annotate_trajectory.plt <- trajectory_no_annotate.plt +  
  annotate("text", x = -5.5, y = -6.6, label = "Tn", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -3, y = 5.5, label = "Treg", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 2.8, y = 4.2, label = "Tex", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 6.2, y = 5, label = "Prolif-like Tex", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 8.3, y = -1.5, label = "Tfh", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 8.3, y = 2.2, label = "Th2", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 2.4, y = -2.8, label = "Quiescent", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -0.5, y = -2.8, label = "Th1", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 0.8, y = 3.7, label = "CTL", 
             color = "black", fontface = "plain", size = 4)

annotate_trajectory.plt
```

```{r, eval=FALSE}
# Save the monocle object
saveRDS(cds, path_to_save_cd4_monocle_obj)

# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/cd4_trajectory.png"), plot=annotate_trajectory.plt, dpi=500, height=5, width=5)
```

## Gene expression across pseudotime

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Group genes into modules
cd4_cds_pr_test_res <- monocle3::graph_test(cds, neighbor_graph = "principal_graph", cores = 4)
monocle_degs <- row.names(subset(cd4_cds_pr_test_res, q_value < 0.05))
gene_modules_df <- find_gene_modules(cds[monocle_degs,], resolution=4e-3)
cell_group_df <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=colData(cds)$level2_annotate)
agg_mat <- aggregate_gene_expression(cds, gene_modules_df, cell_group_df)
row.names(agg_mat) <- paste0("Module ", row.names(agg_mat))

# Save the module results
write.csv(cd4_cds_pr_test_res, here::here("scRNAseq/meta_analysis/supplm_file/cd4_trajectory_gene_expr.csv"))
saveRDS(cd4_cds_pr_test_res, path_to_save_cd4_monocle_dea)
saveRDS(gene_modules_df, path_to_save_cd4_monocle_module)
saveRDS(agg_mat, path_to_save_cd4_monocle_agg_mat)
```

```{r, message=FALSE, warning=FALSE}
# Show the monocle dae
monocle_dea <- readRDS(path_to_save_cd4_monocle_dea)
DT::datatable(monocle_dea, options = list(scrollX = TRUE))
```

```{r, fig.height=9, fig.width=6}
# Plot heatmap 
library(pheatmap)
gene_modules_df <- readRDS(path_to_save_cd4_monocle_module)
agg_mat <- readRDS(path_to_save_cd4_monocle_agg_mat)
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)
pseudotime_heatmap.plt <- pheatmap::pheatmap(agg_mat, scale="column", fontsize = 11,
                                             color = colour_pal,
                                             angle_col = 45,
                                             clustering_method="ward.D2")

pseudotime_heatmap.plt
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Test a range of resolutions to appropriate modules
resolutions <- c(1e-3, 2e-3, 3e-3, 4e-3, 5e-3)

for(res in resolutions){
  modules <- find_gene_modules(cds[monocle_degs,], resolution = res)
  num_mods <- length(unique(modules$module))
  print(paste("Resolution:", res, "->", num_mods, "Modules"))
}
```

```{r}
# Collect gene signatures for fate subtypes based on gene modules
treg_gene <- gene_modules_df %>% filter(module %in% c(18, 29)) %>% pull(id)
tex_gene <- gene_modules_df %>% filter(module %in% c(11, 21, 26)) %>% pull(id)
prolif_tex_gene <- gene_modules_df %>% filter(module %in% c(3, 12, 20, 27)) %>% pull(id)
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 1 Tn > Treg
cells_types_of_interest <- c("Tn", "Treg")
pseudotime_genes <- c("CTLA4", "FOXP3", "HLA-DRA", "PSAP", "VSIR", "DDIT4")

treg_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                           genes = pseudotime_genes, 
                                                           cell_annotations = cells_types_of_interest, 
                                                           pseudotime_boundary = 20,
                                                           facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Tn to Treg") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

treg_genes_pseudotime
```

During Treg development, we observed that this subtype exhibited high expression of HLA-DR genes, suggesting a highly suppressive and terminally differentiated phenotype. Additional TME-associated genes were also identified. For example, PSAP and VSIR have been implicated in CD8⁺ T cell suppression across various cancer types. DDIT4, a hypoxia-responsive gene, was also upregulated, consistent with TME-associated hypoxic conditions.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 2 Tn > Tex
cells_types_of_interest <- c("Tn", "Th1", "CD4+ CTL", "Tex")
pseudotime_genes <- c("PDCD1", "LGALS9", "ISG15", "STAT1", "XAF1", "GLS")

tex_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                           genes = pseudotime_genes, 
                                                           cell_annotations = cells_types_of_interest, 
                                                           pseudotime_boundary = 21,
                                                           facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Tn to Tex") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

tex_genes_pseudotime
```

For the Tex cell fate, upregulation of PDCD1, a canonical exhaustion marker, was observed. Additionally, LGALS9, an immune suppression–associated gene, was also upregulated. Increased expression of ISG15 and STAT1, classically associated with antiviral responses, has been implicated in immune regulation and tumour-promoting processes. Metabolic genes, including GLS, were also upregulated, suggesting metabolic adaptation and potential competition for energy resources within the TME.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 3 Tn > Prolif-like Tex.
cells_types_of_interest <- c("Tn", "Th1", "CD4+ CTL", "Tex")
pseudotime_genes <- c("LAG3", "PCNA", "BAD", "CKS2", "TNFRSF14", "ENO1")

prolif_tex_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                           genes = pseudotime_genes, 
                                                           cell_annotations = cells_types_of_interest, 
                                                           pseudotime_boundary = 25,
                                                           facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Tn to Prolif-like Tex") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

prolif_tex_genes_pseudotime
```

For the Prolif-like Tex fate, we observed increased expression of both exhaustion- and proliferation-associated genes, including LAG3 and PCNA. Concurrent upregulation of the pro-apoptotic gene BAD suggests elevated cellular turnover, potentially contributing to the relatively small population size. Moreover, immune suppression–associated genes, such as CKS2, TNFRSF14, and ENO1, were also upregulated.

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/cd4_treg_gene_flow.png"), plot=treg_genes_pseudotime, dpi=500, height=8, width=6)
ggsave(filename=paste0(path_to_save_figure, "/cd4_tex_gene_flow.png"), plot=tex_genes_pseudotime, dpi=500, height=8, width=6)
ggsave(filename=paste0(path_to_save_figure, "/cd4_prolif_tex_gene_flow.png"), plot=prolif_tex_genes_pseudotime, dpi=500, height=8, width=6)
```

# Pseudobulk gene expression analysis between RT versus CLL

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(DESeq2)
library(limma)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(ggpubr)
```

## Based on Seurat command

```{r}
# Define paths to save and load
path_to_save_seurat_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_seurat_dea.rds")
path_to_save_seurat_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_seurat_hm_gsea.rds")
path_to_save_seurat_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_seurat_go_gsea.rds")
path_to_save_seurat_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_seurat_kegg_gsea.rds")
path_to_save_seurat_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_seurat_reactome_gsea.rds")
path_to_save_seurat_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_seurat_wikipath_gsea.rds")
```

### Find biomarkers of monocytes between RT and CLL conditions

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Find markers between CLL and RT conditions  
Idents(pure_cd4_seurat) <- "new_cond" # we used this variable as it contains RT and CLL
pure_cd4_seurat <- PrepSCTFindMarkers(pure_cd4_seurat) # Prepare SCT assay for DE testing
cd4.ge <- FindMarkers(
  pure_cd4_seurat,
  ident.1 = "RT",
  ident.2 = "CLL",
  test.use = "wilcox",
  only.pos = FALSE,
  logfc.threshold = 0
)
dea.df <- cbind(gene = rownames(cd4.ge), cd4.ge)
avg_exp <- AverageExpression(pure_cd4_seurat, features = dea.df$gene, assay = "SCT") # Calculate the average expression for the genes
dea.df$baseMean <- rowMeans(avg_exp$SCT) # Calculate the mean across all relevant groups (e.g., global mean for the gene)

# Save the expression results
saveRDS(dea.df, path_to_save_seurat_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/cd4_seurat_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Load the saved expression data
dea.df <- readRDS(path_to_save_seurat_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=10}
# Volcano plot for Seurat pseudobulk analysis
genes_of_interest <- c(
  "TNFRSF9", "TIGIT", "PDCD1", "EOMES", "LAYN", "ENTPD1", "FOXP3", "CXCR6",         # RT CD4+ T cells: exhausted and Treg shifting                
  "PTGDR2", "CCR4", "FOS", "JUN", "DUSP1", "TCF7"                                   # CLL CD4+ T cells: Th2, acute stress response, and stemness
)

non_zero_min_pval <- sort(unique(dea.df$p_val_adj), decreasing = FALSE)[2] # Find minimum p-value which is not 0
dea.df$adapted_p_val <- ifelse(dea.df$p_val_adj == 0, non_zero_min_pval, dea.df$p_val_adj) # Change zero adj_p_val to non-zero minimum adj_p_val to avoid plotting error

volcano.plt <- volcano_plot(dea.df, 
                            FC=0.25, 
                            x_col="avg_log2FC",
                            y_col="adapted_p_val",
                            xlim=c(-10,10), 
                            ylim=c(-0.1,100), 
                            genes_of_interest=genes_of_interest,
                            boxedLabels = TRUE) 

volcano.plt
```

```{r, message=FALSE, warning=FALSE}
# MA plot for Seurat pseudobulk analysis
ma_data <- dea.df
colnames(ma_data)[colnames(ma_data) == "avg_log2FC"] <- "log2FoldChange"
colnames(ma_data)[colnames(ma_data) == "p_val_adj"]  <- "padj"

ma.plt <- ggmaplot(ma_data, 
                   fdr = 0.05,            # Threshold for significance
                   fc = 2,                # Threshold for FC
                   size = 0.4,            # Dot size
                   palette = c("#B31B21", "#1465AC", "darkgray"),
                   genenames = as.vector(ma_data$gene),
                   top = 0, label.select = genes_of_interest, # set top = 0, then specify genes using label.select argument
                   legend = "right", 
                   font.label = c("bold", 11), 
                   label.rectangle = TRUE,
                   font.legend = "bold",
                   font.main = "bold",
                   ggtheme = ggplot2::theme_minimal(),
                   xlab = "Mean Expression",
                   ylab = "Log2 Fold Change"
                   )

ma.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/seurat_cd4_volcano.png"), plot=volcano.plt, dpi=500, height=7, width=9)
ggsave(filename=paste0(path_to_save_figure, "/seurat_cd4_ma.png"), plot=ma.plt, dpi=500, height=3.5, width=5)
```

Overall, the gene expression analysis indicates that RT CD4⁺ T cells predominantly shift towards exhausted and regulatory subtypes, whereas CLL CD4⁺ T cells are primarily characterised by Th2, acute stress response, and stem-like or memory-precursor phenotypes.

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea.df$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

### GSEA summary

We then summarised the common terms found from Hallmark, GO:BP, and canonical pathways.

```{r, message=FALSE, warning=FALSE}
# Bar plot of GSEA 
gsea_bp <- rbind(
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_T_CELL_MEDIATED_IMMUNITY",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_REGULATION_OF_LYMPHOCYTE_MEDIATED_IMMUNITY",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_CELL_GROWTH",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_MAPK_SIGNALING_PATHWAY",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_TNFA_SIGNALING_VIA_NFKB",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL1_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL4_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_RANKLRANK_SIGNALING",]
)

gsea_bp$Description <- gsub("HALLMARK_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("GOBP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("KEGG_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("REACTOME_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("WP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("_V1", "", gsea_bp$Description)
gsea_bp$Description <- gsub("T_CELL_MEDIATED_IMMUNITY", "T_CELL_IMMUNITY", gsea_bp$Description)
gsea_bp$Description <- gsub("REGULATION_OF_LYMPHOCYTE_MEDIATED_IMMUNITY", "LYMPHOCYTE_REGULATION", gsea_bp$Description)
gsea_bp$Description <- gsub("CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION", "CYTOKINE_INTERACTION", gsea_bp$Description)
gsea_bp$Description <- gsub("TNFA_SIGNALING_VIA_NFKB", "TNFA_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("RANKLRANK_SIGNALING", "RANKL-RANK_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("MAPK_SIGNALING_PATHWAY", "MAPK_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("_", " ", gsea_bp$Description)

gsea_bp$Description <- factor(gsea_bp$Description, levels = gsea_bp$Description)
gsea_bp$Direction <- "Up"
gsea_bp$Direction[gsea_bp$NES<0] <- "Down"

gsea_bar.plt <- ggplot(gsea_bp, aes(x=Description, y=NES, fill=Direction)) +
  geom_bar(stat = "identity", color="black", width = 0.75) +  
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, color = "black"),
    plot.margin = margin(t = 10, r = 10, b = 100, l = 65, unit = "pt"),
    legend.position = "none"
  ) +
  geom_hline(yintercept = 0) + 
  ylab("NES") + xlab(NULL) + 
  scale_fill_manual(values=c("#357EBD", "#D43F3A"))

gsea_bar.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=12}
# GSEA plots
terms_of_interest <- c(
  "GOBP_T_CELL_MEDIATED_IMMUNITY", "GOBP_REGULATION_OF_LYMPHOCYTE_MEDIATED_IMMUNITY", "KEGG_CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION",
  "GOBP_CELL_GROWTH", "KEGG_MAPK_SIGNALING_PATHWAY", "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "WP_IL1_SIGNALING", "WP_IL4_SIGNALING", "WP_RANKLRANK_SIGNALING"
  )

gsea_plot.lst <- lapply(terms_of_interest, function(term){
  term_system <- strsplit(term, split = "_")[[1]][1]

  adapted_term <- term %>% 
    str_remove("^(HALLMARK|GOBP|KEGG|REACTOME|WP)_") %>%
    str_replace("T_CELL_MEDIATED_IMMUNITY", "T_CELL_IMMUNITY") %>%
    str_replace("REGULATION_OF_LYMPHOCYTE_MEDIATED_IMMUNITY", "LYMPHOCYTE_REGULATION") %>%
    str_replace("CYTOKINE_CYTOKINE_RECEPTOR_INTERACTION", "CYTOKINE_INTERACTION") %>%
    str_replace("TNFA_SIGNALING_VIA_NFKB", "TNFA_SIGNALING") %>%
    str_replace("RANKLRANK_SIGNALING", "RANKL-RANK_SIGNALING") %>%
    str_replace("MAPK_SIGNALING_PATHWAY", "MAPK_SIGNALING") %>%
    str_replace_all("_", " ")
  
  if (term_system == "HALLMARK"){
    gseaplot(HM_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "GOBP"){
    gseaplot(GOBP_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "KEGG") {
    gseaplot(KEGG_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "REACTOME") {
    gseaplot(REACTOME_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else {
    gseaplot(WIKIPATH_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  }
})

gsea.plt <- wrap_plots(gsea_plot.lst, ncol=3)
gsea.plt
```

```{r, eval=FALSE}
# Save GSEA results
saveRDS(HM_gsea, path_to_save_seurat_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_seurat_go_gsea)
saveRDS(KEGG_gsea, path_to_save_seurat_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_seurat_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_seurat_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_seurat_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_seurat_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_seurat_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_seurat_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_seurat_wikipath_gsea.csv"))

# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/cd4_gsea.png"), plot=gsea_bar.plt, dpi=500, height=5, width=7)
ggsave(filename=paste0(path_to_save_figure, "/cd4_gsea_plt.png"), plot=gsea.plt, dpi=500, height=8, width=12)
```

## Based on DESeq2

```{r}
# Define paths to save the results
path_to_save_deseq2_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_deseq2_dea.rds")
path_to_save_deseq2_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_deseq2_hm_gsea.rds")
path_to_save_deseq2_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_deseq2_go_gsea.rds")
path_to_save_deseq2_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_deseq2_kegg_gsea.rds")
path_to_save_deseq2_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_deseq2_reactome_gsea.rds")
path_to_save_deseq2_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd4_deseq2_wikipath_gsea.rds")
```

### Create an DESeq2 object

```{r, message=FALSE, warning=FALSE}
# Prepare single-cell meta-data for DESeq2
pure_cd4_seurat$ref[pure_cd4_seurat$ref == "Hing_2022"] <- "Hing_2023" # Update reference name
pure_cd4_seurat$patient_cond <- paste0("case_", pure_cd4_seurat$case, "_", pure_cd4_seurat$ref, "_", pure_cd4_seurat$new_cond) 
patient_cond_cls <- unique(pure_cd4_seurat$patient_cond)
RT_subjects <- patient_cond_cls[grep("RT", patient_cond_cls)]
RT_subjects <- gsub(pattern = "_", replacement = "-", x=RT_subjects) # change "_" to "-" as DESeq2 converted sample names from "_" to "-"

# Create pseudocount matrix for tumour cell object
cts <- AggregateExpression(pure_cd4_seurat, 
                           group.by = "patient_cond",
                           assays = "RNA",
                           slot = "counts",
                           resturn.seurat = FALSE) 
cts <- cts$RNA
pseudo_count.mat <- as.data.frame(cts)

# Create metadata for DESeq2 object
batch_map <- c("Penter (2021)" = "1", "Nadeu (2022)" = "2", "Rejeski (2022)" = "3", "Hing (2023)" = "4", "Parry (2023a)" = "5", 
               "Parry (2023b)" = "6", "Rigo (2024)" = "7")
colData <- data.frame(samples = colnames(pseudo_count.mat))
colData$condition <- ifelse(colData$samples %in% RT_subjects, "RT", "CLL")
colData$study <- stringr::word(colData$samples, start = -3, end = -2, sep = fixed("-"))
colData$study <- sub("^(.*)-([0-9].*)$", "\\1 (\\2)", colData$study) # Substitiute string like "Parry-2023b" to "Parry (2023b)"
colData$study <- factor(colData$study, level = c("Penter (2021)", "Nadeu (2022)", "Rejeski (2022)", "Hing (2023)", 
                                                 "Parry (2023a)", "Parry (2023b)", "Rigo (2024)"))
colData$batch <- batch_map[colData$study]
colData$batch <- factor(colData$batch)
rownames(colData) <- colData$samples

# Create a DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = pseudo_count.mat, 
                              colData = colData,
                              design = ~ batch + condition)

# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

### Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6.5}
# Variance stabilised transformation
vsdata <- vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
design_matrix <- model.matrix(~ condition, data = colData(dds))
vs_mat_corrected <- limma::removeBatchEffect(
  vs_mat, 
  batch = colData(dds)$batch, 
  design = design_matrix 
)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2               # variance of each PC
pca_var_percent <- pca_var / sum(pca_var) * 100

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     study = colData(dds)$study)

# Plot PCA
pca.plt <- ggplot(pca_df, aes(PC1, PC2, color = study, shape = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  scale_color_manual(values = c(
    "Penter (2021)" = "#1f77b4", 
    "Nadeu (2022)" = "#ff7f0e",
    "Rejeski (2022)" = "#9632B8" ,
    "Hing (2023)" = "#D43F3A" ,
    "Parry (2023a)" = "#5CB85C",
    "Parry (2023b)" = "#FFB900",
    "Rigo (2024)" = "#46B8DA")) +
  scale_shape_manual(values = c("CLL" = 16, "RT" = 17)) +
  guides(
    color = guide_legend(title = "Study"),
    shape = guide_legend(title = "Condition")
  ) +
  theme_bw() +
  theme(
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 11)
  )

pca.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/deseq2_cd4_pca.png"), plot=pca.plt, dpi=500, height=4.5, width=6.5)
```

### Run DESeq2

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Run DESeq2 for the data after outlier removal
keep <- rowSums(counts(dds) >= 10) >= 3 # Keep genes with at least 10 counts in at least 3 samples
noise_genes <- c(
  grep("^IG[HKL][VDJ]", rownames(dds), value = TRUE),  # Ig genes
  grep("^HLA-", rownames(dds), value = TRUE),          # Patient ID (HLA)
  grep("^MT-", rownames(dds), value = TRUE),           # Mitochondrial stress
  "XIST", "RPS4Y1", "KDM5D", "UTY", "DDX3Y",           # Sex genes
  "S100A8", "S100A9", "S100A12", "CXCL8"               # Acute stress response
)

dds <- dds[keep & (!rownames(dds) %in% noise_genes), ]
dds$condition <- relevel(dds$condition, ref = "CLL")
dds <- DESeq(dds)

# Generate a table for gene expression
res <- results(dds, contrast = c("condition", "RT", "CLL"), independentFiltering = FALSE)
res <- lfcShrink(dds, contrast = c("condition", "RT", "CLL"), res = res, type = "normal")
res$direction <- "NS"
res$direction[res$padj < 0.05 & res$log2FoldChange > 0.25] <- "Up"
res$direction[res$padj < 0.05 & res$log2FoldChange < (-0.25)] <- "Down"
dea.df <- as.data.frame(res)

# Save the expression results
saveRDS(dea.df, path_to_save_deseq2_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/cd4_deseq2_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Show the expression results
dea.df <- readRDS(path_to_save_deseq2_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=10}
# Volcano plot
genes_of_interest <- c(
  "BATF", "GIMAP4", "RAC2", "LBH",        # RT CD4+ T cells: exhausted and Treg shifting                
  "GADD45B"                               # CLL CD4+ T cells: Th2, acute stress response, and stemness
)

volcano.plt <- volcano_plot(dea.df, 
                            FC=0.25, 
                            x_col="log2FoldChange",
                            y_col="padj",
                            xlim=c(-1.6,1.6), 
                            ylim=c(-0.1,2), 
                            genes_of_interest=genes_of_interest,
                            boxedLabels = TRUE) 

volcano.plt
```

```{r, message=FALSE, warning=FALSE}
# MA plot 
ma_data <- dea.df

ma.plt <- ggmaplot(ma_data, 
                   fdr = 0.05,            # Threshold for significance
                   fc = 1.19,                # Threshold for FC
                   size = 0.4,            # Dot size
                   palette = c("#B31B21", "#1465AC", "darkgray"),
                   genenames = as.vector(ma_data$gene),
                   top = 0, label.select = genes_of_interest, # set top = 0, then specify genes using label.select argument
                   legend = "right", 
                   font.label = c("bold", 11), 
                   label.rectangle = TRUE,
                   font.legend = "bold",
                   font.main = "bold",
                   ggtheme = ggplot2::theme_minimal(),
                   xlab = "Mean Expression",
                   ylab = "Log2 Fold Change"
                   )

ma.plt
```

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df) 
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save the enrichment results
saveRDS(HM_gsea, path_to_save_deseq2_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_deseq2_go_gsea)
saveRDS(KEGG_gsea, path_to_save_deseq2_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_deseq2_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_deseq2_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_deseq2_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_deseq2_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_deseq2_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_deseq2_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd4_deseq2_wikipath_gsea.csv"))
```

# Session Information

```{r}
sessionInfo()
```














