---
title: "Microenvironment_CD8+_T_cell_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

In this analysis, we focused on CD8+ T cells, from the integrated dataset by performing pseudobulk analysis between RT and CLL in this cell type. Moreover, we also annotated level-2 subtypes of these cells.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tibble)
library(SingleR)
library(celldex)
library(sctransform)
library(cowplot)
library(DT)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_cd8_obj <- here::here("scRNAseq/meta_analysis/R_objects/cd8_seu.rds")
path_to_save_cd8_markers <- here::here("scRNAseq/meta_analysis/R_objects/cd8_marker.rds")
path_to_save_filtered_cd8_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered_cd8_seu.rds")
path_to_save_filtered_cd8_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered_cd8_marker.rds")
path_to_save_pure_cd8_obj <- here::here("scRNAseq/meta_analysis/R_objects/pure_cd8_seu.rds")
path_to_save_pure_cd8_markers <- here::here("scRNAseq/meta_analysis/R_objects/pure_cd8_marker.rds")
path_to_save_figure <- here::here("figures")
```

```{r}
# Thresholds
min_log2FC <- 0.25
alpha <- 0.001
```

```{r}
# Set seed for reproducibility
set.seed(1)
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the CD8+ T cell object
cd8_seurat <- readRDS(path_to_cd8_obj)
```

# Visualisation

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Cluster visualisation
group_by <- c("SCT_snn_res.0.1", "SCT_snn_res.0.2", "SCT_snn_res.0.3")
umap_plots <- lapply(group_by, function (group){
  DimPlot(cd8_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 2,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

cd8_umap.plt <- wrap_plots(umap_plots, ncol=3)
cd8_umap.plt
```

```{r, fig.height=7, fig.width=12, warning=FALSE, message=FALSE}
# Plot based on cell biomarker expression
FeaturePlot(cd8_seurat, 
            features = c("HBM", "HBA1", "CD79A", "CD14", "CD3D", "CD3E", 
                         "CD4", "CD8A", "CD8B", "KLRF1", "FCGR3A", "TYROBP"),
            reduction = "umap") 
```

# CD8+ T cell purification

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(cd8_seurat) <- "SCT_snn_res.1.2"
markers <- FindAllMarkers(cd8_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_cd8_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/cd8_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_cd8_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

From the marker identification we found contaminated clusters, for example, RBC for cluster 8, 13, and 14, and B cells for cluster 5, 7, 10, and 17. We then removed them.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the CD8+ T cell clusters with reprocessing
filtered_cd8_seurat <- subset(cd8_seurat, idents = c(5, 7, 8, 10, 13, 14, 17), invert = TRUE)
filtered_cd8_seurat <- SCTransform(filtered_cd8_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered_cd8_seurat <- RunPCA(filtered_cd8_seurat)
filtered_cd8_seurat <- RunUMAP(filtered_cd8_seurat, dims = 1:30, reduction = "harmony")
filtered_cd8_seurat <- FindNeighbors(filtered_cd8_seurat, dims = 1:30, reduction = "harmony")        
filtered_cd8_seurat <- FindClusters(
  filtered_cd8_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# Save the object
saveRDS(filtered_cd8_seurat, path_to_save_filtered_cd8_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate subclusters for cluster 1 and 2 
filtered_cd8_seurat$seurat_subclusters <- filtered_cd8_seurat$SCT_snn_res.0.2

for (cluster_id in c("1", "2")) {
  Idents(filtered_cd8_seurat) <- "seurat_subclusters"
  
  filtered_cd8_seurat <- FindSubCluster(
    object          = filtered_cd8_seurat,
    cluster         = cluster_id,                  
    graph.name      = "SCT_snn",            
    subcluster.name = "seurat_subclusters", 
    resolution      = 0.2
  )
}

# Save the object
saveRDS(filtered_cd8_seurat, path_to_save_filtered_cd8_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(filtered_cd8_seurat) <- "seurat_subclusters"
markers <- FindAllMarkers(filtered_cd8_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered_cd8_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered_cd8_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered_cd8_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

From the biomarker results, all clustes were CD8+ T cells. We then annotated their subtypes in the next step.

# Subtype annotation

```{r, eval=FALSE}
# Change the obeject and marker name
pure_cd8_seurat <- filtered_cd8_seurat
saveRDS(pure_cd8_seurat, path_to_save_pure_cd8_obj)
saveRDS(markers, path_to_save_pure_cd8_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/pure_cd8_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the clusters
pure_cd8_seurat <- readRDS(path_to_save_pure_cd8_obj)
group_by <- c("SCT_snn_res.0.2", "seurat_subclusters", "SCT_snn_res.0.3")
umap_plots <- lapply(group_by, function (group){
  DimPlot(pure_cd8_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 2,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

pure_cd8_umap.plt <- wrap_plots(umap_plots, ncol=3)
pure_cd8_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the umap
group_by <- c("ref", "tissue", "group")
umap_plots <- lapply(group_by, function (group){
  DimPlot(pure_cd8_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 2,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

pure_cd8_umap.plt <- wrap_plots(umap_plots, ncol=3)
pure_cd8_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=10}
# Visualise the cases
pure_cd8_seurat$patient <- paste0(pure_cd8_seurat$case, "_", pure_cd8_seurat$ref)
case_cd8.plt <- DimPlot(pure_cd8_seurat, 
                       reduction = "umap", 
                       group.by = "patient", 
                       pt.size = 2,
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

case_cd8.plt
```

Based on Parry (2023b), although the original clinical condition for cases MDAC-ACC-17_Parry_2023b, MDAC-ACC-18_Parry_2023b, and MDAC-ACC-29_Parry_2023b was CLL, their tumour cells were annotated as CCND2hi RT based on RT-associated marker expression and predominant S-phase cell-cycle scores (see Tumour Subtype Annotation section).

Similarly, in Penter (2021), tumour cells from case 1 were annotated as BCAT1hiCCND2hi RT, with high G2/M-phase cell-cycle scores. Accordingly, we reclassified the clinical condition of these samples from CLL to RT for downstream analyses.

```{r, eval=FALSE}
# Change the new condition for case MDAC-ACC-17_Parry_2023b, MDAC-ACC-18_Parry_2023b, and MDAC-ACC-29_Parry_2023b
selected_cases <- c("MDAC-ACC-17_Parry_2023b", "MDAC-ACC-18_Parry_2023b", "MDAC-ACC-29_Parry_2023b", "1_Penter_2021")
pure_cd8_seurat$new_cond <- pure_cd8_seurat$group
pure_cd8_seurat@meta.data[pure_cd8_seurat$patient %in% selected_cases, "new_cond"] <- "RT"
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=6}
# Visualise the adapted condition
new_cond_cd8.plt <- DimPlot(pure_cd8_seurat, 
                       reduction = "umap", 
                       group.by = "new_cond", 
                       pt.size = 2,
                       repel=TRUE,
                       raster=TRUE, 
                       label=TRUE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

new_cond_cd8.plt
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_pure_cd8_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We annotated the myeloid subtypes as literature below:

- [CD8+ T cell states in human cancer: insights from single-cell analysis](https://www.nature.com/articles/s41568-019-0235-4)
- [Phenotypic and spatial heterogeneity of CD8+ tumour infiltrating lymphocytes](https://link.springer.com/article/10.1186/s12943-024-02104-w)
- [KLRG+ invariant natural killer T cells are long-lived effectors](https://www.pnas.org/doi/full/10.1073/pnas.1406240111)
- [Human CD8+ T-Cell Populations That Express Natural Killer Receptors](https://immunenetwork.org/DOIx.php?id=10.4110/in.2023.23.e8)
- [Unraveling the phenotypic states of human innate-like T cells: Comparative insights with conventional T cells and mouse models](https://www.cell.com/cell-reports/fulltext/S2211-1247(24)01056-8)
- [Granzyme M: behind enemy lines](https://www.nature.com/articles/cdd2013189#:~:text=This%20highly%20specific%20expression%20of%20hGrM%20by,a%20role%20for%20GrM%20in%20innate%20immunity.)
- [Potential mechanisms of cancer stem-like progenitor T-cell bio-behaviours](https://onlinelibrary.wiley.com/doi/10.1002/ctm2.1817)
- [Developmental Relationships of Four Exhausted CD8+ T Cell Subsets Reveals Underlying Transcriptional and Epigenetic Landscape Control Mechanisms](https://www.sciencedirect.com/science/article/pii/S1074761320301722)

| Cluster | Markers                                                         | Annotation    |
|---------|-----------------------------------------------------------------|---------------|
| 0       | GNLY, GZMB, PRF1                                                | Teff          |
| 1_0     | ZNF683, PRF1, FGFBP2                                            | Tem           |
| 1_1     | ZNF683, PRF1, FGFBP2, FCRL6, NKG7                               | Tem           |
| 1_2     | GZMK, PDCD1, CD28, LAG3, TIGIT, HAVCR2, ENTPD1                  | Pre Tex       |
| 2_0     | ID2, CXCR3, EOMES, GZMK                                         | Tint          |
| 2_1     | TCF7, PDCD1, CD28                                               | Tpex          |
| 2_2     | TCF7, CCR7, IL7R, LTB, SELL, CD27, PTPRCAP                      | Tcm           |
| 3       | CX3CR1, FCGR3A, FGFBP2, GNLY, FCRL6, NKG7, CXCR3, S1PR1, S1PR5  | Temra         |
| 4       | GZMA, IFNG, CCL3, XCL2, MYADM, CAPG, NR4A1, NR4A2, CD69         | Trm-like      |
| 5       | LEF1, TCF7, CCR7, IL7R, S1PR1                                   | Tn            |
| 6       | KLRG1, GZMM, KLRB1, ZBTB16, NCAM1                               | NKT-like      |
| 7       | PDCD1, LAG3, TIGIT, HAVCR2, ENTPD1, IFNG, TOX                   | Tex           |

Abbreviations: Tn, naive T cell; Tcm, central memory T cell; Teff, effector T cell; Tem, effector memory T cell; Temra, terminally differentiated effector memory T cell; Trm-like, tissue-resident memory (Trm)-like T cell; CTL, cytotoxic T cell; NKT, NK-like T cell; Tint, intermediate T cell; Tpex, Progenitor exhausted T cell; Tex, exhausted T cell; Pre Tex, pre-exhausted T cell;

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=5.5}
# Plot UMAP for CD8+ T cell subtypes

# 1. Map Seurat clusters and create the annotation column
cluster_to_annotation <- c(
  "0" = "Teff", "1_0" = "Tem", "1_1" = "Tem", "1_2" = "Pre Tex", "2_0" = "Tint", "2_1" = "Tpex", "2_2" = "Tcm", 
  "3" = "Temra", "4" = "Trm-like", "5" = "Tn", "6" = "NKT-like", "7" = "Tex"
  )

pure_cd8_seurat$level2_annotate <- 
  as.vector(cluster_to_annotation[as.character(pure_cd8_seurat$seurat_subclusters)])

# 2. Set level order
new_levels <- c(
  # Group1: naive and central memory cells 
  "Tn", "Tcm", "Tint", "Tpex", 
  
  # Group2: effector memory, tissue-resident, and CTL  
  "Tem", "Teff", "Temra", "Trm-like",  
  
  # Group3: innate-like T cell
  "NKT-like",

  # Group3: exhausted T cell  
  "Pre Tex", "Tex" 
)

pure_cd8_seurat$level2_annotate <- factor(
  pure_cd8_seurat$level2_annotate,
  levels = new_levels
) 

# 3. Define colours and labels
new_colours <- c(
  "#B3B3B3", "#1F78B4", "#6BAED6", "#A6CEE3",   # Group 1 (4 colours)
  "#4DAF4A", "#008080", "#C28BCD", "#F38D30",   # Group 2 (4 colours)
  "#FFD92F",                                    # Group 3 (1 colour)
  "#E25644", "#B04968"                          # Group 4 (3 colours)
) 

custom_labels <- list(
  bquote("Tn"), bquote("Tcm"), bquote("Tint"), bquote("Tpex"),
  bquote("Tem"), bquote("Teff"), bquote("Temra"), bquote("Trm-like"),
  bquote("NKT-like"), 
  bquote("Pre Tex"), bquote("Tex")
  )

# 4. Plot UMAP
cd8_subtype.plt <- DimPlot(pure_cd8_seurat, 
                          group.by = "level2_annotate", 
                          label = FALSE,
                          repel = TRUE,
                          pt.size = 2,
                          alpha = 1,
                          raster = TRUE
                          ) +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels,
    labels = custom_labels
  ) +
  labs(y = "UMAP 2", x = "UMAP 1") +
  ggtitle("CD8+ T cell subtypes")

cd8_subtype.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Dot plot for biomarker expressions based on annotated cells
features <- c("TCF7", "LEF1", "SELL", "CCR7", "LTB", "GPR183", "EOMES","CD27", "CD28", "GZMK", "GZMA", 
              "GZMB", "PRF1", "GNLY", "NKG7", "CCL3", "CCL4", "S1PR1", "S1PR5", "CX3CR1", "FCGR3A", 
              "FGFBP2", "ZNF683", "GZMM", "KLRB1", "NCAM1", "PDCD1", "LAG3", "TIGIT", "HAVCR2"
              )
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

cd8_markers.plt <- DotPlot(pure_cd8_seurat, features=features, group.by = "level2_annotate") +
  RotatedAxis() +
  scale_colour_gradientn(colours = colour_pal) +
  scale_y_discrete(
      limits = rev(new_levels),
      labels = rev(custom_labels)
    ) +
  labs(
    colour = "Average\nExpression",
    size = "Percent\nExpressed"
  )

cd8_markers.plt
```

```{r, message=FALSE, warning=FALSE}
# Plot subtype population based on tissue and condition
proportion_tissue_df <- pure_cd8_seurat@meta.data %>%
  dplyr::mutate(level2_annotate = factor(level2_annotate, levels = new_levels)) %>%
  dplyr::count(tissue, level2_annotate) %>%
  dplyr::group_by(tissue) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n) # Remove n column

proportion_cond_df <- pure_cd8_seurat@meta.data %>%
  dplyr::mutate(level2_annotate = factor(level2_annotate, levels = new_levels)) %>%
  dplyr::count(new_cond, level2_annotate) %>%
  dplyr::group_by(new_cond) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n) # Remove n column

tissue_cell_pop.plt <- ggplot(proportion_tissue_df, aes(x = tissue, 
                                                    y = proportion, 
                                                    fill = level2_annotate)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "CD4+ T cell subtype" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Tissue", 
    title = "Tissue"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

cond_cell_pop.plt <- ggplot(proportion_cond_df, aes(x = new_cond, 
                                                    y = proportion, 
                                                    fill = level2_annotate)) +
  geom_col(position = "stack", width = 0.5) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "CD8+ T cell subtype" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Condition", 
    title = "Condition"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

# Combine the plots
adapted_tissue_cell_pop.plt <- tissue_cell_pop.plt +
  theme(legend.position = "none",
        axis.title.x = element_blank()) 

adapted_cond_cell_pop.plt <- cond_cell_pop.plt + 
  theme(legend.position = "right", 
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) 

combined_cell_pop.plt <- adapted_tissue_cell_pop.plt + adapted_cond_cell_pop.plt
combined_cell_pop.plt
```

```{r, eval=FALSE}
# Save the CD8+ T cell object
saveRDS(pure_cd8_seurat, path_to_save_pure_cd8_obj)

# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/cd8_umap.png"), plot=cd8_subtype.plt, dpi=500, height=4.5, width=5.5)
ggsave(filename=paste0(path_to_save_figure, "/cd8_dot.png"), plot=cd8_markers.plt, dpi=500, height=4, width=13)
ggsave(filename=paste0(path_to_save_figure, "/cd8_cell_pop.png"), plot=combined_cell_pop.plt , dpi=500, height=4, width=7)
```

# Trajectory analysis

In this section, we used the Monocle 3 package to identify developmental trajectories of CD8+ T cells using naive T cell subtype as a root node.

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(monocle3)
library(igraph)
library(SeuratWrappers)
library(paletteer)
```

```{r}
# Define path to save
path_to_save_cd8_monocle_obj <- here::here("scRNAseq/meta_analysis/R_objects/cd8_monocle_cds.rds")
path_to_save_cd8_monocle_dea <- here::here("scRNAseq/meta_analysis/R_objects/cd8_monocle_dea.rds")
path_to_save_cd8_monocle_module <- here::here("scRNAseq/meta_analysis/R_objects/cd8_monocle_module.rds")
path_to_save_cd8_monocle_agg_mat <- here::here("scRNAseq/meta_analysis/R_objects/cd8_monocle_agg_mat.rds")
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create monocle object and generate graph (use data from seurat)
cds <- as.cell_data_set(pure_cd8_seurat) # Convert the seurat object to monocle object using SeuratWrappers
cds <- preprocess_cds(cds)

# Create gene_short_name
rowData(cds)$gene_short_name <- row.names(rowData(cds))

# Assign seurat clusters to monocle and create a partition
cds <- monocle3::cluster_cells(cds, reduction_method = "UMAP", 
                               cluster_method = "louvain", 
                               resolution = 0.5)
cds@clusters$UMAP$clusters <- pure_cd8_seurat$SCT_snn_res.0.6

# Assign UMAP coordination
cds@int_colData$reducedDims$UMAP <- pure_cd8_seurat@reductions$umap@cell.embeddings
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=17}
# Monocle plot before trajectory analysis
cds <- readRDS(path_to_save_cd8_monocle_obj)
group_by <- c("cluster", "partition", "level2_annotate")
umap_plots <- lapply(group_by, function (group){
  plot_cells(cds, 
             color_cells_by = group, 
             label_groups_by_cluster = FALSE, 
             group_label_size = 5,
             cell_size = 0.1) +  
    theme(legend.position = "right")
})

monocle_umap.plt <- wrap_plots(umap_plots, ncol=3)
monocle_umap.plt
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Learn trajectory
cds <- learn_graph(cds, use_partition = TRUE, close_loop = TRUE,
                   learn_graph_control = list(minimal_branch_len = 5, rann.k = 50))
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=7}
# Plot after learning trajectory
monocle_cluster.plt <- plot_cells(cds, 
                                  color_cells_by = "cluster",
                                  graph_label_size=5,
                                  cell_size = 0.1,
                                  show_trajectory_graph = TRUE
                                  ) + 
  ggtitle("Seurat clusters") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

monocle_cluster.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=7}
# Plot root
root.plt <- plot_cells(cds, label_principal_points = TRUE) + 
  ggtitle("Root principal node") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

root.plt
```

```{r, fig.height=9, fig.width=8, warning=FALSE, message=FALSE}
# Plot based on cell biomarker expression
FeaturePlot(pure_cd8_seurat, 
            features = c("CCR7", "LEF1", "TCF7", "SELL", "GZMK", "LTB"),
            reduction = "umap") 
```

From the Tcm subtype we set Y_200 as a root node as this point expressed the highest TCF7 and SELL.

## Calculate pseudotime

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set a root to order cells in pseudotime
cds <- order_cells(cds, reduction_method = "UMAP", root_pr_nodes = c("Y_666", "Y_200", "Y_959"))
```

```{r, fig.height=5, fig.width=7}
# Plot the pseudotime
monocle_pseudotime.plt <- plot_cells(cds, 
                                     color_cells_by = "pseudotime",
                                     label_cell_groups=FALSE,
                                     label_leaves=FALSE,
                                     label_branch_points=FALSE,
                                     graph_label_size=1.5,
                                     cell_size = 0.1) +
  ggtitle("Pseudotime") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

monocle_pseudotime.plt
```

```{r, fig.height=5, fig.width=5}
# Plot the pseudotime with trajectory path with Seurat style

# 1. Extract principal graph
pg_aux <- principal_graph_aux(cds)$UMAP

# 2. Convert to edge list
edges <- as_data_frame(principal_graph(cds)$UMAP, what = "edges")
nodes <- as.data.frame(t(pg_aux$dp_mst))
nodes$name <- rownames(nodes)
root_node <- cds@principal_graph_aux$UMAP$root_pr_nodes

# 3. Merge edges with coordinates
trajectory_df <- edges %>%
  dplyr::left_join(nodes, by = c("from" = "name")) %>%
  dplyr::rename(
    UMAP_1_from = umap_1,
    UMAP_2_from = umap_2
  ) %>%
  dplyr::left_join(nodes, by = c("to" = "name")) %>%
  dplyr::rename(
    UMAP_1_to = umap_1,
    UMAP_2_to = umap_2
  )

seurat_df <- data.frame(
  Embeddings(pure_cd8_seurat, "umap"),
  pseudotime = pseudotime(cds)
)

root_df <- nodes %>%
  dplyr::filter(name %in% root_node)

# 4. Plot
trajectory_no_annotate.plt <- ggplot(seurat_df, aes(umap_1, umap_2)) +
  geom_point(aes(color = pseudotime), size = 0.5, alpha=1) +
  geom_segment(
    data = trajectory_df,
    aes(
      x = UMAP_1_from,
      y = UMAP_2_from,
      xend = UMAP_1_to,
      yend = UMAP_2_to
    ),
    inherit.aes = FALSE,
    colour = "black",
    linewidth = 0.7
  ) +
  geom_point(
    data = root_df,
    aes(umap_1, umap_2),
    inherit.aes = FALSE,
    shape = 21,          # circle with fill
    size = 6,
    fill = "white",
    colour = "black",
    stroke = 1
  ) +
  geom_text(
    data = root_df,
    aes(umap_1, umap_2),
    label = c("1", "2", "3"),
    inherit.aes = FALSE,
    size = 5,
    fontface = "bold"
  ) +
  scale_color_gradientn(
    name = "Pseudotime",
    colors = paletteer_c("grDevices::RdYlBu", 30, direction = -1)
  ) +
  theme_void() +
  theme(
    legend.position = "top" 
  )  
  
annotate_trajectory.plt <- trajectory_no_annotate.plt +  
  annotate("text", x = -6, y = 5, label = "Tn", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -10, y = -2.2, label = "Temra", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 2.8, y = 14.5, label = "Trm-like", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -2.7, y = 5.5, label = "Tcm", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 0, y = 8.5, label = "Tint", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 5.8, y = 5, label = "Pre Tex", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 5, y = 9, label = "Tex", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 6.8, y = -1, label = "Tem", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -5, y = -8, label = "Teff", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -4.2, y = -2, label = "NKT-like", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -3, y = 2, label = "Tpex", 
             color = "black", fontface = "plain", size = 4) + 
  annotate(
    "segment", 
    x = -2.2, y = 1.9,       # Start point (near the 'Tpex' text)
    xend = 0, yend = 2.5,    # End point (the actual 'Tpex' cells/node)
    colour = "black",
    linewidth = 0.5
  )

annotate_trajectory.plt
```

```{r, eval=FALSE}
# Save the monocle object
saveRDS(cds, path_to_save_cd8_monocle_obj)

# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/cd8_trajectory.png"), plot=annotate_trajectory.plt, dpi=500, height=5, width=5)
```

## Gene expression across pseudotime

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Group genes into modules
cd8_cds_pr_test_res <- monocle3::graph_test(cds, neighbor_graph = "principal_graph", cores = 4)
monocle_degs <- row.names(subset(cd8_cds_pr_test_res, q_value < 0.05))
gene_modules_df <- find_gene_modules(cds[monocle_degs,], resolution=1e-3)
cell_group_df <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=colData(cds)$level2_annotate)
agg_mat <- aggregate_gene_expression(cds, gene_modules_df, cell_group_df)
row.names(agg_mat) <- paste0("Module ", row.names(agg_mat))

# Save the module results
write.csv(cd8_cds_pr_test_res, here::here("scRNAseq/meta_analysis/supplm_file/cd8_trajectory_gene_expr.csv"))
saveRDS(cd8_cds_pr_test_res, path_to_save_cd8_monocle_dea)
saveRDS(gene_modules_df, path_to_save_cd8_monocle_module)
saveRDS(agg_mat, path_to_save_cd8_monocle_agg_mat)
```

```{r, message=FALSE, warning=FALSE}
# Show the monocle dae
monocle_dea <- readRDS(path_to_save_cd8_monocle_dea)
DT::datatable(monocle_dea, options = list(scrollX = TRUE))
```

```{r, fig.height=6, fig.width=6}
# Plot heatmap 
library(pheatmap)
gene_modules_df <- readRDS(path_to_save_cd8_monocle_module)
agg_mat <- readRDS(path_to_save_cd8_monocle_agg_mat)
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)
pseudotime_heatmap.plt <- pheatmap::pheatmap(agg_mat, scale="column", fontsize = 11,
                                             color = colour_pal,
                                             angle_col = 45,
                                             clustering_method="ward.D2")

pseudotime_heatmap.plt
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Test a range of resolutions to appropriate modules
resolutions <- c(1e-3, 2e-3, 3e-3, 4e-3, 5e-3)

for(res in resolutions){
  modules <- find_gene_modules(cds[monocle_degs,], resolution = res)
  num_mods <- length(unique(modules$module))
  print(paste("Resolution:", res, "->", num_mods, "Modules"))
}
```

```{r}
# Collect gene signatures for fate subtypes based on gene modules
tex_gene <- gene_modules_df %>% filter(module %in% c(10, 12)) %>% pull(id)
temra_gene <- gene_modules_df %>% filter(module %in% c(1)) %>% pull(id)
teff_gene <- gene_modules_df %>% filter(module %in% c(2, 5)) %>% pull(id)
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 1 Tcm to Tex
cells_types_of_interest <- c("Tcm", "Tpex", "Tint", "Pre Tex", "Tex")
pseudotime_genes <- c("PDCD1", "HAVCR2", "TIGIT", "ISG15", "PCNA", "NDUFS3")

tex_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                           genes = pseudotime_genes, 
                                                           cell_annotations = cells_types_of_interest, 
                                                           pseudotime_boundary = 11,
                                                           facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Tcm to Tex") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

tex_genes_pseudotime
```

For the Tex fate in CD8+ T cells, this subtype was confirmed by upregulation in exhausted genes (PDCD1, HAVCR2, and TGIT). Moreover, this subtype also exhibited interferon-related and proliferation genes: ISG15 and PCNA. A mitochnondiral gene, such as NDUFS3, was also observed, suggesting that the oxidative phosphorylation was predominated in the exhausted state.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 2 Tcm to Teff
cells_types_of_interest <- c("Tcm", "Tint", "Tem", "Teff")
pseudotime_genes <- c("GZMB", "PRF1", "ZAP70", "MTOR", "PDCD6IP", "EGLN1")

teff_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                           genes = pseudotime_genes, 
                                                           cell_annotations = cells_types_of_interest, 
                                                           pseudotime_boundary = 20,
                                                           facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Tcm to Teff") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

teff_genes_pseudotime
```

From this fate, Teff cells exhibited higher cytotoxic activity, characterised by the upregulation of metabolic-related genes such as mTOR and cytotoxic effector genes including GZMB and PRF1, although the latter two genes declined towards the terminal state. Moreover, this subtype demonstrated enhanced TCR signalling, indicated by increased expression of ZAP70. However, the upregulation of PDCD6IP, an exhaustion-associated gene, together with EGLN1, a hypoxia-response gene, was also observed, suggesting that Teff cells may progressively acquire an exhausted phenotype.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 3 Tn to Temra
cells_types_of_interest <- c("Tn", "Temra")
pseudotime_genes <- c("FCGR3A", "FGFBP2", "GNLY", "NKG7", "CXCR3", "ENO1")

temra_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                           genes = pseudotime_genes, 
                                                           cell_annotations = cells_types_of_interest, 
                                                           pseudotime_boundary = 10,
                                                           facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Tn to Temra") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

temra_genes_pseudotime
```

Based on the results, the Temra fate was confirmed by the expression of canonical markers, including FCGR3A, FGFBP2, and GNLY. Moreover, the increased expression of ENO1 suggested that this subtype may progress towards a senescent-like state.

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/cd8_tex_gene_flow.png"), plot=tex_genes_pseudotime, dpi=500, height=8, width=6)
ggsave(filename=paste0(path_to_save_figure, "/cd8_teff_gene_flow.png"), plot=teff_genes_pseudotime, dpi=500, height=8, width=6)
ggsave(filename=paste0(path_to_save_figure, "/cd8_temra_gene_flow.png"), plot=temra_genes_pseudotime, dpi=500, height=8, width=6)
```

# Gene expression analysis between RT versus CLL

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(DESeq2)
library(limma)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(ggpubr)
```

## Based on Seurat command

```{r}
# Define paths to save and load
path_to_save_seurat_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_seurat_dea.rds")
path_to_save_seurat_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_seurat_hm_gsea.rds")
path_to_save_seurat_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_seurat_go_gsea.rds")
path_to_save_seurat_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_seurat_kegg_gsea.rds")
path_to_save_seurat_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_seurat_reactome_gsea.rds")
path_to_save_seurat_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_seurat_wikipath_gsea.rds")
```

### Find biomarkers of monocytes between RT and CLL conditions

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Find markers between CLL and RT conditions  
Idents(pure_cd8_seurat) <- "new_cond" # we used this variable as it contains RT and CLL
pure_cd8_seurat <- PrepSCTFindMarkers(pure_cd8_seurat) # Prepare SCT assay for DE testing
cd8.ge <- FindMarkers(
  pure_cd8_seurat,
  ident.1 = "RT",
  ident.2 = "CLL",
  test.use = "wilcox",
  only.pos = FALSE,
  logfc.threshold = 0
)
dea.df <- cbind(gene = rownames(cd8.ge), cd8.ge)
avg_exp <- AverageExpression(pure_cd8_seurat, features = dea.df$gene, assay = "SCT") # Calculate the average expression for the genes
dea.df$baseMean <- rowMeans(avg_exp$SCT) # Calculate the mean across all relevant groups (e.g., global mean for the gene)

# Save the expression results
saveRDS(dea.df, path_to_save_seurat_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/cd8_seurat_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Load the saved expression data
dea.df <- readRDS(path_to_save_seurat_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=10}
# Volcano plot for Seurat pseudobulk analysis
genes_of_interest <- c(
  "TIGIT", "PDCD1", "HAVCR2", "GZMK", "CX3CR1", "EOMES",        # RT CD8+ T cells: exhausted gene upregulation               
  "TCF7", "IL7R", "MTOR"                                        # CLL CD8+ T cells: stemness and metabolic adaptation
)

non_zero_min_pval <- sort(unique(dea.df$p_val_adj), decreasing = FALSE)[2] # Find minimum p-value which is not 0
dea.df$adapted_p_val <- ifelse(dea.df$p_val_adj == 0, non_zero_min_pval, dea.df$p_val_adj) # Change zero adj_p_val to non-zero minimum adj_p_val to avoid plotting error

volcano.plt <- volcano_plot(dea.df, 
                            FC=0.25, 
                            x_col="avg_log2FC",
                            y_col="adapted_p_val",
                            xlim=c(-10,10), 
                            ylim=c(-0.1,300), 
                            genes_of_interest=genes_of_interest,
                            boxedLabels = TRUE) 

volcano.plt
```

```{r, message=FALSE, warning=FALSE}
# MA plot for Seurat pseudobulk analysis
ma_data <- dea.df
colnames(ma_data)[colnames(ma_data) == "avg_log2FC"] <- "log2FoldChange"
colnames(ma_data)[colnames(ma_data) == "p_val_adj"]  <- "padj"

ma.plt <- ggmaplot(ma_data, 
                   fdr = 0.05,            # Threshold for significance
                   fc = 2,                # Threshold for FC
                   size = 0.4,            # Dot size
                   palette = c("#B31B21", "#1465AC", "darkgray"),
                   genenames = as.vector(ma_data$gene),
                   top = 0, label.select = genes_of_interest, # set top = 0, then specify genes using label.select argument
                   legend = "right", 
                   font.label = c("bold", 11), 
                   label.rectangle = TRUE,
                   font.legend = "bold",
                   font.main = "bold",
                   ggtheme = ggplot2::theme_minimal(),
                   xlab = "Mean Expression",
                   ylab = "Log2 Fold Change"
                   )

ma.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/seurat_cd8_volcano.png"), plot=volcano.plt, dpi=500, height=7, width=9)
ggsave(filename=paste0(path_to_save_figure, "/seurat_cd8_ma.png"), plot=ma.plt, dpi=500, height=3.5, width=5)
```

Briefly, RT CD8+ T cells upregulated genes involved in exhausted state while CLL CD8+ cells upregulated in stemness and metabolic associated genes.

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea.df$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

### GSEA summary

We then summarised the common terms found from Hallmark, GO:BP, and canonical pathways.

```{r, message=FALSE, warning=FALSE}
# Bar plot of GSEA 
gsea_bp <- rbind(
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_OXIDATIVE_PHOSPHORYLATION",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_PYRUVATE_METABOLISM",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_GLYCOLYSIS_GLUCONEOGENESIS",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_FATTY_ACID_METABOLISM",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_MYC_TARGETS_V1",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_MITOCHONDRIAL_GENE_EXPRESSION",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_METABOLIC_REPROGRAMMING_IN_COLON_CANCER",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_MITOTIC_SPINDLE",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_REGULATION_OF_CELL_DEVELOPMENT",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_CELL_CELL_SIGNALING",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_APOPTOSIS",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_HYPOXIA",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_ADAPTIVE_IMMUNE_RESPONSE",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_INSULIN_SIGNALING",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_WNT_SIGNALING_PATHWAY",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_MAPK_SIGNALING_PATHWAY",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_ERBB_SIGNALING_PATHWAY",],
  KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_MTOR_SIGNALING_PATHWAY",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_KRAS_SIGNALING_DN",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_P53_PATHWAY",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_EGFEGFR_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_NOTCH_SIGNALING_WP61",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_TNFA_SIGNALING_VIA_NFKB",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL1_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL3_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL4_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL5_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL18_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_TOLLLIKE_RECEPTOR_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_PDGF_PATHWAY",]
)

gsea_bp$Description <- gsub("HALLMARK_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("GOBP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("KEGG_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("REACTOME_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("WP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("_V1", "", gsea_bp$Description)
gsea_bp$Description <- gsub("METABOLIC_REPROGRAMMING_IN_COLON_CANCER", "CRC_METABOLIC_REPROGRAMMING", gsea_bp$Description)
gsea_bp$Description <- gsub("CELL_CELL_SIGNALING", "CELL-CELL_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("TNFA_SIGNALING_VIA_NFKB", "TNFA_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("WNT_SIGNALING_PATHWAY", "WNT_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("MAPK_SIGNALING_PATHWAY", "MAPK_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("ERBB_SIGNALING_PATHWAY", "ERBB_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("MTOR_SIGNALING_PATHWAY", "MTOR_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("KRAS_SIGNALING_DN", "KRAS_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("NOTCH_SIGNALING_WP61", "NOTCH_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("TOLLLIKE_RECEPTOR_SIGNALING", "TLR_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("EGFEGFR_SIGNALING", "EGF-EGFR_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("_", " ", gsea_bp$Description)

gsea_bp$Description <- factor(gsea_bp$Description, levels = gsea_bp$Description)
gsea_bp$Direction <- "Up"
gsea_bp$Direction[gsea_bp$NES<0] <- "Down"

gsea_bar.plt <- ggplot(gsea_bp, aes(x=Description, y=NES, fill=Direction)) +
  geom_bar(stat = "identity", color="black", width = 0.75) +  
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, color = "black"),
    plot.margin = margin(t = 10, r = 10, b = 100, l = 65, unit = "pt"),
    legend.position = "none"
  ) +
  geom_hline(yintercept = 0) + 
  ylab("NES") + xlab(NULL) + 
  scale_fill_manual(values=c("#357EBD", "#D43F3A"))

gsea_bar.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=30, fig.width=13}
# GSEA plots
terms_of_interest <- c(
  "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "KEGG_PYRUVATE_METABOLISM", "KEGG_GLYCOLYSIS_GLUCONEOGENESIS", 
  "KEGG_FATTY_ACID_METABOLISM", "HALLMARK_MYC_TARGETS_V1", "GOBP_MITOCHONDRIAL_GENE_EXPRESSION",
  "WP_METABOLIC_REPROGRAMMING_IN_COLON_CANCER", "HALLMARK_MITOTIC_SPINDLE", "GOBP_REGULATION_OF_CELL_DEVELOPMENT",
  "GOBP_CELL_CELL_SIGNALING", "HALLMARK_APOPTOSIS", "HALLMARK_HYPOXIA", "GOBP_ADAPTIVE_IMMUNE_RESPONSE",
  "WP_INSULIN_SIGNALING", "GOBP_WNT_SIGNALING_PATHWAY", "KEGG_MAPK_SIGNALING_PATHWAY", "KEGG_ERBB_SIGNALING_PATHWAY",
  "KEGG_MTOR_SIGNALING_PATHWAY", "HALLMARK_KRAS_SIGNALING_DN", "HALLMARK_P53_PATHWAY", "WP_EGFEGFR_SIGNALING",
  "WP_NOTCH_SIGNALING_WP61", "HALLMARK_TNFA_SIGNALING_VIA_NFKB", "WP_IL1_SIGNALING", "WP_IL3_SIGNALING",
  "WP_IL4_SIGNALING", "WP_IL5_SIGNALING", "WP_IL18_SIGNALING", "WP_TOLLLIKE_RECEPTOR_SIGNALING",
  "WP_PDGF_PATHWAY"
  )

gsea_plot.lst <- lapply(terms_of_interest, function(term){
  term_system <- strsplit(term, split = "_")[[1]][1]

  adapted_term <- term %>% 
    str_remove("^(HALLMARK|GOBP|KEGG|REACTOME|WP)_") %>%
    str_replace("METABOLIC_REPROGRAMMING_IN_COLON_CANCER", "CRC_METABOLIC_REPROGRAMMING") %>%
    str_replace("CELL_CELL_SIGNALING", "CELL-CELL_SIGNALING") %>%
    str_replace("TNFA_SIGNALING_VIA_NFKB", "TNFA_SIGNALING") %>%
    str_replace("WNT_SIGNALING_PATHWAY", "WNT_SIGNALING") %>%
    str_replace("MAPK_SIGNALING_PATHWAY", "MAPK_SIGNALING") %>%
    str_replace("ERBB_SIGNALING_PATHWAY", "ERBB_SIGNALING") %>%
    str_replace("MTOR_SIGNALING_PATHWAY", "MTOR_SIGNALING") %>%
    str_replace("KRAS_SIGNALING_DN", "KRAS_SIGNALING") %>%
    str_replace("NOTCH_SIGNALING_WP61", "NOTCH_SIGNALING") %>%
    str_replace("TOLLLIKE_RECEPTOR_SIGNALING", "TLR_SIGNALING") %>%
    str_replace("EGFEGFR_SIGNALING", "EGF-EGFR_SIGNALING") %>%
    str_replace_all("_", " ")
  
  if (term_system == "HALLMARK"){
    gseaplot(HM_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "GOBP"){
    gseaplot(GOBP_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "KEGG") {
    gseaplot(KEGG_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "REACTOME") {
    gseaplot(REACTOME_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else {
    gseaplot(WIKIPATH_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  }
})

gsea.plt <- wrap_plots(gsea_plot.lst, ncol=3)
gsea.plt
```

```{r, eval=FALSE}
# Save GSEA results
saveRDS(HM_gsea, path_to_save_seurat_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_seurat_go_gsea)
saveRDS(KEGG_gsea, path_to_save_seurat_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_seurat_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_seurat_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_seurat_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_seurat_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_seurat_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_seurat_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_seurat_wikipath_gsea.csv"))

# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/cd8_gsea.png"), plot=gsea_bar.plt, dpi=500, height=5, width=9)
ggsave(filename=paste0(path_to_save_figure, "/cd8_gsea_plt.png"), plot=gsea.plt, dpi=500, height=30, width=13)
```

Overall, RT CD8+ T cells upregulated in metabolic reprogramming to Warburg's effect while downregulated in cell proliferation and differentiation, apoptosis, and immune responses.

## Based on DESeq2

```{r}
# Define paths to save the results
path_to_save_deseq2_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_deseq2_dea.rds")
path_to_save_deseq2_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_deseq2_hm_gsea.rds")
path_to_save_deseq2_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_deseq2_go_gsea.rds")
path_to_save_deseq2_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_deseq2_kegg_gsea.rds")
path_to_save_deseq2_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_deseq2_reactome_gsea.rds")
path_to_save_deseq2_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/cd8_deseq2_wikipath_gsea.rds")
```

### Create an DESeq2 object

```{r, message=FALSE, warning=FALSE}
# Prepare single-cell meta-data for DESeq2
pure_cd8_seurat$ref[pure_cd8_seurat$ref == "Hing_2022"] <- "Hing_2023" # Update reference name
pure_cd8_seurat$patient_cond <- paste0("case_", pure_cd8_seurat$case, "_", pure_cd8_seurat$ref, "_", pure_cd8_seurat$new_cond) 
patient_cond_cls <- unique(pure_cd8_seurat$patient_cond)
RT_subjects <- patient_cond_cls[grep("RT", patient_cond_cls)]
RT_subjects <- gsub(pattern = "_", replacement = "-", x=RT_subjects) # change "_" to "-" as DESeq2 converted sample names from "_" to "-"

# Create pseudocount matrix for tumour cell object
cts <- AggregateExpression(pure_cd8_seurat, 
                           group.by = "patient_cond",
                           assays = "RNA",
                           slot = "counts",
                           resturn.seurat = FALSE) 
cts <- cts$RNA
pseudo_count.mat <- as.data.frame(cts)

# Create metadata for DESeq2 object
batch_map <- c("Penter (2021)" = "1", "Nadeu (2022)" = "2", "Rejeski (2022)" = "3", "Hing (2023)" = "4", "Parry (2023a)" = "5", 
               "Parry (2023b)" = "6", "Rigo (2024)" = "7")
colData <- data.frame(samples = colnames(pseudo_count.mat))
colData$condition <- ifelse(colData$samples %in% RT_subjects, "RT", "CLL")
colData$study <- stringr::word(colData$samples, start = -3, end = -2, sep = fixed("-"))
colData$study <- sub("^(.*)-([0-9].*)$", "\\1 (\\2)", colData$study) # Substitiute string like "Parry-2023b" to "Parry (2023b)"
colData$study <- factor(colData$study, level = c("Penter (2021)", "Nadeu (2022)", "Rejeski (2022)", "Hing (2023)", 
                                                 "Parry (2023a)", "Parry (2023b)", "Rigo (2024)"))
colData$batch <- batch_map[colData$study]
colData$batch <- factor(colData$batch)
rownames(colData) <- colData$samples

# Create a DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = pseudo_count.mat, 
                              colData = colData,
                              design = ~ batch + condition)

# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

### Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6.5}
# Variance stabilised transformation
vsdata <- vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
design_matrix <- model.matrix(~ condition, data = colData(dds))
vs_mat_corrected <- limma::removeBatchEffect(
  vs_mat, 
  batch = colData(dds)$batch, 
  design = design_matrix 
)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2               # variance of each PC
pca_var_percent <- pca_var / sum(pca_var) * 100

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     study = colData(dds)$study)

# Plot PCA
pca.plt <- ggplot(pca_df, aes(PC1, PC2, color = study, shape = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  scale_color_manual(values = c(
    "Penter (2021)" = "#1f77b4", 
    "Nadeu (2022)" = "#ff7f0e",
    "Rejeski (2022)" = "#9632B8" ,
    "Hing (2023)" = "#D43F3A" ,
    "Parry (2023a)" = "#5CB85C",
    "Parry (2023b)" = "#FFB900",
    "Rigo (2024)" = "#46B8DA")) +
  scale_shape_manual(values = c("CLL" = 16, "RT" = 17)) +
  guides(
    color = guide_legend(title = "Study"),
    shape = guide_legend(title = "Condition")
  ) +
  theme_bw() +
  theme(
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 11)
  )

pca.plt
```

From the PCA plot we detected two outliers we then removed it before further analysis

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6.5}
# Plot the PCA plot after outlier removal

# Identify and Remove the Outlier 
outlier_id <- rownames(pca_df)[c(which.max(pca_df$PC1), which.min(pca_df$PC2))] # the outliers are the sample with the highest PC1 value and lowest PC2 value
dds_clean <- dds[, colnames(dds) != outlier_id] # Filter the outlier out of the original dds object

# Re-run VST on the cleaned dataset
vsdata_clean <- vst(dds_clean, blind = FALSE)
vs_mat_clean <- assay(vsdata_clean)

# Adjust for batch (assume batch info is in colData(dds)$batch)
design_matrix <- model.matrix(~ condition, data = colData(dds_clean))
vs_mat_corrected <- limma::removeBatchEffect(
  vs_mat_clean, 
  batch = colData(dds_clean)$batch, 
  design = design_matrix # <--- This argument is key for separation!
)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2
pca_var_percent <- pca_var / sum(pca_var) * 100

# Make a dataframe for ggplot
pca_df_clean <- data.frame(pca$x,
                     condition = colData(dds_clean)$condition,
                     study = colData(dds_clean)$study)

# Plot PCA
corrected_pca.plt <- ggplot(pca_df_clean, aes(PC1, PC2, color = study, shape = condition)) + 
  geom_point(size = 3) +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  scale_color_manual(values = c(
    "Penter (2021)" = "#1f77b4", 
    "Nadeu (2022)" = "#ff7f0e",
    "Rejeski (2022)" = "#9632B8" ,
    "Hing (2023)" = "#D43F3A" ,
    "Parry (2023a)" = "#5CB85C",
    "Parry (2023b)" = "#FFB900",
    "Rigo (2024)" = "#46B8DA")) +
  scale_shape_manual(values = c("CLL" = 16, "RT" = 17)) +
  guides(
    color = guide_legend(title = "Study"),
    shape = guide_legend(title = "Condition")
  ) +
  theme_bw() +
  theme(
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 11)
  ) 

corrected_pca.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/deseq2_cd8_pca.png"), plot=pca.plt, dpi=500, height=4.5, width=6.5)
ggsave(filename=paste0(path_to_save_figure, "/deseq2_cd8_corrected_pca.png"), plot=corrected_pca.plt, dpi=500, height=4.5, width=6.5)
```

### Run DESeq2

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Run DESeq2 for the data after outlier removal
counts_clean <- counts(dds_clean)
colData_clean <- colData(dds_clean)
colData_clean$condition <- factor(colData_clean$condition, levels = c("CLL", "RT"))
colData_clean$batch <- factor(colData_clean$batch) 

dds_final <- DESeqDataSetFromMatrix(
  countData = counts_clean,
  colData = colData_clean,
  design = ~ batch + condition  
)

keep <- rowSums(counts(dds_final) >= 10) >= 3 # Keep genes with at least 10 counts in at least 3 samples
noise_genes <- c(
  grep("^IG[HKL][VDJ]", rownames(dds), value = TRUE),  # Ig genes
  grep("^HLA-", rownames(dds), value = TRUE),          # Patient ID (HLA)
  grep("^MT-", rownames(dds), value = TRUE),           # Mitochondrial stress
  "XIST", "RPS4Y1", "KDM5D", "UTY", "DDX3Y",           # Sex genes
  "S100A8", "S100A9", "S100A12", "CXCL8"               # Acute stress response
)
dds_final <- dds_final[keep & (!rownames(dds_final) %in% noise_genes), ]
dds_final$condition <- relevel(dds_final$condition, ref = "CLL")
dds_final <- DESeq(dds_final)

# Generate a table for gene expression
res <- results(dds_final, contrast = c("condition", "RT", "CLL"), independentFiltering = FALSE)
res <- lfcShrink(dds_final, contrast = c("condition", "RT", "CLL"), res = res, type = "normal")
res$direction <- "NS"
res$direction[res$padj < 0.05 & res$log2FoldChange > 0.25] <- "Up"
res$direction[res$padj < 0.05 & res$log2FoldChange < (-0.25)] <- "Down"
dea.df <- as.data.frame(res)

# Save the expression results
saveRDS(dea.df, path_to_save_deseq2_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/cd8_deseq2_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Show the expression results
dea.df <- readRDS(path_to_save_deseq2_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=10}
# Volcano plot
genes_of_interest <- c(
  "TK1", "NUSAP1", "PCLAF"
)

volcano.plt <- volcano_plot(dea.df, 
                            FC=0.25, 
                            x_col="log2FoldChange",
                            y_col="padj",
                            xlim=c(-1.4,1.4), 
                            ylim=c(-0.1,3.5), 
                            genes_of_interest=genes_of_interest,
                            boxedLabels = TRUE) 

volcano.plt
```

```{r, message=FALSE, warning=FALSE}
# MA plot 
ma_data <- dea.df

ma.plt <- ggmaplot(ma_data, 
                   fdr = 0.05,            # Threshold for significance
                   fc = 1.19,                # Threshold for FC
                   size = 0.4,            # Dot size
                   palette = c("#B31B21", "#1465AC", "darkgray"),
                   genenames = as.vector(ma_data$gene),
                   top = 0, label.select = genes_of_interest, # set top = 0, then specify genes using label.select argument
                   legend = "right", 
                   font.label = c("bold", 11), 
                   label.rectangle = TRUE,
                   font.legend = "bold",
                   font.main = "bold",
                   ggtheme = ggplot2::theme_minimal(),
                   xlab = "Mean Expression",
                   ylab = "Log2 Fold Change"
                   )

ma.plt
```

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df) 
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save the enrichment results
saveRDS(HM_gsea, path_to_save_deseq2_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_deseq2_go_gsea)
saveRDS(KEGG_gsea, path_to_save_deseq2_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_deseq2_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_deseq2_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_deseq2_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_deseq2_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_deseq2_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_deseq2_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/cd8_deseq2_wikipath_gsea.csv"))
```

# Session Information

```{r}
sessionInfo()
```







