---
title: "Microenvironment_immune_cell_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

In this analysis, we focused on immune cells, such as NK, CD4+, and CD8+ T cells, from the integrated dataset by performing pseudobulk analysis between RT and CLL in this cell type. Moreover, we also annotated level-2 subtypes of these cells.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tibble)
library(SingleR)
library(celldex)
library(sctransform)
library(cowplot)
library(DT)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save 
path_to_microenvir_obj <- here::here("scRNAseq/meta_analysis/R_objects/microenvir_seu.rds")
path_to_save_immune_obj <- here::here("scRNAseq/meta_analysis/R_objects/immune_seu.rds")
path_to_save_nk_obj <- here::here("scRNAseq/meta_analysis/R_objects/nk_seu.rds")
path_to_save_t_obj <- here::here("scRNAseq/meta_analysis/R_objects/t_seu.rds")
path_to_save_cd4_obj <- here::here("scRNAseq/meta_analysis/R_objects/cd4_seu.rds")
path_to_save_cd8_obj <- here::here("scRNAseq/meta_analysis/R_objects/cd8_seu.rds")
path_to_save_figure <- here::here("figures")
```

```{r}
# Thresholds
min_log2FC <- 0.25
alpha <- 0.001
```

```{r}
# Set seed for reproducibility
set.seed(1)
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the microenvironment object
microenvir_seurat <- readRDS(path_to_microenvir_obj)
```

# Extract immune cells

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Extract immune cells with reanalysis
immune_seurat <- subset(microenvir_seurat, subset = singleR.labels_main != "Monocytes")
immune_seurat <- SCTransform(immune_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
immune_seurat <- RunPCA(immune_seurat)
immune_seurat <- RunUMAP(immune_seurat, dims = 1:30, reduction = "harmony")
immune_seurat <- FindNeighbors(immune_seurat, dims = 1:30, reduction = "harmony")        
immune_seurat <- FindClusters(
  immune_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 0.9, by = 0.1)
)

# Save the object
saveRDS(immune_seurat, path_to_save_immune_obj)
```

We separated the NK from T cells based on the markers below.

| Marker          | Role                | T Cells (CD4/CD8) | NK Cells                   |
|-----------------|---------------------|-------------------|----------------------------|
| CD3D, CD3E      | T-cell receptor     | High              | Absent                     |
| KLRF1           | NK surface receptor | Low/Absent        | High (The cleanest marker) |
| NCAM1 (CD56)    | Classical NK marker | Absent            | High (in some subsets)     |
| FCGR3A (CD16)   | ADCC receptor       | Low               | High (CD56-dim subset)     |
| TYROBP          | Signaling adaptor   | Absent            | High                       |

```{r, fig.height=7, fig.width=12, warning=FALSE, message=FALSE}
# Plot based on cell biomarker expression
immune_seurat <- readRDS(path_to_save_immune_obj)
FeaturePlot(immune_seurat, 
            features = c("HBM", "HBA1", "CD19", "CD79A", "CD14", "CD3D", "CD3E", 
                         "KLRF1", "NCAM1", "FCGR3A", "TYROBP"),
            reduction = "umap") 
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the immune cell clusters
group_by <- c("SCT_snn_res.0.6", "SCT_snn_res.0.7", "SCT_snn_res.0.8")
umap_plots <- lapply(group_by, function (group){
  DimPlot(immune_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 2,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

immune_umap.plt <- wrap_plots(umap_plots, ncol=3)
immune_umap.plt
```

## Extract NK cells

At SCT_snn_res.0.7, we found that cluster 10 and 12 were NK cells. Then we extracted them for NK cell analyses.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Extract Nk cells with reanalysis
nk_seurat <- subset(immune_seurat, subset = SCT_snn_res.0.7 %in% c(10, 12, 21))
nk_seurat <- SCTransform(nk_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
nk_seurat <- RunPCA(nk_seurat)
nk_seurat <- RunUMAP(nk_seurat, dims = 1:30, reduction = "harmony")
nk_seurat <- FindNeighbors(nk_seurat, dims = 1:30, reduction = "harmony")        
nk_seurat <- FindClusters(
  nk_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 1, by = 0.1)
)

# Save the object
saveRDS(nk_seurat, path_to_save_nk_obj)
```

## Extract T cells

In contrast, we extracted T cells from the remaining clusters apart from cluster 10 and 12.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# Extract T cells with reanalysis
t_seurat <- subset(immune_seurat, subset = !(SCT_snn_res.0.7 %in% c(10, 12)))
t_seurat <- SCTransform(t_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
t_seurat <- RunPCA(t_seurat)
t_seurat <- RunUMAP(t_seurat, dims = 1:30, reduction = "harmony")
t_seurat <- FindNeighbors(t_seurat, dims = 1:30, reduction = "harmony")        
t_seurat <- FindClusters(
  t_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 1, by = 0.1)
)

# Save the object
saveRDS(t_seurat, path_to_save_t_obj)
```

```{r}
# Define paths to load and save 
path_to_save_filtered_nk_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered_nk_seu.rds")
path_to_save_filtered_nk_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered_nk_marker.rds")
path_to_save_filtered2_nk_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered2_nk_seu.rds")
path_to_save_filtered2_nk_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered2_nk_marker.rds")
path_to_save_filtered3_nk_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered3_nk_seu.rds")
path_to_save_filtered3_nk_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered3_nk_marker.rds")
path_to_save_filtered4_nk_obj <- here::here("scRNAseq/meta_analysis/R_objects/filtered4_nk_seu.rds")
path_to_save_filtered4_nk_markers <- here::here("scRNAseq/meta_analysis/R_objects/filtered4_nk_marker.rds")
path_to_save_pure_nk_obj <- here::here("scRNAseq/meta_analysis/R_objects/pure_nk_seu.rds")
path_to_save_pure_nk_markers <- here::here("scRNAseq/meta_analysis/R_objects/pure_nk_marker.rds")
```

# Filtering out contaminated cells

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# NK cell purification
nk_seurat <- readRDS(path_to_save_nk_obj)

# 1.Filter out based on percentage of IG, HB, TCR, and platelet genes 
nk_seurat[["percent.rbc"]] <- PercentageFeatureSet(nk_seurat, pattern = "^HB[AB]")
nk_seurat[["percent.ig"]] <- PercentageFeatureSet(nk_seurat, pattern = "^IG[HKL]")
nk_seurat[["percent.tcr"]] <- PercentageFeatureSet(nk_seurat, pattern = "^TR[ABGD]")
platelet_genes <- c("PF4", "PPBP", "TUBB1")
nk_seurat[["percent.platelet"]] <- PercentageFeatureSet(nk_seurat, features = platelet_genes)
filtered_nk_seurat <- subset(nk_seurat, 
                             subset = percent.rbc < 1 & percent.ig < 1 & percent.tcr < 1 & percent.platelet < 1) 

# 2.Filter out comtaminated T and B cells
filtered_nk_seurat <- subset(filtered_nk_seurat, subset = CD3E < 1 & CD3D < 1 & MS4A1 < 1)

# 3.Filter out comtaminated monocytes
filtered_nk_seurat <- subset(filtered_nk_seurat, subset = CD14 < 1 & LYZ < 2 & S100A9 < 1) # LYS's cutoff is high as it can be found in blood elsewhere

# 4.Reanalysis
filtered_nk_seurat <- SCTransform(filtered_nk_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE) 
filtered_nk_seurat <- RunPCA(filtered_nk_seurat)
filtered_nk_seurat <- RunUMAP(filtered_nk_seurat, dims = 1:30, reduction = "harmony")
filtered_nk_seurat <- FindNeighbors(filtered_nk_seurat, dims = 1:30, reduction = "harmony")        
filtered_nk_seurat <- FindClusters(
  filtered_nk_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# 5.Save the object
saveRDS(filtered_nk_seurat, path_to_save_filtered_nk_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
filtered_nk_seurat <- readRDS(path_to_save_filtered_nk_obj)
Idents(filtered_nk_seurat) <- "SCT_snn_res.1.2"
markers <- FindAllMarkers(filtered_nk_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered_nk_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered_nk_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered_nk_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

From the results we found that cluster 0, 5, 10, 14, 15, and 16 were contaminated cells.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the NK cell clusters with reprocessing
filtered2_nk_seurat <- subset(filtered_nk_seurat, idents = c(0, 1, 2, 4, 6, 8, 9, 11, 15))
filtered2_nk_seurat <- SCTransform(filtered2_nk_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered2_nk_seurat <- RunPCA(filtered2_nk_seurat)
filtered2_nk_seurat <- RunUMAP(filtered2_nk_seurat, dims = 1:30, reduction = "harmony")
filtered2_nk_seurat <- FindNeighbors(filtered2_nk_seurat, dims = 1:30, reduction = "harmony")        
filtered2_nk_seurat <- FindClusters(
  filtered2_nk_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# Save the object
saveRDS(filtered2_nk_seurat, path_to_save_filtered2_nk_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(filtered2_nk_seurat) <- "SCT_snn_res.1.2"
markers <- FindAllMarkers(filtered2_nk_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered2_nk_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered2_nk_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered2_nk_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

Cluster 0, 1, 7, and 11 were contaminated cells, we then removed them.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the NK cell clusters with reprocessing
filtered3_nk_seurat <- subset(filtered2_nk_seurat, idents = c(2, 3, 4, 5, 6, 8, 9, 10))
filtered3_nk_seurat <- SCTransform(filtered3_nk_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered3_nk_seurat <- RunPCA(filtered3_nk_seurat)
filtered3_nk_seurat <- RunUMAP(filtered3_nk_seurat, dims = 1:30, reduction = "harmony")
filtered3_nk_seurat <- FindNeighbors(filtered3_nk_seurat, dims = 1:30, reduction = "harmony")        
filtered3_nk_seurat <- FindClusters(
  filtered3_nk_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# Save the object
saveRDS(filtered3_nk_seurat, path_to_save_filtered3_nk_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(filtered3_nk_seurat) <- "SCT_snn_res.1"
markers <- FindAllMarkers(filtered3_nk_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_filtered3_nk_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/filtered3_nk_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_filtered3_nk_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We found cluster 4 and 6 were contaminated cells.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Keep only the NK cell clusters with reprocessing
filtered4_nk_seurat <- subset(filtered3_nk_seurat, idents = c(0, 1, 2, 3, 5, 7, 8, 9))
filtered4_nk_seurat <- SCTransform(filtered4_nk_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered4_nk_seurat <- RunPCA(filtered4_nk_seurat)
filtered4_nk_seurat <- RunUMAP(filtered4_nk_seurat, dims = 1:30, reduction = "harmony")
filtered4_nk_seurat <- FindNeighbors(filtered4_nk_seurat, dims = 1:30, reduction = "harmony")        
filtered4_nk_seurat <- FindClusters(
  filtered4_nk_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# Save the object
saveRDS(filtered4_nk_seurat, path_to_save_filtered4_nk_obj)
```

In conclusion, we removed and processed the NK cell object for 4 times till we obtained the pure NK cell object for further analysis.

# Subtype annotation

```{r, eval=FALSE}
# Change the obeject name
pure_nk_seurat <- filtered4_nk_seurat
saveRDS(pure_nk_seurat, path_to_save_pure_nk_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the clusters
pure_nk_seurat <- readRDS(path_to_save_pure_nk_obj)
group_by <- c("SCT_snn_res.0.2", "SCT_snn_res.0.3", "SCT_snn_res.0.4")
umap_plots <- lapply(group_by, function (group){
  DimPlot(pure_nk_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 4,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

pure_nk_umap.plt <- wrap_plots(umap_plots, ncol=3)
pure_nk_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=13}
# Visualise the umap
group_by <- c("ref", "tissue", "group")
umap_plots <- lapply(group_by, function (group){
  DimPlot(pure_nk_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 4,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

pure_nk_umap.plt <- wrap_plots(umap_plots, ncol=3)
pure_nk_umap.plt
```

```{r, fig.height=7, fig.width=12, warning=FALSE, message=FALSE}
# Plot based on cell biomarker expression
FeaturePlot(pure_nk_seurat, 
            features = c("NCAM1", "HAVCR2", "FCGR3A", "SELL", "GZMK", "PRF1", "GZMA", "TIGIT",
                         "JUNB", "XCL1", "XCL2", "CD69"),
            reduction = "umap") 
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Calculate markers
Idents(pure_nk_seurat) <- "SCT_snn_res.0.3"
markers <- FindAllMarkers(pure_nk_seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)

# Save the marker results
saveRDS(markers, path_to_save_pure_nk_markers)
write.csv(markers, here::here("scRNAseq/meta_analysis/text_files/pure_nk_markers.csv"), row.names = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Show the marker results
markers <- readRDS(path_to_save_pure_nk_markers)
DT::datatable(markers, options = list(scrollX = TRUE))
```

We annotated the myeloid subtypes as literatures below:

- [Inflammatory-driven NK cell maturation and its impact on pathology](https://www.frontiersin.org/journals/immunology/articles/10.3389/fimmu.2022.1061959/full)
- [Distinct granzyme k expression in immune cells: a single-cell rna-seq meta-analysis](https://link.springer.com/article/10.1007/s13258-024-01555-1)
- [Single-cell data revealed exhaustion of characteristic NK cell subpopulations and T cell subpopulations in hepatocellular carcinoma](https://www.aging-us.com/article/205723/text)
- [Glucocorticoid-induced Tumor Necrosis Factor Receptor Negatively Regulates Activation of Human Primary Natural Killer (NK) Cells by Blocking Proliferative Signals and Increasing NK Cell Apoptosis](https://www.jbc.org/article/S0021-9258(20)52775-0/fulltext)
- [Discovery of specialized NK cell populations infiltrating human melanoma metastases](https://insight.jci.org/articles/view/133103#F2)

| Cluster | Markers             | Annotation                   | 
|---------|---------------------|------------------------------|
| 0       | PRF1, GZMA          | Cytotoxic NK                 | 
| 1       | CD69, JUN, AREG     | TI-like NK                   | 
| 2       | HAVCR2, PRF1, GZMA  | Early exhausted NK           | 
| 3       | GZMK                | Late transitional NK         | 
| 4       | TNFRSF18            | TNFRSF18hi NK                | 
| 5       | CD27, GZMK          | Early transitional NK        | 

* TI-like NK is tumour-infiltrating-like NK

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=7}
# Plot UMAP for NK cell subtypes

# 1. Map Seurat clusters and create the annotation column
cluster_to_annotation <- c(
  "0" = "Cytotoxic NK", "1" = "TI-like NK", "2" = "Early exhausted NK", "3" = "Late transitional NK", 
  "4" = "TNFRSF18hi NK", "5" = "Early transitional NK"
  )

pure_nk_seurat$level2_annotate <- 
  as.vector(cluster_to_annotation[as.character(pure_nk_seurat$SCT_snn_res.0.3)])

# 2. Set level order
new_levels <- c(
  # Group1: transitional NK 
  "Early transitional NK", "Late transitional NK",
  
  # Group2: cytotoxic and early exhausted NK  
  "Cytotoxic NK", "Early exhausted NK",     

  # Group3: other NK 
  "TNFRSF18hi NK", "TI-like NK" 
)

pure_nk_seurat$level2_annotate <- factor(
  pure_nk_seurat$level2_annotate,
  levels = new_levels
) 

# 3. Define colours and labels
new_colours <- c(
  "#A6CEE3", "#1F78B4",  # Group 1 (2 colours)
  "#E25644", "#B04968",  # Group 2 (2 colours)
  "#FFD92F", "#FDBF6F"   # Group 3 (2 colours)
)

custom_labels <- list(
  bquote("Early transitional NK"), bquote("Late transitional NK"),
  bquote("Cytotoxic NK"), bquote("Early exhausted NK"), 
  bquote(TNFRSF18^hi~"NK"), bquote("TI-like NK")
  )

# 4. Plot UMAP
nk_subtype.plt <- DimPlot(pure_nk_seurat, 
                          group.by = "level2_annotate", 
                          label = FALSE,
                          repel = TRUE,
                          pt.size = 1.5,
                          alpha = 1,
                          raster = FALSE
                          ) +
  scale_color_manual(
    values = new_colours,
    breaks = new_levels,
    labels = custom_labels
  ) +
  labs(y = "UMAP 2", x = "UMAP 1") +
  ggtitle("NK subtypes")

nk_subtype.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=9}
# Dot plot for biomarker expressions based on annotated cells
features <- c("NCAM1", "FCGR3A", "SELL", "CD27", "GZMK", "XCL1", "XCL2", "PRF1", "GZMA", "GZMB", "HAVCR2", 
              "TNFRSF18", "CD69", "JUN", "AREG")
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)

nk_markers.plt <- DotPlot(pure_nk_seurat, features=features, group.by = "level2_annotate") +
  RotatedAxis() +
  scale_colour_gradientn(colours = colour_pal) +
  scale_y_discrete(
      limits = rev(new_levels),
      labels = rev(custom_labels)
    ) +
  labs(
    colour = "Average\nExpression",
    size = "Percent\nExpressed"
  )

nk_markers.plt
```

```{r, message=FALSE, warning=FALSE}
# Plot subtype population based on tissue and condition
proportion_tissue_df <- pure_nk_seurat@meta.data %>%
  dplyr::mutate(level2_annotate = factor(level2_annotate, levels = new_levels)) %>%
  dplyr::count(tissue, level2_annotate) %>%
  dplyr::group_by(tissue) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n) # Remove n column

proportion_cond_df <- pure_nk_seurat@meta.data %>%
  dplyr::mutate(level2_annotate = factor(level2_annotate, levels = new_levels)) %>%
  dplyr::count(new_cond, level2_annotate) %>%
  dplyr::group_by(new_cond) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n) # Remove n column

tissue_cell_pop.plt <- ggplot(proportion_tissue_df, aes(x = tissue, 
                                                    y = proportion, 
                                                    fill = level2_annotate)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "NK cell subtype" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Tissue", 
    title = "Tissue"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

cond_cell_pop.plt <- ggplot(proportion_cond_df, aes(x = new_cond, 
                                                    y = proportion, 
                                                    fill = level2_annotate)) +
  geom_col(position = "stack", width = 0.5) +
  scale_fill_manual(
    values = new_colours,
    breaks = new_levels,
    name = "NK cell subtype" # Set legend title
  ) +
  labs(
    y = "Cell proportion",
    x = "Condition", 
    title = "Condition"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

# Combine the plots
adapted_tissue_cell_pop.plt <- tissue_cell_pop.plt +
  theme(legend.position = "none",
        axis.title.x = element_blank()) 

adapted_cond_cell_pop.plt <- cond_cell_pop.plt + 
  theme(legend.position = "right", 
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) 

combined_cell_pop.plt <- adapted_tissue_cell_pop.plt + adapted_cond_cell_pop.plt
combined_cell_pop.plt
```

```{r, eval=FALSE}
# Save the nk object
saveRDS(pure_nk_seurat, path_to_save_pure_nk_obj)

# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/nk_umap.png"), plot=nk_subtype.plt, dpi=500, height=4.5, width=7)
ggsave(filename=paste0(path_to_save_figure, "/nk_dot.png"), plot=nk_markers.plt, dpi=500, height=4, width=9)
ggsave(filename=paste0(path_to_save_figure, "/nk_cell_pop.png"), plot=combined_cell_pop.plt , dpi=500, height=4, width=7)
```

# Trajectory analysis

In this section, we used the Monocle 3 package to identify developmental trajectories of NK cells using early transitional subtype as a root node.

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(monocle3)
library(igraph)
library(SeuratWrappers)
library(paletteer)
```

```{r}
# Define path to save
path_to_save_nk_monocle_obj <- here::here("scRNAseq/meta_analysis/R_objects/nk_monocle_cds.rds")
path_to_save_nk_monocle_dea <- here::here("scRNAseq/meta_analysis/R_objects/nk_monocle_dea.rds")
path_to_save_nk_monocle_module <- here::here("scRNAseq/meta_analysis/R_objects/nk_monocle_module.rds")
path_to_save_nk_monocle_agg_mat <- here::here("scRNAseq/meta_analysis/R_objects/nk_monocle_agg_mat.rds")
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create monocle object and generate graph (use data from seurat)
cds <- as.cell_data_set(pure_nk_seurat) # Convert the seurat object to monocle object using SeuratWrappers
cds <- preprocess_cds(cds)

# Create gene_short_name
rowData(cds)$gene_short_name <- row.names(rowData(cds))

# Assign seurat clusters to monocle and create a partition
cds <- monocle3::cluster_cells(cds, reduction_method = "UMAP", 
                               cluster_method = "louvain", 
                               resolution = 0.5)
cds@clusters$UMAP$clusters <- pure_nk_seurat$SCT_snn_res.0.3

# Assign UMAP coordination
cds@int_colData$reducedDims$UMAP <- pure_nk_seurat@reductions$umap@cell.embeddings
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=17}
# Monocle plot before trajectory analysis
cds <- readRDS(path_to_save_nk_monocle_obj)
group_by <- c("cluster", "partition", "level2_annotate")
umap_plots <- lapply(group_by, function (group){
  plot_cells(cds, 
             color_cells_by = group, 
             label_groups_by_cluster = FALSE, 
             group_label_size = 5,
             cell_size = 0.7) +  
    theme(legend.position = "right")
})

monocle_umap.plt <- wrap_plots(umap_plots, ncol=3)
monocle_umap.plt
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Learn trajectory
cds <- learn_graph(cds, use_partition = TRUE, close_loop = FALSE)
```

```{r, fig.height=5, fig.width=7}
# Plot after learning trajectory
monocle_cluster.plt <- plot_cells(cds, 
                                  color_cells_by = "cluster",
                                  graph_label_size=5,
                                  cell_size = 0.7,
                                  show_trajectory_graph = TRUE
                                  ) + 
  ggtitle("Seurat clusters") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

monocle_cluster.plt
```

```{r, fig.height=5, fig.width=7}
# Plot root
root.plt <- plot_cells(cds, label_principal_points = TRUE) + 
  ggtitle("Root principal node") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

root.plt
```

## Calculate pseudotime

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set a root to order cells in pseudotime
cds <- order_cells(cds, reduction_method = "UMAP", root_pr_nodes = "Y_60")
```

```{r, fig.height=5, fig.width=7}
# Plot the pseudotime
monocle_pseudotime.plt <- plot_cells(cds, 
                                     color_cells_by = "pseudotime",
                                     label_cell_groups=FALSE,
                                     label_leaves=FALSE,
                                     label_branch_points=FALSE,
                                     graph_label_size=1.5,
                                     cell_size = 0.7) +
  ggtitle("Pseudotime") +
  theme(legend.position = "right",
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
        )

monocle_pseudotime.plt
```

```{r, fig.height=5, fig.width=5}
# Plot the pseudotime with trajectory path with Seurat stype

# 1. Extract principal graph
pg_aux <- principal_graph_aux(cds)$UMAP

# 2. Convert to edge list
edges <- as_data_frame(principal_graph(cds)$UMAP, what = "edges")
nodes <- as.data.frame(t(pg_aux$dp_mst))
nodes$name <- rownames(nodes)
root_node <- cds@principal_graph_aux$UMAP$root_pr_nodes

# 3. Merge edges with coordinates
trajectory_df <- edges %>%
  dplyr::left_join(nodes, by = c("from" = "name")) %>%
  dplyr::rename(
    UMAP_1_from = umap_1,
    UMAP_2_from = umap_2
  ) %>%
  dplyr::left_join(nodes, by = c("to" = "name")) %>%
  dplyr::rename(
    UMAP_1_to = umap_1,
    UMAP_2_to = umap_2
  )

seurat_df <- data.frame(
  Embeddings(pure_nk_seurat, "umap"),
  pseudotime = pseudotime(cds)
)

root_df <- nodes %>%
  dplyr::filter(name %in% root_node)

# 4. Plot
trajectory_no_annotate.plt <- ggplot(seurat_df, aes(umap_1, umap_2)) +
  geom_point(aes(color = pseudotime), size = 2, alpha=1) +
  geom_segment(
    data = trajectory_df,
    aes(
      x = UMAP_1_from,
      y = UMAP_2_from,
      xend = UMAP_1_to,
      yend = UMAP_2_to
    ),
    inherit.aes = FALSE,
    colour = "black",
    linewidth = 0.7
  ) +
  geom_point(
    data = root_df,
    aes(umap_1, umap_2),
    inherit.aes = FALSE,
    shape = 21,          # circle with fill
    size = 6,
    fill = "white",
    colour = "black",
    stroke = 1
  ) +
  geom_text(
    data = root_df,
    aes(umap_1, umap_2),
    label = "1",
    inherit.aes = FALSE,
    size = 5,
    fontface = "bold"
  ) +
  scale_color_gradientn(
    name = "Pseudotime",
    colors = paletteer_c("grDevices::RdYlBu", 30, direction = -1)
  ) +
  theme_void() +
  theme(
    legend.position = "bottom" 
  )  
  
annotate_trajectory.plt <- trajectory_no_annotate.plt +  
  annotate("text", x = 1.5, y = -4.3, label = "Transitional NK", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -2.5, y = -6.8, label = "TNFRSF18hi NK", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = -3.7, y = 2.8, label = "Cytotoxic NK", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 2, y = 6.5, label = "Early exhausted NK", 
             color = "black", fontface = "plain", size = 4) +
  annotate("text", x = 7, y = -1.5, label = "TI-like NK", 
             color = "black", fontface = "plain", size = 4)

annotate_trajectory.plt
```

```{r, eval=FALSE}
# Save the monocle object
saveRDS(cds, path_to_save_nk_monocle_obj)

# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/nk_trajectory.png"), plot=annotate_trajectory.plt, dpi=500, height=5, width=5)
```

## Gene expression across pseudotime

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Group genes into modules
nk_cds_pr_test_res <- monocle3::graph_test(cds, neighbor_graph = "principal_graph", cores = 4)
monocle_degs <- row.names(subset(nk_cds_pr_test_res, q_value < 0.05))
gene_modules_df <- find_gene_modules(cds[monocle_degs,], resolution=3e-3)
cell_group_df <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=colData(cds)$level2_annotate)
agg_mat <- aggregate_gene_expression(cds, gene_modules_df, cell_group_df)
row.names(agg_mat) <- paste0("Module ", row.names(agg_mat))

# Save the module results
write.csv(nk_cds_pr_test_res, here::here("scRNAseq/meta_analysis/supplm_file/nk_trajectory_gene_expr.csv"))
saveRDS(nk_cds_pr_test_res, path_to_save_nk_monocle_dea)
saveRDS(gene_modules_df, path_to_save_nk_monocle_module)
saveRDS(agg_mat, path_to_save_nk_monocle_agg_mat)
```

```{r, message=FALSE, warning=FALSE}
# Show the monocle dae
monocle_dea <- readRDS(path_to_save_nk_monocle_dea)
DT::datatable(monocle_dea, options = list(scrollX = TRUE))
```

```{r, fig.height=10, fig.width=8}
# Plot heatmap 
library(pheatmap)
gene_modules_df <- readRDS(path_to_save_nk_monocle_module)
agg_mat <- readRDS(path_to_save_mn_monocle_agg_mat)
colour_pal <- colorRampPalette(c("#053061", "#2171B5", "white", "#FFD92F", "#A50F15"))(100)
pseudotime_heatmap.plt <- pheatmap::pheatmap(agg_mat, scale="column", fontsize = 11,
                                             color = colour_pal,
                                             angle_col = 45,
                                             clustering_method="ward.D2")

pseudotime_heatmap.plt
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Test a range of resolutions to appropriate modules
resolutions <- c(1e-3, 2e-3, 3e-3, 4e-3, 5e-3)

for(res in resolutions){
  modules <- find_gene_modules(cds[monocle_degs,], resolution = res)
  num_mods <- length(unique(modules$module))
  print(paste("Resolution:", res, "->", num_mods, "Modules"))
}
```

```{r}
# Collect gene signatures for fate subtypes based on gene modules
tnfrsf18_gene <- gene_modules_df %>% filter(module %in% c(5, 13)) %>% pull(id)
exhausted_gene <- gene_modules_df %>% filter(module %in% c(3, 11, 12)) %>% pull(id)
ti_like_gene <- gene_modules_df %>% filter(module %in% c(4, 6)) %>% pull(id)
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 1 Transitional NK > TNFRSF18hi NK
cells_types_of_interest <- c("Early transitional NK", "Late transitional NK", "TNFRSF18hi NK")
pseudotime_genes <- c("TNFRSF18", "FCGR3A", "PRF1", "HIGD1A", "BAG5", "LILRB1")

tnfrsf18_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                           genes = pseudotime_genes, 
                                                           cell_annotations = cells_types_of_interest, 
                                                           pseudotime_boundary = 8,
                                                           facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Transitional NK to TNFRSF18hi NK") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

tnfrsf18_genes_pseudotime
```

During the transition from transitional NK cells towards the TNFRSF18^hi NK state, NK cells exhibited features consistent with increased maturation, characterised by the upregulation of FCGR3A and PRF1. However, this trajectory also showed evidence of cellular stress, potentially associated with hypoxic adaptation, as indicated by increased expression of HIGD1A, a gene implicated in promoting cell survival under hypoxic and nutrient-deprived conditions.

In addition, reduced expression of BAG5, an anti-apoptotic regulator, suggests a potential increase in susceptibility to cell death along this developmental path. Notably, this NK cell fate may also contribute to tumour-supportive activity through the upregulation of LILRB1, an inhibitory receptor associated with immune suppression.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 2 Transitional NK > Early exhausted NK
cells_types_of_interest <- c("Early transitional NK", "Late transitional NK", "Cytotoxic NK", "Early exhausted NK")
pseudotime_genes <- c("FCGR3A", "HAVCR2", "CD160", "SMAD3", "CD244", "HIF1A")

exhausted_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                           genes = pseudotime_genes, 
                                                           cell_annotations = cells_types_of_interest, 
                                                           pseudotime_boundary = 15,
                                                           facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Transitional NK to Early exhausted NK") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

exhausted_genes_pseudotime
```

For the transition to the early exhausted NK state, this trajectory exhibited transcriptional features consistent with immune suppression, characterised by increased expression of HAVCR2, SMAD3 (a mediator of TGF-β signalling), and CD244, together with reduced expression of FCGR3A and CD160.

In addition, decreased expression of HIF1A suggests a potential impairment in hypoxic adaptation, indicating that this NK cell subtype may exhibit reduced functional capacity and survival potential within the TME.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=6}
# Branch 3 Transitional NK > TI-like NK
cells_types_of_interest <- c("Early transitional NK", "Late transitional NK", "Cytotoxic NK", "TI-like NK")
pseudotime_genes <- c("CD69", "NR4A2", "DUSP1", "IFNG", "HIF1A", "LDHA")

ti_like_genes_pseudotime <- plot_expression_on_pseudotime(cds_ordered = cds, 
                                                           genes = pseudotime_genes, 
                                                           cell_annotations = cells_types_of_interest, 
                                                           pseudotime_boundary = 20,
                                                           facet_wrap_plot = TRUE) +
  miller_discrete_scale(style = "bars") +
  labs(title = "Transitional NK to TI-like NK") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

ti_like_genes_pseudotime
```

Compared with other cell fates, this trajectory retained features consistent with preserved immune function and proliferative capacity, characterised by increased expression of IFNG. This state also exhibited transcriptional signatures suggestive of aerobic glycolysis, as indicated by the upregulation of LDHA, despite reduced expression of HIF1A.

Notably, several genes associated with early exhaustion or stress responses, including NR4A2 and DUSP1, were also upregulated, suggesting a transitional functional state rather than complete exhaustion.

```{r, eval=FALSE}
# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/nk_tnfrsf18_gene_flow.png"), plot=tnfrsf18_genes_pseudotime, dpi=500, height=8, width=6)
ggsave(filename=paste0(path_to_save_figure, "/nk_exhausted_gene_flow.png"), plot=exhausted_genes_pseudotime, dpi=500, height=8, width=6)
ggsave(filename=paste0(path_to_save_figure, "/nk_ti_like_gene_flow.png"), plot=ti_like_genes_pseudotime, dpi=500, height=8, width=6)
```

# Pseudobulk gene expression analysis between RT versus CLL

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(DESeq2)
library(limma)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(ggpubr)
```

## Based on Seurat command

```{r}
# Define paths to save and load
path_to_save_seurat_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_seurat_dea.rds")
path_to_save_seurat_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_seurat_hm_gsea.rds")
path_to_save_seurat_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_seurat_go_gsea.rds")
path_to_save_seurat_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_seurat_kegg_gsea.rds")
path_to_save_seurat_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_seurat_reactome_gsea.rds")
path_to_save_seurat_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_seurat_wikipath_gsea.rds")
```

### Find biomarkers of monocytes between RT and CLL conditions

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Find markers between CLL and RT conditions  
Idents(pure_nk_seurat) <- "new_cond" # we used this variable as it contains RT and CLL
pure_nk_seurat <- PrepSCTFindMarkers(pure_nk_seurat) # Prepare SCT assay for DE testing
nk.ge <- FindMarkers(
  pure_nk_seurat,
  ident.1 = "RT",
  ident.2 = "CLL",
  test.use = "wilcox",
  only.pos = FALSE,
  logfc.threshold = 0
)
dea.df <- cbind(gene = rownames(nk.ge), nk.ge)
avg_exp <- AverageExpression(pure_nk_seurat, features = dea.df$gene, assay = "SCT") # Calculate the average expression for the genes
dea.df$baseMean <- rowMeans(avg_exp$SCT) # Calculate the mean across all relevant groups (e.g., global mean for the gene)

# Save the expression results
saveRDS(dea.df, path_to_save_seurat_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/nk_seurat_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Load the saved expression data
dea.df <- readRDS(path_to_save_seurat_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=10}
# Volcano plot for Seurat pseudobulk analysis
genes_of_interest <- c(
  "FASLG", "PRF1", "HAVCR2", "FCGR3A", "FOS", "FOSB", "JUND", "GNLY", "TGFB1", "TNF", "IFNG", "LTA"                           
)

non_zero_min_pval <- sort(unique(dea.df$p_val_adj), decreasing = FALSE)[2] # Find minimum p-value which is not 0
dea.df$adapted_p_val <- ifelse(dea.df$p_val_adj == 0, non_zero_min_pval, dea.df$p_val_adj) # Change zero adj_p_val to non-zero minimum adj_p_val to avoid plotting error

volcano.plt <- volcano_plot(dea.df, 
                            FC=0.25, 
                            x_col="avg_log2FC",
                            y_col="adapted_p_val",
                            xlim=c(-10,10), 
                            ylim=c(-0.1,100), 
                            genes_of_interest=genes_of_interest,
                            boxedLabels = TRUE) 

volcano.plt
```

```{r, message=FALSE, warning=FALSE}
# MA plot for Seurat pseudobulk analysis
ma_data <- dea.df
colnames(ma_data)[colnames(ma_data) == "avg_log2FC"] <- "log2FoldChange"
colnames(ma_data)[colnames(ma_data) == "p_val_adj"]  <- "padj"

ma.plt <- ggmaplot(ma_data, 
                   fdr = 0.05,            # Threshold for significance
                   fc = 2,                # Threshold for FC
                   size = 0.4,            # Dot size
                   palette = c("#B31B21", "#1465AC", "darkgray"),
                   genenames = as.vector(ma_data$gene),
                   top = 0, label.select = genes_of_interest, # set top = 0, then specify genes using label.select argument
                   legend = "right", 
                   font.label = c("bold", 11), 
                   label.rectangle = TRUE,
                   font.legend = "bold",
                   font.main = "bold",
                   ggtheme = ggplot2::theme_minimal(),
                   xlab = "Mean Expression",
                   ylab = "Log2 Fold Change"
                   )

ma.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/seurat_nk_volcano.png"), plot=volcano.plt, dpi=500, height=7, width=9)
ggsave(filename=paste0(path_to_save_figure, "/seurat_nk_ma.png"), plot=ma.plt, dpi=500, height=3.5, width=5)
```

For the gene expression analysis, we found that RT NK cells upregulated genes associated with cytotoxicity (FASLG, PRF1, and FCGR3A) and exhaustive state (HAVCR2), suggesting that NK cells may retain cytotoxic potential in the RT condition.

In contrast, the downregulated DEGs in RT NK cells included members of the AP-1 transcription factor family (FOS, FOSB, and JUND), GNLY (a gene related to cytotoxic function), and several regulatory cytokines (TGFB1, TNF, IFNG, and LTA).  

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea.df$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

### GSEA summary 

We then summarised the common terms found from Hallmark, GO:BP, and canonical pathways.

```{r, message=FALSE, warning=FALSE}
# Bar plot of GSEA 
gsea_bp <- rbind(
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_MITOCHONDRIAL_TRANSLATION",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GGOBP_MITOTIC_CELL_CYCLE_PROCESS",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_INTRINSIC_APOPTOTIC_SIGNALING_PATHWAY",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_FC_RECEPTOR_SIGNALING_PATHWAY",],
  GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_REGULATION_OF_B_CELL_PROLIFERATION",],
  HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_TNFA_SIGNALING_VIA_NFKB",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_IL1_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_EGFEGFR_SIGNALING",],
  WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_TOLLLIKE_RECEPTOR_SIGNALING",]
)

gsea_bp$Description <- gsub("HALLMARK_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("GOBP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("KEGG_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("REACTOME_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("WP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("_V1", "", gsea_bp$Description)
gsea_bp$Description <- gsub("MITOTIC_CELL_CYCLE_PROCESS", "MITOTIC_CELL_CYCLE", gsea_bp$Description)
gsea_bp$Description <- gsub("INTRINSIC_APOPTOTIC_SIGNALING_PATHWAY", "INTRINSIC_APOPTOTIC_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("FC_RECEPTOR_SIGNALING_PATHWAY", "FC_RECEPTOR_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("TNFA_SIGNALING_VIA_NFKB", "TNFA_SIGNALING", gsea_bp$Description)
gsea_bp$Description <- gsub("_", " ", gsea_bp$Description)

gsea_bp$Description <- factor(gsea_bp$Description, levels = gsea_bp$Description)
gsea_bp$Direction <- "Up"
gsea_bp$Direction[gsea_bp$NES<0] <- "Down"

gsea_bar.plt <- ggplot(gsea_bp, aes(x=Description, y=NES, fill=Direction)) +
  geom_bar(stat = "identity", color="black", width = 0.75) +  
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, color = "black"),
    plot.margin = margin(t = 10, r = 10, b = 100, l = 65, unit = "pt"),
    legend.position = "none"
  ) +
  geom_hline(yintercept = 0) + 
  ylab("NES") + xlab(NULL) + 
  scale_fill_manual(values=c("#357EBD", "#D43F3A"))

gsea_bar.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=14}
# GSEA plots
terms_of_interest <- c(
  "GOBP_MITOCHONDRIAL_TRANSLATION", "GOBP_MITOTIC_CELL_CYCLE_PROCESS", "GOBP_INTRINSIC_APOPTOTIC_SIGNALING_PATHWAY",
  "GOBP_FC_RECEPTOR_SIGNALING_PATHWAY", "GOBP_REGULATION_OF_B_CELL_PROLIFERATION", "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "WP_IL1_SIGNALING", "WP_EGFEGFR_SIGNALING", "WP_TOLLLIKE_RECEPTOR_SIGNALING"
  )

gsea_plot.lst <- lapply(terms_of_interest, function(term){
  term_system <- strsplit(term, split = "_")[[1]][1]

  adapted_term <- term %>% 
    str_remove("^(HALLMARK|GOBP|KEGG|REACTOME|WP)_") %>%
    str_replace("MITOTIC_CELL_CYCLE_PROCESS", "MITOTIC_CELL_CYCLE") %>%
    str_replace("INTRINSIC_APOPTOTIC_SIGNALING_PATHWAY", "INTRINSIC_APOPTOTIC_SIGNALING") %>%
    str_replace("FC_RECEPTOR_SIGNALING_PATHWAY", "FC_RECEPTOR_SIGNALING") %>%
    str_replace("TNFA_SIGNALING_VIA_NFKB", "TNFA_SIGNALING") %>%
    str_replace_all("_", " ")
  
  if (term_system == "HALLMARK"){
    gseaplot(HM_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "GOBP"){
    gseaplot(GOBP_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "KEGG") {
    gseaplot(KEGG_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else if (term_system == "REACTOME") {
    gseaplot(REACTOME_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  } else {
    gseaplot(WIKIPATH_gsea, 
             by = "runningScore", 
             title = adapted_term, 
             geneSetID = term,
             base_size = 10) + 
      theme(axis.text.x = element_text(angle = 0, hjust = 1))
  }
})

gsea.plt <- wrap_plots(gsea_plot.lst, ncol=3)
gsea.plt
```

```{r, eval=FALSE}
# Save GSEA results
saveRDS(HM_gsea, path_to_save_seurat_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_seurat_go_gsea)
saveRDS(KEGG_gsea, path_to_save_seurat_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_seurat_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_seurat_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_seurat_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_seurat_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_seurat_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_seurat_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_seurat_wikipath_gsea.csv"))

# Save the figures
ggsave(filename=paste0(path_to_save_figure, "/nk_gsea.png"), plot=gsea_bar.plt, dpi=500, height=5, width=7)
ggsave(filename=paste0(path_to_save_figure, "/nk_gsea_plt.png"), plot=gsea.plt, dpi=500, height=10, width=14)
```

## Based on DESeq2

```{r}
# Define paths to save the results
path_to_save_deseq2_dea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_deseq2_dea.rds")
path_to_save_deseq2_hm_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_deseq2_hm_gsea.rds")
path_to_save_deseq2_go_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_deseq2_go_gsea.rds")
path_to_save_deseq2_kegg_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_deseq2_kegg_gsea.rds")
path_to_save_deseq2_reactome_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_deseq2_reactome_gsea.rds")
path_to_save_deseq2_wikipath_gsea <- here::here("scRNAseq/meta_analysis/R_objects/gene_exprs/nk_deseq2_wikipath_gsea.rds")
```

### Create an DESeq2 object

```{r, message=FALSE, warning=FALSE}
# Prepare single-cell meta-data for DESeq2
pure_nk_seurat$ref[pure_nk_seurat$ref == "Hing_2022"] <- "Hing_2023" # Update reference name
pure_nk_seurat$patient_cond <- paste0("case_", pure_nk_seurat$case, "_", pure_nk_seurat$ref, "_", pure_nk_seurat$new_cond) 
patient_cond_cls <- unique(pure_nk_seurat$patient_cond)
RT_subjects <- patient_cond_cls[grep("RT", patient_cond_cls)]
RT_subjects <- gsub(pattern = "_", replacement = "-", x=RT_subjects) # change "_" to "-" as DESeq2 converted sample names from "_" to "-"

# Create pseudocount matrix for tumour cell object
cts <- AggregateExpression(pure_nk_seurat, 
                           group.by = "patient_cond",
                           assays = "RNA",
                           slot = "counts",
                           resturn.seurat = FALSE) 
cts <- cts$RNA
pseudo_count.mat <- as.data.frame(cts)

# Create metadata for DESeq2 object
batch_map <- c("Penter (2021)" = "1", "Nadeu (2022)" = "2", "Rejeski (2022)" = "3", "Hing (2023)" = "4", "Parry (2023a)" = "5", 
               "Parry (2023b)" = "6", "Rigo (2024)" = "7")
colData <- data.frame(samples = colnames(pseudo_count.mat))
colData$condition <- ifelse(colData$samples %in% RT_subjects, "RT", "CLL")
colData$study <- stringr::word(colData$samples, start = -3, end = -2, sep = fixed("-"))
colData$study <- sub("^(.*)-([0-9].*)$", "\\1 (\\2)", colData$study) # Substitiute string like "Parry-2023b" to "Parry (2023b)"
colData$study <- factor(colData$study, level = c("Penter (2021)", "Nadeu (2022)", "Rejeski (2022)", "Hing (2023)", 
                                                 "Parry (2023a)", "Parry (2023b)", "Rigo (2024)"))
colData$batch <- batch_map[colData$study]
colData$batch <- factor(colData$batch)
rownames(colData) <- colData$samples

# Create a DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = pseudo_count.mat, 
                              colData = colData,
                              design = ~ batch + condition)

# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

### Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6.5}
# Variance stabilised transformation
vsdata <- vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
design_matrix <- model.matrix(~ condition, data = colData(dds))
vs_mat_corrected <- limma::removeBatchEffect(
  vs_mat, 
  batch = colData(dds)$batch, 
  design = design_matrix 
)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2               # variance of each PC
pca_var_percent <- pca_var / sum(pca_var) * 100

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     study = colData(dds)$study)

# Plot PCA
pca.plt <- ggplot(pca_df, aes(PC1, PC2, color = study, shape = condition)) +
  geom_point(size = 3) +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  scale_color_manual(values = c(
    "Penter (2021)" = "#1f77b4", 
    "Nadeu (2022)" = "#ff7f0e",
    "Rejeski (2022)" = "#9632B8" ,
    "Hing (2023)" = "#D43F3A" ,
    "Parry (2023a)" = "#5CB85C",
    "Parry (2023b)" = "#FFB900",
    "Rigo (2024)" = "#46B8DA")) +
  scale_shape_manual(values = c("CLL" = 16, "RT" = 17)) +
  guides(
    color = guide_legend(title = "Study"),
    shape = guide_legend(title = "Condition")
  ) +
  theme_bw() +
  theme(
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 11)
  )

pca.plt
```

```{r, eval=FALSE}
# Save the figure
ggsave(filename=paste0(path_to_save_figure, "/deseq2_nk_pca.png"), plot=pca.plt, dpi=500, height=4.5, width=6.5)
```

### Run DESeq2

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Run DESeq2 for the data after outlier removal
keep <- rowSums(counts(dds) >= 10) >= 3 # Keep genes with at least 10 counts in at least 3 samples
noise_genes <- c(
  grep("^IG[HKL][VDJ]", rownames(dds), value = TRUE),  # Ig genes
  grep("^HLA-", rownames(dds), value = TRUE),          # Patient ID (HLA)
  grep("^MT-", rownames(dds), value = TRUE),           # Mitochondrial stress
  "XIST", "RPS4Y1", "KDM5D", "UTY", "DDX3Y",           # Sex genes
  "S100A8", "S100A9", "S100A12", "CXCL8"               # Acute stress response
)

dds <- dds[keep & (!rownames(dds) %in% noise_genes), ]
dds$condition <- relevel(dds$condition, ref = "CLL")
dds <- DESeq(dds)

# Generate a table for gene expression
res <- results(dds, contrast = c("condition", "RT", "CLL"), independentFiltering = FALSE)
res <- lfcShrink(dds, contrast = c("condition", "RT", "CLL"), res = res, type = "normal")
res$direction <- "NS"
res$direction[res$padj < 0.05 & res$log2FoldChange > 0.25] <- "Up"
res$direction[res$padj < 0.05 & res$log2FoldChange < (-0.25)] <- "Down"
dea.df <- as.data.frame(res)

# Save the expression results
saveRDS(dea.df, path_to_save_deseq2_dea)
write.csv(dea.df, here::here("scRNAseq/meta_analysis/supplm_file/nk_deseq2_dea.csv"))
```

```{r, message=FALSE, warning=FALSE}
# Show the expression results
dea.df <- readRDS(path_to_save_deseq2_dea)
DT::datatable(dea.df, options = list(scrollX = TRUE))
```

From the results, we observed that pseudobulk analysis using DESeq2 failed to detect DEGs, likely due to low read counts resulting from NK cell aggregation.

### GSEA

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df) 
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector)

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector)

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector)

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save the enrichment results
saveRDS(HM_gsea, path_to_save_deseq2_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_deseq2_go_gsea)
saveRDS(KEGG_gsea, path_to_save_deseq2_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_deseq2_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_deseq2_wikipath_gsea)

# Save as data frames
write.csv(HM_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_deseq2_hm_gsea.csv"))
write.csv(GOBP_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_deseq2_go_gsea.csv"))
write.csv(KEGG_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_deseq2_kegg_gsea.csv"))
write.csv(REACTOME_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_deseq2_reactome_gsea.csv"))
write.csv(WIKIPATH_GSEA.result, here::here("scRNAseq/meta_analysis/supplm_file/nk_deseq2_wikipath_gsea.csv"))
```

# Session Information

```{r}
sessionInfo()
```
