---
title: "Parry_2023_systematic_review"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

This source code we analysed single-cell RNA-seq for RT and CLL conditions both gene expression and functional enrichment analysis before we will compare the results to other studies for our systematic review. We analysed for [Parry et. al. (2023)](https://www.cell.com/cancer-cell/fulltext/S1535-6108(23)00306-9?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS1535610823003069%3Fshowall%3Dtrue) study which performed single-cell analysis between CLL and RT conditions in three patients. 

Moreover, this study sorted cell types such as CLL and RT tumour cells, and non-tumour cells using flow cytometry before single-cell sequencing. Then, the samples were based on the sorted cells. In this code, we analysed them by assuming that each sorted cells are individual samples. We will merge them all and reannotate the cell types using single-cell pipeline. 

We will also analyse them by considering the flow cytometry sorted method in the next source code in file `2-Parry_2023_analyse_sorted_tumour.qmd`, by selecting only tumour cells and performing gene expression and functional enrichment analysis based on histological type annoated from the flow cytometry without considering origin cases. 

## Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(DoubletFinder)
library(SingleR)
library(celldex)
library(sctransform)
```

```{r}
# Check root path
here::here()
```

## Define paths

```{r}
# Define paths to load and save 
path_to_metadata <- here::here("scRNAseq/systematic_review/4-Parry_2023/sample_metadata.csv")
path_to_count_matrices <- here::here("scRNAseq/phs002458.v2.p1/expression_data/")
path_to_save_obj_lst <- here::here("scRNAseq/systematic_review/4-Parry_2023/R_objects/Parry_seu_lst.rds")
path_to_save_merged_obj <- here::here("scRNAseq/systematic_review/4-Parry_2023/R_objects/Parry_merged_seu.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/systematic_review/4-Parry_2023/R_objects/Parry_integrated_seu.rds")
path_to_save_tumour_obj <- here::here("scRNAseq/systematic_review/4-Parry_2023/R_objects/Parry_tumour_seu.rds")
path_to_save_filtered_tumour_obj <- here::here("scRNAseq/systematic_review/4-Parry_2023/R_objects/Parry_filtered_tumour_seu.rds")
path_to_save_dea <- here::here("scRNAseq/systematic_review/4-Parry_2023/expression_data/Parry_dea.rds")
path_to_save_hm_gsea <- here::here("scRNAseq/systematic_review/4-Parry_2023/enrichment_data/Parry_hm_gsea.rds")
path_to_save_go_gsea <- here::here("scRNAseq/systematic_review/4-Parry_2023/enrichment_data/Parry_go_gsea.rds")
```

## Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r}
# Load the sample metadata
sample_metadata <- read.csv(path_to_metadata)
DT::datatable(sample_metadata, options = list(scrollX = TRUE))
```

We will import count matrices and convert them to Seurat objects with quality control and pre-processing.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create the objects and collect in a list
path_to_download <- str_c(path_to_count_matrices, "/", sample_metadata$SRA_RUN, "/count_matrices/outs/filtered_feature_bc_matrix/")
seurat.lst <- purrr::pmap(
  list(path_to_download, 
       sample_metadata$SRA_RUN, 
       sample_metadata$subject_id, 
       sample_metadata$histological_type,
       sample_metadata$Condition, 
       sample_metadata$Tissue), 
  function(path, sra_run, case, histological_type, cond, tissue){
    count_matrix <- Read10X(path)
    seurat.obj <- CreateSeuratObject(count_matrix, min.cells=10) %>% 
      RenameCells(add.cell.id = sra_run)
    seurat.obj[["ref"]] <- "Parry_2023"
    seurat.obj[["case_id"]] <- sra_run
    seurat.obj[["histological_type"]] <- histological_type
    seurat.obj[["case"]] <- case
    seurat.obj[["cond"]] <- cond
    seurat.obj[["tissue"]] <- tissue
    seurat.obj <- seurat_quality_control(seurat.obj)
    seurat.obj <- seurat_preprocess(seurat.obj)
    if (ncol(seurat.obj) > 500) {  # pass doublet detection if objects have too low cells
      seurat.obj <- doublet_filter(seurat.obj, sct=FALSE)
    } else {
      message("Skipping doublet filtering for object with <500 cells")
    }
    return(seurat.obj)
    }
  )
names(seurat.lst) <- sample_metadata$SRA_RUN
```

```{r, eval=FALSE}
# Save the list
saveRDS(seurat.lst, path_to_save_obj_lst)
```

## Merge and integrate the objects

```{r, eval=FALSE}
# Merge the objects
seurat <- seurat.lst[[1]]
for (i in 2:length(seurat.lst)) {
  seurat <- merge(x = seurat, y = seurat.lst[[i]])
  seurat.lst[[i]] <- NA
}
rm(seurat.lst)
```

```{r, eval=FALSE}
# Save the merged object
saveRDS(seurat, path_to_save_merged_obj)
```

We used Harmony package for data integration and batch correction, and annotated the cell types using SingleR package.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Reanalysis after merging and data integration 
DefaultAssay(seurat) <- "RNA"
seurat <- JoinLayers(seurat)
seurat <- SCTransform(seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
seurat <- RunPCA(seurat)
seurat <- RunHarmony(seurat, group.by.vars = "case_id", dims = 1:30) 
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
seurat <- FindNeighbors(seurat, dims = 1:30)              
seurat <- FindClusters(
    seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 0.9, by = 0.1)
  )
seurat <- singleR_cell_annotation(seurat) 
```

## Save the integrated object

```{r, eval=FALSE}
saveRDS(seurat, path_to_save_integrated_obj)
```

## Visualisation after data integration

```{r}
# Define path
path_to_obj <- path_to_save_integrated_obj

# Load the data
seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE}
# Plot based on case_id
case_id.plt <- DimPlot(seurat, reduction = "umap", group.by = "case_id") + 
  labs(y= "UMAP 2", x = "UMAP 1")
case_id.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=10}
# Plot based on case, histoloigcal_type, conditions, cell types, and clusters
case.plt <- DimPlot(seurat, reduction = "umap", group.by = "case") + 
  labs(y= "UMAP 2", x = "UMAP 1")

ref.plt <- DimPlot(seurat, reduction = "umap", group.by = "ref") + 
  labs(y= "UMAP 2", x = "UMAP 1")

histo.plt <- DimPlot(seurat, reduction = "umap", group.by = "histological_type") + 
  labs(y= "UMAP 2", x = "UMAP 1")

cond.plt <- DimPlot(seurat, reduction = "umap", group.by = "cond") + 
  labs(y= "UMAP 2", x = "UMAP 1")

tissue.plt <- DimPlot(seurat, reduction = "umap", group.by = "tissue") +
  labs(y= "UMAP 2", x = "UMAP 1")

cell.plt <- DimPlot(seurat, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")

(case.plt + ref.plt) / (histo.plt + cond.plt) / (tissue.plt + cell.plt)
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

```{r, fig.height=10, fig.width=10, warning=FALSE, message=FALSE}
# Visualise the varied clusters
varied_clusters.plt <- plot_varying_cluster_resolution(seurat)
varied_clusters.plt
```

```{r, fig.height=4, fig.width=5}
# Plot seurat clusters at resolution=0.1
clusters.plt <- DimPlot(seurat, reduction = "umap", group.by = "SCT_snn_res.0.1", raster=FALSE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

clusters.plt
```

## Tumour cell extraction

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Extract tumour cells with excluding cells from cluster 4
tumour_seurat <- subset(seurat, subset = singleR.labels_main == "B cells")
tumour_seurat <- SCTransform(tumour_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
tumour_seurat <- RunPCA(tumour_seurat)
tumour_seurat <- RunHarmony(tumour_seurat, group.by.vars = "case_id", dims = 1:30) 
tumour_seurat <- RunUMAP(tumour_seurat, dims = 1:30, reduction = "harmony")
tumour_seurat <- FindNeighbors(tumour_seurat, dims = 1:30)              
tumour_seurat <- FindClusters(tumour_seurat, resolution=0.25)
```

```{r, eval=FALSE}
# Save the tumour cells
saveRDS(tumour_seurat, path_to_save_tumour_obj)
```

```{r}
# Define path
path_to_obj <- path_to_save_tumour_obj

# Load the data
tumour_seurat <- readRDS(path_to_obj)
```

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=9.5}
# Plot based on cases, histological types, conditions, and clusters
case.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "case") +
  labs(y= "UMAP 2", x = "UMAP 1")

histo.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "histological_type") +
  labs(y= "UMAP 2", x = "UMAP 1")

cond.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "cond") + 
  labs(y= "UMAP 2", x = "UMAP 1")

clusters.plt <- DimPlot(tumour_seurat, reduction = "umap", group.by = "seurat_clusters") +
  labs(y= "UMAP 2", x = "UMAP 1")

(case.plt + histo.plt) / (cond.plt + clusters.plt)
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(tumour_seurat, 
            features = c("HBM", "HBA1", "CD3D", "CD8A", "CD4", "CD19", "CD79A", "CD14", "NKG7"),
            reduction = "umap") 
```

```{r, fig.height=4, fig.width=5}
# Plot seurat clusters 
clusters.plt <- DimPlot(tumour_seurat, 
                        reduction = "umap", 
                        group.by = "seurat_clusters", 
                        label = TRUE,
                        repel = TRUE,
                        label.size = 4,
                        raster=FALSE) + 
  labs(y= "UMAP 2", x = "UMAP 1")

clusters.plt
```

As there are some erythrocytes and NK cells (cluster 15), we will remove this cluster before psedobulk analysis.

```{r, eval=FALSE}
# Filter out NK cells
filtered_tumour_seurat <- subset(tumour_seurat, subset = seurat_clusters != 15) 

# Save the filtered tumour object
saveRDS(filtered_tumour_seurat, path_to_save_filtered_tumour_obj)
```

## Pseudobulk gene expression analysis and GSEA

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

### Find biomarkers of B cells between RT and CLL conditions

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Find markers between CLL and RT-like conditions  
Idents(filtered_tumour_seurat) <- "cond" # change the default identity
tumour_seurat <- PrepSCTFindMarkers(filtered_tumour_seurat) # Prepare SCT assay for DE testing
tumour.ge <- FindMarkers(
  filtered_tumour_seurat,
  ident.1 = "RT",
  ident.2 = "CLL",
  test.use = "wilcox",
  only.pos = FALSE,
  logfc.threshold = 0
)
dea.df <- cbind(gene = rownames(tumour.ge), tumour.ge)
```

```{r, eval=FALSE}
# Save gene expression data
saveRDS(dea.df, path_to_save_dea)
```

### Gene Set Enrichment Analysis (GSEA)

```{r}
# Load the saved expression data
path_to_dea <- path_to_save_dea
dea.df <- readRDS(path_to_dea)
```

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea.df$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

#### Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)
```

```{r}
# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

No significant terms are detected. We will explore the trends of other non-significant hallmark terms reported from other studies such as Nadeu et al. (2022) and Hing et al (2023) (in original results). From those studies, the terms such as E2F targets, Myc targets, and oxidative phosphorylation (OXPHOS) were upregulated.

```{r, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

#### GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)
```

```{r}
# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

No significant terms are detected. We will explore the trends of other non-significant GO:BP terms reported from other studies such as Nadeu et al. (2022) and Hing et al (2023) (in the original results). From those studies, the terms such as OXPHOS and mitochondrial translation were upregulated while BCR signalling was downregulated.

```{r, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

```{r, eval=FALSE}
# Save gene expression and GSEA data
saveRDS(GOBP_gsea, path_to_save_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_go_gsea)
```

### Pseudobulk analysis using DESeq2

As the gene expression analysis using `FindMarkers()` in Seurat did not provide signicant enrichment terms in hallmark and GO:BP, we will analysis with pseudobulk approach using DESeq2 since we reanalyse from the previous step for these datasets using DESeq2 we found these enrichment terms, such as OXPHOS, E2F, Myc, and BCR signalling pathways, are consistent with Nadeu and Hing's studies.

```{r, message=FALSE, warning=FALSE}
# Import the relevant packages
library(DESeq2)
```

```{r}
# Define paths to save the results
path_to_save_deseq2_dea <- here::here("scRNAseq/systematic_review/4-Parry_2023/pseudo_bulk_deseq2/Parry_deseq2_dea.rds")
path_to_save_deseq2_hm_gsea <- here::here("scRNAseq/systematic_review/4-Parry_2023/pseudo_bulk_deseq2/Parry_deseq2_hm_gsea.rds")
path_to_save_deseq2_go_gsea <- here::here("scRNAseq/systematic_review/4-Parry_2023/pseudo_bulk_deseq2/Parry_deseq2_go_gsea.rds")
```

```{r, eval=FALSE}
# Load the object
filtered_tumour_seurat <- readRDS(path_to_save_filtered_tumour_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Pseudobulk analysis by aggregation based on cases
RT_subjects <- unique(sample_metadata[sample_metadata$Condition == "RT", "subject_id"])
dea.df <- pseudobulk_deseq2(filtered_tumour_seurat, aggregate_group.by = "case", RT_subjects)

# Save the expression results
saveRDS(dea.df, path_to_save_deseq2_dea)
```

```{r}
# Show some expression results
dea.df <- readRDS(path_to_save_deseq2_dea)
head(dea.df)
```

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df) 
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

#### GSEA with Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector)

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

#### GSEA with GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector)

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

These results validate our single-cell downstream analysis although gene expression analysis using `FindMarkers()` in Seurat package did not provide the consistent results compared to other studies. However, to be consistent with our standard pipeline for analysis, we will collect the result from `FindMarkers()` for further steps. 

```{r, eval=FALSE}
# Save the enrichment results
saveRDS(HM_gsea, path_to_save_deseq2_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_deseq2_go_gsea)
```



