---
title: "Hacken_2023_GSE186137_bulkRNA-seq_analysis"
author: "Pakorn Sagulkoo"
format: html
editor: visual
---

In this code, we will analyse mouse bulk RNA-seq from [Hacken et al. (2023)](https://aacrjournals.org/bloodcancerdiscov/article/4/2/150/718388/In-Vivo-Modeling-of-CLL-Transformation-to-Richter). This study employed a mouse experimental model in which a CLL- and RT-like condition was induced using *del*(13q) *Cd19*-Cas9-GFP mice engrafted with LSK cells transduced with sgRNA-expressing lentivirus, followed by multiple loss-of-function alterations. After we performed Salmon converting FASTQ files to pseudocount data in each sample, we will convert them to a count matrix before we use the matrix for DESeq2 for further gene expression analysis.

## Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Import the relavant packages
library(tximport)
library(ensembldb)
library(biomaRt)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(DESeq2)
```

```{r}
# Check root path
here::here()
```

## Define paths

```{r}
# Define paths to load and save files
path_to_samples <- here::here("RNAseq/Hacken_2023_GSE186137")
path_to_sample_metadata <- here::here("RNAseq/systematic_review/1-Hacken_2023_GSE186137/sample_metadata.csv")
path_to_ref_transcriptome <- "/home/pakorns/hc-storage/fastq_files/RNAseq/transcriptome_ref"
path_to_save_txi <- here::here("RNAseq/systematic_review/1-Hacken_2023_GSE186137/count_matrices/Hacken_txi.rds")
path_to_save_dea <- here::here("RNAseq/systematic_review/1-Hacken_2023_GSE186137/expression_data/Hacken_dea.rds")
path_to_save_hm_gsea <- here::here("RNAseq/systematic_review/1-Hacken_2023_GSE186137/enrichment_data/Hacken_hm_gsea.rds")
path_to_save_go_gsea <- here::here("RNAseq/systematic_review/1-Hacken_2023_GSE186137/enrichment_data/Hacken_go_gsea.rds")
```

## Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r, message=FALSE, warning=FALSE}
# Load sample metadata
sample_metadata <- read.csv(path_to_sample_metadata) 
DT::datatable(sample_metadata, options = list(scrollX = TRUE))
```

```{r}
# Create a vector collecting Salmon quant.sf path in each sample
path_to_download <- str_c(path_to_samples, "/", sample_metadata$SRA_RUN, "/quant.sf")
names(path_to_download) <- sample_metadata$SRA_RUN
path_to_download
```

## Convert salmon pseudocounts to an expression matrix

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Load the ensembl mm39 gtf file
path_to_mm39_transcriptome <- str_c(path_to_ref_transcriptome, "/Mus_musculus.GRCm39.115.gtf.gz")
edb <- ensDbFromGtf(path_to_mm39_transcriptome) # Return edb = "./Mus_musculus.GRCm39.115.sqlite"
```

```{r, message=FALSE, warning=FALSE}
# Make the transcript-to-gene map (tx2gene)
edb <- "./Mus_musculus.GRCm39.115.sqlite"
edb_obj <- EnsDb(edb)
tx2gene <- transcripts(edb_obj,
                       columns = c("tx_id","gene_id","gene_name"),
                       return.type="DataFrame")

# convert to a plain data.frame for tximport
tx2gene <- as.data.frame(tx2gene)
head(tx2gene)
```

```{r, message=FALSE, warning=FALSE}
# Load pseudocount files
txi <- tximport(path_to_download, 
                type = "salmon", 
                tx2gene = tx2gene, 
                ignoreTxVersion = TRUE 
               )

# This gives you txi$counts, txi$abundance (TPM), txi$length
head(txi$counts)
```

```{r, eval=FALSE}
# Save the expression file
saveRDS(txi, path_to_save_txi)
```

## Transcriptomics analysis using DESeq2

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(DESeq2)
library(sva)
library(limma)
```

### Create a DESeq2 object

```{r}
# Create metadata for DESeq2 object
colData <- sample_metadata[, c("SRA_RUN", "condition")] %>%
  rename(samples = "SRA_RUN") 
colData$group <- ifelse(colData$condition == "RT", "RT", "Others")
rownames(colData) <- colData$samples
colData$condition <- factor(colData$condition)
DT::datatable(colData, options = list(scrollX = TRUE))
```

```{r, message=FALSE, warning=FALSE}
# Create the object
dds <- DESeqDataSetFromTximport(txi, colData, ~ condition)
```

```{r, message=FALSE, warning=FALSE}
# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
  
# Run DESeq2
dds$condition <- relevel(dds$condition, ref = "CTRL")
dds <- DESeq(dds)
```

### Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6}
# Variance stabilised transformation
vsdata <- DESeq2::vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
vs_mat_corrected <- removeBatchEffect(vs_mat, batch = colData(dds)$batch)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     sample = rownames(colData(dds)))

# Plot
ggplot(pca_df, aes(PC1, PC2, color = condition)) +
  geom_point(size = 3) +
  theme_bw()
```

From the PCA plot we can see that each sample is grouped based on condition although samples in CLL/RS group distribute between RT and CLL groups, consistent with their mixing phenotypes.

### Convert mouse to human ensemble ids

```{r}
# Get orthologs
path_to_ortholog <- str_c(path_to_ref_transcriptome, "/mouse_human_ortholog.txt")
orthologs <- read.csv(path_to_ortholog)
head(orthologs)
```

### RT vs CLL

```{r, message=FALSE, warning=FALSE}
# Generate a table for gene expression
res_RT_vs_CLL <- lfcShrink(dds, contrast = c("condition", "RT", "CLL"), type = "normal")
head(res_RT_vs_CLL)
```

```{r}
# Use left_join to convert mouse to human ensembl ids
human_res_RT_vs_CLL <- as.data.frame(res_RT_vs_CLL) %>%
  rownames_to_column("Mouse_Ensembl_ID") %>%
  left_join(orthologs, by = c("Mouse_Ensembl_ID" = "Mouse.gene.stable.ID"))
```

```{r, eval=FALSE}
# Save the expression result
saveRDS(human_res_RT_vs_CLL, path_to_save_dea)
```

### RT vs others

```{r, message=FALSE, warning=FALSE}
# Create the object and colData based on group (RT and others) instead of condition
dds <- DESeqDataSetFromTximport(txi, colData, ~ group)

# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
  
# Run DESeq2
dds$condition <- relevel(dds$group, ref = "Others")
dds <- DESeq(dds)
```

```{r, message=FALSE, warning=FALSE}
# Generate a table for gene expression
res_RT_vs_others <- lfcShrink(dds, contrast = c("group", "RT", "Others"), type = "normal")
head(res_RT_vs_others)
```

```{r}
# Use left_join to convert mouse to human ensembl ids
human_res_RT_vs_others <- as.data.frame(res_RT_vs_others) %>%
  rownames_to_column("Mouse_Ensembl_ID") %>%
  left_join(orthologs, by = c("Mouse_Ensembl_ID" = "Mouse.gene.stable.ID"))
head(human_res_RT_vs_others)
```

#### Plot a heatmap of gene expression

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(pheatmap)
library(RColorBrewer)
```

```{r}
# Data transformation
vsd <- DESeq2::vst(dds, blind=FALSE)       # Variance-stabilizing transformation
#rld <- rlog(dds_filtered, blind=FALSE)      # Regularized log transformation

# Extract transformed expression matrix
expr_mat <- assay(vsd)

# Calculate z-scores
zscore_mat <- t(scale(t(expr_mat)))
```

```{r}
# Define samples in each condition
CTRL_sample <- colData$samples[colData$condition == "CTRL"]
CLL_sample <- colData$samples[colData$condition == "CLL"]
CLL_RT_sample <- colData$samples[colData$condition == "CLL/RT"]
RT_sample <- colData$samples[colData$condition == "RT"]
```

```{r}
# Extract phenotypes info from column names
conditions <-  ifelse(colnames(zscore_mat) %in% CTRL_sample, "CTRL", 
                      ifelse(colnames(zscore_mat) %in% CLL_sample, "CLL", 
                             ifelse(colnames(zscore_mat) %in% CLL_RT_sample, "CLL/RT", "RT")))

# Create annotation dataframe
annotation_col <- data.frame(
  Condition = factor(conditions, levels = c("CTRL", "CLL", "CLL/RT", "RT"))
)
annotation_col <- annotation_col[, "Condition", drop = FALSE]  # keep as data frame
rownames(annotation_col) <- colnames(zscore_mat)

# Order by Condition and then Case 
order_cols <- order(annotation_col$Condition)
zscore_mat_ordered <- zscore_mat[, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]
```

```{r, fig.height=8, fig.width=4}
# Filter for significant DE genes
sig_genes <- human_res_RT_vs_others[!is.na(human_res_RT_vs_others$padj) & human_res_RT_vs_others$padj < 0.05 & abs(human_res_RT_vs_others$log2FoldChange) > 2, ]

# Order genes by log2FoldChange (descending: up -> down)
gene_order <- sig_genes$Mouse_Ensembl_ID[order(sig_genes$log2FoldChange, decreasing = FALSE)]

# Subset zscore matrix
zscore_mat_sig <- zscore_mat_ordered[gene_order, ]

# Plot heatmap
pheatmap(
  zscore_mat_sig,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  gaps_col = which(diff(as.numeric(factor(annotation_col_ordered$Condition))) != 0)
)
```

## Gene Set Enrichment Analysis (GSEA)

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
``` 

### RT vs CLL

```{r}
# Manipulate the expression data
dea <- human_res_RT_vs_CLL[!is.na(human_res_RT_vs_CLL$Gene.stable.ID), ]
lfc_vector <- dea$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea$Gene.stable.ID
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
lfc_vector <- lfc_vector[!duplicated(names(lfc_vector))]
```

#### Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will explore the trends of hallmark terms reported from other single-cell pseudobulk studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as E2F targets, Myc targets, and oxidative phosphorylation (OXPHOS) were upregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

#### GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will also explore the trends of GO:BP terms reported from other single-cell studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as OXPHOS and mitochondrial translation were upregulated while BCR signalling was downregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

```{r, eval=FALSE}
# Save gene expression and GSEA data
saveRDS(HM_gsea, path_to_save_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_go_gsea)
```

### RT vs others

```{r}
# Manipulate the expression data
dea <- human_res_RT_vs_others[!is.na(human_res_RT_vs_others$Gene.stable.ID), ]
lfc_vector <- dea$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea$Gene.stable.ID
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
lfc_vector <- lfc_vector[!duplicated(names(lfc_vector))]
```

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

From the enrichment results we found that significantly upregulated hallmark terms, such as E2F targets, Myc targets, and G2M checkpoint, were consistent with the original results in Figure 4B. In addition, significantly downregulated terms, such as IFN-gamma response and IL6-JAK-STAT signalling, were also consistent although no IL2-STAT5 signalling was detected. Interestingly, we found OXPHOS was significantly upregulated between RT and others while the term was not significant when comparing only RT and CLL conditions.

