---
title: "Overlapping_pathway_DESeq2"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

After we carried out psedobulk gene expression analysis and GSEA for several RT scRNA-sequences datasets, we will evaluate the consistency of functional enrichment results across those studiescalculated from Pseudobulk analysis using DESeq2. Then we will explore overlapping enrichment terms among those studies.

## Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Import the relavant packages
library(UpSetR)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
```

```{r}
# Check root path
here::here()
```

## Define paths

```{r}
# Define paths to load and save files
path_to_single_cell_analysis <- here::here("scRNAseq/systematic_review")
```

## Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

## Define names used in files and labels

```{r}
# List file names of all studies
study.lst <- c("1-Penter_2021_GSE165087", "2-Nadeu_2022", "3-Hing_2022_GSE205704", "4-Parry_2023b", "5-Rigo_2024")
study.lst
```

```{r}
# Create names for list later
names.lst <- sub(
  "^\\d+-([A-Za-z]+)_([0-9]{4}[A-Za-z]?).*$",  # capture name and year
  "\\1(\\2)",                         # output as Name(Year)
  study.lst
)
names.lst
```

```{r}
# Extract only name of studies
study_names <- sub("^\\d+-([^_0-9]+).*", "\\1", study.lst)
study_names
```

We performed GSEA using five main terms such as Hallmark, GO:BP, KEGG, Reactome, and WikiPath. We will identify these significant terms in both up- and downregulation.

## Overlapping of GSEA results based on DESeq2-based pseudobulk analysis

### Hallmark

```{r}
# Import the results from all studies and collect in a list
path_to_download <- str_c(path_to_single_cell_analysis, "/", study.lst, "/pseudo_bulk_deseq2/", study_names, "_deseq2_hm_gsea.rds")
path_to_download <- path_to_download
hm_gsea.lst <- purrr::pmap(
  list(path_to_download), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)
names(hm_gsea.lst) <- names.lst
```

#### Upregulated terms

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant upregulated hallmark terms
sig_term.lst <- purrr::pmap(
  list(hm_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

In comparison with the results from Seurat-based pseudobulk approach, DESeq2-based result has a higher common terms than the previous approach. Then we will identify those common terms. This will be explored using set operations and UpSet plots. For example, if we have three sets (A, B, and C), the expression ‘members from A ∩ B’ refers to terms that are present in both A and B but excluded from those present in all three sets (A, B, and C).

```{r}
# Find common hallmark terms in all studies
five_common_hm <- Reduce(intersect, sig_term.lst)
five_common_hm
```

```{r}
# Find common hallmark terms in four studies: Nadeu, Hing, Parry, and Rigo
four_common_hm <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Penter(2021)"])
setdiff(four_common_hm, five_common_hm)
```

```{r}
# Find common hallmark terms in four studies: Penter, Nadeu, Hing, and Rigo
four_common_hm <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Parry(2023b)"])
setdiff(four_common_hm, five_common_hm)
```

```{r}
# Find common hallmark terms in four studies: Penter, Nadeu, Hing, and Parry
four_common_hm <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Rigo(2024)"])
setdiff(four_common_hm, five_common_hm)
```

These common upregulated hallmark terms were related to OXOSPHOS, glucose and lipid metabolisms, E2F and Myc signalling, and Cell cycle and proliferation.

#### Downregulated terms

We will exclude Rigo(2024) from the overapping downregulated hallmark terms as no significant downregulated terms were detected.  

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant downregulated hallmark terms
sig_term.lst <- purrr::pmap(
  list(hm_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

Although no common terms were found in all four studies, we will idenity common terms in two and three studies.

```{r}
# Find common hallmark terms in three studies: Penter, Nadeu, and Hing
three_common_hm <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Parry(2023b)"])
three_common_hm
```

```{r}
# Find common hallmark terms in three studies: Nadeu and Hing
two_common_hm <- Reduce(intersect, sig_term.lst[c("Nadeu(2022)", "Hing(2022)")])
setdiff(two_common_hm, three_common_hm)
```

```{r}
# Find common hallmark terms in three studies: Penter and Nadeu
two_common_hm <- Reduce(intersect, sig_term.lst[c("Penter(2021)", "Nadeu(2022)")])
setdiff(two_common_hm, three_common_hm)
```

### GO:BP

```{r}
# Import the results from all studies and collect in a list
path_to_download <- str_c(path_to_single_cell_analysis, "/", study.lst, "/pseudo_bulk_deseq2/", study_names, "_deseq2_go_gsea.rds")
path_to_download <- path_to_download
go_gsea.lst <- purrr::pmap(
  list(path_to_download), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)
names(go_gsea.lst) <- names.lst
```

#### Upregulated terms

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant upregulated GO:BP terms
sig_term.lst <- purrr::pmap(
  list(go_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

```{r}
# Find common GO:BP terms in all studies
five_common_go <- Reduce(intersect, sig_term.lst)
five_common_go
```

```{r}
# Find common GO:BP terms in four studies: Penter, Nadeu, Parry, and Rigo
four_common_go <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Hing(2022)"])
setdiff(four_common_go, five_common_go)
```

```{r}
# Find common GO:BP terms in four studies: Nadeu, Hing, Parry, and Rigo
four_common_go <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Penter(2021)"])
setdiff(four_common_go, five_common_go)
```

```{r}
# Find common GO:BP terms in four studies: Penter, Nadeu, Hing, and Rigo
four_common_go <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Parry(2023b)"])
setdiff(four_common_go, five_common_go)
```

```{r}
# Find common GO:BP terms in four studies: Penter, Nadeu, Hing, and Parry
four_common_go <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Rigo(2024)"])
setdiff(four_common_go, five_common_go)
```

The common upregulated GO:BP terms were Consistent with the common upregulated hallmark terms that the both terms were associated with OXPHOS and nucleotide metabolism. indicating increased cell proliferation activity.  

#### Downregulated terms

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant downregulated GO:BP terms
sig_term.lst <- purrr::pmap(
  list(go_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

Overall, no common downregulated GO:BP terms in all studies were detected. We will idenity common terms in two studies.

```{r}
# Common terms between Penter and Rigo
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Rigo(2024)")])
```

```{r}
# Common terms between Penter and Parry
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Parry(2023b)")])
```

```{r}
# Common terms between Nadeu and Hing
Reduce(intersect, sig_term.lst[c("Nadeu(2022)", "Hing(2022)")])
```

```{r}
# Common terms between Nadeu and Rigo
Reduce(intersect, sig_term.lst[c("Nadeu(2022)", "Rigo(2024)")])
```

```{r}
# Common terms between Nadeu and Parry
Reduce(intersect, sig_term.lst[c("Nadeu(2022)", "Parry(2023b)")])
```

```{r}
# Common terms between Penter and Nadeu
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Nadeu(2022)")])
```

### KEGG pathways

```{r}
# Import the results from all studies and collect in a list
path_to_download <- str_c(path_to_single_cell_analysis, "/", study.lst, "/pseudo_bulk_deseq2/", study_names, "_deseq2_kegg_gsea.rds")
path_to_download <- path_to_download
kegg_gsea.lst <- purrr::pmap(
  list(path_to_download), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)
names(kegg_gsea.lst) <- names.lst
```

#### Upregulated terms

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant upregulated KEGG terms
sig_term.lst <- purrr::pmap(
  list(kegg_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

Although no common terms in all five studies were detected, we will identify common terms in four studies.

```{r}
# Common KEGG terms in four studies: Penter, Nadeu, Hing, and Parry
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Nadeu(2022)", "Hing(2022)", "Parry(2023b)")])
```

```{r}
# Common KEGG terms in four studies: Penter, Nadeu, Hing, and Rigo
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Nadeu(2022)", "Hing(2022)", "Rigo(2024)")])
```

Generally, the common upregulated KEGG terms were related to OXPHOS and nucleotide metabolisms, consistent with hallmark and GO:BP upregulated terms.

#### Downregulated terms

As no significant downregulated KEGG terms in Rigo(2024) are detected, we will not show it.  

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant downregulated KEGG terms
sig_term.lst <- purrr::pmap(
  list(kegg_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

No common terms are detected. We will identify common terms in two studies.

```{r}
# Common terms between Penter and Parry
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Parry(2023b)")])
```

```{r}
# Common terms between Penter and Nadeu
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Nadeu(2022)")])
```

### Reactome pathways

```{r}
# Import the results from all studies and collect in a list
path_to_download <- str_c(path_to_single_cell_analysis, "/", study.lst, "/pseudo_bulk_deseq2/", study_names, "_deseq2_reactome_gsea.rds")
path_to_download <- path_to_download
reactome_gsea.lst <- purrr::pmap(
  list(path_to_download), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)
names(reactome_gsea.lst) <- names.lst
```

#### Upregulated terms

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant upregulated Reactome terms
sig_term.lst <- purrr::pmap(
  list(reactome_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

```{r}
# Common upregulated reactome terms in all studies
five_common_reactome <- Reduce(intersect, sig_term.lst)
five_common_reactome
```

```{r}
# Common upregulated reactome terms in four studies
four_common_reactome <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Parry(2023b)"])
setdiff(four_common_reactome, five_common_reactome)
```

Overall, the upregulated reactome terms were consistent with previous upregulated terms which were related to OXPHOS, nucleotide metabolisms, cell cycle, and cell proliferation.

#### Downregulated terms

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant downregulated Reactome terms
sig_term.lst <- purrr::pmap(
  list(reactome_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

No common terms are detected. We will identify common terms in two studies.

```{r}
# Common terms between Penter and Parry
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Parry(2023b)")])
```

```{r}
# Common terms between Nadeu and Parry
Reduce(intersect, sig_term.lst[c("Nadeu(2022)", "Parry(2023b)")])
```

```{r}
# Common terms between Penter and Nadeu
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Nadeu(2022)")])
```

### WikiPath

```{r}
# Import the results from all studies and collect in a list
path_to_download <- str_c(path_to_single_cell_analysis, "/", study.lst, "/pseudo_bulk_deseq2/", study_names, "_deseq2_wikipath_gsea.rds")
path_to_download <- path_to_download
wikipath_gsea.lst <- purrr::pmap(
  list(path_to_download), 
  function(path){
    gsea <- readRDS(path)
    return(gsea)
  }
)
names(wikipath_gsea.lst) <- names.lst
```

#### Upregulated terms

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant upregulated WikiPath terms
sig_term.lst <- purrr::pmap(
  list(wikipath_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES > 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

Although no common terms in all five studies were detected, we will identify common terms in four and three studies.

```{r}
# Common WikiPath terms in four studies: Penter, Nadeu, Hing, and Rigo
four_common_wikipath <- Reduce(intersect, sig_term.lst[names(sig_term.lst) != "Parry(2023b)"])
four_common_wikipath 
```

```{r}
# Common WikiPath terms in three studies: Hing, Parry, and Rigo
three_common_wikipath <- Reduce(intersect, sig_term.lst[names(sig_term.lst) %in% c("Penter(2021)", "Nadeu(2022)")])
setdiff(three_common_wikipath, four_common_wikipath)
```

```{r}
# Common WikiPath terms in three studies: Nadeu, Hing, and Rigo
three_common_wikipath <- Reduce(intersect, sig_term.lst[!(names(sig_term.lst) %in% c("Penter(2021)", "Parry(2023b)"))])
setdiff(three_common_wikipath, four_common_wikipath)
```

```{r}
# Common WikiPath terms in three studies: Nadeu, Hing, and Parry
three_common_wikipath <- Reduce(intersect, sig_term.lst[!(names(sig_term.lst) %in% c("Penter(2021)", "Rigo(2024)"))])
setdiff(three_common_wikipath, four_common_wikipath)
```

```{r}
# Common WikiPath terms in three studies: Penter, Nadeu, and Hing
three_common_wikipath <- Reduce(intersect, sig_term.lst[!(names(sig_term.lst) %in% c("Parry(2023b)", "Rigo(2024)"))])
setdiff(three_common_wikipath, four_common_wikipath)
```

```{r}
# Common WikiPath terms in three studies: Penter, Hing, and Rigo
three_common_wikipath <- Reduce(intersect, sig_term.lst[!(names(sig_term.lst) %in% c("Nadeu(2022)", "Parry(2023b)"))])
setdiff(three_common_wikipath, four_common_wikipath)
```

Breifly, the upregulated WikiPath terms were consistent with previous upregulated terms which were related to OXPHOS, nucleotide metabolisms, cell cycle, and cell proliferation.

#### Downregulated terms

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=8}
# Extract adjusted significant downregulated WikiPath terms
sig_term.lst <- purrr::pmap(
  list(wikipath_gsea.lst),
  function(gsea){
    selected_term <- gsea@result[gsea@result$NES < 0, "ID"]
    return(selected_term)
  }
)

# Reorder the list first
sig_term.lst <- sig_term.lst %>%
  purrr::discard(~ length(.x) == 0) %>%   # drop empty ones
  rev()                                   # reverse order

# Convert to UpSetR input
upset_data <- fromList(sig_term.lst)

# Plot UpSetR
upset(
  upset_data,
  sets = colnames(upset_data),      # ensures only these sets
  keep.order = TRUE,
  nintersects = NA
)
```

No common terms are detected. We will identify common terms in two studies.

```{r}
# Common terms between Penter and Nadeu
Reduce(intersect, sig_term.lst[c("Penter(2021)", "Nadeu(2022)")])
```

