---
title: "Hing_2023_mouse_cell_cell_communication"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

We have analysed in human integrated scRNA-seq dataset with several methods such as tumour and microenvironmental cell and cell-cell communication. In this section, we analysed cell-cell interaction in the Hing mouse dataset to compare the result with human integrated dataset.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(harmony)
library(dplyr)
library(tibble)
library(sctransform)
library(SingleR)
library(celldex)
library(cowplot)
library(CellChat)
library(ComplexHeatmap)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save
path_to_merged_obj <- here::here("scRNAseq/Hing_2023_GSE183426/R_objects/Hing_2023_GSE183426_merged_seu.rds")
path_to_cell_annotat_ref <- here::here("scRNAseq/Hing_2023_GSE183426/ImmGenData_ref.rds")
path_to_save_integrated_obj <- here::here("scRNAseq/individual_analysis/cell_cell/R_objects/hing_mouse_integrated_seu.rds")
path_to_save_annotated_obj <- here::here("scRNAseq/individual_analysis/cell_cell/R_objects/hing_mouse_annotated_seu.rds")
path_to_save_filtered_obj <- here::here("scRNAseq/individual_analysis/cell_cell/R_objects/hing_mouse_filtered_seu.rds")
path_to_save_cll_obj <- here::here("scRNAseq/individual_analysis/cell_cell/R_objects/hing_mouse_cll_seu.rds")
path_to_save_rt_obj <- here::here("scRNAseq/individual_analysis/cell_cell/R_objects/hing_mouse_rt_seu.rds") 
path_to_save_cll_cellchat <- here::here("scRNAseq/individual_analysis/cell_cell/R_objects/hing_mouse_cll_cellchat.rds") 
path_to_save_rt_cellchat <- here::here("scRNAseq/individual_analysis/cell_cell/R_objects/hing_mouse_rt_cellchat.rds")
path_to_save_merged_cellchat_list <- here::here("scRNAseq/individual_analysis/cell_cell/R_objects/hing_mouse_merged_cellchat_lst.rds") 
path_to_save_merged_cellchat <- here::here("scRNAseq/individual_analysis/cell_cell/R_objects/hing_mouse_merged_cellchat.rds") 
path_to_save_figure <- here::here("figures")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r, eval=FALSE}
# Load the merge dataset
merged_seurat <- readRDS(path_to_merged_obj)
```

# Data integration

We integrated the mouse datasets with keeping only Eu-hTCL1 and Eu-hPRMT5/hTCL1 models that represent CLL and RT.

```{r, eval=FALSE}
# Set the limit to 15 GB (15 * 1024^2 bytes)
options(future.globals.maxSize = 15 * 1024^3)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Data integration and reprocessing 
DefaultAssay(merged_seurat) <- "RNA"
merged_seurat$ref <- "Hing_2023" # correct Hing_2022 to Hing_2023
integrated_seurat <- subset(merged_seurat, subset = model %in% c("Eµ-hPRMT5/hTCL1", "Eµ-hTCL1"))
integrated_seurat <- JoinLayers(integrated_seurat)
integrated_seurat <- SCTransform(integrated_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
integrated_seurat <- RunPCA(integrated_seurat)
integrated_seurat <- RunHarmony(integrated_seurat, group.by.vars = "case_id", dims = 1:30) 
integrated_seurat <- RunUMAP(integrated_seurat, dims = 1:30, reduction = "harmony")
integrated_seurat <- FindNeighbors(integrated_seurat, dims = 1:30, reduction = "harmony")              
integrated_seurat <- FindClusters(
    integrated_seurat,
    random.seed = 12345,
    verbose = FALSE,
    resolution = seq(0.1, 2, by = 0.1)
  )

# Cell annotation
ref <- readRDS(path_to_cell_annotat_ref)
integrated_seurat <- singleR_cell_annotation(integrated_seurat, ref=ref, annotation = "both") 

# Save the integrated object
saveRDS(integrated_seurat, path_to_save_integrated_obj)
```

# Visualisation after data integration

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=16}
# Plot based on case_id, conditions, and cell types
integrated_seurat <- readRDS(path_to_save_integrated_obj)

group_by <- c("case_id", "model", "singleR.labels_main")
umap_plots <- lapply(group_by, function (group){
  DimPlot(integrated_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 2,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

integrated_umap.plt <- wrap_plots(umap_plots, ncol=3)
integrated_umap.plt
```

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=16}
# Plot based on clusters
group_by <- c("SCT_snn_res.0.1", "SCT_snn_res.0.2", "SCT_snn_res.0.3")
umap_plots <- lapply(group_by, function (group){
  DimPlot(integrated_seurat, 
          reduction = "umap", 
          group.by = group, 
          pt.size = 2,
          repel=TRUE,
          raster=TRUE, 
          label=TRUE) +
    labs(y= "UMAP 2", x = "UMAP 1")
})

integrated_umap.plt <- wrap_plots(umap_plots, ncol=3)
integrated_umap.plt
```

```{r, fig.height=7, fig.width=9, warning=FALSE, message=FALSE}
# Plot based on gene expression
FeaturePlot(integrated_seurat, 
            features = c("Hbb-bt", "Hba-a1", "Cd3d", "Cd8a", "Cd4", "Cd19", "Cd79a", "Cd14", "Ncr1"),
            reduction = "umap") 
```

# Cell type annotation

We manually annotated main cell types based on SingleR annotation and marker feature plots.

```{r, eval=FALSE}
# Define new cluster labels
new.cluster.ids <- c("Cd8+ T cell", "Tumour cell", "NK cell", "Tumour cell", "Cd4+ T cell", "Tumour cell", "Tumour cell", 
                     "Monocyte", "Cd8+ T cell", "Tumour cell", "Tumour cell", "Tumour cell", "Tumour cell", "Tumour cell", "Tumour cell", 
                     "Tumour cell", "Cd8+ T cell", "Tumour cell", "Cd8+ T cell", "Tumour cell", "NK cell", "Tumour cell", 
                     "Erythrocyte", "Tumour cell")

# Map these to the cluster identities (i.e., Idents of the Seurat object)
Idents(integrated_seurat) <- "SCT_snn_res.0.3"
names(new.cluster.ids) <- levels(integrated_seurat)

# Now map cluster IDs to the manual labels
integrated_seurat$level1_annotate <- plyr::mapvalues(
  x = as.character(Idents(integrated_seurat)),
  from = names(new.cluster.ids),
  to = new.cluster.ids
)
```

```{r, fig.height=3, fig.width=18}
# Dot plot for biomarker expression in each cell type
integrated_seurat <- readRDS(path_to_save_annotated_obj)

# 1. Create meta-data to concatenate clusters and cell types
integrated_seurat$cluster_celltype <- paste0("C", integrated_seurat$SCT_snn_res.0.3, " - ", integrated_seurat$level1_annotate)

# 2. Extract a data frame from Seurat dot plot
features = c("Cd3d", "Cd8a", "Cd4", "Ncr1", "Pdcd1", "Foxp3", "Cd19", "Itgam", "Cd177", "Cd14", "Hbb-bt")
dotdata <- DotPlot(
  integrated_seurat,
  features = features,
  group.by = "cluster_celltype"  
)$data

# 3. Preprocess the data frame compatible with ggplot
dotdata <- dotdata %>% 
  tidyr::separate(col = id, into = c("cluster", "celltype"), sep = " - ") %>%
  dplyr::mutate(proportion=pct.exp/100) %>%
  dplyr::rename(
    gene_name = features.plot,
    expression = avg.exp.scaled
  )

# 4. Factor levels
dotdata$cluster <- factor(dotdata$cluster, levels = unique(dotdata$cluster))
dotdata$gene_name <- factor(dotdata$gene_name, levels = rev(features))
celltype_order <- c("Tumour cell", "Cd8+ T cell", "Cd4+ T cell", "NK cell", "Monocyte", "Erythrocyte")
dotdata$celltype <- factor(dotdata$celltype, levels = celltype_order)

# 5. ggplot for the dot plot
biomarker_dot.plt <- ggplot(dotdata, aes(x = cluster, 
                                         y = gene_name, 
                                         color = expression,
                                         size = proportion)) +
  geom_point() +
  scale_size_area() +
  scale_color_viridis_c() +
  facet_wrap(~celltype, scales = "free_x", nrow=1) +
  theme_minimal(base_size = 8) +
  theme(strip.background = ggh4x::element_part_rect(side = "b", colour = "black", fill = "transparent")) +
  theme(axis.text.y = element_text(face = "italic")) +
  labs(x = NULL, y = NULL, size = "Proportion", color = "Expression")

biomarker_dot.plt
```

```{r, eval=FALSE}
# Save the annotated object
saveRDS(integrated_seurat, path_to_save_annotated_obj)
```

We filtered out contaminated RBCs before downstream analysis.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Filtered out RBCs with reprocessing
filtered_seurat <- subset(integrated_seurat, subset = level1_annotate != "Erythrocyte")
filtered_seurat <- SCTransform(filtered_seurat, vst.flavor = "v2", method = "glmGamPoi", verbose = FALSE)
filtered_seurat <- RunPCA(filtered_seurat)
filtered_seurat <- RunUMAP(filtered_seurat, dims = 1:30, reduction = "harmony")
filtered_seurat <- FindNeighbors(filtered_seurat, dims = 1:30, reduction = "harmony")        
filtered_seurat <- FindClusters(
  filtered_seurat,
  random.seed = 12345,
  verbose = FALSE,
  resolution = seq(0.1, 2, by = 0.1)
)

# Save the object
saveRDS(filtered_seurat, path_to_save_filtered_obj)
```

# Prepare the data for CellChat

As CellChat prefers the normalised RNA slot, we changed the default assay from SCT to RNA instead.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set assay default to RNA and normalise the values
DefaultAssay(filtered_seurat) <- "RNA"
Idents(filtered_seurat) <- "level1_annotate"
filtered_seurat <- NormalizeData(filtered_seurat, normalization.method="LogNormalize", scale.factor=10000)
```

After that, we split the object into RT and CLL according the CellChat recommendation.

```{r, eval=FALSE}
# Split the object based on the conditions
rt_seurat <- subset(filtered_seurat, subset = model == "Eµ-hPRMT5/hTCL1") 
cll_seurat <- subset(filtered_seurat, subset = model == "Eµ-hTCL1")

# Save the objects
saveRDS(rt_seurat, path_to_save_rt_obj)
saveRDS(cll_seurat, path_to_save_cll_obj)
```

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Set the ligand-receptor interaction database
future::plan(sequential) # do sequential
CellChatDB <- CellChatDB.mouse
CellChatDB.use <- subsetDB(CellChatDB, search = "Secreted Signaling", key = "annotation") # use Secreted Signaling
```

# Cell-cell communication analysis with CellChat

For cell–cell interaction analysis, CLL and RT samples were analysed separately as individual objects. The results were subsequently integrated to enable joint visualisation and comparative analysis.

## CLL cell-cell communication

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create cellchat object
data.input <- cll_seurat[["RNA"]]$data
labels <- Idents(cll_seurat)
meta <- data.frame(labels = labels, row.names = names(labels)) # create a dataframe of the cell labels
cellchat <- createCellChat(object = cll_seurat, group.by = "ident", assay = "RNA")
cellchat@DB <- CellChatDB.use

# Preprocessing the expression data for cell-cell communication analysis
cellchat <- subsetData(cellchat) 
#future::plan("multisession", workers = 4) # do parallel
cellchat <- identifyOverExpressedGenes(cellchat, do.fast = FALSE)
cellchat <- identifyOverExpressedInteractions(cellchat)

# Compute the communication probability and infer cellular communication network
cellchat <- computeCommunProb(cellchat, type = "triMean")
cellchat <- filterCommunication(cellchat, min.cells = 10) # filter out few cells in certain cell groups
cellchat <- computeCommunProbPathway(cellchat) # Infer a signaling pathway level
cellchat <- aggregateNet(cellchat) # Calculate the aggregated cell-cell communication network
cellchat <- netAnalysis_computeCentrality(cellchat, slot.name = "netP") # Compute the network centrality scores

# Save the cellchat object
saveRDS(cellchat, path_to_save_cll_cellchat)
```

```{r, message=FALSE, warning=FALSE}
# Import the cellchat object
cellchat <- readRDS(path_to_save_cll_cellchat)
```

```{r, message=FALSE, warning=FALSE}
# Plot the number/weight of interactions
groupSize <- as.numeric(table(cellchat@idents))
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize, weight.scale = TRUE, label.edge= FALSE, title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize, weight.scale = TRUE, label.edge= FALSE, title.name = "Interaction weights/strength")
```

```{r}
# Plot the signaling sending to each cell group
mat <- cellchat@net$weight
par(mfrow = c(2, 3), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```

```{r}
# Plot the signaling received by each cell group
mat <- t(cellchat@net$weight)
par(mfrow = c(2, 3), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```

We showed all ligand-receptor interactions found in all cell types for the CLL conditions.

```{r, fig.height=6, fig.width=11, message=FALSE, warning=FALSE}
# Show all the significant interactions (L-R pairs) 
cll_bubble.plt <- netVisual_bubble(cellchat, remove.isolate = FALSE) +
  labs(x = "Source-target cell type", y = "Ligand-receptor interaction") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(size = 10, margin = margin(t = 12)), 
    axis.title.y = element_text(size = 10, angle = 90, margin = margin(r = 20)), 
    plot.margin = unit(c(1, 1, 1, 1), "cm") # Add space around plot so titles aren't cut off
  ) 

cll_bubble.plt
```

```{r, eval=FALSE}
# Save the bubble plot
ggsave(filename=paste0(path_to_save_figure, "/cellchat_mouse_cll_bubble.png"), plot=cll_bubble.plt, dpi=500, height=6, width=11)
```

From the bubble plot, we can see that the tumour cells in the CLL conditions had the autocrine signalling via Mif->(Cd74+Cxcr4) and Mif->(Cd74+Cd44) interactions. The tumour cells also received with the same interaction from NK and T cells. Moreover, the tumour cells were received via Lgals9-based interactions from monocytes.

For the role as the sender, the tumour cells interacted with Cd4+ T cells via Il16->Cd4.

```{r, fig.height=8, fig.width=8}
# Plot the chord diagram
cll_chord.plt <- netVisual_chord_gene(cellchat, lab.cex = 0.5,legend.pos.y = 30) 
```

```{r, eval=FALSE}
# Save the chord plot
png(filename = paste0(path_to_save_figure, "/cellchat_mouse_cll_chord.png"), 
    width = 8, height = 8, units = "in", res = 500)

par(mar = c(1, 1, 1, 1)) 
netVisual_chord_gene(cellchat, lab.cex = 0.5, legend.pos.y = 30)

dev.off()
```

```{r, fig.height=4, fig.width=5, message=FALSE, warning=FALSE}
# Visualize dominant senders (sources) and receivers (targets) in a 2D space
dominant_2d.plt <- netAnalysis_signalingRole_scatter(cellchat)
dominant_2d.plt
```

From the plot in the CLL condition, we can see that NK and Cd8+ T cells contributed the most as the senders while tumour cells provided the highest role as the receivers.

```{r, fig.height=6, fig.width=11, message=FALSE, warning=FALSE}
# Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
out_heatmap.plt <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing")
in_heatmap.plt <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming")
commun_heatmap.plt <- out_heatmap.plt + in_heatmap.plt
commun_heatmap.plt
```

```{r, eval=FALSE}
# Save the heatmap
png(filename = paste0(path_to_save_figure, "/cellchat_mouse_cll_heatmap.png"), width = 11, height = 6, units = "in", res = 500)
draw(commun_heatmap.plt)
dev.off()
```

## RT cell-cell communication

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Create cellchat object
data.input <- rt_seurat[["RNA"]]$data
labels <- Idents(rt_seurat)
meta <- data.frame(labels = labels, row.names = names(labels)) # create a dataframe of the cell labels
cellchat <- createCellChat(object = rt_seurat, group.by = "ident", assay = "RNA")
cellchat@DB <- CellChatDB.use

# Preprocessing the expression data for cell-cell communication analysis
cellchat <- subsetData(cellchat) 
#future::plan("multisession", workers = 4) # do parallel
cellchat <- identifyOverExpressedGenes(cellchat, do.fast = FALSE)
cellchat <- identifyOverExpressedInteractions(cellchat)

# Compute the communication probability and infer cellular communication network
cellchat <- computeCommunProb(cellchat, type = "triMean")
cellchat <- filterCommunication(cellchat, min.cells = 10) # filter out few cells in certain cell groups
cellchat <- computeCommunProbPathway(cellchat) # Infer a signaling pathway level
cellchat <- aggregateNet(cellchat) # Calculate the aggregated cell-cell communication network
cellchat <- netAnalysis_computeCentrality(cellchat, slot.name = "netP") # Compute the network centrality scores

# Save the cellchat object
saveRDS(cellchat, path_to_save_rt_cellchat)
```

```{r, message=FALSE, warning=FALSE}
# Import the cellchat object
cellchat <- readRDS(path_to_save_rt_cellchat)
```

```{r, message=FALSE, warning=FALSE}
# Plot the number/weight of interactions
groupSize <- as.numeric(table(cellchat@idents))
par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize, weight.scale = TRUE, label.edge= FALSE, title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize, weight.scale = TRUE, label.edge= FALSE, title.name = "Interaction weights/strength")
```

```{r}
# Plot the signaling sending to each cell group
mat <- cellchat@net$weight
par(mfrow = c(2, 3), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```

```{r}
# Plot the signaling received by each cell group
mat <- t(cellchat@net$weight)
par(mfrow = c(2, 3), xpd=TRUE)
for (i in 1:nrow(mat)) {
  mat2 <- matrix(0, nrow = nrow(mat), ncol = ncol(mat), dimnames = dimnames(mat))
  mat2[i, ] <- mat[i, ]
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE, edge.weight.max = max(mat), title.name = rownames(mat)[i])
}
```

We showed all ligand-receptor interactions found in all cell types for the RT conditions.

```{r, fig.height=6, fig.width=11, message=FALSE, warning=FALSE}
# Show all the significant interactions (L-R pairs) 
rt_bubble.plt <- netVisual_bubble(cellchat, remove.isolate = FALSE) +
  labs(x = "Source-target cell type", y = "Ligand-receptor interaction") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(size = 10, margin = margin(t = 12)), 
    axis.title.y = element_text(size = 10, angle = 90, margin = margin(r = 20)), 
    plot.margin = unit(c(1, 1, 1, 1), "cm") # Add space around plot so titles aren't cut off
  ) 

rt_bubble.plt
```

```{r, eval=FALSE}
# Save the bubble plot
ggsave(filename=paste0(path_to_save_figure, "/cellchat_mouse_rt_bubble.png"), plot=rt_bubble.plt, dpi=500, height=6, width=11)
```

Compared with CLL, LR interactions in RT were decreased, supporting that RT tumour become more autonomous.

```{r, fig.height=8, fig.width=8}
# Plot the chord diagram
rt_chord.plt <- netVisual_chord_gene(cellchat, lab.cex = 0.5,legend.pos.y = 30) 
```

```{r, eval=FALSE}
# Save the chord plot
png(filename = paste0(path_to_save_figure, "/cellchat_mouse_rt_chord.png"), 
    width = 8, height = 8, units = "in", res = 500)

par(mar = c(1, 1, 1, 1)) 
netVisual_chord_gene(cellchat, lab.cex = 0.5, legend.pos.y = 30)

dev.off()
```

```{r, fig.height=4, fig.width=5, message=FALSE, warning=FALSE}
# Visualize dominant senders (sources) and receivers (targets) in a 2D space
dominant_2d.plt <- netAnalysis_signalingRole_scatter(cellchat)
dominant_2d.plt
```

Compared with the CLL condition, tumour cells in RT become less incoming and outgoing interactions while monocytes and NK cells are dominant in incoming and outgoing interactions, respectively.

```{r, fig.height=6, fig.width=11, message=FALSE, warning=FALSE}
# Signaling role analysis on the aggregated cell-cell communication network from all signaling pathways
out_heatmap.plt <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "outgoing")
in_heatmap.plt <- netAnalysis_signalingRole_heatmap(cellchat, pattern = "incoming")
commun_heatmap.plt <- out_heatmap.plt + in_heatmap.plt
commun_heatmap.plt
```

```{r, eval=FALSE}
# Save the heatmap
png(filename = paste0(path_to_save_figure, "/cellchat_mouse_rt_heatmap.png"), width = 11, height = 6, units = "in", res = 500)
draw(commun_heatmap.plt)
dev.off()
```

## Comparative analysis

After we analysed the cell-cell communication in both CLL and RT conditions with the different cellchat objects, we performed the comparative analysis for the interaction in this section.

```{r, eval=FALSE}
# Load the objects and merge into one object
cll_cellchat <- readRDS(path_to_save_cll_cellchat)
rt_cellchat <- readRDS(path_to_save_rt_cellchat)
cellchat.lst <- list(CLL = cll_cellchat, RT = rt_cellchat)
cellchat <- mergeCellChat(cellchat.lst, add.names = names(cellchat.lst)) 

# Save the files
saveRDS(cellchat.lst, path_to_save_merged_cellchat_list)
saveRDS(cellchat, path_to_save_merged_cellchat)
```

### Number of interactions and interaction strength

```{r}
# Load the merged object and list
cellchat.lst <- readRDS(path_to_save_merged_cellchat_list)
cellchat <- readRDS(path_to_save_merged_cellchat)
cll_cellchat <- cellchat.lst$CLL
rt_cellchat <- cellchat.lst$RT

# Compare the total number of interactions and interaction strength
assigned_colours <- c("#1465AC", "#B31B21")
compare_number.plt <- compareInteractions(cellchat, show.legend = FALSE, group = c(1,2), color.use = assigned_colours)
compare_weight.plt <- compareInteractions(cellchat, show.legend = FALSE, group = c(1,2), measure = "weight", 
                                          color.use = assigned_colours)
compare_number.plt + compare_weight.plt
```

```{r}
# Circle plots for differential number of interactions or interaction strength
par(mfrow = c(1,2), xpd=TRUE)
netVisual_diffInteraction(cellchat, weight.scale = TRUE)
netVisual_diffInteraction(cellchat, weight.scale = TRUE, measure = "weight")
```

For the circle plots representing the number of interactions and interaction strength, the red and blue line indicate the edge for RT and CLL, respectively. In comparison between RT and CLL, the tumour cells had the higher number of interactions and the strength.

```{r, message=FALSE, warning=FALSE}
# Heatmap for differential number of interactions or interaction strength
compare_number_heatmap.plt <- netVisual_heatmap(cellchat)
compare_weigth_heatmap.plt <- netVisual_heatmap(cellchat, measure = "weight")
compare_heatmap.plt <- compare_number_heatmap.plt + compare_weigth_heatmap.plt
compare_heatmap.plt
```

```{r, eval=FALSE}
# Save the figure
png(filename = paste0(path_to_save_figure, "/cellchat_mouse_compare_interaction_heatmap.png"), width = 8, height = 4, units = "in", res = 500)
draw(compare_heatmap.plt)
dev.off()
```

In the heatmaps, colour intensity represents relative changes in both the number of interactions and interaction strength between the RT and CLL conditions. In addition, the top coloured bar plot represents the column-wise sum of the absolute relative values, indicating the overall interaction magnitude for receiver cell types. The coloured bar on the right shows the row-wise summation, representing the overall interaction magnitude for sender cell types.

```{r, message=FALSE, warning=FALSE}
# Circle plots compare the number of interactions 
weight.max <- getMaxWeight(cellchat.lst, attribute = c("idents","count"))
par(mfrow = c(1,2), xpd=TRUE)
for (i in 1:length(cellchat.lst)) {
  netVisual_circle(cellchat.lst[[i]]@net$count, 
                   weight.scale = TRUE, label.edge= FALSE, 
                   edge.weight.max = weight.max[2], edge.width.max = 12, 
                   title.name = paste0("Number of interactions - ", names(cellchat.lst)[i]))
}
```

```{r, message=FALSE, warning=FALSE}
# Circle plots compare the interaction strength 
par(mfrow = c(1,2), xpd=TRUE)
for (i in 1:length(cellchat.lst)) {
  netVisual_circle(cellchat.lst[[i]]@net$weight, 
                   weight.scale = TRUE, label.edge= FALSE, edge.width.max = 12, 
                   title.name = paste0("Interaction strength - ", names(cellchat.lst)[i]))
}
```

For the heatmaps and circle pltos, tumour cells in the RT conditions exhibited lower interaction activity while kept communications to monocytes and NK cells.

```{r, message=FALSE, warning=FALSE, fig.height=3, fig.width=8}
# Identify cell populations with significant changes in sending or receiving signals
num.link <- sapply(cellchat.lst, function(x) {rowSums(x@net$count) + colSums(x@net$count)-diag(x@net$count)})
weight.MinMax <- c(min(num.link), max(num.link)) # control the dot size in the different datasets
gg <- list()
for (i in 1:length(cellchat.lst)) {
  gg[[i]] <- netAnalysis_signalingRole_scatter(cellchat.lst[[i]], title = names(cellchat.lst)[i], weight.MinMax = weight.MinMax) +
    xlim(0, 0.4) +  
    ylim(0, 0.6)    
}
patchwork::wrap_plots(plots = gg)
```

Consistent with the previous plots, tumour cells in RT reduced incoming interactions while monocytes become predominant instead. NK cells remained predominant in outgoing interactions for both conditions.

### Signalling comparison

```{r, message=FALSE, warning=FALSE, fig.height=7, fig.width=13}
# Identify the signaling changes of specific cell populations
assigned_colours <- c("Shared" = "black",           
                      "CLL specific" = "#1465AC",   
                      "RT specific" = "#B31B21")    

tumour_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Tumour cell")
monocyte_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Monocyte")
nk_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "NK cell")
cd4_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Cd4+ T cell")
cd8_signal_diff.plt <- netAnalysis_signalingChanges_scatter(cellchat, idents.use = "Cd8+ T cell")

all_plots <- list(tumour_signal_diff.plt, 
                  monocyte_signal_diff.plt, 
                  nk_signal_diff.plt, 
                  cd4_signal_diff.plt, 
                  cd8_signal_diff.plt)

all_plots_coloured <- lapply(all_plots, function(p) {
  p + 
    scale_color_manual(values = assigned_colours) +
    scale_fill_manual(values = assigned_colours)
})

patchwork::wrap_plots(plots = all_plots_coloured)
```

```{r, message=FALSE, warning=FALSE}
# Compare the overall information flow of each signaling pathway or ligand-receptor pair
assigned_colours <- c("#1465AC", "#B31B21")

compare_inform_stacked.plt <- rankNet(cellchat, 
                                      mode = "comparison", 
                                      measure = "weight", 
                                      sources.use = NULL, 
                                      targets.use = NULL, 
                                      stacked = TRUE, 
                                      do.stat = TRUE,
                                      color.use = assigned_colours) +
  theme(legend.position = "bottom")

compare_inform_bar.plt <- rankNet(cellchat, 
                                  mode = "comparison", 
                                  measure = "weight", 
                                  sources.use = NULL, 
                                  targets.use = NULL, 
                                  stacked = FALSE, 
                                  do.stat = TRUE,
                                  color.use = assigned_colours) +
  theme(legend.position = "bottom")

compare_inform_stacked.plt + compare_inform_bar.plt
```

```{r, fig.height=6, fig.width=12, message=FALSE, warning=FALSE}
# Compare outgoing and incoming signalling patterns 
pathway.union <- union(cll_cellchat@netP$pathways, rt_cellchat@netP$pathways)
cll_outgoing_heatmap <- netAnalysis_signalingRole_heatmap(cll_cellchat, pattern = "outgoing", signaling = pathway.union, title = "CLL", width = 5, height = 6)
rt_outgoing_heatmap <- netAnalysis_signalingRole_heatmap(rt_cellchat, pattern = "outgoing", signaling = pathway.union, title = "RT", width = 5, height = 6)
cll_incoming_heatmap <- netAnalysis_signalingRole_heatmap(cll_cellchat, pattern = "incoming", signaling = pathway.union, title = "CLL", width = 5, height = 6)
rt_incoming_heatmap <- netAnalysis_signalingRole_heatmap(rt_cellchat, pattern = "incoming", signaling = pathway.union, title = "RT", width = 5, height = 6)

draw(cll_outgoing_heatmap + cll_incoming_heatmap + rt_outgoing_heatmap + rt_incoming_heatmap, ht_gap = unit(0.5, "cm"))
```

```{r, eval=FALSE}
# Save the heatmap
png(filename = paste0(path_to_save_figure, "/cellchat_mouse_compare_signal_heatmap.png"), width = 12, height = 6, units = "in", res = 500)
draw(cll_outgoing_heatmap + cll_incoming_heatmap + rt_outgoing_heatmap + rt_incoming_heatmap, ht_gap = unit(0.5, "cm"))
dev.off()
```

```{r}
# Circle plot for the CXCL signalling in the RT conditions
pathways.show <- c("CXCL") 
par(mfrow=c(1,1))
netVisual_aggregate(rt_cellchat, 
                    signaling = pathways.show, 
                    layout = "circle", 
                    signaling.name = pathways.show)
```

The results showed increased CXCL signalling in the RT condition, with NK cells acting as receivers and monocytes as the primary senders.

### Ligand-receptor analysis

```{r, fig.height=7, fig.width=17, message=FALSE, warning=FALSE}
# Identify dysfunctional signaling by comparing the communication probabities 
compare_prob_bubble.plt <- netVisual_bubble(cellchat, 
                                            comparison = c(1, 2), 
                                            remove.isolate = TRUE)

x_axis_labels <- levels(compare_prob_bubble.plt$data$source.target)
axis_colours <- rep("black", length(x_axis_labels))
axis_colours[str_detect(x_axis_labels, "\\(CLL\\)")] <- "#1465AC"
axis_colours[str_detect(x_axis_labels, "\\(RT\\)")] <- "#B31B21"

compare_prob_bubble.plt <- compare_prob_bubble.plt +
  labs(x = "Source-target cell type", y = "Ligand-receptor interaction") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, colour = axis_colours),
    axis.title.x = element_text(size = 10, margin = margin(t = 12)), 
    axis.title.y = element_text(size = 10, angle = 90, margin = margin(r = 20)), 
    plot.margin = unit(c(1, 1, 1, 1), "cm") # Add space around plot so titles aren't cut off
  ) 

compare_prob_bubble.plt
```

```{r, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
# Identify dysfunctional signaling by comparing the communication probabities of tumour->non-tumour and non-tumour->tumour
compare_tumour_prob_bubble.plt <- netVisual_bubble_custom(cellchat, 
                                                          comparison = c(1, 2), 
                                                          focus.cell = "Tumour cell", 
                                                          remove.isolate = TRUE)

compare_tumour_prob_bubble.plt$data <- compare_tumour_prob_bubble.plt$data[!is.na(compare_tumour_prob_bubble.plt$data$prob), ]
compare_tumour_prob_bubble.plt$data$source.target <- droplevels(compare_tumour_prob_bubble.plt$data$source.target)
x_axis_labels <- levels(compare_tumour_prob_bubble.plt$data$source.target)
axis_colours <- rep("black", length(x_axis_labels))
axis_colours[str_detect(x_axis_labels, "\\(CLL\\)")] <- "#1465AC"
axis_colours[str_detect(x_axis_labels, "\\(RT\\)")] <- "#B31B21"

compare_tumour_prob_bubble.plt <- compare_tumour_prob_bubble.plt + 
  labs(x = "Source-target cell type", y = "Ligand-receptor interaction") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, colour = axis_colours),
        axis.title.x = element_text(size = 10, margin = margin(t = 12)), 
        axis.title.y = element_text(size = 10, angle = 90, margin = margin(r = 20)), 
        plot.margin = unit(c(1, 1, 1, 1), "cm")) # Add space around plot so titles aren't cut off

compare_tumour_prob_bubble.plt
```

```{r, eval=FALSE}
# Save the bubble plots
ggsave(filename=paste0(path_to_save_figure, "/cellchat_mouse_compare_all_bubble.png"), 
       plot=compare_prob_bubble.plt, dpi=500, height=7, width=17)
ggsave(filename=paste0(path_to_save_figure, "/cellchat_mouse_compare_tumour_bubble.png"), 
       plot=compare_tumour_prob_bubble.plt, dpi=500, height=5, width=8)
```

The bubble plot showed that, in the RT condition, LR interactions are less than in CLL.

```{r, fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
# Identify dysfunctional signaling from and to tumour
group_tumour <- "Tumour cell"  
all_groups <- levels(cellchat@idents)
group_others <- setdiff(all_groups, group_tumour)

plot_outgoing <- netVisual_bubble(cellchat, 
                                  sources.use = group_tumour, 
                                  targets.use = all_groups, 
                                  comparison = c(1, 2), 
                                  remove.isolate = TRUE) +
                  labs(title = "Signaling from Tumour") + 
                  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

plot_incoming <- netVisual_bubble(cellchat, 
                                  sources.use = group_others, 
                                  targets.use = group_tumour, 
                                  comparison = c(1, 2), 
                                  remove.isolate = TRUE) +
                  labs(title = "Signaling to Tumour") + 
                  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

tumour_bubble.plt <- plot_outgoing / plot_incoming
tumour_bubble.plt
```

```{r, fig.height=9, fig.width=20, message=FALSE, warning=FALSE}
# Identify the up-regulated (increased) and down-regulated (decreased) signaling ligand-receptor pairs
LR_up_prob_bubble.plt <- netVisual_bubble(cellchat, 
                                          max.dataset = 2,
                                          comparison = c(1, 2),
                                          title.name = "Increased signaling in RT", 
                                          angle.x = 45, 
                                          remove.isolate = TRUE)

LR_down_prob_bubble.plt <- netVisual_bubble(cellchat, 
                                            max.dataset = 1,
                                            comparison = c(1, 2),
                                            title.name = "Decreased signaling in RT", 
                                            angle.x = 45, 
                                            remove.isolate = TRUE)

LR_up_prob_bubble.plt / LR_down_prob_bubble.plt
```

```{r, message=FALSE, warning=FALSE}
# Calculate differential expression analysis between RT and CLL for each cell type
pos.dataset = "RT"
features.name = paste0(pos.dataset, ".merged")
cellchat <- identifyOverExpressedGenes(cellchat, 
                                       group.dataset = "datasets", 
                                       pos.dataset = pos.dataset, 
                                       features.name = features.name, 
                                       only.pos = FALSE, 
                                       thresh.pc = 0.1, thresh.fc = 0.05,thresh.p = 0.05, 
                                       do.fast = FALSE,
                                       group.DE.combined = FALSE) 
net <- netMappingDEG(cellchat, features.name = features.name, variable.all = TRUE)
net.up <- subsetCommunication(cellchat, net = net, datasets = "RT",ligand.logFC = 0.05, receptor.logFC = NULL)
net.down <- subsetCommunication(cellchat, net = net, datasets = "CLL",ligand.logFC = -0.05, receptor.logFC = NULL)
```

```{r, eval=FALSE}
# Save the ligand-receptor interaction results as the csv files
write.csv(
  net.up, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/cellchat_mouse_rt_up_lr.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)

write.csv(
  net.down, 
  file = here::here("scRNAseq/meta_analysis/supplm_file/cellchat_mouse_rt_down_lr.csv"), 
  row.names = FALSE,   # Removes the row numbers 
  na = ""              # Makes empty cells blank instead of writing "NA"
)
```

```{r, fig.height=9, fig.width=18, message=FALSE, warning=FALSE}
# Plot the bubble plot 
pairLR.use.up = net.up[, "interaction_name", drop = FALSE]
LR_up_dea_bubble.plt <- netVisual_bubble(cellchat, 
                        pairLR.use = pairLR.use.up, 
                        comparison = c(1,2),
                        angle.x = 45, 
                        remove.isolate = TRUE, 
                        title.name = "Up-regulated signaling in RT")
                        
pairLR.use.down = net.down[, "interaction_name", drop = FALSE]
LR_down_dea_bubble.plt <- netVisual_bubble(cellchat, 
                        pairLR.use = pairLR.use.down, 
                        comparison = c(1,2),
                        angle.x = 45, 
                        remove.isolate = TRUE, 
                        title.name = "Down-regulated signaling in RT")

LR_up_dea_bubble.plt / LR_down_dea_bubble.plt
```

```{r, message=FALSE, warning=FALSE}
# Violin plot ligand-receptor expression in CXCL signalling
cellchat@meta$datasets = factor(cellchat@meta$datasets, levels = c("CLL", "RT")) # set factor level
plotGeneExpression(cellchat, signaling = "CXCL", split.by = "datasets", colors.ggplot = TRUE, type = "violin")
```

# Session Information

```{r}
sessionInfo()
```




