---
title: "Nadeu_2022_reanalysis_qc_object"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

## Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Attach the packages you will need for the analysis.
library(Seurat)
library(ggpubr)
library(tidyverse)
library(DT)
```

```{r}
# Set working path
setwd("/mnt/hp-storage/users/pakorns/scRNAseq/Nadeu2022_NatMed_scRNAseq_data/")
getwd()
```

```{r}
# Define paths
path_to_obj <- here::here("reanalysis/results/R_objects/2.seurat_unfiltered.rds")
path_to_project_metadata <- here::here("metadata/richter_metadata.csv")
path_to_sample_metadata <- here::here("metadata/sample_metadata.csv")
path_to_save <- here::here("reanalysis/results/R_objects/3.seurat_filtered.rds")
```

```{r}
# Functions
source(here::here("bin/utils.R"))
```

## Set parameters

```{r}
# Thresholds
min_lib_size <- 900
min_n_genes <- 250
max_pct_mt <- 22.5
min_cells <- 3
```

## Import the data

```{r}
# Read merged Seurat object
seurat <- readRDS(path_to_obj)
seurat

# Read project and donor metadata
project_metadata_df <- read_csv(path_to_project_metadata)
DT::datatable(project_metadata_df)
sample_metadata_df <- read_csv(path_to_sample_metadata)
DT::datatable(sample_metadata_df)
```

## Add smaple metadata

```{r}
sample_metadata_df$donor_id <- as.character(sample_metadata_df$donor_id)
seurat$donor_id <- seurat$donor_id %>%
  str_remove("ICGC_") %>%
  str_remove("^0")
seurat$subproject <- ifelse(seurat$is_hashed, "BCLLATLAS_10", "BCLLATLAS_29")
seurat$sample_id <- seurat$hash.ID
singlets <- colnames(seurat)[!(seurat$hash.ID %in% c("Negative", "Doublet"))]
seurat <- subset(seurat, cells = singlets)
seurat$sample_id <- seurat$sample_id %>%
  str_remove("(ICGC-012-|ICGC-019|ICGC-365-|ICGC-3299-)") %>%
  str_remove("^-")
seurat$sample_id[seurat$sample_id == "NA"] <- seurat$library_name[seurat$sample_id == "NA"] 
seurat$sample_id <- str_replace(seurat$sample_id, "(-|_)", "/")
seurat$sample_id[seurat$sample_id == "365/07"] <- "365-07"
seurat$sample_id[seurat$sample_id == "19/06"] <- "19-06"
seurat$sample_id[seurat$sample_id == "19/12"] <- "19-12"
sample_metadata_df[sample_metadata_df$sample_id == "19/2768B", ]


test_identifiers <- all(unique(seurat$sample_id) %in% unique(sample_metadata_df$sample_id))
if (test_identifiers) {
  sample_metadata_df <- dplyr::select(sample_metadata_df, - "donor_id")
  new_metadata <- seurat@meta.data %>%
    left_join(sample_metadata_df, by = "sample_id") %>%
    as.data.frame()
  rownames(new_metadata) <- colnames(seurat)
  seurat@meta.data <- new_metadata

} else {
  stop("Identifiers are different!")
}


# Remove hashing-specific columns
removed_vars <- c("HTO_margin", "HTO_classification", "HTO_classification.global")
seurat@meta.data <- seurat@meta.data[, !(colnames(seurat@meta.data) %in% removed_vars)]
```

```{r}
# Find samples with low quality in the other techniques
exclude_fn <- which(
  seurat$Exclude_FN == "Sample excluded due to low quality in the other techniques"
)
unique(seurat$sample_id[exclude_fn])
```

```{r}
# Exclude that smaple from the object
seurat
keep_cells <- colnames(seurat)[seurat$sample_id != "14/2944"]
seurat <- subset(seurat, cells = keep_cells)
seurat
```

## Distribution of QC variables

### Signal-to-noise ratio (cell hashing)

```{r, warning=FALSE}
# Calculate mean SNR
snr_means_df <- seurat@meta.data %>%
  dplyr::filter(is_hashed) %>%
  dplyr::group_by(library_name) %>%
  dplyr::summarise(
    mean_snr = round(mean(hashing_snr), 2),
    sd_snr = round(sd(hashing_snr), 2)
  ) %>%
  dplyr::arrange(desc(mean_snr))
DT::datatable(snr_means_df)


# Plot SNR distribution
breaks <- c(0, 2.5, 5, 7.5)
snr_gg <- seurat@meta.data %>%
  dplyr::filter(is_hashed) %>%
  horizontal_boxplot(
    categorical_var = "library_name",
    continuous_var = "hashing_snr",
    fill = "donor_id",
    ylab = "Signal-to-Noise (hashing)"
  ) +
  scale_y_continuous(limits = c(0, 7.5), breaks = breaks)
snr_gg
```

### Library complexity

```{r}
# Calculate mean library size
seurat_list <- SplitObject(seurat, split.by = "donor_id")
n_features_means_dfs <- purrr::map(seurat_list, function(seurat_obj) {
  df <- seurat_obj@meta.data %>%
    dplyr::group_by(time_point, sample_id) %>%
    dplyr::summarise(
      mean_n_features = round(mean(nFeature_RNA), 2),
      sd_n_features = round(sd(nFeature_RNA), 2)) %>%
    dplyr::arrange(time_point)
  df
})
n_features_means_dfs


# Plot library size distribution
n_features_ggs <- purrr::map(names(seurat_list), function(x) {
  p <- seurat_list[[x]]@meta.data %>%
    ggplot(aes(time_point, nFeature_RNA, fill = subproject)) +
      geom_violin() +
      scale_y_log10() +
      labs(title = x, x = "", y = "Number of Detected Genes") +
      theme_bw() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 13),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 13),
        legend.title = element_blank()
      ) 
  p
})
names(n_features_ggs) <- names(seurat_list)
n_features_ggs
```

## Filter out cells

### Library Size

```{r, fig.wide=TRUE, warning=FALSE}
lib_size_hist <- seurat@meta.data %>%
  plot_histogram_qc(x = "nCount_RNA", x_lab = "Library Size (log10(total UMI))") +
  geom_vline(xintercept = min_lib_size, linetype = "dashed", color = "red")
lib_size_hist1 <- lib_size_hist +
    scale_x_log10()
lib_size_hist2 <- lib_size_hist +
    scale_x_continuous(limits = c(0, 4000)) +
    xlab("Library Size (total UMI)") +
    theme_pubr()
lib_size_hist1 + lib_size_hist2
```

### Number of detected genes

```{r, fig.wide=TRUE, warning=FALSE}
n_genes_hist1 <- seurat@meta.data %>%
  plot_histogram_qc(x = "nFeature_RNA", x_lab = "Number of Detected Genes") +
  geom_vline(xintercept = min_n_genes, linetype = "dashed", color = "red")
n_genes_hist2 <- n_genes_hist1 +
  scale_x_continuous(limits = c(0, 2000)) 
n_genes_hist1 + n_genes_hist2
```

### Fraction of mitochondrial expression

```{r, warning=FALSE}
# Calculate percentage of mitochondrial and ribosomal expression
seurat$pct_mt <- PercentageFeatureSet(seurat, pattern = "^MT-")

# Plot
pct_mt_hist <- seurat@meta.data %>%
  plot_histogram_qc(x = "pct_mt", x_lab = "% Mitochondrial Expression") +
  geom_vline(xintercept = max_pct_mt, linetype = "dashed", color = "red") +
  scale_x_continuous(limits = c(0, 100))
pct_mt_hist
```

### Fraction of ribosomal expression

```{r}
seurat$pct_ribosomal <- PercentageFeatureSet(
  seurat,
  pattern = "^RPS"
)
pct_ribosomal_hist <- seurat@meta.data %>%
  plot_histogram_qc(x = "pct_ribosomal", x_lab = "% Ribosomal Expression")
pct_ribosomal_hist
```

### Joint QC metrics

```{r, warning=FALSE}
# Number of detected genes VS library size
n_genes_vs_lib_size <- FeatureScatter(
  seurat,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA",
  pt.size = 0.15,
  cols = rep("black", length(levels(Idents(seurat))))
)
n_genes_vs_lib_size <- n_genes_vs_lib_size +
  labs(x = "Library Size (total UMI)", y = "Number of Detected Genes") +
  theme(legend.position = "none", plot.title = element_blank())
n_genes_vs_lib_size +
  geom_vline(xintercept = min_lib_size, linetype = "dashed", color = "red") +
  geom_hline(yintercept = min_n_genes, linetype = "dashed", color = "red")

  
# % mitochondrial expression VS library size
pct_mt_vs_lib_size <- FeatureScatter(
  seurat,
  feature1 = "nCount_RNA",
  feature2 = "pct_mt",
  pt.size = 0.15,
  cols = rep("black", length(levels(Idents(seurat))))
)
pct_mt_vs_lib_size <- pct_mt_vs_lib_size +
  scale_x_log10()+
  labs(x = "Library Size (total UMI)", y = "% Mitochondrial Expression") +
  theme(legend.position = "none", plot.title = element_blank())
pct_mt_vs_lib_size +
  geom_vline(xintercept = min_lib_size, linetype = "dashed", color = "red") +
  geom_hline(yintercept = max_pct_mt, linetype = "dashed", color = "red")
```

### Subset empty droplets and lysed cells

```{r}
metadata_before_qc <- seurat@meta.data
is_low_quality <- 
  seurat$nCount_RNA < min_lib_size |
  seurat$nFeature_RNA < min_n_genes |
  seurat$pct_mt > max_pct_mt
table(is_low_quality)
seurat
seurat <- subset(seurat, cells = colnames(seurat)[!is_low_quality])
seurat
metadata_after_qc <- seurat@meta.data
```

### QC summary table

```{r, warning=FALSE, message=FALSE}
qc_before <- metadata_before_qc %>%
  group_by(donor_id, sample_id, time_point) %>% 
  summarise(num_cells_before_qc = n())
qc_after <- metadata_after_qc %>%
  group_by(donor_id, sample_id, time_point) %>%
  summarise(
    num_cells_after_qc = n(),
    average_library_size = mean(nCount_RNA),
    average_num_detected_genes = mean(nFeature_RNA),
    average_mitochondrial_fraction = mean(pct_mt)
  )
qc_table <- left_join(qc_before, qc_after, by = "sample_id")
DT::datatable(qc_table)
```

## Filter out genes

### Set threshold

```{r fig.wide=TRUE, warning=FALSE}
n_cells <- Matrix::rowSums(LayerData(seurat[["RNA"]], layer = "counts") > 0)
gene_qc <- n_cells %>% 
  as.data.frame() %>% 
  ggplot(aes(n_cells)) + 
    geom_histogram(bins = 100, alpha = 0.75) +
    scale_x_log10("Number of cells") +
    theme_bw() 
gene_qc +
  geom_vline(xintercept = min_cells, linetype = "dashed", color = "red")
```

### Plot genes with highest expression

```{r fig.height=10}
top_50_genes <- sort(n_cells, decreasing = TRUE)[1:50]
top_50_genes_df <- data.frame(
  gene = names(top_50_genes),
  n_cells = top_50_genes
)
top_50_genes_gg <- top_50_genes_df %>%
  ggplot(aes(fct_reorder(gene, n_cells), n_cells)) +
    geom_point() +
    labs(x = "", y = "Number of expressing cells") +
    coord_flip()
top_50_genes_gg
```


### Filter

```{r}
kept_genes <- rownames(seurat)[n_cells > min_cells]
table(n_cells > min_cells)
seurat
seurat <- subset(seurat, features = kept_genes)
seurat
```


## Save filtered Seurat object

```{r}
saveRDS(seurat, path_to_save)
```







