---
title: "Nadeu_2022_bulkRNA-seq_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

In this code we will analyse bulk RNA-seq between Richter's transformation (RT) and chronic lymphocytic leukaemia (CLL) patients. The data comes from [Nadeu et al. (2022)](https://www.nature.com/articles/s41591-022-01927-8#data-availability). The original article performed kallisto to preprocess raw files to pseudocount files. We downloaded the original kallisto pseudocount files from this [GitHub](https://github.com/ferrannadeu/RichterTransformation/tree/main).

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Import the relavant packages
library(tximport)
library(ensembldb)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(DESeq2)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save files
path_to_samples <- here::here("RNAseq/Nadeu_2022/kallisto")
path_to_sample_metadata <- here::here("RNAseq/systematic_review/1-Nadeu_2022/sample_metadata.tsv")
path_to_ref_transcriptome <- "/home/pakorns/hc-storage/fastq_files/RNAseq/transcriptome_ref"
path_to_save_txi <- here::here("RNAseq/systematic_review/1-Nadeu_2022/count_matrices/Nadeu_txi.rds")
path_to_save_dea <- here::here("RNAseq/systematic_review/1-Nadeu_2022/expression_data/Nadeu_dea.rds")
path_to_save_hm_gsea <- here::here("RNAseq/systematic_review/1-Nadeu_2022/enrichment_data/Nadeu_hm_gsea.rds")
path_to_save_go_gsea <- here::here("RNAseq/systematic_review/1-Nadeu_2022/enrichment_data/Nadeu_go_gsea.rds")
path_to_save_kegg_gsea <- here::here("RNAseq/systematic_review/1-Nadeu_2022/enrichment_data/Nadeu_kegg_gsea.rds")
path_to_save_reactome_gsea <- here::here("RNAseq/systematic_review/1-Nadeu_2022/enrichment_data/Nadeu_reactome_gsea.rds")
path_to_save_wikipath_gsea <- here::here("RNAseq/systematic_review/1-Nadeu_2022/enrichment_data/Nadeu_wikipath_gsea.rds")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r, message=FALSE, warning=FALSE}
# Load sample metadata
sample_metadata <- read.delim(path_to_sample_metadata, header = TRUE, stringsAsFactors = FALSE) 
DT::datatable(sample_metadata, options = list(scrollX = TRUE))
```

```{r}
# Create a vector collecting kallisto results in each sample
path_to_download <- str_c(path_to_samples, "/", sample_metadata$Sample, "_abundance.tsv")
names(path_to_download) <- sample_metadata$Sample
path_to_download
```

# Convert kallisto results to an expression matrix

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Load the ensembl hg38 gtf file
path_to_hg38_transcriptome <- str_c(path_to_ref_transcriptome, "/Homo_sapiens.GRCh38.115.gtf.gz")
edb <- ensDbFromGtf(path_to_hg38_transcriptome) # Return edb = "./Homo_sapiens.GRCh38.115.sqlite"
```

```{r, message=FALSE, warning=FALSE}
# Make the transcript-to-gene map (tx2gene)
edb <- here::here("RNAseq/systematic_review/1-Nadeu_2022/Homo_sapiens.GRCh38.115.sqlite")
edb_obj <- EnsDb(edb)
tx2gene <- transcripts(edb_obj,
                       columns = c("tx_id","gene_id","gene_name"),
                       return.type="DataFrame")

# convert to a plain data.frame for tximport
tx2gene <- as.data.frame(tx2gene)
head(tx2gene)
```

```{r, message=FALSE, warning=FALSE}
# Load pseudocount files
txi <- tximport(path_to_download, 
                type = "kallisto", 
                tx2gene = tx2gene, 
                ignoreTxVersion = TRUE 
               )

# This gives you txi$counts, txi$abundance (TPM), txi$length
head(txi$counts)
```

```{r, eval=FALSE}
# Save the expression file
saveRDS(txi, path_to_save_txi)
```

# Transcriptomics analysis using DESeq2

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(DESeq2)
library(limma)
```

## Create a DESeq2 object

```{r}
# Create metadata for DESeq2 object
colData <- sample_metadata[, c("Sample", "Diagnosis", "Case")] %>%
  rename(
    samples = "Sample",
    condition = "Diagnosis",
    case = "Case"
  ) %>%
  mutate(
    batch = ifelse(case == "63", 1, 
                   ifelse(case == "365", 2, 
                          ifelse(case == "3299", 3, 
                                 ifelse(case == "19", 4, 
                                        ifelse(case == "12", 5, 6)))))
  )

rownames(colData) <- colData$samples
colData$condition <- factor(colData$condition)
colData$batch <- factor(colData$batch)
colData
```

```{r, message=FALSE, warning=FALSE}
# Create the object
dds <- DESeqDataSetFromTximport(txi, colData, ~ batch + condition)
```

```{r, message=FALSE, warning=FALSE}
# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
  
# Run DESeq2
dds$condition <- relevel(dds$condition, ref = "CLL")
dds <- DESeq(dds)
```

## Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6}
# Variance stabilised transformation
vsdata <- vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
vs_mat_corrected <- removeBatchEffect(vs_mat, batch = colData(dds)$batch)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2               # variance of each PC
pca_var_percent <- pca_var / sum(pca_var) * 100

# Create case name
case <- sub("^0*", "", sub("-.*$", "", rownames(colData(dds))))

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     sample = case)

# Plot
pca.plt <- ggplot(pca_df, aes(PC1, PC2, color=condition, label=sample)) +
  geom_point(size = 3) +
  geom_text_repel(size=3.25, color = "black") +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  theme_bw()

pca.plt
```

After batch correction using `limma()`, we can see that the samples were grouped based on the conditions rather than the case origin.

## Differential gene expression analysis

```{r, message=FALSE, warning=FALSE}
# Generate a table for gene expression
res <- lfcShrink(dds, 
                 contrast = c("condition", "RT", "CLL"), 
                 type = 'normal')
head(res)
```

```{r}
# Add gene names to a column in expression result
dea.df <- res
gene_map <- unique(tx2gene[, c("gene_id", "gene_name")])
dea.df$gene_id <- rownames(dea.df)
dea.df$gene_name <- gene_map$gene_name[match(rownames(dea.df), gene_map$gene_id)]
head(dea.df)
```

```{r, eval=FALSE}
# Save the expression result
saveRDS(dea.df, path_to_save_dea)
```

## Plot heatmap of gene expression

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(pheatmap)
library(RColorBrewer)
```

```{r}
# Data transformation
vsd <- vst(dds, blind=FALSE)       # Variance-stabilizing transformation
#rld <- rlog(dds_filtered, blind=FALSE)      # Regularized log transformation

# Extract transformed expression matrix
expr_mat <- assay(vsd)

# Calculate z-scores
zscore_mat <- t(scale(t(expr_mat)))
```

```{r}
# Extract main and sub- clusters info from column names
RT_sample <- sample_metadata[sample_metadata$Diagnosis == "RT", "Sample"] 
conditions <-  ifelse(colnames(zscore_mat) %in% RT_sample, "RT", "CLL")
patient_id <- sub("^0*", "", sub("-.*$", "", colnames(zscore_mat)))

# Create annotation dataframe
annotation_col <- data.frame(
  Condition = factor(conditions, levels = c("CLL", "RT")),
  Case = factor(patient_id, levels = c("63","365","3299","19","12", "816"))
)
annotation_col <- annotation_col[, c("Case", "Condition")] 
rownames(annotation_col) <- colnames(zscore_mat)

# Order by Condition and then Case 
order_cols <- order(annotation_col$Condition, annotation_col$Case)
zscore_mat_ordered <- zscore_mat[, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]
```

```{r, fig.height=8, fig.width=4}
# Filter for significant DE genes
sig_genes <- dea.df[!is.na(dea.df$padj) & dea.df$padj < 0.05, ]

# Order genes by log2FoldChange (descending: up -> down)
gene_order <- sig_genes$gene_id[order(sig_genes$log2FoldChange, decreasing = TRUE)]

# Subset zscore matrix
zscore_mat_sig <- zscore_mat_ordered[gene_order, ]

# Plot heatmap
pheatmap(
  zscore_mat_sig,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  gaps_col = which(diff(as.numeric(factor(annotation_col_ordered$Condition))) != 0)
)
```

## Volcano plot

```{r, message=FALSE, warning=FALSE}
# Import the package
library(EnhancedVolcano)
```

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=10}
# Plot volcano 
genes <- c("MKI67", "CDK1", "CDK4","CDK5","CDK6","CDKN3","CDK17",
           "WNT3", "WNT5A", "WNT6", "WNT16", "TLR8", "TLR9", "TLR10", 
           "AICDA", "POLH", "MYC", "MYCBP", "CCND2", "CCND3")

dea_plt <- as.data.frame(dea.df)
idx <- match(genes, dea_plt$gene_name)
rownames(dea_plt)[idx] <- dea_plt$gene_name[idx]

volcano.plt <- volcano_plot(dea_plt, 
                            xlim=c(-7,7), 
                            ylim=c(-0.1,21), 
                            genes_of_interest = genes,
                            boxedLabels = TRUE) 
  
volcano.plt
```

# Gene Set Enrichment Analysis (GSEA)

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

```{r}
# Manipulate the expression data
lfc_vector <- dea.df$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- rownames(dea.df)
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

### Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will explore the trends of hallmark terms reported from other single-cell pseudobulk studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as E2F targets, Myc targets, and oxidative phosphorylation (OXPHOS) were upregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

From the plots, terms such as E2F and Myc targets were consistent with the two single-cell studies while the OXPHOS trend was not clear.

### GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will also explore the trends of GO:BP terms reported from other single-cell studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as OXPHOS and mitochondrial translation were upregulated while BCR signalling was downregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

We also performed GSEA with additional enrichment systems such as KEGG pathways, Reactome, and WikiPath.

### KEGG pathway

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

### Reactome

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

### WikiPath

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r, eval=FALSE}
# Save gene expression and GSEA data
saveRDS(HM_gsea, path_to_save_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_go_gsea)
saveRDS(KEGG_gsea, path_to_save_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_wikipath_gsea)
```

## Bar plot summary

```{r}
# Bar plot of GSEA 
gsea_bp <- rbind(HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_OXIDATIVE_PHOSPHORYLATION",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_FATTY_ACID_METABOLISM",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == "REACTOME_DNA_REPAIR",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_MITOCHONDRIAL_TRANSLATION",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_CELL_CYCLE_PHASE_TRANSITION",],
                 WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_GLYCOLYSIS_AND_GLUCONEOGENESIS",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_MTORC1_SIGNALING",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_MYC_TARGETS_V1",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_E2F_TARGETS",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_G2M_CHECKPOINT",],
                 KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_B_CELL_RECEPTOR_SIGNALING_PATHWAY",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_INFLAMMATORY_RESPONSE",]
)

gsea_bp$Description <- gsub("HALLMARK_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("GOBP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("KEGG_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("REACTOME_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("WP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("_V1", "", gsea_bp$Description)
gsea_bp$Description <- gsub("OXIDATIVE_PHOSPHORYLATION", "OXPHOS", gsea_bp$Description)
gsea_bp$Description <- gsub("REACTIVE_OXYGEN_SPECIES_PATHWAY", "ROS_PATHWAY", gsea_bp$Description)
gsea_bp$Description <- gsub("B_CELL_RECEPTOR_SIGNALING_PATHWAY", "BCR_PATHWAY", gsea_bp$Description)
gsea_bp$Description <- gsub("_", " ", gsea_bp$Description)

gsea_bp$Description <- factor(gsea_bp$Description, levels = gsea_bp$Description)
gsea_bp$Direction <- "Up"
gsea_bp$Direction[gsea_bp$NES<0] <- "Down"

gsea_bar.plt <- ggplot(gsea_bp, aes(x=Description, y=NES, fill=Direction)) +
  geom_bar(stat = "identity", color="black", width = 0.75) +  
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_hline(yintercept = 0) + 
  ylab("NES") + xlab(NULL) + 
  scale_fill_manual(values=c("#8882BA", "#EF7F6A")) + theme(legend.position = "none")

gsea_bar.plt
```

## GSEA plot summary

```{r, message=FALSE, warning=FALSE, fig.height=6, fig.width=9}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "OXPHOS", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "E2F targets", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "MYC targets", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_gsea.plt <- gseaplot(KEGG_gsea, 
                          by = "runningScore", 
                          title = "BCR pathway", 
                          geneSetID = "KEGG_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

gsea.plt <- OXPHOS_gsea.plt + E2F_gsea.plt + Myc_gsea.plt + BCR_gsea.plt +
  plot_layout(ncol = 2)

gsea.plt
```

# Session Information

```{r}
sessionInfo()
```





