---
title: "Hing_2023_GSE183425_bulkRNA-seq_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

In this code, we will analyse mouse bulk RNA-seq from [Hing et al. (2023)](https://www.nature.com/articles/s41467-022-35778-1). This study employed a mouse experimental model using Eµ-hTCL1 and Eµ-hPRMT5/hTCL1 mouse model for CLL and RT, respectively. The experiment was performed in two during between 3 and 6 months after conducting the models. After we performed Salmon converting FASTQ files to pseudocount data in each sample, we will convert them to a count matrix before we use the matrix for DESeq2 for further gene expression analysis.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Import the relavant packages
library(tximport)
library(ensembldb)
library(biomaRt)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
library(DESeq2)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save files
path_to_samples <- here::here("RNAseq/Hing_2023_GSE183425")
path_to_sample_metadata <- here::here("RNAseq/systematic_review/4-Hing_2023_GSE183425/sample_metadata.csv")
path_to_ref_transcriptome <- "/home/pakorns/hc-storage/fastq_files/RNAseq/transcriptome_ref"
path_to_save_txi <- here::here("RNAseq/systematic_review/4-Hing_2023_GSE183425/count_matrices/Hing_GSE183425_txi.rds")
path_to_save_dea <- here::here("RNAseq/systematic_review/4-Hing_2023_GSE183425/expression_data/Hing_GSE183425_dea.rds")
path_to_save_hm_gsea <- here::here("RNAseq/systematic_review/4-Hing_2023_GSE183425/enrichment_data/Hing_GSE183425_hm_gsea.rds")
path_to_save_go_gsea <- here::here("RNAseq/systematic_review/4-Hing_2023_GSE183425/enrichment_data/Hing_GSE183425_go_gsea.rds")
path_to_save_kegg_gsea <- here::here("RNAseq/systematic_review/4-Hing_2023_GSE183425/enrichment_data/Hing_GSE183425_kegg_gsea.rds")
path_to_save_reactome_gsea <- here::here("RNAseq/systematic_review/4-Hing_2023_GSE183425/enrichment_data/Hing_GSE183425_reactome_gsea.rds")
path_to_save_wikipath_gsea <- here::here("RNAseq/systematic_review/4-Hing_2023_GSE183425/enrichment_data/Hing_GSE183425_wikipath_gsea.rds")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r, message=FALSE, warning=FALSE}
# Load sample metadata
sample_metadata <- read.csv(path_to_sample_metadata) 
DT::datatable(sample_metadata, options = list(scrollX = TRUE))
```

From the original article, Eµ-hTCL1 mice is a model for CLL and the author used Eµ-hPRMT5 mice for a model for CLL transformed to aggressive lymphoma. Although Eµ-hPRMT5 mice had some features of RT, the RT phenotype is not clear when compared to Eµ-hPRMT5/hTCL1. We then assign Eµ-hTCL1 and Eµ-hPRMT5 mice as CLL group and assign Eµ-hPRMT5/hTCL1 mice as RT group.

```{r}
# Create a vector collecting Salmon quant.sf path in each sample
path_to_download <- str_c(path_to_samples, "/", sample_metadata$GEO_Accession, "/quant.sf")
names(path_to_download) <- sample_metadata$GEO_Accession
path_to_download
```

# Convert salmon pseudocounts to an expression matrix

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Load the ensembl mm39 gtf file
path_to_mm39_transcriptome <- str_c(path_to_ref_transcriptome, "/Mus_musculus.GRCm39.115.gtf.gz")
edb <- ensDbFromGtf(path_to_mm39_transcriptome) # Return edb = "./Mus_musculus.GRCm39.115.sqlite"
```

```{r, message=FALSE, warning=FALSE}
# Make the transcript-to-gene map (tx2gene)
edb <- here::here("RNAseq/systematic_review/4-Hing_2023_GSE183425/Mus_musculus.GRCm39.115.sqlite")
edb_obj <- EnsDb(edb)
tx2gene <- transcripts(edb_obj,
                       columns = c("tx_id","gene_id","gene_name"),
                       return.type="DataFrame")

# convert to a plain data.frame for tximport
tx2gene <- as.data.frame(tx2gene)
head(tx2gene)
```

```{r, message=FALSE, warning=FALSE}
# Load pseudocount files
txi <- tximport(path_to_download, 
                type = "salmon", 
                tx2gene = tx2gene, 
                ignoreTxVersion = TRUE 
               )

# This gives you txi$counts, txi$abundance (TPM), txi$length
head(txi$counts)
```

```{r, eval=FALSE}
# Save the expression file
saveRDS(txi, path_to_save_txi)
```

# Transcriptomics analysis using DESeq2

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(DESeq2)
library(limma)
```

## Create a DESeq2 object

```{r}
# Create metadata for DESeq2 object
colData <- data.frame(sample = sample_metadata$GEO_Accession)
colData$condition <- sample_metadata$cond
colData$transgene <- sample_metadata$transgene
rownames(colData) <- colData$sample
colData$condition <- factor(colData$condition)
colData
```

```{r, message=FALSE, warning=FALSE}
# Create the object
dds <- DESeqDataSetFromTximport(txi, colData, ~ condition)
```

```{r, message=FALSE, warning=FALSE}
# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
  
# Run DESeq2
dds$condition <- relevel(dds$condition, ref = "CLL")
dds <- DESeq(dds)
```

## Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6}
# Variance stabilised transformation
vsdata <- vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
vs_mat_corrected <- removeBatchEffect(vs_mat, batch = colData(dds)$batch)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2               # variance of each PC
pca_var_percent <- pca_var / sum(pca_var) * 100

# Create case name
case <- sub("^0*", "", sub("-.*$", "", rownames(colData(dds))))

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     sample = case)
pca_df <- pca_df %>%
  dplyr::inner_join(colData %>% dplyr::select(sample, transgene), by = "sample")

# Plot
pca.plt <- ggplot(pca_df, aes(PC1, PC2, color=condition, label=sample, shape=transgene)) +
  geom_point(size = 3) +
  geom_text_repel(size=3.25, color = "black") +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  theme_bw()

pca.plt
```

After batch correction using `limma()`, the samples from Eµ-hPRMT5 mice (GSM5557497 to GSM5557500) overlapped with the samples from Eµ-hPRMT5/hTCL1 mice (GSM5557501 to GSM5557503) and the sample from Eµ-hTCL1 mouse (GSM5557504), indicating that Eµ-hPRMT5 mouse model shares phenotype between RT and CLL conditions.

## Differential gene expression analysis (DEA)

```{r, message=FALSE, warning=FALSE}
# Generate a table for gene expression
res <- lfcShrink(dds, 
                 contrast = c("condition", "RT", "CLL"), 
                 type = 'normal')
head(res)
```

## Convert mouse to human ensemble ids

```{r}
# Get orthologs
path_to_ortholog <- str_c(path_to_ref_transcriptome, "/mouse_human_ortholog.txt")
orthologs <- read.csv(path_to_ortholog)
head(orthologs)
```

```{r}
# Use left_join to convert mouse to human ensembl ids
human_res <- as.data.frame(res) %>%
  rownames_to_column("Mouse_Ensembl_ID") %>%
  left_join(orthologs, by = c("Mouse_Ensembl_ID" = "Mouse.gene.stable.ID"))

head(human_res)
```

```{r, eval=FALSE}
# Save the expression result
saveRDS(human_res, path_to_save_dea)
```

## Plot heatmap of gene expression

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(pheatmap)
library(RColorBrewer)
```

```{r}
# Data transformation
vsd <- vst(dds, blind=FALSE)       # Variance-stabilizing transformation
#rld <- rlog(dds_filtered, blind=FALSE)      # Regularized log transformation

# Extract transformed expression matrix
expr_mat <- assay(vsd)

# Calculate z-scores
zscore_mat <- t(scale(t(expr_mat)))
```

```{r}
# Extract main and sub- clusters info from column names
RT_sample <- colData[colData$condition == "RT", "samples"]
conditions <-  ifelse(colnames(zscore_mat) %in% RT_sample, "RT", "CLL")

# Create annotation dataframe
annotation_col <- data.frame(
  Condition = factor(conditions, levels = c("CLL", "RT"))
)
annotation_col <- annotation_col[, "Condition", drop = FALSE] 
rownames(annotation_col) <- colnames(zscore_mat)

# Order by Condition and then Case 
order_cols <- order(annotation_col$Condition)
zscore_mat_ordered <- zscore_mat[, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]
```

```{r, fig.height=4, fig.width=4.5}
# Filter for significant DE genes
sig_genes <- human_res[!is.na(human_res$padj) & human_res$padj < 0.05 & abs(human_res$log2FoldChange) > 0.5, ]

# Order genes by log2FoldChange (descending: up -> down)
gene_order <- sig_genes$Mouse_Ensembl_ID[order(sig_genes$log2FoldChange, decreasing = FALSE)]

# Subset zscore matrix
zscore_mat_sig <- zscore_mat_ordered[gene_order, ]

# Plot heatmap
pheatmap(
  zscore_mat_sig,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = TRUE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  gaps_col = which(diff(as.numeric(factor(annotation_col_ordered$Condition))) != 0)
)
```

# Gene Set Enrichment Analysis (GSEA)

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

```{r}
# Manipulate the expression data
dea <- human_res[!is.na(human_res$Gene.stable.ID), ]
lfc_vector <- dea$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea$Gene.stable.ID
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
lfc_vector <- lfc_vector[!duplicated(names(lfc_vector))]
```

### Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will explore the trends of hallmark terms reported from other single-cell pseudobulk studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as E2F targets, Myc targets, and oxidative phosphorylation (OXPHOS) were upregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

### GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will also explore the trends of GO:BP terms reported from other single-cell studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as OXPHOS and mitochondrial translation were upregulated while BCR signalling was downregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

We also performed GSEA with additional enrichment systems such as KEGG pathways, Reactome, and WikiPath.

### KEGG pathway

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

### Reactome

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

### WikiPath

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

Overall, from the results for both DEA and GSEA we can see that grouping Eµ-hPRMT5 mice into CLL condition is not appropriate as their phenotype overlapped between RT and CLL groups, causing the DEA and GSEA inconsistent with CLL versus RT in previous datasets.

```{r, eval=FALSE}
# Save gene expression and GSEA data
saveRDS(HM_gsea, path_to_save_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_go_gsea)
saveRDS(KEGG_gsea, path_to_save_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_wikipath_gsea)
```

# Session Information

```{r}
sessionInfo()
```





