---
title: "Rompietti_2025_bulkRNA-seq_analysis"
author: "Pakorn Sagulkoo"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
editor: visual
---

In this code, we will analyse mouse bulk RNA-seq from [Rompietti et al. (2025)](https://www.nature.com/articles/s41375-025-02557-y). This study employed a mouse experimental model in which a CLL- and RT-like condition was induced using Eμ-*TCL1 Bcor^-/-^* mice model. We will analyse the gene expression from the count matrices from original article.

# Import the relevant packages and set working path directory

```{r, setup, message=FALSE, warning=FALSE}
# Import the relavant packages
library(ensembldb)
library(biomaRt)
library(dplyr)
library(tibble)
library(stringr)
library(ggplot2)
library(ggrepel)
```

```{r}
# Check root path
here::here()
```

# Define paths

```{r}
# Define paths to load and save files
path_to_samples <- here::here("RNAseq/Rompietti_2025")
path_to_ref_transcriptome <- "/home/pakorns/hc-storage/fastq_files/RNAseq/transcriptome_ref"
path_to_save_dea <- here::here("RNAseq/systematic_review/3-Rompietti_2025/expression_data/Rompietti_dea.rds")
path_to_save_hm_gsea <- here::here("RNAseq/systematic_review/3-Rompietti_2025/enrichment_data/Rompietti_hm_gsea.rds")
path_to_save_go_gsea <- here::here("RNAseq/systematic_review/3-Rompietti_2025/enrichment_data/Rompietti_go_gsea.rds")
path_to_save_kegg_gsea <- here::here("RNAseq/systematic_review/3-Rompietti_2025/enrichment_data/Rompietti_kegg_gsea.rds")
path_to_save_reactome_gsea <- here::here("RNAseq/systematic_review/3-Rompietti_2025/enrichment_data/Rompietti_reactome_gsea.rds")
path_to_save_wikipath_gsea <- here::here("RNAseq/systematic_review/3-Rompietti_2025/enrichment_data/Rompietti_wikipath_gsea.rds")
```

# Load data and functions

```{r, message=FALSE, warning=FALSE}
# Source functions
source(here::here("bin/utils.R"))
```

```{r, message=FALSE, warning=FALSE}
# List the files and paths
sample_list <- c("TCL1", "TCL2", "TCL3", "TCL1BCOR1", "TCL1BCOR2", "TCL1BCOR3")
path_to_download <- str_c(path_to_samples, "/", sample_list, ".counts.txt")
names(path_to_download) <- sample_list
path_to_download
```

# Construct an expression matrix

```{r}
# Read some table to explore their structures
examp_expr.df <- read.table(path_to_download[[1]], header = TRUE)
head(examp_expr.df)
```

We can see that the transcript ids were alreadey converted to gene id and the count values is the last column (naming `TCL1`). We will extract Geneid and TCL1 column and do the same thing as other datasets and then merge them all.

```{r, message=FALSE, warning=FALSE}
# Create a list to collect expression dataframes
expr_df.lst <- purrr::map2(path_to_download, sample_list, function(path, sample){
  df <- read.table(path, header = TRUE)
  df <- df %>%
    dplyr::select(Geneid, sample)
  return(df)
})
names(expr_df.lst) <- sample_list
```

```{r}
# Join the dataframe
expr.df <- purrr::reduce(expr_df.lst, full_join, by = "Geneid")
head(expr.df)
```

```{r}
# Construct an expression matrix
expr_mat <- expr.df %>%
  column_to_rownames("Geneid") %>%
  as.matrix()
head(expr_mat)
```

# Transcriptomics analysis using DESeq2

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(DESeq2)
library(limma)
```

## Create a DESeq2 object

```{r}
# Create metadata for DESeq2 object
colData <- data.frame(samples = sample_list)
colData$condition <- ifelse(grepl("BCOR", colData$samples), "RT", "CLL")
rownames(colData) <- colData$samples
colData$condition <- factor(colData$condition)
colData
```

```{r, message=FALSE, warning=FALSE}
# Create the object
dds <- DESeqDataSetFromMatrix(expr_mat, colData, ~ condition)
```

```{r, message=FALSE, warning=FALSE}
# Keep genes with at least 10 counts across all samples
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
  
# Run DESeq2
dds$condition <- relevel(dds$condition, ref = "CLL")
dds <- DESeq(dds)
```

## Plot the PCA based on the expression data

```{r, message=FALSE, warning=FALSE, fig.height=4.5, fig.width=6}
# Variance stabilised transformation
vsdata <- vst(dds, blind = FALSE)

# Extract matrix
vs_mat <- assay(vsdata)

# Adjust for batch (assume batch info is in colData(dds)$batch)
vs_mat_corrected <- removeBatchEffect(vs_mat, batch = colData(dds)$batch)

# PCA on corrected data
pca <- prcomp(t(vs_mat_corrected))

# Calculate % variance explained
pca_var <- pca$sdev^2               # variance of each PC
pca_var_percent <- pca_var / sum(pca_var) * 100

# Create case name
case <- sub("^0*", "", sub("-.*$", "", rownames(colData(dds))))

# Make a dataframe for ggplot
pca_df <- data.frame(pca$x,
                     condition = colData(dds)$condition,
                     sample = case)

# Plot
pca.plt <- ggplot(pca_df, aes(PC1, PC2, color=condition, label=sample)) +
  geom_point(size = 3) +
  geom_text_repel(size=3.25, color = "black") +
  xlab(paste0("PC1 (", round(pca_var_percent[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(pca_var_percent[2], 1), "%)")) +
  theme_bw()

pca.plt
```

After batch correction using `limma()`, we can see that the samples were grouped based on the conditions.

## Differential gene expression analysis

```{r, message=FALSE, warning=FALSE}
# Generate a table for gene expression
res <- lfcShrink(dds, 
                 contrast = c("condition", "RT", "CLL"), 
                 type = 'normal')
head(res)
```

## Convert mouse to human ensemble ids

```{r}
# Get orthologs
path_to_ortholog <- str_c(path_to_ref_transcriptome, "/mouse_human_ortholog.txt")
orthologs <- read.csv(path_to_ortholog)
head(orthologs)
```

```{r}
# Use left_join to convert mouse to human ensembl ids
human_res <- as.data.frame(res) %>%
  rownames_to_column("Mouse_Ensembl_ID") %>%
  left_join(orthologs, by = c("Mouse_Ensembl_ID" = "Mouse.gene.stable.ID"))

head(human_res)
```

```{r, eval=FALSE}
# Save the expression result
saveRDS(human_res, path_to_save_dea)
```

## Plot heatmap of gene expression

```{r, message=FALSE, warning=FALSE}
# Import the relevant package
library(pheatmap)
library(RColorBrewer)
```

```{r}
# Data transformation
vsd <- vst(dds, blind=FALSE)       # Variance-stabilizing transformation
#rld <- rlog(dds_filtered, blind=FALSE)      # Regularized log transformation

# Extract transformed expression matrix
expr_mat <- assay(vsd)

# Calculate z-scores
zscore_mat <- t(scale(t(expr_mat)))
```

```{r}
# Extract main and sub- clusters info from column names
RT_sample <- sample_list[grepl("BCOR", sample_list)]
conditions <-  ifelse(colnames(zscore_mat) %in% RT_sample, "RT", "CLL")

# Create annotation dataframe
annotation_col <- data.frame(
  Condition = factor(conditions, levels = c("CLL", "RT"))
)
annotation_col <- annotation_col[, "Condition", drop = FALSE] 
rownames(annotation_col) <- colnames(zscore_mat)

# Order by Condition and then Case 
order_cols <- order(annotation_col$Condition)
zscore_mat_ordered <- zscore_mat[, order_cols]
annotation_col_ordered <- annotation_col[order_cols, , drop = FALSE]
```

```{r}
# List genes of interest
gene_name.lst <- c("Ccnd2", "Ccnb1", "Ccnd3", "Cdk1", "Cdk6", "Ccl22", "Ccl17", "Il6ra", "Cx3cl1", "Tgfbr1", 
                   "Cdkn1a", "Cdkn2b", "H2-Q2", "Lair1") # ENSMUSG00000091705 is H2-Q2
gene_id.lst <- c("ENSMUSG00000000184", "ENSMUSG00000041431", "ENSMUSG00000034165", "ENSMUSG00000019942", "ENSMUSG00000040274", 
                 "ENSMUSG00000031779", "ENSMUSG00000031780", "ENSMUSG00000027947", "ENSMUSG00000031778", "ENSMUSG00000007613", 
                 "ENSMUSG00000023067", "ENSMUSG00000073802", "ENSMUSG00000091705", "ENSMUSG00000055541")
names(gene_id.lst) <- gene_name.lst
gene_id.lst
```

```{r, fig.height=3.5, fig.width=3}
# Subset zscore matrix
zscore_mat_interested <- zscore_mat_ordered[gene_id.lst, ]
rownames(zscore_mat_interested) <- names(gene_id.lst)

# Plot heatmap
pheatmap(
  zscore_mat_interested,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = TRUE,
  color = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(100),
  annotation_col = annotation_col_ordered,
  fontsize = 8,
  fontsize_row = 7,
  fontsize_col = 7,
  gaps_col = which(diff(as.numeric(factor(annotation_col_ordered$Condition))) != 0)
)
```

# Gene Set Enrichment Analysis (GSEA)

```{r, message=FALSE, warning=FALSE}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

```{r}
# Manipulate the expression data
dea <- human_res[!is.na(human_res$Gene.stable.ID), ]
lfc_vector <- dea$log2FoldChange # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- dea$Gene.stable.ID
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
lfc_vector <- lfc_vector[!duplicated(names(lfc_vector))]
```

## Hallmark

```{r, message=FALSE, warning=FALSE}
# Hallmark
HM_gsea <- HM_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
HM_GSEA.result <- HM_gsea@result
HM_GSEA_selected.result <- HM_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(HM_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will explore the trends of hallmark terms reported from other single-cell pseudobulk studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as E2F targets, Myc targets, and oxidative phosphorylation (OXPHOS) were upregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for HALLMARK_E2F_TARGETS, HALLMARK_MYC_TARGETS_V1, and HALLMARK_OXIDATIVE_PHOSPHORYLATION

E2F_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_E2F_TARGETS", 
                          geneSetID = "HALLMARK_E2F_TARGETS",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

Myc_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_gsea.plt <- gseaplot(HM_gsea, 
                          by = "runningScore", 
                          title = "HALLMARK_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "HALLMARK_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

E2F_gsea.plt + Myc_gsea.plt + OXPHOS_gsea.plt
```

## GO:BP

```{r, message=FALSE, warning=FALSE}
# GO:BP
GOBP_gsea <- GO_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the GSEA result
GOBP_GSEA.result <- GOBP_gsea@result
GOBP_GSEA_selected.result <- GOBP_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(GOBP_GSEA_selected.result, options = list(scrollX = TRUE))
```

We will also explore the trends of GO:BP terms reported from other single-cell studies such as Nadeu et al. (2022) and Hing et al (2023). From those studies, the terms such as OXPHOS and mitochondrial translation were upregulated while BCR signalling was downregulated.

```{r, message=FALSE, warning=FALSE, fig.height=4, fig.width=18}
# GSEA plot for GOBP_OXIDATIVE_PHOSPHORYLATION, GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY, and GOBP_MITOCHONDRIAL_TRANSLATION

OXPHOS_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_OXIDATIVE_PHOSPHORYLATION", 
                          geneSetID = "GOBP_OXIDATIVE_PHOSPHORYLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

BCR_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY", 
                          geneSetID = "GOBP_B_CELL_RECEPTOR_SIGNALING_PATHWAY",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

mitochondrial_GO_gsea.plt <- gseaplot(GOBP_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MITOCHONDRIAL_TRANSLATION", 
                          geneSetID = "GOBP_MITOCHONDRIAL_TRANSLATION",
                          base_size = 10) +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))

OXPHOS_GO_gsea.plt + BCR_GO_gsea.plt + mitochondrial_GO_gsea.plt
```

We also performed GSEA with additional enrichment systems such as KEGG pathways, Reactome, and WikiPath.

## KEGG pathway

```{r, message=FALSE, warning=FALSE}
# KEGG pathway
KEGG_gsea <- KEGG_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
KEGG_GSEA.result <- KEGG_gsea@result
KEGG_GSEA_selected.result <- KEGG_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(KEGG_GSEA_selected.result, options = list(scrollX = TRUE))
```

## Reactome

```{r, message=FALSE, warning=FALSE}
# REACTOME
REACTOME_gsea <- REACTOME_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
REACTOME_GSEA.result <- REACTOME_gsea@result
REACTOME_GSEA_selected.result <- REACTOME_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(REACTOME_GSEA_selected.result, options = list(scrollX = TRUE))
```

## WikiPath

```{r, message=FALSE, warning=FALSE}
# WikiPath
WIKIPATH_gsea <- WIKIPATH_GSEA(lfc_vector, gene_ident = "ensembl_gene")

# Collect the HM GSEA result
WIKIPATH_GSEA.result <- WIKIPATH_gsea@result
WIKIPATH_GSEA_selected.result <- WIKIPATH_GSEA.result %>%
  as_tibble() %>%
  dplyr::select(c(Description, NES, pvalue, p.adjust))

DT::datatable(WIKIPATH_GSEA_selected.result, options = list(scrollX = TRUE))
```

```{r}
# Bar plot of GSEA 
gsea_bp <- rbind(HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_OXIDATIVE_PHOSPHORYLATION",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == 
                                                 "REACTOME_AEROBIC_RESPIRATION_AND_RESPIRATORY_ELECTRON_TRANSPORT",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_MITOCHONDRIAL_TRANSLATION",],
                 WIKIPATH_GSEA_selected.result[WIKIPATH_GSEA_selected.result$Description == "WP_GLYCOLYSIS_AND_GLUCONEOGENESIS",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_MTORC1_SIGNALING",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_E2F_TARGETS",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_G2M_CHECKPOINT",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_MYC_TARGETS_V1",],
                 HM_GSEA_selected.result[HM_GSEA_selected.result$Description == "HALLMARK_INFLAMMATORY_RESPONSE",],
                 GOBP_GSEA_selected.result[GOBP_GSEA_selected.result$Description == "GOBP_ADAPTIVE_IMMUNE_RESPONSE",],
                 KEGG_GSEA_selected.result[KEGG_GSEA_selected.result$Description == "KEGG_TOLL_LIKE_RECEPTOR_SIGNALING_PATHWAY",],
                 REACTOME_GSEA_selected.result[REACTOME_GSEA_selected.result$Description == 
                                                 "REACTOME_SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR",]
                 )

gsea_bp$Description <- gsub("HALLMARK_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("GOBP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("KEGG_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("REACTOME_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("WP_", "", gsea_bp$Description)
gsea_bp$Description <- gsub("_V1", "", gsea_bp$Description)
gsea_bp$Description <- gsub("OXIDATIVE_PHOSPHORYLATION", "OXPHOS", gsea_bp$Description)
gsea_bp$Description <- gsub("AEROBIC_RESPIRATION_AND_RESPIRATORY_ELECTRON_TRANSPORT", "RESPIRATORY_ELECTRON_TRANSPORT", gsea_bp$Description)
gsea_bp$Description <- gsub("SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR", "BCR_PATHWAY", gsea_bp$Description)
gsea_bp$Description <- gsub("_V1", "", gsea_bp$Description)
gsea_bp$Description <- gsub("_", " ", gsea_bp$Description)

gsea_bp$Description <- factor(gsea_bp$Description, levels = gsea_bp$Description)
gsea_bp$Direction <- "Up"
gsea_bp$Direction[gsea_bp$NES<0] <- "Down"

gsea_bar.plt <- ggplot(gsea_bp, aes(x=Description, y=NES, fill=Direction)) +
  geom_bar(stat = "identity", color="black", width = 0.75) +  
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  geom_hline(yintercept = 0) + 
  ylab("NES") + xlab(NULL) + 
  scale_fill_manual(values=c("#8882BA", "#EF7F6A")) + theme(legend.position = "none")

gsea_bar.plt
```

```{r, eval=FALSE}
# Save gene expression and GSEA data
saveRDS(HM_gsea, path_to_save_hm_gsea)
saveRDS(GOBP_gsea, path_to_save_go_gsea)
saveRDS(KEGG_gsea, path_to_save_kegg_gsea)
saveRDS(REACTOME_gsea, path_to_save_reactome_gsea)
saveRDS(WIKIPATH_gsea, path_to_save_wikipath_gsea)
```

# Session Information

```{r}
sessionInfo()
```





