---
title: "GSE201704_reanalysis_with_protein"
author: "Pakorn Sagulkoo"
format:
  html:
    embed-resources: true
editor: visual
---

## Import relevant packages

```{r}
library(Seurat)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(SingleR)
library(celldex)
```

## Set a working path directory

```{r}
setwd("/mnt/hp-storage/users/pakorns/scRNAseq/GSE201704/GSM6069857/with_protein/")
```

## Import the object

```{r}
GSM6069857_merged.seu <- readRDS("GSM6069857_merged.rds")
```

## Downstream analysis

```{r}
# Downstream analysis pipeline
GSM6069857_merged.seu <- JoinLayers(GSM6069857_merged.seu)
#GSM6069857_merged.seu <- NormalizeData(GSM6069857_merged.seu)
#GSM6069857_merged.seu <- FindVariableFeatures(object = GSM6069857_merged.seu, nfeatures = 1000)
GSM6069857_merged.seu <- ScaleData(object = GSM6069857_merged.seu)
GSM6069857_merged.seu <- RunPCA(object = GSM6069857_merged.seu)
```

## Clustering and running t-SNE and UMAP

```{r}
# Graph-based clustering 
GSM6069857_merged.seu <- FindNeighbors(GSM6069857_merged.seu, dims = 1:20)
GSM6069857_merged.seu <- FindClusters(GSM6069857_merged.seu, resolution = 0.35)
```

```{r}
# Running UMAP and tSNE
GSM6069857_merged.seu <- RunUMAP(GSM6069857_merged.seu, dims = 1:20)
GSM6069857_merged.seu <- RunTSNE(GSM6069857_merged.seu, dims = 1:20)
```

## Data visualisation
### UMAP and tSNE based on clustering

```{r fig.height=4.5, fig.width=10}
# Plot UMAP based on clusters 
UMAP_merged_cluster.plt <- DimPlot(GSM6069857_merged.seu, reduction = "umap") + labs(y= "UMAP 2", x = "UMAP 1", title = "Cell clustering")

# Plot tSNE based on clusters 
tSNE_merged_cluster.plt <- DimPlot(GSM6069857_merged.seu, reduction = "tsne") + labs(y= "tSNE 2", x = "tSNE 1", title = "Cell clustering")

UMAP_merged_cluster.plt + tSNE_merged_cluster.plt
```

### UMAP and tSNE based on gene expression

```{r fig.height=7, fig.width=9}
# Plot UMAP based on gene expression
FeaturePlot(GSM6069857_merged.seu, 
            features = c("CD8A", "IL7R", "CD4", "CD19", "CD79A", "CD79B", "CD14", "GNLY", "NKG7"),
            reduction = "umap") 
```

```{r fig.height=7, fig.width=9}
# Plot UMAP based on gene expression
FeaturePlot(GSM6069857_merged.seu, 
            features = c("CD8A", "IL7R", "CD4", "CD19", "CD79A", "CD79B", "CD14", "GNLY", "NKG7"),
            reduction = "tsne")
```

### UMAP and tSNE based on surface protein expression

```{r}
# Change default assay to antibody capture and normalisation
DefaultAssay(GSM6069857_merged.seu) <- "ADT"
```


```{r fig.height=7, fig.width=6}
# Plot UMAP based on surface protein expression
FeaturePlot(GSM6069857_merged.seu, 
            features = c("CD3", "CD4.1", "CD8", "CD19.1", "CD11c", "CD56"),
            reduction = "umap") 
```

```{r fig.height=7, fig.width=6}
# Plot tSNE based on surface protein expression
DefaultAssay(GSM6069857_merged.seu) <- "ADT"
FeaturePlot(GSM6069857_merged.seu, 
            features = c("CD3", "CD4.1", "CD8", "CD19.1", "CD11c", "CD56"),
            reduction = "tsne") 
```

### UMAP and tSNE based on HTO

```{r fig.height=4.5, fig.width=11}
# Plot UMAP based on HTO 
UMAP_merged_HTO.plt <- DimPlot(GSM6069857_merged.seu, reduction = "umap", group.by = 'HTO') + labs(y= "UMAP 2", x = "UMAP 1")

# Plot tSNE based on HTO 
tSNE_merged_HTO.plt <- DimPlot(GSM6069857_merged.seu, reduction = "tsne", group.by = 'HTO') + labs(y= "tSNE 2", x = "tSNE 1")

UMAP_merged_HTO.plt + tSNE_merged_HTO.plt
```

## Cell annotation
### Annotation using SingleR

```{r}
DefaultAssay(GSM6069857_merged.seu) <- "RNA"
ref <- celldex::DatabaseImmuneCellExpressionData()
seurat_counts <- GetAssayData(GSM6069857_merged.seu, layer = 'data')

pred1 <- SingleR(test = seurat_counts,
                ref = ref,
                labels = ref$label.main)

GSM6069857_merged.seu$singleR.labels_main <- pred1$labels[match(rownames(GSM6069857_merged.seu@meta.data), rownames(pred1))]
```

```{r fig.height=4.5, fig.width=11}
# Plot UMAP based on SingleR 
UMAP_merged_SingleR.plt <- DimPlot(GSM6069857_merged.seu, reduction = "umap", group.by = "singleR.labels_main") +
  labs(y= "UMAP 2", x = "UMAP 1")

# Plot tSNE based on SingleR 
tSNE_merged_SingleR.plt <- DimPlot(GSM6069857_merged.seu, reduction = "tsne", group.by = "singleR.labels_main") +
  labs(y= "tSNE 2", x = "tSNE 1")

UMAP_merged_SingleR.plt + tSNE_merged_SingleR.plt
```

### Annotated based on manually clustering and biomarker identification

```{r}
# Define new cluster labels
new.cluster.ids <- c("T cell", "NK cell", "Myeloid cell", "B cell", "NK cell", "Myeloid cell", "T cell", "Myeloid cell", 
                     "B cell", "Myeloid cell", "T cell", "B cell", "Myeloid cell", "Myeloid cell", "NK cell")

# Map these to the cluster identities (i.e., Idents of the Seurat object)
names(new.cluster.ids) <- levels(GSM6069857_merged.seu)

# Now map cluster IDs to the manual labels
GSM6069857_merged.seu$manually_annotated <- plyr::mapvalues(
  x = as.character(Idents(GSM6069857_merged.seu)),
  from = names(new.cluster.ids),
  to = new.cluster.ids
)
```

```{r fig.height=4.5, fig.width=11}
# Plot UMAP based on manually annotation 
UMAP_merged_annotated.plt <- DimPlot(GSM6069857_merged.seu, reduction = "umap", group.by = "manually_annotated", label = TRUE) +
  labs(y= "UMAP 2", x = "UMAP 1")

# Plot tSNE based on manually annotation 
tSNE_merged_annotated.plt <- DimPlot(GSM6069857_merged.seu, reduction = "tsne", group.by = "manually_annotated", label = TRUE) +
  labs(y= "tSNE 2", x = "tSNE 1")

UMAP_merged_annotated.plt + tSNE_merged_annotated.plt
```

## Gene expression and functional enrichment analysis between pre- and post CAR-T treatment

```{r}
# Import the related libraries in functional enrichment analysis
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

```{r}
# Hallmark set
hs_hm_sets <- read.csv("/mnt/hp-storage/users/pakorns/scRNAseq/GSE201704/GSM6069857/hs_hm_sets.csv")

# GO:BP set
hs_go_sets <- read.csv("/mnt/hp-storage/users/pakorns/scRNAseq/GSE201704/GSM6069857/hs_go_sets.csv")
```

### T cells

#### GSEA with Hallmarks

```{r}
# Extract only T cells
Tcell.seu <- subset(GSM6069857_merged.seu, subset = manually_annotated == 'T cell')
```

```{r}
# Find markers between pre- and post-treatment  
Idents(Tcell.seu) <- "HTO" # change the default identity
Tcell_pre_post.ge <- FindMarkers(Tcell.seu, ident.1 = 'CMO306', ident.2 = 'CMO305')
```

```{r}
# Manipulate the expression data
Tcell.ge <- cbind(gene = rownames(Tcell_pre_post.ge), Tcell_pre_post.ge)
lfc_vector <- Tcell.ge$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- Tcell.ge$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r}
# Show some gene expression results
head(Tcell.ge)
```

```{r}
# Run GSEA (Hallmarks)
set.seed(2020)
HM_gse <- GSEA(
  geneList = lfc_vector, # Ordered ranked gene list
  minGSSize = 25, # Minimum gene set size
  maxGSSize = 500, # Maximum gene set set
  pvalueCutoff = 0.05, # p-value cutoff
  eps = 0, # Boundary for calculating the p value
  seed = TRUE, # Set seed to make results reproducible
  pAdjustMethod = "BH", # Benjamini-Hochberg correction
  TERM2GENE = dplyr::select(
    hs_hm_sets,
    gs_name,
    gene_symbol
  )
)
```

```{r}
# Collect the GSEA result
Tcell_GSEA.result <- HM_gse@result
Tcell_GSEA.result
```

```{r fig.height=4.5, fig.width=18}
# GSEA plot for HALLMARK_IL2_STAT5_SIGNALING, HALLMARK_MYC_TARGETS_V1, and HALLMARK_INFLAMMATORY_RESPONSE

IL2_STAT5_gsea.plot <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_IL2_STAT5_SIGNALING", 
                          geneSetID = "HALLMARK_IL2_STAT5_SIGNALING",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

Myc_gsea.plot <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_MYC_TARGETS_V1", 
                          geneSetID = "HALLMARK_MYC_TARGETS_V1",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

inflam_response_gsea.plot <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_INFLAMMATORY_RESPONSE", 
                          geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

IL2_STAT5_gsea.plot + Myc_gsea.plot + inflam_response_gsea.plot
```

```{r}
# Measure the expression trends compared to the original results: HALLMARK_IL2_STAT5_SIGNALING
IL2_STAT5_genes <- hs_hm_sets[hs_hm_sets$gs_name == "HALLMARK_IL2_STAT5_SIGNALING", "gene_symbol"]
IL2_STAT5_genes_expression <- Tcell.ge[Tcell.ge$gene %in% IL2_STAT5_genes, ]
head(IL2_STAT5_genes_expression)
```

```{r}
# Measure the expression trends compared to the original results: HALLMARK_INFLAMMATORY_RESPONSE
inflam_response_genes <- hs_hm_sets[hs_hm_sets$gs_name == "HALLMARK_INFLAMMATORY_RESPONSE", "gene_symbol"]
inflam_response_genes_expression <- Tcell.ge[Tcell.ge$gene %in% inflam_response_genes, ]

# Measure the expression trends compared to the original results: HALLMARK_MYC_TARGETS_V1
Myc_genes <- hs_hm_sets[hs_hm_sets$gs_name == "HALLMARK_MYC_TARGETS_V1", "gene_symbol"]
Myc_genes_expression <- Tcell.ge[Tcell.ge$gene %in% Myc_genes, ]
```

```{r}
# Box plot for IL2_STAT5_genes_expression, inflam_response_genes_expression, and Myc_genes_expression
IL2_STAT5_box.plot <- ggplot(IL2_STAT5_genes_expression, aes(y = avg_log2FC)) +
  geom_boxplot() + 
  labs(title = "IL2_STAT5")

inflam_response_box.plot <- ggplot(inflam_response_genes_expression, aes(y = avg_log2FC)) +
  geom_boxplot() + 
  labs(title = "Inflam_Response")

Myc_box.plot <- ggplot(Myc_genes_expression, aes(y = avg_log2FC)) +
  geom_boxplot() + 
  labs(title = "Myc")

IL2_STAT5_box.plot + inflam_response_box.plot + Myc_box.plot
```

### Myeloid cells

#### GSEA with Hallmarks

```{r}
# Extract only myeloid cells
Mcell.seu <- subset(GSM6069857_merged.seu, subset = manually_annotated == 'Myeloid cell')
```

```{r}
# Find markers between pre- and post-treatment  
Idents(Mcell.seu) <- "HTO" # change the default identity
Mcell_pre_post.ge <- FindMarkers(Mcell.seu, ident.1 = 'CMO306', ident.2 = 'CMO305')
```

```{r}
# Manipulate the expression data
Mcell.ge <- cbind(gene = rownames(Mcell_pre_post.ge), Mcell_pre_post.ge)
lfc_vector <- Mcell.ge$avg_log2FC # Create a named vector ranked based on the log2 fold change
names(lfc_vector) <- Mcell.ge$gene
lfc_vector <- na.omit(lfc_vector) # Omit any NA values
lfc_vector <- sort(lfc_vector, decreasing = TRUE) # Sort the log2 fold change in descending order
```

```{r}
# Run GSEA (Hallmarks)
set.seed(2020)
HM_gse <- GSEA(
  geneList = lfc_vector, # Ordered ranked gene list
  minGSSize = 25, # Minimum gene set size
  maxGSSize = 500, # Maximum gene set set
  pvalueCutoff = 0.05, # p-value cutoff
  eps = 0, # Boundary for calculating the p value
  seed = TRUE, # Set seed to make results reproducible
  pAdjustMethod = "BH", # Benjamini-Hochberg correction
  TERM2GENE = dplyr::select(
    hs_hm_sets,
    gs_name,
    gene_symbol
  )
)
```

```{r}
# Collect the GSEA hallmarks result
Mcell_HmGSEA.result <- HM_gse@result
Mcell_HmGSEA.result
```

```{r fig.height=4.5, fig.width=18}
# GSEA plot for HALLMARK_ALLOGRAFT_REJECTION, HALLMARK_IL2_STAT5_SIGNALING, and HALLMARK_INFLAMMATORY_RESPONSE

allograft_reject_gsea.plot <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_ALLOGRAFT_REJECTION", 
                          geneSetID = "HALLMARK_ALLOGRAFT_REJECTION",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

IL2_STAT5_gsea.plot <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_IL2_STAT5_SIGNALING", 
                          geneSetID = "HALLMARK_IL2_STAT5_SIGNALING",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

inflam_response_gsea.plot <- gseaplot(HM_gse, 
                          by = "runningScore", 
                          title = "HALLMARK_INFLAMMATORY_RESPONSE", 
                          geneSetID = "HALLMARK_INFLAMMATORY_RESPONSE",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

allograft_reject_gsea.plot + IL2_STAT5_gsea.plot + inflam_response_gsea.plot
```

```{r}
# Measure the expression trends compared to the original results: HALLMARK_ALLOGRAFT_REJECTION, HALLMARK_IL2_STAT5_SIGNALING, and HALLMARK_INFLAMMATORY_RESPONSE

allograft_reject_genes <- hs_hm_sets[hs_hm_sets$gs_name == "HALLMARK_ALLOGRAFT_REJECTION", "gene_symbol"]
allograft_reject_genes_expression <- Mcell.ge[Mcell.ge$gene %in% allograft_reject_genes, ]

IL2_STAT5_genes <- hs_hm_sets[hs_hm_sets$gs_name == "HALLMARK_IL2_STAT5_SIGNALING", "gene_symbol"]
IL2_STAT5_genes_expression <- Mcell.ge[Mcell.ge$gene %in% IL2_STAT5_genes, ]

inflam_response_genes <- hs_hm_sets[hs_hm_sets$gs_name == "HALLMARK_INFLAMMATORY_RESPONSE", "gene_symbol"]
inflam_response_genes_expression <- Mcell.ge[Mcell.ge$gene %in% inflam_response_genes, ]
```

```{r}
# Box plot for allograft_reject_expression, IL2_STAT5_genes_expression, and inflam_response_genes_expression

allograft_reject_box.plot <- ggplot(allograft_reject_genes_expression, aes(y = avg_log2FC)) +
  geom_boxplot() + 
  labs(title = "Allograft_reject")

IL2_STAT5_box.plot <- ggplot(IL2_STAT5_genes_expression, aes(y = avg_log2FC)) +
  geom_boxplot() + 
  labs(title = "IL2_STAT5")

inflam_response_box.plot <- ggplot(inflam_response_genes_expression, aes(y = avg_log2FC)) +
  geom_boxplot() + 
  labs(title = "Inflam_Response")

allograft_reject_box.plot + IL2_STAT5_box.plot + inflam_response_box.plot
```

#### GSEA with GO:BP

```{r}
# GSEA in GO-BP term
set.seed(2020)
go_gsea <- GSEA(
  geneList = lfc_vector, # Ordered ranked gene list
  minGSSize = 25, # Minimum gene set size
  maxGSSize = 500, # Maximum gene set set
  pvalueCutoff = 0.05, # p-value cutoff
  eps = 0, # Boundary for calculating the p value
  seed = TRUE, # Set seed to make results reproducible
  pAdjustMethod = "BH", # Benjamini-Hochberg correction
  TERM2GENE = dplyr::select(
    hs_go_sets,
    gs_name,
    gene_symbol
  )
)
```

```{r}
# Collect the GSEA GO:BP result
Mcell_GoGSEA.result <- go_gsea@result
Mcell_GoGSEA.result
```

```{r fig.height=4, fig.width=15}
# GSEA plot for GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN and GOBP_MYELOID_LEUKOCYTE_MEDIATED_IMMUNITY

antigen_present_gsea.plot <- gseaplot(go_gsea, 
                          by = "runningScore", 
                          title = "GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN", 
                          geneSetID = "GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

myeloid_mediated_gsea.plot <- gseaplot(go_gsea, 
                          by = "runningScore", 
                          title = "GOBP_MYELOID_LEUKOCYTE_MEDIATED_IMMUNITY", 
                          geneSetID = "GOBP_MYELOID_LEUKOCYTE_MEDIATED_IMMUNITY",
                          base_size = 10) +
  scale_x_continuous(breaks = seq(0, 10000, by = 2000)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

antigen_present_gsea.plot + myeloid_mediated_gsea.plot 
```

```{r}
# Measure the expression trends compared to the original results: GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN and GOBP_MYELOID_LEUKOCYTE_MEDIATED_IMMUNITY

antigen_present_genes <- hs_go_sets[hs_go_sets$gs_name == "GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION_OF_PEPTIDE_ANTIGEN", "gene_symbol"]
antigen_present_genes_expression <- Mcell.ge[Mcell.ge$gene %in% antigen_present_genes, ]

myeloid_mediated_genes <- hs_go_sets[hs_go_sets$gs_name == "GOBP_MYELOID_LEUKOCYTE_MEDIATED_IMMUNITY", "gene_symbol"]
myeloid_mediated_genes_expression <- Mcell.ge[Mcell.ge$gene %in% myeloid_mediated_genes, ]
```

```{r}
# Box plot for antigen_present_expression and myeloid_mediated_expression

antigen_present_box.plot <- ggplot(antigen_present_genes_expression, aes(y = avg_log2FC)) +
  geom_boxplot() + 
  labs(title = "Antigen_present")

myeloid_mediated_box.plot <- ggplot(myeloid_mediated_genes_expression, aes(y = avg_log2FC)) +
  geom_boxplot() + 
  labs(title = "Myeloid_mediated")

antigen_present_box.plot + myeloid_mediated_box.plot 
```





